### 2017-09-26 ms.tex
temp push, going to remodel c

\documentclass[12pt]{article}
\usepackage[]{graphicx}
\usepackage[]{color}
%% use \xspace to allow for space after a macro as necessary
\usepackage{xspace}
\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
\usepackage{pdfpages}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{commentcol}{rgb}{0.345, 0.345, 0.945}
\newcommand{\comment}[1]{\textcolor{commentcol}{[[#1]]}}%

% general-purpose mathrm macros
\newcommand{\symsub}[2]{\ensuremath{#1_{\tiny \mathrm{#2}}}\xspace}
\newcommand{\mrm}[1]{\ensuremath{\mathrm{#1}}
\xspace}

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{randomcolor}{rgb}{.97, .27, .67}
\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[sort]{natbib}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage{hyperref}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{Fitting phylogenetic generalized linear mixed-effect models using lme4 }
\author{Michael Li and Ben Bolker}
\date{\today}

\providecommand{\keywords}[1]{\textbf{\textit{Keywords:}} #1}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}

%% don't typeset BMB comments
\newcommand{\bmbhide}[1]{}
\newcommand{\bmb}[1]{{\color{blue} BB: #1}}

\newcommand{\fref}[1]{Figure~\ref{fig:#1}}

\newcommand{\ml}[1]{{\color{red} ML: #1}}

\newcommand{\add}[1]{{\color{blue} ADD: #1}}

%\SweaveOpts{concordance=TRUE}
%\SweaveOpts{concordance=TRUE}
\maketitle

\doublespacing

\keywords{phyloglmm ... }

\section{Introduction}

Something about phylogenetic applications. This growing field has produced a number of statistical tools and software corde to implement them ... but this code is disparate and handles data differently, making routine data analyese challenging. ...

Here we present 
\section{Methods}

\end{document}

### 2017-09-26 dune_lme4.R

@@ -5,8 +5,8 @@ library(Matrix)
 library(lme4)
 library(dplyr)
 
-debug(phylo_lmm)
-debug(modify_phylo_retrms)
+# debug(phylo_lmm)
+# debug(modify_phylo_retrms)
 
 dd <- data.frame(dat)
 print(dd %>% count(site))
@@ -24,7 +24,7 @@ phyZ <- phylo.to.Z(phy)
 dat <- (dat
   %>% rowwise()
   %>% mutate(obs = sp
-    , standard_site = (as.numeric(site) - mean(1:20))/sd(1:20)
+    # , standard_site = (as.numeric(site) - mean(1:20))/sd(1:20)
   )
 )	
 
@@ -34,12 +34,12 @@ lme4time_1 <- system.time(
   lme4fit_1 <- phylo_lmm(Y ~ 1 + log.sla + annual 
 		+ (1|obs) 
 		+ (1|sp)
-    + (1 | site:sp)
+    + (1 | sp:site)
 		# + (0 + site|sp)
 		+ (0 + log.sla | site)
 		+ (1|site) 
 		, data=dat
-		, phylonm = c("sp","site:sp")
+		, phylonm = c("sp","sp:site")
 		, nsp = 28
 		, phylo = phy
 		, phyloZ=phyZ
@@ -47,15 +47,18 @@ lme4time_1 <- system.time(
   )
 )
 
+print(summary(lme4fit_1))
+
 
 lme4time_2 <- system.time(
   lme4fit_2 <- phylo_lmm(Y ~ 1 + log.sla + annual
-		# + (1|obs)
-		# + (1|sp)
-		+ (0 + site|sp)
-		# + (1|site)
+		+ (1|obs)
+		+ (1|sp)
+		+ (1 | sp:site)
+		# + (0 + site|sp)
+		+ (1|site)
 		, data=dat
-		, phylonm = "sp"
+		, phylonm = c("sp","sp:site")
 		, nsp = 28
 		, phylo = phy
 		, phyloZ=phyZ
@@ -135,3 +138,20 @@ print(peztime_2)
 print(summary(pezfit_2))
 print(lme4time_2)
 print(summary(lme4fit_2))
+
+
+lme4fit_3 <- phylo_lmm(Y ~ 1
+                       + (1|obs) 
+                       + (1|sp)
+                       + (1 | sp:site)
+                       # + (0 + site|sp)
+                       + (1|site) 
+                       , data=dat
+                       , phylonm = c("sp","sp:site")
[…]

### 2017-09-26 phyloglmm.R

@@ -14,8 +14,8 @@ dat <- (dat
 	%>% mutate(obs = sp)
 )	
 
-debug(phylo_lmm)
-debug(modify_phylo_retrms)
+# debug(phylo_lmm)
+# debug(modify_phylo_retrms)
 
 
 lme4timecor <- system.time(
@@ -34,18 +34,21 @@ lme4timecor <- system.time(
 )
 
 print(summary(lme4_interactionRE))
-# 
-# lme4_fullRE <- phylo_lmm(Y ~ site_name 
-#                                   # + (1|obs) 
-#                                   # + (1|sp) 
-#                                   + (0+site_name|sp) 
-#                                   # + (1|site_name:sp)
-#                                   , data=dat
-#                                   , phylonm = c("sp","site_name:sp")
-#                                   , phylo = phy
-#                                   , phyloZ=phyZ
-#                                   , nsp = nspp
-#                                   , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
-# )
-# 
-# print(summary(lme4_fullRE))
+
+lme4_fullRE <- phylo_lmm(Y ~ site_name 
+                                  # + (1|obs) 
+                                  # + (1|sp) 
+                                  + (0+site_name|sp) 
+                                  # + (1|site_name:sp)
+                                  , data=dat
+                                  , phylonm = c("sp","site_name:sp")
+                                  , phylo = phy
+                                  , phyloZ=phyZ
+                                  , nsp = nspp
+                                  , control=lmerControl(check.nobs.vs.nlev="ignore"
+                                                        , check.nobs.vs.nRE="ignore"
+                                                        , optCtrl=list(maxfun=1e5))
+                                
+)
+
+print(summary(lme4_fullRE))

### 2017-09-26 phyloglmm_setup.R

@@ -57,13 +57,15 @@ modify_phylo_retrms <- function(rt,phylo,phylonm,phyloZ,nsp){
     , function(i){
       (length(i)*(length(i)+1))/2
     })
-
+  Lind_list <- list()
 	 ### lFormula is creating the all RE index w.r.t nsp lengths into a simple vector, we have to use the function above to split properly
-	Lind_list <- split(rt[["Lind"]],rep(seq_along(Lind_split_length),Lind_split_length*nsp))
-  if(names(rt[["cnms"]][1]) == "sp:site_name"){ ### hack
-	  Lind_list[[1]] <- rep(1,9)
-    Lind_list[[2]] <- rep(2,3)
-  }
+	# Lind_list <- split(rt[["Lind"]],rep(seq_along(Lind_split_length),Lind_split_length*nsp))
+  # if(names(rt[["cnms"]][1]) == "sp:site_name"){ ### hack
+  
+  for(i in 1:length(rt$cnms)){
+	  Lind_list[[i]] <- rep(i,sum(rt$Lind==i))
+   }
+  # }
 	## Lambdat: replace block-diagonal element in Lambdat with a
 	## larger diagonal matrix
 	Lambdat_list <- split_blkMat(rt[["Lambdat"]],inds)
@@ -71,13 +73,14 @@ modify_phylo_retrms <- function(rt,phylo,phylonm,phyloZ,nsp){
 	for(i in phylo.pos){
 		## each sp is being rep w.r.t the complexity of RE in their respective Zt
 		repterms <- length(rt[["cnms"]][[i]])
-		if(names(rt[["cnms"]][i]) == "sp:site_name"){
-		  repterms <- 3 ### Hacked number, need to think about how to do this
+		if(names(rt[["cnms"]][i]) == "sp:site"){
+		  repterms <- 20 ### Hacked number, need to think about how to do this
 		  n.edge <- n.edge*repterms
 		}
 		## reconstitute Zt from new Ztlist
 		## We have to rep the same number of sp terms and edges in phyloZ to match Zt 
 		rt[["Ztlist"]][[i]] <- t(kronecker(phyloZ,diag(repterms)))%*% rt[["Ztlist"]][[i]]
+		## switch places inside kronecker
 		Gpdiff_new[i] <- n.edge  ## replace
 
 		## We have to create the Lind to match the theta field with larger reps w.r.t n.edges
[…]

### 2017-09-27 dune_lme4.R
need to clean up repo before moving forward

@@ -5,8 +5,8 @@ library(Matrix)
 library(lme4)
 library(dplyr)
 
-# debug(phylo_lmm)
-# debug(modify_phylo_retrms)
+debug(phylo_lmm)
+debug(modify_phylo_retrms)
 
 dd <- data.frame(dat)
 print(dd %>% count(site))
@@ -33,9 +33,8 @@ dat <- (dat
 lme4time_1 <- system.time(
   lme4fit_1 <- phylo_lmm(Y ~ 1 + log.sla + annual 
 		+ (1|obs) 
-		+ (1|sp)
-    + (1 | sp:site)
-		# + (0 + site|sp)
+		# + (1|sp)
+    # + (1 | sp:site)
 		+ (0 + log.sla | site)
 		+ (1|site) 
 		, data=dat
@@ -55,7 +54,6 @@ lme4time_2 <- system.time(
 		+ (1|obs)
 		+ (1|sp)
 		+ (1 | sp:site)
-		# + (0 + site|sp)
 		+ (1|site)
 		, data=dat
 		, phylonm = c("sp","sp:site")
@@ -96,16 +94,16 @@ peztime_1 <- system.time(
 	, sp = dat$sp
 	, site = dat$site
 	, random.effects = list(re.sp
-		, re.sp.phy
-		, re.nested.phy
+		# , re.sp.phy
+		# , re.nested.phy
 		, re.sla
 		, re.site
 	)
 	, REML = T
 	, verbose = F
 	, s2.init = c(1.5, rep(0.01, 3))
-	, reltol = 10e-20
-	, maxit = 10000
+	, reltol = 10e-10
+	, maxit = 1000
 	)
 )
 
@@ -123,8 +121,8 @@ peztime_2 <- system.time(
 	, REML = T
 	, verbose = F
 	, s2.init = c(1.5, rep(0.01, 3))
-	, reltol = 10e-20
-	, maxit = 10000
+	, reltol = 10e-10
+	, maxit = 1000
 	)
 )
 
@@ -139,19 +137,3 @@ print(summary(pezfit_2))
 print(lme4time_2)
 print(summary(lme4fit_2))
 
-
-lme4fit_3 <- phylo_lmm(Y ~ 1
-                       + (1|obs) 
-                       + (1|sp)
-                       + (1 | sp:site)
-                       # + (0 + site|sp)
-                       + (1|site) 
-                       , data=dat
-                       , phylonm = c("sp","sp:site")
-                       , nsp = 28
[…]

### 2017-09-27 phyloglmm.R

@@ -19,13 +19,13 @@ dat <- (dat
 
 
 lme4timecor <- system.time(
-	lme4_interactionRE <- phylo_lmm(Y ~ site_name 
+	lme4_interactionRE <- phylo_lmm(Y ~ site 
 		# + (1|obs) 
 		+ (1|sp) 
 		# + (0+site_name|sp) 
-		+ (1|sp:site_name)
+		+ (1|sp:site)
 		, data=dat
-		, phylonm = c("sp","sp:site_name")
+		, phylonm = c("sp","sp:site")
 		, phylo = phy
 		, phyloZ=phyZ
 		, nsp = nspp
@@ -35,20 +35,3 @@ lme4timecor <- system.time(
 
 print(summary(lme4_interactionRE))
 
-lme4_fullRE <- phylo_lmm(Y ~ site_name 
-                                  # + (1|obs) 
-                                  # + (1|sp) 
-                                  + (0+site_name|sp) 
-                                  # + (1|site_name:sp)
-                                  , data=dat
-                                  , phylonm = c("sp","site_name:sp")
-                                  , phylo = phy
-                                  , phyloZ=phyZ
-                                  , nsp = nspp
-                                  , control=lmerControl(check.nobs.vs.nlev="ignore"
-                                                        , check.nobs.vs.nRE="ignore"
-                                                        , optCtrl=list(maxfun=1e5))
-                                
-)
-
-print(summary(lme4_fullRE))

### 2017-09-27 phyloglmm_setup.R

@@ -42,6 +42,7 @@ modify_phylo_retrms <- function(rt,phylo,phylonm,phyloZ,nsp){
 	}
 	
 	## need to know number of number of speices to split index
+	## Fixme, the current index spliting is broken
 	if(is.null(nsp)){
 	nsp <- nrow(rt[["Lambdat"]])/length(rt[["cnms"]][[phylonm]]) 
 	}
@@ -58,6 +59,9 @@ modify_phylo_retrms <- function(rt,phylo,phylonm,phyloZ,nsp){
       (length(i)*(length(i)+1))/2
     })
   Lind_list <- list()
+  ##Fixme, manually hacked Lind_list with a for loop. This will break if we have 
+  ## terms on the left side of the bar/pipe
+  
 	 ### lFormula is creating the all RE index w.r.t nsp lengths into a simple vector, we have to use the function above to split properly
 	# Lind_list <- split(rt[["Lind"]],rep(seq_along(Lind_split_length),Lind_split_length*nsp))
   # if(names(rt[["cnms"]][1]) == "sp:site_name"){ ### hack
@@ -74,8 +78,8 @@ modify_phylo_retrms <- function(rt,phylo,phylonm,phyloZ,nsp){
 		## each sp is being rep w.r.t the complexity of RE in their respective Zt
 		repterms <- length(rt[["cnms"]][[i]])
 		if(names(rt[["cnms"]][i]) == "sp:site"){
-		  repterms <- 20 ### Hacked number, need to think about how to do this
-		  n.edge <- n.edge*repterms
+		  repterms <- 20 ### Hacked number sites, need to think about how to do this
+		  n.edge <- n.edge*repterms ## This is simply a term to create correct dim for Lind and Lambdat
 		}
 		## reconstitute Zt from new Ztlist
 		## We have to rep the same number of sp terms and edges in phyloZ to match Zt 
@@ -105,10 +109,10 @@ modify_phylo_retrms <- function(rt,phylo,phylonm,phyloZ,nsp){
 	rt[["Lambdat"]] <- Matrix::.bdiag(Lambdat_list)
 	## flist: Not sure how this part is being used.
 	rt[["flist"]] <- as.list(rt[["flist"]])
-	n.edge <- 4
-	for(i in 1:length(rt[["flist"]])){
-	  rt[["flist"]][i] <- factor(paste0("edge_",seq(n.edge)))
-	}
[…]

### 2017-09-27 simulate_lme4.R

@@ -1,0 +1,3 @@
+### Simulating via lme4 object
+
+bb <- simulate(aa,newparams=list(theta=c(2,0,0),beta=c(0,0,0)))

### 2017-10-17 ms.tex
need to clean up repo before moving forward
all cases are done, single case is not that stable

@@
 
-Something about phylogenetic applications.
-This growing field has produced a number of statistical tools and software corde to implement them ...
-but this code is disparate and handles data differently, making routine data analyese challenging.
-...
+Given a phylogenetic tree, phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with the evolutionary processes that generate species and their traits. 
+Ecologist have long considered the effects of functional traits and environmental conditions on pheotypic traits. 
+For example, regression analyses of body temperature as a function of body size for animal within a clade, species compositions in different sites, etc.
 
-Here we present 
 \section{Methods}
 
+The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
+
+An alternative approach is to model the phylogenetic correlation as a gaussian process.
+In particular, suppose that the evolutionary process is a Brownian motion.
+In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
+If we set up the Z matrix appropriately, we can model everythign with a sequence of independent errors, rather than having to do a fancy things to impose a correlation structure on the random effects.
+
 \end{document}

### 2017-10-17 lme4_simtest.R

@@ -29,7 +29,7 @@ for(i in 1:simnum){
 	sdcor_ss <- cov_ssdf$sdcor
 	sdcor_ms <- cov_msdf$sdcor
 	sd0vec_ss[i] <- sdcor_ss[1]
-	residvec_ss[i] <- sdcor_ss[4]
+	residvec_ss[i] <- sdcor_ss[2]
 	sd0vec_ms[i] <- sdcor_ms[1]
 	sd1vec_ms[i] <- sdcor_ms[2]
 	corvec_ms[i] <- sdcor_ms[3]

### 2017-10-17 parameters.R

@@ -1,8 +1,7 @@
 ### parameters for phylogenetic tree simulations
-seed <- 1000
-simnum <- 100
+simnum <- 300
 # Generate simulated data for nspp species and nsite sites
-nspp <- 200
+nspp <- 300
 
 nsite <- 1
 

### 2017-10-17 phyloglmm.R

@@ -22,7 +22,9 @@ print(head(dat))
 
 lme4time <- system.time(
 	lme4fit <- phylo_lmm(Y ~ X
-		+ (1 + X|sp)
+		+ (1|sp)
+		+ (0 + X|sp)
+#		+ (1+X|sp)
 		, data=dat
 		, phylonm = c("sp","sp:site")
 		, phylo = phy

### 2017-10-17 simulate_tree.R

@@ -47,6 +47,11 @@ b1 <- b.all[(nspp+1):(2*nspp)]
 
 y <- matrix(outer(b0, array(1, dim = c(1, nsite))), nrow = nspp,
             ncol = nsite) + matrix(outer(b1, X), nrow = nspp, ncol = nsite)
+
+if(nsite == 1){
+y <- matrix(outer(b0, array(1, dim = c(1, nsite))), nrow = nspp,
+	ncol = nsite) + matrix(b1*X, nrow = nspp, ncol = nsite)
+}
 e <- rnorm(nspp * nsite, sd = sd.resid) # add residual variance 
 y <- y + matrix(e, nrow = nspp, ncol = nsite)
 y <- matrix(y, nrow = nspp * nsite, ncol = 1)
@@ -62,8 +67,12 @@ colnames(Y) <- 1:nsite
 # Transform data matrices into "long" form, and generate a data frame
 YY <- matrix(Y, nrow = nspp * nsite, ncol = 1)
 
-#XX <- matrix(kronecker(X, matrix(1, nrow = nspp, ncol = 1)), nrow =
-#               nspp * nsite, ncol = 1)
+XX <- matrix(kronecker(X, matrix(1, nrow = nspp, ncol = 1)), nrow =
+               nspp * nsite, ncol = 1)
+
+if(nsite == 1){
+	XX <- matrix(X,nrow=nspp,ncol=1)
+}
 
 site <- matrix(kronecker(1:nsite, matrix(1, nrow = nspp, ncol =
                                            1)), nrow = nspp * nsite, ncol = 1)
@@ -73,5 +82,4 @@ sp <- matrix(kronecker(matrix(1, nrow = nsite, ncol = 1), 1:nspp),
 dat <- data.frame(Y = YY, X, site = as.factor(site), sp = as.factor(sp),site_name = rep(site_name,each=nspp))
 
 
-print(head(dat))
 

### 2017-10-30 ms.tex
pushing random text and updating parameters for single site simulation test
pushing draft looking THING

@@
 \section{Introduction}
+\ml{I will fill in appropriate citation later}
 
-Given a phylogenetic tree, phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with the evolutionary processes that generate species and their traits. 
-Ecologist have long considered the effects of functional traits and environmental conditions on pheotypic traits. 
-For example, regression analyses of body temperature as a function of body size for animal within a clade, species compositions in different sites, etc.
+Given a phylogenetic tree, phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes that generate species and exhibit in their traits.
+Ecologist have long considered the effects of species functional traits and environmental conditions on their phenotypic traits and also incorportating historical relationships of species.
+Decades of work have created various different methods that enables researchers to analyze species relativeness problems/systems for different types of data and questions.
+%For example, regression analyses of body temperature as a function of body size for animal within a clade, species compositions in different sites, etc. 
+\ml{more examples...and reference}
+
+The most widely used statistical method to model/fit related species data that accounts for phylogenetic structure is the phylogenetic regression (Felsenstein 1985). 
+More recently, people use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
+
+Despite the choice of modeling strageties, there isn't a convient platform to do all and limited in model complexities additional complexities in the systems.
+
+In this paper, we model it using lme4 and test it with simulated cases across platforms and fit it to dun (maybe more examples if we can find it) example.
+
+We also compare different platforms.
 
@@
 
+We generate test data using a simple glmm framework that varies in complexity in random effect.
+
+\section{Phylogenetic Tree Transformation}
 The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
@@
 
+\subsection{Platforms}
+
+...
+
+\section{Results}
+
+\section{Discussion}
+
+\section{Conclusion}
+
 \end{document}

### 2017-10-31 ms.tex

@@
 Decades of work have created various different methods that enables researchers to analyze species relativeness problems/systems for different types of data and questions.
+But many challenges remain.
+In particular, incorportating phylogenetic signals in complex/large/multi dimensional systems is always challenging, especially with large phylogenies where computation powers are limited.
+In such case, researchers often use simpler models with minimal phylogenetic signals limiting at the species/individuals level and neglect additional phylogenetic signals present in the system/model.
+
 %For example, regression analyses of body temperature as a function of body size for animal within a clade, species compositions in different sites, etc. 
-\ml{more examples...and reference}
-
 The most widely used statistical method to model/fit related species data that accounts for phylogenetic structure is the phylogenetic regression (Felsenstein 1985). 
-More recently, people use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
-
+More recently, researchers use (turn to?) linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures. 
+\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here.}
 Despite the choice of modeling strageties, there isn't a convient platform to do all and limited in model complexities additional complexities in the systems.
 
-In this paper, we model it using lme4 and test it with simulated cases across platforms and fit it to dun (maybe more examples if we can find it) example.
-
+In this paper, we model it using lme4 and test it with simulated cases and fit it to dun (maybe more examples if we can find it) example. 
 We also compare different platforms.
 
-\section{Methods}
+\section{Materials and Methods}
 
-We generate test data using a simple glmm framework that varies in complexity in random effect.
+We generate test data using a simple glmm framework that varies in complexity in random effect. All simulations and fits were performed with R.
 
-\section{Phylogenetic Tree Transformation}
+\subsection{Simulations}
+
+We simulated two phenotypic traits. Trait $Y$ of species $i$ ($i$ = 1,...,$n$) that depends on a trait X using the model:
+
+\ml{Y~X + (1+X|phylo) and the 3 forms in ols/gls} 
+\begin{align}
+Y & = \alpha_{0} + \alpha_{1}X + \beta (1+X) + \epsilon 
+\\
+\beta & \sim \mathrm{MVN}(0,\Sigma)
+\\
+\epsilon & \sim \mathrm{N}(0, \sigma^2_{n})
+\end{align}
+
+where
+
+\begin{align}
+\Sigma & = \begin{pmatrix} \sigma^2_{sp} & \rho \sigma_{sp} \sigma_{X} \\ \rho \sigma_{sp} \sigma_{X} & \sigma^2_{X} \end{pmatrix}
+\end{align}
+We assumed both phenotypic trait are measured. X (e.g. ...) affects
+
+\ml{metabolism example?}
+
+\ml{ditto above with multiple sites}
+\subsection{Phylogenetic Tree Transformation}
 The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
@@
 
-\subsection{Platforms}
+\subsection{Model Fitting}
+\ml{platform/model equivalence}
 

### 2017-11-07 ms.tex
master equation

@@
 \bibliographystyle{chicago}
-\title{Fitting phylogenetic generalized linear mixed-effect models using lme4 }
+\title{Fitting phylogenetic generalized linear mixed models}
 \author{Michael Li and Ben Bolker}
@@
 
-We simulated two phenotypic traits. Trait $Y$ of species $i$ ($i$ = 1,...,$n$) that depends on a trait X using the model:
-
-\ml{Y~X + (1+X|phylo) and the 3 forms in ols/gls} 
-\begin{align}
-Y & = \alpha_{0} + \alpha_{1}X + \beta (1+X) + \epsilon 
-\\
-\beta & \sim \mathrm{MVN}(0,\Sigma)
-\\
-\epsilon & \sim \mathrm{N}(0, \sigma^2_{n})
-\end{align}
-
-where
+%We simulated two phenotypic traits (X andY). 
+The typical phylogenetic linear mixed model analysis is of the form:
 
 \begin{align}
-\Sigma & = \begin{pmatrix} \sigma^2_{sp} & \rho \sigma_{sp} \sigma_{X} \\ \rho \sigma_{sp} \sigma_{X} & \sigma^2_{X} \end{pmatrix}
+(Y \mid B=b) & \sim N(X\beta + Zb, \sigma^{2}),
 \end{align}
-We assumed both phenotypic trait are measured. X (e.g. ...) affects
+where the conditional distribution of Y given $ B = b$ is a $n \times 1$ vector random response variable, whose observed value is y of some measureable phenotypic trait; 
+X is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independent variables (traits);
+Z is an $n \times (m + 1)$ model matrix for the $m + 1$ -- dimensional vector-valued random effects variable; 
+$\beta$ is a $ m + 1 $ coefficient vector.
+The parameters of the model are the coefficients $\beta$ and the scale parameter $\sigma$.
+The unconditional distribution of B is multivariate normal with mean zero and a parameterized $(m + 1) \times (m + 1)$ variance-covariance matrix
+\begin{align}
+B & \sim \mathrm{MVN}(0,\Sigma),
+\end{align}
+and
+\begin{align}
+\Sigma & = \begin{pmatrix} \sigma^2_{sp} & \rho \sigma_{sp} \sigma_{X} \\ \rho \sigma_{sp} \sigma_{X} & \sigma^2_{X} \end{pmatrix},
+\end{align}
+where $\sigma^{2}_{sp}$ and $\sigma^{2}_{X}$ is the phylogenetic variance of species and phenotype X; 
+and $\rho$ is the correlation of \ml{...?}.
 
-\ml{metabolism example?}
-
-\ml{ditto above with multiple sites}
 \subsection{Phylogenetic Tree Transformation}

### 2017-11-13 ms.tex
temp push

@@
 \bibliographystyle{chicago}
-\title{Fitting phylogenetic generalized linear mixed models}
+\title{Need a good title.}
 \author{Michael Li and Ben Bolker}
@@
 The most widely used statistical method to model/fit related species data that accounts for phylogenetic structure is the phylogenetic regression (Felsenstein 1985).
-More recently, researchers use (turn to?) linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures. 
-\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here.}
-Despite the choice of modeling strageties, there isn't a convient platform to do all and limited in model complexities additional complexities in the systems.
+The typical phylogenetic regression is of the form:
+\begin{align}
+y & = XB + \epsilon \\
+\epsilon & \sim MVN(0,\sigma^{2}C),
+\end{align}
+where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m \times 1$ coefficient vector; $\espilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{ \epsilon} C$ in which $C$ is the phylogenetic correlation (PC).
+\ml{Need to say something like $\epsilon$ explains all the phylogenetic corelation and error or zero residual}
 
-In this paper, we model it using lme4 and test it with simulated cases and fit it to dun (maybe more examples if we can find it) example. 
-We also compare different platforms.
+More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
+\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
+An alternative modelling approach is to use linear mixed effects modelling framework.
+The typical phylogenetic linear mixed model has the form:
+\begin{align}
+Y & \sim N(XB + Zb, \sigma^{2}),
+\end{align}
+Despite the choice of modeling strageties, there isn't a convient platform to do all and limited in model complexities additional complexities in the systems and . 
+\ml{This sounds cheesy}
+
+In this paper, we took a simple, yet effective alternative approach model to model phylogenetic correlation where we ...
+.
+We used lme4 and test it with simulated cases and fit it to dun (maybe more examples if we can find it) example.
+We will compare our simulations and example fits to different platforms.
 

### 2017-11-14 ms.tex
intro is coming along nicely

@@
 \epsilon & \sim MVN(0,\sigma^{2}C),
+\label{eq:gls}
 \end{align}
-where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m \times 1$ coefficient vector; $\espilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{ \epsilon} C$ in which $C$ is the phylogenetic correlation (PC).
+where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m + 1$ coefficient vector; $\espilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{ \epsilon} C$ in which $C$ is the phylogenetic correlation (PC).
 \ml{Need to say something like $\epsilon$ explains all the phylogenetic corelation and error or zero residual}
 
-More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
-\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
+% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
+%\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
 An alternative modelling approach is to use linear mixed effects modelling framework.
-The typical phylogenetic linear mixed model has the form:
+The typical linear mixed model has the form:
 \begin{align}
-Y & \sim N(XB + Zb, \sigma^{2}),
+Y & = XB + Zb + \epsilon_{mm}
 \end{align}
-Despite the choice of modeling strageties, there isn't a convient platform to do all and limited in model complexities additional complexities in the systems and . 
-\ml{This sounds cheesy}
+where $Z$ is an $n \times (m+1)$ model matrix for the $(m+1)$ -- dimensional vector-valued $m$ independment variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
+Analogously, the phylogenetic regression given by \ref{eq:ls} can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$ and $Z=Z^{c}$ where the PC is incorportated in the random effect matrix.
+
+Recent years, researchers have begun to adpot these phylogenetic-based methods in various applications. 
+Despite the choice of modeling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis \ml{add examples .. model selection and etc}.
+There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematic studies of the comparative performance of comparative phylogenetic methods.
 

### 2017-11-14 ms.tex
the intro is starting to come along

@@
 Ecologist have long considered the effects of species functional traits and environmental conditions on their phenotypic traits and also incorportating historical relationships of species.
-Decades of work have created various different methods that enables researchers to analyze species relativeness problems/systems for different types of data and questions.
+Decades of work have created various different methods that enables researchers to analyze species relativeness problems for different types of data and questions.
 But many challenges remain.
-In particular, incorportating phylogenetic signals in complex/large/multi dimensional systems is always challenging, especially with large phylogenies where computation powers are limited.
-In such case, researchers often use simpler models with minimal phylogenetic signals limiting at the species/individuals level and neglect additional phylogenetic signals present in the system/model.
+In particular, incorportating phylogenetic signals in high dimensional systems is always challenging, especially with large phylogenies where computation powers are limited.
+In such case, researchers often use simpler models with minimal phylogenetic signals limiting at the individuals level and neglect additional phylogenetic signals present in the system.
 
 %For example, regression analyses of body temperature as a function of body size for animal within a clade, species compositions in different sites, etc. 
-The most widely used statistical method to model/fit related species data that accounts for phylogenetic structure is the phylogenetic regression (Felsenstein 1985).
+The most widely used statistical method to model related species data that accounts for phylogenetic structure is the phylogenetic regression (Felsenstein 1985).
 The typical phylogenetic regression is of the form:
@@
 where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m + 1$ coefficient vector; $\espilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{ \epsilon} C$ in which $C$ is the phylogenetic correlation (PC).
-\ml{Need to say something like $\epsilon$ explains all the phylogenetic corelation and error or zero residual}
 
@@
 \begin{align}
-Y & = XB + Zb + \epsilon_{mm}
+Y & \sim F(\mu) \\
+g(\mu) & = XB + Zb + \epsilon_{mm} \\
+\epsilon_{mm} & \sim MVN(0,\sigma^2)
 \end{align}
 where $Z$ is an $n \times (m+1)$ model matrix for the $(m+1)$ -- dimensional vector-valued $m$ independment variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
-Analogously, the phylogenetic regression given by \ref{eq:ls} can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$ and $Z=Z^{c}$ where the PC is incorportated in the random effect matrix.
+Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$ and $Z=Z^{c}$ where the PC is incorportated in the random effect matrix.
+\ml{where can I add a line about independent contrast? maybe in the discussion?}
 
 Recent years, researchers have begun to adpot these phylogenetic-based methods in various applications. 
-Despite the choice of modeling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis \ml{add examples .. model selection and etc}.
-There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematic studies of the comparative performance of comparative phylogenetic methods.
+Despite the choice of modeling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis (i.e model selection and etc). 
+There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematically look at comparative performance of comparative phylogenetic methods.
 
-In this paper, we took a simple, yet effective alternative approach model to model phylogenetic correlation where we ...
-.
-We used lme4 and test it with simulated cases and fit it to dun (maybe more examples if we can find it) example.
-We will compare our simulations and example fits to different platforms.
+In this paper, we propose an alternative, yet straightforward but little-used method of constructing $Z^C$ directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
+We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorportate various complexity of phylogenetic signals in the model. 
+We compare model approaches and explore various complex modelling strategies in the mixed model framework and a flexible platform for researches to explore biologically interesting questions.
+
+We also compare three different R packages: nlme, pez, and lme4. 
+\ml{Not sure if we should include the line "we are not using Bayesian approaches" here but defintely in the discussions.}
 

### 2017-11-30 ms.tex
need a quick chat

@@
 
-%For example, regression analyses of body temperature as a function of body size for animal within a clade, species compositions in different sites, etc. 
-The most widely used statistical method to model related species data that accounts for phylogenetic structure is the phylogenetic regression (Felsenstein 1985).
+Recent years, researchers have begun to incorportate phylogenetic-based methods in various modelling applications \ml{cite}.
+The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.
+Despite the choice of modelling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis.
+There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematically look at comparative performance of comparative phylogenetic methods.
+
+In this paper, we propose an alternative, yet straightforward but little-used method of constructing phylogenetic correlations directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
+We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorportate various complexity of phylogenetic signals in the model. 
+We compare model approaches and explore various complex modelling strategies in the mixed model framework and a flexible platform for researches to explore biologically interesting questions.
+
+We also compare three different R packages: nlme, pez, and lme4. 
+\ml{Not sure if we should include the line "we are not using Bayesian approaches" here but defintely in the discussions.}
+
+\section{Materials and Methods}
+
 The typical phylogenetic regression is of the form:
@@
 Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$ and $Z=Z^{c}$ where the PC is incorportated in the random effect matrix.
-\ml{where can I add a line about independent contrast? maybe in the discussion?}
 
-Recent years, researchers have begun to adpot these phylogenetic-based methods in various applications. 
-Despite the choice of modeling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis (i.e model selection and etc). 
-There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematically look at comparative performance of comparative phylogenetic methods.
+\subsection{Phylogenetic Tree Transformation}
+The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
+Phylogenetic independent constrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
+Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.
 
-In this paper, we propose an alternative, yet straightforward but little-used method of constructing $Z^C$ directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
-We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorportate various complexity of phylogenetic signals in the model. 
-We compare model approaches and explore various complex modelling strategies in the mixed model framework and a flexible platform for researches to explore biologically interesting questions.
-
-We also compare three different R packages: nlme, pez, and lme4. 
-\ml{Not sure if we should include the line "we are not using Bayesian approaches" here but defintely in the discussions.}
-
-\section{Materials and Methods}
-
-We generate test data using a simple glmm framework that varies in complexity in random effect. All simulations and fits were performed with R.
+An alternative approach is to model the phylogenetic correlation as a \textit{gaussian process}. 
+In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk and daughter species after speciation are all independent.  
+In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
+Thus, the evolutionary history for each species can be model with a sequence of independent errors, rather than having to impose a correlation structure on the random effects.
 
@@
 
-%We simulated two phenotypic traits (X andY). 
-The typical phylogenetic linear mixed model analysis is of the form:
-
-\begin{align}
-(Y \mid B=b) & \sim N(X\beta + Zb, \sigma^{2}),
-\end{align}
-where the conditional distribution of Y given $ B = b$ is a $n \times 1$ vector random response variable, whose observed value is y of some measureable phenotypic trait; 
-X is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independent variables (traits);
-Z is an $n \times (m + 1)$ model matrix for the $m + 1$ -- dimensional vector-valued random effects variable; 
-$\beta$ is a $ m + 1 $ coefficient vector.
-The parameters of the model are the coefficients $\beta$ and the scale parameter $\sigma$.
-The unconditional distribution of B is multivariate normal with mean zero and a parameterized $(m + 1) \times (m + 1)$ variance-covariance matrix
-\begin{align}
-B & \sim \mathrm{MVN}(0,\Sigma),
-\end{align}
-and
-\begin{align}
-\Sigma & = \begin{pmatrix} \sigma^2_{sp} & \rho \sigma_{sp} \sigma_{X} \\ \rho \sigma_{sp} \sigma_{X} & \sigma^2_{X} \end{pmatrix},
-\end{align}
-where $\sigma^{2}_{sp}$ and $\sigma^{2}_{X}$ is the phylogenetic variance of species and phenotype X; 
-and $\rho$ is the correlation of \ml{...?}.
-
-\subsection{Phylogenetic Tree Transformation}
-The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
-
-An alternative approach is to model the phylogenetic correlation as a gaussian process. 
-In particular, suppose that the evolutionary process is a Brownian motion. 
-In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
-If we set up the Z matrix appropriately, we can model everythign with a sequence of independent errors, rather than having to do a fancy things to impose a correlation structure on the random effects.
+We generate test data using a simple glmm framework that varies in complexity in random effect. All simulations and fits were performed with R.
 

### 2017-11-30 ms.tex
need to think about dimension problems

@@
 \end{align}
-where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m + 1$ coefficient vector; $\espilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{ \epsilon} C$ in which $C$ is the phylogenetic correlation (PC).
+where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m + 1$ coefficient vector; $\espilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{ \epsilon} C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
 

### 2017-12-08 ms.tex
temp pushing ms

@@
 
+\ml{$Z^{C}$ = $phyZ^{T}$ * Z}
+
+The random-effect model matrix $Z$ can be decomposed into term-wise model matrix $Z_{i}$ as described in \ml{in text citation lme4}.
+Thus, the phylogenetic correlated random-effect matrix $Z^{C}_{i}$ is
+
+\begin{align}
+Z^{C} & = BS Z ,
+\end{align}
+
+where $BS$ is an $b \times s$ matrix of branch--species relation.
+
+\ml{for correlated intercept and slope it is $Z^{C} = Kron(BS,diag(repterm))^T * Z$}
+
 \subsection{Simulations}

### 2017-12-08 fit_phyloglmm.R

@@ -50,6 +50,7 @@ if(platform == "pez"){
 if(platform == "lme4"){
 	fittime <- system.time(
 		fitmod <- phylo_lmm(Y ~ X
+		  + (1 |obs)
 			+ (1 + X|sp)
 			, data=dat
 			, phylonm = c("sp","sp:site")

### 2017-12-08 ms.tex
pushing table

@@
 
-\begin{align}
-Z^{C} & = BS Z ,
-\end{align}
+\begin{equation}
+Z^{C}_{i} = (BS \bigotimes I_{q_i \times s})^T Z_{i} ,
+\end{equation}
 
-where $BS$ is an $b \times s$ matrix of branch--species relation.
-
-\ml{for correlated intercept and slope it is $Z^{C} = Kron(BS,diag(repterm))^T * Z$}
+where $BS$ is an $b \times s$ matrix of branch--species relations; $I$ is an identity matrix size $q_i \times s$; $q_i$ is the number of columns contributed by $Z_i$; s is the number of sites; and  $\bigotimes$ is the Kronecker product \ml{cite kronecker}.
 
@@
 
-We generate test data using a simple glmm framework that varies in complexity in random effect. All simulations and fits were performed with R.
+\begin{tabular}{|c|c|c|}
+\hline
+Type of RE	& Formula	& Platform	\\
+\hline
+Single site intercept										&	1 \mid sp						& gls, lme4 \\
+Single site intercept and uncorrelated slopes		&  1 \mid sp + X \mid sp		& lme4 \\
+Single site intercept and correlated slopes 			& 1 + X \mid sp					& lme4 \\
+\hline
+Multiple site intercept 									&  1 \mid sp 						& pez, lme4 \\
+Multiple site intercept and uncorrelated slopes 	&  1 \mid sp + X \mid sp 		& pez, lme4 \\
+Multiple site intercept and uncorrelated slopes 	& 	1 + X \mid sp					& lme4 \\
+ditto with site interaction								& ditto above \mid sp:site		& ditto above \\
+\hline
+
+\end{tabular}
+\\
+
+We generate test data using a simple glmm framework that varies in complexity in random effect described in Table 1 \ml{macro this}. All simulations and fits were performed with R.
 

### 2017-12-08 ms.tex
fixed formula 6 in ms

@@
 
-\ml{$Z^{C}$ = $phyZ^{T}$ * Z}
-
 The random-effect model matrix $Z$ can be decomposed into term-wise model matrix $Z_{i}$ as described in \ml{in text citation lme4}.
@@
 \begin{equation}
-Z^{C}_{i} = (BS \bigotimes I_{q_i \times s})^T Z_{i} ,
+Z^{C}_{i} = Z_{i}(SB \bigotimes I_{p_i \times s}) ,
 \end{equation}
 
-where $BS$ is an $b \times s$ matrix of branch--species relations; $I$ is an identity matrix size $q_i \times s$; $q_i$ is the number of columns contributed by $Z_i$; s is the number of sites; and  $\bigotimes$ is the Kronecker product \ml{cite kronecker}.
+\ml{double check if $l and p$ are the same with lme4}
+
+where $SB$ is an $l \times b$ matrix of species--branch relations; $I$ is an identity matrix size $p_i \times s$; $p_i$ is the number of columns contributed by $Z_i$; s is the number of sites; and  $\bigotimes$ is the Kronecker product \ml{cite kronecker}.
 

### 2017-12-08 fit_phyloglmm.R

@@ -50,7 +50,7 @@ if(platform == "pez"){
 if(platform == "lme4"){
 	fittime <- system.time(
 		fitmod <- phylo_lmm(Y ~ X
-		  + (1 |obs)
+		  + (1 |sp)
 			+ (1 + X|sp)
 			, data=dat
 			, phylonm = c("sp","sp:site")

### 2018-02-14 ms.tex
fixed formulas in the ms

--- 2017-12-08 ms.tex
@@
 \end{align}
-where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m + 1$ coefficient vector; $\espilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{ \epsilon} C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
+where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m + 1$ coefficient vector; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
 
@@
 
-\begin{align}
-Z^{C} & = BS Z ,
-\end{align}
+\begin{equation}
+Z^{C}_{i} = (BS^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T},
+\end{equation}
 
-where $BS$ is an $b \times s$ matrix of branch--species relation.
+\ml{double check if $l and p$ are the same with lme4}
 
-\ml{for correlated intercept and slope it is $Z^{C} = Kron(BS,diag(repterm))^T * Z$}
+where $\ast$ is the Khatri-Rao product, $BS_{i}$ is an $l_{i} \times b_{i}$ matrix of species--branch relations; $J_{i}$ the indicator matrix of grouping factors indices matrix size $n \times l_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
+
+\ml{double kronecker for interactions}
 
@@
 
-We generate test data using a simple glmm framework that varies in complexity in random effect. All simulations and fits were performed with R.
+\begin{tabular}{|c|c|c|}
+\hline
+Type of RE	& Formula	& Platform	\\
+\hline
+
+Single site intercept	  &	1 $\mid$ sp	& gls, lme4 \\
+Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + X $\mid$ sp		& lme4 \\
+Single site intercept and correlated slopes 			& 1 + X $\mid$ sp					& lme4 \\
+\hline
+Multiple site intercept 									&  1 $\mid$ sp 						& pez, lme4 \\
+Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + X $\mid$ sp 		& pez, lme4 \\
+Multiple site intercept and uncorrelated slopes 	& 	1 + X $\mid$ sp					& lme4 \\
+ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
+\hline
+% \vspace{1in}
+\end{tabular}
+
+We generate test data using a simple glmm framework that varies in complexity in random effect described in Table 1 \ml{macro this}. All simulations and fits were performed with R.
 

### 2018-02-14 collect.R

@@ -170,7 +170,7 @@ lme4cs_results <- function(pp,tt){
   return(lme4cs_list)
 }
 
-#lme4cs_data <- bind_rows(lme4cs_results(lme4cs_path,lme4cs_res))
+lme4cs_data <- bind_rows(lme4cs_results(lme4cs_path,lme4cs_res))
 
 #### Collect pez cs results ----
 
@@ -201,7 +201,7 @@ pez_cs_results <- function(pp,tt){
   return(pez_list)
 }
 
-#pez_cs_data <- bind_rows(pez_cs_results(pez_cs_path,pez_res))
+pez_cs_data <- bind_rows(pez_cs_results(pez_cs_path,pez_res))
 
 #### lme4 cs slopes NOT DONE, MISSING uncor ----
 
@@ -265,22 +265,21 @@ lme4cs_slopeuncor_results <- function(pp,tt){
   return(lme4cs_list)
 }
 
-#lme4cs_slopecor_data <- bind_rows(lme4cs_slopecor_results(lme4cs_slope_path,lme4cs_slopecor_res))
-#lme4cs_slopeuncor_data <- bind_rows(lme4cs_slopeuncor_results(lme4cs_slope_path,lme4cs_slopeuncor_res))
+lme4cs_slopecor_data <- bind_rows(lme4cs_slopecor_results(lme4cs_slope_path,lme4cs_slopecor_res))
+lme4cs_slopeuncor_data <- bind_rows(lme4cs_slopeuncor_results(lme4cs_slope_path,lme4cs_slopeuncor_res))
 
-#lme4cs_slope_data <- rbind(lme4cs_slopecor_data,lme4cs_slopeuncor_data)
+lme4cs_slope_data <- rbind(lme4cs_slopecor_data,lme4cs_slopeuncor_data)
 #### Collect pez cs slope ----
 ## phylo and phyloX out of order?
 pez_cs_slope_path <- "./datadir/pez/pez_cs/slope/"
 
 pez_cs_slope_res <- list.files(pez_cs_slope_path,pattern="uncor")
 
-#pez_cs_slope_data <- bind_rows(pez_cs_results(pez_cs_slope_path,pez_cs_slope_res))
+pez_cs_slope_data <- bind_rows(pez_cs_results(pez_cs_slope_path,pez_cs_slope_res))
 
 
 #### Save results ----
 
[…]

### 2018-02-14 plot.R

@@ -3,7 +3,9 @@
 library(dplyr)
 library(ggplot2)
 library(tidyr)
+library(grid)
 theme_set(theme_bw())
+zmargin <- theme(panel.margin=grid::unit(0,"lines"))
 
 data_list <- readRDS("./datadir/result_list.RDS")
 
@@ -14,10 +16,10 @@ lme4ms_data <- data_list[[4]]
 pez_data <- data_list[[5]]
 lme4ms_slope_data <- data_list[[6]]
 pez_slope_data <- data_list[[7]]
-#lme4cs_data <- data_list[[8]]
-#pez_cs_data <- data_list[[9]]
-#lme4cs_slope_data <- data_list[[10]]
-#pez_cs_slope_data <- data_list[[11]]
+lme4cs_data <- data_list[[8]]
+pez_cs_data <- data_list[[9]]
+lme4cs_slope_data <- data_list[[10]]
+pez_cs_slope_data <- data_list[[11]]
 
 
 gls_data2 <- (gls_data
@@ -75,6 +77,7 @@ gg_ss <- (ggplot(data=ss_data,aes(x=size,y=sd,fill=Platform))
   + scale_y_log10(limits=c(0.5,50),breaks=c(1,4,10,50))
   + ggtitle("Single Site using GLS and lme4")
   + ylab("Standard Deviation")
+  + zmargin
 )
 
 print(gg_ss)
@@ -108,6 +111,7 @@ gg_ss_slope <- (ggplot(data=ss_slope_data,aes(x=size,y=sd))
   + scale_y_log10(limits=c(0.5,20),breaks=c(0.1,1,2,4,10,15))
   + ylab("Standard Deviation")
   + ggtitle("LME4 Single site slope (Need to draw a separate correlation plot")
+  + zmargin
 )
 
 print(gg_ss_slope)
@@ -137,6 +141,7 @@ gg_ms <- (ggplot(data=ms_data,aes(x=size,y=sd,fill=Platform))
   + scale_y_continuous(limits=c(0,12),breaks=c(0,2,4,6,8,10,12))
   + ggtitle("Multiple Site using pez and lme4")
   + ylab("Standard Deviation")
+  + zmargin
 )
 
 print(gg_ms)
@@ -177,6 +182,7 @@ gg_ms_slope <- (ggplot(data=ms_slope_data,aes(x=size,y=sd,fill=Platform))
   + geom_hline(data=ss_slope_dummy,aes(yintercept = Y))
[…]

### 2018-02-14 ms.tex
temp pushing before meeting

@@
 
+% \begin{table}
+% \begin{center}
 \begin{tabular}{|c|c|c|}
@@
 Single site intercept	  &	1 $\mid$ sp	& gls, lme4 \\
-Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + X $\mid$ sp		& lme4 \\
-Single site intercept and correlated slopes 			& 1 + X $\mid$ sp					& lme4 \\
+Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& lme4 \\
+Single site intercept and correlated slopes 			& 1 + trait $\mid$ sp					& lme4 \\
 \hline
 Multiple site intercept 									&  1 $\mid$ sp 						& pez, lme4 \\
-Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + X $\mid$ sp 		& pez, lme4 \\
-Multiple site intercept and uncorrelated slopes 	& 	1 + X $\mid$ sp					& lme4 \\
-ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
+Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + trait $\mid$ sp 		& pez, lme4 \\
+Multiple site intercept and correlated slopes 	& 	1 + trait $\mid$ sp					& lme4 \\
 \hline
+\end{tabular}
 % \vspace{1in}
-\end{tabular}
+% ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
+% \end{center}
+% \end{table}
 
-We generate test data using a simple glmm framework that varies in complexity in random effect described in Table 1 \ml{macro this}. All simulations and fits were performed with R.
+The simulation models is based on glmm framework described in \ml{eq labels 3-5}.   
+To construct the simulation model, let $n$ be the number of species distributed among $m$ sites.
+Let $Y$ be the $mn \times 1$ vectors of response trait.  
+We generate test data using three different $n$ (20, 100, 500) and two different $m$ (1, 20) using a simple glmm framework that varies in complexity in random effect described in \ml{Table 1, macro this with labels}.
+We also extend the multiple site models described in \ml{Table 1} with species by site interaction random effects. 
+All simulations and fits were performed with R.
 
 \subsection{Model Fitting}
-\ml{platform/model equivalence}
-
-...
 

### 2018-02-14 plot.R

@@ -192,6 +192,7 @@ gg_ms_slope_time <- (ggplot(data=ms_slope_data,aes(x=size,y=time,fill=platform))
               + geom_violin(position=position_dodge(width=0),alpha=0.4)
               + scale_y_log10()
               + ggtitle("Multiple Site slope timing")
+              + zmargin
 )
 
 print(gg_ms_slope_time)
@@ -213,6 +214,7 @@ gg_cs <- (ggplot(data=cs_data,aes(x=size,y=sd,fill=platform))
  + facet_grid(.~sdtype, scale="free_y")
  + geom_violin(position=position_dodge(width=0.2),alpha=0.4)
  + ggtitle("Compound symmetric using pez and lme4")
+ + zmargin
 )
 
 print(gg_cs)
@@ -221,6 +223,7 @@ gg_cstime <- (ggplot(data=cs_data,aes(x=size,y=time,fill=platform))
  + geom_violin(position=position_dodge(width=0),alpha=0.4)
  + scale_y_log10()
  + ggtitle("CS timing")
+ + zmargin
 )
 
 print(gg_cstime)
@@ -251,6 +254,7 @@ gg_cs_slope <- (ggplot(data=cs_slope_data,aes(x=size,y=sd,fill=platform))
                + geom_violin(position=position_dodge(width=0.2),alpha=0.4)
                # + scale_y_log10()
                + ggtitle("CS slope using pez and lme4")
+               + zmargin
 )
 
 print(gg_cs_slope)
@@ -260,6 +264,7 @@ gg_cs_slope_time <- (ggplot(data=cs_slope_data,aes(x=size,y=time,fill=platform))
                     + geom_violin(position=position_dodge(width=0),alpha=0.4)
                     + scale_y_log10()
                     + ggtitle("CS slope timing")
+                    + zmargin
 )
 
 print(gg_cs_slope_time)

### 2018-02-18 ms.tex
temp pushing look up table, it is currently too repetative. working on presentation

@@
 \usepackage[]{color}
+\usepackage{tabularx}
 %% use \xspace to allow for space after a macro as necessary
@@
 
+\newpage
+\section{translation}
+
+$1 \mid Sp_{phylo}$, $0 + X_{E} \mid Sp_{phylo}$ and $X_{E} \mid Sp_{phylo}$ means phylogenetically related have similar response,
+
+\begin{tabularx}{\textwidth}{|l|X|X|}
+\hline
+Formula & Statistics & Biology \\
+\hline
+$1 \mid Sp$ &
+random species intercept; variation within species in mean response across all factors &
+variation of how species respond \\
+\hline
+
+$0 + X_{E} \mid Sp$ &
+random slope of environment factor within species; variation in coefficient within species for the environmental factor &
+variation of how species respond to the same environmental factor \\
+\hline
+
+$1 + X_{E} \mid Sp$ &
+random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental factor with correlated mean response across all other factors &
+variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
+\hline
+
+$1 \mid Site:Sp $ &
+random variation in intercept among species within sites &
+variation of how species respond within sites \\
+\hline
+
+$1 | Sp_{Phylo} $ &
+variation among species in mean response across all factors demonstrate phylogenetic signal &
+phylogenetically related species respond similarly \\
+\hline
+
+$0 + X_{E} \mid Sp_{Phylo}$ &
+variation among species for environmental factors demostrate phylogenetic signal &
+phylogenetically related species respond similar (share common response) to the same environmental factor \\
+\hline
+\end{tabularx}
 \end{document}

### 2018-02-18 parameters.R

@@ -17,7 +17,7 @@ beta1 <- 0
 # magnitude of random effects
 sd.B0 <- 4
 sd.B1 <- 2
-sd.B1 <- 0.00000001
+#sd.B1 <- 0.00000001
 
 rho.B01 <- 0.7
 rho.B01 <- 0

### 2018-02-18 phyloglmm.R

@@ -22,9 +22,9 @@ print(head(dat))
 
 lme4time <- system.time(
 	lme4fit <- phylo_lmm(Y ~ X
-#		+ (1|sp)
-#		+ (0 + X|sp)
-		+ (1+X|sp)
+		+ (1|sp)
+		+ (0 + X|sp)
+#		+ (1+X|sp)
 		, data=dat
 		, phylonm = c("sp","site:sp")
 		, phylo = phy
@@ -43,4 +43,4 @@ print(summary(lme4fit))
 
 lme4_list <- list(lme4fit, lme4time)
 
-saveRDS(lme4_list, file=paste("datadir/lme4_mmslope_cor",size,seed,"rds",sep="."))
+saveRDS(lme4_list, file=paste("datadir/lme4_test",size,seed,"rds",sep="."))

### 2018-02-18 simulate_tree.R

@@ -13,14 +13,18 @@ phy <- compute.brlen(phy, method = "Grafen", power = 0.5)
 Vphy <- vcv(phy)
 # Vphy <- Vphy/(det(Vphy)^(1/nspp))
 
+iD <- t(chol(Vphy))
+Xsd <- 20
+
 # Generate environmental site variable and standardize it
 
 if(nsite == 1){
-	X <- rnorm(n=nspp,sd=1)
+	X <- rnorm(n=nspp,sd=Xsd)
+	X <- iD %*% rnorm(n=nspp,sd=Xsd)
 }
 if(nsite > 1){
 X <- matrix(1:nsite, nrow = 1, ncol = nsite)
-X <- (X - mean(X))/sd(X)
+# X <- (X - mean(X))/sd(X)
 }
 site_name <- sapply(1:nsite, function(x){paste("site","_",x,sep="")})
 

### 2018-02-26 ms.tex
tweaking intro

--- 2018-02-14 ms.tex
@@
 \bibliographystyle{chicago}
-\title{Need a good title.}
+\title{Need a good title. Submit to Methods of Ecology and Evolution or Evolutionary Biology}
 \author{Michael Li and Ben Bolker}
@@
 \section{Introduction}
-\ml{I will fill in appropriate citation later}
 
-Given a phylogenetic tree, phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes that generate species and exhibit in their traits.
-Ecologist have long considered the effects of species functional traits and environmental conditions on their phenotypic traits and also incorportating historical relationships of species.
+Ecologists and evolutionary biologists have long used various comparative techniques to study general patterns exhibited by species in response to its own phenotypic traits or different environmental and evolutionary changes.
+Phylogenetic compariative methods aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes by incorportating the histroical relationships shared across species via phylogenetic (evolutionary) trees.
+\ml{parsimonious can be a confusing word here}
+%Given a phylogenetic tree, phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes that generate species and exhibit in their traits.
+%Ecologist have long considered the effects of species phenotypic traits and environmental conditions on their responses and also incorportating historical relationships of species.
 Decades of work have created various different methods that enables researchers to analyze species relativeness problems for different types of data and questions.
 But many challenges remain.
-In particular, incorportating phylogenetic signals in high dimensional systems is always challenging, especially with large phylogenies where computation powers are limited.
-In such case, researchers often use simpler models with minimal phylogenetic signals limiting at the individuals level and neglect additional phylogenetic signals present in the system.
+In particular, accounting for phylogenetic signals in high dimensional systems is always challenging, especially with large phylogenies where computation powers are limited.
+\ml{define phylogenetic signal?}
+In such case, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
 
-Recent years, researchers have begun to incorportate phylogenetic-based methods in various modelling applications \ml{cite}.
+Recent years, researchers have begun to incorportate phylogenetic-based methods in various modelling applications \ml{cite, look up phylogenetic meta-analysis as well}.
 The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.
+
 Despite the choice of modelling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis.
@@
 
+\subsection{sups}
+
+\begin{tabular}{|c|c|c|}
+\hline
+Formula & Statistics & Biology \\
+\hline
+$1 \mid Sp$ &
+random species intercept; variation within species in average response across all factors &
+variation of how species respond \\
+
+$0 + X_{E} \mid Sp$ &
+random slope of environment factor within species; variation in coefficient within species for the environmental factor &
+variation of how species respond to the same environmental factor
+
+$1 + X_{E} \mid Sp$ &
+random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental with correlated average response across all other factors &
+variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
+
+$1 \mid Site:Sp $ &
+random variation in intercept among species within sites &
+variation of how species respond within sites \\
+
+$1 | Sp_{Phylo} $ &
+variation among species in average response across all factors demonstrate phylogenetic signal &
+phylogenetically related species respond similarly \\
+
+$0 + X_{E} \mid Sp_{Phylo}$ &
+variation among species for environmental factors demostrate phylogenetic signal &
+phylogenetically related species respond similar (share common response) to the same environmental factor
+
+\end{tabular}
+
 \subsection{Simulations}

### 2018-02-28 ms.tex
what happen to my look up table

@@
 
-Ecologists and evolutionary biologists have long used various comparative techniques to study general patterns exhibited by species in response to its own phenotypic traits or different environmental and evolutionary changes.
-Phylogenetic compariative methods aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes by incorportating the histroical relationships shared across species via phylogenetic (evolutionary) trees.
+Phylogenetic compariative methods have been widely used to study species relativeness problems.
+%Ecologists and evolutionary biologists have long used various comparative techniques to study general patterns exhibited by species in response to its own phenotypic traits or different environmental and evolutionary changes.
+Phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes by incorportating the histroical relationships shared across species via phylogenetic (evolutionary) trees.
 \ml{parsimonious can be a confusing word here}
+Although such models attemp to account for species correlations, they often make strong and over-simplified assumptions of the evolutionary process. 
 %Given a phylogenetic tree, phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes that generate species and exhibit in their traits.
 %Ecologist have long considered the effects of species phenotypic traits and environmental conditions on their responses and also incorportating historical relationships of species.
-Decades of work have created various different methods that enables researchers to analyze species relativeness problems for different types of data and questions.
+Decades of work have created various different methods that enables researchers to incorportate phylogenetic effects in analyzing species relativeness applications.
 But many challenges remain.
-In particular, accounting for phylogenetic signals in high dimensional systems is always challenging, especially with large phylogenies where computation powers are limited.
-\ml{define phylogenetic signal?}
-In such case, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
+In particular, accounting for phylogenetic effects in high dimensional systems is always challenging, especially for models incorportating multiple forms of phylogenetic effects, and especially with large phylogenies where computation powers are limited.
+\ml{Define phylogenetic effects. PE can mean a lot of different things (see Pagel's tree transformation). What we mean here is strictly phylogenetic signal.}
+Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \ml{Li and Ives 2017 and independence thing} --
+in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
 
-Recent years, researchers have begun to incorportate phylogenetic-based methods in various modelling applications \ml{cite, look up phylogenetic meta-analysis as well}.
+%In the past few decades, there have been an accelerating increase capabilities in discovering evolutionary history of species and methods to reconstruct phylogenetic tree are making species historical evolutionary data avalible at higher resolution.
+A lot of multi species analysis, and meta analysis. Accounting for species level variation, study and time, very few accounts for phylogenetic signal variation.
+
+researchers have begun to incorportate phylogenetic-based methods in various modelling applications \ml{cite, look up phylogenetic meta-analysis as well}.
 The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.

### 2018-02-28 main.tex
temp pushing, going to clean up and delete some text

--- 2018-02-28 ms.tex
@@
-\documentclass[12pt]{article}
-\usepackage[]{graphicx}
-\usepackage[]{color}
-%% use \xspace to allow for space after a macro as necessary
-\usepackage{xspace}
-\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
-\usepackage{pdfpages}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{commentcol}{rgb}{0.345, 0.345, 0.945}
-\newcommand{\comment}[1]{\textcolor{commentcol}{[[#1]]}}%
-
-% general-purpose mathrm macros
-\newcommand{\symsub}[2]{\ensuremath{#1_{\tiny \mathrm{#2}}}\xspace}
-\newcommand{\mrm}[1]{\ensuremath{\mathrm{#1}}
-\xspace}
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{randomcolor}{rgb}{.97, .27, .67}
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\usepackage[sort]{natbib}
-\usepackage{amsmath}
-\usepackage{alltt}
-\usepackage{hyperref}
-\usepackage[utf8]{inputenc} % for accented characters
-%% stuff for editing
-%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
-%% to suppress notes & comments: \usepackage[final]{changes}
-\usepackage[backgroundcolor=lightgray]{todonotes}
-\usepackage{setspace}
-\bibliographystyle{chicago}
-\title{Need a good title. Submit to Methods of Ecology and Evolution or Evolutionary Biology}
-\author{Michael Li and Ben Bolker}
-\date{\today}
-
-\providecommand{\keywords}[1]{\textbf{\textit{Keywords:}} #1}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}
-
-%% don't typeset BMB comments
-\newcommand{\bmbhide}[1]{}
-\newcommand{\bmb}[1]{{\color{blue} BB: #1}}
-
-\newcommand{\fref}[1]{Figure~\ref{fig:#1}}
-
-\newcommand{\ml}[1]{{\color{red} ML: #1}}
-
-\newcommand{\add}[1]{{\color{blue} ADD: #1}}
-
-%\SweaveOpts{concordance=TRUE}
-%\SweaveOpts{concordance=TRUE}
-\maketitle
-
-\doublespacing
-
-\keywords{phyloglmm ... }
-
 \section{Introduction}
@@
 \ml{Define phylogenetic effects. PE can mean a lot of different things (see Pagel's tree transformation). What we mean here is strictly phylogenetic signal.}
-Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \ml{Li and Ives 2017 and independence thing} --
-in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
+Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \ml{Li and Ives 2017 and independence thing} -- in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
 
-%In the past few decades, there have been an accelerating increase capabilities in discovering evolutionary history of species and methods to reconstruct phylogenetic tree are making species historical evolutionary data avalible at higher resolution.
+State of the art of phylogenetic comparative analysis.
+
+In the past few decades, there have been an accelerating increase in discovering evolutionary history of species making historical evolutionary data accessible and at higher resolutions.
+Furthermore, there have been a steadily advance of methodological research (e.g.
+reconstruct phylogenetic tree methods, and phylogenetic compariative methods) and observational (and meta-analysis) studies on collections of species.
+Despite the widespread use of these methods and accounting for variability across species, groups, and different studies, there have been relatively few studies accounting for phylogenetic signals.
+
 A lot of multi species analysis, and meta analysis. Accounting for species level variation, study and time, very few accounts for phylogenetic signal variation.
@@
 
-We also compare three different R packages: nlme, pez, and lme4. 
+We also compare three different R packages: nlme, pez, phylolm, and lme4.
 \ml{Not sure if we should include the line "we are not using Bayesian approaches" here but defintely in the discussions.}
@@
 \end{align}
-where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m + 1$ coefficient vector; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
+where $y$ is an $n \times 1$ vector of measurable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independent variable (traits); $B$ is an $m + 1$ coefficient vector; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
 
@@
 \end{align}
-where $Z$ is an $n \times (m+1)$ model matrix for the $(m+1)$ -- dimensional vector-valued $m$ independment variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
-Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$ and $Z=Z^{c}$ where the PC is incorportated in the random effect matrix.
+where $Z$ is an $n \times (m+1)$ model matrix for the $(m+1)$ -- dimensional vector-valued $m$ independent variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
+Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$ and $Z=Z^{c}$ where the PC is incorporated in the random effect matrix.
 
@@
 The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
-Phylogenetic independent constrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
+Phylogenetic independent contrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
 Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.
@@
 \ml{double kronecker for interactions}
-
-\subsection{sups}
-
-\begin{tabular}{|c|c|c|}
-\hline
-Formula & Statistics & Biology \\
-\hline
-$1 \mid Sp$ &
-random species intercept; variation within species in average response across all factors &
-variation of how species respond \\
-
-$0 + X_{E} \mid Sp$ &
-random slope of environment factor within species; variation in coefficient within species for the environmental factor &
-variation of how species respond to the same environmental factor
-
-$1 + X_{E} \mid Sp$ &
-random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental with correlated average response across all other factors &
-variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
-
-$1 \mid Site:Sp $ &
-random variation in intercept among species within sites &
-variation of how species respond within sites \\
-
-$1 | Sp_{Phylo} $ &
-variation among species in average response across all factors demonstrate phylogenetic signal &
-phylogenetically related species respond similarly \\
-
-$0 + X_{E} \mid Sp_{Phylo}$ &
-variation among species for environmental factors demostrate phylogenetic signal &
-phylogenetically related species respond similar (share common response) to the same environmental factor
-
-\end{tabular}
 
@@
 \section{Discussion}
+\section{Conclusion}
+\newpage
+\section{translation}
 
-\section{Conclusion}
+$1 \mid Sp_{phylo}$, $0 + X_{E} \mid Sp_{phylo}$ and $X_{E} \mid Sp_{phylo}$ means phylogenetically related have similar response,
 
-\end{document}
+\begin{tabularx}{\textwidth}{|l|X|X|}
+\hline
+Formula & Statistics & Biology \\
+\hline
+$1 \mid Sp$ &
+random species intercept; variation within species in mean response across all factors &
+variation of how species respond \\
+\hline
+
+$0 + X_{E} \mid Sp$ &
+random slope of environment factor within species; variation in coefficient within species for the environmental factor &
+variation of how species respond to the same environmental factor \\
+\hline
+
+$1 + X_{E} \mid Sp$ &
+random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental factor with correlated mean response across all other factors &
+variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
+\hline
+
+$1 \mid Site:Sp $ &
+random variation in intercept among species within sites &
+variation of how species respond within sites \\
+\hline
+
+$1 | Sp_{Phylo} $ &
+variation among species in mean response across all factors demonstrate phylogenetic signal &
+phylogenetically related species respond similarly \\
+\hline
+
+$0 + X_{E} \mid Sp_{Phylo}$ &
+variation among species for environmental factors demostrate phylogenetic signal &
+phylogenetically related species respond similar (share common response) to the same environmental factor \\
+\hline
+\end{tabularx}

### 2018-02-28 ms.tex
temp pushing, going to clean up and delete some text

--- 2018-02-18 ms.tex
@@
 \usepackage{tabularx}
-%% use \xspace to allow for space after a macro as necessary
+% use \xspace to allow for space after a macro as necessary
 \usepackage{xspace}
@@
 
-\section{Introduction}
-\ml{I will fill in appropriate citation later}
+\include{main}
 
-Given a phylogenetic tree, phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes that generate species and exhibit in their traits.
-Ecologist have long considered the effects of species functional traits and environmental conditions on their phenotypic traits and also incorportating historical relationships of species.
-Decades of work have created various different methods that enables researchers to analyze species relativeness problems for different types of data and questions.
-But many challenges remain.
-In particular, incorportating phylogenetic signals in high dimensional systems is always challenging, especially with large phylogenies where computation powers are limited.
-In such case, researchers often use simpler models with minimal phylogenetic signals limiting at the individuals level and neglect additional phylogenetic signals present in the system.
-
-Recent years, researchers have begun to incorportate phylogenetic-based methods in various modelling applications \ml{cite}.
-The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.
-Despite the choice of modelling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis.
-There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematically look at comparative performance of comparative phylogenetic methods.
-
-In this paper, we propose an alternative, yet straightforward but little-used method of constructing phylogenetic correlations directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
-We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorportate various complexity of phylogenetic signals in the model. 
-We compare model approaches and explore various complex modelling strategies in the mixed model framework and a flexible platform for researches to explore biologically interesting questions.
-
-We also compare three different R packages: nlme, pez, and lme4. 
-\ml{Not sure if we should include the line "we are not using Bayesian approaches" here but defintely in the discussions.}
-
-\section{Materials and Methods}
-
-The typical phylogenetic regression is of the form:
-\begin{align}
-y & = XB + \epsilon \\
-\epsilon & \sim MVN(0,\sigma^{2}C),
-\label{eq:gls}
-\end{align}
-where $y$ is an $n \times 1$ vector of measureable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independment variable (traits); $B$ is an $m + 1$ coefficient vector; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
-
-% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
-%\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
-An alternative modelling approach is to use linear mixed effects modelling framework.
-The typical linear mixed model has the form:
-\begin{align}
-Y & \sim F(\mu) \\
-g(\mu) & = XB + Zb + \epsilon_{mm} \\
-\epsilon_{mm} & \sim MVN(0,\sigma^2)
-\end{align}
-where $Z$ is an $n \times (m+1)$ model matrix for the $(m+1)$ -- dimensional vector-valued $m$ independment variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
-Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$ and $Z=Z^{c}$ where the PC is incorportated in the random effect matrix.
-
-\subsection{Phylogenetic Tree Transformation}
-The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
-Phylogenetic independent constrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
-Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.
-
-An alternative approach is to model the phylogenetic correlation as a \textit{gaussian process}. 
-In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk and daughter species after speciation are all independent.  
-In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
-Thus, the evolutionary history for each species can be model with a sequence of independent errors, rather than having to impose a correlation structure on the random effects.
-
-The random-effect model matrix $Z$ can be decomposed into term-wise model matrix $Z_{i}$ as described in \ml{in text citation lme4}.
-Thus, the phylogenetic correlated random-effect matrix $Z^{C}_{i}$ is
-
-\begin{equation}
-Z^{C}_{i} = (BS^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T},
-\end{equation}
-
-\ml{double check if $l and p$ are the same with lme4}
-
-where $\ast$ is the Khatri-Rao product, $BS_{i}$ is an $l_{i} \times b_{i}$ matrix of species--branch relations; $J_{i}$ the indicator matrix of grouping factors indices matrix size $n \times l_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
-
-\ml{double kronecker for interactions}
-
-\subsection{Simulations}
-
-% \begin{table}
-% \begin{center}
-\begin{tabular}{|c|c|c|}
-\hline
-Type of RE	& Formula	& Platform	\\
-\hline
-
-Single site intercept	  &	1 $\mid$ sp	& gls, lme4 \\
-Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& lme4 \\
-Single site intercept and correlated slopes 			& 1 + trait $\mid$ sp					& lme4 \\
-\hline
-Multiple site intercept 									&  1 $\mid$ sp 						& pez, lme4 \\
-Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + trait $\mid$ sp 		& pez, lme4 \\
-Multiple site intercept and correlated slopes 	& 	1 + trait $\mid$ sp					& lme4 \\
-\hline
-\end{tabular}
-% \vspace{1in}
-% ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
-% \end{center}
-% \end{table}
-
-The simulation models is based on glmm framework described in \ml{eq labels 3-5}.   
-To construct the simulation model, let $n$ be the number of species distributed among $m$ sites.
-Let $Y$ be the $mn \times 1$ vectors of response trait.  
-We generate test data using three different $n$ (20, 100, 500) and two different $m$ (1, 20) using a simple glmm framework that varies in complexity in random effect described in \ml{Table 1, macro this with labels}.
-We also extend the multiple site models described in \ml{Table 1} with species by site interaction random effects. 
-All simulations and fits were performed with R.
-
-\subsection{Model Fitting}
-
-\section{Results}
-
-\section{Discussion}
-
-\section{Conclusion}
-
-\newpage
-\section{translation}
-
-$1 \mid Sp_{phylo}$, $0 + X_{E} \mid Sp_{phylo}$ and $X_{E} \mid Sp_{phylo}$ means phylogenetically related have similar response,
-
-\begin{tabularx}{\textwidth}{|l|X|X|}
-\hline
-Formula & Statistics & Biology \\
-\hline
-$1 \mid Sp$ &
-random species intercept; variation within species in mean response across all factors &
-variation of how species respond \\
-\hline
-
-$0 + X_{E} \mid Sp$ &
-random slope of environment factor within species; variation in coefficient within species for the environmental factor &
-variation of how species respond to the same environmental factor \\
-\hline
-
-$1 + X_{E} \mid Sp$ &
-random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental factor with correlated mean response across all other factors &
-variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
-\hline
-
-$1 \mid Site:Sp $ &
-random variation in intercept among species within sites &
-variation of how species respond within sites \\
-\hline
-
-$1 | Sp_{Phylo} $ &
-variation among species in mean response across all factors demonstrate phylogenetic signal &
-phylogenetically related species respond similarly \\
-\hline
-
-$0 + X_{E} \mid Sp_{Phylo}$ &
-variation among species for environmental factors demostrate phylogenetic signal &
-phylogenetically related species respond similar (share common response) to the same environmental factor \\
-\hline
-\end{tabularx}
 \end{document}

### 2018-02-28 main.tex
need to straighten some parts out

@@
 Phylogenetic compariative methods have been widely used to study species relativeness problems.
-%Ecologists and evolutionary biologists have long used various comparative techniques to study general patterns exhibited by species in response to its own phenotypic traits or different environmental and evolutionary changes.
 Phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes by incorportating the histroical relationships shared across species via phylogenetic (evolutionary) trees.
@@
 Although such models attemp to account for species correlations, they often make strong and over-simplified assumptions of the evolutionary process. 
-%Given a phylogenetic tree, phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes that generate species and exhibit in their traits.
-%Ecologist have long considered the effects of species phenotypic traits and environmental conditions on their responses and also incorportating historical relationships of species.
 Decades of work have created various different methods that enables researchers to incorportate phylogenetic effects in analyzing species relativeness applications.
@@
 
-State of the art of phylogenetic comparative analysis.
+Incorportating phylogenetic links between species and combine evolutionary changes with statistical processes advances the development of phylogenetic comparative methods.
+The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion. 
+Although the Brownian motion is an over-simplified model of evolution, many of the existing phylogenetic comparative methods are built upon.
+For example, phylogenetic regression via independent contrasts, phylogenetic generalized least squares, Pagel's $\lambda$, Blomberg's $K$ are methods developed in the last couple of decades and have provided important insights in this field.
+Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.
+Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes, Pagel's $\delta$ and $\kappa$ transformations where it explores variation of evolutionary rates and speciational changes through time respectively).
 
@@
 reconstruct phylogenetic tree methods, and phylogenetic compariative methods) and observational (and meta-analysis) studies on collections of species.
-Despite the widespread use of these methods and accounting for variability across species, groups, and different studies, there have been relatively few studies accounting for phylogenetic signals.
+Despite the widespread use of these methods and ways to account for variability across species, groups, and different studies, there have been relatively few studies accounting for phylogenetic signals.
 
-A lot of multi species analysis, and meta analysis. Accounting for species level variation, study and time, very few accounts for phylogenetic signal variation.
+%The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.
 
-researchers have begun to incorportate phylogenetic-based methods in various modelling applications \ml{cite, look up phylogenetic meta-analysis as well}.
-The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.
-
-Despite the choice of modelling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis.
-There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematically look at comparative performance of comparative phylogenetic methods.
+%Despite the choice of modelling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis.
+%There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematically look at comparative performance of comparative phylogenetic methods.
 

### 2018-03-04 main.tex
bib not working

@@
 \ml{Define phylogenetic effects. PE can mean a lot of different things (see Pagel's tree transformation). What we mean here is strictly phylogenetic signal.}
-Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \ml{Li and Ives 2017 and independence thing} -- in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
+Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \cite{felsenstein1985phylogenies, li2017statistical} -- in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
 
 Incorportating phylogenetic links between species and combine evolutionary changes with statistical processes advances the development of phylogenetic comparative methods.
-The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion. 
+The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}. 
 Although the Brownian motion is an over-simplified model of evolution, many of the existing phylogenetic comparative methods are built upon.
-For example, phylogenetic regression via independent contrasts, phylogenetic generalized least squares, Pagel's $\lambda$, Blomberg's $K$ are methods developed in the last couple of decades and have provided important insights in this field.
+For example, phylogenetic regression via independent contrasts \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares \citep{grafen1989phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, Blomberg's $K$ \citep{blomgerg2003} are methods developed in the last couple of decades and have provided important insights in this field.
 Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.

### 2018-03-04 main.tex
temp pushing simulation section

@@
 \ml{Define phylogenetic effects. PE can mean a lot of different things (see Pagel's tree transformation). What we mean here is strictly phylogenetic signal.}
-Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \cite{felsenstein1985phylogenies, li2017statistical} -- in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
+Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \citep{felsenstein1985phylogenies, li2017statistical} -- in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
 
@@
 \end{align}
-where $y$ is an $n \times 1$ vector of measurable response phenotypic trait; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independent variable (traits); $B$ is an $m + 1$ coefficient vector; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
+where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independent variable (phenotypic traits or environmental variable); $B$ is an $m + 1$ coefficient vector; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
 
@@
 %\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
-An alternative modelling approach is to use linear mixed effects modelling framework.
+An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
 The typical linear mixed model has the form:
@@
 \begin{equation}
-Z^{C}_{i} = (BS^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T},
+Z^{C}_{i} = (S^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T},
 \end{equation}
@@
 
-where $\ast$ is the Khatri-Rao product, $BS_{i}$ is an $l_{i} \times b_{i}$ matrix of species--branch relations; $J_{i}$ the indicator matrix of grouping factors indices matrix size $n \times l_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
+where $\ast$ is the Khatri-Rao product, $S_{i}$ is an $l_{i} \times b_{i}$ matrix of species--branch relations; $J_{i}$ the indicator matrix of grouping factors indices matrix size $n \times l_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
 
-\ml{double kronecker for interactions}
+\ml{note: $SS^T = C$ under BM assumption}
+%\ml{double kronecker for interactions}
 
-\subsection{Simulations}
+\subsection{Simulation}
+
+We generated test data based on the simple mixed model formulation \ml{ref equation (3-5)} with a single response variable $Y$ and a single predictor continuous variable $X$ from $n$ species and $m$ sites.
+To account for multiple levels of phylogenetic signal, we assume phylogenetic signal in the intercept $b_0$, and slope $b_1$.
+We further assume that random slope and intercept are correlated.
+
+\ml{talk about platforms and single/multiple sites}
 

### 2018-03-05 main.tex
multiple obs works now for non-interaction RE

@@
 In particular, accounting for phylogenetic effects in high dimensional systems is always challenging, especially for models incorportating multiple forms of phylogenetic effects, and especially with large phylogenies where computation powers are limited.
-\ml{Define phylogenetic effects. PE can mean a lot of different things (see Pagel's tree transformation). What we mean here is strictly phylogenetic signal.}
+\ml{Define phylogenetic effects.
+PE can mean a lot of different things (see Pagel's tree transformation).
+What we mean here is strictly phylogenetic signal-- extent to which trait/response values are statistically related to phylogeny.}
 Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \citep{felsenstein1985phylogenies, li2017statistical} -- in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
@@
 In this paper, we propose an alternative, yet straightforward but little-used method of constructing phylogenetic correlations directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
+
+\ml{flexabilitiy: multiple obs per species and multiple levels of phylogenetic signal}
 We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorportate various complexity of phylogenetic signals in the model. 
@@
 
-We also compare three different R packages: nlme, pez, phylolm, and lme4.
+We also compare three different R packages: nlme:gls, pez, phylolm, and lme4.
+In principle, any given valid method should converge/get the same result/conclusion.
+However, even with the relatively simple models considered here, methods that are not flexible/parameter constrinted can experience problems.
+Furthermore, even when different methods converge to essentially the same result, they may show large differences in computational efficiency.
 \ml{Not sure if we should include the line "we are not using Bayesian approaches" here but defintely in the discussions.}
@@
 
-\ml{note: $SS^T = C$ under BM assumption}
+\ml{note: $SS^T = Cov(phylo)$ under BM assumption}
 %\ml{double kronecker for interactions}
@@
 
-We generated test data based on the simple mixed model formulation \ml{ref equation (3-5)} with a single response variable $Y$ and a single predictor continuous variable $X$ from $n$ species and $m$ sites.
+We generated test data based on the simple mixed model formulation \ml{ref equation (3-5)} with a single response variable $Y$ and a single predictor continuous variable $X$ from $n$ (20, 100, and 500) species, and $m$ (1 and 20) sites.
 To account for multiple levels of phylogenetic signal, we assume phylogenetic signal in the intercept $b_0$, and slope $b_1$.
 We further assume that random slope and intercept are correlated. 
+For models with multiple sites, we also assume for species and site interactions random effects, an equilvalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
 
-\ml{talk about platforms and single/multiple sites}
+Many platforms 
+- single observations per species (IC and pgls)
+- single observation per species per site (pez)
+- lme4 does everything
 

### 2018-03-05 dune_lme4.R

@@ -34,16 +34,19 @@ dat <- (dat
   %>% rowwise()
   %>% mutate(obs = sp
   )
-)	
+)
+
+dat2 <- rbind(dat,dat[560,])
+dat2[561,"Y"]
 
 lme4time_1 <- system.time(
   lme4fit_1 <- phylo_lmm2(Y ~ 1 + log.sla + annual 
-		# + (1|obs) 
+		 + (1|obs) 
 		+ (1|sp)
-    # + (1 | sp:site)
-		 # + (0 + log.sla | site)
-		# + (1|site) 
-		, data=dat
+     # + (1 | sp:site)
+		  + (0 + log.sla | site)
+		 + (1|site) 
+		, data=dat2
 		, phylonm = c("sp","sp:site","site:sp")
 		, nsp = 28
 		, phylo = phy
@@ -91,7 +94,7 @@ REs <- get_RE(veg.long = dune.veg2
 	, trait.re = c("log.sla")
 	, phylo = dune.phylo2
 	, trans = "log"
-	, stand = TRUE
+	, stand = FALSE
 )
 
 # get_RE returns a list of random effects in this order:
@@ -108,21 +111,21 @@ re.sla = list(unname(unlist(dat["log.sla"])), site = dat$site, covar = diag(nsit
 
 
 peztime_1 <- system.time(
-  pezfit_1 <-  communityPGLMM(formula = "Y ~ 1 + log.sla + annual"
-    , data = dat
-    , family = "gaussian"
-    , sp = dat$sp
-    , site = dat$site
-    , random.effects = list(#re.sp
-    re.sp.phy
-      # re.nested.phy
-      #, re.sla
-      #, re.site
-      )
-    , REML = F
-    , verbose = F
-    , s2.init = c(1.5, rep(0.01, 4))
-  )
+ pezfit_1 <-  communityPGLMM(formula = "Y ~ 1 + log.sla + annual"
+   , data = dat
+   , family = "gaussian"
+   , sp = dat$sp
+   , site = dat$site
+   , random.effects = list(re.sp
+   , re.sp.phy
+     # , re.nested.phy
+     , re.sla
+     , re.site
+     )
+   , REML = F
+   , verbose = F
+   , s2.init = c(1.5, rep(0.01, 4))
+ )
 )
 # 
 # peztime_2 <- system.time(

### 2018-03-05 new_phylo_setup.R

@@ -2,7 +2,7 @@
 
 phylo_lmm2 <- function(formula,data,phylo,phylonm=NULL,phyloZ=NULL,nsp=NULL,control,REML){
   lmod <- lFormula2(formula=formula,data = data,control=control, REML=REML,phylonm=phylonm, phyloZ=phyloZ)
-  lmod$reTrms <- modify_phylo_retrms2(lmod$reTrms,phylo,phylonm,phyloZ,nsp)
+  # lmod$reTrms <- modify_phylo_retrms2(lmod$reTrms,phylo,phylonm,phyloZ,nsp)
   devfun <- do.call(mkLmerDevfun, lmod)
   opt <- optimizeLmer(devfun,control=control$optCtrl)
   mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
@@ -222,7 +222,8 @@ modify_phylo_retrms2 <- function(rt,phylo,phylonm,phyloZ,nsp){
   ## terms on the left side of the bar/pipe
   
   ### lFormula is creating the all RE index w.r.t nsp lengths into a simple vector, we have to use the function above to split properly
-  Lind_list <- split(rt[["Lind"]],rep(seq_along(Lind_split_length),Lind_split_length*nsp))
+  # Lind_list <- split(rt[["Lind"]],rep(seq_along(Lind_split_length),Lind_split_length*nsp))
+  Lind_list <- split(rt[["Lind"]],rep(seq_along(Gpdiff),Gpdiff))
   if(names(rt[["cnms"]][1]) %in% c("sp:site","site:sp")){ ### hack
     for(i in 1:length(rt$cnms)){
       Lind_list[[i]] <- rep(i,sum(rt$Lind==i))

### 2018-03-07 main.tex
finish up in the am

@@
 Although the Brownian motion is an over-simplified model of evolution, many of the existing phylogenetic comparative methods are built upon.
-For example, phylogenetic regression via independent contrasts \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares \citep{grafen1989phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, Blomberg's $K$ \citep{blomgerg2003} are methods developed in the last couple of decades and have provided important insights in this field.
+For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods developed in the last couple of decades and have provided important insights in this field.
 Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.
@@
 
+\ml{write a paragraph on pglmm?
+Flexbility paragraph and say PIC/GLS assumes the tpoplogy and branch lengths are error free, Pagel's lambda and Blomberg's K accounts for additional tip variations of the phylogeny, to account for additional variation such as measurement error and such, people turn to Bayesian mixed models \cite{de2012bayesian, hadfield2010general}}
+
 %The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.
@@
 In this paper, we propose an alternative, yet straightforward but little-used method of constructing phylogenetic correlations directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
-
 \ml{flexabilitiy: multiple obs per species and multiple levels of phylogenetic signal}
-We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorportate various complexity of phylogenetic signals in the model. 
+We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorportate various complexity of phylogenetic signals in the model and measurement errors. 
 We compare model approaches and explore various complex modelling strategies in the mixed model framework and a flexible platform for researches to explore biologically interesting questions.
@@
 For models with multiple sites, we also assume for species and site interactions random effects, an equilvalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
-
-Many platforms 
-- single observations per species (IC and pgls)
-- single observation per species per site (pez)
-- lme4 does everything
+Table \ml{ref something} provides the statistially equilvalent fitting capabilities of common/existing phylogenetic methods.
 
@@
 
-The simulation models is based on glmm framework described in \ml{eq labels 3-5}.   
-To construct the simulation model, let $n$ be the number of species distributed among $m$ sites.
-Let $Y$ be the $mn \times 1$ vectors of response trait.  
-We generate test data using three different $n$ (20, 100, 500) and two different $m$ (1, 20) using a simple glmm framework that varies in complexity in random effect described in \ml{Table 1, macro this with labels}.
-We also extend the multiple site models described in \ml{Table 1} with species by site interaction random effects. 
-All simulations and fits were performed with R.
-
-\subsection{Model Fitting}
-
 \section{Results}
 
+\begin{center}
+\begin{figure}[h]
+\includegraphics[scale=0.8,page=1]{./git_push/ssplot.pdf}
+\label{ssplot}
+\end{figure}
+\end{center}
+
+The full model (which matches the simulations model that incorportates phylogenetic signal on the intercept, slope and correlation) provides good estimates of the different levels of the phylogenetic variation signal (Figure 1 and 2).
+
+In general, as the number of species (or phylogeny) increases, all phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) except PGLS in the single site simulations.
+It is unclear how different levels of phylogenetic signal is being confounded under the error free (without tip variation) assumption for the PGLS models.
+The estimated phylogenic signal estimates appear to be increasing as the number of species increase.
+
+\begin{center}
+\begin{figure}[h]
+\includegraphics[scale=0.8,page=1]{./csplot.pdf}
+\end{figure}
+\end{center}
+
+\begin{center}
+\begin{figure}[h]
+\includegraphics[scale=0.8,page=2]{./csplot.pdf}
+\end{figure}
+\end{center}
+
+% \begin{center}
+% \begin{figure}[h]
+% \includegraphics[scale=1]{./ssplot.pdf}
+% \end{figure}
+% \end{center}
+
+\newpage
+
 \section{Discussion}
+
+large datasets
+limitations (unbalance datasets)
+this trick works in bayesian/mcmc framework
+
+\newpage
+
 \section{Conclusion}
+
 \newpage
+
 \section{translation}

### 2018-03-07 main.tex
does it compile?

@@
 % \begin{center}
-\begin{tabular}{|c|c|c|}
+\begin{tabularx}{\textwidth}{|X|X|X|}
 \hline
@@
 
-Single site intercept	  &	1 $\mid$ sp	& gls, lme4 \\
+Single site intercept	  &	1 $\mid$ sp	& gls (if residual = 0), phyloglm, lme4 \\
 Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& lme4 \\
@@
 \hline
-\end{tabular}
+\end{tabularx}
 % \vspace{1in}
@@
 It is unclear how different levels of phylogenetic signal is being confounded under the error free (without tip variation) assumption for the PGLS models.
-The estimated phylogenic signal estimates appear to be increasing as the number of species increase.
+Models with only two parameters/levels of variation captures the phylogenetic intercept and tip variation reasonably well to the simulated parameters from the full model dispite not including phylogenetic random slope and slope intercept correlation.
+
+Platforms using the mixed model framework for multiple site simulations follows similar behaviour as single site simulation (i.e.
+as the number of species increase, differnt levels of phylogenetic signals and variations are closer to the simulated parameters).
+There are no clear difference with phylogenetic signamodels without random slope and intercept correlation
 
@@
 
+\subsection{over coming limitation}
 large datasets
@@
 
+\subsection{conflation of phylogenetic process and observation process error}
+
+clarity
+
+\subsection{rethinking phylogenetic comparative methods and statistical test}
+
+\ml{cite Li et al and the other paper with model selection before adding RE}
 \newpage

### 2018-03-13 main.tex
updating ms

@@
 
-% \begin{table}
-% \begin{center}
-\begin{tabularx}{\textwidth}{|X|X|X|}
-\hline
-Type of RE	& Formula	& Platform	\\
-\hline
-
-Single site intercept	  &	1 $\mid$ sp	& gls (if residual = 0), phyloglm, lme4 \\
-Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& lme4 \\
-Single site intercept and correlated slopes 			& 1 + trait $\mid$ sp					& lme4 \\
-\hline
-Multiple site intercept 									&  1 $\mid$ sp 						& pez, lme4 \\
-Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + trait $\mid$ sp 		& pez, lme4 \\
-Multiple site intercept and correlated slopes 	& 	1 + trait $\mid$ sp					& lme4 \\
-\hline
-\end{tabularx}
-% \vspace{1in}
-% ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
-% \end{center}
-% \end{table}
+% % \begin{table}
+% % \begin{center}
+% 
+% \begin{tabularx}{\textwidth}{|X|X|X|}
+% \hline
+% Type of RE	& Formula	& Platform	\\
+% \hline
+% 
+% Single site intercept	  &	1 $\mid$ sp	& gls (if residual = 0), phyloglm, lme4 \\
+% Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& lme4 \\
+% Single site intercept and correlated slopes 			& 1 + trait $\mid$ sp					& lme4 \\
+% \hline
+% Multiple site intercept 									&  1 $\mid$ sp 						& pez, lme4 \\
+% Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + trait $\mid$ sp 		& pez, lme4 \\
+% Multiple site intercept and correlated slopes 	& 	1 + trait $\mid$ sp					& lme4 \\
+% \hline
+% \end{tabularx}
+% % \vspace{1in}
+% % ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
+% % \end{center}
+% % \end{table}
 
@@
 as the number of species increase, differnt levels of phylogenetic signals and variations are closer to the simulated parameters).
-There are no clear difference with phylogenetic signamodels without random slope and intercept correlation
+All parameter estimates are much more robust for the multiple site than the single site simulations especially slope-intercept correlations. 
+There are no clear difference with phylogenetic signal models without random slope and intercept correlation.
+However, dispite the similarities of robustness estimation, the efficiency (measured by computational speed) of sampling independent errors in lme4 is dramatically better (orders of magitude faster than pez.
+Furthermore, it was not practical for pez to fit large sample size simulation data as the computational speed did not scale linearly.
 
@@
 
-\subsection{over coming limitation}
+We have fitted models including maximum capabilities to model varying levels of variation/phylogenetic signal to simulated data with ....
+Using models that can include/account for multiple levels of variation is necessary for robust fits, but models that include phylogenetic signal in the intercept level can work as well as the full model.
+Under the Brownian motion evolution process assumption, modelling via independent errors is much more computational efficient than imposing equilvalent correlation structures.
+
+\subsection{Process and observation errors}
+
+Another way to think about various forms/sources of phylogenetic signals and uncertainties is characterizing into process and observation errors.
+
+\subsection{Model Flexibilities}
+
 large datasets
@@
 this trick works in bayesian/mcmc framework
-
-\subsection{conflation of phylogenetic process and observation process error}
 

### 2018-03-14 main.tex
first pass complete

@@
 
-Phylogenetic compariative methods have been widely used to study species relativeness problems.
-Phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes by incorportating the histroical relationships shared across species via phylogenetic (evolutionary) trees.
+Phylogenetic comparative methods have been widely used to study species relativeness problems.
+Phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes by incorporating the historical relationships shared across species via phylogenetic (evolutionary) trees.
 \ml{parsimonious can be a confusing word here}
-Although such models attemp to account for species correlations, they often make strong and over-simplified assumptions of the evolutionary process. 
-Decades of work have created various different methods that enables researchers to incorportate phylogenetic effects in analyzing species relativeness applications.
+Although such models attempt to account for species correlations, they often make strong and over-simplified assumptions of the evolutionary process. 
+Decades of work have created various different methods that enables researchers to incorporate phylogenetic effects in analyzing species relativeness applications.
 But many challenges remain.
-In particular, accounting for phylogenetic effects in high dimensional systems is always challenging, especially for models incorportating multiple forms of phylogenetic effects, and especially with large phylogenies where computation powers are limited.
+In particular, accounting for phylogenetic effects in high dimensional systems is always challenging, especially for models incorporating multiple forms of phylogenetic effects, and especially with large phylogenies where computation powers are limited.
 \ml{Define phylogenetic effects.
@@
 
-Incorportating phylogenetic links between species and combine evolutionary changes with statistical processes advances the development of phylogenetic comparative methods.
+Incorporating phylogenetic links between species and combine evolutionary changes with statistical processes advances the development of phylogenetic comparative methods.
 The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}. 
@@
 Furthermore, there have been a steadily advance of methodological research (e.g.
-reconstruct phylogenetic tree methods, and phylogenetic compariative methods) and observational (and meta-analysis) studies on collections of species.
+reconstruct phylogenetic tree methods, and phylogenetic comparative methods) and observational (and meta-analysis) studies on collections of species.
 Despite the widespread use of these methods and ways to account for variability across species, groups, and different studies, there have been relatively few studies accounting for phylogenetic signals.
@@
 \ml{write a paragraph on pglmm?
-Flexbility paragraph and say PIC/GLS assumes the tpoplogy and branch lengths are error free, Pagel's lambda and Blomberg's K accounts for additional tip variations of the phylogeny, to account for additional variation such as measurement error and such, people turn to Bayesian mixed models \cite{de2012bayesian, hadfield2010general}}
+Flexbility paragraph and say PIC/GLS assumes the topology and branch lengths are error free, Pagel's lambda and Blomberg's K accounts for additional tip variations of the phylogeny, to account for additional variation such as measurement error and such, people turn to Bayesian mixed models \cite{de2012bayesian, hadfield2010general}}
 
@@
 In this paper, we propose an alternative, yet straightforward but little-used method of constructing phylogenetic correlations directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
-\ml{flexabilitiy: multiple obs per species and multiple levels of phylogenetic signal}
-We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorportate various complexity of phylogenetic signals in the model and measurement errors. 
+\ml{flexibility: multiple obs per species and multiple levels of phylogenetic signal}
+We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorporate various complexity of phylogenetic signals in the model and measurement errors. 
 We compare model approaches and explore various complex modelling strategies in the mixed model framework and a flexible platform for researches to explore biologically interesting questions.
@@
 In principle, any given valid method should converge/get the same result/conclusion.
-However, even with the relatively simple models considered here, methods that are not flexible/parameter constrinted can experience problems.
+However, even with the relatively simple models considered here, methods that are not flexible/parameter constrained can experience problems.
 Furthermore, even when different methods converge to essentially the same result, they may show large differences in computational efficiency.
-\ml{Not sure if we should include the line "we are not using Bayesian approaches" here but defintely in the discussions.}
+\ml{Not sure if we should include the line "we are not using Bayesian approaches" here but definitely in the discussions.}
 
@@
 
-An alternative approach is to model the phylogenetic correlation as a \textit{gaussian process}. 
+An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
 In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk and daughter species after speciation are all independent.  
@@
 We further assume that random slope and intercept are correlated. 
-For models with multiple sites, we also assume for species and site interactions random effects, an equilvalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
-Table \ml{ref something} provides the statistially equilvalent fitting capabilities of common/existing phylogenetic methods.
+For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
+Table \ml{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
 
@@
 
-The full model (which matches the simulations model that incorportates phylogenetic signal on the intercept, slope and correlation) provides good estimates of the different levels of the phylogenetic variation signal (Figure 1 and 2).
+The full model (which matches the simulations model that incorporates phylogenetic signal on the intercept, slope and correlation) provides good estimates of the different levels of the phylogenetic variation signal (Figure 1 and 2).
 
@@
 It is unclear how different levels of phylogenetic signal is being confounded under the error free (without tip variation) assumption for the PGLS models.
-Models with only two parameters/levels of variation captures the phylogenetic intercept and tip variation reasonably well to the simulated parameters from the full model dispite not including phylogenetic random slope and slope intercept correlation.
+Models with only two parameters/levels of variation captures the phylogenetic intercept and tip variation reasonably well to the simulated parameters from the full model despite not including phylogenetic random slope and slope intercept correlation.
 
 Platforms using the mixed model framework for multiple site simulations follows similar behaviour as single site simulation (i.e.
-as the number of species increase, differnt levels of phylogenetic signals and variations are closer to the simulated parameters).
+as the number of species increase, different levels of phylogenetic signals and variations are closer to the simulated parameters).
 All parameter estimates are much more robust for the multiple site than the single site simulations especially slope-intercept correlations. 
 There are no clear difference with phylogenetic signal models without random slope and intercept correlation.
-However, dispite the similarities of robustness estimation, the efficiency (measured by computational speed) of sampling independent errors in lme4 is dramatically better (orders of magitude faster than pez.
+However, despite the similarities of robustness estimation, the efficiency (measured by computational speed) of sampling independent errors in lme4 is dramatically better (orders of magnitude faster than pez.
 Furthermore, it was not practical for pez to fit large sample size simulation data as the computational speed did not scale linearly.
@@
 Using models that can include/account for multiple levels of variation is necessary for robust fits, but models that include phylogenetic signal in the intercept level can work as well as the full model.
-Under the Brownian motion evolution process assumption, modelling via independent errors is much more computational efficient than imposing equilvalent correlation structures.
+Under the Brownian motion evolution process assumption, modelling via independent errors is much more computational efficient than imposing equivalent correlation structures.
 
@@
 Another way to think about various forms/sources of phylogenetic signals and uncertainties is characterizing into process and observation errors.
+Process error or in this context, variation contributed from the evolutionary matches nicely with the variation of phylogenetic signal. 
+Observation/measurement error on the other hand is harder to match in this context and data dependent.
+In the classic examples, observation error is (assume to be?) zero for PIC and PGLS; single-site observation model using phylolm (or any method with two variables describing the variance (Pagel's $\lambda$, Blomberg's K)) often match the observation/measurement error with the tip variation of the phylogeny.
+For multiple observations per species, the tip variation becomes one of the process errors with phylogenetic signal and the observation error is the variation among species and additional other non-process errors.
 
-\subsection{Model Flexibilities}
+\subsection{Extension and alternatives}
 
-large datasets
-limitations (unbalance datasets)
-this trick works in bayesian/mcmc framework
-
-clarity
-
-\subsection{rethinking phylogenetic comparative methods and statistical test}
-
-\ml{cite Li et al and the other paper with model selection before adding RE}
-\newpage
+Our analysis covers the classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models).
+Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, and phylogenetic meta analysis.
+More broadly, the simple independence error approach we developed here offers a more efficient equivalent way to do Brownian motion evolution phylogenetic comparative analysis. 
+This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using Bayesian frameworks such as MCMCglmm. 
+More importantly, this implementation in lme4 allows users to fit phylogenetic mixed models to the fullest that other frequentest platforms?
+cannot (i.e.
+large data, unbalance species observations, complex random effects) and explore new ideas.
 
@@
 
+We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical pcm to simulated data. 
+First, this new approach is magnitudes faster compared to existing phylogenetic mixed models without losing robustness/accuracy and can easily combined with any statistically mixed model framework. 
+Last and most importantly, it is more flexible in fitting large phylogenies, unbalanced datasets, and complex random effects such as slope correlations.
+
 \newpage
 
-\section{translation}
+\section{Supplements -- translation}
 
@@
 $0 + X_{E} \mid Sp_{Phylo}$ &
-variation among species for environmental factors demostrate phylogenetic signal &
+variation among species for environmental factors demonstrate phylogenetic signal &
 phylogenetically related species respond similar (share common response) to the same environmental factor \\

### 2018-04-01 main.tex
temp pushing

@@
 
-Phylogenetic comparative methods have been widely used to study species relativeness problems.
-Phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes by incorporating the historical relationships shared across species via phylogenetic (evolutionary) trees.
-\ml{parsimonious can be a confusing word here}
-Although such models attempt to account for species correlations, they often make strong and over-simplified assumptions of the evolutionary process. 
+Phylogenetic comparative methods have been widely used to study the relationships among traits of species within an evolutionary lineage.
+These approaches aim to find parsimonious models that link traits and ecological phenomena with evolutionary processes by incorporating the historical relationships shared among taxa as represented by phylogenetic (evolutionary) trees.
+Although such models attempt to account for correlations among species, \ml{they often make strong and over-simplified assumptions of the evolutionary process.REWRITE THIS..
+something like complicated and comp inefficient and often neglict.}
 Decades of work have created various different methods that enables researchers to incorporate phylogenetic effects in analyzing species relativeness applications.

### 2018-04-01 names.R

@@ -6,6 +6,6 @@ seed <- as.numeric(targetname[4])
 
 if(size == "small"){nspp <- 20}
 if(size == "med"){nspp <- 100}
-if(size == "large"){nspp <- 500}
+if(size == "large"){nspp <- 200}
 
 datadir <- "datadir/"

### 2018-04-01 phyloglmm_setup.R

@@ -1,6 +1,6 @@
 #### Helper functions for phyloglmm
 
-phylo.to.Z <- function(r){
+phylo.to.Z <- function(r,stand=FALSE){
 	ntip <- length(r$tip.label)
 	Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
 	nodes <- (ntip+1):max(r$edge)
@@ -12,8 +12,10 @@ phylo.to.Z <- function(r){
 			Zid[i,ce] <- 1   ## set Zid to 1
 			cn <- r$edge[ce,1]            ## find previous node
 		}
-	}                                      
+	}
+	sig <- det(vcv(r))^(1/ntip)
 	Z <- t(sqrt(r$edge.length) * t(Zid))
+	if(stand){Z <- t(sqrt(r$edge.length/sig) * t(Zid))}
 	rownames(Z) <- r$tip.label
 	colnames(Z) <- 1:length(r$edge.length)
 	return(Z)                                  

### 2018-04-01 simulate_tree.R

@@ -14,13 +14,13 @@ Vphy <- vcv(phy)
 # Vphy <- Vphy/(det(Vphy)^(1/nspp))
 
 iD <- t(chol(Vphy))
-Xsd <- 20
+Xsd <- 1
 
 # Generate environmental site variable and standardize it
 
 if(nsite == 1){
 	X <- rnorm(n=nspp,sd=Xsd)
-	X <- iD %*% rnorm(n=nspp,sd=Xsd)
+#	X <- iD %*% rnorm(n=nspp,sd=Xsd)
 }
 if(nsite > 1){
 X <- matrix(1:nsite, nrow = 1, ncol = nsite)

### 2018-04-04 main.tex
temp pushing first 2 paragraphs

--- 2018-03-14 main.tex
@@
 
-Phylogenetic comparative methods have been widely used to study species relativeness problems.
-Phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes by incorporating the historical relationships shared across species via phylogenetic (evolutionary) trees.
-\ml{parsimonious can be a confusing word here}
-Although such models attempt to account for species correlations, they often make strong and over-simplified assumptions of the evolutionary process. 
-Decades of work have created various different methods that enables researchers to incorporate phylogenetic effects in analyzing species relativeness applications.
-But many challenges remain.
+Phylogenetic comparative methods (PCM) aim to find parsimonious models that link traits and ecological phenomena with evolutionary processes by incorporating the historical relationships shared among taxa as represented by phylogenetic (evolutionary) trees.
+The primary approach to incorporate this link is by transforming the phylogenetic relationship in the evolutionary tree into species phylogenetic correlation matrices.
+Phylogenetic generalized least square regression (PGLS) and phylogenetic generalized linear mixed models (PGLMM) are classic methods that uses this approach; they have been widely and effectively used.
+However, despite the widespread use of these classic PCMs, many analytic tools available are inflexible and computationally inefficient.
+As a result, researchers have often overlooked and avoided using PCMs. 
+In this paper, we proposed an more flexible and computationally efficient alternative method that incorporates phylogenetic correlations as a Gaussian process, modeling the evolutionary changes along each branch of a phylogenetic tree as independent (possibly multivariate) random Normal deviates.
+
+Ever-increasing capabilities of genetic sequencing techniques are making genomic data available over an ever-wider range of species and also higher resolutions in their evolutionary relationships.
+As a result, researchers are interested in exploring more complicated systems and fitting more complicated models to their data. 
+Given a phylogenetic tree, it is relatively simple to compute a corresponding correlation matrix, which can be incorporated in statistical methods.
+The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}. 
+Although the Brownian motion is an over-simplified model of evolution, many of the existing phylogenetic comparative methods are built upon.
+For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods developed in the last couple of decades and have provided important insights in this field.
+But for technical reasons, all of these approaches are slow (computationally inefficient) or inflexible in terms of degrees of phylogenetic variabilitity.
+
+______
+
+Phylogenetic comparative methods (PCM) aim to find parsimonious models that link traits and ecological phenomena with evolutionary processes by incorporating the historical relationships shared among taxa as represented by phylogenetic (evolutionary) trees.
+Decades of work have created various different methods that enables researchers to incorporate phylogenetic effects in analyzing species relativeness applications, yet existing methods are still have strong contraints.
+In particular, the current stat
+% But many challenges remain.
 In particular, accounting for phylogenetic effects in high dimensional systems is always challenging, especially for models incorporating multiple forms of phylogenetic effects, and especially with large phylogenies where computation powers are limited.
-\ml{Define phylogenetic effects.
+% \ml{Define phylogenetic effects.
 PE can mean a lot of different things (see Pagel's tree transformation).
@@
 y & = XB + \epsilon \\
-\epsilon & \sim MVN(0,\sigma^{2}C),
+\epsilon & \sim \textrm{MVN}(0,\sigma^{2}C),
 \label{eq:gls}

### 2018-04-18 main.tex
intro done for now, fixing methods atm

--- 2018-03-14 main.tex
@@
 
-Phylogenetic comparative methods have been widely used to study species relativeness problems.
-Phylogenetic models aim to find parsimonious methods that can be used to link ecological phenomena with evolutionary processes by incorporating the historical relationships shared across species via phylogenetic (evolutionary) trees.
-\ml{parsimonious can be a confusing word here}
-Although such models attempt to account for species correlations, they often make strong and over-simplified assumptions of the evolutionary process. 
-Decades of work have created various different methods that enables researchers to incorporate phylogenetic effects in analyzing species relativeness applications.
-But many challenges remain.
-In particular, accounting for phylogenetic effects in high dimensional systems is always challenging, especially for models incorporating multiple forms of phylogenetic effects, and especially with large phylogenies where computation powers are limited.
-\ml{Define phylogenetic effects.
-PE can mean a lot of different things (see Pagel's tree transformation).
-What we mean here is strictly phylogenetic signal-- extent to which trait/response values are statistically related to phylogeny.}
-Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \citep{felsenstein1985phylogenies, li2017statistical} -- in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
+Ever-increasing capabilities of data collection techniques (i.e.
+genetic sequencing, geographic information systems monitoring, and etc) and sampling efforts from field biologist are making large volumes of ecological/biological data available over an ever-wider range of species and also higher resolution (i.e.
+measurable phenotypic traits to genetic sequence data).
+As a result, researchers are interested in exploring more complicated systems and fitting more models to data or synthetic databases of species traits.
+One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis; that is, analyzing relationships among species when estimating the effects of species traits within a community.
+Although there are statistical tools available and widely used for this type of problem, these tools are either inflexible, or computationally inefficient and unpractical, or both when fitting large volumes of data or complex models.
+In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relations through random effects.
 
-Incorporating phylogenetic links between species and combine evolutionary changes with statistical processes advances the development of phylogenetic comparative methods.
-The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}. 
-Although the Brownian motion is an over-simplified model of evolution, many of the existing phylogenetic comparative methods are built upon.
-For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods developed in the last couple of decades and have provided important insights in this field.
-Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.
-Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes, Pagel's $\delta$ and $\kappa$ transformations where it explores variation of evolutionary rates and speciational changes through time respectively).
+In the last few decides, researcher have constantly develop new phylogenetic comparative methods and have shown importance of including phylogenetic relationships in these type of problems \cite{felsenstein1985phylogenies, li2017statistical}.
+The core of the existing techniques are based on different modelling assumptions of the evolutionary process using various parametrizations of phylogenetic correlations (PC) of taxa; that is matrices inferred from the topology of evolutionary trees by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods.
+For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods assuming Brownian motion (pure genetic draft) process.
+Despite the wide usage of these methods, there are a handful of recognizable limitations with these approaches and noticeably many published examples in the literature are constrained to small balanced data sets and simple models; that is, phylogenetic variability/signal present in residual variance.
+\citep{li2017canfun}.\ml{add more, should be easy.
+I don't really like this line, skipping for now}
 
-In the past few decades, there have been an accelerating increase in discovering evolutionary history of species making historical evolutionary data accessible and at higher resolutions.
-Furthermore, there have been a steadily advance of methodological research (e.g.
-reconstruct phylogenetic tree methods, and phylogenetic comparative methods) and observational (and meta-analysis) studies on collections of species.
-Despite the widespread use of these methods and ways to account for variability across species, groups, and different studies, there have been relatively few studies accounting for phylogenetic signals.
-
-\ml{write a paragraph on pglmm?
-Flexbility paragraph and say PIC/GLS assumes the topology and branch lengths are error free, Pagel's lambda and Blomberg's K accounts for additional tip variations of the phylogeny, to account for additional variation such as measurement error and such, people turn to Bayesian mixed models \cite{de2012bayesian, hadfield2010general}}
-
-%The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.
-
-%Despite the choice of modelling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis.
-%There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematically look at comparative performance of comparative phylogenetic methods.
-
-In this paper, we propose an alternative, yet straightforward but little-used method of constructing phylogenetic correlations directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
-\ml{flexibility: multiple obs per species and multiple levels of phylogenetic signal}
-We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorporate various complexity of phylogenetic signals in the model and measurement errors. 
-We compare model approaches and explore various complex modelling strategies in the mixed model framework and a flexible platform for researches to explore biologically interesting questions.
-
-We also compare three different R packages: nlme:gls, pez, phylolm, and lme4.
-In principle, any given valid method should converge/get the same result/conclusion.
-However, even with the relatively simple models considered here, methods that are not flexible/parameter constrained can experience problems.
-Furthermore, even when different methods converge to essentially the same result, they may show large differences in computational efficiency.
-\ml{Not sure if we should include the line "we are not using Bayesian approaches" here but definitely in the discussions.}
+In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches. 
+More specifically, we will show our new formulation is more flexible and can be easily extended along with the features of mixed model such as phylogenetically correlated random slopes (i.e.
+the relationship among taxa on changes of predictors/covariates), random interactions \ml{why not?
+we are already using this for the alternative formulation of nested RE} and nested random effect models.
+We will compare our technique with existing methods to show robustness and efficiency via simulations.
+We will end off by discussing opportunities and practicalities of our method and reevaluate the state of art for this area of research.
 
@@
 y & = XB + \epsilon \\
-\epsilon & \sim MVN(0,\sigma^{2}C),
+\epsilon & \sim \textrm{MVN}(0,\sigma^{2}C),
 \label{eq:gls}
 \end{align}
-where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times (m + 1)$ model matrix containing 1's in the first column and the $m$ independent variable (phenotypic traits or environmental variable); $B$ is an $m + 1$ coefficient vector; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
+where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times (m)$ model matrix containing 1's in the first column and the $m$ predictor variables (phenotypic traits or environmental variables); $B$ is an $m$ vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
 
@@
 g(\mu) & = XB + Zb + \epsilon_{mm} \\
-\epsilon_{mm} & \sim MVN(0,\sigma^2)
+\epsilon_{mm} & \sim \textrm{MVN}(0,\sigma^2)
 \end{align}
-where $Z$ is an $n \times (m+1)$ model matrix for the $(m+1)$ -- dimensional vector-valued $m$ independent variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
+where $Z$ is an $n \times (m)$ model matrix for the $(m)$ -- dimensional vector-valued $m$ predictor variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
 Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$ and $Z=Z^{c}$ where the PC is incorporated in the random effect matrix.
 
-\subsection{Phylogenetic Tree Transformation}
+\ml{need to becareful, we actually define Z^C later}
+
+\subsection{Reformulating phylogenetic correlation matrix}
 The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
@@
 
+% 
+% Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.
+Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes, Pagel's $\delta$ and $\kappa$ transformations where it explores variation of evolutionary rates and speciational changes through time respectively).
+% 
+% 
+% 
+% Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.
+Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes, Pagel's $\delta$ and $\kappa$ transformations where it explores variation of evolutionary rates and speciational changes through time respectively).
+% 
+% 
+% 
+% 
+% In particular, when estimating the effects of species traits on their probability of occurrence in a particular site, it is important to account for correlation due to phylogenetic relationships.
+% While this is an interesting and important addition for researchers when fitting more complex models, they often neglect the addition of phylogenetic relationships. \ml{find examples}
+% 
+% In the last few decades, there have been extensive amount of ongoing work in this area of ecological/evolutionary statistics.
+% In and many new methods to incorporate phylogeny information into statistical frameworks. 
+% The classic approach to this problem is to create a matrix that captures all the evolutionary changes that occurred on all the branches inferred from the evolutionary tree.
+% This matrix is often called the "phylogenetic correlation" (PC), which can be incorportated in statistical methods.
+% For example, the classic most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}.
+% 
+% 
+% 
+% 
+% 
+% 
+% 
+% 
+% Phylogenetic comparative methods (PCM) aim to find parsimonious models that link traits and ecological phenomena with evolutionary processes by incorporating the historical relationships shared among taxa as represented by phylogenetic (evolutionary) trees.
+% The primary approach to incorporate this link is by transforming the phylogenetic relationship in the evolutionary tree into species phylogenetic correlation matrices.
+% 
+% Phylogenetic generalized least square regression (PGLS) and phylogenetic generalized linear mixed models (PGLMM) are classic methods that uses this approach; they have been widely and effectively used.
+% However, despite the widespread use of these classic PCMs, many analytic tools available are inflexible and computationally inefficient.
+% As a result, researchers have often overlooked and avoided using PCMs. 
+% 
+% 
+% 
+% In this paper, we proposed an more flexible and computationally efficient alternative method that incorporates phylogenetic correlations as a Gaussian process, modeling the evolutionary changes along each branch of a phylogenetic tree as independent (possibly multivariate) random Normal deviates.
+% 
+% Ever-increasing capabilities of genetic sequencing techniques are making genomic data available over an ever-wider range of species.
+% As a result, researchers are interested in exploring more complicated systems and fitting more models to their data. 
+% Given a phylogenetic tree, it is relatively simple to compute a corresponding correlation matrix, which can be incorporated in statistical methods.
+% The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}. 
+% Although the Brownian motion is an over-simplified model of evolution, many of the existing phylogenetic comparative methods are built upon.
+% For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods developed in the last couple of decades and have provided important insights in this field.
+% But for technical reasons, all of these approaches are slow (computationally inefficient) or inflexible in terms of degrees of phylogenetic variabilitity. 
+% 
+% 
+% Phylogenetic comparative methods (PCM) aim to find parsimonious models that link traits and ecological phenomena with evolutionary processes by incorporating the historical relationships shared among taxa as represented by phylogenetic (evolutionary) trees.
+% Decades of work have created various different methods that enables researchers to incorporate phylogenetic effects in analyzing species relativeness applications, yet existing methods are still have strong contraints.
+% In particular, the current stat
+% % But many challenges remain.
+% In particular, accounting for phylogenetic effects in high dimensional systems is always challenging, especially for models incorporating multiple forms of phylogenetic effects, and especially with large phylogenies where computation powers are limited.
+% % \ml{Define phylogenetic effects.
+PE can mean a lot of different things (see Pagel's tree transformation).
+What we mean here is strictly phylogenetic signal-- extent to which trait/response values are statistically related to phylogeny.}
+% Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \citep{felsenstein1985phylogenies, li2017statistical} -- in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
+% 
+% Incorporating phylogenetic links between species and combine evolutionary changes with statistical processes advances the development of phylogenetic comparative methods.
+% The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}. 
+% Although the Brownian motion is an over-simplified model of evolution, many of the existing phylogenetic comparative methods are built upon.
+% For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods developed in the last couple of decades and have provided important insights in this field.
+% Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.
+Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes, Pagel's $\delta$ and $\kappa$ transformations where it explores variation of evolutionary rates and speciational changes through time respectively).
+% 
+% In the past few decades, there have been an accelerating increase in discovering evolutionary history of species making historical evolutionary data accessible and at higher resolutions.
+% Furthermore, there have been a steadily advance of methodological research (e.g.
+reconstruct phylogenetic tree methods, and phylogenetic comparative methods) and observational (and meta-analysis) studies on collections of species.
+% Despite the widespread use of these methods and ways to account for variability across species, groups, and different studies, there have been relatively few studies accounting for phylogenetic signals.
+% 
+% \ml{write a paragraph on pglmm?
+Flexbility paragraph and say PIC/GLS assumes the topology and branch lengths are error free, Pagel's lambda and Blomberg's K accounts for additional tip variations of the phylogeny, to account for additional variation such as measurement error and such, people turn to Bayesian mixed models \cite{de2012bayesian, hadfield2010general}}
+% 
+% %The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.
+% 
+% 
+% %Despite the choice of modelling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis.
+% %There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematically look at comparative performance of comparative phylogenetic methods.
+% 
+% In this paper, we propose an alternative, yet straightforward but little-used method of constructing phylogenetic correlations directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
+% \ml{flexibility: multiple obs per species and multiple levels of phylogenetic signal}
+% We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorporate various complexity of phylogenetic signals in the model and measurement errors. 
+% We compare model approaches and explore various complex modelling strategies in the mixed model framework and a flexible platform for researches to explore biologically interesting questions.
+% 
+% We also compare three different R packages: nlme:gls, pez, phylolm, and lme4.
+% In principle, any given valid method should converge/get the same result/conclusion.
+% However, even with the relatively simple models considered here, methods that are not flexible/parameter constrained can experience problems.
+% Furthermore, even when different methods converge to essentially the same result, they may show large differences in computational efficiency.
+% \ml{Not sure if we should include the line "we are not using Bayesian approaches" here but definitely in the discussions.}
+%
+
 \section{Conclusion}

### 2018-04-18 main.tex
need to finish platforms and read over results/discussion

@@
 For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods assuming Brownian motion (pure genetic draft) process.
-Despite the wide usage of these methods, there are a handful of recognizable limitations with these approaches and noticeably many published examples in the literature are constrained to small balanced data sets and simple models; that is, phylogenetic variability/signal present in residual variance.
+Despite the wide usage of these methods, there are a handful of recognizable limitations with these approaches and noticeably many published examples in the literature are constrained to small balanced data sets and simple models; that is, simply looking at phylogenetic relationship among species via traits.
 \citep{li2017canfun}.\ml{add more, should be easy.
@@
 More specifically, we will show our new formulation is more flexible and can be easily extended along with the features of mixed model such as phylogenetically correlated random slopes (i.e.
-the relationship among taxa on changes of predictors/covariates), random interactions \ml{why not?
+the relationship among taxa via changes of predictors/covariates on traits), random interactions \ml{why not?
 we are already using this for the alternative formulation of nested RE} and nested random effect models.
-We will compare our technique with existing methods to show robustness and efficiency via simulations.
+We will compare our technique with existing methods to show robustness and efficiency via simulations and data.
 We will end off by discussing opportunities and practicalities of our method and reevaluate the state of art for this area of research.
@@
 \end{align}
-where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times (m)$ model matrix containing 1's in the first column and the $m$ predictor variables (phenotypic traits or environmental variables); $B$ is an $m$ vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
+where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times m$ model matrix containing 1's in the first column and the $m$ predictor variables (phenotypic traits or environmental variables); $B$ is an $m$ vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
 
@@
 \end{align}
-where $Z$ is an $n \times (m)$ model matrix for the $(m)$ -- dimensional vector-valued $m$ predictor variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
-Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$ and $Z=Z^{c}$ where the PC is incorporated in the random effect matrix.
+where $Z$ is an $n \times m$ model matrix for the $m$ -- dimensional vector-valued $m$ predictor variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
+Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$, $Z$ to an $n \times 1$ matrix such that:
 
-\ml{need to becareful, we actually define Z^C later}
+\begin{align}
+Z & \sim \textrm{MVN}(0,\sigma^{2}C).
+\end{align}
 
 \subsection{Reformulating phylogenetic correlation matrix}
-The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
-Phylogenetic independent contrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
-Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.
+% The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
+% Phylogenetic independent contrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
+% Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.
 
@@
 In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
-Thus, the evolutionary history for each species can be model with a sequence of independent errors, rather than having to impose a correlation structure on the random effects.
+Thus, the evolutionary history for each species can be modelled with a sequence of independent errors with species--branch matrix $S$, rather than having to impose a correlation $C$. 
+This is mathematically equvilent under BM assumption where we can reconstruct C by matrix multiplying $S$ by itself.
 
-The random-effect model matrix $Z$ can be decomposed into term-wise model matrix $Z_{i}$ as described in \ml{in text citation lme4}.
+The random-effect model matrix $Z$ can be decomposed into term-wise model matrix $Z_{i}$ as described in \citet{bates2015fitting}.
 Thus, the phylogenetic correlated random-effect matrix $Z^{C}_{i}$ is
@@
 
-\ml{double check if $l and p$ are the same with lme4}
-
 where $\ast$ is the Khatri-Rao product, $S_{i}$ is an $l_{i} \times b_{i}$ matrix of species--branch relations; $J_{i}$ the indicator matrix of grouping factors indices matrix size $n \times l_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
-
-\ml{note: $SS^T = Cov(phylo)$ under BM assumption}
-%\ml{double kronecker for interactions}
 
@@
 
-We generated test data based on the simple mixed model formulation \ml{ref equation (3-5)} with a single response variable $Y$ and a single predictor continuous variable $X$ from $n$ (20, 100, and 500) species, and $m$ (1 and 20) sites.
-To account for multiple levels of phylogenetic signal, we assume phylogenetic signal in the intercept $b_0$, and slope $b_1$.
+We generated test data based on the simple mixed model formulation \ml{ref equation (3-5)} with a single response variable $Y$ and a single predictor continuous variable $X$ for $n$ (20, 100, and 500) species, and $m$ (1 and 20) sites.
+To account for multiple levels of phylogenetic relations, we assume phylogenetic relations in the intercept $b_0$, and slope $b_1$.
 We further assume that random slope and intercept are correlated. 
 For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
-Table \ml{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
+
+\subsection{Platforms}
+
+Our algorithmic approach is general and could be implemented in a wide range of computational platformsthat support independent latent variables, we implemented our approach using the lme4 R package \citep{bates2015fitting}.
+We have included three additional R packages that can fit phylogenetic compariative models, and they are nlme \citep{pinheiro2014r}, phylolm \citep{ho2014phylolm}, and pez \citep{pearse2015pez}.
+Phylogenetic generalized least squares (PGLS) in nlme, is one of the most widely used techniques in phylogenetic compariative analysis; it
+
+% Table \ml{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
 
@@
 
-The full model (which matches the simulations model that incorporates phylogenetic signal on the intercept, slope and correlation) provides good estimates of the different levels of the phylogenetic variation signal (Figure 1 and 2).
+The full model (which matches the simulation model that incorporates phylogenetic signal on the intercept, slope and correlation) provides good estimates of the different levels of the phylogenetic variation signal (Figure 1 and 2).
 
-In general, as the number of species (or phylogeny) increases, all phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) except PGLS in the single site simulations.
+In general, as the number of species increases, all phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) except PGLS in the single site simulations.
 It is unclear how different levels of phylogenetic signal is being confounded under the error free (without tip variation) assumption for the PGLS models.

### 2018-04-18 main.tex
working on results

@@
 As a result, researchers are interested in exploring more complicated systems and fitting more models to data or synthetic databases of species traits.
-One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis; that is, analyzing relationships among species when estimating the effects of species traits within a community.
+One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis; that is, analyzing relationships (signals) among species when estimating the effects of species traits within a community.
 Although there are statistical tools available and widely used for this type of problem, these tools are either inflexible, or computationally inefficient and unpractical, or both when fitting large volumes of data or complex models.
-In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relations through random effects.
+In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic signal via mixed effect models, highlighting various complexities of phylogenetic signals through random effects.
 
-In the last few decides, researcher have constantly develop new phylogenetic comparative methods and have shown importance of including phylogenetic relationships in these type of problems \cite{felsenstein1985phylogenies, li2017statistical}.
+In the last few decides, researcher have constantly develop new phylogenetic comparative methods and have shown importance of including phylogenetic signal in these type of problems \cite{felsenstein1985phylogenies, li2017statistical}.
 The core of the existing techniques are based on different modelling assumptions of the evolutionary process using various parametrizations of phylogenetic correlations (PC) of taxa; that is matrices inferred from the topology of evolutionary trees by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods.
 For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods assuming Brownian motion (pure genetic draft) process.
-Despite the wide usage of these methods, there are a handful of recognizable limitations with these approaches and noticeably many published examples in the literature are constrained to small balanced data sets and simple models; that is, simply looking at phylogenetic relationship among species via traits.
-\citep{li2017canfun}.\ml{add more, should be easy.
-I don't really like this line, skipping for now}
+Despite the wide usage of these methods, there are a handful of recognizable limitations with these approaches and noticeably many published examples in the literature are constrained to small balanced data sets and simple models; that is, simply looking at phylogenetic signal among species via traits.
+\citep{li2017canfun, freckleton2002phylogenetic}.
 
@@
 \begin{align}
-y & = XB + \epsilon \\
-\epsilon & \sim \textrm{MVN}(0,\sigma^{2}C),
-\label{eq:gls}
+y & = XB + \epsilon \label{eq:gls1} \\ 
+\epsilon & \sim \textrm{MVN}(0,\sigma^{2}C), \label{eq:gls2}
 \end{align}
@@
 \begin{align}
-Y & \sim F(\mu) \\
-g(\mu) & = XB + Zb + \epsilon_{mm} \\
-\epsilon_{mm} & \sim \textrm{MVN}(0,\sigma^2)
+Y & \sim F(\mu) \label{eq:glmm1} \\
+g(\mu) & = XB + Zb + \epsilon_{mm} \label{eq:glmm2} \\
+\epsilon_{mm} & \sim \textrm{MVN}(0,\sigma^2) \label{eq:glmm3}
 \end{align}
@@
 \begin{align}
-Z & \sim \textrm{MVN}(0,\sigma^{2}C).
+Z & \sim \textrm{MVN}(0,\sigma^{2}C). \label{eq:glmmgls}
 \end{align}
@@
 In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
-Thus, the evolutionary history for each species can be modelled with a sequence of independent errors with species--branch matrix $S$, rather than having to impose a correlation $C$. 
-This is mathematically equvilent under BM assumption where we can reconstruct C by matrix multiplying $S$ by itself.
+Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $S$, rather than having to impose a correlation $C$. 
+This is mathematically equivalent under BM assumption where we can reconstruct C by matrix multiplying $S$ by itself.
 
@@
 \begin{equation}
-Z^{C}_{i} = (S^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T},
+Z^{C}_{i} = (S^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T}, \label{eq:ZC}
 \end{equation}
@@
 
-We generated test data based on the simple mixed model formulation \ml{ref equation (3-5)} with a single response variable $Y$ and a single predictor continuous variable $X$ for $n$ (20, 100, and 500) species, and $m$ (1 and 20) sites.
-To account for multiple levels of phylogenetic relations, we assume phylogenetic relations in the intercept $b_0$, and slope $b_1$.
+We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $Y$ and a single predictor continuous variable $X$ for $n$ (20, 100, and 500) species, and $m$ (1 and 20) sites/groups.
+To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$.
 We further assume that random slope and intercept are correlated. 
@@
 
-Our algorithmic approach is general and could be implemented in a wide range of computational platformsthat support independent latent variables, we implemented our approach using the lme4 R package \citep{bates2015fitting}.
-We have included three additional R packages that can fit phylogenetic compariative models, and they are nlme \citep{pinheiro2014r}, phylolm \citep{ho2014phylolm}, and pez \citep{pearse2015pez}.
-Phylogenetic generalized least squares (PGLS) in nlme, is one of the most widely used techniques in phylogenetic compariative analysis; it
+Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables, we implemented our approach using the lme4 R package \citep{bates2015fitting}.
+We have included three additional R packages that can fit phylogenetic comparative models under BM process assumption, and they are nlme \citep{pinheiro2014r}, phylolm \citep{ho2014phylolm}, and pez \citep{pearse2015pez}.
+Phylogenetic generalized least squares (PGLS) in \textit{nlme}, is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
+Phylogenetic generalized linear model (PGLM) in \textit{phylolm}, is a slightly more flexible variation of PGLS that can allow for within species variation (sometimes called the tip variation). 
+Neither PGLS nor PGLM can fit random slope or multiple sites. 
+One of the relatively few packages that currently fit random slopes or multiple sites is \textit{pez}, which can handle adding additional uncorrelated random slopes with intercept and species in multiple sites/groups.
 
@@
 First, this new approach is magnitudes faster compared to existing phylogenetic mixed models without losing robustness/accuracy and can easily combined with any statistically mixed model framework. 
-Last and most importantly, it is more flexible in fitting large phylogenies, unbalanced datasets, and complex random effects such as slope correlations.
+Last and most importantly, it is more flexible in fitting large phylogenies, unbalanced data sets, and complex random effects such as slope correlations.
 

### 2018-04-18 main.tex
pushing to bb

@@
 
-In general, as the number of species increases, all phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) except PGLS in the single site simulations.
-It is unclear how different levels of phylogenetic signal is being confounded under the error free (without tip variation) assumption for the PGLS models.
-Models with only two parameters/levels of variation captures the phylogenetic intercept and tip variation reasonably well to the simulated parameters from the full model despite not including phylogenetic random slope and slope intercept correlation.
+In general, all phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) with more data, except PGLS in the single site simulations. 
+It is unclear how different levels of phylogenetic signal are being confounded under the error free (without tip variation) assumption for the PGLS models.
+Models with only two parameters/levels of variation capture the phylogenetic intercept and tip variation reasonably well to the simulated parameters despite not including phylogenetic random slope and slope intercept correlation.
 
 Platforms using the mixed model framework for multiple site simulations follows similar behaviour as single site simulation (i.e.
-as the number of species increase, different levels of phylogenetic signals and variations are closer to the simulated parameters).
-All parameter estimates are much more robust for the multiple site than the single site simulations especially slope-intercept correlations. 
+different levels of phylogenetic signals and variations are closer to the simulated parameters with more data).
+All parameter estimates are much more robust for the multiple site than the single site simulations, especially slope-intercept correlations. 
 There are no clear difference with phylogenetic signal models without random slope and intercept correlation.
@@
 
-We have fitted models including maximum capabilities to model varying levels of variation/phylogenetic signal to simulated data with ....
+We have fitted models including maximum capabilities to model varying levels of variation/phylogenetic signal to simulated data with our simple model with a single response and predictor variable.
 Using models that can include/account for multiple levels of variation is necessary for robust fits, but models that include phylogenetic signal in the intercept level can work as well as the full model.
-Under the Brownian motion evolution process assumption, modelling via independent errors is much more computational efficient than imposing equivalent correlation structures.
+Under the Brownian motion evolution process assumption, modelling via independent errors is much more computationally efficient than imposing equivalent correlation structures.
 
@@
 Our analysis covers the classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models).
-Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, and phylogenetic meta analysis.
-More broadly, the simple independence error approach we developed here offers a more efficient equivalent way to do Brownian motion evolution phylogenetic comparative analysis. 
-This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using Bayesian frameworks such as MCMCglmm. 
-More importantly, this implementation in lme4 allows users to fit phylogenetic mixed models to the fullest that other frequentest platforms?
-cannot (i.e.
+Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-BM evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \cite{butler2004phylogenetic}), Bayesian approaches \cite{hadfield2010general}.
+More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis.
+This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as MCMCglmm \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
+More importantly, this implementation in lme4 allows users to fit phylogenetic mixed models to the fullest that other frequentest platforms lack of (i.e.
 large data, unbalance species observations, complex random effects) and explore new ideas.
-
-% 
-% Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.
-Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes, Pagel's $\delta$ and $\kappa$ transformations where it explores variation of evolutionary rates and speciational changes through time respectively).
-% 
-% 
-% 
-% Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.
-Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes, Pagel's $\delta$ and $\kappa$ transformations where it explores variation of evolutionary rates and speciational changes through time respectively).
-% 
-% 
-% 
-% 
-% In particular, when estimating the effects of species traits on their probability of occurrence in a particular site, it is important to account for correlation due to phylogenetic relationships.
-% While this is an interesting and important addition for researchers when fitting more complex models, they often neglect the addition of phylogenetic relationships. \ml{find examples}
-% 
-% In the last few decades, there have been extensive amount of ongoing work in this area of ecological/evolutionary statistics.
-% In and many new methods to incorporate phylogeny information into statistical frameworks. 
-% The classic approach to this problem is to create a matrix that captures all the evolutionary changes that occurred on all the branches inferred from the evolutionary tree.
-% This matrix is often called the "phylogenetic correlation" (PC), which can be incorportated in statistical methods.
-% For example, the classic most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}.
-% 
-% 
-% 
-% 
-% 
-% 
-% 
-% 
-% Phylogenetic comparative methods (PCM) aim to find parsimonious models that link traits and ecological phenomena with evolutionary processes by incorporating the historical relationships shared among taxa as represented by phylogenetic (evolutionary) trees.
-% The primary approach to incorporate this link is by transforming the phylogenetic relationship in the evolutionary tree into species phylogenetic correlation matrices.
-% 
-% Phylogenetic generalized least square regression (PGLS) and phylogenetic generalized linear mixed models (PGLMM) are classic methods that uses this approach; they have been widely and effectively used.
-% However, despite the widespread use of these classic PCMs, many analytic tools available are inflexible and computationally inefficient.
-% As a result, researchers have often overlooked and avoided using PCMs. 
-% 
-% 
-% 
-% In this paper, we proposed an more flexible and computationally efficient alternative method that incorporates phylogenetic correlations as a Gaussian process, modeling the evolutionary changes along each branch of a phylogenetic tree as independent (possibly multivariate) random Normal deviates.
-% 
-% Ever-increasing capabilities of genetic sequencing techniques are making genomic data available over an ever-wider range of species.
-% As a result, researchers are interested in exploring more complicated systems and fitting more models to their data. 
-% Given a phylogenetic tree, it is relatively simple to compute a corresponding correlation matrix, which can be incorporated in statistical methods.
-% The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}. 
-% Although the Brownian motion is an over-simplified model of evolution, many of the existing phylogenetic comparative methods are built upon.
-% For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods developed in the last couple of decades and have provided important insights in this field.
-% But for technical reasons, all of these approaches are slow (computationally inefficient) or inflexible in terms of degrees of phylogenetic variabilitity. 
-% 
-% 
-% Phylogenetic comparative methods (PCM) aim to find parsimonious models that link traits and ecological phenomena with evolutionary processes by incorporating the historical relationships shared among taxa as represented by phylogenetic (evolutionary) trees.
-% Decades of work have created various different methods that enables researchers to incorporate phylogenetic effects in analyzing species relativeness applications, yet existing methods are still have strong contraints.
-% In particular, the current stat
-% % But many challenges remain.
-% In particular, accounting for phylogenetic effects in high dimensional systems is always challenging, especially for models incorporating multiple forms of phylogenetic effects, and especially with large phylogenies where computation powers are limited.
-% % \ml{Define phylogenetic effects.
-PE can mean a lot of different things (see Pagel's tree transformation).
-What we mean here is strictly phylogenetic signal-- extent to which trait/response values are statistically related to phylogeny.}
-% Ignoring phylogenetic effects/signals can be problematic both biologically and statistically \citep{felsenstein1985phylogenies, li2017statistical} -- in such cases, researchers often use simpler models accounting for minimal phylogenetic signals by limiting at the individuals level and neglect additional phylogenetic signals present in the system.
-% 
-% Incorporating phylogenetic links between species and combine evolutionary changes with statistical processes advances the development of phylogenetic comparative methods.
-% The most widely used statistical process in phylogenetic comparative analysis to model evolutionary process is the pure random drift process, Brownian motion \citep{felsenstein1973maximum}. 
-% Although the Brownian motion is an over-simplified model of evolution, many of the existing phylogenetic comparative methods are built upon.
-% For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods developed in the last couple of decades and have provided important insights in this field.
-% Nevertheless, there are also a handful of selection beyond the pure random drift (BM) process where they probe specific aspects of the evolutionary process (i.e.
-Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes, Pagel's $\delta$ and $\kappa$ transformations where it explores variation of evolutionary rates and speciational changes through time respectively).
-% 
-% In the past few decades, there have been an accelerating increase in discovering evolutionary history of species making historical evolutionary data accessible and at higher resolutions.
-% Furthermore, there have been a steadily advance of methodological research (e.g.
-reconstruct phylogenetic tree methods, and phylogenetic comparative methods) and observational (and meta-analysis) studies on collections of species.
-% Despite the widespread use of these methods and ways to account for variability across species, groups, and different studies, there have been relatively few studies accounting for phylogenetic signals.
-% 
-% \ml{write a paragraph on pglmm?
-Flexbility paragraph and say PIC/GLS assumes the topology and branch lengths are error free, Pagel's lambda and Blomberg's K accounts for additional tip variations of the phylogeny, to account for additional variation such as measurement error and such, people turn to Bayesian mixed models \cite{de2012bayesian, hadfield2010general}}
-% 
-% %The most widely used statistical method to model related species data that accounts for the phylogenetic structure is the \textit{phylogenetic regression} via independent constrast \ml{cite Felsenstein 1985} and further generalizations to generalized least square regression (PGLS) and generalized linear mixed models (PGLMM) to allow for different models of trait evolution \ml{or types of response?}.
-% 
-% 
-% %Despite the choice of modelling strageties, there are many limitations in the current state of art of statistical phylogenetic analysis.
-% %There are a handful of studies discussing the importance of incorportating phylogenetic signals but relatively few systematically look at comparative performance of comparative phylogenetic methods.
-% 
-% In this paper, we propose an alternative, yet straightforward but little-used method of constructing phylogenetic correlations directly from phylogenetic tree by summing the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
-% \ml{flexibility: multiple obs per species and multiple levels of phylogenetic signal}
-% We apply relatively simple phylogenetic comparative approaches to data from simulated phylogenies that incorporate various complexity of phylogenetic signals in the model and measurement errors. 
-% We compare model approaches and explore various complex modelling strategies in the mixed model framework and a flexible platform for researches to explore biologically interesting questions.
-% 
-% We also compare three different R packages: nlme:gls, pez, phylolm, and lme4.
-% In principle, any given valid method should converge/get the same result/conclusion.
-% However, even with the relatively simple models considered here, methods that are not flexible/parameter constrained can experience problems.
-% Furthermore, even when different methods converge to essentially the same result, they may show large differences in computational efficiency.
-% \ml{Not sure if we should include the line "we are not using Bayesian approaches" here but definitely in the discussions.}
-%
 
@@
 
-We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical pcm to simulated data. 
+We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical PCM to simulated data. 
 First, this new approach is magnitudes faster compared to existing phylogenetic mixed models without losing robustness/accuracy and can easily combined with any statistically mixed model framework. 
-Last and most importantly, it is more flexible in fitting large phylogenies, unbalanced data sets, and complex random effects such as slope correlations.
+Last and most importantly, it is more flexible in fitting large phylogenies, large volume of data, unbalanced data sets, and complex random effects such as slope correlations.
 

### 2018-05-02 main.tex
temp pushing bits and pieces

@@
 As a result, researchers are interested in exploring more complicated systems and fitting more models to data or synthetic databases of species traits.
-One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis; that is, analyzing relationships (signals) among species when estimating the effects of species traits within a community.
-Although there are statistical tools available and widely used for this type of problem, these tools are either inflexible, or computationally inefficient and unpractical, or both when fitting large volumes of data or complex models.
+One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis using eco-evolutionary regression; that is, analyzing evolutionary relationships (signals) among species between species traits and environmental variables when estimating the effects of traits within a community.
+Although there are sophisticated statistical tools specially designed for this type of problem, the current state of the art of these tools are either inflexible, or computationally inefficient and unpractical when fitting large volumes of data.
+In particular, modelling and understanding different levels of variation/errors is always challenging both biologically and computationally, especially for models with many predictor variables, and especially accounting for phylogenetically relationships with multiple predictor variables.
+In such case, researchers often try to find ways to simplify or revert to simpler models for practical uses which can lead to both statiscially and biological issues.  
 In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic signal via mixed effect models, highlighting various complexities of phylogenetic signals through random effects.
+
+\ml{add a paragraph on different levels of variations with random slope examples listed below}
+- random slope but not random intercept\cite{nowakowski2018phylogenetic}
+- can put phylo but did not\cite{bunnefeld2012island}
+- taxa grouping \cite{ord2010adaptation}
 

### 2018-05-02 vignette.Rmd
temp pushing bits and pieces

---
title: "An introduction to pglmm in lme4"
author: "Michael Li and Ben Bolker"
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
output: pdf_document
---

# 1 Preamble

You can install lme4 by the developer version on github using devtools ``devtools:::install_github(...)''

# Phylogenetic Generalised Linear Mixed Models

What follows is a simple example of how to simulate, and then analyse, the most basic kind of PGLMM.

## Simulate data

```{r}
### simulate phylogenetic tree

library(ape)
## library(MASS)  ## for mvrnorm() ## don't load so we don't screw up dplyr::select()
seed <- 101
set.seed(seed)

# Basic parameters
nspp <- 3

# # fixed effects
beta0 <- 0
beta1 <- 0

# random effects and residual variance (set to zero for binary data)
sd.B0 <- 1
sd.B1 <- 2
sd.resid <- 10
rho.B01 <- 0

# Simulate phylogeny
phy <- rtree(n = nspp)
phy <- compute.brlen(phy, method = "Grafen", power = 0.5)

Vphy <- vcv(phy)

iD <- t(chol(Vphy))
Xsd <- 1

# Generate environmental variable 
x <- rnorm(n=nspp,sd=Xsd)

cormat <- matrix(c(1,rho.B01,rho.B01,1),2,2)
sdvec <- c(sd.B0,sd.B1)
varmat <- sdvec %*% t(sdvec)
covmat <- varmat * cormat

#print(covmat)

Sigma <- kronecker(covmat,Vphy)

#if((signal.B0==signal.B1) & (signal.B0 == FALSE)){
#	Sigma <- kronecker(covmat,diag(nspp))
#}

b.all <- MASS::mvrnorm(n=1
	, mu=rep(c(beta0,beta1),each=nspp)
	, Sigma=Sigma
)

b0 <- b.all[1:nspp] 
b1 <- b.all[(nspp+1):(2*nspp)]

y <- matrix(outer(b0, array(1, dim = c(1, 1))), nrow = nspp,
	ncol = 1) + matrix(b1*x, nrow = nspp, ncol = 1)

e <- rnorm(nspp, sd = sd.resid) # add residual variance

Y <- y + matrix(e, nrow = nspp, ncol = 1)

# name the simulated species 1:nspp and sites 1:nsites
rownames(Y) <- 1:nspp

X <- matrix(x,nrow=nspp,ncol=1)

sp <- matrix(kronecker(matrix(1, nrow = 1, ncol = 1), 1:nspp),
             nrow = nspp, ncol = 1)

dat <- data.frame(Y = Y, X=X, sp = as.factor(sp))

print(dim(dat))

```

## Species branch matrix

- call this SBmat

```{r}

library(Matrix)
library(lme4)
library(dplyr)

source("phyloglmm_setup.R")

phyZ <- phylo.to.Z(phy)

print(phyZ)

```

## Fitting phyloglmm

- I kind of want to keep the orginal name (lmer and glmer).

```{r}

source("new_phylo_setup.R")

dat <- (dat %>% mutate(obs = sp))

## everything can be reduce down to formula,data, and a phy object

fitmod <- phylo_lmm2(Y ~ X + (1 |sp) + (1 + X|sp)
			, data=dat 
			, phylonm = "sp" # wishy-washy
			, phyloZ = phyZ # Maybe we just want the phy object and compute this internally
			, nsp = nspp # make default check
			, control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")  ## make default
			, REML = FALSE # make default 
)

summary(fitmod)

```

## Simulating using lme4's simulation function

- first fit the object and hack the parameters

### 2018-05-02 parameters.R

@@ -12,7 +12,7 @@ nsite <- 20
 #nsite <- 1
 
 # residual variance (set to zero for binary data)
- sd.resid <- 10
+sd.resid <- 10
 # sd.resid <- 0.000000001
 # # fixed effects
 beta0 <- 0

### 2018-05-07 main.tex
temp pushing for now, rewriting the intro

@@
 measurable phenotypic traits to genetic sequence data).
-As a result, researchers are interested in exploring more complicated systems and fitting more models to data or synthetic databases of species traits.
-One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis using eco-evolutionary regression; that is, analyzing evolutionary relationships (signals) among species between species traits and environmental variables when estimating the effects of traits within a community.
+As a result, researchers are interested in exploring ways to study more complicated systems and fitting more models to data or synthetic databases of species traits.
+One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression; that is, analyzing evolutionary relationships (signals) among species between species traits and environmental variables when estimating the effects of traits within a community.
 Although there are sophisticated statistical tools specially designed for this type of problem, the current state of the art of these tools are either inflexible, or computationally inefficient and unpractical when fitting large volumes of data.
-In particular, modelling and understanding different levels of variation/errors is always challenging both biologically and computationally, especially for models with many predictor variables, and especially accounting for phylogenetically relationships with multiple predictor variables.
-In such case, researchers often try to find ways to simplify or revert to simpler models for practical uses which can lead to both statiscially and biological issues.  
-In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic signal via mixed effect models, highlighting various complexities of phylogenetic signals through random effects.
+In particular, modelling and understanding relationships between response and multiple predictor variables, and accounting for different levels of variation (errors) to complicated systems with large data is always challenging both biologically and statistically, and adding phylogenetic relationships will further increases the difficulties.
+In such case, researchers often try to find ways to simplify or revert to simpler models (i.e.
+model reduction or ignore phylogenetic relationships, or both) for practical uses which can lead to both statistical and biological issues \cite{felsenstein1985phylogenies, li2017statistical}.
+\ml{In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
+We will use this modelling framework to present and discuss importance of different levels of variations and their considerations and interpretations in phylogenetic comparative analysis.}
+\ml{move this to the last paragraph in the intro?}
+
+Residual variation (or variation in residuals) is a serious concern for most comparative studies (e.g., \cite{pagel1988recent, kelly2004comparative,ives2007within} and it has different, yet correct interpretations in various phylogenetic comparative literature.
+For example, the standard/classic phylogenetic regression is a correlated-residual model using phylogenetically independent contrasts (PICs), where the residuals evolve as a Brownian motion process \citep{felsenstein1985phylogenies}; in other words, the variation of residuals are phylogenetically correlated.
+Many of the more recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} -- are built upon Felsentein's PICs model by making additional sophisticated assumptions to correct for possible bias where separates/partition phylogenetically correlated residual variations into phylogenetically uncorrelated residual variation (observation error or tip variation) and phylogenetic signal (biological/evolutionary process error) \cite{hansen2012interpreting}.
+Lastly, the standard statistical residual variation of repeated measures or multiple observations of the same taxon, where phylogenetic signal and tip variation are now both part of the evolutionary process error.
+The last case is straightforward statistically, but rarely seen in the literature and existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
+
+--- stop here for now.
 

### 2018-05-08 main.tex
first pass of the new intro, it is at the right length and a new story

--- 2018-04-18 main.tex
@@
 measurable phenotypic traits to genetic sequence data).
-As a result, researchers are interested in exploring more complicated systems and fitting more models to data or synthetic databases of species traits.
-One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis; that is, analyzing relationships (signals) among species when estimating the effects of species traits within a community.
-Although there are statistical tools available and widely used for this type of problem, these tools are either inflexible, or computationally inefficient and unpractical, or both when fitting large volumes of data or complex models.
-In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic signal via mixed effect models, highlighting various complexities of phylogenetic signals through random effects.
+As a result, researchers are interested in exploring ways to study more complicated systems and fitting more models to data or synthetic databases of species traits.
+One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression; that is, analyzing evolutionary relationships (signals) among species between species traits and environmental variables when estimating the effects of traits within a community.
+Although there are sophisticated statistical tools specially designed for this type of problem, the current state of the art of these tools are either inflexible, or computationally inefficient and unpractical when fitting large volumes of data.
+In particular, modelling and understanding relationships between response and multiple predictor variables, and accounting for different levels of variation (errors) to complicated systems with large data is always challenging both biologically and statistically, and adding phylogenetic relationships will further increases the difficulties.
+In such case, researchers often try to find ways to simplify or revert to simpler models (i.e.
+model reduction or ignore phylogenetic relationships, or both) for practical uses \cite{bunnefeld2012island, ord2010adaptation}, which can lead to both statistical and biological issues \cite{felsenstein1985phylogenies, li2017statistical}.
+In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
+We will use this modelling framework to present and discuss importance of different levels of variations and their considerations and interpretations in phylogenetic comparative analysis.
 
-In the last few decides, researcher have constantly develop new phylogenetic comparative methods and have shown importance of including phylogenetic signal in these type of problems \cite{felsenstein1985phylogenies, li2017statistical}.
-The core of the existing techniques are based on different modelling assumptions of the evolutionary process using various parametrizations of phylogenetic correlations (PC) of taxa; that is matrices inferred from the topology of evolutionary trees by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods.
-For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods assuming Brownian motion (pure genetic draft) process.
-Despite the wide usage of these methods, there are a handful of recognizable limitations with these approaches and noticeably many published examples in the literature are constrained to small balanced data sets and simple models; that is, simply looking at phylogenetic signal among species via traits.
-\citep{li2017canfun, freckleton2002phylogenetic}.
+There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
+First, a method for linking phylogenetic relationships among species when estimating the effects of traits. 
+The standard/classic phylogenetic regression is a correlated-residual model using phylogenetically independent contrasts (PICs), where the residuals evolve as a Brownian motion process \citep{felsenstein1985phylogenies}; in other words, the variation of residuals are phylogenetically correlated.
+Many of the more recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} -- are built upon Felsentein's PICs model by making additional sophisticated assumptions allowing extra parameters to correct for possible bias where separates/partitions the phylogenetically correlated residual variation into phylogenetically uncorrelated residual variation (observation error or tip variation) and phylogenetic signal (biological/evolutionary process error) \cite{hansen2012interpreting}.
+However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error and residual variation is associated with observation level variation (the standard statistical definition).
+Despite the addition of multiple observations being straightforward statistically, it is rarely seen in the literature as existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
+
+The second challenge is flexibly-incorportating modelling phylogenetic relationships to other common statistical features with many predictor variables in phylogenetic regression (i.e.
+phylogenetically correlated slopes, interactions, nested designs and etc.)
+For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
+The tools available for extending phylogenetic relationships to common statistical regression features are very limited, inflexible and computationally inefficient; thus, sophisticated biologist would turn to more flexible settings (e.g.
+Bayesian models).
 
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches. 
-More specifically, we will show our new formulation is more flexible and can be easily extended along with the features of mixed model such as phylogenetically correlated random slopes (i.e.
-the relationship among taxa via changes of predictors/covariates on traits), random interactions \ml{why not?
+More specifically, we will show our new formulation is more flexible and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetically phylogenetic relationship on other predictor variables with random slopes (i.e.
+phylogenetic signal in response to change in the predictor variable), random interactions \ml{why not?
 we are already using this for the alternative formulation of nested RE} and nested random effect models.

### 2018-05-11 main.tex
pushing another version

@@
 The standard/classic phylogenetic regression is a correlated-residual model using phylogenetically independent contrasts (PICs), where the residuals evolve as a Brownian motion process \citep{felsenstein1985phylogenies}; in other words, the variation of residuals are phylogenetically correlated.
-Many of the more recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} -- are built upon Felsentein's PICs model by making additional sophisticated assumptions allowing extra parameters to correct for possible bias where separates/partitions the phylogenetically correlated residual variation into phylogenetically uncorrelated residual variation (observation error or tip variation) and phylogenetic signal (biological/evolutionary process error) \cite{hansen2012interpreting}.
+Many of the more recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} -- are built upon Felsentein's PICs model by making additional sophisticated assumptions allowing extra parameters to correct for possible bias where they separate/partition the phylogenetically correlated residual variation into phylogenetically uncorrelated residual variation (observation error or tip variation) and phylogenetic signal (biological/evolutionary process error) \cite{hansen2012interpreting}.
 However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error and residual variation is associated with observation level variation (the standard statistical definition).
@@
 
-The second challenge is flexibly-incorportating modelling phylogenetic relationships to other common statistical features with many predictor variables in phylogenetic regression (i.e.
+The second challenge is flexibly-incorporating modelling phylogenetic relationships to other common statistical features with many predictor variables in phylogenetic regression (i.e.
 phylogenetically correlated slopes, interactions, nested designs and etc.)
@@
 where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times m$ model matrix containing 1's in the first column and the $m$ predictor variables (phenotypic traits or environmental variables); $B$ is an $m$ vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
+The PC matrix is inferred from the topology of the evolutionary tree by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods.
 
@@
 
+\ml{add a paragraph about dune fits on real data}
+
 \newpage
@@
 
-\subsection{Process and observation errors}
+\subsection{Process and residual (observation) variabilities and confoundings}
 
-Another way to think about various forms/sources of phylogenetic signals and uncertainties is characterizing into process and observation errors.
-Process error or in this context, variation contributed from the evolutionary matches nicely with the variation of phylogenetic signal. 
-Observation/measurement error on the other hand is harder to match in this context and data dependent.
-In the classic examples, observation error is (assume to be?) zero for PIC and PGLS; single-site observation model using phylolm (or any method with two variables describing the variance (Pagel's $\lambda$, Blomberg's K)) often match the observation/measurement error with the tip variation of the phylogeny.
-For multiple observations per species, the tip variation becomes one of the process errors with phylogenetic signal and the observation error is the variation among species and additional other non-process errors.
+Process and observation errors are highly confounded in phylogenetic regression.
+The effects of models that include both process and residual variations (i.e.
+pglmm, pglm) instead of the less realistic model that assumes variation due to evolution and zero observation (gls) was large.
+In particular, models that did not include residual variation tended to over-estimate the amount of variabilities in the evolution process. 
+Thus, any model that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e.
+Pagel's lambda, Blomberg's K).
+
+\subsection{Random Slopes}
+
+In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2013nested, cleasby2015quantifying}.
+When analyzing evolutionary process among species between species traits and environmental variables, it is entirely plausible that the relationship across different traits and predictor variables are varying in a phylogenetically correlated way.
+Although the effects of including phylogenetically correlated random slopes are relatively low compared to process and residual variation in the intercept, it is still useful and informative than ignoring it from the model.
+It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \cite{ord2010adaptation}) - the PGLMM collapses to a standard random-slopes model.
+\ml{which would only be identifiable if ...}
+However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be *theoretically* identifiable.
 
@@
 This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as MCMCglmm \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
-More importantly, this implementation in lme4 allows users to fit phylogenetic mixed models to the fullest that other frequentest platforms lack of (i.e.
+However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit.
+\ml{This note should go somewhere here: how much data do we need in order to practically estimate the random slopes?
+Are we making a mistake (cf Schielzeth) by ignoring random slopes?
+What should we do when we don't have enough data?
+(cf.
+Barr et al 2013 "keep it maximal"; Bates, Vashisth, etc etc who don't like that, possibility of Bayesian approaches, etc etc etc)}
+More importantly, this implementation in lme4 allows users to fit phylogenetic mixed models to the fullest ( that other frequentest platforms lack of (i.e.
 large data, unbalance species observations, complex random effects) and explore new ideas.

### 2018-05-16 main.tex
working on ms

@@
 
-Ever-increasing capabilities of data collection techniques (i.e.
-genetic sequencing, geographic information systems monitoring, and etc) and sampling efforts from field biologist are making large volumes of ecological/biological data available over an ever-wider range of species and also higher resolution (i.e.
-measurable phenotypic traits to genetic sequence data).
+Ever-increasing data collection capabilities (such as genomic sequencing, telemetry studies of animal behaviour, or remote sensing of environmental variables), in combination with large-scale synthesis of existing data bases on species occurrence and phenotypic traits, are making large volumes of biological data available over an ever-wider taxonomic range.
 As a result, researchers are interested in exploring ways to study more complicated systems and fitting more models to data or synthetic databases of species traits.
-One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression; that is, analyzing evolutionary relationships (signals) among species between species traits and environmental variables when estimating the effects of traits within a community.
-Although there are sophisticated statistical tools specially designed for this type of problem, the current state of the art of these tools are either inflexible, or computationally inefficient and unpractical when fitting large volumes of data.
-In particular, modelling and understanding relationships between response and multiple predictor variables, and accounting for different levels of variation (errors) to complicated systems with large data is always challenging both biologically and statistically, and adding phylogenetic relationships will further increases the difficulties.
+One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression.
+Comparative analyses explore relationships among species traits (including occurrence within communities) while taking evolutionary relationships into account; in some cases, researchers focus on exploring the evolutionary signals in the traits rather than simply controlling for them.
+Although a wide range of sophisticated statistical tools for comparative analyses exist, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
 In such case, researchers often try to find ways to simplify or revert to simpler models (i.e.
-model reduction or ignore phylogenetic relationships, or both) for practical uses \cite{bunnefeld2012island, ord2010adaptation}, which can lead to both statistical and biological issues \cite{felsenstein1985phylogenies, li2017statistical}.
-In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
-We will use this modelling framework to present and discuss importance of different levels of variations and their considerations and interpretations in phylogenetic comparative analysis.
+model reduction or ignore phylogenetic relationships, or both) for practical uses \cite{bunnefeld2012island, ord2010adaptation},
+\todo{what do these papers do?}
+which can lead to both statistical and biological issues \cite{felsenstein1985phylogenies, li2017statistical}. 
+In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
 
 There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
-First, a method for linking phylogenetic relationships among species when estimating the effects of traits. 
-The standard/classic phylogenetic regression is a correlated-residual model using phylogenetically independent contrasts (PICs), where the residuals evolve as a Brownian motion process \citep{felsenstein1985phylogenies}; in other words, the variation of residuals are phylogenetically correlated.
-Many of the more recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} -- are built upon Felsentein's PICs model by making additional sophisticated assumptions allowing extra parameters to correct for possible bias where they separate/partition the phylogenetically correlated residual variation into phylogenetically uncorrelated residual variation (observation error or tip variation) and phylogenetic signal (biological/evolutionary process error) \cite{hansen2012interpreting}.
+First, we need a method for incorporating phylogenetic relationships among species into the model of the relationships among traits
+The classic phylogenetic regression assumes that phylogenetic correlation among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; this assumption leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
+More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon Felsenstein's PICs by allowing extra parameters to correct for possible bias.
+They partition the phylogenetically correlated residual variation into phylogenetically uncorrelated residual variation (observation error or tip variation) and phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting} \todo{do PGLMMs do this, or just Blomberg/Pagel?
+How do B/P differ?}
 However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error and residual variation is associated with observation level variation (the standard statistical definition).
@@
 
-The second challenge is flexibly-incorporating modelling phylogenetic relationships to other common statistical features with many predictor variables in phylogenetic regression (i.e.
-phylogenetically correlated slopes, interactions, nested designs and etc.)
-For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
-The tools available for extending phylogenetic relationships to common statistical regression features are very limited, inflexible and computationally inefficient; thus, sophisticated biologist would turn to more flexible settings (e.g.
-Bayesian models).
+The second challenge is flexibly incorporating models of phylogenetic relationships along with other common features of statistical models, such as variation among multiple different grouping variables or variation in effects (slopes) as well as correlated residual variation.
+For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
+The tools available for extending phylogenetic relationships to common statistical regression features are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \todo{cite brms, MCMCglmm}
 
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches.
-More specifically, we will show our new formulation is more flexible and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetically phylogenetic relationship on other predictor variables with random slopes (i.e.
-phylogenetic signal in response to change in the predictor variable), random interactions \ml{why not?
+However, our new formulation is more flexible and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, random interactions \ml{why not?
 we are already using this for the alternative formulation of nested RE} and nested random effect models.
 We will compare our technique with existing methods to show robustness and efficiency via simulations and data.
-We will end off by discussing opportunities and practicalities of our method and reevaluate the state of art for this area of research.
+We end by discussing opportunities and practicalities of our method and considering the state of art for this area of research.
 
@@
 More importantly, this implementation in lme4 allows users to fit phylogenetic mixed models to the fullest ( that other frequentest platforms lack of (i.e.
-large data, unbalance species observations, complex random effects) and explore new ideas.
+large data, unbalanced species observations, complex random effects) and explore new ideas.
 

### 2018-05-23 main.tex
bmb changes

@@
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches.
-However, our new formulation is more flexible and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, random interactions \ml{why not?
-we are already using this for the alternative formulation of nested RE} and nested random effect models.
+However, our new formulation is more flexible and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
 We will compare our technique with existing methods to show robustness and efficiency via simulations and data.

### 2018-06-07 main.tex
reference not working

@@
 \todo{what do these papers do?}
+\todo{bunnefeld2012island : used taxon for random slopes but did not include phylogenetic information.
+ord2010adaptation : used OU procress with means (single obs per taxa)}
 which can lead to both statistical and biological issues \cite{felsenstein1985phylogenies, li2017statistical}. 
@@
 There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
-First, we need a method for incorporating phylogenetic relationships among species into the model of the relationships among traits
+First, we need a method for incorporating phylogenetic relationships among species into the model of the relationships among traits.
 The classic phylogenetic regression assumes that phylogenetic correlation among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; this assumption leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.

### 2018-06-07 main.tex
finish comments in the intro, moving on to methods

@@
 As a result, researchers are interested in exploring ways to study more complicated systems and fitting more models to data or synthetic databases of species traits.
-One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression.
+One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression \cite{hansen2012interpreting}. 
 Comparative analyses explore relationships among species traits (including occurrence within communities) while taking evolutionary relationships into account; in some cases, researchers focus on exploring the evolutionary signals in the traits rather than simply controlling for them.
@@
 More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon Felsenstein's PICs by allowing extra parameters to correct for possible bias.
-They partition the phylogenetically correlated residual variation into phylogenetically uncorrelated residual variation (observation error or tip variation) and phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting} \todo{do PGLMMs do this, or just Blomberg/Pagel?
+They partition the phylogenetically correlated residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting} \todo{do PGLMMs do this, or just Blomberg/Pagel?
 How do B/P differ?}
+\todo{They all do this. B's K is a scaled ratio of of variance, and P's lambda is a scaling parameter of the correlation.} 
 However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error and residual variation is associated with observation level variation (the standard statistical definition).
@@
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
-The tools available for extending phylogenetic relationships to common statistical regression features are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \todo{cite brms, MCMCglmm}
+The tools available for extending phylogenetic relationships to common statistical regression features are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \cite{hadfield2010mcmc, burkner2016brms}.
 

### 2018-06-08 main.tex
I think I am done with methods for now

@@
 First, we need a method for incorporating phylogenetic relationships among species into the model of the relationships among traits.
-The classic phylogenetic regression assumes that phylogenetic correlation among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; this assumption leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
+The classic phylogenetic regression assumes that phylogenetic correlation (PC) among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; this assumption leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
 More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon Felsenstein's PICs by allowing extra parameters to correct for possible bias.
@@
 \end{align}
-where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times m$ model matrix containing 1's in the first column and the $m$ predictor variables (phenotypic traits or environmental variables); $B$ is an $m$ vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).
+where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times m$ model matrix containing 1's in the first column and the $m$ predictor variables (phenotypic traits or environmental variables); $B$ is an $m$ vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation.
 The PC matrix is inferred from the topology of the evolutionary tree by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods.
@@
 Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $S$, rather than having to impose a correlation $C$.
+
+$S$ is a matrix product of an $n \times b$ indicator matrix $S_{id}$ of branch indices, and an adjusted branch length vector $B_{adj}$:
+
+\begin{align}
+B_{adj} & = (\frac{Bl}{\omega})^{1/2} \\
+\omega & = \textrm{det}(\Sigma)^{1/n},
+\end{align}
+where $Bl$ is the branch length vector, $\omega$ is a scaling factor, and $\Sigma$ is the variance-covariance matrix of the phylogenetic tree.
 This is mathematically equivalent under BM assumption where we can reconstruct C by matrix multiplying $S$ by itself.

### 2018-06-08 main.tex
finish with results, need to draft supplement
temp pushing, almost done discussion

@@
 
-\ml{add a paragraph about dune fits on real data}
+We used our method and fitted the full model of the dune meadow data recently used with pez \cite(li2017canfun}. 
+We have obtained identical results for fix and random effect estimates and approximately 120 times faster compared to pez. 
+Codes for fitting dune meadow analysis using lme4 and pez are provided in the supplements.
 
@@
 
-In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2013nested, cleasby2015quantifying}.
+In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying}.
 When analyzing evolutionary process among species between species traits and environmental variables, it is entirely plausible that the relationship across different traits and predictor variables are varying in a phylogenetically correlated way.
 Although the effects of including phylogenetically correlated random slopes are relatively low compared to process and residual variation in the intercept, it is still useful and informative than ignoring it from the model.
-It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \cite{ord2010adaptation}) - the PGLMM collapses to a standard random-slopes model.
-\ml{which would only be identifiable if ...}
+It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \cite{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
 However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be *theoretically* identifiable.

### 2018-06-24 main.tex
fix citation typo
done editing ms, making slides for ISEC now

@@
 
-We used our method and fitted the full model of the dune meadow data recently used with pez \cite(li2017canfun}. 
+We used our method and fitted the full model of the dune meadow data recently used with pez \cite{li2017canfun}. 
 We have obtained identical results for fix and random effect estimates and approximately 120 times faster compared to pez. 
@@
 The effects of models that include both process and residual variations (i.e.
-pglmm, pglm) instead of the less realistic model that assumes variation due to evolution and zero observation (gls) was large.
+PGLMM, PGLM) instead of the less realistic model that assumes variation due to evolution and zero observation (PGLS) was large.
 In particular, models that did not include residual variation tended to over-estimate the amount of variabilities in the evolution process. 
 Thus, any model that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e.
-Pagel's lambda, Blomberg's K).
+Pagel's $\lambda$, Blomberg's $K$).
 
@@
 It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \cite{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
+Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
+and Are we making a mistake by ignoring random slopes \cite{schielzeth2008conclusions}?
 However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be *theoretically* identifiable.
@@
 This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as MCMCglmm \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
-However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit.
-\ml{This note should go somewhere here: how much data do we need in order to practically estimate the random slopes?
-Are we making a mistake (cf Schielzeth) by ignoring random slopes?
+However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e.
 What should we do when we don't have enough data?
-(cf.
-Barr et al 2013 "keep it maximal"; Bates, Vashisth, etc etc who don't like that, possibility of Bayesian approaches, etc etc etc)}
+Should we use a complex model and overfit \cite{barr2013random}, the right balance of complexity and data \cite{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
 More importantly, this implementation in lme4 allows users to fit phylogenetic mixed models to the fullest ( that other frequentest platforms lack of (i.e.

### 2018-07-10 main.tex
working on ms

@@
 One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression \cite{hansen2012interpreting}. 
-Comparative analyses explore relationships among species traits (including occurrence within communities) while taking evolutionary relationships into account; in some cases, researchers focus on exploring the evolutionary signals in the traits rather than simply controlling for them.
+Comparative analyses explore relationships among species traits (including occurrence within communities) while taking evolutionary relationships into account; in some cases, researchers are interested in quantifying the evolutionary signals underly traits rather than simply controlling for them.
 Although a wide range of sophisticated statistical tools for comparative analyses exist, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
-In such case, researchers often try to find ways to simplify or revert to simpler models (i.e.
-model reduction or ignore phylogenetic relationships, or both) for practical uses \cite{bunnefeld2012island, ord2010adaptation},
-\todo{what do these papers do?}
-\todo{bunnefeld2012island : used taxon for random slopes but did not include phylogenetic information.
-ord2010adaptation : used OU procress with means (single obs per taxa)}
-which can lead to both statistical and biological issues \cite{felsenstein1985phylogenies, li2017statistical}. 
+In such cases, researchers may try to find ways to simplify their analyses (i.e. simplifying the model or ignoring phylogenetic relationships, or both) \citep{bunnefeld2012island, ord2010adaptation},
+%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
+%% ord2010adaptation : used OU process with means (single obs per taxa),
+%%  i.e. ignoring within-taxon variation?
+which can lead to both statistical and biological issues
+\cite{felsenstein1985phylogenies, li2017statistical}.
+\bmb{What precise ``issues'' do you have in mind?
+Don't think we need to state them here, but we should know.
+Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness?
+(2) not sure whether taking taxon-level means (Ord 2010) is actually a problem.
+Ignores some variation,
+but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
+
 In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
+\bmb{nice!}
 
@@
 First, we need a method for incorporating phylogenetic relationships among species into the model of the relationships among traits.
-The classic phylogenetic regression assumes that phylogenetic correlation (PC) among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; this assumption leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
-More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon Felsenstein's PICs by allowing extra parameters to correct for possible bias.
-They partition the phylogenetically correlated residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting} \todo{do PGLMMs do this, or just Blomberg/Pagel?
-How do B/P differ?}
-\todo{They all do this. B's K is a scaled ratio of of variance, and P's lambda is a scaling parameter of the correlation.} 
+\bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
+You go on to talk about the tip-variation/residual-variation confounding issue.
+Maybe replace with something talking more specifically about resolving tip variation?}
+The classic phylogenetic regression assumes that phylogenetic correlation (PC) among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the traits are normally distributed and observed without any additional error or within-species variation, this leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
+More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon Felsenstein's PICs by considering different response distributions and allowing extra parameters to account for additional complexities in the evolutionary process.
+They partition the phylogenetically correlated residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting}.
 However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error and residual variation is associated with observation level variation (the standard statistical definition).
-Despite the addition of multiple observations being straightforward statistically, it is rarely seen in the literature as existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
+\bmb{This is a good point, well expressed. Does PGLMM lump tip variation
+  and within-species variation, or is this primarily an issue with Blomberg/Pagel?
+Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots}
+Despite the addition of multiple observations per tip taxon (species) being  statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
+\bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
 
 The second challenge is flexibly incorporating models of phylogenetic relationships along with other common features of statistical models, such as variation among multiple different grouping variables or variation in effects (slopes) as well as correlated residual variation.
+\bmb{maybe we can focus on random slopes here?
+Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
@@
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches.
-However, our new formulation is more flexible and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
+However, our new formulation is more flexible; it can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
 We will compare our technique with existing methods to show robustness and efficiency via simulations and data.

### 2018-07-10 main.tex
bmb finished (for now)

@@
 
-The typical phylogenetic regression is of the form:
+\newcommand{\bX}{{\boldmath X}}
+\newcommand{\bbeta}{{\boldsymbol \beta}}
+\newcommand{\bmu}{{\boldsymbol \mu}}
+\newcommand{\by}{{\boldmath y}}
+\newcommand{\bC}{{\boldmath C}}
+
+The typical phylogenetic regression is of the form \bmb{boldface matrices/vectors i.e. $\bX \bbeta$? or not worth the trouble?
+How about $\bmu=\bX \bbeta$; $\by \sim \textrm{MVN}(\bmu,\sigma^2 \bC)$ for greater generality?
+(OK, I see that you move in this direction later.
+But I'm not sure it's worth the detour of using two different notations?
+even if you are going to extend to include fixed effects, $Z$, etc.
+later ...}:
 \begin{align}
@@
 \end{align}
-where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times m$ model matrix containing 1's in the first column and the $m$ predictor variables (phenotypic traits or environmental variables); $B$ is an $m$ vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation.
+where $y$ is an $n \times 1$ response vector; $X$ is an $n \times m$ model matrix containing $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $B$ is an $m$-vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ where is the $n \times n$ $C$ is the $n \times n$ phylogenetic correlation matrix.
 The PC matrix is inferred from the topology of the evolutionary tree by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods.
+\bmb{give a ref for this? maybe Paradis's ape book, or Garamszegi's book, or ?}
 
@@
 
+\bmb{worth including the MCMCglmm/brms inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
+
 \subsection{Reformulating phylogenetic correlation matrix}
@@
 \end{align}
+\bmb{would like to avoid multi-letter notation if possible; multi-letter subscripts should be text, e.g.
+${\mathbf B}_{\mbox{\tiny adj}}$ (using macros to define stuff in one place \ldots) Also consider $\ell$ as a replacement for $l$}
 where $Bl$ is the branch length vector, $\omega$ is a scaling factor, and $\Sigma$ is the variance-covariance matrix of the phylogenetic tree.
@@
 
-where $\ast$ is the Khatri-Rao product, $S_{i}$ is an $l_{i} \times b_{i}$ matrix of species--branch relations; $J_{i}$ the indicator matrix of grouping factors indices matrix size $n \times l_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
+where $\ast$ is the Khatri-Rao product \bmb{ref?}, $S_{i}$ is an $l_{i} \times b_{i}$ matrix of species--branch relations; $J_{i}$ the indicator matrix of grouping factors indices matrix size $n \times l_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
 
@@
 To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$.
-We further assume that random slope and intercept are correlated. 
+\bmb{clarify what you mean by ``multiple levels of phylog.
+signals'' here?}
+We further assume that random slope and intercept are correlated. \bmb{, ``i.e. ...'' it will help readers unfamiliar with these ideas to give slightly more detailed examples}
 For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
@@
 
-Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables, we implemented our approach using the lme4 R package \citep{bates2015fitting}.
-We have included three additional R packages that can fit phylogenetic comparative models under BM process assumption, and they are nlme \citep{pinheiro2014r}, phylolm \citep{ho2014phylolm}, and pez \citep{pearse2015pez}.
-Phylogenetic generalized least squares (PGLS) in \textit{nlme}, is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
-Phylogenetic generalized linear model (PGLM) in \textit{phylolm}, is a slightly more flexible variation of PGLS that can allow for within species variation (sometimes called the tip variation). 
-Neither PGLS nor PGLM can fit random slope or multiple sites. 
-One of the relatively few packages that currently fit random slopes or multiple sites is \textit{pez}, which can handle adding additional uncorrelated random slopes with intercept and species in multiple sites/groups.
+\newcommand{\pkg}[1]{{\tt #1}}
+\newcommand{\code}[1]{{\tt #1}}
+
+Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables.
+We implemented our approach using the \pkg{lme4} R package \citep{bates2015fitting}.
+We compare our approach with three other R packages that can fit phylogenetic comparative models under the assumption that evolution follows BM process; are \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, and \pkg{pez} \citep{pearse2015pez}.
+\bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
+\begin{verbatim}
+grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
+invert=TRUE,value=TRUE)
+\end{verbatim}
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
+Phylogenetic generalized linear models (PGLM) (\bmb{function name?} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow for within-species variation (sometimes called the tip variation).
+\bmb{does it handle non-Normal responses? I'd think so if it's ``generalized''. Worth establishing our terminology for ``tip variation''/''within-species variation'' etc. earlier?}
+Neither PGLS nor PGLM can fit random-slope models or multiple sites \bmb{what does ``multiple sites'' mean in the model context?}
+One of the relatively few packages that currently fit random slopes or multiple sites is \pkg{pez}, which can handle adding additional uncorrelated random slopes with intercept and species in multiple sites/groups.
+\bmb{I like the commented-out table here (maybe in supp material? or maybe useful enough that it's worth including in the main text?) although it could take a lot of fussing to get it right}
 
@@
   \includegraphics[scale=0.8,page=1]{./git_push/ssplot.pdf}
+  \caption{\bmb{caption ??}}
 \label{ssplot}
@@
 
+\bmb{Need to think a little harder about the important points/order here (might be time to do the exercise where you take the results and boil them back down to short bullet points, in order to figure out whether you've said the important stuff, in the right order).
+Maybe lead by saying that all platforms give similar answers - to the extent that they are fitting the same statistical model (this might even be emphasized/explained above, in Methods).
+Am I correct in believing that phylolm=lme4 but without intercept/slope correlation, and gls is additionally missing random slopes?}
+
+\bmb{add refs to figures \ldots}
+
 The full model (which matches the simulation model that incorporates phylogenetic signal on the intercept, slope and correlation) provides good estimates of the different levels of the phylogenetic variation signal (Figure 1 and 2).
+\bmb{delete ``signal''?
+or ``variation''?
+what does ``phylogenetic variation signal'' mean?}
 
-In general, all phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) with more data, except PGLS in the single site simulations. 
+In general, the phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) with more data, except PGLS in the single site simulations.
+\bmb{(2) please define ``robustness'' -- usually means lack of sensitivity to model misspecification.
+Are you talking about the variance of the estimates?
+I'd prefer not to use ``robust'' in a non-standard way if possible (2) what are small/medium/large?
+I see that $n=\{20,100,500\}$ and $m=\{1,20\}$, but not sure how these match up to ``small'', ``medium'', ``large''; is Figure 1 all $m=1$?
+Might be worth just labeling by $n=??$ rather than using small/medium/large}
 It is unclear how different levels of phylogenetic signal are being confounded under the error free (without tip variation) assumption for the PGLS models.
-Models with only two parameters/levels of variation capture the phylogenetic intercept and tip variation reasonably well to the simulated parameters despite not including phylogenetic random slope and slope intercept correlation.
+\bmb{can we figure this out, even in an empirical/crude way??? it's tough to have results without knowing the true value they're aiming for ...}
+Models with only two parameters/levels of variation capture the true phylogenetic intercept and tip variation reasonably well despite neglecting phylogenetic variation in the slopes (\pkg{gls}) and correlation between the intercepts and slopes (\pkg{phylolm}).
+\bmb{say something about the implications of this [e.g.
+good enough for getting an overall impression of phylog.
+whatever-it-is but not enough for detailed description/inference about the evolutionary process?] (or maybe that should be saved for the Discussion?)}
 
-Platforms using the mixed model framework for multiple site simulations follows similar behaviour as single site simulation (i.e.
+Platforms using the mixed model framework for multiple site simulations follow similar behaviour as the single-site simulation (i.e.
 different levels of phylogenetic signals and variations are closer to the simulated parameters with more data).
 All parameter estimates are much more robust for the multiple site than the single site simulations, especially slope-intercept correlations.
-There are no clear difference with phylogenetic signal models without random slope and intercept correlation.
-However, despite the similarities of robustness estimation, the efficiency (measured by computational speed) of sampling independent errors in lme4 is dramatically better (orders of magnitude faster than pez.
-Furthermore, it was not practical for pez to fit large sample size simulation data as the computational speed did not scale linearly.
+The lack of phylogenetic correlation in \pkg{pez}'s statistical models does not appear to hurt its ability to estimate other aspects of phylogenetic signal; in fact, its slightly simpler model may be responsible for its slightly less variable estimates.
+\bmb{Can you identify the extreme cases in lme4's estimate?
+Are these singular models?
+Would they be diagnosable?
+Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem?
+(If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
+However, despite the similarities of robustness estimation, the efficiency (measured by computational speed) of sampling independent errors in \pkg{lme4} is dramatically better (orders of magnitude faster) than \pkg{pez}.
+\bmb{be more specific, e.g. ``median of 8 vs 800 seconds for the medium-size example''. Clarify what ``independent errors'' means here?}
+Furthermore, it was not practical for \pkg{pez} to fit large sample size simulation data as the computational speed did not scale linearly.
+\bmb{Can you put a lower bound on pez's computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
+  or (easier) do a brute-force series of increasing sample size and
+  compute a log-log regression of time vs sample size to find the
+  approximate power?}
 
@@
   \includegraphics[scale=0.8,page=1]{./csplot.pdf}
+  \caption{\bmb{caption??}}
 \end{figure}
@@
 \includegraphics[scale=0.8,page=2]{./csplot.pdf}
+\caption{\bmb{caption??}}
 \end{figure}
@@
 
+\bmb{Can you give more motivation for this? Why is it here?}
+
 We used our method and fitted the full model of the dune meadow data recently used with pez \cite{li2017canfun}. 
@@
 
-We have fitted models including maximum capabilities to model varying levels of variation/phylogenetic signal to simulated data with our simple model with a single response and predictor variable.
+We have fitted models including maximum capabilities \bmb{???} to model varying levels of variation/phylogenetic signal to simulated data with our simple model with a single response and predictor variable.
 Using models that can include/account for multiple levels of variation is necessary for robust fits, but models that include phylogenetic signal in the intercept level can work as well as the full model.
+\bmb{clarify?}
 Under the Brownian motion evolution process assumption, modelling via independent errors is much more computationally efficient than imposing equivalent correlation structures.
+\bmb{clarify?
+you mean that assuming int/slope independence is much faster?
+Don't remember seeing such an example in the results?}
 
@@
 PGLMM, PGLM) instead of the less realistic model that assumes variation due to evolution and zero observation (PGLS) was large.
-In particular, models that did not include residual variation tended to over-estimate the amount of variabilities in the evolution process. 
+In particular, models that did not include residual variation tended to over-estimate the amount of variability in the evolutionary process. 
 Thus, any model that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e.
 Pagel's $\lambda$, Blomberg's $K$).
+\bmb{again, we/you need to think carefully about what these coarse measures are good for; just avoiding mis-estimating the strength of trait/trait or trait/environment parameters?
+Is it worth doing some sims on that, or do we basically believe if the total amount of phylogenetic signal is correctly assessed, that we get the right answer for the trait effect?
+Is \cite{schielzeth2009conclusions} relevant/a useful analogy?
+(You cite it below, but there are two separate questions: (1) whether including random slopes gives us more insight into evol processes; (2) whether omitting random slopes gives us bad answers even when we're not interested in them.) Worth citing/discussing \cite{uyeda2018rethinking} (as an example of even deeper analysis of evolutionary processes, beyond BM etc.)?}
 
@@
 
+\bmb{I think you need a lot more here; what does a random slopes model mean?
+You have one sentence that says ``people should be thinking about random slopes''.
+Then you talk about correlation (which is a much more subtle/less important issue, I think).}
 In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying}.
@@
 and Are we making a mistake by ignoring random slopes \cite{schielzeth2008conclusions}?
-However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be *theoretically* identifiable.
+However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
 
@@
 Should we use a complex model and overfit \cite{barr2013random}, the right balance of complexity and data \cite{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
-More importantly, this implementation in lme4 allows users to fit phylogenetic mixed models to the fullest ( that other frequentest platforms lack of (i.e.
+More importantly, this implementation in \pkg{lme4} allows users to fit phylogenetic mixed models to the fullest ( that other frequentist platforms lack of (i.e.
 large data, unbalanced species observations, complex random effects) and explore new ideas.

### 2018-08-15 main.tex
temp push

@@
 Although a wide range of sophisticated statistical tools for comparative analyses exist, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
-In such cases, researchers may try to find ways to simplify their analyses (i.e. simplifying the model or ignoring phylogenetic relationships, or both) \citep{bunnefeld2012island, ord2010adaptation},
+In such cases, researchers may try to find ways to simplify their analyses (i.e.
+simplifying the model by making assumptions of the evolutionary process) \citep{bunnefeld2012island, ord2010adaptation},
 %% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
@@
 %%  i.e. ignoring within-taxon variation?
-which can lead to both statistical and biological issues
+which can lead to statistical issues and erroneous conclusions.
 \cite{felsenstein1985phylogenies, li2017statistical}.
-\bmb{What precise ``issues'' do you have in mind?
+% \bmb{What precise ``issues'' do you have in mind?
 Don't think we need to state them here, but we should know.
@@
 Ignores some variation,
-but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
-
+% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
 In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
-\bmb{nice!}
+% \bmb{nice!}
 
 There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
+Separating 
 First, we need a method for incorporating phylogenetic relationships among species into the model of the relationships among traits.
+
 \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
@@
 Maybe replace with something talking more specifically about resolving tip variation?}
+
 The classic phylogenetic regression assumes that phylogenetic correlation (PC) among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the traits are normally distributed and observed without any additional error or within-species variation, this leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
@@
 Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots}
+
 Despite the addition of multiple observations per tip taxon (species) being  statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
 \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
+\ml{todo}
 

### 2018-08-17 main.tex
finish fixing intro

@@
 %%  i.e. ignoring within-taxon variation?
-which can lead to statistical issues and erroneous conclusions.
+which can lead to statistical issues and erroneous conclusions
 \cite{felsenstein1985phylogenies, li2017statistical}.
@@
 There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
-Separating 
-First, we need a method for incorporating phylogenetic relationships among species into the model of the relationships among traits.
-
-\bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
+% Phylogenetic signal : the magnitude of phylogenetic correlation
+The first challenge is segregating phylogenetic signals \citep{blomberg2003testing} and other sources of variation that are present in residuals.  
+% \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
 You go on to talk about the tip-variation/residual-variation confounding issue.
 Maybe replace with something talking more specifically about resolving tip variation?}
-
 The classic phylogenetic regression assumes that phylogenetic correlation (PC) among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the traits are normally distributed and observed without any additional error or within-species variation, this leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
@@
 Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots}
+\ml{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
+Despite the addition of multiple observations per tip taxon (species) being statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
+% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
 
-Despite the addition of multiple observations per tip taxon (species) being  statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
-\bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
-\ml{todo}
-
-The second challenge is flexibly incorporating models of phylogenetic relationships along with other common features of statistical models, such as variation among multiple different grouping variables or variation in effects (slopes) as well as correlated residual variation.
+The second challenge is flexibly model phylogenetic relationships along with other common features of statistical models, such as variation among multiple different grouping variables or variation in effects (random slopes) \citep{schielzeth2008conclusions} as well as correlated residual variation.
 \bmb{maybe we can focus on random slopes here?
 Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
+Several recent studies have tried to incorporate phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
 The tools available for extending phylogenetic relationships to common statistical regression features are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \cite{hadfield2010mcmc, burkner2016brms}.
+Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.
+
+\begin{tabularx}{\textwidth}{|X|X|X|X|}
+\hline
+Model & Method & Data & Platform \\
+\hline
+\hline
+GLM (Simple)  & No Residual & Single Observation & GLS \\
+\hline
+              & Residual    & Single Observation & Pagel's $\lambda$ \\
+              & + phylo intercept &                    & Blomberg's $K$ \\ 
+              &             &                    & phylolm \\
+\hline
+GLMM (Complex)& Random effects & Single Observation & pez \\ 
+              &                & Balanced Design &  \\
+\hline
+              &                & Unrestricted  & lme4, glmmTMB \\
+\hline
+Bayesian GLMM &                & Balanced Design & MCMCglmm \\ 
+\hline        
+              &                & Unrestricted   & brms \\
+\hline
+\end{tabularx}
 

### 2018-08-29 main.tex
happy with the intro for now

@@
 One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression \cite{hansen2012interpreting}. 
-Comparative analyses explore relationships among species traits (including occurrence within communities) while taking evolutionary relationships into account; in some cases, researchers are interested in quantifying the evolutionary signals underly traits rather than simply controlling for them.
-Although a wide range of sophisticated statistical tools for comparative analyses exist, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
+Comparative analyses explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; in some cases, researchers are interested in quantifying the evolutionary signals underly traits rather than simply controlling for them.
+Unlike ordinary regression, where biologist look at the change in the of predictor variables (for example, observable species trait and environmental factors) on the response variable; the evolutionary relationship between species are unobservable.
+While evolutionary relationships are biologically important and interesting, they often get neglected in multi-species analyses. 
+Recently, biologist have started to incorportate phylogenetic relationships to multi-species modelling, and there exist a range of statistical tools for comparative analyses, but many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
 In such cases, researchers may try to find ways to simplify their analyses (i.e.
@@
 However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error and residual variation is associated with observation level variation (the standard statistical definition).
-\bmb{This is a good point, well expressed. Does PGLMM lump tip variation
-  and within-species variation, or is this primarily an issue with Blomberg/Pagel?
-Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots}
-\ml{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
+%\bmb{This is a good point, well expressed.
+Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel?
+Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
+%\ml{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
 Despite the addition of multiple observations per tip taxon (species) being statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
@@
 
-The second challenge is flexibly model phylogenetic relationships along with other common features of statistical models, such as variation among multiple different grouping variables or variation in effects (random slopes) \citep{schielzeth2008conclusions} as well as correlated residual variation.
+The second challenge is flexibly model phylogenetic relationships to other predictor variables, that is, the variation among multiple different grouping predictor variables or variation in effects (random slopes) as well as correlated variation.
+While it is reasonable to examine evolutionary relationships between species with their response, it is plausiable the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated.
 \bmb{maybe we can focus on random slopes here?
@@
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
-The tools available for extending phylogenetic relationships to common statistical regression features are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \cite{hadfield2010mcmc, burkner2016brms}.
+The tools available for extending phylogenetic relationships to predictor variables are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \cite{hadfield2010mcmc, burkner2016brms}.
 Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.
@@
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches.
-However, our new formulation is more flexible; it can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
-We will compare our technique with existing methods to show robustness and efficiency via simulations and data.
-We end by discussing opportunities and practicalities of our method and considering the state of art for this area of research.
+Our new formulation is more flexible; it can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
+We will compare our technique coded in R package lme4 and glmmTMB with existing methods/platforms (i.e.
+nlme, phylolm, pez, MCMCglmm, and brms) to data from simulated model that incorportates phylogenetic signal from both predictors and tips/species, as well as residual variation.
+In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
+We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.
 

### 2018-08-29 main.tex
temp pushing

@@
 They partition the phylogenetically correlated residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting}.
-However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error and residual variation is associated with observation level variation (the standard statistical definition).
+However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/interept level variation) and residual variation is associated with observation level variation (the standard statistical definition).
+\ml{make a diagram!}
 %\bmb{This is a good point, well expressed.
@@
 
-The second challenge is flexibly model phylogenetic relationships to other predictor variables, that is, the variation among multiple different grouping predictor variables or variation in effects (random slopes) as well as correlated variation.
+The second challenge is flexibly model phylogenetic relationships to other predictor variables (predictor level variation), that is, the variation among multiple different grouping predictor variables or variation in effects (random slopes) (i.e phylogenetically related species may behave/ have similar changes in effect or in similar conditions) as well as correlated variation.
 While it is reasonable to examine evolutionary relationships between species with their response, it is plausiable the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated.
-\bmb{maybe we can focus on random slopes here?
+%\bmb{maybe we can focus on random slopes here?
 Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
@@
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
-The tools available for extending phylogenetic relationships to predictor variables are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \cite{hadfield2010mcmc, burkner2016brms}.
+The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \cite{hadfield2010mcmc, burkner2016brms}.
 Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.
+
+\newcommand{\pkg}[1]{{\tt #1}}
+\newcommand{\code}[1]{{\tt #1}}
 
@@
 \hline
-GLMM (Complex)& Random effects & Single Observation & pez \\ 
+GLMM (Complex)& Random effects & Single Observation & \pkg{pez} \\ 
               &                & Balanced Design &  \\
 \hline
-              &                & Unrestricted  & lme4, glmmTMB \\
+              &                & Unrestricted  & \pkg{lme4}, \pkg{glmmTMB} \\
 \hline
-Bayesian GLMM &                & Balanced Design & MCMCglmm \\ 
+Bayesian GLMM &                & Balanced Design & \pkg{MCMCglmm} \\ 
 \hline        
-              &                & Unrestricted   & brms \\
+              &                & Unrestricted   & \pkg{brms} \\
 \hline
@@
 Our new formulation is more flexible; it can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
-We will compare our technique coded in R package lme4 and glmmTMB with existing methods/platforms (i.e.
-nlme, phylolm, pez, MCMCglmm, and brms) to data from simulated model that incorportates phylogenetic signal from both predictors and tips/species, as well as residual variation.
+We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing methods/platforms (i.e.
+\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{MCMCglmm} \cite{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) to data from simulated model that incorportates phylogenetic signal from both predictors and tips/species, as well as residual variation.
 In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
@@
 
-\newcommand{\bX}{{\boldmath X}}
+\newcommand{\bX}{{\mathbf X}}
 \newcommand{\bbeta}{{\boldsymbol \beta}}
 \newcommand{\bmu}{{\boldsymbol \mu}}
-\newcommand{\by}{{\boldmath y}}
-\newcommand{\bC}{{\boldmath C}}
+\newcommand{\bY}{{\mathbf Y}}
+\newcommand{\bC}{{\mathbf C}}
+\newcommand{\bZ}{{\mathbf Z}}
+\newcommand{\bb}{{\mathbf b}}
+\newcommand{\bSigma}{{\boldsymbol \Sigma}}
 
-The typical phylogenetic regression is of the form \bmb{boldface matrices/vectors i.e. $\bX \bbeta$? or not worth the trouble?
-How about $\bmu=\bX \bbeta$; $\by \sim \textrm{MVN}(\bmu,\sigma^2 \bC)$ for greater generality?
+The typical phylogenetic regression is of the form
+
+%\bmb{boldface matrices/vectors i.e.
+$\bX \bbeta$?
+or not worth the trouble?
+How about $\bmu=\bX \bbeta$; $\bY \sim \textrm{MVN}(\bmu,\sigma^2 \bC)$ for greater generality?
 (OK, I see that you move in this direction later.
@@
 later ...}:
+
 \begin{align}
-y & = XB + \epsilon \label{eq:gls1} \\ 
-\epsilon & \sim \textrm{MVN}(0,\sigma^{2}C), \label{eq:gls2}
+\bmu & = \bX \bbeta  \label{eq:gls1} \\ 
+\bY & \sim \textrm{MVN}(\bmu,\sigma^{2} \bC), \label{eq:gls2}
 \end{align}
-where $y$ is an $n \times 1$ response vector; $X$ is an $n \times m$ model matrix containing $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $B$ is an $m$-vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ where is the $n \times n$ $C$ is the $n \times n$ phylogenetic correlation matrix.
-The PC matrix is inferred from the topology of the evolutionary tree by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods.
-\bmb{give a ref for this? maybe Paradis's ape book, or Garamszegi's book, or ?}
+
+where $\bX$ is an $n \times m$ model matrix containing $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector and assumed to be multivariate normally distributed with mean $\bmu$ and a variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation matrix.
+The PC matrix is inferred from the topology of the evolutionary tree by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods \citep{garamszegi2014modern}.
+
+%\bmb{give a ref for this? maybe Paradis's ape book, or Garamszegi's book, or ?}
+%\ml{chapter 7 in garamszegi 2014}
 
@@
 \begin{align}
-Y & \sim F(\mu) \label{eq:glmm1} \\
-g(\mu) & = XB + Zb + \epsilon_{mm} \label{eq:glmm2} \\
-\epsilon_{mm} & \sim \textrm{MVN}(0,\sigma^2) \label{eq:glmm3}
+\bY & \sim \textrm{Distr}(\bmu) \label{eq:glmm1} \\
+g(\bmu) & = \bX \bbeta + \bZ \bb \label{eq:glmm2} \\
+\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3}
 \end{align}
-where $Z$ is an $n \times m$ model matrix for the $m$ -- dimensional vector-valued $m$ predictor variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
-Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$, $Z$ to an $n \times 1$ matrix such that:
+where $Z$ is an $n \times m$ model matrix for the $m$ -- dimensional vector-valued $m$ predictor variables; $\bb$ is the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$.
+Analogously, the phylogenetic regression given by \ref{eq:gls1} can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$.
 
-\begin{align}
-Z & \sim \textrm{MVN}(0,\sigma^{2}C). \label{eq:glmmgls}
-\end{align}
-
-\bmb{worth including the MCMCglmm/brms inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
+\bmb{worth including the \pkg{MCMCglmm}/\pkg{brms} inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
+\ml{inverse VCV for BLUP? an alternative way to find the inverse of the vcv(phy)}
 
@@
 
+\newcommand{\bS}{{\mathbf S}}
+\newcommand{\bJ}{{\mathbf J}}
+\newcommand{\bBadj}{{\mathbf B}_{\mbox{\tiny adj}}}
+\newcommand{\bomega}{{\boldsymbol \omega}}
+\newcommand{\bell}{{\boldsymbol \ell}}
+
 An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
@@
 In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
-Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $S$, rather than having to impose a correlation $C$.
+Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $\bS$, rather than having to impose a correlation $\bC$.
 
-$S$ is a matrix product of an $n \times b$ indicator matrix $S_{id}$ of branch indices, and an adjusted branch length vector $B_{adj}$:
+$\bS$ is a matrix product of an $n \times b$ indicator matrix $\bS_{id}$ of branch indices, and an adjusted branch length vector $\bBadj$:
 
 \begin{align}
-B_{adj} & = (\frac{Bl}{\omega})^{1/2} \\
-\omega & = \textrm{det}(\Sigma)^{1/n},
+\bBadj & = (\frac{\bell}{\bomega})^{1/2} \\
+\bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
 \end{align}
-\bmb{would like to avoid multi-letter notation if possible; multi-letter subscripts should be text, e.g.
+% \bmb{would like to avoid multi-letter notation if possible; multi-letter subscripts should be text, e.g.
 ${\mathbf B}_{\mbox{\tiny adj}}$ (using macros to define stuff in one place \ldots) Also consider $\ell$ as a replacement for $l$}
-where $Bl$ is the branch length vector, $\omega$ is a scaling factor, and $\Sigma$ is the variance-covariance matrix of the phylogenetic tree.
-This is mathematically equivalent under BM assumption where we can reconstruct C by matrix multiplying $S$ by itself.
+where $\bell$ is the branch length vector, $\bomega$ is a scaling factor, and $\bSigma_{phy}$ is the variance-covariance matrix of the phylogenetic tree.
+This is mathematically equivalent under BM assumption where we can reconstruct $\bC$ by matrix multiplying $\bS$ by itself.
 
-The random-effect model matrix $Z$ can be decomposed into term-wise model matrix $Z_{i}$ as described in \citet{bates2015fitting}.
-Thus, the phylogenetic correlated random-effect matrix $Z^{C}_{i}$ is
+The random-effect model matrix $\bZ$ can be decomposed into term-wise model matrix $\bZ_{i}$ as described in \citet{bates2015fitting}.
+Thus, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is
 
 \begin{equation}
-Z^{C}_{i} = (S^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T}, \label{eq:ZC}
+\bZ^{C}_{i} = (\bS^{T}_{i}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
 \end{equation}
-
-where $\ast$ is the Khatri-Rao product \bmb{ref?}, $S_{i}$ is an $l_{i} \times b_{i}$ matrix of species--branch relations; $J_{i}$ the indicator matrix of grouping factors indices matrix size $n \times l_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
+\ml{need a matrix diagram on the linear algebra}
+where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{i}$ is an $\ell_{i} \times b_{i}$ matrix of species--branch relations; $\bJ_{i}$ the indicator matrix of grouping factors indices matrix size $n \times \ell_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
 
@@
 
-We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $Y$ and a single predictor continuous variable $X$ for $n$ (20, 100, and 500) species, and $m$ (1 and 20) sites/groups.
+We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species, and $m$ (1 and 25) sites/groups.
 To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$.
@@
 
-\newcommand{\pkg}[1]{{\tt #1}}
-\newcommand{\code}[1]{{\tt #1}}
+Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables.
+We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \ml{citation}.
+We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{MCMCglmm} \cite{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}.
+% \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
+% \begin{verbatim}
+% grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
+% invert=TRUE,value=TRUE)
+% \end{verbatim}
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow observational residuals.
+Both packages can do other evolutionary models and different correlation structures (i.e. Pagel's $\lambda$, Blombergs $K$ and etc.)
+Neither PGLS nor PGLM can multiple observations or predictor level variation \ml{intercept level vs predictor level?}. 
+One of the relatively few packages that currently fit phylogenetic correlations to predictor level variationis is \pkg{pez}, which can handle adding additional uncorrelated random slopes and interactions with intercept.
 
-Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables.
-We implemented our approach using the \pkg{lme4} R package \citep{bates2015fitting}.
-We compare our approach with three other R packages that can fit phylogenetic comparative models under the assumption that evolution follows BM process; are \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, and \pkg{pez} \citep{pearse2015pez}.
-\bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
-\begin{verbatim}
-grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
-invert=TRUE,value=TRUE)
-\end{verbatim}
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
-Phylogenetic generalized linear models (PGLM) (\bmb{function name?} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow for within-species variation (sometimes called the tip variation).
-\bmb{does it handle non-Normal responses? I'd think so if it's ``generalized''. Worth establishing our terminology for ``tip variation''/''within-species variation'' etc. earlier?}
-Neither PGLS nor PGLM can fit random-slope models or multiple sites \bmb{what does ``multiple sites'' mean in the model context?}
-One of the relatively few packages that currently fit random slopes or multiple sites is \pkg{pez}, which can handle adding additional uncorrelated random slopes with intercept and species in multiple sites/groups.
 \bmb{I like the commented-out table here (maybe in supp material? or maybe useful enough that it's worth including in the main text?) although it could take a lot of fussing to get it right}
@@
 % 
-% Single site intercept	  &	1 $\mid$ sp	& gls (if residual = 0), phyloglm, lme4 \\
-% Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& lme4 \\
-% Single site intercept and correlated slopes 			& 1 + trait $\mid$ sp					& lme4 \\
+% Single site intercept	  &	1 $\mid$ sp	& gls (if residual = 0), phyloglm, \pkg{lme4} \\
+% Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& \pkg{lme4} \\
+% Single site intercept and correlated slopes 			& 1 + trait $\mid$ sp					& \pkg{lme4} \\
 % \hline
-% Multiple site intercept 									&  1 $\mid$ sp 						& pez, lme4 \\
-% Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + trait $\mid$ sp 		& pez, lme4 \\
-% Multiple site intercept and correlated slopes 	& 	1 + trait $\mid$ sp					& lme4 \\
+% Multiple site intercept 									&  1 $\mid$ sp 						& \pkg{pez}, \pkg{lme4} \\
+% Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + trait $\mid$ sp 		& \pkg{pez}, \pkg{lme4} \\
+% Multiple site intercept and correlated slopes 	& 	1 + trait $\mid$ sp					& \pkg{lme4} \\
 % \hline
@@
 Maybe lead by saying that all platforms give similar answers - to the extent that they are fitting the same statistical model (this might even be emphasized/explained above, in Methods).
-Am I correct in believing that phylolm=lme4 but without intercept/slope correlation, and gls is additionally missing random slopes?}
+Am I correct in believing that phylolm=\pkg{lme4} but without intercept/slope correlation, and gls is additionally missing random slopes?}
 
@@
 The lack of phylogenetic correlation in \pkg{pez}'s statistical models does not appear to hurt its ability to estimate other aspects of phylogenetic signal; in fact, its slightly simpler model may be responsible for its slightly less variable estimates.
-\bmb{Can you identify the extreme cases in lme4's estimate?
+\bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate?
 Are these singular models?
@@
 Furthermore, it was not practical for \pkg{pez} to fit large sample size simulation data as the computational speed did not scale linearly.
-\bmb{Can you put a lower bound on pez's computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
+\bmb{Can you put a lower bound on \pkg{pez}'s computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
   or (easier) do a brute-force series of increasing sample size and
@@
 
-We used our method and fitted the full model of the dune meadow data recently used with pez \cite{li2017canfun}. 
-We have obtained identical results for fix and random effect estimates and approximately 120 times faster compared to pez. 
-Codes for fitting dune meadow analysis using lme4 and pez are provided in the supplements.
+We used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \cite{li2017canfun}. 
+We have obtained identical results for fix and random effect estimates and approximately 120 times faster compared to \pkg{pez}. 
+Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.
 
@@
 More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis.
-This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as MCMCglmm \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
+This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as \pkg{MCMCglmm} \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
 However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e.

### 2018-09-05 main.tex
bmb comment about what to look at
adding matrix diagrams

@@
+\bmb{stop at Results until further notice}
 \section{Introduction}
@@
 Comparative analyses explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; in some cases, researchers are interested in quantifying the evolutionary signals underly traits rather than simply controlling for them.
-Unlike ordinary regression, where biologist look at the change in the of predictor variables (for example, observable species trait and environmental factors) on the response variable; the evolutionary relationship between species are unobservable.
+Unlike ordinary regression, where biologist look at the change in the of predictor variables (for example, species trait and environmental factors) on the response variable that are often observable; in phylogenetic regression, the evolutionary relationship between species are unobservable and assumed to evolve in some underlying evolutionary process \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 While evolutionary relationships are biologically important and interesting, they often get neglected in multi-species analyses. 
@@
 % Phylogenetic signal : the magnitude of phylogenetic correlation
-The first challenge is segregating phylogenetic signals \citep{blomberg2003testing} and other sources of variation that are present in residuals.  
+The first challenge is to identify and segregate phylogenetic signals \citep{blomberg2003testing} and other sources of variation that are present in residuals.  
 % \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
@@
 However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/interept level variation) and residual variation is associated with observation level variation (the standard statistical definition).
-\ml{make a diagram!}
+\mli{make a diagram!}
 %\bmb{This is a good point, well expressed.
@@
 Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
-%\ml{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
+%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
 Despite the addition of multiple observations per tip taxon (species) being statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
@@
 
-The second challenge is flexibly model phylogenetic relationships to other predictor variables (predictor level variation), that is, the variation among multiple different grouping predictor variables or variation in effects (random slopes) (i.e phylogenetically related species may behave/ have similar changes in effect or in similar conditions) as well as correlated variation.
+The second challenge is flexibly model phylogenetic relationships to other predictor variables (predictor level variation), that is, the variation among multiple different grouping predictor variables (random levels) or variation in effects (random slopes) (i.e phylogenetically related species may behave/ have similar changes in effect or in similar conditions) as well as correlated variation.
 While it is reasonable to examine evolutionary relationships between species with their response, it is plausiable the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated.
@@
 %\bmb{give a ref for this? maybe Paradis's ape book, or Garamszegi's book, or ?}
-%\ml{chapter 7 in garamszegi 2014}
+%\mli{chapter 7 in garamszegi 2014}
 
 % More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
-%\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
+%\mli{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
 An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
@@
 \bmb{worth including the \pkg{MCMCglmm}/\pkg{brms} inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
-\ml{inverse VCV for BLUP? an alternative way to find the inverse of the vcv(phy)}
+\mli{inverse VCV for BLUP? an alternative way to find the inverse of the vcv(phy)}
 
@@
 \newcommand{\bJ}{{\mathbf J}}
+\newcommand{\bB}{{\mathbf B}}
 \newcommand{\bBadj}{{\mathbf B}_{\mbox{\tiny adj}}}
@@
 
-$\bS$ is a matrix product of an $n \times b$ indicator matrix $\bS_{id}$ of branch indices, and an adjusted branch length vector $\bBadj$:
+\subsubsection*{Constructing the species--branch matrix}
+
+$\bS$ is a matrix product of an $n \times b$ indicator matrix $\bS_{id}$ of branch indices, and a branch length vector $\bB$.
+For example, \mli{ref fig:tree} is a phylogenetic tree with three species:
+
+\mli{example.Rout}
+\begin{align*}
+\bS_{id} = \begin{bmatrix}
+1 & 1 & 0 & 0 \\ 
+1 & 0 & 1 & 0 \\ 
+0 & 0 & 0 & 1
+\end{bmatrix} , 
+\bB = \begin{bmatrix}
+L1 \\
+L2 \\
+L3 \\
+L4 
+\end{bmatrix}
+\\
+\bS = \bS_{id} \bB = \begin{bmatrix}
+L1 & L2 & 0 & 0 \\
+L1 &  0 &L3 & 0 \\
+0  &  0 & 0  & L4
+\end{bmatrix}
+\end{align*}
 
@@
 where $\bell$ is the branch length vector, $\bomega$ is a scaling factor, and $\bSigma_{phy}$ is the variance-covariance matrix of the phylogenetic tree.
+For example, 
 This is mathematically equivalent under BM assumption where we can reconstruct $\bC$ by matrix multiplying $\bS$ by itself.
+
+\subsubsection*{Constructing the random effects model matrix}
 
@@
 \end{equation}
-\ml{need a matrix diagram on the linear algebra}
 where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{i}$ is an $\ell_{i} \times b_{i}$ matrix of species--branch relations; $\bJ_{i}$ the indicator matrix of grouping factors indices matrix size $n \times \ell_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
@@
 Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables.
-We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \ml{citation}.
+We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \mli{citation}.
 We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{MCMCglmm} \cite{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}.
@@
 Both packages can do other evolutionary models and different correlation structures (i.e. Pagel's $\lambda$, Blombergs $K$ and etc.)
-Neither PGLS nor PGLM can multiple observations or predictor level variation \ml{intercept level vs predictor level?}. 
+Neither PGLS nor PGLM can multiple observations or predictor level variation \mli{intercept level vs predictor level?}. 
 One of the relatively few packages that currently fit phylogenetic correlations to predictor level variationis is \pkg{pez}, which can handle adding additional uncorrelated random slopes and interactions with intercept.
@@
 
-% Table \ml{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
+% Table \mli{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
 

### 2018-09-05 main.tex
find citition for scaling factor
temp pushing

@@
 \newcommand{\bell}{{\boldsymbol \ell}}
+\newcommand{\e}{{ \epsilon}}
 
@@
 Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $\bS$, rather than having to impose a correlation $\bC$.
+For example, \mli{ref fig:tree} is a phylogenetic tree with three species, the corresponding $\bS$.
 
-\subsubsection*{Constructing the species--branch matrix}
+\begin{align*}
+\bS = \begin{bmatrix}
+\ell_1 \e_1 & \ell_2 \e_2 & 0 & 0 \\
+\ell_1 \e_1 &  0 & \ell_3 \e_3 & 0 \\
+0  &  0 & 0  & \ell_4 \e_4
+\end{bmatrix}
+\end{align*}
+The error value corresponding to tip 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i$ is the square root of the branch length and the $\ell_i$ are independent, homoscedastic Normal variates.
+
+\subsubsection*{Constructing the random effects model matrix}
 
@@
 
-\mli{example.Rout}
 \begin{align*}
@@
 \bB = \begin{bmatrix}
-L1 \\
-L2 \\
-L3 \\
-L4 
-\end{bmatrix}
-\\
-\bS = \bS_{id} \bB = \begin{bmatrix}
-L1 & L2 & 0 & 0 \\
-L1 &  0 &L3 & 0 \\
-0  &  0 & 0  & L4
+\ell_1 \\
+\ell_2 \\
+\ell_3 \\
+\ell_4 
 \end{bmatrix}
 \end{align*}
+
+\mli{cite ives book for standardizing VCV with the determinate scaling factor}
 
@@
 This is mathematically equivalent under BM assumption where we can reconstruct $\bC$ by matrix multiplying $\bS$ by itself.
-
-\subsubsection*{Constructing the random effects model matrix}
 

### 2018-09-05 main.tex
temp pushing

@@
 
-\subsection{Reformulating phylogenetic correlation matrix}
+\subsection{Reformulating phylogenetic covariance matrix}
 % The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
@@
 
-\subsubsection*{Constructing the random effects model matrix}
+\subsubsection*{Constructing the species--branch random effects model matrix}
 
-$\bS$ is a matrix product of an $n \times b$ indicator matrix $\bS_{id}$ of branch indices, and a branch length vector $\bB$.
-For example, \mli{ref fig:tree} is a phylogenetic tree with three species:
+The error-free $\bS$ is a matrix product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices, and a branch length vector $\bell$.
 
 \begin{align*}
-\bS_{id} = \begin{bmatrix}
+\bS_{ind} = \begin{bmatrix}
 1 & 1 & 0 & 0 \\ 
@@
 \end{bmatrix} , 
-\bB = \begin{bmatrix}
+\bell = \begin{bmatrix}
 \ell_1 \\
@@
 \end{align*}
+The $\bS$ matrix has a very nice property that $\bS \bS^T$ gives the variance-covariance matrix of the phylogeny.
 
@@
 \begin{align}
-\bBadj & = (\frac{\bell}{\bomega})^{1/2} \\
+\bell_{adj} & = (\frac{\bell}{\bomega})^{1/2} \\
 \bomega & = \textrm{det}(\bSigma_{phy})^{1/n},

### 2018-09-11 main.tex
pushing current version with SGV. Removing it from the paper

@@
 Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $\bS$, rather than having to impose a correlation $\bC$.
-For example, \mli{ref fig:tree} is a phylogenetic tree with three species, the corresponding $\bS$.
+For example, figure \ref{fig:tree} is a phylogenetic tree with three species, the corresponding $\bS$ is in the form:
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=1]{./git_push/tree.pdf}
+  \caption{\bmb{caption ??}}
+\label{fig:tree}
+\end{figure}
+\end{center}
 
@@
 
-\begin{align*}
+\begin{align}
 \bS_{ind} = \begin{bmatrix}
@@
 \end{bmatrix}
-\end{align*}
+\label{eq:S}
+\end{align}
 The $\bS$ matrix has a very nice property that $\bS \bS^T$ gives the variance-covariance matrix of the phylogeny.
-
-\mli{cite ives book for standardizing VCV with the determinate scaling factor}
-
-\begin{align}
-\bell_{adj} & = (\frac{\bell}{\bomega})^{1/2} \\
-\bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
-\end{align}
-% \bmb{would like to avoid multi-letter notation if possible; multi-letter subscripts should be text, e.g.
-${\mathbf B}_{\mbox{\tiny adj}}$ (using macros to define stuff in one place \ldots) Also consider $\ell$ as a replacement for $l$}
-where $\bell$ is the branch length vector, $\bomega$ is a scaling factor, and $\bSigma_{phy}$ is the variance-covariance matrix of the phylogenetic tree.
-For example, 
-This is mathematically equivalent under BM assumption where we can reconstruct $\bC$ by matrix multiplying $\bS$ by itself.
 
@@
 \begin{equation}
-\bZ^{C}_{i} = (\bS^{T}_{i}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
+\bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
 \end{equation}
-where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{i}$ is an $\ell_{i} \times b_{i}$ matrix of species--branch relations; $\bJ_{i}$ the indicator matrix of grouping factors indices matrix size $n \times \ell_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
+where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{adji}$ is an $m_{i} \times b_{i}$ matrix of the scaled species--branch matrix; $\bJ_{i}$ the indicator matrix of grouping factors indices matrix size $n \times m_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
+
+\mli{move example to the supp}
+% For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix,
+% 
+% \begin{align}
+% \bX = \begin{bmatrix}
+% 1 & t_1  \\ 
+% 1 & t_2  \\ 
+% 1 & t_3 
+% \end{bmatrix} 
+% \end{align}
+% then the term-wise phylogenetic random effects model matrix is,
+% 
+% \begin{align}
+% \bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
+% \Bigg[
+% \Bigg(
+% \begin{bmatrix}
+% \ell_{adj1} & \ell_{adj1}  & 0 \\
+% \ell_{adj2} &  0  & 0 \\
+% 0  &  \ell_{adj3} & 0 \\
+% 0 & 0 &  \ell_{adj4} 
+% \end{bmatrix}
+% \begin{bmatrix}
+% 1 & 0  & 0 \\
+% 0 & 1  & 0 \\
+% 0 & 0  & 1  
+% \end{bmatrix}
+% \Bigg)
+% \ast
+% \begin{bmatrix}
+% 1   & 1   & 1  \\ 
+% t_1 & t_2 & t_3
+% \end{bmatrix} 
+% \Bigg]^T
+% \\
+% = \begin{bmatrix}
+% \ell_{adj1} & \ell_{adj1}t_1 & \ell_{adj2} & \ell_{adj2}t_1 & 0 & 0 & 0 & 0 \\
+% \ell_{adj1} & \ell_{adj1}t_2 & 0 & 0 & \ell_{adj3} & \ell_{adj3}t_2 & 0 & 0 \\
+% 0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
+% \end{bmatrix}
+% \end{align}
+
+The scaled species--branch matrix $\bS_{adji}$ is scaled version of $\bS$ via the standardized generalized variance (SGV).
+SGVs are used as an intrinsic/natural way to scale down the variance-covariance matrix measured in different scales (for example, phylogenetic distance, years and etc) compariable to other variation in the model \citep{sengupta1987tests}.
+SGV is the $nth$ root of the generalized variance (GV), or the determinant of the variance-covariance matrix, where GV measures the overall size or volume of the variance-covariance matrix.  
+The SGV is given by:
+\begin{align}
+\bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
+\end{align}
+where $n$ is the number of species in the phylogeny.
+
+The analog in the $\bS$ is to scale the branch vector $\bell$ by square root of the SGV:
+\begin{align}
+\bell_{adj} & = \frac{\bell}{\sqrt{\bomega}}
+\end{align} 
+and the $\bS_{adj}$ is the matrix product of $\bS_{ind}$ and $\bell_{adj}$.
 
@@
 
-We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species, and $m$ (1 and 25) sites/groups.
+We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species and $m$ (1 and 25) sites/groups.
+For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species for every site.  
+For single-site model, the full simulation model is the following:
+
+\begin{align}
+\bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + b_{obs}\\
+(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
+\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
+\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
+\end{bmatrix}
+\Bigg) \\
+b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
+\end{align}
+
 To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$.

### 2018-09-11 main.tex
adding simulation full model formulas

@@
 \begin{equation}
-\bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
+\bZ^{C}_{i} = (\bS^{T}_{i}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
 \end{equation}
-where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{adji}$ is an $m_{i} \times b_{i}$ matrix of the scaled species--branch matrix; $\bJ_{i}$ the indicator matrix of grouping factors indices matrix size $n \times m_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
+where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{i}$ is an $m_{i} \times b_{i}$ matrix of the species--branch matrix; $\bJ_{i}$ the indicator matrix of grouping factors indices matrix size $n \times m_{i}$; $X_{i}$ is the raw random effects model matrix size $n \times p_{i}$.
 
@@
 
-The scaled species--branch matrix $\bS_{adji}$ is scaled version of $\bS$ via the standardized generalized variance (SGV).
-SGVs are used as an intrinsic/natural way to scale down the variance-covariance matrix measured in different scales (for example, phylogenetic distance, years and etc) compariable to other variation in the model \citep{sengupta1987tests}.
-SGV is the $nth$ root of the generalized variance (GV), or the determinant of the variance-covariance matrix, where GV measures the overall size or volume of the variance-covariance matrix.  
-The SGV is given by:
-\begin{align}
-\bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
-\end{align}
-where $n$ is the number of species in the phylogeny.
-
-The analog in the $\bS$ is to scale the branch vector $\bell$ by square root of the SGV:
-\begin{align}
-\bell_{adj} & = \frac{\bell}{\sqrt{\bomega}}
-\end{align} 
-and the $\bS_{adj}$ is the matrix product of $\bS_{ind}$ and $\bell_{adj}$.
+% 
+% The scaled species--branch matrix $\bS_{adji}$ is scaled version of $\bS$ via the standardized generalized variance (SGV).
+% SGVs are used as an intrinsic/natural way to scale down the variance-covariance matrix measured in different scales (for example, phylogenetic distance, years and etc) compariable to other variation in the model \citep{sengupta1987tests}.
+% SGV is the $nth$ root of the generalized variance (GV), or the determinant of the variance-covariance matrix, where GV measures the overall size or volume of the variance-covariance matrix.  
+% The SGV is given by:
+% \begin{align}
+% \bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
+% \end{align}
+% where $n$ is the number of species in the phylogeny.
+% 
+% The analog in the $\bS$ is to scale the branch vector $\bell$ by square root of the SGV:
+% \begin{align}
+% \bell_{adj} & = \frac{\bell}{\sqrt{\bomega}}
+% \end{align} 
+% and the $\bS_{adj}$ is the matrix product of $\bS_{ind}$ and $\bell_{adj}$.
 
@@
 
-To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$.
+The full single site simulation model contains two fixed effects parameters ($\beta_0$ and $\beta_1$), and four random effect parameters (phylogenetic random intercept, phylogenetic random slope, correlation between phylogenetic random slope and intercept and observation variance).
+Predictor and intercept level random variation/effect of species are not applicable in this simulation setting with single observation per species because it will be confounded with the observational level variation (residuals).
+
+The full multi-site simulation model :
+
+\begin{align}
+\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site interaction} + b_{obs}\\
+(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
+\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
+\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
+\end{bmatrix}
+\Bigg) \\
+(b_{sp_{int}}, b_{sp_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
+\sigma^2_{sp_{int}} & \sigma_{sp_{int},phy_{slope}} \\ 
+\sigma_{sp_{int},sp_{slope}} & \sigma^2_{sp_{slope}}
+\end{bmatrix}
+\Bigg) \\
+b_{site} & \sim \textrm{MVN} ( 0 , \Sigma_{site}) \\
+b_{sp:site} & \sim \textrm{MVN} (0, kron(\bf{I}_{site}, \sigma^2_{phy})) \\
+b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
+\end{align}
+
+% 
+% To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$.
 \bmb{clarify what you mean by ``multiple levels of phylog.
 signals'' here?}
-We further assume that random slope and intercept are correlated. \bmb{, ``i.e. ...'' it will help readers unfamiliar with these ideas to give slightly more detailed examples}
+% We further assume that random slope and intercept are correlated.
+
+\bmb{, ``i.e. ...'' it will help readers unfamiliar with these ideas to give slightly more detailed examples}
 For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.

### 2018-09-12 main.tex
full version of everything before results

@@
 
-We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species and $m$ (1 and 25) sites/groups.
-For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species for every site.  
+We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species.
+For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
 For single-site model, the full simulation model is the following:
-
 \begin{align}
@@
 \end{align}
+where it contains two fixed effect parameters ($\beta_0$ and $\beta_1$), and four random effect parameters (phylogenetic random intercept, phylogenetic random slope, correlation between phylogenetic random slope and intercept and observation variance).
+Predictor and intercept level random variation/effect of species are not applicable in this simulation setting with single observation per species because it will confound with the observational level variation (residuals).
 
-The full single site simulation model contains two fixed effects parameters ($\beta_0$ and $\beta_1$), and four random effect parameters (phylogenetic random intercept, phylogenetic random slope, correlation between phylogenetic random slope and intercept and observation variance).
-Predictor and intercept level random variation/effect of species are not applicable in this simulation setting with single observation per species because it will be confounded with the observational level variation (residuals).
-
-The full multi-site simulation model :
-
+We extend the simulation model by adding multiple sites where each site will have one observation per species. The multiple-site full model is the following: 
 \begin{align}
-\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site interaction} + b_{obs}\\
+\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site} + b_{obs}\\
 (b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
@@
 b_{site} & \sim \textrm{MVN} ( 0 , \Sigma_{site}) \\
-b_{sp:site} & \sim \textrm{MVN} (0, kron(\bf{I}_{site}, \sigma^2_{phy})) \\
+b_{sp:site} & \sim \textrm{MVN} (0, \textrm{Kron}(\bf{I}_{site}, \sigma^2_{phy})) \\
 b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
 \end{align}
-
-% 
-% To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$.
-\bmb{clarify what you mean by ``multiple levels of phylog.
-signals'' here?}
-% We further assume that random slope and intercept are correlated.
-
-\bmb{, ``i.e. ...'' it will help readers unfamiliar with these ideas to give slightly more detailed examples}
-For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
+The multiple-site full simulation model will have five additional random effects (predictor and intercept level random effect of species and their correlation, random intercept of site and random intercept of species site interaction) compared to the single-site full model.
+Predictor and intercept level random effect of species are now applicable in the multiple-site model because there are multiple (one in every site) observations per species. 
+Random intercept of species and site interactions is an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
 
@@
 Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow observational residuals.
-Both packages can do other evolutionary models and different correlation structures (i.e. Pagel's $\lambda$, Blombergs $K$ and etc.)
+Both packages can do other evolutionary models and different correlation structures (i.e.
+Pagel's $\lambda$, Blombergs $K$ and etc.), but we restrict our PGLS fits to the simple error-free BM correlation.
 Neither PGLS nor PGLM can multiple observations or predictor level variation \mli{intercept level vs predictor level?}. 

### 2018-09-13 main.tex
pushing yet another chart

@@
 
-% % \begin{table}
-% % \begin{center}
-% 
-% \begin{tabularx}{\textwidth}{|X|X|X|}
-% \hline
-% Type of RE	& Formula	& Platform	\\
-% \hline
-% 
-% Single site intercept	  &	1 $\mid$ sp	& gls (if residual = 0), phyloglm, \pkg{lme4} \\
-% Single site intercept and uncorrelated slopes		&  1 $\mid$ sp + trait $\mid$ sp		& \pkg{lme4} \\
-% Single site intercept and correlated slopes 			& 1 + trait $\mid$ sp					& \pkg{lme4} \\
-% \hline
-% Multiple site intercept 									&  1 $\mid$ sp 						& \pkg{pez}, \pkg{lme4} \\
-% Multiple site intercept and uncorrelated slopes 	&  1 $\mid$ sp + trait $\mid$ sp 		& \pkg{pez}, \pkg{lme4} \\
-% Multiple site intercept and correlated slopes 	& 	1 + trait $\mid$ sp					& \pkg{lme4} \\
-% \hline
-% \end{tabularx}
-% % \vspace{1in}
-% % ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
-% % \end{center}
-% % \end{table}
+% \begin{table}
+% \begin{center}
+
+\begin{tabularx}{\textwidth}{|c||c|c|c|c|c|c|}
+\hline
+Package	& \small{nlme}	& \small{phylolm}	& \small{lme4/glmmTMB} & \small{pez} & \small{brms} & \small{MCMCglmm} \\
+\hline
+\hline
+Single Site & & & & & & \\
+Random effects & X & X & X & & X & X \\
+\hline
+Phylo intercept & X & X & X &  & X & X \\
+Phylo slope &  &  & X &  & X & X \\
+Phylo correlation &  &  & X &  & X & X \\
+Residual & & X & X &  & X & X \\
+\hline
+\hline
+Multiple Site & & & & & & \\
+Random effects &  &  & X & X & X & X \\
+\hline
+Phylo intercept &  &  & X & X & X & X \\
+Phylo slope &  &  & X & X & X & X \\
+Phylo correlation &  &  & X &  & X & X \\
+Phylo interaction & &  & X & X & X &  \\
+Species intercept & &  & X & X & X & X \\
+Species slope & & & X & X & X & X \\
+Species correlation & &  & X &  & X & X \\
+Residual & &  & X & X & X & X \\
+\hline
+
+\end{tabularx}
+% \vspace{1in}
+% ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
+% \end{center}
+% \end{table}
 

### 2018-09-18 main.tex
pushing ms

@@
   \includegraphics[scale=0.8,page=1]{./git_push/ssplot.pdf}
-  \caption{\bmb{caption ??}}
+  \caption{Old}
+\end{figure}
+\end{center}
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=1]{./git_push/plot.pdf}
+  \caption{New}
 \label{ssplot}
@@
 \bmb{add refs to figures \ldots}
+
+\begin{itemize}
+\item{full model provides good estimates (except for correlation)}
+\item{reduced models try to fit with whatever parameters they have available (over-estimates the simulation parameters)}
+\item{no slope vs slope}
+\item{no correlation vs correlation}
+\item{as size/data/number of species increase, the parameter estimates gets better}
+\item{single site models are fast}
+\item{estimates gets closer to simulation parameter as number of species increase}
+\item{lme4/tmb are fast, everything else is slow for multiple sites}
+\end{itemize}
 
@@
   \includegraphics[scale=0.8,page=1]{./csplot.pdf}
-  \caption{\bmb{caption??}}
+  \caption{old}
 \end{figure}
@@
 \begin{figure}[h]
-\includegraphics[scale=0.8,page=2]{./csplot.pdf}
-\caption{\bmb{caption??}}
+  \includegraphics[scale=0.8,page=2]{./git_push/plot.pdf}
+  \caption{new}
 \end{figure}
 \end{center}
+
+% \begin{center}
+% \begin{figure}[h]
+% \includegraphics[scale=0.8,page=2]{./csplot.pdf}
+% \caption{\bmb{caption??}}
+% \end{figure}
+% \end{center}
 

### 2018-09-24 main.tex
pushing pseudo results to finish the paper

@@
 \begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./git_push/ssplot.pdf}
-  \caption{Old}
+  \includegraphics[scale=0.8,page=1]{./git_push/plot.Rout.pdf}
+  \caption{New}
+\label{ssplot}
 \end{figure}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./git_push/plot.pdf}
+  \includegraphics[scale=0.8,page=2]{./git_push/plot.Rout.pdf}
+  \caption{New}
+\label{ssplot}
+\end{figure}
+\end{center}
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=3]{./git_push/plot.Rout.pdf}
   \caption{New}
@@
 
-The full model (which matches the simulation model that incorporates phylogenetic signal on the intercept, slope and correlation) provides good estimates of the different levels of the phylogenetic variation signal (Figure 1 and 2).
-\bmb{delete ``signal''?
-or ``variation''?
-what does ``phylogenetic variation signal'' mean?}
+The full model (which matches the simulation model that incorporates phylogenetic signal on the intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameters
 

### 2018-09-24 collect.R

@@ -6,6 +6,7 @@ library(dplyr)
 library(ggplot2)
 library(brms)
 library(tidyr)
+library(phylolm)
 
 #### Collect gls results ----
 
@@ -174,19 +175,19 @@ ssdat <- rbind(gls_data, phylolm_data, lme4ss_data, brmsss_data)
 lme4_path <- "./datadir/lme4/"
 lme4ms_res <- list.files(path = lme4_path, pattern = "ms")
 lme4ms_results <- function(tt){
-	lme4ms_df <- data.frame(resid = numeric(900)
-		, phylo_X = numeric(900)
-		, phylo_int = numeric(900)
-		, phylo_cor = numeric(900)
-		, phylo_interaction = numeric(900)
-		, species_X = numeric(900)
-		, species_int = numeric(900)
-		, species_cor = numeric(900)
-		, site_int = numeric(900)
-		, B0 = numeric(900)
-		, B1 = numeric(900)
-		, model = numeric(900)
-		, time = numeric(900)
+	lme4ms_df <- data.frame(resid = numeric(310)
+		, phylo_X = numeric(310)
+		, phylo_int = numeric(310)
+		, phylo_cor = numeric(310)
+		, phylo_interaction = numeric(310)
+		, species_X = numeric(310)
+		, species_int = numeric(310)
+		, species_cor = numeric(310)
+		, site_int = numeric(310)
+		, B0 = numeric(310)
+		, B1 = numeric(310)
+		, model = numeric(310)
+		, time = numeric(310)
 		)
 	for(i in 1:length(tt)){
 	  lme4_obj <- readRDS(paste(lme4_path,tt[i],sep=""))
@@ -306,9 +307,87 @@ pez_data <- pez_results(pez_res)
 
 
 
-msdat <- rbind(lme4ms_data,pez_data)
+lme4pez_path <- "./datadir/lme4pez/"
+lme4pez_res <- list.files(path = lme4pez_path, pattern = "ms")
+lme4pez_results <- function(tt){
+  lme4ms_df <- data.frame(resid = numeric(300)
+                          , phylo_X = numeric(300)
+                          , phylo_int = numeric(300)
+                          , phylo_cor = NA
+                          , phylo_interaction = numeric(300)
+                          , species_X = numeric(300)
+                          , species_int = numeric(300)
+                          , species_cor = NA
+                          , site_int = numeric(300)
+                          , B0 = numeric(300)
+                          , B1 = numeric(300)
[…]

### 2018-09-24 compare_pez.R

@@ -1,104 +1,112 @@
-## Compare pez
-
-
-### rep li and ives example 
-
+## simulate and fit compound symmetric case
 library(ape)
 library(Matrix)
 library(lme4)
 library(dplyr)
+library(pez)
 
-maxit <- 10000000
-reltol <- 0.00000001
 
+dat <- (dat
+	%>% mutate(obs = sp
+		, site = site_name
+		)
+	%>% arrange(sp,site)
+)
+
+print(dat)
+phyZ <- phylo.to.Z(phy)
+# 
 # debug(phylo_lmm)
 # debug(modify_phylo_retrms)
 
-dd <- data.frame(dat)
-print(dd %>% count(site))
-## 28 sp in each site
+Vphy <- Vphy[levels(dat$sp),levels(dat$sp)]
 
-phy <- get_phylo(veg.long = dune.veg2
-                 , trait = dune.traits2[c(1, 2)]
-                 , trait.re = c("log.sla")
-                 , phylo = dune.phylo2
-                 , trans = "log"
+# random intercept with species independent
+sp.int <- list(1, sp = dat$sp, covar = diag(nspp))
+
+# random intercept with species showing phylogenetic covariances
+phy.int <- list(1, sp = dat$sp, covar = Vphy)
+
+# random slope with species independent
+sp.X <- list(dat$X, sp = dat$sp, covar = diag(nspp))
+
+# random slope with species showing phylogenetic covariances
+phy.X <- list(dat$X, sp = dat$sp, covar = Vphy)
+
+# sp:site
+phy.interaction <- list(1, sp = dat$sp, covar = Vphy, site = dat$site)
+
+site.int <- list(1, site=dat$site, covar = diag(nsite))
+
+tempmod <- phylo_lmm(Y ~ X
+	+ (1 | sp:site)
+	+ (1 + X | sp)
+	+ (1 + X | obs)
+	+ (1 | site)
+	, data=dat
+	, phylonm = c("sp", "sp:site")
+	, phylo = phy
+	, phyloZ = phyZ
+	, control = lmerControl(check.nobs.vs.nlev = "ignore", check.nobs.vs.nRE = "ignore")
+	, REML = TRUE
 )
 
-phyZ <- phylo.to.Z(phy)
[…]

### 2018-09-24 fit_pez.R

@@ -10,12 +10,16 @@ dat <- (dat
 	%>% mutate(obs = sp
 		, site = site_name
 		)
+	%>% arrange(sp,site)
 )
 
+print(dat)
 phyZ <- phylo.to.Z(phy)
 # 
 # debug(phylo_lmm)
 # debug(modify_phylo_retrms)
+
+Vphy <- Vphy[levels(dat$sp),levels(dat$sp)]
 
 # random intercept with species independent
 sp.int <- list(1, sp = dat$sp, covar = diag(nspp))
@@ -32,7 +36,7 @@ phy.X <- list(dat$X, sp = dat$sp, covar = Vphy)
 # sp:site
 phy.interaction <- list(1, sp = dat$sp, covar = Vphy, site = dat$site)
 
-sp.site <- list(1, site=dat$site, covar = diag(nsite))
+site.int <- list(1, site=dat$site, covar = diag(nsite))
 
 tempmod <- phylo_lmm(Y ~ X
 	+ (1 | sp:site)
@@ -72,9 +76,10 @@ fittime <- system.time(
 		, sp = dat$sp
 		, site = dat$site
 		, random.effects = list(phy.interaction
-			, phy.int, phy.X
-			, sp.int, sp.X
-			, sp.site)
+			 , phy.int, phy.X
+			 , sp.int, sp.X
+			, site.int
+			)
 		, REML = TRUE
 		, verbose = FALSE
 	)

### 2018-09-24 lme4pez.R

@@ -1,0 +1,100 @@
+#### Fitting phyloglmm via lme4
+
+library(ape)
+library(Matrix)
+library(lme4)
+library(dplyr)
+
+print(covmat)
+print(cov2cor(covmat))
+
+t1 <- proc.time()
+
+phyZ <- phylo.to.Z(phy,stand=FALSE)
+
+dat <- (dat
+        %>% mutate(obs = sp
+                   , site = site_name
+                   , y_na = NA)
+)	
+
+t2 <- proc.time()
+#debug(phylo_lmm)
+#debug(modify_phylo_retrms)
+
+if(numsite == "ss"){
+  lme4fit <- phylo_lmm(Y ~ X + (1+X|sp)
+                       , data=dat
+                       , phylonm = c("sp","site:sp")
+                       , phylo = phy
+                       , phyloZ=phyZ
+                       , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore"
+                       )
+                       , REML = TRUE
+  )
+  t3 <- proc.time()
+  lme4time <- t3-t1
+}
+
+if(numsite == "ms"){
+  
+  tempmod <- phylo_lmm(Y ~ X
+                       + (1 | sp:site)
+                       + (1 + X | sp)
+                       + (1 + X | obs)
+                       + (1 | site)
+                       , data=dat
+                       , phylonm = c("sp","sp:site")
+                       , phylo = phy
+                       , phyloZ=phyZ
+                       # , nsp = nsite
+                       , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
+                       , REML = TRUE
+  )
+  
+  t1 <- sd.B0 
+  t2 <- rho.B01*sd.B1 
+  t3 <- sqrt(sd.B1^2 - t2^2)
+  
+  t4 <- sd.tip 
+  t5 <- rho.slopetip*sd.slope 
+  t6 <- sqrt(sd.slope^2 - t5^2)
+  new_y <- simulate(tempmod
+                    , newparams=list(theta=c(ss
+                                             , t1, t2, t3
+                                             , t4, t5, t6
+                                             , sd.site
+                    )/sd.resid
+                    , beta = c(beta0,beta1)
+                    , sigma = sd.resid))
+  dat$new_y <- new_y[[1]]
+  
+  t4 <- proc.time()
+  lme4fit <- phylo_lmm(new_y ~ X
+                       + (1 | sp)
[…]

### 2018-09-24 names.R

@@ -12,6 +12,6 @@ if(size == "xlarge"){nspp <- 1000}
 
 
 if(numsite == "ss"){nsite <- 1}
-if(numsite == "ms"){nsite <- 25}
+if(numsite == "ms"){nsite <- 20}
 
 datadir <- "datadir/"

### 2018-09-24 phyloglmm.R

@@ -71,10 +71,10 @@ if(numsite == "ms"){
   
 t4 <- proc.time()
     lme4fit <- phylo_lmm(new_y ~ X
-      + (1 | sp:site)
       + (1 + X | sp)
-		 + (1 + X | obs)
+	 	+ (1 + X | obs)
       + (1 | site)
+		+ (1 | sp:site)
       , data=dat
       , phylonm = c("sp","sp:site")
       , phylo = phy

### 2018-09-24 plot.R

@@ -7,17 +7,21 @@ library(grid)
 theme_set(theme_bw())
 zmargin <- theme(panel.margin=grid::unit(0,"lines"))
 
-#data_list <- readRDS("./datadir/result_list.RDS")
+data_list <- readRDS("./datadir/collect.RDS")
+ssdat_raw <- data_list[[1]]
+msdat_raw <- data_list[[2]]
 
-ssdat <- (ssdat
+ssdat <- (ssdat_raw
 	%>% separate(model,c("platform", "sites", "size", "seed","saveformat"), "[.]")
-	%>% select(time, platform, size, seed, resid, phylo_int, phylo_X, phylo_cor)
-	%>% gather(key=sdtype, value=sd, -c(platform, size, seed, time))
+	%>% select(time, platform, size, seed, resid, phylo_int, phylo_X, phylo_cor, B0, B1)
+	%>% gather(key = sdtype, value = sd, -c(platform, size, seed, time, B0, B1))
 	%>% mutate(size = factor(size
 	      , levels=c("small","med","large","xlarge"), labels=c("25","50","100","1000")
 			)
+			, platform = factor(platform, levels=c("gls", "phylolm","lme4", "brms"))
 		)
 )
+
 
 gg_ss <- (ggplot(data=ssdat, aes(x=size, y=sd, fill=platform))
   + facet_wrap(~sdtype, scale="free_y")
@@ -27,24 +31,76 @@ gg_ss <- (ggplot(data=ssdat, aes(x=size, y=sd, fill=platform))
 
 print(gg_ss)
 
-msdat <- (msdat
+gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, fill=platform))
+	+ geom_violin(position=position_dodge(width=0.2),alpha=0.4)
+	+ scale_y_log10()
+	+ ggtitle("Single site time")
+)
+
+print(gg_sstime)
+
+ss_coverage <- (ssdat
+	%>% group_by(platform, size)
+	%>% summarise(B0_coverage = mean(B0, na.rm=TRUE)
+		, B1_coverage = mean(B1, na.rm=TRUE)
+		)
+	%>% gather(key=fixed_parameter, value=coverage, -c(platform, size))
+)
+
+gg_sscoverage <- (ggplot(data=ss_coverage
+	, aes(x=size, y=coverage, shape=fixed_parameter, colour=platform)
+	)
+	+ facet_wrap(~platform, nrow = 1)
+	+ geom_point(size=4)
[…]

### 2018-09-24 simulate_tree.R

@@ -7,11 +7,12 @@ set.seed(seed)
 phy <- rtree(n = nspp)
 
 ## Compute branch lengths using other methods (optional)
-#phy2 <- compute.brlen(phy, method = "Grafen", power = 0.5)
+#phy <- compute.brlen(phy, method = "Grafen", power = 0.5)
 
 # standardize the phylogenetic covariance matrix to have determinant 1 (optional)
 Vphy <- vcv(phy)
-# Vphy <- Vphy/(det(Vphy)^(1/nspp)) ## for MCMCglmm?
+#Vphy <- Vphy/max(Vphy)
+#Vphy <- Vphy/(det(Vphy)^(1/nspp)) ## for MCMCglmm?
 
 iD <- t(chol(Vphy))
 Xsd <- 1

### 2018-10-11 collect.R
almost ready to look at results again

@@ -13,14 +13,14 @@ library(phylolm)
 gls_path <- "./datadir/gls/"
 gls_res <- list.files(path = gls_path)
 gls_results <- function(tt){
-  gls_df <- data.frame(resid = numeric(1200)
-    , phylo_int = numeric(1200)
+  gls_df <- data.frame(resid = numeric(300)
+    , phylo_int = numeric(300)
     , phylo_X = NA
     , phylo_cor = NA
-    , B0 = numeric(1200)
-    , B1 = numeric(1200)
-    , model = numeric(1200)
-    , time = numeric(1200)
+    , B0 = numeric(300)
+    , B1 = numeric(300)
+    , model = numeric(300)
+    , time = numeric(300)
   )
   for(i in 1:length(tt)){
     gls_obj <- readRDS(paste(gls_path,tt[i],sep=""))
@@ -44,14 +44,14 @@ gls_data <- bind_rows(gls_results(gls_res))
 lme4_path <- "./datadir/lme4/"
 lme4ss_res <- list.files(path = lme4_path, pattern = "ss")
 lme4ss_results <- function(tt){
-  lme4ss_df <- data.frame(resid = numeric(1200)
-    , phylo_X = numeric(1200)
-    , phylo_int = numeric(1200)
-    , phylo_cor = numeric(1200)
-    , B0 = numeric(1200)
-    , B1 = numeric(1200)
-    , model = numeric(1200)
-    , time = numeric(1200)
+  lme4ss_df <- data.frame(resid = numeric(300)
+    , phylo_X = numeric(300)
+    , phylo_int = numeric(300)
+    , phylo_cor = numeric(300)
+    , B0 = numeric(300)
+    , B1 = numeric(300)
+    , model = numeric(300)
+    , time = numeric(300)
   )
   for(i in 1:length(tt)){
     lme4_obj <- readRDS(paste(lme4_path,tt[i],sep=""))
@@ -140,14 +140,14 @@ brmsss_data <- brmsss_results(brmsss_res)
 phylolm_path <- "./datadir/phylolm/"
 phylolm_res <- list.files(path = phylolm_path)
 phylolm_results <- function(tt){
-  phylolm_df <- data.frame(resid = numeric(1200)
-    , phylo_int = numeric(1200)
+  phylolm_df <- data.frame(resid = numeric(300)
+    , phylo_int = numeric(300)
     , phylo_X = NA
     , phylo_cor = NA
[…]

### 2018-10-11 compare_pez.R

@@ -8,7 +8,6 @@ library(phyr)
 
 dat <- (dat
 	%>% mutate(obs = sp
-		, site = site_name
 		)
 	%>% arrange(sp,site)
 )
@@ -22,46 +21,10 @@ phyZ <- phylo.to.Z(phy)
 Vphy <- Vphy[levels(dat$sp),levels(dat$sp)]
 
 
-tempmod <- phylo_lmm(Y ~ X
-	+ (1 | sp:site)
-	+ (1 + X | sp)
-	+ (1 + X | obs)
-	+ (1 | site)
-	, data=dat
-	, phylonm = c("sp", "sp:site")
-	, phylo = phy
-	, phyloZ = phyZ
-	, control = lmerControl(check.nobs.vs.nlev = "ignore", check.nobs.vs.nRE = "ignore")
-	, REML = TRUE
-)
 
-
-rho.B01 <- 0
-rho.slopetip <- 0
-
-t1 <- sd.B0
-t2 <- rho.B01*sd.B1
-t3 <- sqrt(sd.B1^2 - t2^2)
-
-t4 <- sd.tip
-t5 <- rho.slopetip*sd.slope
-t6 <- sqrt(sd.slope^2 - t5^2)
-
-new_y <- simulate(tempmod
-	, newparams=list(theta=c(ss
-	, t1, t2, t3
-	, t4, t5, t6
-	, sd.site)/sd.resid
-	, beta = c(beta0,beta1)
-	, sigma = sd.resid
-	)
-)
-
-dat$new_y <- new_y[[1]]
-
-#phyrfit <- communityPGLMM(new_y ~ X
-#	+ (1 | sp__)
-#	+ (X | sp__)
+phyrfit <- communityPGLMM(new_y ~ X
+	+ (1 | sp__)
+	+ (X | sp__)
 	+ (1 | site)
 	+ (1 | sp__@site)
 	, data = dat
@@ -91,6 +54,6 @@ print(phyrfit)
 print(summary(lme4fit))
 res_list <- list(phyrfit, lme4fit)
 
-#saveRDS(res_list, file=paste("datadir/compare",numsite,size,seed,"rds",sep="."))
+saveRDS(res_list, file=paste("datadir/compare",numsite,size,tree_seed,"rds",sep="."))
 
 #rdnosave()

### 2018-10-11 compare_pez_plot.R

@@ -9,21 +9,21 @@ library(tidyr)
 library(phylolm)
 library(cowplot)
 
-lme4_path <- "./datadir/compare_ultra"
-lme4ms_res <- list.files(path = lme4_path, pattern = "med")
+lme4_path <- "./datadir/"
+lme4ms_res <- list.files(path = lme4_path, pattern = "rds")
 lme4ms_results <- function(tt){
-  lme4ms_df <- data.frame(resid = numeric(100)
-                          , phylo_X = numeric(100)
-                          , phylo_int = numeric(100)
-                          , phylo_cor = numeric(100)
-                          , phylo_interaction = numeric(100)
-                          , species_X = numeric(100)
-                          , species_int = numeric(100)
-                          , species_cor = numeric(100)
-                          , site_int = numeric(100)
-                          , B0 = numeric(100)
-                          , B1 = numeric(100)
-                          , model = numeric(100)
+  lme4ms_df <- data.frame(resid = numeric(10)
+                          , phylo_X = numeric(10)
+                          , phylo_int = numeric(10)
+                          , phylo_cor = numeric(10)
+                          , phylo_interaction = numeric(10)
+                          , species_X = numeric(10)
+                          , species_int = numeric(10)
+                          , species_cor = numeric(10)
+                          , site_int = numeric(10)
+                          , B0 = numeric(10)
+                          , B1 = numeric(10)
+                          , model = numeric(10)
                           , platform = "lme4"
   )
   for(i in 1:length(tt)){
@@ -102,21 +102,21 @@ lme4ms_data <- lme4ms_results(lme4ms_res)
 
 
 pez_path <- "./datadir/"
-pez_res <- list.files(path = pez_path, pattern = "med")
+pez_res <- list.files(path = pez_path, pattern = "rds")
 pez_results <- function(tt){
-  pez_df <- data.frame(resid = numeric(100)
-                       , phylo_X = numeric(100)
-                       , phylo_int = numeric(100)
+  pez_df <- data.frame(resid = numeric(10)
+                       , phylo_X = numeric(10)
+                       , phylo_int = numeric(10)
                        , phylo_cor = NA
-                       , phylo_interaction = numeric(100)
-                       , species_X = numeric(100)
-                       , species_int = numeric(100)
+                       , phylo_interaction = numeric(10)
+                       , species_X = numeric(10)
+                       , species_int = numeric(10)
                        , species_cor = NA
-                       , site_int = numeric(100)
-                       , B0 = numeric(100)
[…]

### 2018-10-11 fit_gls.R

@@ -12,7 +12,7 @@ phy_var <- diag(vcv(phy))
 
 varweights <- varFixed(~phy_var)
 
-tt <- system.time(fit_gls <- gls(Y~X
+tt <- system.time(fit_gls <- gls(new_y~X
 	, data=dat
 	, correlation=corBrownian(phy=phy)
 	, weights = varweights
@@ -25,6 +25,6 @@ print(fit_gls)
 print(summary(fit_gls))
 
 gls_list <- list(fit_gls,tt)
-saveRDS(gls_list,file=paste("datadir/gls",numsite,size,seed,"rds",sep="."))
+saveRDS(gls_list,file=paste("datadir/gls",numsite,size,tree_seed,"rds",sep="."))
 
 #rdnosave()

### 2018-10-11 fit_pez.R

@@ -8,16 +8,14 @@ library(pez)
 
 dat <- (dat
 	%>% mutate(obs = sp
-		, site = site_name
+		, site = factor(site)
 		)
-	%>% arrange(sp,site)
+	%>% arrange(site,sp)
 )
 
-print(dat)
-phyZ <- phylo.to.Z(phy)
-# 
-# debug(phylo_lmm)
-# debug(modify_phylo_retrms)
+print(dat$site)
+
+print(dat$sp)
 
 Vphy <- Vphy[levels(dat$sp),levels(dat$sp)]
 
@@ -38,37 +36,6 @@ phy.interaction <- list(1, sp = dat$sp, covar = Vphy, site = dat$site)
 
 site.int <- list(1, site=dat$site, covar = diag(nsite))
 
-tempmod <- phylo_lmm(Y ~ X
-	+ (1 | sp:site)
-	+ (1 + X | sp)
-	+ (1 + X | obs)
-	+ (1 | site)
-	, data=dat
-	, phylonm = c("sp","sp:site")
-	, phylo = phy
-	, phyloZ=phyZ
-	, control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
-	, REML = TRUE)
-
-
-t1 <- sd.B0
-t2 <- rho.B01*sd.B1
-t3 <- sqrt(sd.B1^2 - t2^2)
-
-t4 <- sd.tip
-t5 <- rho.slopetip*sd.slope
-t6 <- sqrt(sd.slope^2 - t5^2)
-
-new_y <- simulate(tempmod
-	, newparams=list(theta=c(ss
-	, t1, t2, t3
-	, t4, t5, t6
-	, sd.site)/sd.resid
-, beta = c(beta0,beta1)
-, sigma = sd.resid))
-
-dat$new_y <- new_y[[1]]
-
 fittime <- system.time(
 	fitpez <- communityPGLMM(new_y ~ X
 		, data = dat
@@ -80,7 +47,7 @@ fittime <- system.time(
 			 , sp.int, sp.X
 			, site.int
 			)
-		, REML = TRUE
+		, REML = FALSE
 		, verbose = FALSE
 	)
 )
@@ -88,6 +55,6 @@ fittime <- system.time(
 print(fittime)
 print(summary(fitpez))
 
-saveRDS(list(fittime,fitpez),file=paste("datadir/pez",numsite,size,seed,"rds",sep="."))
[…]

### 2018-10-11 fit_phylolm.R

@@ -21,6 +21,6 @@ print(summary(fit_phylolm))
 
 
 phylolm_list <- list(fit_phylolm,tt)
-saveRDS(phylolm_list,file=paste("datadir/phylolm",numsite,size,seed,"rds",sep="."))
+saveRDS(phylolm_list,file=paste("datadir/phylolm",numsite,size,tree_seed,"rds",sep="."))
 
 #rdnosave()

### 2018-10-11 fit_phyr.R

@@ -1,0 +1,29 @@
+library(ape)
+library(phyr)
+library(Matrix)
+library(dplyr)
+
+fittime <- system.time(
+	fitphyr <- communityPGLMM(new_y ~ X
+		+ (1 | sp__)
+		+ (X | sp__)
+		+ (1 | site)
+		+ (1 | sp__@site)
+		, data = dat
+		, family = "gaussian"
+		, tree = phy
+		, REML = FALSE
+		, cpp = TRUE
+		)
+)
+
+print(fittime)
+
+print(fitphyr)
+
+fit_list <- list(fitphyr, fittime)
+
+saveRDS(fit_list, file=paste("datadir/phyr",numsite, size, tree_seed,"rds",sep="."))
+
+# rdnosave()
+

### 2018-10-11 glmmTMBhacked2.R

@@ -1,0 +1,249 @@
+glmmTMBhacked <- function (formula, data = NULL, family = gaussian(), ziformula = ~0, 
+          dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL, 
+          phylonm = NULL,
+          na.action = na.fail, se = TRUE, verbose = FALSE, doFit = TRUE, 
+          control = glmmTMBControl(), REML = FALSE) 
+{
+  call <- mf <- mc <- match.call()
+  if (is.character(family)) {
+    if (family == "beta") {
+      family <- "beta_family"
+      warning("please use ", sQuote("beta_family()"), " rather than ", 
+              sQuote("\"beta\""), " to specify a Beta-distributed response")
+    }
+    family <- get(family, mode = "function", envir = parent.frame())
+  }
+  if (is.function(family)) {
+    family <- family()
+  }
+  if (is.null(family$family)) {
+    print(family)
+    stop("'family' not recognized")
+  }
+  fnames <- names(family)
+  if (!all(c("family", "link") %in% fnames)) 
+    stop("'family' must contain at least 'family' and 'link' components")
+  if (length(miss_comp <- setdiff(c("linkfun", "variance"), 
+                                  fnames)) > 0) {
+    warning("some components missing from ", sQuote("family"), 
+            ": downstream methods may fail")
+  }
+  if (grepl("^quasi", family$family)) 
+    stop("\"quasi\" families cannot be used in glmmTMB")
+  link <- family$link
+  environment(formula) <- parent.frame()
+  call$formula <- mc$formula <- formula
+  if (!is.null(eval(substitute(offset), data, enclos = environment(formula)))) {
+    formula <- addForm0(formula, makeOp(substitute(offset), 
+                                        op = quote(offset)))
+  }
+  environment(ziformula) <- environment(formula)
+  call$ziformula <- ziformula
+  environment(dispformula) <- environment(formula)
+  call$dispformula <- dispformula
+  m <- match(c("data", "subset", "weights", "na.action", "offset"), 
+             names(mf), 0L)
+  mf <- mf[c(1L, m)]
+  mf$drop.unused.levels <- TRUE
+  mf[[1]] <- as.name("model.frame")
+  if (inForm(ziformula, quote(.))) {
[…]

### 2018-10-11 names.R

@@ -3,7 +3,7 @@ targetname <- unlist(strsplit(rtargetname,"[.]"))
 platform <- targetname[2]
 numsite <- targetname[3]
 size <- targetname[4]
-seed <- as.numeric(targetname[5])
+tree_seed <- as.numeric(targetname[5])
 
 if(size == "small"){nspp <- 25}
 if(size == "med"){nspp <- 50}

### 2018-10-11 parameters.R

@@ -1,38 +1,41 @@
 ### parameters for phylogenetic tree simulations
 
-set.seed(seed)
-# simnum <- 200
-# Generate simulated data for nspp species and nsite sites
+set.seed(tree_seed)
 
-#nspp <- 50
-#nsite <- 10
+
+seed <- 0122
+
+nrep <- 1 ## only lme4 can do number reps atm
 
 # residual variance (set to zero for binary data)
-sd.resid <- 0.0000001
+sd.resid <- 1
+sd.site <- 5
+#sd.site <- 0
+Xsd <- 1
+
 # sd.resid <- 0.000000001
 # # fixed effects
 beta0 <- 0
 beta1 <- 0
 
 # magnitude of phylogenetic random effects
-sd.B0 <- 10
-sd.B1 <- 2
-rho.B01 <- 0.0
+physd.B0 <- 10
+physd.B1 <- 10
+phyrho.B01 <- 0.7
 
 # magnitude of random effects
 
-sd.site <- 8
-sd.tip <- 100
-sd.slope <- 3
-rho.slopetip <- 0.0
+sd.B0 <- 3
+sd.B1 <- 3
+rho.B01 <- 0.4
 
-# whether or not to include phylogenetic signal in B0 and B1
-signal.B0 <- TRUE
-signal.B1 <- TRUE
+#sd.B0 <- 0
+#sd.B1 <- 0
+#rho.B01 <- 0
 
 # compound symmetric parameters 
-ss <- 6 ## sp:site 
-
+sd.interaction  <- 5 ## sp:site 
+#sd.interaction <- 0.000001
 
 # MCMC iterations
 

### 2018-10-11 peztest.R

@@ -1,0 +1,246 @@
+library(ape)
+library(Matrix)
+library(lme4)
+library(dplyr)
+library(pez)
+
+seed <- 101
+source("parameters.R")
+nspp <- 30
+nsite <- 20
+rho.B01 <- 0.7
+source("simulate_tree.R")
+source("new_phylo_setup.R")
+
+## pez and lme4 setup ----
+
+sp.int <- list(1, sp = dat$sp, covar = diag(nspp))
+
+# random intercept with species showing phylogenetic covariances
+phy.int <- list(1, sp = dat$sp, covar = Vphy)
+
+# random slope with species independent
+sp.X <- list(dat$X, sp = dat$sp, covar = diag(nspp))
+
+# random slope with species showing phylogenetic covariances
+phy.X <- list(dat$X, sp = dat$sp, covar = Vphy)
+
+phyZ <- phylo.to.Z(phy,stand=FALSE)
+
+dat <- (dat %>% mutate(obs = sp))
+
+## test 1 random int (good, all match up)----
+
+pez1 <- communityPGLMM(Y~X
+  , data = dat
+  , sp = dat$sp
+  , site = dat$site
+  , random.effects = list(sp.int)
+  , REML = TRUE
+  , verbose = FALSE
+)
+
+lme41 <- lmer(Y ~ X + (1|sp), data=dat)
+
+phylolmm1 <- phylo_lmm(Y ~ X + (1|obs)
+  , data=dat
+  , phylonm = c("sp","site:sp")
+  , phylo = phy
+  , phyloZ=phyZ
+  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore"
+  )
+  , REML = TRUE
+)
+
+print(summary(pez1))
+print(summary(lme41))
+print(summary(phylolmm1))
+
+
+### test 2 random slope (pez slope optim control is different?) ----
+
+pez2 <- communityPGLMM(Y~X
+                       , data = dat
+                       , sp = dat$sp
+                       , site = dat$site
+                       , random.effects = list(sp.int, sp.X)
+                       , REML = TRUE
+                       , verbose = FALSE
+)
+
+lme42 <- lmer(Y ~ X + (1|sp) + (0 + X | sp), data=dat)
+
+phylolmm2 <- phylo_lmm(Y ~ X + (1|obs) + (0 + X|obs)
+                       , data=dat
+                       , phylonm = c("sp","site:sp")
+                       , phylo = phy
[…]

### 2018-10-11 phyloglmm.R

@@ -5,17 +5,13 @@ library(Matrix)
 library(lme4)
 library(dplyr)
 
-print(covmat)
-print(cov2cor(covmat))
 
 t1 <- proc.time()
 
 phyZ <- phylo.to.Z(phy,stand=FALSE)
 
 dat <- (dat
-	%>% mutate(obs = sp
-    , site = site_name
-    , y_na = NA)
+	%>% mutate(obs = sp)
 )	
 
 t2 <- proc.time()
@@ -37,37 +33,6 @@ lme4time <- t3-t1
 }
 
 if(numsite == "ms"){
-  
-  tempmod <- phylo_lmm(Y ~ X
-    + (1 | sp:site)
-    + (1 + X | sp)
-    + (1 + X | obs)
-    + (1 | site)
-    , data=dat
-    , phylonm = c("sp","sp:site")
-    , phylo = phy
-    , phyloZ=phyZ
-    # , nsp = nsite
-    , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
-    , REML = TRUE
-  )
-  
-  t1 <- sd.B0 
-  t2 <- rho.B01*sd.B1 
-  t3 <- sqrt(sd.B1^2 - t2^2)
-  
-  t4 <- sd.tip 
-  t5 <- rho.slopetip*sd.slope 
-  t6 <- sqrt(sd.slope^2 - t5^2)
-  new_y <- simulate(tempmod
-  	, newparams=list(theta=c(ss
-   	, t1, t2, t3
-		, t4, t5, t6
-     	, sd.site
-  )/sd.resid
-  , beta = c(beta0,beta1)
-  , sigma = sd.resid))
-  dat$new_y <- new_y[[1]]
   
 t4 <- proc.time()
     lme4fit <- phylo_lmm(new_y ~ X
@@ -93,6 +58,6 @@ print(summary(lme4fit))
 
 lme4_list <- list(lme4fit, lme4time)
 
-saveRDS(lme4_list, file=paste("datadir/lme4",numsite,size,seed,"rds",sep="."))
+saveRDS(lme4_list, file=paste("datadir/lme4",numsite,size,tree_seed,"rds",sep="."))
 
 #rdnosave()

### 2018-10-11 scale.R

@@ -4,7 +4,7 @@ library(MCMCglmm)
 
 seed <- 202
 set.seed(seed)
-nspp <- 5
+nspp <- 10
 
 phy <- rtree(nspp)
 plot(phy)
@@ -18,5 +18,6 @@ plot(uphy)
 uvc <- vcv(uphy)
 print(det(uvc))
 
+inverseA(phy,scale=TRUE)
 inverseA(phy,scale=FALSE)
-inverseA(uphy)
+aa <- inverseA(uphy)

### 2018-10-11 simulate_tree.R

@@ -1,92 +1,100 @@
 ### simulate phylogenetic tree
 
 library(ape)
+library(MASS)
+library(sparseMVN)
+library(Matrix)
+library(dplyr)
 ## library(MASS)  ## for mvrnorm() ## don't load so we don't screw up dplyr::select()
-set.seed(seed)
+set.seed(tree_seed)
 
-phy <- rcoal(n = nspp)
+phy <- rtree(n = nspp)
 
-## Compute branch lengths using other methods (optional)
-#phy <- compute.brlen(phy, method = "Grafen", power = 0.5)
-
-# standardize the phylogenetic covariance matrix to have determinant 1 (optional)
 Vphy <- vcv(phy)
-#Vphy <- Vphy/max(Vphy)
-#Vphy <- Vphy/(det(Vphy)^(1/nspp)) ## for MCMCglmm?
 
-iD <- t(chol(Vphy))
-Xsd <- 1
+# Generate environmental site variable
 
-# Generate environmental site variable and standardize it
+X <- rnorm(n=nspp*nsite*nrep, sd=Xsd)
 
-if(nsite == 1){
-	X <- rnorm(n=nspp,sd=Xsd)
-#	X <- iD %*% rnorm(n=nspp,sd=Xsd)
-}
-if(nsite > 1){
-X <- rnorm(n = nspp*nsite, sd=Xsd)
-B_site <- matrix(1:nsite, nrow = 1, ncol = nsite)
-# X <- (X - mean(X))/sd(X)
-}
-site_name <- sapply(1:nsite, function(x){paste("site","_",x,sep="")})
+# Generate phylogenetic intercept and slope
 
-cormat <- matrix(c(1,rho.B01,rho.B01,1),2,2)
-sdvec <- c(sd.B0,sd.B1)
+phycormat <- matrix(c(1,phyrho.B01,phyrho.B01,1),2,2)
+physdvec <- c(physd.B0,physd.B1)
+phyvarmat <- physdvec %*% t(physdvec)
+phycovmat <- phyvarmat * phycormat
+
+
+phySigma <- kronecker(phycovmat,Vphy)
+
+b_phy <- MASS::mvrnorm(n=1
+	, mu=rep(c(beta0,beta1),each=nspp)
+	, Sigma=phySigma
+)
+
+Y.phy <- rep(head(b_phy, nspp), each = nrep*nsite) 
+X.phy <- rep(tail(b_phy, nspp), each = nrep*nsite)*X
+
+
[…]

### 2018-10-11 main.tex
minor typo in methods

@@
 They partition the phylogenetically correlated residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting}.
-However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/interept level variation) and residual variation is associated with observation level variation (the standard statistical definition).
-\mli{make a diagram!}
+However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/intercept level variation) and residual variation is associated with observation level variation (the standard statistical definition).
+\mli{See shiny}
 %\bmb{This is a good point, well expressed.
@@
 \end{align}
-where it contains two fixed effect parameters ($\beta_0$ and $\beta_1$), and four random effect parameters (phylogenetic random intercept, phylogenetic random slope, correlation between phylogenetic random slope and intercept and observation variance).
+where it contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept, phylogenetic random slope and correlation between phylogenetic random slope and intercept) and observation/residual variance.
 Predictor and intercept level random variation/effect of species are not applicable in this simulation setting with single observation per species because it will confound with the observational level variation (residuals).
@@
 (b_{sp_{int}}, b_{sp_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
-\sigma^2_{sp_{int}} & \sigma_{sp_{int},phy_{slope}} \\ 
+\sigma^2_{sp_{int}} & \sigma_{sp_{int},sp_{slope}} \\ 
 \sigma_{sp_{int},sp_{slope}} & \sigma^2_{sp_{slope}}

### 2018-10-12 main.tex
temp pushing results (not done)

@@
 
-The full model (which matches the simulation model that incorporates phylogenetic signal on the intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameters
+The full single site model using \pkg{lme4} (which matches the signle simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$) and have nominal coverage for the fixed effect parameters ($\beta_0$ and $\beta_1$) as number of species increase.
 
-In general, the phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) with more data, except PGLS in the single site simulations.
+In general, models that cannot match the simulation model (gls and phylolm) will try to fit the data with the available parameters.
+Phylolm (which lack the phylogenetic slope parameter) provides good estimates for phylogenetic intercept parameter ($\sigma_{phy}_{int}$) but overestimates the residual/observation standard deviation and have a bad coverage for the fixed slope parameter.
+
+the phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) with more data, except PGLS in the single site simulations.
 \bmb{(2) please define ``robustness'' -- usually means lack of sensitivity to model misspecification.

### 2018-10-13 collect.R
lme4 is definitely faster (good)

@@ -267,19 +267,19 @@ lme4ms_data <- lme4ms_results(lme4ms_res)
 pez_path <- "./datadir/pez/"
 pez_res <- list.files(path = pez_path)
 pez_results <- function(tt){
-  pez_df <- data.frame(resid = numeric(40)
-    , phylo_X = numeric(40)
-    , phylo_int = numeric(40)
+  pez_df <- data.frame(resid = numeric(174)
+    , phylo_X = numeric(174)
+    , phylo_int = numeric(174)
     , phylo_cor = NA
-    , phylo_interaction = numeric(40)
-    , species_X = numeric(40)
-    , species_int = numeric(40)
+    , phylo_interaction = numeric(174)
+    , species_X = numeric(174)
+    , species_int = numeric(174)
     , species_cor = NA
-    , site_int = numeric(40)
-    , B0 = numeric(40)
-    , B1 = numeric(40)
-    , model = numeric(40)
-    , time = numeric(40)
+    , site_int = numeric(174)
+    , B0 = numeric(174)
+    , B1 = numeric(174)
+    , model = numeric(174)
+    , time = numeric(174)
   )
   for(i in 1:length(tt)){
     pez_obj <- readRDS(paste(pez_path,tt[i],sep=""))
@@ -309,19 +309,19 @@ pez_data <- pez_results(pez_res)
 phyr_path <- "./datadir/phyr/"
 phyr_res <- list.files(path = phyr_path)
 phyr_results <- function(tt){
-  phyr_df <- data.frame(resid = numeric(40)
-                       , phylo_X = numeric(40)
-                       , phylo_int = numeric(40)
+  phyr_df <- data.frame(resid = numeric(58)
+                       , phylo_X = numeric(58)
+                       , phylo_int = numeric(58)
                        , phylo_cor = NA
-                       , phylo_interaction = numeric(40)
-                       , species_X = numeric(40)
-                       , species_int = numeric(40)
+                       , phylo_interaction = numeric(58)
+                       , species_X = numeric(58)
+                       , species_int = numeric(58)
                        , species_cor = NA
-                       , site_int = numeric(40)
-                       , B0 = numeric(40)
-                       , B1 = numeric(40)
-                       , model = numeric(40)
-                       , time = numeric(40)
+                       , site_int = numeric(58)
+                       , B0 = numeric(58)
+                       , B1 = numeric(58)
+                       , model = numeric(58)
+                       , time = numeric(58)
   )
   for(i in 1:length(tt)){
[…]

### 2018-10-13 main.tex

--- 2018-10-11 main.tex
@@
 
-\begin{itemize}
-\item{full model provides good estimates (except for correlation)}
-\item{reduced models try to fit with whatever parameters they have available (over-estimates the simulation parameters)}
-\item{no slope vs slope}
-\item{no correlation vs correlation}
-\item{as size/data/number of species increase, the parameter estimates gets better}
-\item{single site models are fast}
-\item{estimates gets closer to simulation parameter as number of species increase}
-\item{lme4/tmb are fast, everything else is slow for multiple sites}
-\end{itemize}
+% \begin{itemize}
+% \item{full model provides good estimates (except for correlation)}
+% \item{reduced models try to fit with whatever parameters they have available (over-estimates the simulation parameters)}
+% \item{no slope vs slope}
+% \item{no correlation vs correlation}
+% \item{as size/data/number of species increase, the parameter estimates gets better}
+% \item{single site models are fast}
+% \item{estimates gets closer to simulation parameter as number of species increase}
+% \item{lme4/tmb are fast, everything else is slow for multiple sites}
+% \end{itemize}
 
-The full model (which matches the simulation model that incorporates phylogenetic signal on the intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameters
+The full single site model using phyloglmm (which matches the signle simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$) and fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases.
 
-In general, the phylogenetic signal estimates are more robust for all models (including models without random slope and correlation) with more data, except PGLS in the single site simulations.
-\bmb{(2) please define ``robustness'' -- usually means lack of sensitivity to model misspecification.
+In general, models that cannot match the simulation model (PGLS and PGLM) will try to fit the data with the available parameters.
+PGLM (which lacks the phylogenetic slope parameter) provides good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation and have a bad coverage for the fixed slope parameter.
+PGLS, with only one parameter available, confounds all the other variations (phylogenetic slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept as well as large RMSE and overcovering for $\beta_0$, and bad estimates for $\beta_1$.
+\mli{double check this, what happens when X is also phylogenetically correlated?}
+
+% \bmb{(2) please define ``robustness'' -- usually means lack of sensitivity to model.
 Are you talking about the variance of the estimates?
@@
 Might be worth just labeling by $n=??$ rather than using small/medium/large}
-It is unclear how different levels of phylogenetic signal are being confounded under the error free (without tip variation) assumption for the PGLS models.
-\bmb{can we figure this out, even in an empirical/crude way??? it's tough to have results without knowing the true value they're aiming for ...}
-Models with only two parameters/levels of variation capture the true phylogenetic intercept and tip variation reasonably well despite neglecting phylogenetic variation in the slopes (\pkg{gls}) and correlation between the intercepts and slopes (\pkg{phylolm}).
+% It is unclear how different levels of phylogenetic signal are being confounded under the error free (without tip variation) assumption for the PGLS models.
+% \bmb{can we figure this out, even in an empirical/crude way??? it's tough to have results without knowing the true value they're aiming for ...}
+
 \bmb{say something about the implications of this [e.g.
@@
 (If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
-However, despite the similarities of robustness estimation, the efficiency (measured by computational speed) of sampling independent errors in \pkg{lme4} is dramatically better (orders of magnitude faster) than \pkg{pez}.
+\mli{save it for random slope discussion}
+
+However, despite the similarities of estimation, the efficiency (measured by computational speed) of sampling independent errors in \pkg{lme4} is  better than \pkg{pez}.
+For example, the median of 50 vs 300 seconds for the medium-size example.
 \bmb{be more specific, e.g. ``median of 8 vs 800 seconds for the medium-size example''. Clarify what ``independent errors'' means here?}
@@
 
-\bmb{Can you give more motivation for this? Why is it here?}
+% \bmb{Can you give more motivation for this? Why is it here?}
 

### 2018-10-13 main.tex
temp pushing, moving onto discussion (finally)

@@
-\bmb{stop at Results until further notice}
 \section{Introduction}
@@
 
-The full single site model using phyloglmm (which matches the signle simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$) and fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases.
+The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$).
+Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases.
+\mli{go back and define coverage and how its used}
 
-In general, models that cannot match the simulation model (PGLS and PGLM) will try to fit the data with the available parameters.
-PGLM (which lacks the phylogenetic slope parameter) provides good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation and have a bad coverage for the fixed slope parameter.
-PGLS, with only one parameter available, confounds all the other variations (phylogenetic slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept as well as large RMSE and overcovering for $\beta_0$, and bad estimates for $\beta_1$.
+In general, models that cannot match the simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
+PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation.
+$\beta_0$ estimates slightly undercover the nominal coverage (90\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
+PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept as well as large RMSE and overcovering for $\beta_0$, and bad estimates for $\beta_1$.
 \mli{double check this, what happens when X is also phylogenetically correlated?}
@@
 
-\bmb{say something about the implications of this [e.g.
-good enough for getting an overall impression of phylog.
-whatever-it-is but not enough for detailed description/inference about the evolutionary process?] (or maybe that should be saved for the Discussion?)}
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=4]{./git_push/plot.Rout.pdf}
+  \caption{new}
+\end{figure}
+\end{center}
 
-Platforms using the mixed model framework for multiple site simulations follow similar behaviour as the single-site simulation (i.e.
-different levels of phylogenetic signals and variations are closer to the simulated parameters with more data).
-All parameter estimates are much more robust for the multiple site than the single site simulations, especially slope-intercept correlations.
-The lack of phylogenetic correlation in \pkg{pez}'s statistical models does not appear to hurt its ability to estimate other aspects of phylogenetic signal; in fact, its slightly simpler model may be responsible for its slightly less variable estimates.
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=6]{./git_push/plot.Rout.pdf}
+  \caption{new}
+\end{figure}
+\end{center}
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=7]{./git_push/plot.Rout.pdf}
+  \caption{new}
+\end{figure}
+\end{center}
+
+Multiple site fits on the other hand are a lot more similar across platforms (as there are less deviates between the fitted and simulation model) compared to the single site fits.
+Similar to the single site fits, phyloglmm which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
+\mli{either keep it or take it out, because  will affect other parameters}
+The lack of correlation in \pkg{phyr}'s statistical models does not appear to have a large effect on its ability to estimate other parameters, whereas \pkg{pez} underestimates (lower estimates) many parameters.
 \bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate?
@@
 
-However, despite the similarities of estimation, the efficiency (measured by computational speed) of sampling independent errors in \pkg{lme4} is  better than \pkg{pez}.
-For example, the median of 50 vs 300 seconds for the medium-size example.
-\bmb{be more specific, e.g. ``median of 8 vs 800 seconds for the medium-size example''. Clarify what ``independent errors'' means here?}
-Furthermore, it was not practical for \pkg{pez} to fit large sample size simulation data as the computational speed did not scale linearly.
+However, despite the similarities of parameter estimates across platforms/methods for the multiple sites simulations fits, the efficiency (measured by computational speed) are different across playforms/methods and sample size.
+For example, the median time for \pkg{lme4} to fit 50/100 species model takes $~$ 50/100 seconds versus $~$ 200/2000 seconds for \pkg{pez} and \pkg{phyr}. 
+Lastly, it takes $~$ 2000 seconds for lme4 to fit 1000 species model and it was not practical for \pkg{pez} and \pkg{phyr} to fit 1000 species models where the computational speed did not scale linearly with sample size 25/50/100.
+\mli{put a lower bound to give up}
+
 \bmb{Can you put a lower bound on \pkg{pez}'s computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
@@
   approximate power?}
-
-\begin{center}
-\begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./csplot.pdf}
-  \caption{old}
-\end{figure}
-\end{center}
-
-\begin{center}
-\begin{figure}[h]
-  \includegraphics[scale=0.8,page=2]{./git_push/plot.pdf}
-  \caption{new}
-\end{figure}
-\end{center}
-
-% \begin{center}
-% \begin{figure}[h]
-% \includegraphics[scale=0.8,page=2]{./csplot.pdf}
-% \caption{\bmb{caption??}}
-% \end{figure}
-% \end{center}
-
-% \begin{center}
-% \begin{figure}[h]
-% \includegraphics[scale=1]{./ssplot.pdf}
-% \end{figure}
-% \end{center}
-
-% \bmb{Can you give more motivation for this? Why is it here?}
 
@@
 \section{Discussion}
+
+% \bmb{say something about the implications of this [e.g.
+good enough for getting an overall impression of phylog.
+whatever-it-is but not enough for detailed description/inference about the evolutionary process?] (or maybe that should be saved for the Discussion?)}
 

### 2018-10-13 main.tex
pushing ms

@@
 
-We have fitted models including maximum capabilities \bmb{???} to model varying levels of variation/phylogenetic signal to simulated data with our simple model with a single response and predictor variable.
-Using models that can include/account for multiple levels of variation is necessary for robust fits, but models that include phylogenetic signal in the intercept level can work as well as the full model.
-\bmb{clarify?}
-Under the Brownian motion evolution process assumption, modelling via independent errors is much more computationally efficient than imposing equivalent correlation structures.
-\bmb{clarify?
-you mean that assuming int/slope independence is much faster?
-Don't remember seeing such an example in the results?}
+The evolutionary model simulations presented in this paper are an oversimplification of real evolutionary processes in nature, but it is analogous to how researchers fit models to study complicated systems (as it is impossible to create a model to "match" the complexities of nature).
+We have fitted models using different platforms/methods that best match the simulation models. 
+Models that cannot match the simulation (full) model are expected to underperform; it is important to understand the limitations and performance of these simpler models/assumptions that are restricted by the commonly used methods available.
+Using models that include some form of phylogenetic variation and residual variation is necessary to \mli{correctly/robustly?} detect evolutionary process.
 
@@
 
-Process and observation errors are highly confounded in phylogenetic regression.
+Process and observation errors are highly confounded in phylogenetic regression if it is not handled correctly. 
 The effects of models that include both process and residual variations (i.e.
 PGLMM, PGLM) instead of the less realistic model that assumes variation due to evolution and zero observation (PGLS) was large.
-In particular, models that did not include residual variation tended to over-estimate the amount of variability in the evolutionary process. 
-Thus, any model that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e.
+% In particular, models that did not include residual variation tended to over-estimate the amount of variability in the evolutionary process. 
+Thus, for single measurement per species model, any method that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e.
 Pagel's $\lambda$, Blomberg's $K$).
+However, if multiple measurements/observations are considered/allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation.
+
+\mli{this is specifically targeting boettiger's critque on Pagel's lambda.
+https://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html
+The difference between the two figures is what it thinks the "tip" variation is.
+For the second plot, if we treat the sister taxas as "multiple" observation with some observation error on it, then you will get back to the first plot.}
+
 \bmb{again, we/you need to think carefully about what these coarse measures are good for; just avoiding mis-estimating the strength of trait/trait or trait/environment parameters?
@@
 In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying}.
-When analyzing evolutionary process among species between species traits and environmental variables, it is entirely plausible that the relationship across different traits and predictor variables are varying in a phylogenetically correlated way.
-Although the effects of including phylogenetically correlated random slopes are relatively low compared to process and residual variation in the intercept, it is still useful and informative than ignoring it from the model.
+When analyzing relationships among species between traits and environmental variables, it is entirely plausible that change in effect of predictor variables have on traits are changing in a phylogenetically correlated way.
+In fact, it is easier to interpret random slopes in the evolutionary process context (i.e, are size of different species phylogenetically correlated, or growth rate through time of different species phylogenetically correlated?)
+\mli{I like to think of it in a longitudinal setting of observing plant length/height.
+At time zero, they are all still seeds, one can imagine the growth rate are different plant species grow in a phylogenetically correlated way (similar species have similar growth rates).}
 It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \cite{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.

### 2018-10-16 main.tex
add ord 2010 ref on random slopes

@@
 Then you talk about correlation (which is a much more subtle/less important issue, I think).}
-In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying}.
+In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
 When analyzing relationships among species between traits and environmental variables, it is entirely plausible that change in effect of predictor variables have on traits are changing in a phylogenetically correlated way.

### 2018-10-30 phyloglmm_ms.tex
pushing to bb
temp pushing bits and pieces
first pass of the new intro, it is at the right length and a new story
working on ms
finish comments in the intro, moving on to methods
trying to fix bib?
adding matrix diagrams
pushing pseudo results to finish the paper
made most of the changes before simulation section

--- 2018-02-28 ms.tex
@@
 \documentclass[12pt]{article}
-\usepackage[]{graphicx}
-\usepackage[]{color}
+\usepackage{graphicx}
+\usepackage{grffile}
+\usepackage{color}
+\usepackage{soul}
 \usepackage{tabularx}
@@
 %% to suppress notes & comments: \usepackage[final]{changes}
-\usepackage[backgroundcolor=lightgray]{todonotes}
 \usepackage{setspace}
+\usepackage{changes}
+\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
+
 \bibliographystyle{chicago}
-\title{Need a good title. Submit to Methods of Ecology and Evolution or Evolutionary Biology}
+\title{Reformulating phylogenetic mixed models to improve flexibility and speed}
 \author{Michael Li and Ben Bolker}
@@
 
-\newcommand{\ml}[1]{{\color{red} ML: #1}}
+\newcommand{\mli}[1]{{\color{red} ML: #1}}
 
@@
 
-\keywords{phyloglmm ... }
+\keywords{phyloglmm, species--branch matrix, phylogenetic correlation, phylogenetic comparative methods}
 
-\include{main}
+\section{Introduction}
+
+Ever-increasing data collection capabilities (such as genomic sequencing, telemetry studies of animal behaviour, or remote sensing of environmental variables), in combination with large-scale synthesis of existing data bases on species occurrence and phenotypic traits, are making large volumes of biological data available over an ever-wider taxonomic range.
+In response, researchers are studying more complicated systems and fitting more models to larger data sets describing species occurrence and traits.
+One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting}. 
+Comparative analyses explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; researchers may be interested in controlling for the statistical effects of phylogenetic correlation, quantifying the evolutionary signals that underly traits or both.
+Unlike ordinary regression, where all of the variables of interest (for example, species trait and environmental factors) are directly observable, in phylogenetic regression, the evolutionary relationship between species are unobservable and assumed to evolve according to some underlying (latent) evolutionary process \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
+While evolutionary relationships are biologically important and interesting, they often get neglected in multi-species analyses \citep{bunnefeld2012island}. 
+Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling, and there exist a range of statistical tools for comparative analyses, but many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
+In such cases, researchers may try to find ways to simplify their analyses (i.e.
+simplifying the model by making assumptions of the evolutionary process) \citep{bunnefeld2012island, ord2010adaptation},
+%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
+%% ord2010adaptation : used OU process with means (single obs per taxa),
+%%  i.e. ignoring within-taxon variation?
+which can lead to statistical issues and erroneous conclusions
+\citep{felsenstein1985phylogenies, li2017statistical}.
+% \bmb{What precise ``issues'' do you have in mind?
+Don't think we need to state them here, but we should know.
+Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness?
+(2) not sure whether taking taxon-level means (Ord 2010) is actually a problem.
+Ignores some variation,
+% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
+In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
+% \bmb{nice!}
+
+There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
+% Phylogenetic signal : the magnitude of phylogenetic correlation
+The first challenge is to identify and segregate phylogenetic signals \citep{blomberg2003testing} and other sources of variation that are present in the data.  
+% \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
+You go on to talk about the tip-variation/residual-variation confounding issue.
+Maybe replace with something talking more specifically about resolving tip variation?}
+The classic phylogenetic regression assumes that phylogenetic correlation among the residuals from a regression between two species-level traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the residuals are normally distributed and observed without any additional error or within-species variation, we can use Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
+More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different response distributions and allowing extra parameters to account for additional complexities in the evolutionary process.
+In particular, they can partition residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
+However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/intercept level variation) and residual variation is associated with observation level variation within-species.
+\mli{See shiny}
+%\bmb{This is a good point, well expressed.
+Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel?
+Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
+%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
+Despite multiple observations per tip taxon (species) as typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
+% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
+
+%% FIXME
+The second challenge is flexibly model phylogenetic relations.
+While it is reasonable to examine evolutionary relationships between species with their response, it is plausible the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated.
+%\bmb{maybe we can focus on random slopes here?
+Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
+Several recent studies have incorporated phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
+For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
+The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
+Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.
+
+\newcommand{\pkg}[1]{{\tt #1}}
+\newcommand{\code}[1]{{\tt #1}}
+
+\begin{tabularx}{\textwidth}{|X|X|X|X|}
+\hline
+Model & Method & Data & Platform \\
+\hline
+\hline
+GLM (Simple)  & No Residual & Single Observation & GLS, PIC \\
+\hline
+              & Residual    & Single Observation & Pagel's $\lambda$ \\
+              & + phylo intercept &                    & Blomberg's $K$ \\ 
+              &             &                    & via phylolm, GLS \\
+\hline
+GLMM (Complex)& Random effects & Single Observation & \pkg{pez} \\ 
+              &                & Balanced Design &  \\
+\hline
+              &                & Unrestricted  & \pkg{lme4}, \pkg{glmmTMB} \\
+\hline
+Bayesian GLMM &                & Balanced Design & \pkg{MCMCglmm} \\ 
+\hline        
+              &                & Unrestricted   & \pkg{brms} \\
+\hline
+\end{tabularx}
+
+However, in this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches.
+Our new formulation is more flexible; it can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
+We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing methods/platforms (i.e.
+\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
+We also expect that it will be faster, because \pkg{lme4} and \pkg{glmmTMB} are optimized in performance using advanced linear algebra techniques. 
+\mli{fix me}
+In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
+We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.
+
+\section{Materials and Methods}
+
+\newcommand{\bX}{{\mathbf X}}
+\newcommand{\bbeta}{{\boldsymbol \beta}}
+\newcommand{\bmu}{{\boldsymbol \mu}}
+\newcommand{\bY}{{\mathbf Y}}
+\newcommand{\bC}{{\mathbf C}}
+\newcommand{\bZ}{{\mathbf Z}}
+\newcommand{\bb}{{\mathbf b}}
+\newcommand{\bSigma}{{\boldsymbol \Sigma}}
+
+The standard phylogenetic regression takes the form
+
+%\bmb{boldface matrices/vectors i.e.
+$\bX \bbeta$?
+or not worth the trouble?
+How about $\bmu=\bX \bbeta$; $\bY \sim \textrm{MVN}(\bmu,\sigma^2 \bC)$ for greater generality?
+(OK, I see that you move in this direction later.
+But I'm not sure it's worth the detour of using two different notations?
+even if you are going to extend to include fixed effects, $Z$, etc.
+later ...}:
+
+\begin{align}
+\bmu & = \bX \bbeta  \label{eq:gls1} \\ 
+\bY & \sim \textrm{MVN}(\bmu,\sigma^{2} \bC), \label{eq:gls2}
+\end{align}
+where $\bX$ is an $n \times m$ model matrix containing $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector which is assumed to be multivariate normally distributed with mean $\bmu$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation matrix.
+The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.
+
+%\bmb{give a ref for this? maybe Paradis's ape book, or Garamszegi's book, or ?}
+%\mli{chapter 7 in garamszegi 2014}
+
+% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
+%\mli{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
+An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
+The typical linear mixed model has the form:
+\begin{align}
+\bY & \sim \textrm{Distr}(\bmu) \label{eq:glmm1} \\
+g(\bmu) & = \bX \bbeta + \bZ \bb \label{eq:glmm2} \\
+\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3}
+\end{align}
+where $Z$ is an $n \times m$ model matrix for the $m$ -- dimensional vector-valued $m$ predictor variables; $\bb$ represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$.
+Analogously, the phylogenetic regression given by \ref{eq:gls1} can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$.
+
+\bmb{worth including the \pkg{MCMCglmm}/\pkg{brms} inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
+\mli{inverse VCV for BLUP? an alternative way to find the inverse of the vcv(phy)}
+\mli{G vs R, PIC and simple variations (phylolm ...) uses R and no G}
+
+\subsection{Reformulating phylogenetic covariance matrix}
+% The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
+% Phylogenetic independent contrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
+% Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.
+
+\newcommand{\bS}{{\mathbf S}}
+\newcommand{\bJ}{{\mathbf J}}
+\newcommand{\bB}{{\mathbf B}}
+\newcommand{\bBadj}{{\mathbf B}_{\mbox{\tiny adj}}}
+\newcommand{\bomega}{{\boldsymbol \omega}}
+\newcommand{\bell}{{\boldsymbol \ell}}
+\newcommand{\e}{{ \epsilon}}
+
+An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
+In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk with daughter species evolving independently after speciation.
+In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
+Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $\bS$, equivalently to imposing a correlation $\bC$.
+For example, as shown in figure \ref{fig:tree}, the corresponding $\bS$ is in the form:
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=1]{./git_push/tree.pdf}
+  \caption{\bmb{caption ??}}
+\label{fig:tree}
+\end{figure}
+\end{center}
+
+\begin{align*}
+\bS = \begin{bmatrix}
+\ell_1 \e_1 & \ell_2 \e_2 & 0 & 0 \\
+\ell_1 \e_1 &  0 & \ell_3 \e_3 & 0 \\
+0  &  0 & 0  & \ell_4 \e_4
+\end{bmatrix}
+\end{align*}
+The error value corresponding to tip 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i$ is the square root of the branch length and the $\ell_i$ are independent, homoscedastic Normal variates (i.e.
+variance is $\in (\ell_1 \e_1 + \ell_2 \e2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$, linear in branch length)
+
+\subsubsection*{Constructing the species--branch random effects model matrix}
+
+The $\bS$ is a matrix product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices, and a branch length vector $\bell$.
+
+\begin{align}
+\bS_{ind} = \begin{bmatrix}
+1 & 1 & 0 & 0 \\ 
+1 & 0 & 1 & 0 \\ 
+0 & 0 & 0 & 1
+\end{bmatrix} , 
+\bell = \begin{bmatrix}
+\ell_1 \\
+\ell_2 \\
+\ell_3 \\
+\ell_4 
+\end{bmatrix}
+\label{eq:S}
+\end{align}
+$\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
+$\bS \bS^T$ gives the variance-covariance matrix of the phylogeny.
+
+The random-effect model matrix $\bZ$ can be decomposed into term-wise model matrix $\bZ_{i}$ as described in \citet{bates2015fitting}.
+Thus, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is
+
+\begin{equation}
+\bZ^{C}_{i} = (\bS^{T}_{i}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
+\end{equation}
+where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{i}$ is the $m_{i} \times b_{i}$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m_i$ indicator matrix of grouping factors; $X_{i}$ is the $n \times p_{i}$ raw random effects model matrix.
+
+\mli{move example to the supp}
+% For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix,
+% 
+% \begin{align}
+% \bX = \begin{bmatrix}
+% 1 & t_1  \\ 
+% 1 & t_2  \\ 
+% 1 & t_3 
+% \end{bmatrix} 
+% \end{align}
+% then the term-wise phylogenetic random effects model matrix is,
+% 
+% \begin{align}
+% \bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
+% \Bigg[
+% \Bigg(
+% \begin{bmatrix}
+% \ell_{adj1} & \ell_{adj1}  & 0 \\
+% \ell_{adj2} &  0  & 0 \\
+% 0  &  \ell_{adj3} & 0 \\
+% 0 & 0 &  \ell_{adj4} 
+% \end{bmatrix}
+% \begin{bmatrix}
+% 1 & 0  & 0 \\
+% 0 & 1  & 0 \\
+% 0 & 0  & 1  
+% \end{bmatrix}
+% \Bigg)
+% \ast
+% \begin{bmatrix}
+% 1   & 1   & 1  \\ 
+% t_1 & t_2 & t_3
+% \end{bmatrix} 
+% \Bigg]^T
+% \\
+% = \begin{bmatrix}
+% \ell_{adj1} & \ell_{adj1}t_1 & \ell_{adj2} & \ell_{adj2}t_1 & 0 & 0 & 0 & 0 \\
+% \ell_{adj1} & \ell_{adj1}t_2 & 0 & 0 & \ell_{adj3} & \ell_{adj3}t_2 & 0 & 0 \\
+% 0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
+% \end{bmatrix}
+% \end{align}
+
+% 
+% The scaled species--branch matrix $\bS_{adji}$ is scaled version of $\bS$ via the standardized generalized variance (SGV).
+% SGVs are used as an intrinsic/natural way to scale down the variance-covariance matrix measured in different scales (for example, phylogenetic distance, years and etc) compariable to other variation in the model \citep{sengupta1987tests}.
+% SGV is the $nth$ root of the generalized variance (GV), or the determinant of the variance-covariance matrix, where GV measures the overall size or volume of the variance-covariance matrix.  
+% The SGV is given by:
+% \begin{align}
+% \bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
+% \end{align}
+% where $n$ is the number of species in the phylogeny.
+% 
+% The analog in the $\bS$ is to scale the branch vector $\bell$ by square root of the SGV:
+% \begin{align}
+% \bell_{adj} & = \frac{\bell}{\sqrt{\bomega}}
+% \end{align} 
+% and the $\bS_{adj}$ is the matrix product of $\bS_{ind}$ and $\bell_{adj}$.
+
+\subsection{Simulation}
+
+We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species.
+For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
+For single-site model, the full simulation model is the following:
+\begin{align}
+\bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + b_{obs}\\
+(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
+\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
+\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
+\end{bmatrix}
+\Bigg) \\
+b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
+\end{align}
+where it contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept, phylogenetic random slope and correlation between phylogenetic random slope and intercept) and observation/residual variance.
+Predictor and intercept level random variation/effect of species are not applicable in this simulation setting with single observation per species because it will confound with the observational level variation (residuals).
+
+We extend the simulation model by adding multiple sites where each site will have one observation per species. The multiple-site full model is the following: 
+\begin{align}
+\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site} + b_{obs}\\
+(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
+\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
+\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
+\end{bmatrix}
+\Bigg) \\
+(b_{sp_{int}}, b_{sp_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
+\sigma^2_{sp_{int}} & \sigma_{sp_{int},sp_{slope}} \\ 
+\sigma_{sp_{int},sp_{slope}} & \sigma^2_{sp_{slope}}
+\end{bmatrix}
+\Bigg) \\
+b_{site} & \sim \textrm{MVN} ( 0 , \Sigma_{site}) \\
+b_{sp:site} & \sim \textrm{MVN} (0, \textrm{Kron}(\bf{I}_{site}, \sigma^2_{phy})) \\
+b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
+\end{align}
+The multiple-site full simulation model will have five additional random effects (predictor and intercept level random effect of species and their correlation, random intercept of site and random intercept of species site interaction) compared to the single-site full model.
+Predictor and intercept level random effect of species are now applicable in the multiple-site model because there are multiple (one in every site) observations per species. 
+Random intercept of species and site interactions is an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
+
+\subsection{Platforms}
+
+Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables. 
+We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \mli{citation}.
+We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{MCMCglmm} \cite{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}.
+% \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
+% \begin{verbatim}
+% grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
+% invert=TRUE,value=TRUE)
+% \end{verbatim}
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow observational residuals.
+Both packages can do other evolutionary models and different correlation structures (i.e.
+Pagel's $\lambda$, Blombergs $K$ and etc.), but we restrict our PGLS fits to the simple error-free BM correlation.
+Neither PGLS nor PGLM can multiple observations or predictor level variation \mli{intercept level vs predictor level?}. 
+One of the relatively few packages that currently fit phylogenetic correlations to predictor level variationis is \pkg{pez}, which can handle adding additional uncorrelated random slopes and interactions with intercept.
+
+\bmb{I like the commented-out table here (maybe in supp material? or maybe useful enough that it's worth including in the main text?) although it could take a lot of fussing to get it right}
+
+% Table \mli{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
+
+% \begin{table}
+% \begin{center}
+
+\begin{tabularx}{\textwidth}{|c||c|c|c|c|c|c|}
+\hline
+Package	& \small{nlme}	& \small{phylolm}	& \small{lme4/glmmTMB} & \small{pez} & \small{brms} & \small{MCMCglmm} \\
+\hline
+\hline
+Single Site & & & & & & \\
+Random effects & X & X & X & & X & X \\
+\hline
+Phylo intercept & X & X & X &  & X & X \\
+Phylo slope &  &  & X &  & X & X \\
+Phylo correlation &  &  & X &  & X & X \\
+Residual & & X & X &  & X & X \\
+\hline
+\hline
+Multiple Site & & & & & & \\
+Random effects &  &  & X & X & X & X \\
+\hline
+Phylo intercept &  &  & X & X & X & X \\
+Phylo slope &  &  & X & X & X & X \\
+Phylo correlation &  &  & X &  & X & X \\
+Phylo interaction & &  & X & X & X &  \\
+Species intercept & &  & X & X & X & X \\
+Species slope & & & X & X & X & X \\
+Species correlation & &  & X &  & X & X \\
+Residual & &  & X & X & X & X \\
+\hline
+
+\end{tabularx}
+% \vspace{1in}
+% ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
+% \end{center}
+% \end{table}
+
+\section{Results}
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=1]{./git_push/plot.Rout.pdf}
+  \caption{New}
+\label{ssplot}
+\end{figure}
+\end{center}
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=2]{./git_push/plot.Rout.pdf}
+  \caption{New}
+\label{ssplot}
+\end{figure}
+\end{center}
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=3]{./git_push/plot.Rout.pdf}
+  \caption{New}
+\label{ssplot}
+\end{figure}
+\end{center}
+
+\bmb{Need to think a little harder about the important points/order here (might be time to do the exercise where you take the results and boil them back down to short bullet points, in order to figure out whether you've said the important stuff, in the right order).
+Maybe lead by saying that all platforms give similar answers - to the extent that they are fitting the same statistical model (this might even be emphasized/explained above, in Methods).
+Am I correct in believing that phylolm=\pkg{lme4} but without intercept/slope correlation, and gls is additionally missing random slopes?}
+
+\bmb{add refs to figures \ldots}
+
+% \begin{itemize}
+% \item{full model provides good estimates (except for correlation)}
+% \item{reduced models try to fit with whatever parameters they have available (over-estimates the simulation parameters)}
+% \item{no slope vs slope}
+% \item{no correlation vs correlation}
+% \item{as size/data/number of species increase, the parameter estimates gets better}
+% \item{single site models are fast}
+% \item{estimates gets closer to simulation parameter as number of species increase}
+% \item{lme4/tmb are fast, everything else is slow for multiple sites}
+% \end{itemize}
+
+The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$).
+Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases.
+\mli{go back and define coverage and how its used}
+
+In general, models that cannot match the simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
+PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation.
+$\beta_0$ estimates slightly undercover the nominal coverage (90\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
+PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept as well as large RMSE and overcovering for $\beta_0$, and bad estimates for $\beta_1$.
+\mli{double check this, what happens when X is also phylogenetically correlated?}
+
+% \bmb{(2) please define ``robustness'' -- usually means lack of sensitivity to model.
+Are you talking about the variance of the estimates?
+I'd prefer not to use ``robust'' in a non-standard way if possible (2) what are small/medium/large?
+I see that $n=\{20,100,500\}$ and $m=\{1,20\}$, but not sure how these match up to ``small'', ``medium'', ``large''; is Figure 1 all $m=1$?
+Might be worth just labeling by $n=??$ rather than using small/medium/large}
+% It is unclear how different levels of phylogenetic signal are being confounded under the error free (without tip variation) assumption for the PGLS models.
+% \bmb{can we figure this out, even in an empirical/crude way??? it's tough to have results without knowing the true value they're aiming for ...}
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=4]{./git_push/plot.Rout.pdf}
+  \caption{new}
+\end{figure}
+\end{center}
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=6]{./git_push/plot.Rout.pdf}
+  \caption{new}
+\end{figure}
+\end{center}
+
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8,page=7]{./git_push/plot.Rout.pdf}
+  \caption{new}
+\end{figure}
+\end{center}
+
+Multiple site fits on the other hand are a lot more similar across platforms (as there are less deviates between the fitted and simulation model) compared to the single site fits.
+Similar to the single site fits, phyloglmm which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
+\mli{either keep it or take it out, because  will affect other parameters}
+The lack of correlation in \pkg{phyr}'s statistical models does not appear to have a large effect on its ability to estimate other parameters, whereas \pkg{pez} underestimates (lower estimates) many parameters.
+\bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate?
+Are these singular models?
+Would they be diagnosable?
+Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem?
+(If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
+\mli{save it for random slope discussion}
+
+However, despite the similarities of parameter estimates across platforms/methods for the multiple sites simulations fits, the efficiency (measured by computational speed) are different across playforms/methods and sample size.
+For example, the median time for \pkg{lme4} to fit 50/100 species model takes $~$ 50/100 seconds versus $~$ 200/2000 seconds for \pkg{pez} and \pkg{phyr}. 
+Lastly, it takes $~$ 2000 seconds for lme4 to fit 1000 species model and it was not practical for \pkg{pez} and \pkg{phyr} to fit 1000 species models where the computational speed did not scale linearly with sample size 25/50/100.
+\mli{put a lower bound to give up}
+
+\bmb{Can you put a lower bound on \pkg{pez}'s computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
+  or (easier) do a brute-force series of increasing sample size and
+  compute a log-log regression of time vs sample size to find the
+  approximate power?}
+
+We used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \cite{li2017canfun}. 
+We have obtained identical results for fix and random effect estimates and approximately 120 times faster compared to \pkg{pez}. 
+Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.
+
+\newpage
+
+\section{Discussion}
+
+% \bmb{say something about the implications of this [e.g.
+good enough for getting an overall impression of phylog.
+whatever-it-is but not enough for detailed description/inference about the evolutionary process?] (or maybe that should be saved for the Discussion?)}
+
+The evolutionary model simulations presented in this paper are an oversimplification of real evolutionary processes in nature, but it is analogous to how researchers fit models to study complicated systems (as it is impossible to create a model to "match" the complexities of nature).
+We have fitted models using different platforms/methods that best match the simulation models. 
+Models that cannot match the simulation (full) model are expected to underperform; it is important to understand the limitations and performance of these simpler models/assumptions that are restricted by the commonly used methods available.
+Using models that include some form of phylogenetic variation and residual variation is necessary to \mli{correctly/robustly?} detect evolutionary process.
+
+\subsection{Process and residual (observation) variabilities and confoundings}
+
+Process and observation errors are highly confounded in phylogenetic regression if it is not handled correctly. 
+The effects of models that include both process and residual variations (i.e.
+PGLMM, PGLM) instead of the less realistic model that assumes variation due to evolution and zero observation (PGLS) was large.
+% In particular, models that did not include residual variation tended to over-estimate the amount of variability in the evolutionary process. 
+Thus, for single measurement per species model, any method that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e.
+Pagel's $\lambda$, Blomberg's $K$).
+However, if multiple measurements/observations are considered/allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation.
+
+\mli{this is specifically targeting boettiger's critque on Pagel's lambda.
+https://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html
+The difference between the two figures is what it thinks the "tip" variation is.
+For the second plot, if we treat the sister taxas as "multiple" observation with some observation error on it, then you will get back to the first plot.}
+
+\bmb{again, we/you need to think carefully about what these coarse measures are good for; just avoiding mis-estimating the strength of trait/trait or trait/environment parameters?
+Is it worth doing some sims on that, or do we basically believe if the total amount of phylogenetic signal is correctly assessed, that we get the right answer for the trait effect?
+Is \cite{schielzeth2009conclusions} relevant/a useful analogy?
+(You cite it below, but there are two separate questions: (1) whether including random slopes gives us more insight into evol processes; (2) whether omitting random slopes gives us bad answers even when we're not interested in them.) Worth citing/discussing \cite{uyeda2018rethinking} (as an example of even deeper analysis of evolutionary processes, beyond BM etc.)?}
+
+\subsection{Random Slopes}
+
+\bmb{I think you need a lot more here; what does a random slopes model mean?
+You have one sentence that says ``people should be thinking about random slopes''.
+Then you talk about correlation (which is a much more subtle/less important issue, I think).}
+In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+When analyzing relationships among species between traits and environmental variables, it is entirely plausible that change in effect of predictor variables have on traits are changing in a phylogenetically correlated way.
+In fact, it is easier to interpret random slopes in the evolutionary process context (i.e, are size of different species phylogenetically correlated, or growth rate through time of different species phylogenetically correlated?)
+\mli{I like to think of it in a longitudinal setting of observing plant length/height.
+At time zero, they are all still seeds, one can imagine the growth rate are different plant species grow in a phylogenetically correlated way (similar species have similar growth rates).}
+It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \cite{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
+Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
+and Are we making a mistake by ignoring random slopes \cite{schielzeth2008conclusions}?
+However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
+
+\subsection{Extension and alternatives}
+
+Our analysis covers the classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models).
+Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-BM evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \cite{butler2004phylogenetic}), Bayesian approaches \cite{hadfield2010general}.
+More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis.
+This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as \pkg{MCMCglmm} \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
+However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e.
+What should we do when we don't have enough data?
+Should we use a complex model and overfit \cite{barr2013random}, the right balance of complexity and data \cite{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
+More importantly, this implementation in \pkg{lme4} allows users to fit phylogenetic mixed models to the fullest ( that other frequentist platforms lack of (i.e.
+large data, unbalanced species observations, complex random effects) and explore new ideas.
+
+\section{Conclusion}
+
+We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical PCM to simulated data. 
+First, this new approach is magnitudes faster compared to existing phylogenetic mixed models without losing robustness/accuracy and can easily combined with any statistically mixed model framework. 
+Last and most importantly, it is more flexible in fitting large phylogenies, large volume of data, unbalanced data sets, and complex random effects such as slope correlations.
+
+\newpage
+
+\section{Supplements -- translation}
+
+$1 \mid Sp_{phylo}$, $0 + X_{E} \mid Sp_{phylo}$ and $X_{E} \mid Sp_{phylo}$ means phylogenetically related have similar response,
+
+\begin{tabularx}{\textwidth}{|l|X|X|}
+\hline
+Formula & Statistics & Biology \\
+\hline
+$1 \mid Sp$ &
+random species intercept; variation within species in mean response across all factors &
+variation of how species respond \\
+\hline
+
+$0 + X_{E} \mid Sp$ &
+random slope of environment factor within species; variation in coefficient within species for the environmental factor &
+variation of how species respond to the same environmental factor \\
+\hline
+
+$1 + X_{E} \mid Sp$ &
+random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental factor with correlated mean response across all other factors &
+variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
+\hline
+
+$1 \mid Site:Sp $ &
+random variation in intercept among species within sites &
+variation of how species respond within sites \\
+\hline
+
+$1 | Sp_{Phylo} $ &
+variation among species in mean response across all factors demonstrate phylogenetic signal &
+phylogenetically related species respond similarly \\
+\hline
+
+$0 + X_{E} \mid Sp_{Phylo}$ &
+variation among species for environmental factors demonstrate phylogenetic signal &
+phylogenetically related species respond similar (share common response) to the same environmental factor \\
+\hline
+\end{tabularx}
+
+\bibliography{phyloglmm}
 

### 2018-11-01 phyloglmm_ms.tex
adding another subsection before results

@@
 Unlike ordinary regression, where all of the variables of interest (for example, species trait and environmental factors) are directly observable, in phylogenetic regression, the evolutionary relationship between species are unobservable and assumed to evolve according to some underlying (latent) evolutionary process \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
-While evolutionary relationships are biologically important and interesting, they often get neglected in multi-species analyses \citep{bunnefeld2012island}. 
-Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling, and there exist a range of statistical tools for comparative analyses, but many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
+% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
+Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic} , and recently, a demend of integration of evolutionary knowledge in studying ecological applications \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}
+%such as ecological conser, the effects of climate change  \citep{santamaria2012evolution}. \mli{cite more here}.
+While there exist a range of statistical tools for comparative analyses, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
 In such cases, researchers may try to find ways to simplify their analyses (i.e.
-simplifying the model by making assumptions of the evolutionary process) \citep{bunnefeld2012island, ord2010adaptation},
+controlling for individual-varation of species and neglecting phylogenetic correlation between species \citep{bunnefeld2012island}, comparing species family \citep{tella1999habitat}, neglecting within-species variation \citep{ord2010adaptation}, and etc).
 %% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
@@
 %%  i.e. ignoring within-taxon variation?
-which can lead to statistical issues and erroneous conclusions
-\citep{felsenstein1985phylogenies, li2017statistical}.
+%% fix me!?
+% which can lead to statistical issues and erroneous conclusions
+% \citep{felsenstein1985phylogenies, li2017statistical}.
 % \bmb{What precise ``issues'' do you have in mind?
@@
 % Phylogenetic signal : the magnitude of phylogenetic correlation
-The first challenge is to identify and segregate phylogenetic signals \citep{blomberg2003testing} and other sources of variation that are present in the data.  
+The first challenge is to identify and segregate phylogenetic signals \citep{blomberg2003testing} from other sources of variation that are present in the data.  
 % \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
@@
 Maybe replace with something talking more specifically about resolving tip variation?}
-The classic phylogenetic regression assumes that phylogenetic correlation among the residuals from a regression between two species-level traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the residuals are normally distributed and observed without any additional error or within-species variation, we can use Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
+The classic phylogenetic regression assumes that phylogenetic correlation among the residuals from a regression between two species-level traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the residuals are normally distributed and observed without any additional error or within-species variation, we can use Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies, nicolakakis2000forebrain}.
 More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different response distributions and allowing extra parameters to account for additional complexities in the evolutionary process.
@@
 However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/intercept level variation) and residual variation is associated with observation level variation within-species.
-\mli{See shiny}
 %\bmb{This is a good point, well expressed.
@@
 %\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
-Despite multiple observations per tip taxon (species) as typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
+Despite having multiple observations per tip taxon (species), a typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
 % \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
@@
 Several recent studies have incorporated phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
-For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance.
+For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach.
 The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
@@
 We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing methods/platforms (i.e.
-\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
-We also expect that it will be faster, because \pkg{lme4} and \pkg{glmmTMB} are optimized in performance using advanced linear algebra techniques. 
-\mli{fix me}
+\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
+We also expect that it will be faster, because \pkg{lme4} and \pkg{glmmTMB} are optimized in performance using efficient computational linear algebra techniques. 
 In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
@@
 \newcommand{\bb}{{\mathbf b}}
+\newcommand{\be}{{\mathbf e}}
 \newcommand{\bSigma}{{\boldsymbol \Sigma}}
@@
 \bY & \sim \textrm{Distr}(\bmu) \label{eq:glmm1} \\
-g(\bmu) & = \bX \bbeta + \bZ \bb \label{eq:glmm2} \\
-\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3}
+g(\bmu) & = \bX \bbeta + \bZ \bb + \be \label{eq:glmm2} \\
+\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
+\be & \sim \textrm{MVN}(0, \sigma^2_{residual} \label{eq:glmm3})
 \end{align}
-where $Z$ is an $n \times m$ model matrix for the $m$ -- dimensional vector-valued $m$ predictor variables; $\bb$ represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$.
-Analogously, the phylogenetic regression given by \ref{eq:gls1} can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$.
+where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\be$ (sometimes referred as "R" effect) represents the residuals.
+Analogously, the phylogenetic regression given by \ref{eq:gls1} can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^2_{residual} = 0$.
 
-\bmb{worth including the \pkg{MCMCglmm}/\pkg{brms} inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
-\mli{inverse VCV for BLUP? an alternative way to find the inverse of the vcv(phy)}
-\mli{G vs R, PIC and simple variations (phylolm ...) uses R and no G}
+% \bmb{worth including the \pkg{MCMCglmm}/\pkg{brms} inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
+% \mli{inverse VCV for BLUP? an alternative way to find the inverse of the vcv(phy)}
+% \mli{G vs R, PIC and simple variations (phylolm ...) uses R and no G}
 
@@
 \end{align*}
-The error value corresponding to tip 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i$ is the square root of the branch length and the $\ell_i$ are independent, homoscedastic Normal variates (i.e.
-variance is $\in (\ell_1 \e_1 + \ell_2 \e2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$, linear in branch length)
+The error value corresponding to tip 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i$ is the square root of the branch length $L_i$ in figure \ref{fig:tree} and the $\ell_i$ are independent, homoscedastic Normal variates with zero mean and $\sigma^2$ variance (i.e.
+variance for tip 1 is $(\ell_1 \e_1 + \ell_2 \e_2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$, linear in branch length)
 
@@
 
-The $\bS$ is a matrix product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices, and a branch length vector $\bell$.
+The $\bS$ matrix is the matrix product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices, and a branch length vector $\bell$.
 
@@
 \label{eq:S}
-\end{align}
+\end{align}.
 $\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
@@
 
-We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species.
+We generated test data based on the simple mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
 For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
@@
 \begin{align}
-\bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + b_{obs}\\
+\bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + e_{residual}  \label{eq:ss_sim}\\\
 (b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
@@
 \Bigg) \\ 
-b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
+e_{residual} & \sim \textrm{MVN} ( 0 , \Sigma) .
 \end{align}
-where it contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept, phylogenetic random slope and correlation between phylogenetic random slope and intercept) and observation/residual variance.
-Predictor and intercept level random variation/effect of species are not applicable in this simulation setting with single observation per species because it will confound with the observational level variation (residuals).
+The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance, phylogenetic random slope variance and covariance between phylogenetic random slope and intercept) and observation/residual variance.
+Predictor and intercept level random effect of species are not applicable in this simulation setting with single observation per species because it will be confounded with the observational level variation (residuals).
 
-We extend the simulation model by adding multiple sites where each site will have one observation per species. The multiple-site full model is the following: 
+We extend the simulation model by adding multiple sites where each site has one observation per species. The multiple-site full model is the following: 
 \begin{align}
-\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site} + b_{obs}\\
+\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site} + e_{residual} \label{eq:ms_sim}\\
 (b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
@@
 b_{sp:site} & \sim \textrm{MVN} (0, \textrm{Kron}(\bf{I}_{site}, \sigma^2_{phy})) \\
-b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
+e_{residual} & \sim \textrm{MVN} ( 0 , \Sigma)
 \end{align}
-The multiple-site full simulation model will have five additional random effects (predictor and intercept level random effect of species and their correlation, random intercept of site and random intercept of species site interaction) compared to the single-site full model.
+The multiple-site full simulation model has five additional random effects (predictor and intercept level random effect of species variance and their covariance, random intercept of site and random intercept of species-site interaction) compared to the single--site full model.
 Predictor and intercept level random effect of species are now applicable in the multiple-site model because there are multiple (one in every site) observations per species. 
-Random intercept of species and site interactions is an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
+Random intercept of species--site interactions examines whether the species within a comunity are likely to contain stronger phylogenetic relatedness than expected by chance; which is equivalent to phylogenetic attraction described in \cite{helmus2007separating}.
 
@@
 
-Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables. 
-We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \mli{citation}.
-We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{MCMCglmm} \cite{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}.
+Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as part of a linear or generalized linear model. 
+We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}.
+We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, and \pkg{brms} \citep{burkner2016brms}.
 % \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
@@
 % \end{verbatim}
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
-Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow observational residuals.
-Both packages can do other evolutionary models and different correlation structures (i.e.
-Pagel's $\lambda$, Blombergs $K$ and etc.), but we restrict our PGLS fits to the simple error-free BM correlation.
-Neither PGLS nor PGLM can multiple observations or predictor level variation \mli{intercept level vs predictor level?}. 
-One of the relatively few packages that currently fit phylogenetic correlations to predictor level variationis is \pkg{pez}, which can handle adding additional uncorrelated random slopes and interactions with intercept.
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it is simily fitting a linear model where the covariance structure between species is the expected Brownian motion evolution process (or other evolutionary process) on the tree instead of independent and identically distributed.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow additional observational residuals and also non-Gaussian response variables.
+Both packages can model other evolutionary processes and different correlation structures (i.e. Pagel's $\lambda$, Blomberg's $K$ etc.), but we restrict our PGLS fits to the simple BM correlation. 
+Neither PGLS nor PGLM can handle multiple observations or predictor level variation.
+One of the relatively few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle adding additional uncorrelated random slopes and interactions with intercept.
+Lastly, bayesian phylogenetic GLMM using Markov chain monte carlo (MCMC) are avalible and can handle all cases described above. 
+However, MCMC is often computationally expensive for GLMMs compared to platforms using optimization.
+\pkg{MCMCglmm} \citep{hadfield2010general} is the most commonly used bayesian phylogenetic GLMM, but we will use \pkg{brms} instead, which uses a more powerful MCMC technique called Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid} that is more efficient \citep{li2018fitting}.
 
@@
 % \begin{center}
-
+\begin{table}
 \begin{tabularx}{\textwidth}{|c||c|c|c|c|c|c|}
@@
 \hline
-Single Site & & & & & & \\
-Random effects & X & X & X & & X & X \\
+Single Site & X & X & X & & X & X \\
 \hline
@@
 \hline
-Multiple Site & & & & & & \\
-Random effects &  &  & X & X & X & X \\
+Multiple Site &  &  & X & X & X & X \\
 \hline
@@
 \hline
-
 \end{tabularx}
@@
 % \end{center}
-% \end{table}
+\caption{bla bla bla \label{model_variant}}
+\end{table}
+
+\subsection{Simulation and evaluations}
+We simulated 100 phylogenetic trees for each sample size (25,50,100) and incorporated with simulation parameters (illustrated in figure 2 and 5) to simulate responses via the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
+All model variants via different platforms were used to fit each realization.
+Tables \ref{model_variant} shows the parameters that are estimable for each model/platform. 
+We combined two convergence criteria to assess convergence for the main parameters (R0, Peff, Prep): we required a value of the
+Gelman and Rubin statistic and an effective sample size (ESS) greater than 400 for each replication. For
+each replication we sample four chains starting with 4000 iterations; w
 
@@
 
-\bmb{Need to think a little harder about the important points/order here (might be time to do the exercise where you take the results and boil them back down to short bullet points, in order to figure out whether you've said the important stuff, in the right order).
+% \bmb{Need to think a little harder about the important points/order here (might be time to do the exercise where you take the results and boil them back down to short bullet points, in order to figure out whether you've said the important stuff, in the right order).
 Maybe lead by saying that all platforms give similar answers - to the extent that they are fitting the same statistical model (this might even be emphasized/explained above, in Methods).
@@
 Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases.
-\mli{go back and define coverage and how its used}
 

### 2018-11-06 phyloglmm_ms.tex
temp pushing, glmmTMB fits are missing

@@
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
-Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic} , and recently, a demend of integration of evolutionary knowledge in studying ecological applications \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}
+Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic} , and recently, a demend of integration of evolutionary knowledge in studying ecological applications such as biodiversity conservation, effects of climate change, and many more \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
 %such as ecological conser, the effects of climate change  \citep{santamaria2012evolution}. \mli{cite more here}.
@@
 \mli{move example to the supp}
-% For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix,
-% 
-% \begin{align}
-% \bX = \begin{bmatrix}
-% 1 & t_1  \\ 
-% 1 & t_2  \\ 
-% 1 & t_3 
-% \end{bmatrix} 
-% \end{align}
-% then the term-wise phylogenetic random effects model matrix is,
-% 
-% \begin{align}
-% \bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
-% \Bigg[
-% \Bigg(
-% \begin{bmatrix}
-% \ell_{adj1} & \ell_{adj1}  & 0 \\
-% \ell_{adj2} &  0  & 0 \\
-% 0  &  \ell_{adj3} & 0 \\
-% 0 & 0 &  \ell_{adj4} 
-% \end{bmatrix}
-% \begin{bmatrix}
-% 1 & 0  & 0 \\
-% 0 & 1  & 0 \\
-% 0 & 0  & 1  
-% \end{bmatrix}
-% \Bigg)
-% \ast
-% \begin{bmatrix}
-% 1   & 1   & 1  \\ 
-% t_1 & t_2 & t_3
-% \end{bmatrix} 
-% \Bigg]^T
-% \\
-% = \begin{bmatrix}
-% \ell_{adj1} & \ell_{adj1}t_1 & \ell_{adj2} & \ell_{adj2}t_1 & 0 & 0 & 0 & 0 \\
-% \ell_{adj1} & \ell_{adj1}t_2 & 0 & 0 & \ell_{adj3} & \ell_{adj3}t_2 & 0 & 0 \\
-% 0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
-% \end{bmatrix}
-% \end{align}
+For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix,
+
+\begin{align}
+\bX = \begin{bmatrix}
+1 & t_1  \\ 
+1 & t_2  \\ 
+1 & t_3 
+\end{bmatrix} 
+\end{align}
+then the term-wise phylogenetic random effects model matrix is,
+
+\begin{align}
+\bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
+\Bigg[
+\Bigg(
+\begin{bmatrix}
+\ell_{adj1} & \ell_{adj1}  & 0 \\
+\ell_{adj2} &  0  & 0 \\
+0  &  \ell_{adj3} & 0 \\
+0 & 0 &  \ell_{adj4} 
+\end{bmatrix}
+\begin{bmatrix}
+1 & 0  & 0 \\
+0 & 1  & 0 \\
+0 & 0  & 1  
+\end{bmatrix}
+\Bigg)
+\ast
+\begin{bmatrix}
+1   & 1   & 1  \\ 
+t_1 & t_2 & t_3
+\end{bmatrix} 
+\Bigg]^T
+\\
+= \begin{bmatrix}
+\ell_{adj1} & \ell_{adj1}t_1 & \ell_{adj2} & \ell_{adj2}t_1 & 0 & 0 & 0 & 0 \\
+\ell_{adj1} & \ell_{adj1}t_2 & 0 & 0 & \ell_{adj3} & \ell_{adj3}t_2 & 0 & 0 \\
+0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
+\end{bmatrix}
+\end{align}
 
@@
 \subsection{Simulation and evaluations}
-We simulated 100 phylogenetic trees for each sample size (25,50,100) and incorporated with simulation parameters (illustrated in figure 2 and 5) to simulate responses via the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
+We simulated 100 phylogenetic trees for each sample size (25,50,100, 500) and incorporated with simulation parameters (illustrated in figure 2 and 5) to simulate responses via the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
 All model variants via different platforms were used to fit each realization.
-Tables \ref{model_variant} shows the parameters that are estimable for each model/platform. 
-We combined two convergence criteria to assess convergence for the main parameters (R0, Peff, Prep): we required a value of the
-Gelman and Rubin statistic and an effective sample size (ESS) greater than 400 for each replication. For
-each replication we sample four chains starting with 4000 iterations; w
+Tables \ref{model_variant} shows the parameters that are estimable for each platform. 
+All model must pass its respective convergence test before evaluating goodness of fit. 
+For bayesian GLMM, we combined two convergence criteria to assess convergence: we required a value of the
+Gelman and Rubin statistic and an effective sample size (ESS) greater than 1000 for each replication. 
+For each replication we sample two chains starting with 10000 iterations.
+
+We used coverage to assess model fit.
+Coverage refers to the frequency with which the computed confidence intervals include the true values of
+parameters.
+We used 90\% (Wald and profile) confidence intervals and quantile-based intervals (for bayesian GLMM)
+to evaluate coverage.
+We also compare computational speed between different platforms.
 
@@
 
-The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except for random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$).
-Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases.
+The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except the random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$).
+Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases for \pkg{lme4} but not others.
 
 In general, models that cannot match the simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
-PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation.
-$\beta_0$ estimates slightly undercover the nominal coverage (90\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
-PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept as well as large RMSE and overcovering for $\beta_0$, and bad estimates for $\beta_1$.
-\mli{double check this, what happens when X is also phylogenetically correlated?}
+PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation of the intercept ($\beta_0$) estimates slightly undercover the nominal coverage (90\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
+PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept and overcovering for $\beta_0$, and bad estimates for $\beta_1$.
+\mli{refer back to Schielzeth and Forstmeier}
 
@@
 
-Multiple site fits on the other hand are a lot more similar across platforms (as there are less deviates between the fitted and simulation model) compared to the single site fits.
+Multiple site fits on the other hand, are much more similar across platforms (as the fitted and simulation models are more similar).
 Similar to the single site fits, phyloglmm which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
@@
 For example, the median time for \pkg{lme4} to fit 50/100 species model takes $~$ 50/100 seconds versus $~$ 200/2000 seconds for \pkg{pez} and \pkg{phyr}. 
-Lastly, it takes $~$ 2000 seconds for lme4 to fit 1000 species model and it was not practical for \pkg{pez} and \pkg{phyr} to fit 1000 species models where the computational speed did not scale linearly with sample size 25/50/100.
+Lastly, it takes $\sim$ 2000 seconds for lme4 to fit 1000 species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 1000 species models where the computational speed did not scale linearly with sample size 25/50/100.
 \mli{put a lower bound to give up}
@@
 We used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \cite{li2017canfun}. 
-We have obtained identical results for fix and random effect estimates and approximately 120 times faster compared to \pkg{pez}. 
+We have obtained identical results for fix and random effect estimates, our code is approximately 120 times faster than \pkg{pez}. 
 Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.

### 2018-11-06 collect.R

@@ -113,6 +113,7 @@ brmsss_results <- function(tt){
     , B1 = numeric(200)
     , model = numeric(200)
     , time = numeric(200)
+    , convergence = numeric(200)
   )
   for(i in 1:length(tt)){
     brms_obj <- readRDS(paste(brms_path,tt[i],sep=""))

### 2018-11-06 fit_brms.R

@@ -29,6 +29,7 @@ brms_time1 <- proc.time()
 			, iter = stan_nitt
 			, chains = 2
 		)
+
 brms_time2 <- proc.time()
 
 print(summary(brms_fit))
@@ -37,6 +38,6 @@ brms_time <- brms_time2 - brms_time1
 print(summary(brms_fit))
 
 brms_list <- list(brms_fit,brms_time)
-saveRDS(brms_list,file=paste("datadir/brms",numsite,size,seed,"rds",sep="."))
+saveRDS(brms_list,file=paste("datadir/brms",numsite,size,tree_seed,"rds",sep="."))
 
 #rdnosave()

### 2018-11-06 plot.R

@@ -4,39 +4,53 @@ library(dplyr)
 library(ggplot2)
 library(tidyr)
 library(grid)
-theme_set(theme_bw())
+# theme_set(theme_bw())
 zmargin <- theme(panel.margin=grid::unit(0,"lines"))
 
 data_list <- readRDS("./datadir/collect.RDS")
-ssdat_raw <- data_list[[1]]
+ssdat_raw <- rbind(data_list[[1]],brmsss_data)
 msdat_raw <- data_list[[2]]
+
+tree_seed = 1
+source("parameters.R")
+
+sspar_df <- data.frame(
+  sdtype = c("resid", "phylo_int", "phylo_X", "phylo_cor")
+  , y_int = c(sd.resid, physd.B0, physd.B1, phyrho.B01)
+)
 
 ssdat <- (ssdat_raw
 	%>% separate(model,c("platform", "sites", "size", "seed","saveformat"), "[.]")
 	%>% select(time, platform, size, seed, resid, phylo_int, phylo_X, phylo_cor, B0, B1)
 	%>% gather(key = sdtype, value = sd, -c(platform, size, seed, time, B0, B1))
 	%>% mutate(size = factor(size
-	      , levels=c("small","med","large","xlarge"), labels=c("25","50","100","1000")
+	      , levels=c("small","med","large","xlarge"), labels=c("25","50","100","500")
 			)
 			, platform = factor(platform, levels=c("gls", "phylolm","lme4", "brms"))
 		)
-	%>% filter(platform != "brms")
+	%>% left_join(.,par_df)
+	# %>% filter(platform != "brms")
 )
 
 
 gg_ss <- (ggplot(data=ssdat, aes(x=size, y=sd, fill=platform))
   + facet_wrap(~sdtype, scale="free_y")
   + geom_violin(position=position_dodge(width=0.2),alpha=0.4)
-  + geom_hline(aes(yintercept = yint))
-  + ggtitle("Single site")
+  + geom_hline(aes(yintercept = y_int))
+  # + scale_fill_manual(values=c("Dark Red","Dark Blue","Dark Green","Purple"))
+  + scale_fill_brewer(palette = "Dark2")
+  # + ggtitle("Single site")
+  + theme_bw()
 )
 
 print(gg_ss)
 
 gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, fill=platform))
[…]

### 2018-11-06 phyloglmm_ms.tex
updated figures

@@
 \begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=0.8,page=1]{./figure/ssplot.pdf}
   \caption{New}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.8,page=2]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=0.8,page=1]{./figure/sstime.pdf}
   \caption{New}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.8,page=3]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=0.8,page=1]{./figure/sscoverage.pdf}
   \caption{New}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.8,page=4]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=0.8,page=1]{./figure/msplot.pdf}
   \caption{new}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.8,page=6]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=0.8,page=1]{./figure/mstime.pdf}
   \caption{new}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.8,page=7]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=0.8,page=1]{./figure/mscoverage.pdf}
   \caption{new}

### 2018-11-06 phyloglmm_ms.tex
fixing up discussion

@@
 The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except the random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$).
-Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases for \pkg{lme4} but not others.
+Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not others.
 
 In general, models that cannot match the simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
-PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation of the intercept ($\beta_0$) estimates slightly undercover the nominal coverage (90\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
-PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept and overcovering for $\beta_0$, and bad estimates for $\beta_1$.
-\mli{refer back to Schielzeth and Forstmeier}
+PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation of the intercept ($\beta_0$) estimates slightly undercover the nominal coverage (95\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
+PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept and overcovering for $\beta_0$, and undercovering for $\beta_1$.
 
@@
 Multiple site fits on the other hand, are much more similar across platforms (as the fitted and simulation models are more similar).
-Similar to the single site fits, phyloglmm which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
-\mli{either keep it or take it out, because  will affect other parameters}
-The lack of correlation in \pkg{phyr}'s statistical models does not appear to have a large effect on its ability to estimate other parameters, whereas \pkg{pez} underestimates (lower estimates) many parameters.
-\bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate?
+Similar to the single site fits, \pkg{lme4} and \pkg{glmmTMB} which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
+The lack of correlations in \pkg{pez} and \pkg{phyr} statistical models does not appear to have a large effect on its ability to estimate other parameters, whereas \pkg{pez} underestimates (lower estimates) many parameters.
+% \bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate?
 Are these singular models?
@@
 (If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
-\mli{save it for random slope discussion}
+% \mli{save it for random slope discussion}
 
-However, despite the similarities of parameter estimates across platforms/methods for the multiple sites simulations fits, the efficiency (measured by computational speed) are different across playforms/methods and sample size.
+However, despite the similarities of parameter estimates across platforms for the multiple sites simulation fits, the efficiency (measured by computational speed) are different across platforms and sample size.
 For example, the median time for \pkg{lme4} to fit 50/100 species model takes $~$ 50/100 seconds versus $~$ 200/2000 seconds for \pkg{pez} and \pkg{phyr}. 
-Lastly, it takes $\sim$ 2000 seconds for lme4 to fit 1000 species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 1000 species models where the computational speed did not scale linearly with sample size 25/50/100.
-\mli{put a lower bound to give up}
+Lastly, it takes $\sim$ 1000 seconds for lme4 to fit 500 species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500 species for our study, where the computational speed did not scale linearly with sample size 25/50/100.
+\pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
 
-\bmb{Can you put a lower bound on \pkg{pez}'s computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
-  or (easier) do a brute-force series of increasing sample size and
-  compute a log-log regression of time vs sample size to find the
-  approximate power?}
+% \bmb{Can you put a lower bound on \pkg{pez}'s computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
+%   or (easier) do a brute-force series of increasing sample size and
+%   compute a log-log regression of time vs sample size to find the
+%   approximate power?}
 
@@
 The evolutionary model simulations presented in this paper are an oversimplification of real evolutionary processes in nature, but it is analogous to how researchers fit models to study complicated systems (as it is impossible to create a model to "match" the complexities of nature).
-We have fitted models using different platforms/methods that best match the simulation models. 
-Models that cannot match the simulation (full) model are expected to underperform; it is important to understand the limitations and performance of these simpler models/assumptions that are restricted by the commonly used methods available.
-Using models that include some form of phylogenetic variation and residual variation is necessary to \mli{correctly/robustly?} detect evolutionary process.
+We have fitted models using different platforms that best match the simulation models. 
+Models that cannot match the simulation (full) model are expected to underperform; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods available.
+Using models that include some form of phylogenetic variation and residual variation is necessary to detect evolutionary process.
+Here are a few points researchers should think about when fitting phylogenetic regression models.
 
@@
 Process and observation errors are highly confounded in phylogenetic regression if it is not handled correctly. 
-The effects of models that include both process and residual variations (i.e.
-PGLMM, PGLM) instead of the less realistic model that assumes variation due to evolution and zero observation (PGLS) was large.
-% In particular, models that did not include residual variation tended to over-estimate the amount of variability in the evolutionary process. 
-Thus, for single measurement per species model, any method that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e.
+For single measurement per species model, any method that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e.
 Pagel's $\lambda$, Blomberg's $K$).
-However, if multiple measurements/observations are considered/allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation.
-
-\mli{this is specifically targeting boettiger's critque on Pagel's lambda.
-https://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html
-The difference between the two figures is what it thinks the "tip" variation is.
-For the second plot, if we treat the sister taxas as "multiple" observation with some observation error on it, then you will get back to the first plot.}
-
-\bmb{again, we/you need to think carefully about what these coarse measures are good for; just avoiding mis-estimating the strength of trait/trait or trait/environment parameters?
-Is it worth doing some sims on that, or do we basically believe if the total amount of phylogenetic signal is correctly assessed, that we get the right answer for the trait effect?
-Is \cite{schielzeth2009conclusions} relevant/a useful analogy?
-(You cite it below, but there are two separate questions: (1) whether including random slopes gives us more insight into evol processes; (2) whether omitting random slopes gives us bad answers even when we're not interested in them.) Worth citing/discussing \cite{uyeda2018rethinking} (as an example of even deeper analysis of evolutionary processes, beyond BM etc.)?}
+If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation.
+However, one can summarise multiple measurements to a single measurement (for example, mean or median) per species to avoid confounding residual and tip variations.
 
@@
 In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+Neglicting random slopes can lead to erroneous fixed effect estimates \citep{schielzeth2008conclusions} (parameters of interest) as shown in the simulations above. 
 When analyzing relationships among species between traits and environmental variables, it is entirely plausible that change in effect of predictor variables have on traits are changing in a phylogenetically correlated way.
 In fact, it is easier to interpret random slopes in the evolutionary process context (i.e, are size of different species phylogenetically correlated, or growth rate through time of different species phylogenetically correlated?)
-\mli{I like to think of it in a longitudinal setting of observing plant length/height.
-At time zero, they are all still seeds, one can imagine the growth rate are different plant species grow in a phylogenetically correlated way (similar species have similar growth rates).}
-It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \cite{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
+It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
 Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
-and Are we making a mistake by ignoring random slopes \cite{schielzeth2008conclusions}?
-However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
+and Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}?
+% However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
 
@@
 More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis.
-This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as \pkg{MCMCglmm} \cite{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
+This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as \pkg{MCMCglmm} \citep{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
 However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e.
 What should we do when we don't have enough data?
-Should we use a complex model and overfit \cite{barr2013random}, the right balance of complexity and data \cite{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
+Should we use a complex model and overfit \citep{barr2013random}, the right balance of complexity and data \citep{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
 More importantly, this implementation in \pkg{lme4} allows users to fit phylogenetic mixed models to the fullest ( that other frequentist platforms lack of (i.e.

### 2018-11-19 phyloglmm_ms.tex
working on intro ...

@@
 
-Ever-increasing data collection capabilities (such as genomic sequencing, telemetry studies of animal behaviour, or remote sensing of environmental variables), in combination with large-scale synthesis of existing data bases on species occurrence and phenotypic traits, are making large volumes of biological data available over an ever-wider taxonomic range.
-In response, researchers are studying more complicated systems and fitting more models to larger data sets describing species occurrence and traits.
-One particular area of growth is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting}. 
-Comparative analyses explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; researchers may be interested in controlling for the statistical effects of phylogenetic correlation, quantifying the evolutionary signals that underly traits or both.
-Unlike ordinary regression, where all of the variables of interest (for example, species trait and environmental factors) are directly observable, in phylogenetic regression, the evolutionary relationship between species are unobservable and assumed to evolve according to some underlying (latent) evolutionary process \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
+Ever-increasing data collection capabilities (such as genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic data bases of species occurrence and phenotypic traits, are making large volumes of biological data available over an ever-wider taxonomic range.
+Researchers are exploiting these data to fit complex models describing species occurrence and traits.
+In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is
+becoming ever more popular \bmb{might be interesting to some bibliometry, although absolutely not necessary \ldots}
+PCAs explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; they can be used to control for the statistical effects of phylogenetic correlation, to quantify the phylogenetic signals in trait distributions, or both.
+Unlike ordinary regression, where all of the variables of interest (for example, species trait and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
@@
 \bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
-\be & \sim \textrm{MVN}(0, \sigma^2_{residual} \label{eq:glmm3})
+\be & \sim \textrm{MVN}(0, \sigma^2_{residual} \label{eq:glmm4})
 \end{align}

### 2018-11-27 phyloglmm_ms.tex
bmb tweaks

@@
 In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is
-becoming ever more popular \bmb{might be interesting to some bibliometry, although absolutely not necessary \ldots}
+becoming ever more popular \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
+how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature?
+\ldots}
 PCAs explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; they can be used to control for the statistical effects of phylogenetic correlation, to quantify the phylogenetic signals in trait distributions, or both.
-Unlike ordinary regression, where all of the variables of interest (for example, species trait and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
+Unlike ordinary regression, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
-Over the last few decades, biologists have incorporated phylogenetic relationships to multi-species modelling \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic} , and recently, a demend of integration of evolutionary knowledge in studying ecological applications such as biodiversity conservation, effects of climate change, and many more \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
-%such as ecological conser, the effects of climate change  \citep{santamaria2012evolution}. \mli{cite more here}.
-While there exist a range of statistical tools for comparative analyses, many of these tools are either inflexible, or too computationally expensive to analyze large volumes of data.
-In such cases, researchers may try to find ways to simplify their analyses (i.e.
-controlling for individual-varation of species and neglecting phylogenetic correlation between species \citep{bunnefeld2012island}, comparing species family \citep{tella1999habitat}, neglecting within-species variation \citep{ord2010adaptation}, and etc).
+Ecologists have incorporated phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently, there is increased demand for integrating evolutionary knowledge to address ecological applications such as biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
+While there are a wide range tools for comparative analyses, they may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
+In such cases, researchers must find ways to simplify their analyses (i.e.
+controlling for individual variation of species \bmb{??} and neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness/treating taxon as a strictly hierarchical description \citep{tella1999habitat}; neglecting within-species variation \citep{ord2010adaptation}; etc).
 %% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
@@
 In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
-% \bmb{nice!}
+
+\bmb{somewhere, cite/frame our approach in terms of the recent ``basis functions in ecology'' paper}
 
@@
 
-\bmb{I like the commented-out table here (maybe in supp material? or maybe useful enough that it's worth including in the main text?) although it could take a lot of fussing to get it right}
-
 % Table \mli{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
@@
 
+\bmb{explain: classical scenario for PIC is when we have species-level data (i.e., species averages or ``typical'' values).
+What do we do when we have multiple observations per species?
+Why can't we just average?
+(cf.
+\cite{murtaugh2007simplicity}) with a nested, balanced design, averaging by group and doing 1-way ANOVA gets the same answer for fixed effects as LMMs).
+\textbf{answers}: (1) unbalanced data (can handle by weighting by $n$, equivalent to assuming homoscedasticity and using inverse-variance weights); (2) non-Gaussian responses (averaging doesn't make sense) [can we use offsets to handle this case?
+don't know]; (3) maybe for equivalent of randomized block designs?
+(4) when within-species variance is actually of interest}
 Process and observation errors are highly confounded in phylogenetic regression if it is not handled correctly. 

### 2018-11-27 lme4_profile_example.R

@@ -11,5 +11,11 @@ print(lme4obj[[1]])
 lme4_profile <- confint(lme4obj[[1]]
   , method = "profile"
   , parm = c("(Intercept)","X")
-)
+    )
 
+pp <- profile(lme4obj[[1]],
+              which = c("(Intercept)","X"),
+              verbose=20)
+confint(pp)
+
+sessionInfo()

### 2018-12-11 phyloglmm_ms.tex
temp pushing, still lots to be done

@@
 
-Ever-increasing data collection capabilities (such as genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic data bases of species occurrence and phenotypic traits, are making large volumes of biological data available over an ever-wider taxonomic range.
-Researchers are exploiting these data to fit complex models describing species occurrence and traits.
+Ever-increasing data collection capabilities (e.g.
+genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic data bases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
+Researchers use these data to fit complex models describing species occurrence and traits.
 In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is
-becoming ever more popular \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
+becoming increasingly popular % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
 how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature?
 \ldots}
-PCAs explore the relationships among species traits (including occurrence within communities) while taking the underlying/latent evolutionary relationships of the species into account; they can be used to control for the statistical effects of phylogenetic correlation, to quantify the phylogenetic signals in trait distributions, or both.
-Unlike ordinary regression, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
+PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals in trait distributions, or both.
+Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
-Ecologists have incorporated phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently, there is increased demand for integrating evolutionary knowledge to address ecological applications such as biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
-While there are a wide range tools for comparative analyses, they may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
-In such cases, researchers must find ways to simplify their analyses (i.e.
-controlling for individual variation of species \bmb{??} and neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness/treating taxon as a strictly hierarchical description \citep{tella1999habitat}; neglecting within-species variation \citep{ord2010adaptation}; etc).
+Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently, they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
+While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
+In such cases, researchers must find ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
 %% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
@@
 % but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
-In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
-
-\bmb{somewhere, cite/frame our approach in terms of the recent ``basis functions in ecology'' paper}
+In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models through random effects. 
+This method allow researchers to easily incorporate various complexities without sacrificing speed.
 
@@
 % Phylogenetic signal : the magnitude of phylogenetic correlation
-The first challenge is to identify and segregate phylogenetic signals \citep{blomberg2003testing} from other sources of variation that are present in the data.  
+The first challenge is to identify and separate phylogenetic signals from other sources of variation that are present in the data \citep{blomberg2003testing}.  
 % \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
@@
 Maybe replace with something talking more specifically about resolving tip variation?}
-The classic phylogenetic regression assumes that phylogenetic correlation among the residuals from a regression between two species-level traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the residuals are normally distributed and observed without any additional error or within-species variation, we can use Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies, nicolakakis2000forebrain}.
-More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different response distributions and allowing extra parameters to account for additional complexities in the evolutionary process.
-In particular, they can partition residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
-However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/intercept level variation) and residual variation is associated with observation level variation within-species.
+The classic phylogenetic regression uses a statistical model that is equivalent to assuming that phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the trait that is the dependent variable evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model.
+If the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
+More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for additional complexities in the evolutionary process.
+In particular, they can partition residual variation into (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
+However, each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call intercept-level variation) while the among species residual error is associated with among observations variation within each species.
 %\bmb{This is a good point, well expressed.
@@
 %\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
-Despite having multiple observations per tip taxon (species), a typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
+Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such observations to partition variablility more finely.
+Indeed, many existing methods cannot accommodate multiple observations per species, instead requiring users to collapse the data by taking species averages.
+%Despite having multiple observations per tip taxon (species), a typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
 % \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
@@
 %% FIXME
-The second challenge is flexibly model phylogenetic relations.
+The second challenge is flexibly modeling phylogenetic relations.
 While it is reasonable to examine evolutionary relationships between species with their response, it is plausible the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated.
@@
 Several recent studies have incorporated phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
-For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach.
-The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
+For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach.
+The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
 Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.
@@
 \hline
-GLM (Simple)  & No Residual & Single Observation & GLS, PIC \\
+GLM (Simple)  & Correlated residual & Single Observation & GLS, PIC \\
 \hline

### 2019-01-28 phyloglmm_ms.tex
stopping for now

@@
 \usepackage{tabularx}
+\usepackage{multirow}
 % use \xspace to allow for space after a macro as necessary
@@
 \usepackage[sort]{natbib}
+\usepackage{blkarray}
 \usepackage{amsmath}
@@
 Researchers use these data to fit complex models describing species occurrence and traits.
-In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is
-becoming increasingly popular % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
+In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is becoming increasingly popular.
+% \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
 how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature?
 \ldots}
-PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals in trait distributions, or both.
+PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
 Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
@@
 
-%% FIXME
+%% FIXME: second sentence
 The second challenge is flexibly modeling phylogenetic relations.
@@
 
-\begin{tabularx}{\textwidth}{|X|X|X|X|}
+%% FIXME: chart looks ugly
+\begin{table}[]
+\begin{tabular}{|l|l|l|l|}
 \hline
-Model & Method & Data & Platform \\
-\hline
-\hline
-GLM (Simple)  & Correlated residual & Single Observation & GLS, PIC \\
-\hline
-              & Residual    & Single Observation & Pagel's $\lambda$ \\
-              & + phylo intercept &                    & Blomberg's $K$ \\ 
-              &             &                    & via phylolm, GLS \\
-\hline
-GLMM (Complex)& Random effects & Single Observation & \pkg{pez} \\ 
-              &                & Balanced Design &  \\
-\hline
-              &                & Unrestricted  & \pkg{lme4}, \pkg{glmmTMB} \\
-\hline
-Bayesian GLMM &                & Balanced Design & \pkg{MCMCglmm} \\ 
-\hline        
-              &                & Unrestricted   & \pkg{brms} \\
-\hline
-\end{tabularx}
+\textbf{Model}                                                                                    & \textbf{Method}                                                       & \textbf{Data}                                                                & \textbf{Platform}                                                                              \\ \hline
+\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)\end{tabular}}        & Correlated residual                                                   & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular}                                   \\ \cline{2-4}
+                                                                                                  & \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
+\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single Observation\\ Balanced Design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4}
+                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}                                                                                  \\ \hline
+\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced Design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4}
+                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
+\end{tabular}
+\end{table}
 
-However, in this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches.
-Our new formulation is more flexible; it can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
-We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing methods/platforms (i.e.
-\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
-We also expect that it will be faster, because \pkg{lme4} and \pkg{glmmTMB} are optimized in performance using efficient computational linear algebra techniques. 
-In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
-We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.
+%% FIXME: second sentence
+In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
+Our new formulation can be incorporated in any framework that allows for random effect without the need to implement special correlation structures.
+%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
+We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
+\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
+As well as being flexible, we also expect that our approach will be fast, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
+%% FIXME: move to methods/simulations
+% In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
+% We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.
 
@@
 
-The standard phylogenetic regression takes the form
+\subsection{Phylogenetic regression}
+
+The standard phylogenetic regression can be formulated as
 
@@
 \end{align}
-where $\bX$ is an $n \times m$ model matrix containing $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector which is assumed to be multivariate normally distributed with mean $\bmu$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation matrix.
+where $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector which is assumed to be multivariate normally distributed with mean $\bmu$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation matrix.
 The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.
@@
 %\mli{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
-An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
-The typical linear mixed model has the form:
+
+\subsection{Phylogenetic generalized linear mixed model}
+Alternatively, one can use the generalized linear mixed effects modelling (GLMM) framework \citep{lynch1991methods}.
+The classic generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorportate more than one kind of variability through random effects and for non-gaussian responses.
+The typical GLMM has the form:
 \begin{align}
-\bY & \sim \textrm{Distr}(\bmu) \label{eq:glmm1} \\
-g(\bmu) & = \bX \bbeta + \bZ \bb + \be \label{eq:glmm2} \\
+\bY & \sim \mathbf{f}(\bmu,\bSigma) \label{eq:glmm1} \\
+\bmu & = \bX \bbeta + \bZ \bb  \label{eq:glmm2} \\
 \bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
-\be & \sim \textrm{MVN}(0, \sigma^2_{residual} \label{eq:glmm4})
 \end{align}
-where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\be$ (sometimes referred as "R" effect) represents the residuals.
-Analogously, the phylogenetic regression given by \ref{eq:gls1} can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^2_{residual} = 0$.
+where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G-side" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\be$ (sometimes referred as "R" effect) represents the residuals.
+%% FIXME: see basis-fn paper
+Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^2_{residual} = 0$.
+
+There are serveral advantages in modelling phylogenetic correlated traits in the mixed model framework.
+First, random intercepts can allow the response trait to vary across groups/levels within grouping predictors (e.g. patch, site, experiment, and etc). 
+Second, it can also allow the response (trait or distribution) to vary according to the phylogenetic relationships across species (i.e.
+similar species will have similiar responses), and between species (tip variation).
+The third type of variation that is often neglected or rarely considered are random slopes.
+Random slopes can allow fixed effects (size of the effect in continuous predictor variables) to vary for each species. 
+Analogous to modelling the repsonse dependency via phylogenetic relationship among species, phylogenetic random slopes can allow fixed effects to account for the phylogenetic relations (i.e.
+similar species will have similar size of the effects on the response than distinct pairs).
 
@@
 
-\subsection{Reformulating phylogenetic covariance matrix}
+\subsection{Reformulating the phylogenetic covariance matrix}
 % The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
@@
 An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
-In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk with daughter species evolving independently after speciation.
-In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
-Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $\bS$, equivalently to imposing a correlation $\bC$.
-For example, as shown in figure \ref{fig:tree}, the corresponding $\bS$ is in the form:
+In particular, suppose that the evolutionary process is a Brownian motion, which means that evolution of a continuous trait evolve independently, following a standard Brownian motion, along each branch of the phylogeny.
+In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
+Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$, is equivalent to imposing a correlation $\bC$.
+For example, for the phylogeny in figure \ref{fig:tree}, the corresponding $\bS$ takes the form:
 
+%% FIXME: need top/side labels
 \begin{center}
@@
   \includegraphics[scale=0.8,page=1]{./git_push/tree.pdf}
-  \caption{\bmb{caption ??}}
+  \caption{Three species phylogenetic tree.}
 \label{fig:tree}
@@
 
-\begin{align*}
-\bS = \begin{bmatrix}
-\ell_1 \e_1 & \ell_2 \e_2 & 0 & 0 \\
-\ell_1 \e_1 &  0 & \ell_3 \e_3 & 0 \\
-0  &  0 & 0  & \ell_4 \e_4
-\end{bmatrix}
-\end{align*}
-The error value corresponding to tip 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i$ is the square root of the branch length $L_i$ in figure \ref{fig:tree} and the $\ell_i$ are independent, homoscedastic Normal variates with zero mean and $\sigma^2$ variance (i.e.
-variance for tip 1 is $(\ell_1 \e_1 + \ell_2 \e_2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$, linear in branch length)
+\[
+\begin{blockarray}{ccccc}
+ & L_1 & L_2 & L_3 & L_4  \\
+\begin{block}{c(cccc)}
+  t_1 & \ell_1 \e_1 & \ell_2 \e_2 & 0           & 0 \\
+  t_2 & \ell_1 \e_1 &  0          & \ell_3 \e_3 & 0 \\
+  t_3 & 0           &  0          & 0           & \ell_4 \e_4 \\
+\end{block}
+\end{blockarray}
+ \]
+
+%% FIXME: double check this
+The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree} and the $\ell_i$ are independent, homoscedastic Normal variates with zero mean and $\sigma^2$ variance (i.e.
+the variance for species 1 is $(\ell_1 \e_1 + \ell_2 \e_2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$, linear in branch length).
+%% FIXME: see BB's cmt
 
@@
 
-The $\bS$ matrix is the matrix product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices, and a branch length vector $\bell$.
+The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and a branch length vector $\bell$.
 
@@
 \ell_4 
-\end{bmatrix}
+\end{bmatrix} .
 \label{eq:S}
-\end{align}.
+\end{align}
 $\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
@@
 
-The random-effect model matrix $\bZ$ can be decomposed into term-wise model matrix $\bZ_{i}$ as described in \citet{bates2015fitting}.
-Thus, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is
+The random-effect model matrix $\bZ$ can be decomposed into term-wise model matrices $\bZ_{i}$ as described in \citet{bates2015fitting}.
+Analogous to the procedure described in \citet{bates2015fitting}, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is
 
 \begin{equation}
-\bZ^{C}_{i} = (\bS^{T}_{i}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
+\bZ^{C}_{i} = (\bS^{T}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
 \end{equation}
-where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{i}$ is the $m_{i} \times b_{i}$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m_i$ indicator matrix of grouping factors; $X_{i}$ is the $n \times p_{i}$ raw random effects model matrix.
+where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; and $\bX_{i}$ is the $n \times p_{i}$ raw random effects model matrix.
 
-\mli{move example to the supp}
-For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix,
+For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix corresponding to a random-slope model,
 
@@
 \bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
-\Bigg[
-\Bigg(
+\left[
+\left(
 \begin{bmatrix}
@@
 \end{bmatrix}
-\Bigg)
+\right)
 \ast
@@
 \end{bmatrix} 
-\Bigg]^T
+\right]^T
 \\
@@
 0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
-\end{bmatrix}
+\end{bmatrix}.
 \end{align}
@@
 
-We generated test data based on the simple mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
-For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
-For single-site model, the full simulation model is the following:
+We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
+For simplicity, the response variable $\bY$ is conditional normally distributed (i.e. gaussian distribution with an idenitiy link function) , which is a linear mixed effect model. 
+For the first set of simulation, we simulate one observation per species.
+% For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
+The full simulation model is the following:
 \begin{align}
 \bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + e_{residual}  \label{eq:ss_sim}\\\
-(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
+(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
 \sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
@@
 \end{bmatrix} 
-\Bigg) \\ 
+\right) \\ 
 e_{residual} & \sim \textrm{MVN} ( 0 , \Sigma) .
 \end{align}
-The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance, phylogenetic random slope variance and covariance between phylogenetic random slope and intercept) and observation/residual variance.
+The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance, phylogenetic random slope variance and covariance between phylogenetic random slope and intercept) and residual variance.
+The covariance between phylognetic random intercept and slope measures the correlation of phylogenetic dependency variability of the size in effect and outcome.
+\mli{con't here!}
 Predictor and intercept level random effect of species are not applicable in this simulation setting with single observation per species because it will be confounded with the observational level variation (residuals).
 
-We extend the simulation model by adding multiple sites where each site has one observation per species. The multiple-site full model is the following: 
+We extend the simulation model by adding multiple groups where each group has one observation per species. The multiple-group full model is the following: 
 \begin{align}
-\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site} + e_{residual} \label{eq:ms_sim}\\
-(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
+\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{group}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:group} + e_{residual} \label{eq:ms_sim}\\
+(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
 \sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
@@
 \end{bmatrix}
-\Bigg) \\
-(b_{sp_{int}}, b_{sp_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
+\right) \\
+(b_{sp_{int}}, b_{sp_{slope}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
 \sigma^2_{sp_{int}} & \sigma_{sp_{int},sp_{slope}} \\ 
@@
 \end{bmatrix}
-\Bigg) \\
+\right) \\
 b_{site} & \sim \textrm{MVN} ( 0 , \Sigma_{site}) \\

### 2019-02-05 phyloglmm_ms.tex
everything before simulation and evaluation is ready for BMB to read

@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 
+\subsection*{Challenges in modelling evolutionary process}
+
 There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
 % Phylogenetic signal : the magnitude of phylogenetic correlation
-The first challenge is to identify and separate phylogenetic signals from other sources of variation that are present in the data \citep{blomberg2003testing}.  
+The first challenge is to identify and separate phylogenetic signals from other sources of variation present in the data \citep{blomberg2003testing}.  
 % \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
@@
 In particular, they can partition residual variation into (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
-However, each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call intercept-level variation) while the among species residual error is associated with among observations variation within each species.
+However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call intercept-level variation) while the among species residual error is associated with among observations variation within each species.
 %\bmb{This is a good point, well expressed.
@@
 
+\begin{center}
+\begin{figure}[h]
+  \includegraphics[scale=0.8]{./phylo_diagram.png}
+  \caption{Visual categorization of various forms of variation.}
+\label{fig:variation}
+\end{figure}
+\end{center}
+
 %% FIXME: second sentence
-The second challenge is flexibly modeling phylogenetic relations.
-While it is reasonable to examine evolutionary relationships between species with their response, it is plausible the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated.
+The second challenge is flexibly modeling phylogenetic relations across different levels of complexity.
+Classic PCA usually examines the response (trait or distribution) to vary according to the phylogenetic relationship across species, but it is plausible the changes in effect (or levels) of the predictor variables
+may vary according to the phylogenetic relationship across species as well.
 %\bmb{maybe we can focus on random slopes here?
@@
 
-%% FIXME: second sentence
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
-Our new formulation can be incorporated in any framework that allows for random effect without the need to implement special correlation structures.
+Our new formulation can be incorporated in any framework that allows for random effect without the need to implement special correlation structures by incorportating phylogenetic structures as part of the mean model.
 %% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
@@
 
-\section{Materials and Methods}
+\section*{Materials and Methods}
 
@@
 \newcommand{\bb}{{\mathbf b}}
-\newcommand{\be}{{\mathbf e}}
+\newcommand{\besp}{{\boldsymbol \epsilon}}
 \newcommand{\bSigma}{{\boldsymbol \Sigma}}
 
-\subsection{Phylogenetic regression}
+\subsection*{Phylogenetic regression}
 
+We begin by describing the classic PCA in a linear regression setting.
+Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors $\bX$. 
 The standard phylogenetic regression can be formulated as
@@
 
-\subsection{Phylogenetic generalized linear mixed model}
+\subsection*{Phylogenetic generalized linear mixed model}
 Alternatively, one can use the generalized linear mixed effects modelling (GLMM) framework \citep{lynch1991methods}.
-The classic generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorportate more than one kind of variability through random effects and for non-gaussian responses.
+The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorportate more than one kind of variability through random effects and for non-gaussian responses.
 The typical GLMM has the form:
 \begin{align}
-\bY & \sim \mathbf{f}(\bmu,\bSigma) \label{eq:glmm1} \\
+\bY & \sim \mathbf{f}(\bmu) + \besp \label{eq:glmm1} \\
 \bmu & = \bX \bbeta + \bZ \bb  \label{eq:glmm2} \\
 \bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
+\besp & \sim \textrm{N}(0,\sigma^{2}_{\epsilon})
 \end{align}
-where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G-side" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\be$ (sometimes referred as "R" effect) represents the residuals.
-%% FIXME: see basis-fn paper
-Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^2_{residual} = 0$.
+where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G-side" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
+Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the generalized mixed model framework with Gaussian responses by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^{2}_{\epsilon} = 0$.
 
@@
 
-\subsection{Reformulating the phylogenetic covariance matrix}
+\subsection*{Reformulating the phylogenetic covariance matrix}
 % The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
@@
 An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
-In particular, suppose that the evolutionary process is a Brownian motion, which means that evolution of a continuous trait evolve independently, following a standard Brownian motion, along each branch of the phylogeny.
+In particular, suppose that the evolutionary process is a Brownian-motion, which means that evolution of a continuous trait evolve independently, following a standard Brownian-motion, along each branch of the phylogeny.
 In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
@@
 The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree} and the $\ell_i$ are independent, homoscedastic Normal variates with zero mean and $\sigma^2$ variance (i.e.
-the variance for species 1 is $(\ell_1 \e_1 + \ell_2 \e_2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$, linear in branch length).
-%% FIXME: see BB's cmt
+the variance for species 1 is $(\ell_1 \e_1 + \ell_2 \e_2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$).
+The variance of Brownian-motion, is the evolutionary rate, where it is linear in evolution time.
 
@@
 
-The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and a branch length vector $\bell$.
+The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and $\bell$ is a vector of square root branch lengths.
 
@@
 \end{bmatrix} , 
+\qquad
 \bell = \begin{bmatrix}
@@
 \begin{bmatrix}
-\ell_{adj1} & \ell_{adj1}  & 0 \\
-\ell_{adj2} &  0  & 0 \\
-0  &  \ell_{adj3} & 0 \\
-0 & 0 &  \ell_{adj4} 
+\ell_{1} & \ell_{1}  & 0 \\
+\ell_{2} &  0  & 0 \\
+0  &  \ell_{3} & 0 \\
+0 & 0 &  \ell_{4} 
 \end{bmatrix}
@@
 = \begin{bmatrix}
-\ell_{adj1} & \ell_{adj1}t_1 & \ell_{adj2} & \ell_{adj2}t_1 & 0 & 0 & 0 & 0 \\
-\ell_{adj1} & \ell_{adj1}t_2 & 0 & 0 & \ell_{adj3} & \ell_{adj3}t_2 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
+\ell_{1} & \ell_{1}t_1 & \ell_{2} & \ell_{2}t_1 & 0 & 0 & 0 & 0 \\
+\ell_{1} & \ell_{1}t_2 & 0 & 0 & \ell_{3} & \ell_{3}t_2 & 0 & 0 \\
+0 & 0 & 0 & 0 & 0 & 0 & \ell_{4} & \ell_{4}t_3
 \end{bmatrix}.
@@
 
-\subsection{Simulation}
+\subsection*{Simulation}
+\subsubsection*{Single group model}
 
@@
 \begin{align}
-\bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + e_{residual}  \label{eq:ss_sim}\\\
-(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
-\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
+\bY & = (\beta_0 + b_{\mathrm{phy_{int}}}) + (\beta_1 + b_{\mathrm{phy_{slope}}}) X + \besp  \label{eq:ss_sim}\\\
+(b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
+\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
+\sigma_{\mathrm{phy_{int},phy_{slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
 \end{bmatrix} 
 \right) \\ 
-e_{residual} & \sim \textrm{MVN} ( 0 , \Sigma) .
+\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2) .
 \end{align}
@@
 The covariance between phylognetic random intercept and slope measures the correlation of phylogenetic dependency variability of the size in effect and outcome.
-\mli{con't here!}
-Predictor and intercept level random effect of species are not applicable in this simulation setting with single observation per species because it will be confounded with the observational level variation (residuals).
+Predictor and intercept level random effect of species are not applicable in this simulation setting because there is no variation at the species level (i.e. single observation per species).
 
-We extend the simulation model by adding multiple groups where each group has one observation per species. The multiple-group full model is the following: 
+\subsubsection*{Multi-group model}
+
+We extend the simulation model by adding multiple groups where each group has one observation per species. The multi-group full model is the following: 
 \begin{align}
-\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{group}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:group} + e_{residual} \label{eq:ms_sim}\\
-(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
-\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
+\bY & = (\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}) + (\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}) X + b_{\mathrm{sp:group}} + \besp \label{eq:ms_sim}\\
+(b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
+\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
+\sigma_{\mathrm{phy_{int},phy_{slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
 \end{bmatrix}
 \right) \\
-(b_{sp_{int}}, b_{sp_{slope}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\sigma^2_{sp_{int}} & \sigma_{sp_{int},sp_{slope}} \\ 
-\sigma_{sp_{int},sp_{slope}} & \sigma^2_{sp_{slope}}
+(b_{\mathrm{sp_{int}}}, b_{\mathrm{sp_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
+\sigma^2_{\mathrm{sp_{int}}} & \sigma_{\mathrm{sp_{int},sp_{slope}}} \\ 
+\sigma_{\mathrm{sp_{int},sp_{slope}}} & \sigma^2_{\mathrm{sp_{slope}}}
 \end{bmatrix}
 \right) \\
-b_{site} & \sim \textrm{MVN} ( 0 , \Sigma_{site}) \\
-b_{sp:site} & \sim \textrm{MVN} (0, \textrm{Kron}(\bf{I}_{site}, \sigma^2_{phy})) \\
-e_{residual} & \sim \textrm{MVN} ( 0 , \Sigma)
+b_{\mathrm{group}} & \sim \textrm{MVN} ( 0 , \Sigma_{\mathrm{group}}) \\
+b_{\mathrm{sp:group}} & \sim \textrm{MVN} (0, \bf{I}_{\mathrm{group}} \otimes \sigma^2_{phy}) \\
+\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
 \end{align}
-The multiple-site full simulation model has five additional random effects (predictor and intercept level random effect of species variance and their covariance, random intercept of site and random intercept of species-site interaction) compared to the single--site full model.
-Predictor and intercept level random effect of species are now applicable in the multiple-site model because there are multiple (one in every site) observations per species. 
-Random intercept of species--site interactions examines whether the species within a comunity are likely to contain stronger phylogenetic relatedness than expected by chance; which is equivalent to phylogenetic attraction described in \cite{helmus2007separating}.
+where $\otimes$ is the Kronecker product.
 
-\subsection{Platforms}
+The multi-group full simulation model has five additional random effects (predictor and intercept level random effect of species variance and their covariance, random intercept of site and random intercept of species-site interaction) compared to the single--group full model.
+Predictor and intercept level random effect of species are now applicable in the multi-site model setting because there are multiple (one in each group) observations per species, thus there exist some form of variation at the species level.
+Random intercept of species--group interactions describes whether the species within a group are more closely related on average than occurring by chance; which is equivalent to phylogenetic attraction as described in \cite{helmus2007separating}.
 
-Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as part of a linear or generalized linear model. 
+\subsection*{Platforms}
+
+Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as part of a linear or random effects (Z matrix) in generalized linear model.
 We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}.
@@
 % \end{verbatim}
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it is simily fitting a linear model where the covariance structure between species is the expected Brownian motion evolution process (or other evolutionary process) on the tree instead of independent and identically distributed.
-Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) is a slightly more flexible variation of PGLS that can allow additional observational residuals and also non-Gaussian response variables.
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it simply fits a linear model where the covariance structure between species is equivalent to assuming a Brownian-motion evolutionary process (or other evolutionary process) on the tree instead of treating the residual variation as independent and identically distributed.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow observational residuals and as well as non-Gaussian response variables.
 Both packages can model other evolutionary processes and different correlation structures (i.e. Pagel's $\lambda$, Blomberg's $K$ etc.), but we restrict our PGLS fits to the simple BM correlation. 
-Neither PGLS nor PGLM can handle multiple observations or predictor level variation.
-One of the relatively few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle adding additional uncorrelated random slopes and interactions with intercept.
-Lastly, bayesian phylogenetic GLMM using Markov chain monte carlo (MCMC) are avalible and can handle all cases described above. 
-However, MCMC is often computationally expensive for GLMMs compared to platforms using optimization.
-\pkg{MCMCglmm} \citep{hadfield2010general} is the most commonly used bayesian phylogenetic GLMM, but we will use \pkg{brms} instead, which uses a more powerful MCMC technique called Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid} that is more efficient \citep{li2018fitting}.
+Neither PGLS nor PGLM can handle multiple observations within a species or random slopes.
+One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle adding additional uncorrelated random slopes and interactions with intercept.
+Lastly, Bayesian phylogenetic GLMM using Markov chain Monte Carlo (MCMC) are available and can handle all of the cases described above. 
+However, MCMC is often computationally expensive for GLMMs compared to platforms using determinstic optimization.
+\pkg{MCMCglmm} \citep{hadfield2010general} is the most commonly used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid} algorithm \citep{li2018fitting}.
 
 % Table \mli{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
-
-% \begin{table}
-% \begin{center}
-\begin{table}
-\begin{tabularx}{\textwidth}{|c||c|c|c|c|c|c|}
+\begin{table}[]
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline
-Package	& \small{nlme}	& \small{phylolm}	& \small{lme4/glmmTMB} & \small{pez} & \small{brms} & \small{MCMCglmm} \\
-\hline
-\hline
-Single Site & X & X & X & & X & X \\
-\hline
-Phylo intercept & X & X & X &  & X & X \\
-Phylo slope &  &  & X &  & X & X \\
-Phylo correlation &  &  & X &  & X & X \\
-Residual & & X & X &  & X & X \\
-\hline
-\hline
-Multiple Site &  &  & X & X & X & X \\
-\hline
-Phylo intercept &  &  & X & X & X & X \\
-Phylo slope &  &  & X & X & X & X \\
-Phylo correlation &  &  & X &  & X & X \\
-Phylo interaction & &  & X & X & X &  \\
-Species intercept & &  & X & X & X & X \\
-Species slope & & & X & X & X & X \\
-Species correlation & &  & X &  & X & X \\
-Residual & &  & X & X & X & X \\
-\hline
-\end{tabularx}
-% \vspace{1in}
-% ditto with site interaction								& ditto above $\mid$ sp:site		& ditto above \\
-% \end{center}
-\caption{bla bla bla \label{model_variant}}
+\textbf{Package} & \pkg{nlme} & \pkg{phylolm} & \pkg{lme4}/\pkg{glmmTMB} & \pkg{pez} & \pkg{brms} & \pkg{MCMCglmm} \\ \hline
+Single Group & X & X & X &  & X & X \\ \hline
+\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-Intercept correlation\\ Residual\end{tabular} & X & \begin{tabular}[c]{@{}c@{}}X\\ \\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} \\ \hline
+Multi-group &  &  & X & X & X & X \\ \hline
+\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-intercept correlation\\ Phylo Species-group interaction\\ Species intercept\\ Species Slope \\ Species Slope-intercept correlation\\ Residual\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ \\ X\\ X\\ X\\ X\end{tabular} \\ \hline
+\end{tabular}%
+}
 \end{table}
 
-\subsection{Simulation and evaluations}
+%% STOP HERE, MLI will need to do everything below this comment. Everything above this comment is ready for BB to read.
+
+\subsection*{Simulation and evaluations}
 We simulated 100 phylogenetic trees for each sample size (25,50,100, 500) and incorporated with simulation parameters (illustrated in figure 2 and 5) to simulate responses via the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
@@
 
-\section{Results}
+\section*{Results}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 

### 2019-02-11 phyloglmm_ms.tex
results section complete, need to manually load brms results from Mac's side

@@
 \subsection*{Simulation and evaluations}
-We simulated 100 phylogenetic trees for each sample size (25,50,100, 500) and incorporated with simulation parameters (illustrated in figure 2 and 5) to simulate responses via the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
-All model variants via different platforms were used to fit each realization.
+We simulated 100 phylogenetic trees for each sample size (n = 25,50,100, 500) and incorporated simulation parameters (illustrated in figure 2 and 5) to simulate responses from the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
+Each realization were fitted to all model variant using different platforms.
 Tables \ref{model_variant} shows the parameters that are estimable for each platform. 
-All model must pass its respective convergence test before evaluating goodness of fit. 
-For bayesian GLMM, we combined two convergence criteria to assess convergence: we required a value of the
-Gelman and Rubin statistic and an effective sample size (ESS) greater than 1000 for each replication. 
+We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
+For Bayesian GLMM, we combined two criteria to assess convergence: we required a value of the Gelman and Rubin statistic less than 1.1 and an effective sample size (ESS) greater than 1000 the fixed effect parameters for each replication.
 For each replication we sample two chains starting with 10000 iterations. 
-
-We used coverage to assess model fit.
+%% TODO MAKE SURE THIS IS CORRECT!
+We first evaluate our estimates by looking at the distribution of the estimated values -- this gives us an idea of bias and variationce (i.e. quality of the point estimate).
+Then we look at coverage, to assess quality of confidence intervals.
 Coverage refers to the frequency with which the computed confidence intervals include the true values of
@@
 to evaluate coverage.
-We also compare computational speed between different platforms.
+We also compare computational speed between different platforms to evaluate the efficiency of the platforms and methods.
 
 \section*{Results}
+
+We used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \citep{li2017canfun}. 
+\pkg{lme4} and \pkg{pez} give identical results for fix and random effect estimates, our code is approximately 120 times faster than \pkg{pez}. 
+Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.
+
+\subsection{Single Group models simulations}
 
@@
 
-\bmb{add refs to figures \ldots}
-
 % \begin{itemize}
@@
 
-The full single site model using phyloglmm (which matches the single simulation model that incorporates phylogenetic intercept, slope and correlation) provides good estimates for all parameters except the random correlation parameter ($\sigma_{phy_{int},phy_{slope}}$).
-Fixed effect parameters ($\beta_0$ and $\beta_1$) estimates approaches nominal coverage as number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not others.
+The full fitted model (which matches the simulation model that incorporates phylogenetic intercept, slope and correlation) provides estimates with low bias and variance for all parameters. 
+Estimates for fixed effect parameters ($\beta_0$ and $\beta_1$) approach nominal coverage as the number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not for other packages. 
+\mli{we know why brms over covers}
+%% \mli{Is there a figure for this?....Ok, maybe We don't need it.}
 
-In general, models that cannot match the simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
-PGLM (which lacks the phylogenetic slope parameter) provides reasonable good estimates for phylogenetic intercept parameter ($\sigma_{phy_{int}}$) but overestimates the residual/observation standard deviation of the intercept ($\beta_0$) estimates slightly undercover the nominal coverage (95\% with 100 species) and have a bad coverage ($<$ 60\%) for the fixed slope parameter.
-PGLS, with only one parameter available, confounds all variations (phylogenetic intercept, slope and residual variation) into the single parameter (phylogenetic intercept), resulting in overestimating the phylogenetic intercept and overcovering for $\beta_0$, and undercovering for $\beta_1$.
+In general, models that are insufficiently flexible to match the true simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
+PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for phylogenetic intercept parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly conservative (90\% coverage of the 95\% CI with 100 species) and the fixed slope parameter ($\beta_1$) have poor coverage ($<$ 60\%).
+PGLS, with only one parameter available, confounds all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept and overcovering for $\beta_0$, and undercovering for $\beta_1$.
 
@@
 % \bmb{can we figure this out, even in an empirical/crude way??? it's tough to have results without knowing the true value they're aiming for ...}
+
+\subsection*{Multi-group model simulations}
 
@@
 
-Multiple site fits on the other hand, are much more similar across platforms (as the fitted and simulation models are more similar).
-Similar to the single site fits, \pkg{lme4} and \pkg{glmmTMB} which matches the simulation model provides good estimates for all parameters except for the random correlation parameter ($\sigma_{sp_{int,slope}}$).
-The lack of correlations in \pkg{pez} and \pkg{phyr} statistical models does not appear to have a large effect on its ability to estimate other parameters, whereas \pkg{pez} underestimates (lower estimates) many parameters.
+Multi-group model fits on the other hand, are much more similar across platforms (only the more powerful subset of platforms can fit these models at all, and the fitting models are closer to the true simulation model).
+Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides and good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int,slope}}}$).
+The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, whereas \pkg{pez} underestimates many parameters.
 % \bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate?
@@
 
-However, despite the similarities of parameter estimates across platforms for the multiple sites simulation fits, the efficiency (measured by computational speed) are different across platforms and sample size.
-For example, the median time for \pkg{lme4} to fit 50/100 species model takes $~$ 50/100 seconds versus $~$ 200/2000 seconds for \pkg{pez} and \pkg{phyr}. 
-Lastly, it takes $\sim$ 1000 seconds for lme4 to fit 500 species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500 species for our study, where the computational speed did not scale linearly with sample size 25/50/100.
+Although the parameter estimates are similar across platforms for the multi-group simulation fits, computational efficiency varies enormously across platforms and sample size.
+For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
+Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 (100) species model takes $\approx$ 9 (20) versus $\approx$ 200 (2000) seconds for \pkg{pez} and \pkg{phyr}.
+\pkg{glmmTMB} takes $\approx$ 125 seconds for lme4 to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species for our study, where the computational speed did not scale linearly with sample size 25/50/100.
 \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
@@
 %   approximate power?}
-
-We used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \cite{li2017canfun}. 
-We have obtained identical results for fix and random effect estimates, our code is approximately 120 times faster than \pkg{pez}. 
-Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.
 
@@
 
-The evolutionary model simulations presented in this paper are an oversimplification of real evolutionary processes in nature, but it is analogous to how researchers fit models to study complicated systems (as it is impossible to create a model to "match" the complexities of nature).
+The simulations presented in this paper are an oversimplification of the real evolutionary processes in nature, but they are no simpler than what researchers typically use.
+In fact, our simulations showed that even simple single predictor--response multi-species model can incorporate multiple levels of complexity (e.g.
+including random slopes, intercept-slope correlations, and the phylogenetic analog).
 We have fitted models using different platforms that best match the simulation models. 
-Models that cannot match the simulation (full) model are expected to underperform; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods available.
+Models that cannot match the simulation (full) model are expected to perform poorly; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods available.
 Using models that include some form of phylogenetic variation and residual variation is necessary to detect evolutionary process.
@@
 
-\subsection{Process and residual (observation) variabilities and confoundings}
+\subsection{Process and observation variabilities and confoundings}
 
-\bmb{explain: classical scenario for PIC is when we have species-level data (i.e., species averages or ``typical'' values).
+% \bmb{explain: classical scenario for PIC is when we have species-level data (i.e., species averages or ``typical'' values).
 What do we do when we have multiple observations per species?

### 2019-02-11 collect_brms.R

@@ -1,0 +1,39 @@
+library(brms)
+library(dplyr)
+
+brms_path <- "./datadir/brms/"
+brmsss_res <- list.files(path = brms_path, pattern = "ss")
+brmsss_results <- function(tt){
+  brms_df <- data.frame(resid = numeric(200)
+                        , phylo_X = numeric(200)
+                        , phylo_int = numeric(200)
+                        , phylo_cor = numeric(200)
+                        , B0 = numeric(200)
+                        , B1 = numeric(200)
+                        , model = numeric(200)
+                        , time = numeric(200)
+                        , convergence = NA
+  )
+  for(i in 1:length(tt)){
+    brms_obj <- readRDS(paste(brms_path,tt[i],sep=""))
+    sd_dat <- as.data.frame(posterior_samples(brms_obj[[1]],c("^sigma","^sd_","^cor_")))
+    brms_df[i,"resid"] <- median(sd_dat[,"sigma"])
+    brms_df[i, "phylo_X"] <- median(sd_dat[,"sd_sp__X"])
+    brms_df[i, "phylo_int"] <- median(sd_dat[,"sd_sp__Intercept"])
+    brms_df[i,"phylo_cor"] <- median(sd_dat[,"cor_sp__Intercept__X"])
+    b_dat <- as.data.frame(posterior_samples(brms_obj[[1]], c("^b")))
+    brms_df[i, "B0"] <- as.numeric(between(0
+                                           , quantile(b_dat[,"b_Intercept"], 0.025)
+                                           , quantile(b_dat[,"b_Intercept"], 0.975)
+    ))
+    brms_df[i,"B1"] <- as.numeric(between(0
+                                          , quantile(b_dat[,"b_X"], 0.025)
+                                          , quantile(b_dat[,"b_X"], 0.975)
+    ))
+    brms_df[i,"model"] <- tt[i]
+    brms_df[i,"time"] <- brms_obj[[2]][[1]]
+  }
+  return(brms_df)
+}
+
+brmsss_data <- brmsss_results(brmsss_res)

### 2019-02-11 plot.R

@@ -26,16 +26,16 @@ ssdat <- (ssdat_raw
 	%>% mutate(size = factor(size
 	      , levels=c("small","med","large","xlarge"), labels=c("25","50","100","500")
 			)
-			, platform = factor(platform, levels=c("gls", "phylolm","lme4", "brms"))
+			, platform = factor(platform, levels=c("gls", "phylolm","lme4", "brms", "glmmTMB"))
 		)
-	%>% left_join(.,par_df)
+	%>% left_join(.,sspar_df)
 	# %>% filter(platform != "brms")
 )
 
 
 gg_ss <- (ggplot(data=ssdat, aes(x=size, y=sd, fill=platform))
   + facet_wrap(~sdtype, scale="free_y")
-  + geom_violin(position=position_dodge(width=0.2),alpha=0.4)
+  + geom_violin(position=position_dodge(width=0.5),alpha=0.4)
   + geom_hline(aes(yintercept = y_int))
   # + scale_fill_manual(values=c("Dark Red","Dark Blue","Dark Green","Purple"))
   + scale_fill_brewer(palette = "Dark2")
@@ -51,6 +51,7 @@ gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, fill=platform))
 	+ scale_y_log10()
 	+ ylab("Time (seconds)")
 	# + ggtitle("Single site time")
+  + theme_bw()
 )
 
 print(gg_sstime)
@@ -87,7 +88,7 @@ mspar_df <- data.frame(
 msdat <- (msdat_raw
 	%>% separate(model,c("platform", "sites", "size", "seed", "saveformat"),"[.]")
 	%>% select(-c(sites,seed,saveformat))
-	%>% filter(convergence != 1)
+	%>% filter((convergence != 1) | is.na(convergence))
 	%>% select(-convergence)
 	%>% gather(key=sdtype, value=sd, -c(platform,size,time,B0,B1))
 	%>% mutate(size = factor(size,
@@ -111,7 +112,7 @@ print(gg_ms
 
 gg_mstime <- (gg_sstime
 	%+% msdat
-	+ ggtitle("multiple site time")
+	# + ggtitle("multiple site time")
 )
 
 print(gg_mstime)
@@ -138,5 +139,33 @@ ms_coverage <- (msdat
                 %>% gather(key=fixed_parameter, value=coverage, -c(platform, size))
 )
 
[…]

### 2019-04-08 phyloglmm_ms.tex
fix some typos

@@
 \usepackage{graphicx}
+\usepackage{float}
 \usepackage{grffile}
@@
 %\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
-Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such observations to partition variablility more finely.
+Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such observations to partition variability more finely.
 Indeed, many existing methods cannot accommodate multiple observations per species, instead requiring users to collapse the data by taking species averages.
@@
 \begin{center}
-\begin{figure}[h]
+\begin{figure}[H]
   \includegraphics[scale=0.8]{./phylo_diagram.png}
@@
 The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
-Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.
+Table \ref{table:model} provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.
 
@@
 \end{tabular}
+\caption{Model and platforms}
+\label{table:model}
 \end{table}
@@
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
-Our new formulation can be incorporated in any framework that allows for random effect without the need to implement special correlation structures by incorportating phylogenetic structures as part of the mean model.
+Our new formulation can be incorporated in any framework that allows for random effect without the need to implement special correlation structures by incorporating phylogenetic structures as part of the mean model.
 %% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
@@
 Alternatively, one can use the generalized linear mixed effects modelling (GLMM) framework \citep{lynch1991methods}.
-The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorportate more than one kind of variability through random effects and for non-gaussian responses.
+The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorporate more than one kind of variability through random effects and for non-Gaussian responses.
 The typical GLMM has the form:
@@
 
-There are serveral advantages in modelling phylogenetic correlated traits in the mixed model framework.
+There are several advantages in modelling phylogenetic correlated traits in the mixed model framework.
 First, random intercepts can allow the response trait to vary across groups/levels within grouping predictors (e.g. patch, site, experiment, and etc). 
 Second, it can also allow the response (trait or distribution) to vary according to the phylogenetic relationships across species (i.e.
-similar species will have similiar responses), and between species (tip variation).
+similar species will have similar responses), and between species (tip variation).
 The third type of variation that is often neglected or rarely considered are random slopes.
 Random slopes can allow fixed effects (size of the effect in continuous predictor variables) to vary for each species. 
-Analogous to modelling the repsonse dependency via phylogenetic relationship among species, phylogenetic random slopes can allow fixed effects to account for the phylogenetic relations (i.e.
+Analogous to modelling the response dependency via phylogenetic relationship among species, phylogenetic random slopes can allow fixed effects to account for the phylogenetic relations (i.e.
 similar species will have similar size of the effects on the response than distinct pairs).
@@
 \begin{center}
-\begin{figure}[h]
+\begin{figure}[H]
   \includegraphics[scale=0.8,page=1]{./git_push/tree.pdf}
@@
 We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
-For simplicity, the response variable $\bY$ is conditional normally distributed (i.e. gaussian distribution with an idenitiy link function) , which is a linear mixed effect model. 
+For simplicity, the response variable $\bY$ is conditional normally distributed (i.e. Gaussian distribution with an identity link function) , which is a linear mixed effect model. 
 For the first set of simulation, we simulate one observation per species.
@@
 Lastly, Bayesian phylogenetic GLMM using Markov chain Monte Carlo (MCMC) are available and can handle all of the cases described above. 
-However, MCMC is often computationally expensive for GLMMs compared to platforms using determinstic optimization.
+However, MCMC is often computationally expensive for GLMMs compared to platforms using deterministic optimization.
 \pkg{MCMCglmm} \citep{hadfield2010general} is the most commonly used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid} algorithm \citep{li2018fitting}.
@@
 }
+\caption{Parameters}
+\label{table:platform}
 \end{table}
-
-%% STOP HERE, MLI will need to do everything below this comment. Everything above this comment is ready for BB to read.
 
@@
 Each realization were fitted to all model variant using different platforms.
-Tables \ref{model_variant} shows the parameters that are estimable for each platform. 
+Tables \ref{table:platform} shows the parameters that are estimable for each platform. 
 We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
-For Bayesian GLMM, we combined two criteria to assess convergence: we required a value of the Gelman and Rubin statistic less than 1.1 and an effective sample size (ESS) greater than 1000 the fixed effect parameters for each replication.
+For Bayesian GLMM, we combined two criteria to assess convergence: we required a value of the Gelman and Rubin statistic less than 1.1 and an effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0} \ \textrm{and} \ \beta_{1}$) for each replication.
 For each replication we sample two chains starting with 10000 iterations. 
 %% TODO MAKE SURE THIS IS CORRECT!
-We first evaluate our estimates by looking at the distribution of the estimated values -- this gives us an idea of bias and variationce (i.e. quality of the point estimate).
-Then we look at coverage, to assess quality of confidence intervals.
-Coverage refers to the frequency with which the computed confidence intervals include the true values of
-parameters.
-We used 90\% (Wald and profile) confidence intervals and quantile-based intervals (for bayesian GLMM)
-to evaluate coverage.
+We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms -- this gives us an idea of bias and variation (i.e.
+quality of the point estimate).
+Then we look at frequentist coverage, to assess quality of confidence intervals.
+Coverage refers to the frequency with which the computed confidence intervals include the true values of parameters.
+We used 90\% (Wald and profile) confidence intervals and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
 We also compare computational speed between different platforms to evaluate the efficiency of the platforms and methods.
@@
 
-\subsection{Single Group models simulations}
+\subsection*{Single Group models simulations}
 
 \begin{center}
-\begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./figure/ssplot.pdf}
+\begin{figure}[H]
+  \includegraphics[scale=0.8,page=1]{./git_push/plot.Rout.pdf}
   \caption{New}
@@
 \begin{center}
-\begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./figure/sstime.pdf}
+\begin{figure}[H]
+  \includegraphics[scale=0.8,page=2]{./git_push/plot.Rout.pdf}
   \caption{New}
@@
 \begin{center}
-\begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./figure/sscoverage.pdf}
+\begin{figure}[H]
+  \includegraphics[scale=0.8,page=3]{./git_push/plot.Rout.pdf}
   \caption{New}
@@
 PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for phylogenetic intercept parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly conservative (90\% coverage of the 95\% CI with 100 species) and the fixed slope parameter ($\beta_1$) have poor coverage ($<$ 60\%).
-PGLS, with only one parameter available, confounds all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept and overcovering for $\beta_0$, and undercovering for $\beta_1$.
+PGLS, with only one parameter available, confounds all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept and over-covering for $\beta_0$, and under-covering for $\beta_1$.
 
@@
 \begin{center}
-\begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./figure/msplot.pdf}
+\begin{figure}[H]
+  \includegraphics[scale=0.8,page=4]{./git_push/plot.Rout.pdf}
   \caption{new}
@@
 \begin{center}
-\begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./figure/mstime.pdf}
+\begin{figure}[H]
+  \includegraphics[scale=0.8,page=5]{./git_push/plot.Rout.pdf}
   \caption{new}
@@
 \begin{center}
-\begin{figure}[h]
-  \includegraphics[scale=0.8,page=1]{./figure/mscoverage.pdf}
+\begin{figure}[H]
+  \includegraphics[scale=0.8,page=6]{./git_push/plot.Rout.pdf}
   \caption{new}
@@
 
-\subsection{Process and observation variabilities and confoundings}
+\subsection{Process and observation variability and confounding}
 
@@
 If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation.
-However, one can summarise multiple measurements to a single measurement (for example, mean or median) per species to avoid confounding residual and tip variations.
+However, one can summarize multiple measurements to a single measurement (for example, mean or median) per species to avoid confounding residual and tip variations.
 
@@
 In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
-Neglicting random slopes can lead to erroneous fixed effect estimates \citep{schielzeth2008conclusions} (parameters of interest) as shown in the simulations above. 
+Neglecting random slopes can lead to erroneous fixed effect estimates \citep{schielzeth2008conclusions} (parameters of interest) as shown in the simulations above. 
 When analyzing relationships among species between traits and environmental variables, it is entirely plausible that change in effect of predictor variables have on traits are changing in a phylogenetically correlated way.

### 2019-04-09 phyloglmm_ms.tex
filled in all figure and table captions

@@
 Indeed, many existing methods cannot accommodate multiple observations per species, instead requiring users to collapse the data by taking species averages.
+Figure \ref{fig:variation} illustrates a visual categorization of various forms of variation for three species. 
 %Despite having multiple observations per tip taxon (species), a typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
@@
   \includegraphics[scale=0.8]{./phylo_diagram.png}
-  \caption{Visual categorization of various forms of variation.}
+  \caption{Visual categorization of various forms of phylogenetic variation of three species.
+Top left shows pure noise (i.e.
+assuming a star phylogeny); top right accounts for phylogeny, but without noise; bottom left accounts both phylogeny and noise at the tip, and finally bottom right shows multiple observations (with noise) at the tips.}
 \label{fig:variation}
@@
 \end{tabular}
-\caption{Model and platforms}
+\caption{List of phylogenetic generalized linear models and R packages.}
 \label{table:model}
@@
 \bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
-\besp & \sim \textrm{N}(0,\sigma^{2}_{\epsilon})
+\besp & \sim \textrm{N}(0,\sigma^{2}_{\epsilon}) \label{eq:glmm4}
 \end{align}
@@
 }
-\caption{Parameters}
+\caption{List of estimable parameters for each R package.}
 \label{table:platform}
@@
 \subsection*{Simulation and evaluations}
-We simulated 100 phylogenetic trees for each sample size (n = 25,50,100, 500) and incorporated simulation parameters (illustrated in figure 2 and 5) to simulate responses from the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
+We simulated 100 phylogenetic trees for each sample size (n = 25,50,100 and additional n = 500 for multi-group model) and incorporated simulation parameters (illustrated in Figure \ref{ssplot} and Figure \ref{msplot}) to simulate responses from the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
 Each realization were fitted to all model variant using different platforms.
@@
   \includegraphics[scale=0.8,page=1]{./git_push/plot.Rout.pdf}
-  \caption{New}
+  \caption{Comparsion of single group model parameter estimates across different R packages in Table \ref{table:platform}.
+The horizontal line shows the true value of the parameters in the simulation model.
+Models capable of fitting all parameters (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) fit well for all parameters.}
 \label{ssplot}
@@
   \includegraphics[scale=0.8,page=2]{./git_push/plot.Rout.pdf}
-  \caption{New}
-\label{ssplot}
+  \caption{Comparison of speed for all fitting packages variants: layout of the packages as in Figure \ref{ssplot}}
+\label{ssplot_speed}
 \end{figure}
@@
   \includegraphics[scale=0.8,page=3]{./git_push/plot.Rout.pdf}
-  \caption{New}
-\label{ssplot}
+  \caption{Comparison of coverage probablility for fixed effect parameters.
+Models matching the simulation model (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) have coverage near the nominal value of 0.95.
+The black line shows the nominal coverage, and the grey ribbon the 95\% binomial confidence interval based on 100 simulated fits.}
+\label{ssplot_coverage}
 \end{figure}
@@
 
-% \bmb{Need to think a little harder about the important points/order here (might be time to do the exercise where you take the results and boil them back down to short bullet points, in order to figure out whether you've said the important stuff, in the right order).
-Maybe lead by saying that all platforms give similar answers - to the extent that they are fitting the same statistical model (this might even be emphasized/explained above, in Methods).
-Am I correct in believing that phylolm=\pkg{lme4} but without intercept/slope correlation, and gls is additionally missing random slopes?}
-
-% \begin{itemize}
-% \item{full model provides good estimates (except for correlation)}
-% \item{reduced models try to fit with whatever parameters they have available (over-estimates the simulation parameters)}
-% \item{no slope vs slope}
-% \item{no correlation vs correlation}
-% \item{as size/data/number of species increase, the parameter estimates gets better}
-% \item{single site models are fast}
-% \item{estimates gets closer to simulation parameter as number of species increase}
-% \item{lme4/tmb are fast, everything else is slow for multiple sites}
-% \end{itemize}
-
-The full fitted model (which matches the simulation model that incorporates phylogenetic intercept, slope and correlation) provides estimates with low bias and variance for all parameters. 
+The full fitted model (which matches the simulation model that incorporates phylogenetic intercept, slope and correlation) provides estimates with low bias (the difference between the parameter estimates and true simulation parameters) and variance for all parameters.
 Estimates for fixed effect parameters ($\beta_0$ and $\beta_1$) approach nominal coverage as the number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not for other packages.
-\mli{we know why brms over covers}
+\pkg{brms} have higher coverage than the nominal coverage at 95 \% because the prior distributions for the simulation parameters are cenetered at the true values.
 %% \mli{Is there a figure for this?....Ok, maybe We don't need it.}
@@
 
-% \bmb{(2) please define ``robustness'' -- usually means lack of sensitivity to model.
-Are you talking about the variance of the estimates?
-I'd prefer not to use ``robust'' in a non-standard way if possible (2) what are small/medium/large?
-I see that $n=\{20,100,500\}$ and $m=\{1,20\}$, but not sure how these match up to ``small'', ``medium'', ``large''; is Figure 1 all $m=1$?
-Might be worth just labeling by $n=??$ rather than using small/medium/large}
-% It is unclear how different levels of phylogenetic signal are being confounded under the error free (without tip variation) assumption for the PGLS models.
-% \bmb{can we figure this out, even in an empirical/crude way??? it's tough to have results without knowing the true value they're aiming for ...}
-
 \subsection*{Multi-group model simulations}
@@
   \includegraphics[scale=0.8,page=4]{./git_push/plot.Rout.pdf}
-  \caption{new}
+  \caption{Comparsion of multi group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
+  \label{msplot}
 \end{figure}
@@
   \includegraphics[scale=0.8,page=5]{./git_push/plot.Rout.pdf}
-  \caption{new}
+  \caption{Comparsion of multi group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
+  \label{msplot_time}
 \end{figure}
@@
   \includegraphics[scale=0.8,page=6]{./git_push/plot.Rout.pdf}
-  \caption{new}
+  \caption{Comparsion of multi group model coverage across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
+  \label{msplot_coverage}
 \end{figure}

### 2019-04-09 phyloglmm_ms.tex
finally reach the discussion .... again...

@@
   \includegraphics[scale=0.8,page=1]{./git_push/plot.Rout.pdf}
-  \caption{Comparsion of single group model parameter estimates across different R packages in Table \ref{table:platform}.
+  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
 The horizontal line shows the true value of the parameters in the simulation model.
@@
   \includegraphics[scale=0.8,page=4]{./git_push/plot.Rout.pdf}
-  \caption{Comparsion of multi group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
+  \caption{Comparison of multi-group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
   \label{msplot}
@@
   \includegraphics[scale=0.8,page=5]{./git_push/plot.Rout.pdf}
-  \caption{Comparsion of multi group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
+  \caption{Comparison of multi-group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
   \label{msplot_time}
@@
   \includegraphics[scale=0.8,page=6]{./git_push/plot.Rout.pdf}
-  \caption{Comparsion of multi group model coverage across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
+  \caption{Comparison of multi-group model coverage across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
   \label{msplot_coverage}
@@
 Multi-group model fits on the other hand, are much more similar across platforms (only the more powerful subset of platforms can fit these models at all, and the fitting models are closer to the true simulation model).
-Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides and good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int,slope}}}$).
+Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides and good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int,slope}}}$) for few number of species (i.e.
+n = 25 and 50).
 The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, whereas \pkg{pez} underestimates many parameters.
-% \bmb{Can you identify the extreme cases in \pkg{lme4}'s estimate?
-Are these singular models?
-Would they be diagnosable?
-Can you say anything about when the assumption of independence of slope and intercept \textbf{would} be a problem?
-(If you can think of such a scenario, you could run sims for one such example - don't need to go into great detail, but being able to say that you confirmed that it's a problem in such-and-such a case would be important.)}
-% \mli{save it for random slope discussion}
 
@@
 Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 (100) species model takes $\approx$ 9 (20) versus $\approx$ 200 (2000) seconds for \pkg{pez} and \pkg{phyr}.
-\pkg{glmmTMB} takes $\approx$ 125 seconds for lme4 to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species for our study, where the computational speed did not scale linearly with sample size 25/50/100.
+\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species for our study, where the computational speed did not scale linearly with sample size 25/50/100.
 \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.

### 2019-04-09 phylog.rmd
almost done?

---
title: "Phylogenetic comparative methods in the lme4-verse"
author: Michael Li and Ben Bolker
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
output:
    html_document:
    mathjax: default
---

```{r pkgs,message=FALSE, warning=FALSE}
library(ape)
library(lme4)
library(Matrix)
library(MCMCglmm)
library(MASS) 
library(pez)
library(glmmTMB)
library(dplyr)
library(coda)
library(lattice)
library(broom)
library(dotwhisker)
```

- The standard problem of *phylogenetic comparative methods* is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree - for example, regression analyses of body temperature as a function of body size for animals within a clade
- More generally, we can frame this in the usual GLMM way as
$$
\begin{split}
y & \sim D(\mu,\phi) \\
\mu & = g^{-1}(\eta) = g^{-1}(X \beta + Z b) \\
b & \sim \textrm{MVN}(0,\Sigma)
\end{split}
$$
where the part that makes it specifically phylogenetic is that $\Sigma$ captures the *phylogenetic correlation*. The PC is the correlation among observations due to relatedness; recently diverged taxa have higher correlation than more anciently diverged taxa. In the extreme case of a *star phylogeny* (all taxa diverged from each other simultaneously at some point in the past) the phylogenetic correlation collapses to a diagonal matrix and we get back to the simple, uncorrelated regression.

Various P(G)LMM (phylogenetic [generalized] linear mixed model] approaches have been proposed. Many depend on Pagel's lambda transformation, which gives the correlation matrix a particularly simple form (but has been [criticized](http://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html) ...)

An alternative approach is to model the phylogenetic correlation as a *Gaussian process*. In particular, suppose that the evolutionary process is a Brownian motion (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on). In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to do fancy things to impose a correlation structure on the random effects.

## Nuts and bolts: from a phylogeny to a $Z$ matrix for the GP

```{r setup}
library(ape)
set.seed(101)
r <- makeNodeLabel(rtree(4))
plot(r,show.node.label=TRUE)
```

Information in a `phylo` object is contained in the *edge matrix*:

>    edge: a two-column matrix of mode numeric where each row represents
          an edge of the tree; the nodes and the tips are symbolized
          with numbers; the tips are numbered 1, 2, ..., and the nodes
          are numbered after the tips. For each row, the first column
          gives the ancestor.

```{r edge}
t(r$edge)
```

and a list of *edge lengths*

```{r edgelen}
r$edge.length
```

Inspecting this tree, we can figure out (see `$tip.label` and `$node.label` for label-to-number correspondences):

- tips are 1-4, nodes are 5-7
- tip 1 (`t1`) involves branches 2 (6 $\to$ 1) and 1 (5 $\to$ 6).
- tip 2 (`t3`) involves branches 3 (6 $\to$ 2) and 1 (5 $\to$ 6)
- tip 3 (`t2`) involves branches 5 (7 $\to$ 3) and 4 (5 $\to$ 7)
- tip 4 (`t4`) involves branches 6 (7 $\to$ 4) and 4 (5 $\to$ 7)

So, for example, we can say that the 'error' value corresponding to tip 1 is
$\ell_1 b_1 + \ell_2 b_2$,  where $\ell_i$ is the (square root of??) the branch length
and the $b_i$ are independent, homoscedastic Normal variates. Alternately, the $Z$ matrix is

$$
\begin{pmatrix}
\ell_1 & \ell_2 &   0     & 0 & 0 & 0 \\
\ell_1 &    0   &  \ell_3 & 0 & 0 & 0 \\
0 & 0 &  0  & \ell_4 & \ell_5 & 0 \\
0 & 0 &  0  & \ell_4 & 0 & \ell_6
\end{pmatrix}
$$
where $\ell_i$ is the length of the $i^\textrm{th}$ branch,
so that the species effects are $Z b$.

If we can build the corresponding $Z$ matrix, then we can insert it in
the `lme4` modular model-fitting process (see `?modular`).

Here's a (probably not very efficent) way to construct the Z matrix.
(There must be a way to not walk the tree multiple times from every tip ...

```{r phylo.to.Z}
phylo.to.Z <- function(r,stand=FALSE){
  ntip <- length(r$tip.label)
  Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
  nodes <- (ntip+1):max(r$edge)
  root <- nodes[!(nodes %in% r$edge[,2])]
  for (i in 1:ntip){
    cn <- i  ## current node
    while (cn != root){
      ce <- which(r$edge[,2]==cn)   ## find current edge
      Zid[i,ce] <- 1   ## set Zid to 1
      cn <- r$edge[ce,1]            ## find previous node
    }
  }
  V <- vcv(r)
  # V <- V/max(V)
  sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
  # sig <- det(V)^(1/ntip)
  Z <- t(sqrt(r$edge.length) * t(Zid))
  if(stand){Z <- t(sqrt(r$edge.length/sig) * t(Zid))}
  rownames(Z) <- r$tip.label
  colnames(Z) <- 1:length(r$edge.length)
  return(Z)                                  
}
phylo.to.Z(r)
```

(This could benefit from the repeated-entry sparse matrix class that Steve Walker wrote.)

On the other hand, it only takes a few seconds to run for a 200-species phylogeny (see below).

## constructing a GP PGLMM with lme4: machinery

"All" we need to do is (1) call `(g)lFormula`, with a formula that includes a `(1|phylo)` term, to build the basic (wrong) structure; (2) modify the `reTrms` component of the structure appropriately; (3) go through the rest of the modular procedure for building a (G)LMM.

```{r phylocode}
source("lme4_phylo_setup.R")
```

## glmmTMB fits: nuts and bolts

`glmmTMB` can be deconstructed in a similar way. In fact, we can re-use a lot of the
machinery. Being able to use `glmmTMB` means we can use a broader range of distributions,
zero-inflation, etc. (machinery below assumes phylogenetic structure only in the
conditional distribution). This is also a little clunky, some adjustment on the `glmmTMB`
side might make it a bit easier.

```{r tmbfit0}
source("glmmTMB_phylo_setup.R")
```

## Example

### get data

From [chapter 11](http://www.mpcm-evolution.com/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of Garamszegi (ed.): data are [here](http://mpcm-evolution.com/OPM/Chapter11_OPM/data.zip)

```{r getdat}
if (!file.exists("data/phylo.nex")) {
    dir.create("data")
    download.file("http://mpcm-evolution.com/OPM/Chapter11_OPM/data.zip",
                  dest="data/OPM_ch11_data.zip")
    setwd("data")
    untar("OPM_ch11_data.zip")
    setwd("..")
}
phylo <- read.nexus("data/phylo.nex")
```

Compute appropriate $Z$ matrix up front, to measure speed (also
reusable in a few places below):

```{r calc_phyloZ,cache=TRUE}
system.time(phyloZ <- phylo.to.Z(phylo))
```

## Result comparison with Gaussian example in chapter 11

```{r G_results}
datG <- read.table("data/data_simple.txt",header=TRUE)
datG$obs <- factor(seq(nrow(datG)))
datG <- datG %>% mutate(sp = phylo)
phylo_lmm_fit <- phylo_lmm(phen~cofactor+(1|sp)
  , data=datG
  , phylonm = "sp"
  , phylo = phylo
  , phyloZ=phyloZ
  , REML = TRUE
  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
)

print(summary(phylo_lmm_fit))

```

## Result comparison with Gaussian with repeated measures example in chapter 11

```{r GR_results}
datR <- read.table("data/data_repeat.txt",header=TRUE)
datR$obs <- factor(seq(nrow(datR)))
datR <- (datR 
  %>% mutate(sp = species
        , animals = phylo
        )
)
datR$spec_mean_cf <- sapply(split(datR$cofactor,datR$phylo),mean)[datR$phylo]
datR$within_spec_cf <- datR$cofactor-datR$spec_mean_cf
phylo_lmm_fit <- phylo_lmm(phen~spec_mean_cf+within_spec_cf+(1|sp) + (1|animals)
  , data=datR
  , phylonm = "sp"
  , phylo = phylo
  , phyloZ=phyloZ
  , REML = FALSE
  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
)

print(summary(phylo_lmm_fit))

```

## Result comparison with non-Gaussian example in chapter 11

```{r P_results}
dat <- read.table("data/data_pois.txt",header=TRUE)
dat$obs <- factor(seq(nrow(dat)))

dat <- dat %>% mutate(sp=phylo) 
phylo_glmm_fit <- phylo_glmm(phen_pois~cofactor+(1|sp)+(1|obs)
  , data=dat
  , phylonm = "sp"
  , family = poisson
  , phylo = phylo
  , phyloZ=phyloZ
  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
)

summary(phylo_glmm_fit)

```

### 2019-04-09 vignette.Rmd
almost done?

@@
 
-```{r}
+```{r simdata}
 ### simulate phylogenetic tree
@@
 # Basic parameters
-nspp <- 3
+nspp <- 1000
 
@@
 
-```{r}
+```{r phyloZ}
 
@@
 
-source("phyloglmm_setup.R")
+source("lme4_phylo_setup.R")
 
@@
 
-print(phyZ)
+# print(phyZ)
 
@@
 
-```{r}
+```{r fit}
 
-source("new_phylo_setup.R")
+dat2 <- (dat %>% mutate(obs = sp, sp = paste("t",sp,sep="")))
 
-dat <- (dat %>% mutate(obs = sp))
-
-## everything can be reduce down to formula,data, and a phy object
-
-fitmod <- phylo_lmm2(Y ~ X + (1 |sp) + (1 + X|sp)
-			, data=dat 
-			, phylonm = "sp" # wishy-washy
-			, phyloZ = phyZ # Maybe we just want the phy object and compute this internally
-			, nsp = nspp # make default check
-			, control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")  ## make default
-			, REML = FALSE # make default 
+fitmod <- phylo_lmm(Y ~ X + (1 + X|sp)
+			, data=dat2
+			, phylonm = "sp" 
+			, phyloZ = phyZ 
+			, control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")  
+			, REML = FALSE 
 )

### 2019-04-09 example.rmd
pushing branch

--- 2019-04-09 phylog.rmd
@@
 ---
-
-```{r pkgs,message=FALSE, warning=FALSE}
-library(ape)
-library(lme4)
-library(Matrix)
-library(MCMCglmm)
-library(MASS) 
-library(pez)
-library(glmmTMB)
-library(dplyr)
-library(coda)
-library(lattice)
-library(broom)
-library(dotwhisker)
-```
 
@@
 
-An alternative approach is to model the phylogenetic correlation as a *Gaussian process*.
+An alternative approach is to model the phylogenetic correlation as a *Gaussian process* (GP).
 In particular, suppose that the evolutionary process is a Brownian motion (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on).
@@
 library(ape)
+library(Matrix)
 set.seed(101)
@@
 Here's a (probably not very efficent) way to construct the Z matrix.
-(There must be a way to not walk the tree multiple times from every tip ...
+(There must be a way to not walk the tree multiple times from every tip ...)
 
@@
   V <- vcv(r)
-  # V <- V/max(V)
   sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
-  # sig <- det(V)^(1/ntip)
   Z <- t(sqrt(r$edge.length) * t(Zid))
@@
 
+```{r pkgs,message=FALSE, warning=FALSE}
+library(ape)
+library(lme4)
+library(MCMCglmm)
+library(MASS) 
+library(pez)
+library(glmmTMB)
+library(dplyr)
+library(coda)
+library(lattice)
+library(broom)
+library(dotwhisker)
+```
+
+The phylo--to--Z function is already in the source code.
 ```{r phylocode}

### 2019-04-10 example.rmd
done

--- 2019-04-09 phylog.rmd
@@
 ---
-title: "Phylogenetic comparative methods in the lme4-verse"
-author: Michael Li and Ben Bolker
-date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
+title: "Example"
+author: "Michael Li and Ben Bolker"
 output:
-    html_document:
+  pdf_document: default
   mathjax: default
+  html_document: null
 ---
+
+We will reproduce the examples in [chapter 11](http://www.mpcm-evolution.com/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of Garamszegi (ed.) using lme4 and glmmTMB.
+
+To fit we need a random effect in the formula that includes a `(...|phylo)` term, to build the basic random effect structure multipled by the phyloZ matrix (see vignette for more details on phyloZ matrix).
 
@@
 library(ape)
+library(Matrix)
 library(lme4)
-library(Matrix)
 library(MCMCglmm)
@@
 
-- The standard problem of *phylogenetic comparative methods* is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree - for example, regression analyses of body temperature as a function of body size for animals within a clade
-- More generally, we can frame this in the usual GLMM way as
-$$
-\begin{split}
-y & \sim D(\mu,\phi) \\
-\mu & = g^{-1}(\eta) = g^{-1}(X \beta + Z b) \\
-b & \sim \textrm{MVN}(0,\Sigma)
-\end{split}
-$$
-where the part that makes it specifically phylogenetic is that $\Sigma$ captures the *phylogenetic correlation*.
-The PC is the correlation among observations due to relatedness; recently diverged taxa have higher correlation than more anciently diverged taxa.
-In the extreme case of a *star phylogeny* (all taxa diverged from each other simultaneously at some point in the past) the phylogenetic correlation collapses to a diagonal matrix and we get back to the simple, uncorrelated regression.
-
-Various P(G)LMM (phylogenetic [generalized] linear mixed model] approaches have been proposed.
-Many depend on Pagel's lambda transformation, which gives the correlation matrix a particularly simple form (but has been [criticized](http://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html) ...)
-
-An alternative approach is to model the phylogenetic correlation as a *Gaussian process*.
-In particular, suppose that the evolutionary process is a Brownian motion (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on).
-In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
-If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to do fancy things to impose a correlation structure on the random effects.
-
-## Nuts and bolts: from a phylogeny to a $Z$ matrix for the GP
-
-```{r setup}
-library(ape)
-set.seed(101)
-r <- makeNodeLabel(rtree(4))
-plot(r,show.node.label=TRUE)
-```
-
-Information in a `phylo` object is contained in the *edge matrix*:
-
->    edge: a two-column matrix of mode numeric where each row represents
-          an edge of the tree; the nodes and the tips are symbolized
-          with numbers; the tips are numbered 1, 2, ..., and the nodes
-          are numbered after the tips. For each row, the first column
-          gives the ancestor.
-
-```{r edge}
-t(r$edge)
-```
-
-and a list of *edge lengths*
-
-```{r edgelen}
-r$edge.length
-```
-
-Inspecting this tree, we can figure out (see `$tip.label` and `$node.label` for label-to-number correspondences):
-
-- tips are 1-4, nodes are 5-7
-- tip 1 (`t1`) involves branches 2 (6 $\to$ 1) and 1 (5 $\to$ 6).
-- tip 2 (`t3`) involves branches 3 (6 $\to$ 2) and 1 (5 $\to$ 6)
-- tip 3 (`t2`) involves branches 5 (7 $\to$ 3) and 4 (5 $\to$ 7)
-- tip 4 (`t4`) involves branches 6 (7 $\to$ 4) and 4 (5 $\to$ 7)
-
-So, for example, we can say that the 'error' value corresponding to tip 1 is
-$\ell_1 b_1 + \ell_2 b_2$,  where $\ell_i$ is the (square root of??) the branch length
-and the $b_i$ are independent, homoscedastic Normal variates. Alternately, the $Z$ matrix is
-
-$$
-\begin{pmatrix}
-\ell_1 & \ell_2 &   0     & 0 & 0 & 0 \\
-\ell_1 &    0   &  \ell_3 & 0 & 0 & 0 \\
-0 & 0 &  0  & \ell_4 & \ell_5 & 0 \\
-0 & 0 &  0  & \ell_4 & 0 & \ell_6
-\end{pmatrix}
-$$
-where $\ell_i$ is the length of the $i^\textrm{th}$ branch,
-so that the species effects are $Z b$.
-
-If we can build the corresponding $Z$ matrix, then we can insert it in
-the `lme4` modular model-fitting process (see `?modular`).
-
-Here's a (probably not very efficent) way to construct the Z matrix.
-(There must be a way to not walk the tree multiple times from every tip ...
-
-```{r phylo.to.Z}
-phylo.to.Z <- function(r,stand=FALSE){
-  ntip <- length(r$tip.label)
-  Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
-  nodes <- (ntip+1):max(r$edge)
-  root <- nodes[!(nodes %in% r$edge[,2])]
-  for (i in 1:ntip){
-    cn <- i  ## current node
-    while (cn != root){
-      ce <- which(r$edge[,2]==cn)   ## find current edge
-      Zid[i,ce] <- 1   ## set Zid to 1
-      cn <- r$edge[ce,1]            ## find previous node
-    }
-  }
-  V <- vcv(r)
-  # V <- V/max(V)
-  sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
-  # sig <- det(V)^(1/ntip)
-  Z <- t(sqrt(r$edge.length) * t(Zid))
-  if(stand){Z <- t(sqrt(r$edge.length/sig) * t(Zid))}
-  rownames(Z) <- r$tip.label
-  colnames(Z) <- 1:length(r$edge.length)
-  return(Z)                                  
-}
-phylo.to.Z(r)
-```
-
-(This could benefit from the repeated-entry sparse matrix class that Steve Walker wrote.)
-
-On the other hand, it only takes a few seconds to run for a 200-species phylogeny (see below).
-
-## constructing a GP PGLMM with lme4: machinery
-
-"All" we need to do is (1) call `(g)lFormula`, with a formula that includes a `(1|phylo)` term, to build the basic (wrong) structure; (2) modify the `reTrms` component of the structure appropriately; (3) go through the rest of the modular procedure for building a (G)LMM.
-
 ```{r phylocode}
 source("lme4_phylo_setup.R")
-```
-
-## glmmTMB fits: nuts and bolts
-
-`glmmTMB` can be deconstructed in a similar way. In fact, we can re-use a lot of the
-machinery. Being able to use `glmmTMB` means we can use a broader range of distributions,
-zero-inflation, etc. (machinery below assumes phylogenetic structure only in the
-conditional distribution). This is also a little clunky, some adjustment on the `glmmTMB`
-side might make it a bit easier.
-
-```{r tmbfit0}
 source("glmmTMB_phylo_setup.R")
@@
 
-## Example
-
-### get data
+### Get data
 

### 2019-04-10 vignette.rmd
done

--- 2019-04-09 phylog.rmd
@@
 title: "Phylogenetic comparative methods in the lme4-verse"
-author: Michael Li and Ben Bolker
-date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
+author: "Michael Li and Ben Bolker"
 output:
-    html_document:
+  pdf_document: default
   mathjax: default
+  html_document: null
 ---
 
-```{r pkgs,message=FALSE, warning=FALSE}
-library(ape)
-library(lme4)
-library(Matrix)
-library(MCMCglmm)
-library(MASS) 
-library(pez)
-library(glmmTMB)
-library(dplyr)
-library(coda)
-library(lattice)
-library(broom)
-library(dotwhisker)
-```
-
-- The standard problem of *phylogenetic comparative methods* is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree - for example, regression analyses of body temperature as a function of body size for animals within a clade
-- More generally, we can frame this in the usual GLMM way as
+The goal of *phylogenetic comparative methods* is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree - for example, regression analyses of body temperature as a function of body size for animals within a clade
+More generally, we can frame this in the usual GLMM way as
 $$
@@
 $$
-where the part that makes it specifically phylogenetic is that $\Sigma$ captures the *phylogenetic correlation*.
+where the part that makes it specifically phylogenetic is that $\Sigma$ captures the *phylogenetic correlation* (PC).
 The PC is the correlation among observations due to relatedness; recently diverged taxa have higher correlation than more anciently diverged taxa.
@@
 
-Various P(G)LMM (phylogenetic [generalized] linear mixed model] approaches have been proposed.
+Various P(G)LMM (phylogenetic [generalized] linear mixed model) approaches have been proposed.
 Many depend on Pagel's lambda transformation, which gives the correlation matrix a particularly simple form (but has been [criticized](http://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html) ...)
 
-An alternative approach is to model the phylogenetic correlation as a *Gaussian process*.
-In particular, suppose that the evolutionary process is a Brownian motion (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on).
+An alternative approach is to model the phylogenetic correlation as a *Gaussian process* (GP).
+In particular, suppose that the evolutionary process is a Brownian motion process (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on).
 In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past.
-If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to do fancy things to impose a correlation structure on the random effects.
+If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to impose a correlation structure on the random effects.
 
@@
 So, for example, we can say that the 'error' value corresponding to tip 1 is
-$\ell_1 b_1 + \ell_2 b_2$,  where $\ell_i$ is the (square root of??) the branch length
+$\ell_1 b_1 + \ell_2 b_2$,  where $\ell_i$ is the square root of the branch length
 and the $b_i$ are independent, homoscedastic Normal variates. Alternately, the $Z$ matrix is
@@
 
-Here's a (probably not very efficent) way to construct the Z matrix.
-(There must be a way to not walk the tree multiple times from every tip ...
+Here's a (probably not very efficient) way to construct the Z matrix.
+(There must be a way to not walk the tree multiple times from every tip ...)
 
@@
 
-(This could benefit from the repeated-entry sparse matrix class that Steve Walker wrote.)
-
 On the other hand, it only takes a few seconds to run for a 200-species phylogeny (see below).
 
-## constructing a GP PGLMM with lme4: machinery
-
-"All" we need to do is (1) call `(g)lFormula`, with a formula that includes a `(1|phylo)` term, to build the basic (wrong) structure; (2) modify the `reTrms` component of the structure appropriately; (3) go through the rest of the modular procedure for building a (G)LMM.
-
-```{r phylocode}
-source("lme4_phylo_setup.R")
-```
-
-## glmmTMB fits: nuts and bolts
-
+To fit in lme4 and glmmTMB, we need random effect terms in the formula that includes a `(...|phylo)`.
+Using the lme4 machinery, it will build the appropriate random effect structure multiplied by the phyloZ matrix.
 `glmmTMB` can be deconstructed in a similar way. In fact, we can re-use a lot of the
@@
 side might make it a bit easier.
-
-```{r tmbfit0}
-source("glmmTMB_phylo_setup.R")
-```
-
-## Example
-
-### get data
-
-From [chapter 11](http://www.mpcm-evolution.com/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of Garamszegi (ed.): data are [here](http://mpcm-evolution.com/OPM/Chapter11_OPM/data.zip)
-
-```{r getdat}
-if (!file.exists("data/phylo.nex")) {
-    dir.create("data")
-    download.file("http://mpcm-evolution.com/OPM/Chapter11_OPM/data.zip",
-                  dest="data/OPM_ch11_data.zip")
-    setwd("data")
-    untar("OPM_ch11_data.zip")
-    setwd("..")
-}
-phylo <- read.nexus("data/phylo.nex")
-```
-
-Compute appropriate $Z$ matrix up front, to measure speed (also
-reusable in a few places below):
-
-```{r calc_phyloZ,cache=TRUE}
-system.time(phyloZ <- phylo.to.Z(phylo))
-```
-
-## Result comparison with Gaussian example in chapter 11
-
-```{r G_results}
-datG <- read.table("data/data_simple.txt",header=TRUE)
-datG$obs <- factor(seq(nrow(datG)))
-datG <- datG %>% mutate(sp = phylo)
-phylo_lmm_fit <- phylo_lmm(phen~cofactor+(1|sp)
-  , data=datG
-  , phylonm = "sp"
-  , phylo = phylo
-  , phyloZ=phyloZ
-  , REML = TRUE
-  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
-)
-
-print(summary(phylo_lmm_fit))
-
-```
-
-## Result comparison with Gaussian with repeated measures example in chapter 11
-
-```{r GR_results}
-datR <- read.table("data/data_repeat.txt",header=TRUE)
-datR$obs <- factor(seq(nrow(datR)))
-datR <- (datR 
-  %>% mutate(sp = species
-        , animals = phylo
-        )
-)
-datR$spec_mean_cf <- sapply(split(datR$cofactor,datR$phylo),mean)[datR$phylo]
-datR$within_spec_cf <- datR$cofactor-datR$spec_mean_cf
-phylo_lmm_fit <- phylo_lmm(phen~spec_mean_cf+within_spec_cf+(1|sp) + (1|animals)
-  , data=datR
-  , phylonm = "sp"
-  , phylo = phylo
-  , phyloZ=phyloZ
-  , REML = FALSE
-  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
-)
-
-print(summary(phylo_lmm_fit))
-
-```
-
-## Result comparison with non-Gaussian example in chapter 11
-
-```{r P_results}
-dat <- read.table("data/data_pois.txt",header=TRUE)
-dat$obs <- factor(seq(nrow(dat)))
-
-dat <- dat %>% mutate(sp=phylo) 
-phylo_glmm_fit <- phylo_glmm(phen_pois~cofactor+(1|sp)+(1|obs)
-  , data=dat
-  , phylonm = "sp"
-  , family = poisson
-  , phylo = phylo
-  , phyloZ=phyloZ
-  , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
-)
-
-summary(phylo_glmm_fit)
-
-```

### 2019-04-10 example.rmd
minor example stuff

@@
   html_document: null
+bibliography: phyloglmm.bib
+urlcolor: blue
 ---
 
-We will reproduce the examples in [chapter 11](http://www.mpcm-evolution.com/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of Garamszegi (ed.) using lme4 and glmmTMB.
+We will reproduce the examples in [chapter 11](http://www.mpcm-evolution.com/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of @garamszegi2014modern using phylogenetic GLMMs based on `lme4` and `glmmTMB`.
 
-To fit we need a random effect in the formula that includes a `(...|phylo)` term, to build the basic random effect structure multipled by the phyloZ matrix (see vignette for more details on phyloZ matrix).
+To fit we need a random effect in the formula that includes a `(...|phylo)` term, to build the basic random effect structure multiplied by the `phyloZ` matrix (see vignette for more details on the `phyloZ` matrix).
 
@@
 library(glmmTMB)
-library(dplyr)
 library(coda)
@@
 
-From [chapter 11](http://www.mpcm-evolution.com/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of Garamszegi (ed.): data are [here](http://mpcm-evolution.com/OPM/Chapter11_OPM/data.zip)
+From [chapter 11](http://www.mpcm-evolution.com/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of @garamszegi2014modern: data are [here](http://mpcm-evolution.com/OPM/Chapter11_OPM/data.zip)
 
@@
 }
-phylo <- read.nexus("data/phylo.nex")
+phylo <- ape::read.nexus("data/phylo.nex")
 ```
@@
 datG <- read.table("data/data_simple.txt",header=TRUE)
+## create observation-level grouping variable
 datG$obs <- factor(seq(nrow(datG)))
-datG <- datG %>% mutate(sp = phylo)
+datG$sp <- datG$phylo
 phylo_lmm_fit <- phylo_lmm(phen~cofactor+(1|sp)
@@
 )
-
 print(summary(phylo_lmm_fit))
-
 ```
@@
 ```
+
+## References

### 2019-04-10 phyloglmm_ms.tex
minor edits

@@
 The simulations presented in this paper are an oversimplification of the real evolutionary processes in nature, but they are no simpler than what researchers typically use.
-In fact, our simulations showed that even simple single predictor--response multi-species model can incorporate multiple levels of complexity (e.g.
-including random slopes, intercept-slope correlations, and the phylogenetic analog).
+In fact, our results showed that even simple single predictor--response multi-species model can incorporate multiple levels of complexity (e.g.
+including random slopes, intercept-slope correlations, and the phylogenetic analog) and fitting
+
 We have fitted models using different platforms that best match the simulation models. 

### 2019-04-10 example.rmd
minor change
push glmmTMB example

@@
 library(lme4)
-library(MCMCglmm)
 library(MASS) 
-library(pez)
 library(glmmTMB)
@@
 library(broom)
-library(dotwhisker)
+library(dplyr)
 ```
@@
 source("lme4_phylo_setup.R")
-source("glmmTMB_phylo_setup.R")
+source("glmmTMBhacked.R")
 ```
@@
 
+Similarily, fitting using glmmTMB:
+
+```{r glmmTMB}
+glmmTMB_fit <- glmmTMBhacked(phen~cofactor+(1|sp)
+    , data=datG
+    , phyloZ=phyloZ
+    , phylonm = "sp"
+    , doFit=TRUE
+    , dispformula = ~1
+    , REML = FALSE
+  ) 
+print(summary(glmmTMB_fit))
+```
+
 ## Result comparison with Gaussian with repeated measures example in chapter 11
@@
 
+glmmTMB_fit2 <- glmmTMBhacked(phen~spec_mean_cf+within_spec_cf+(1|sp) + (1|animals)
+    , data=datR
+    , phyloZ=phyloZ
+    , phylonm = "sp"
+    , doFit=TRUE
+    , dispformula = ~1
+    , REML = FALSE
+  ) 
+print(summary(glmmTMB_fit2))
+
 ```
@@
 
+glmmTMB_fit3 <- glmmTMBhacked(phen_pois~cofactor+(1|sp)+(1|obs)
+    , data=dat
+    , family = poisson
+    , phyloZ=phyloZ
+    , phylonm = "sp"
+    , doFit=TRUE
+    , dispformula = ~1
+    , REML = FALSE
+  ) 
+print(summary(glmmTMB_fit3))
+
 ```

### 2019-04-10 glmmTMBhacked.R

@@ -1,0 +1,276 @@
+glmmTMBhacked <- function (formula, data = NULL, family = gaussian(), ziformula = ~0, 
+                           dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL, 
+                           phylonm = NULL,
+                           na.action = na.fail, se = TRUE, verbose = FALSE, doFit = TRUE, 
+                           control = glmmTMBControl(), REML = FALSE) 
+{
+  call <- mf <- mc <- match.call()
+  if (is.character(family)) {
+    if (family == "beta") {
+      family <- "beta_family"
+      warning("please use ", sQuote("beta_family()"), " rather than ", 
+              sQuote("\"beta\""), " to specify a Beta-distributed response")
+    }
+    family <- get(family, mode = "function", envir = parent.frame())
+  }
+  if (is.function(family)) {
+    family <- family()
+  }
+  if (is.null(family$family)) {
+    print(family)
+    stop("'family' not recognized")
+  }
+  fnames <- names(family)
+  if (!all(c("family", "link") %in% fnames)) 
+    stop("'family' must contain at least 'family' and 'link' components")
+  if (length(miss_comp <- setdiff(c("linkfun", "variance"), 
+                                  fnames)) > 0) {
+    warning("some components missing from ", sQuote("family"), 
+            ": downstream methods may fail")
+  }
+  if (grepl("^quasi", family$family)) 
+    stop("\"quasi\" families cannot be used in glmmTMB")
+  link <- family$link
+  environment(formula) <- parent.frame()
+  call$formula <- mc$formula <- formula
+  if (!is.null(eval(substitute(offset), data, enclos = environment(formula)))) {
+    formula <- addForm0(formula, makeOp(substitute(offset), 
+                                        op = quote(offset)))
+  }
+  environment(ziformula) <- environment(formula)
+  call$ziformula <- ziformula
+  environment(dispformula) <- environment(formula)
+  call$dispformula <- dispformula
+  m <- match(c("data", "subset", "weights", "na.action", "offset"), 
+             names(mf), 0L)
+  mf <- mf[c(1L, m)]
+  mf$drop.unused.levels <- TRUE
+  mf[[1]] <- as.name("model.frame")
[…]

### 2019-04-11 phylotreelike.rmd
pushing tree-like example

---
title: "How Tree-like are the data?"
author: "Michael Li"
date: "April 10, 2019"
output: html_document
---

I am trying to finish the discussion of the phyloglmm paper and focus on writing about why we / ecologist care/ would want to do this.
The terms "phylogenetic variation" or "phylogenetic signal" have always been confusing to me.
The definition that is often used is "Phylogenetic signal is a measure of the statistical dependence among species' trait values due to their phylogenetic relationships". 
After hearing Susan Holmes' talk, I really like how she phrased the question: "Are the data tree-like?"
Instead of looking at the data and ask if they are tree-like or what kind of tree we can build from the observed "trait values/abundance and etc", I am interested in the question (for the sake of finishing phyloglmm): Given a known phylogenetic tree, what kind of data can we generate and it is obvious they look "tree-like"?

I still find the definition vague and I often take it as: "species that are closely related have similar trait values", which is more or less true.
However, the opposite of that statement might not be true.

## Example

Let's assume evolutionary process is a Brownian-motion process, which means that evolution of a trait evolve independently, following a standard Brownian-motion, along each branch of the phylogeny. The phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history.

Consider this phylogenetic tree:

```{r crazy tree, echo=FALSE, message=FALSE}
library(ape)
library(ggplot2)
library(dplyr)
library(tidyr)

nspp <- 3
ncomb <- 1e5
dist <- 1
set.seed(101)
phy <- rtree(n = nspp)
phy$edge.length <- c(50,1,1,1)

Vphy <- vcv(phy)

plot(phy)

```

"Tree-like" data from this extreme example should be "very"easy to identify. 
We expect t1 and t2 to look similar (as stated above) and t3 to look different than t1 and t2.

Simulating one set of data that corresponds to this tree (this is what people usually have, a set of measurements for a bunch of species) look like this. All the examples here do not include observation error (we have observation error in the phyloglmm paper) for simplicity.

```{r simdat, message=FALSE, echo=FALSE}

Y_phy <- MASS::mvrnorm(n=ncomb
  , mu=rep(0,each=nspp)
  , Sigma=Vphy
)

dd <- data.frame(Y_phy, sim=1:ncomb)

dd2 <- dd %>% gather(key = "tip", value = "obs", -sim)

dist <- sqrt(min(diag(Vphy)))

gg <- (ggplot(dd2, aes(x=tip, y=obs, color=tip,group=sim))
  + geom_point()
  + geom_line()
  + theme_bw()
  + scale_color_manual(values=c("black","red","blue"))
  + ylab("observed measurements")
)

print(gg %+% filter(dd2,sim==2453))

```

Ok... maybe there is a bug in the code and MLi doesn't know what he is doing (very common), because it doesn't look like the tree. 
Now let's look at 100 sets of observations from the same tree.

```{r 50trees, echo=FALSE}
print(gg %+% filter(dd2,sim<100))
```

Ok, that looks a lot better! The lines between t1 and t2 are more "horizontal" because we expect high correlation between them. the lines connected to t3 are all over the place because t3 have zero correlation with t1 and t2. 
Going back to the definition, phylogenetic variability is the "potential" to look different from the ancestral node. 
t3 has a very short branch, so it doesn't have large potential to look very different, whereas t1 and t2 have a long branch, can potentially look very different.

```{r hacky half violion code, echo=FALSE}
## flat violin

"%||%" <- function(a, b) {
  if (!is.null(a)) a else b
}

geom_flat_violin <- function(mapping = NULL, data = NULL, stat = "ydensity",
                             position = "dodge", trim = TRUE, scale = "area",
                             show.legend = NA, inherit.aes = TRUE, ...) {
  layer(
    data = data,
    mapping = mapping,
    stat = stat,
    geom = GeomFlatViolin,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      trim = trim,
      scale = scale,
      ...
    )
  )
}

GeomFlatViolin <-
  ggproto("GeomFlatViolin", Geom,
          setup_data = function(data, params) {
            data$width <- data$width %||%
              params$width %||% (resolution(data$x, FALSE) * 0.9)

            # ymin, ymax, xmin, and xmax define the bounding rectangle for each group
            data %>%
              group_by(group) %>%
              mutate(ymin = min(y),
                     ymax = max(y),
                     xmin = x - width / 2,
                     xmax = x)
          },

          draw_group = function(data, panel_scales, coord) {
            # Find the points for the line to go all the way around
            data <- transform(data, 
                              xmaxv = x,
                              xminv = x + violinwidth * (xmin - x))

            # Make sure it's sorted properly to draw the outline
            newdata <- rbind(plyr::arrange(transform(data, x = xminv), y),
                             plyr::arrange(transform(data, x = xmaxv), -y))

            # Close the polygon: set first and last point the same
            # Needed for coord_polar and such
            newdata <- rbind(newdata, newdata[1,])

            ggplot2:::ggname("geom_flat_violin", GeomPolygon$draw_panel(newdata, panel_scales, coord))
          },

          draw_key = draw_key_polygon,

          default_aes = aes(weight = 1, colour = "grey20", fill = "white", size = 0.5,
                            alpha = NA, linetype = "solid"),

          required_aes = c("x", "y")
  )

```

Drawing the potential sampling distributions side by side for each species look like this:

```{r distribution, echo=FALSE}
gg_distr <- (ggplot(dd2, aes(x=tip, y=obs, color=tip,group=tip))
  + geom_flat_violin()
  + theme_bw()
  + scale_color_manual(values=c("black","red","blue"))

)

print(gg_distr)

```

## Star phylogeny

Repeating the same thing for a star phylogeny:

```{r star, echo=FALSE}
phy$edge.length <- c(0,1,1,1)

Vphy <- vcv(phy)

plot(phy)

dist <- sqrt(min(diag(Vphy)))

stardat <- data.frame(obs = rnorm(ncomb*nspp)
  , tip = rep(c("t1","t2","t3"),each=ncomb)
  , sim = rep(1:ncomb,nspp)
)

print(gg %+% filter(stardat,sim<100))
print(gg_distr %+% stardat)

```

## Random tree

Repeating the same thing for a random tree.

```{r random, echo=FALSE}
phy <- rtree(n = nspp)

Vphy <- vcv(phy)

plot(phy)

dist <- sqrt(min(diag(Vphy)))

Y_phy <- MASS::mvrnorm(n=ncomb
  , mu=rep(0,each=nspp)
  , Sigma=Vphy
)

dd <- data.frame(Y_phy, sim=1:ncomb)

dat <- dd %>% gather(key = "tip", value = "obs", -sim)

print(gg %+% filter(dat,sim<100))
print(gg_distr %+% dat)
# 
#   + annotate("rect", xmin=0, xmax=4
#              , ymin=-2*dist
#              , ymax=2*dist , alpha=0.2)

```

## Thoughts and tons of questions

Going back to the question: "Are the data tree-like?" (I don't know) or "Does the data look like my tree?" (I don't know).
We can see in the above examples, we can generate data that are *not* tree-like from the tree. 
The danger is, these *non-tree-like* simulations look like noise and given we did not add observation error, it is almost (I think) impossible to disentangle "phylogenetic noise" vs observation error if observation errors are present.

So what proportion of measurements simulated from the "null/true tree" are "tree-like"?
Another way to think about it is what proportion are "not tree-like"?
A first guess is near the center.
For example:

```{r region, echo=FALSE}

print(gg %+% filter(dd2,sim<100)
  + annotate("rect", xmin=0, xmax=4
             , ymin=-2*1
             , ymax=2*1 , alpha=0.2)
)

print(gg %+% filter(dat,sim<100)
  + annotate("rect", xmin=0, xmax=4
             , ymin=-2*dist
             , ymax=2*dist , alpha=0.2)
)

```

This gray ribbon-band is two standard deviation away from the center using the minimum variance from the phylogenetic variance matrix. 
This looks good at first, but this is also the region with the highest probability evaluated at NVM(0,phylo_var).

### 2019-04-11 phylotreelike.rmd
syncing

@@
 library(tidyr)
+library(mvtnorm)
 
@@
 This looks good at first, but this is also the region with the highest probability evaluated at NVM(0,phylo_var).
+
+## Likelihood surface
+
+```{r likelihood surface, warning=FALSE, echo=FALSE, message=FALSE}
+
+pp <- expand.grid(t1 = seq(-10,10,length=31)
+  , t2 = seq(-10,10,length=31)    ## non-focal parameter
+  , t3 = seq(-3,3,length=9)
+)
+
+ll <- sapply(1:nrow(pp),function(x){mvtnorm::dmvnorm(pp[x,1:3], mean =c(0,0,0), sigma = Vphy, log = TRUE)})
+
+pp$ll <- ll
+gg0 <- (ggplot(pp,aes(t1,t2))
+    + facet_wrap(~t3,labeller=label_both)
+    + geom_raster()
+    + scale_fill_viridis_c()
+#     + scale_x_continuous(expand=c(0,0))
+#     + scale_y_continuous(expand=c(0,0))
+    + theme(panel.spacing=grid::unit(0,"lines"))
+)
+
+gg0L <- gg0+ geom_raster(aes(fill=ll))
+
+print(gg0L)
+
+squash <- function(vals, matchvals) {
+    sapply(vals,function(x) matchvals[which.min(abs(x-matchvals))])
+}
+
+dd_squash <- dd
+dd_squash$t3 <- squash(dd$t3,unique(pp$t3))
+samp_dd <- head(dd_squash,1000)
+samp_dd$ll <- sapply(1:nrow(samp_dd),function(x){mvtnorm::dmvnorm(samp_dd[x,1:3], mean =c(0,0,0), sigma = Vphy, log = TRUE)})
+
+print(gg0L 
+  + geom_point(data=samp_dd, aes(x=t1, y=t2),size=0.1)
+  + xlim(c(-10,10))
+  + ylim(c(-10,10))
+)
+```

### 2019-04-12 phylotreelike.rmd
cool results, need to fill in text

--- 2019-04-11 phylotreelike.rmd
@@
 The definition that is often used is "Phylogenetic signal is a measure of the statistical dependence among species' trait values due to their phylogenetic relationships". 
-After hearing Susan Holmes' talk, I really like how she phrased the question: "Are the data tree-like?"
-Instead of looking at the data and ask if they are tree-like or what kind of tree we can build from the observed "trait values/abundance and etc", I am interested in the question (for the sake of finishing phyloglmm): Given a known phylogenetic tree, what kind of data can we generate and it is obvious they look "tree-like"?
+After hearing Dr. Susan Holmes' talk, I really like how she phrased the question: "Are the data tree-like?"
+Instead of looking at the data and ask if they are tree-like or what kind of tree we can build from the observed "trait values/abundance and etc", I am interested in the question (for the sake of finishing phyloglmm): Given a known phylogenetic tree, what kind of data can we generate and is it obvious that they look "tree-like"?
 
-I still find the definition vague and I often take it as: "species that are closely related have similar trait values", which is more or less true.
-However, the opposite of that statement might not be true.
+I still find the definition vague and I often take it as: "species that are closely related have similar trait values", which is both true and not true to some degree.
 
@@
 
-"Tree-like" data from this extreme example should be "very"easy to identify. 
+"Tree-like" data from this extreme example should be "very" easy to identify. 
 We expect t1 and t2 to look similar (as stated above) and t3 to look different than t1 and t2.
@@
 Now let's look at 100 sets of observations from the same tree.
+Note, each "line" represents a trait that is observed across species/tips.
 
@@
 This looks good at first, but this is also the region with the highest probability evaluated at NVM(0,phylo_var).
+
+## Likelihood surface approach
+
+Note: I am certain I am reinventing the wheel, but let just see how far I can go starting from the basics.
+
+Using the extreme tree presented above, given a set of observed trait (triplets, simulated from the tree), we can compute the likelihood of the given the variance of the phylogenetic tree.
+We need to decide on a null model/tree to compare it with. 
+The most logical null/tree is the tree that preserves the branch lengths (i.e. the diagonals remain the same) and assume there isn't any shared information (i.e. the off diagonals are all zero).
+The null tree looks like this:
+
+```{r null_tree, echo=FALSE}
+
+phy$edge.length <- c(0,51,51,1)
+
+plot(phy)
+Vnull <- vcv(phy)
+
+```
+
+The way to think about "phylogenetic signal" is the amount of correlation we see each pair of species (distance?). 
+The distribution plot above just shows the amount of change each species evolved compared to the ancestral node.
+Thus, the null tree will have the same species level trait distribution, but what we are interested in are the "crosses".
+
+Given a set of traits, we can now calculate the likelihood under the known/true tree (with the covariance) and the null tree (without the covariance).
+We can do a likelihood ratio test and compute the Chisq statistic.
+
+```{r likelihood surface, warning=FALSE, echo=FALSE, message=FALSE}
+
+pp <- expand.grid(t1 = seq(-10,10,length=21)
+  , t2 = seq(-10,10,length=21)    ## non-focal parameter
+  , t3 = seq(-3,3,length=9)
+)
+
+ll <- sapply(1:nrow(pp),function(x){mvtnorm::dmvnorm(pp[x,1:3], mean =c(0,0,0), sigma = Vphy, log = TRUE)})
+
+pp$ll <- ll
+
+Vnull <- diag(diag(Vphy))
+
+c <- qchisq(0.95,df=1)
+
+pp$null <- sapply(1:nrow(pp),function(x){mvtnorm::dmvnorm(pp[x,1:3], mean =c(0,0,0), sigma = Vnull, log = TRUE)})
+
+pp2 <- (pp
+  %>% rowwise()
+  %>% mutate(LR = ll-null
+             # , LRT = max(as.numeric(LR > log(c)),0.5)
+             , LRT = 2*LR > log(c)
+
+             )
+  # %>% arrange(LR)
+)
+
+gg0 <- (ggplot(pp2,aes(t1,t2))
+    + facet_wrap(~t3,labeller=label_both)
+    + geom_raster()
+    + scale_fill_viridis_c()
+#     + scale_x_continuous(expand=c(0,0))
+#     + scale_y_continuous(expand=c(0,0))
+    + theme(panel.spacing=grid::unit(0,"lines"))
+)
+
+gg0L <- gg0+ geom_raster(aes(fill=ll)) + ggtitle("Likelihood surface of the true tree")
+gg0null <- gg0+ geom_raster(aes(fill=null)) + ggtitle("Likelihood surface of the null tree")
+gg0LR <- gg0 + geom_raster(aes(fill=LR)) + ggtitle("Chisq")
+print(gg0L)
+print(gg0null)
+print(gg0LR)
+
+squash <- function(vals, matchvals) {
+    sapply(vals,function(x) matchvals[which.min(abs(x-matchvals))])
+}
+
+print(gg0 
+  + geom_raster(aes(fill=LR,alpha=LRT)) + ggtitle("Tree-like data")
+)
+
+print(mean(pp2$LRT))
+
+```
+
+Approximately 30\% of the traits are are "tree-like" by LRT.
+
+## More Data!
+
+Stats 101, can we see it more clearly if we have more data?
+There are two ways to increase data:
+
+- increase the number of species 
+- observe more traits (more lines)
+
+### Increase more species
+
+Before I simulate a larger tree, I just want to test the less extreme trees.
+
+```{r likelihood surface new tree, warning=FALSE, echo=FALSE, message=FALSE, cache=TRUE}
+
+phy <- rtree(n = nspp)
+plot(phy)
+Vphy <- vcv(phy)
+
+pp <- expand.grid(t1 = seq(-10,10,length=9)
+  , t2 = seq(-10,10,length=9)    ## non-focal parameter
+  , t3 = seq(-10,10,length=9)
+)
+
+ll <- sapply(1:nrow(pp),function(x){mvtnorm::dmvnorm(pp[x,1:3], mean =c(0,0,0), sigma = Vphy, log = TRUE)})
+
+pp$ll <- ll
+
+Vnull <- diag(diag(Vphy))
+
+c <- qchisq(0.95,df=1)
+
+pp$null <- sapply(1:nrow(pp),function(x){mvtnorm::dmvnorm(pp[x,1:3], mean =c(0,0,0), sigma = Vnull, log = TRUE)})
+
+pp2 <- (pp
+  %>% rowwise()
+  %>% mutate(LR = ll-null
+             # , LRT = max(as.numeric(LR > log(c)),0.5)
+             , LRT = 2*LR > log(c)
+
+             )
+  # %>% arrange(LR)
+)
+
+gg0 <- (ggplot(pp2,aes(t1,t2))
+    + facet_wrap(~t3,labeller=label_both)
+    + geom_raster()
+    + scale_fill_viridis_c()
+#     + scale_x_continuous(expand=c(0,0))
+#     + scale_y_continuous(expand=c(0,0))
+    + theme(panel.spacing=grid::unit(0,"lines"))
+)
+
+gg0L <- gg0+ geom_raster(aes(fill=ll)) + ggtitle("Likelihood surface of the true tree")
+gg0null <- gg0+ geom_raster(aes(fill=null)) + ggtitle("Likelihood surface of the null tree")
+gg0LR <- gg0 + geom_raster(aes(fill=LR)) + ggtitle("Chisq")
+print(gg0L)
+print(gg0null)
+print(gg0LR)
+
+squash <- function(vals, matchvals) {
+    sapply(vals,function(x) matchvals[which.min(abs(x-matchvals))])
+}
+
+print(gg0 
+  + geom_raster(aes(fill=LR,alpha=LRT)) + ggtitle("Tree-like data")
+)
+
+print(mean(pp2$LRT))
+
+```
+
+Note: Maybe the extreme examples are not good?
+
+```{r more species, warning=FALSE, echo=FALSE, message=FALSE,cache=TRUE}
+
+ddprop <- data.frame(nspp=3:6, prop=NA)
+ddprop$prop[1] <- mean(pp2$LRT)
+rep_pp <- pp
+for(i in 4:6){
+phy <- rtree(n = i)
+Vphy <- vcv(phy)
+
+temp_dimension <- expand.grid(rep(1,nrow(rep_pp)),new=seq(-10,10,length=9))
+rep_pp <- rbind(rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp)
+newgrid <- cbind(rep_pp,temp_dimension[,"new"])
+colnames(newgrid)[i] <- paste("t",i,sep="_")
+
+newgrid$ll <- sapply(1:nrow(newgrid),function(x){mvtnorm::dmvnorm(newgrid[x,1:i], mean =rep(0,i), sigma = Vphy, log = TRUE)})
+c <- qchisq(0.95,df=i-2)
+
+Vnull <- diag(diag(vcv(phy)))
+
+newgrid$null <- sapply(1:nrow(newgrid),function(x){mvtnorm::dmvnorm(newgrid[x,1:i], mean =rep(0,i), sigma = Vnull, log = TRUE)})
+
+c <- qchisq(0.95,df=i-2)
+
+newgrid2 <- (newgrid
+  %>% rowwise()
+  %>% mutate(LR = ll-null
+             # , LRT = max(as.numeric(LR > log(c)),0.5)
+             , LRT = 2*LR > log(c)
+             )
+  # %>% arrange(LR)
+)
+
+ddprop$prop[i-2] <- mean(newgrid2$LRT)
+print(i)
+rep_pp <- newgrid2[,1:i]
+}
+
+print(ddprop)
+```
+
+Brute forcing the grid is not a good idea and very inefficient. Can we get away with just sampling the tree?
+
+```{r, sampling, echo=FALSE,cache=TRUE}
+ddprop <- data.frame(nspp=3:40, prop=NA)
+rep_pp <- pp
+for(i in 3:40){
+phy <- rtree(n = i)
+Vphy <- vcv(phy)
+
+Y_phy <- MASS::mvrnorm(n=1000
+  , mu=rep(0,each=i)
+  , Sigma=Vphy
+)
+
+dd <- data.frame(Y_phy)
+
+dd$ll <- sapply(1:nrow(dd),function(x){mvtnorm::dmvnorm(dd[x,1:i], mean =rep(0,i), sigma = Vphy, log = TRUE)})
+
+c <- qchisq(0.95,df=i-2)
+
+Vnull <- diag(diag(vcv(phy)))
+
+dd$null <- sapply(1:nrow(dd),function(x){mvtnorm::dmvnorm(dd[x,1:i], mean =rep(0,i), sigma = Vnull, log = TRUE)})
+
+c <- 1.96
+
+dd2 <- (dd
+  %>% rowwise()
+  %>% mutate(LR = ll-null
+             # , LRT = max(as.numeric(LR > log(c)),0.5)
+             , LRT = LR > log(c)
+             )
+  # %>% arrange(LR)
+)
+
+ddprop$prop[i-2] <- mean(dd2$LRT)
+}
+
+plot(ddprop)
+
+```
+
+Wow, it is surprising how fast it levels off.
+
+## Increasing number of traits
+
+The other way to increase data is to collect different types of traits.
+
+```{r increasing traits, echo=FALSE,cache=TRUE}
+
+phy <- rtree(n = 3)
+Vphy <- vcv(phy)
+
+Y_phy <- MASS::mvrnorm(n=1000
+  , mu=rep(0,each=3)
+  , Sigma=Vphy
+)
+
+dd <- data.frame(Y_phy,line=1:nrow(Y_phy))
+
+dd$ll <- sapply(1:nrow(dd),function(x){mvtnorm::dmvnorm(dd[x,1:3], mean =rep(0,3), sigma = Vphy, log = TRUE)})
+
+Vnull <- diag(diag(vcv(phy)))
+
+dd$null <- sapply(1:nrow(dd),function(x){mvtnorm::dmvnorm(dd[x,1:3], mean =rep(0,3), sigma = Vnull, log = TRUE)})
+
+c <- 1.96
+
+dd2 <- (dd
+  %>% mutate(LR = ll-null # line log ratio
+       , sumLL = cumsum(ll)
+       , sumnull = cumsum(null)
+  )
+  %>% rowwise()
+  %>% mutate(LRT = LR > log(c)
+      , sumLR = sumLL - sumnull
+      , sumLRT = sumLR > log(c)
+      )
+  # %>% arrange(LR)
+)
+
+ddsort <- (dd
+  %>% mutate(LR = ll-null) # line log ratio
+  %>% arrange(LR)
+  %>% mutate(sumLL = cumsum(ll)
+       , sumnull = cumsum(null)
+  )
+  %>% rowwise()
+  %>% mutate(LRT = LR > log(c)
+      , sumLR = sumLL - sumnull
+      , sumLRT = sumLR > log(c)
+      )
+  # %>% arrange(LR)
+)
+
+plot(dd2$line,dd2$LRT)
+plot(dd2$line,dd2$sumLRT)
+
+plot(ddsort$LRT)
+plot(ddsort$sumLRT)
+
+```
+
+I think I got a good idea now.

### 2019-04-12 phylotreelike.rmd
syncing
git pushing

@@
 The way to think about "phylogenetic signal" is the amount of correlation we see each pair of species (distance?). 
-The distribution plot above just shows the amount of change each species evolved compared to the ancestral node.
-Thus, the null tree will have the same species level trait distribution, but what we are interested in are the "crosses".
+The distribution plot above just shows the amount of change each species evolve compared to the ancestral node.
+Thus, the null tree will have the same species level trait distribution, but what we are interested in are the correlation (crosses).
 
@@
 We can do a likelihood ratio test and compute the Chisq statistic.
+TODO: What is the "right" statistic? I am using Chisq for the easy examples and using Z when I am lazy.
 
-```{r likelihood surface, warning=FALSE, echo=FALSE, message=FALSE}
+```{r likelihood surface, warning=FALSE, echo=FALSE, message=FALSE, cache=TRUE}
 
-pp <- expand.grid(t1 = seq(-10,10,length=21)
-  , t2 = seq(-10,10,length=21)    ## non-focal parameter
+pp <- expand.grid(t1 = seq(-10,10,length=51)
+  , t2 = seq(-10,10,length=51)    ## non-focal parameter
   , t3 = seq(-3,3,length=9)
@@
 
-Stats 101, can we see it more clearly if we have more data?
+Statistics 101, can we see it more clearly if we have more data?
 There are two ways to increase data:
@@
 
-pp <- expand.grid(t1 = seq(-10,10,length=9)
-  , t2 = seq(-10,10,length=9)    ## non-focal parameter
-  , t3 = seq(-10,10,length=9)
+pp <- expand.grid(t1 = seq(-10,10,length=25)
+  , t2 = seq(-10,10,length=25)    ## non-focal parameter
+  , t3 = seq(-10,10,length=25)
 )
@@
 Note: Maybe the extreme examples are not good? 
+We still see a large area in the likelihood surface that aren't tree-like.
 
-```{r more species, warning=FALSE, echo=FALSE, message=FALSE,cache=TRUE}
+Brute forcing the grid is not a good idea and very inefficient when we increase the dimension. 
+The grid itself is not a good representation of the sampling distribution anyway, thus when calculating the proportion of tree-like, it decreases as we increase the number of species.
 
-ddprop <- data.frame(nspp=3:6, prop=NA)
-ddprop$prop[1] <- mean(pp2$LRT)
-rep_pp <- pp
-for(i in 4:6){
-phy <- rtree(n = i)
-Vphy <- vcv(phy)
-
-temp_dimension <- expand.grid(rep(1,nrow(rep_pp)),new=seq(-10,10,length=9))
-rep_pp <- rbind(rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp)
-newgrid <- cbind(rep_pp,temp_dimension[,"new"])
-colnames(newgrid)[i] <- paste("t",i,sep="_")
-
-newgrid$ll <- sapply(1:nrow(newgrid),function(x){mvtnorm::dmvnorm(newgrid[x,1:i], mean =rep(0,i), sigma = Vphy, log = TRUE)})
-c <- qchisq(0.95,df=i-2)
-
-Vnull <- diag(diag(vcv(phy)))
-
-newgrid$null <- sapply(1:nrow(newgrid),function(x){mvtnorm::dmvnorm(newgrid[x,1:i], mean =rep(0,i), sigma = Vnull, log = TRUE)})
-
-c <- qchisq(0.95,df=i-2)
-
-newgrid2 <- (newgrid
-  %>% rowwise()
-  %>% mutate(LR = ll-null
-             # , LRT = max(as.numeric(LR > log(c)),0.5)
-             , LRT = 2*LR > log(c)
-             )
-  # %>% arrange(LR)
-)
-
-ddprop$prop[i-2] <- mean(newgrid2$LRT)
-print(i)
-rep_pp <- newgrid2[,1:i]
-}
-
-print(ddprop)
-```
-
-Brute forcing the grid is not a good idea and very inefficient. Can we get away with just sampling the tree?
+Can we get away with just sampling traits from the tree?
 
@@
 
-Y_phy <- MASS::mvrnorm(n=1000
+Y_phy <- MASS::mvrnorm(n=10000
   , mu=rep(0,each=i)
@@
 
-Wow, it is surprising how fast it levels off.
+Here I sample 10,000 points from the n-dimension space using a MVN with the phylogenetic variance. 
+It is surprising how fast it levels off.
 
@@
 
-```{r increasing traits, echo=FALSE,cache=TRUE}
+```{r increasing traits, echo=FALSE,cache=TRUE,eval=FALSE}
 
@@
 
-I think I got a good idea now.
+Here, I simulate 1000 sets of traits (lines) and did the LRT for each line. 
+The first plot is consistent with the result before we see a small proportion of tree-like data.
+In order to "increase" more data, I computated did the likelihood ratio test given we saw multiple sets of traits.
+We can easily do this by adding the log likelihood for each set of traits under the NVM under the true tree and null tree and do the corresponding LRT. 
+Note: Order does matter! We need a random order!
+
+We can see that if we combine multiple traits, it increases our chance to detect if the data are tree-like via LRT.
+
+Just a exteme example, what if we combine a bunch of non-tree-like traits? 
+By ordering the

### 2019-04-12 phylotreelike.rmd
need to ggplot the figures

@@
 
-Can we get away with just sampling traits from the tree?
+Can we get away with just sampling traits from the tree? (Yes!)
 
@@
 
+```{r ggprop, echo=FALSE}
+
+ggprop <- (ggplot(ddprop, aes(x=nspp,y=prop))
+  + geom_point()
+  + geom_line()
+  + theme_bw()
+  + ylab("Proportion of Tree-like traits")
+  + xlab("Number of species")
+)
+
+print(ggprop)
+
+```
+
 Here I sample 10,000 points from the n-dimension space using a MVN with the phylogenetic variance. 
 It is surprising how fast it levels off.
+TODO: Maybe add confidence interval bands.
 
@@
 
-```{r increasing traits, echo=FALSE,cache=TRUE,eval=FALSE}
+```{r increasing traits, echo=FALSE,cache=TRUE,warning=FALSE}
 
@@
 
-plot(dd2$line,dd2$LRT)
+# plot(dd2$line,dd2$LRT)
+
+gg1 <- (ggplot(dd2, aes(x=line, y=LRT))
+  + geom_point()
+  + theme_bw()
+  + ylab("Tree like")
+  + xlab("Trait")
+)
+
+print(gg1)
+
+dd3 <- dd2 %>% gather(key = "tip", value = "obs", -c(line,ll,null,LR,sumLL,sumnull,LRT,sumLR,sumLRT))
+
+dist <- sqrt(min(diag(Vphy)))
+
+gg2 <- (ggplot(data=(dd3 %>% filter(line<50)), aes(x=tip, y=obs, color=tip,group=line,alpha=LRT))
+  + geom_point()
+  + geom_line()
+  + theme_bw()
+  + scale_color_manual(values=c("black","red","blue"))
+  + ylab("observed measurements")
+)
+
+print(gg2)
+
+gg3 <- (ggplot(data=(dd3 %>% filter(line<15)), aes(x=tip, y=obs, color=tip,group=line,alpha=sumLRT))
+  + geom_point()
+  + geom_line()
+  + theme_bw()
+  + scale_color_manual(values=c("black","red","blue"))
+  + ylab("observed measurements")
+)
+
+print(gg3)
+
 plot(dd2$line,dd2$sumLRT)
@@
 Just a exteme example, what if we combine a bunch of non-tree-like traits?
-By ordering the

### 2019-04-15 example.rmd
push

--- 2019-04-10 example.rmd
@@
 
+Similarily, fitting using glmmTMB:
+
+```{r glmmTMB}
+glmmTMB_fit <- glmmTMBphylo(phen~cofactor+(1|sp)
+    , data=datG
+    , phyloZ=phyloZ
+    , phylonm = "sp"
+    , doFit=TRUE
+    , dispformula = ~1
+    , REML = FALSE
+  ) 
+print(summary(glmmTMB_fit))
+```
+
 ## Result comparison with Gaussian with repeated measures example in chapter 11
@@
 
+glmmTMB_fit2 <- glmmTMBphylo(phen~spec_mean_cf+within_spec_cf+(1|sp) + (1|animals)
+    , data=datR
+    , phyloZ=phyloZ
+    , phylonm = "sp"
+    , doFit=TRUE
+    , dispformula = ~1
+    , REML = FALSE
+  ) 
+print(summary(glmmTMB_fit2))
+
 ```
@@
 
+glmmTMB_fit3 <- glmmTMBphylo(phen_pois~cofactor+(1|sp)+(1|obs)
+    , data=dat
+    , family = poisson
+    , phyloZ=phyloZ
+    , phylonm = "sp"
+    , doFit=TRUE
+    , dispformula = ~1
+    , REML = FALSE
+  ) 
+print(summary(glmmTMB_fit3))
+
 ```

### 2019-04-15 glmmTMB_phylo_setup.R

@@ -1,28 +1,276 @@
-### TMB Helper functions
-
-modify_TMBstruc <- function(tmbstruc,phylo,phylonm,
-                            phyloZ) {
-  n.edge <- nrow(phylo$edge)
-  ## stuff in tmbstruc (but not within data.tmb) is (maybe) necessary
-  ##  for cosmetics, proper reporting
-  tmbstruc$condList$reTrms <- modify_phylo_retrms(tmbstruc$condList$reTrms, #tmbstruc$condList$reTrms,
-                                                  phylo,phylonm,phyloZ)
-  tmbstruc$condReStruc$`1 + X | sp`$blockReps <- n.edge
-  tmbstruc$condList$Z <- t(tmbstruc$condList$reTrms$Zt)
-  ## data *inside* data.tmb is actually the most critical to allow correct fit
-  tmbstruc$data.tmb$terms$`1 + X | sp`$blockReps <- n.edge
-  tmbstruc$data.tmb$Z <- t(tmbstruc$condList$reTrms$Zt)
-  tmbstruc$parameters$b <- rep(0,ncol(tmbstruc$data.tmb$Z))
-  return(tmbstruc)
+glmmTMBphylo <- function (formula, data = NULL, family = gaussian(), ziformula = ~0, 
+                           dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL, 
+                           phylonm = NULL,
+                           na.action = na.fail, se = TRUE, verbose = FALSE, doFit = TRUE, 
+                           control = glmmTMBControl(), REML = FALSE) 
+{
+  call <- mf <- mc <- match.call()
+  if (is.character(family)) {
+    if (family == "beta") {
+      family <- "beta_family"
+      warning("please use ", sQuote("beta_family()"), " rather than ", 
+              sQuote("\"beta\""), " to specify a Beta-distributed response")
+    }
+    family <- get(family, mode = "function", envir = parent.frame())
+  }
+  if (is.function(family)) {
+    family <- family()
+  }
+  if (is.null(family$family)) {
+    print(family)
+    stop("'family' not recognized")
+  }
+  fnames <- names(family)
+  if (!all(c("family", "link") %in% fnames)) 
+    stop("'family' must contain at least 'family' and 'link' components")
+  if (length(miss_comp <- setdiff(c("linkfun", "variance"), 
+                                  fnames)) > 0) {
+    warning("some components missing from ", sQuote("family"), 
[…]

### 2019-04-15 glmmTMBhacked.R

@@ -1,276 +1,0 @@
-glmmTMBhacked <- function (formula, data = NULL, family = gaussian(), ziformula = ~0, 
-                           dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL, 
-                           phylonm = NULL,
-                           na.action = na.fail, se = TRUE, verbose = FALSE, doFit = TRUE, 
-                           control = glmmTMBControl(), REML = FALSE) 
-{
-  call <- mf <- mc <- match.call()
-  if (is.character(family)) {
-    if (family == "beta") {
-      family <- "beta_family"
-      warning("please use ", sQuote("beta_family()"), " rather than ", 
-              sQuote("\"beta\""), " to specify a Beta-distributed response")
-    }
-    family <- get(family, mode = "function", envir = parent.frame())
-  }
-  if (is.function(family)) {
-    family <- family()
-  }
-  if (is.null(family$family)) {
-    print(family)
-    stop("'family' not recognized")
-  }
-  fnames <- names(family)
-  if (!all(c("family", "link") %in% fnames)) 
-    stop("'family' must contain at least 'family' and 'link' components")
-  if (length(miss_comp <- setdiff(c("linkfun", "variance"), 
-                                  fnames)) > 0) {
-    warning("some components missing from ", sQuote("family"), 
-            ": downstream methods may fail")
-  }
-  if (grepl("^quasi", family$family)) 
-    stop("\"quasi\" families cannot be used in glmmTMB")
-  link <- family$link
-  environment(formula) <- parent.frame()
-  call$formula <- mc$formula <- formula
-  if (!is.null(eval(substitute(offset), data, enclos = environment(formula)))) {
-    formula <- addForm0(formula, makeOp(substitute(offset), 
-                                        op = quote(offset)))
-  }
-  environment(ziformula) <- environment(formula)
-  call$ziformula <- ziformula
-  environment(dispformula) <- environment(formula)
-  call$dispformula <- dispformula
-  m <- match(c("data", "subset", "weights", "na.action", "offset"), 
-             names(mf), 0L)
-  mf <- mf[c(1L, m)]
-  mf$drop.unused.levels <- TRUE
-  mf[[1]] <- as.name("model.frame")
[…]

### 2019-04-30 phyloglmm_ms.tex
minor change

@@
 Researchers use these data to fit complex models describing species occurrence and traits.
-In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is becoming increasingly popular.
+In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or distributions.
 % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
@@
 \ldots}
-PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
-Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
+Given a known phylogenetic tree, PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
+Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the amount of unobserved process of trait evolution (i.e how "tree-like" are the observed data given the phylogenetic tree) \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
-Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently, they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
+Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; and more recently becoming increasingly popular, they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
 While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.

### 2019-04-30 phylotreelike.rmd
minor change

@@
 ```{r, sampling, echo=FALSE,cache=TRUE}
-ddprop <- data.frame(nspp=3:40, prop=NA)
+ddprop <- data.frame(nspp=3:100, prop=NA)
 rep_pp <- pp
-for(i in 3:40){
+for(i in 3:100){
 phy <- rtree(n = i)
@@
 
-Y_phy <- MASS::mvrnorm(n=10000
+Y_phy <- MASS::mvrnorm(n=2000
   , mu=rep(0,each=i)
@@
 
-c <- 1.96
-
 dd2 <- (dd
@@
              # , LRT = max(as.numeric(LR > log(c)),0.5)
-             , LRT = LR > log(c)
+             , LRT = abs(LR) > c/2
              )
@@
 
-c <- 1.96
+c <- qchisq(0.95,df=3-2)
 
@@
   %>% rowwise()
-  %>% mutate(LRT = LR > log(c)
+  %>% mutate(LRT = abs(LR) > c/2
       , sumLR = sumLL - sumnull
-      , sumLRT = sumLR > log(c)
+      , sumLRT = sumLR > c/2
       )
@@
   %>% rowwise()
-  %>% mutate(LRT = LR > log(c)
+  %>% mutate(LRT = LR > c/2
       , sumLR = sumLL - sumnull
-      , sumLRT = sumLR > log(c)
+      , sumLRT = sumLR > c/2
       )

### 2019-05-02 phyloglmm_ms.tex
Do we care about the magic theshold?

@@
 Given a known phylogenetic tree, PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
-Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the amount of unobserved process of trait evolution (i.e how "tree-like" are the observed data given the phylogenetic tree) \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
+Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the amount of unobserved process of trait evolution (i.e how ``tree-like" are the observed data given the phylogenetic tree) \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
@@
 
+\subsection*{How ``tree-like" are the data?}
+
+Multispecies data sets are often very challenging to model using basic statistical frameworks due to the complexity and non-normality data. 
+Most researchers typically use basic regression based methods such as generalized-linear models (GLMs) models and suffocated ecologist would use generalized-linear mixed effect models \citep{bolker2009generalized} to control for different levels of variation.
+Most of the basic statistical techniques assume species are independent, and indentically distributed (if species are not accounted for in the model), which are false from a evolutionary perspective.
+However, traits
+
 \subsection*{Challenges in modelling evolutionary process}
+
+\mli{this part doesn't make sense. Going back to the tree-like example, we can generate two test: cor vs no cor and tip vs noise}
 

### 2019-05-02 phylotreelike.rmd
Do we care about the magic theshold?

@@
 ```{r, sampling, echo=FALSE,cache=TRUE}
-ddprop <- data.frame(nspp=3:100, prop=NA)
+ddprop <- data.frame(nspp=3:50, cortest=NA, nulltest=NA, startest=NA)
 rep_pp <- pp
-for(i in 3:100){
+for(i in 3:50){
 phy <- rtree(n = i)
 Vphy <- vcv(phy)
+
+wrong_tree <- rtree(n = i)
+wrongVphy <- vcv(wrong_tree)
 
@@
 
+dd$wll <- sapply(1:nrow(dd),function(x){mvtnorm::dmvnorm(dd[x,1:i], mean =rep(0,i), sigma = wrongVphy, log = TRUE)})
+
 c <- qchisq(0.95,df=i-2)
@@
 dd$null <- sapply(1:nrow(dd),function(x){mvtnorm::dmvnorm(dd[x,1:i], mean =rep(0,i), sigma = Vnull, log = TRUE)})
+dd$null2 <- sapply(1:nrow(dd),function(x){logLik(lm(unlist(dd[x,1:i])~-1))})
 
@@
              # , LRT = max(as.numeric(LR > log(c)),0.5)
-             , LRT = abs(LR) > c/2
+             , LR2 = ll-null2
+             , LR3 = null-null2
+             , LRT = ll > null # abs(LR) > c/2
+             , LRT1 = ll > wll
+             , LRT2 = ll > null2 # abs(LR2) > (qchisq(0.95,df=(2*i-1))/2)
+             , LRT3 = null > null2 # abs(LR3) > (qchisq(0.95,df=(i-1))/2)
+             , LRT4 = wll > null
+             , LRT5 = wll > null2
              )
@@
 
-ddprop$prop[i-2] <- mean(dd2$LRT)
+ddprop$tree_cor[i-2] <- mean(dd2$LRT)
+ddprop$wrong_tree[i-2] <- mean(dd2$LRT1)
+ddprop$tree_test[i-2] <- mean(dd2$LRT2)
+ddprop$star_test[i-2] <- mean(dd2$LRT3)
+ddprop$wrongtree_cor[i-2] <- mean(dd2$LRT4)
+ddprop$wrongtree_test[i-2] <- mean(dd2$LRT5)
 }
-
-plot(ddprop)
 
@@
 
-ggprop <- (ggplot(ddprop, aes(x=nspp,y=prop))
+ddprop2 <- (ddprop
+  %>% gather(key="LRT", value="proportion", -nspp)
+  %>% na.omit()
+)
+
+ggprop <- (ggplot(ddprop2, aes(x=nspp,y=proportion,color=LRT))
   + geom_point()

### 2019-05-05 phyloglmm_ms.tex
pushing

--- 2019-04-30 phyloglmm_ms.tex
@@
 Researchers use these data to fit complex models describing species occurrence and traits.
-In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or distributions.
+In particular, phylogenetic comparative methods (PCM) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or distributions.
 % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
@@
 \ldots}
-Given a known phylogenetic tree, PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
-Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the amount of unobserved process of trait evolution (i.e how "tree-like" are the observed data given the phylogenetic tree) \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
+Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
+Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs use phylogenetic relationships to estimate the amount of unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
@@
 
-\subsection*{Challenges in modelling evolutionary process}
+\subsection*{Challenges in modeling phylogenetic process in multispecies data}
 
-There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
-% Phylogenetic signal : the magnitude of phylogenetic correlation
-The first challenge is to identify and separate phylogenetic signals from other sources of variation present in the data \citep{blomberg2003testing}.  
-% \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph?
-You go on to talk about the tip-variation/residual-variation confounding issue.
-Maybe replace with something talking more specifically about resolving tip variation?}
-The classic phylogenetic regression uses a statistical model that is equivalent to assuming that phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the trait that is the dependent variable evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model.
+Multispecies data sets are often very challenging to model using basic statistical frameworks due to the complexity and non-normality data. 
+Most researchers typically use basic regression based methods such as generalized-linear models (GLMs) and suffocated ecologist would use generalized-linear mixed effect models \citep{bolker2009generalized} to control for different levels of variation.
+Most of the basic statistical techniques assume species are independent, which are false from a evolutionary perspective.
+The classic phylogenetic regression uses a statistical model that is equivalent to assuming that phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the trait that is the dependent variable evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
 If the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
 More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for additional complexities in the evolutionary process.
-In particular, they can partition residual variation into (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
-However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call intercept-level variation) while the among species residual error is associated with among observations variation within each species.
+In particular, these methods can partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
+However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call tip variation as intercept-level variation) while the among species residual error is associated with among observations variation within each species.
 %\bmb{This is a good point, well expressed.
@@
 
-%% FIXME: second sentence
-The second challenge is flexibly modeling phylogenetic relations across different levels of complexity.
-Classic PCA usually examines the response (trait or distribution) to vary according to the phylogenetic relationship across species, but it is plausible the changes in effect (or levels) of the predictor variables
-may vary according to the phylogenetic relationship across species as well.
+Classic PCMs usually examines the response (trait or distribution) to vary according to the phylogenetic relationship across species, but it is plausible the changes in effect (or levels) of the predictor variables may vary according to the phylogenetic relationship across species as well.
+Suppose we have exmained a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's example \citep{felsenstein1985phylogenies}) using mixed-effect model.
+It is plausible that one group is simply larger than the other (random intercepts) with similar ratio, or alternatively, one group having a larger ratio (random slopes). 
 %\bmb{maybe we can focus on random slopes here?
@@
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
-Our new formulation can be incorporated in any framework that allows for random effect without the need to implement special correlation structures by incorporating phylogenetic structures as part of the mean model.
+Our new formulation can be incorporated in any framework that allows for random effect (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures by incorporating phylogenetic structures as part of the mean model.
 %% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
@@
 
-We begin by describing the classic PCA in a linear regression setting.
+We begin by describing the classic PCM in a linear regression setting.
 Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors $\bX$. 
@@
 \subsection*{Phylogenetic generalized linear mixed model}
-Alternatively, one can use the generalized linear mixed effects modelling (GLMM) framework \citep{lynch1991methods}.
+Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework \citep{lynch1991methods}.
 The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorporate more than one kind of variability through random effects and for non-Gaussian responses.
@@
 
-There are several advantages in modelling phylogenetic correlated traits in the mixed model framework.
+There are several advantages in modeling phylogenetic correlated traits in the mixed model framework.
 First, random intercepts can allow the response trait to vary across groups/levels within grouping predictors (e.g. patch, site, experiment, and etc). 
@@
 Random slopes can allow fixed effects (size of the effect in continuous predictor variables) to vary for each species. 
-Analogous to modelling the response dependency via phylogenetic relationship among species, phylogenetic random slopes can allow fixed effects to account for the phylogenetic relations (i.e.
+Analogous to modeling the response dependency via phylogenetic relationship among species, phylogenetic random slopes can allow fixed effects to account for the phylogenetic relations (i.e.
 similar species will have similar size of the effects on the response than distinct pairs).
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8,page=1]{./git_push/tree.pdf}
+  \includegraphics[scale=0.8,page=1]{./figure/phylotree.png}
   \caption{Three species phylogenetic tree.}
@@
 
-The multi-group full simulation model has five additional random effects (predictor and intercept level random effect of species variance and their covariance, random intercept of site and random intercept of species-site interaction) compared to the single--group full model.
-Predictor and intercept level random effect of species are now applicable in the multi-site model setting because there are multiple (one in each group) observations per species, thus there exist some form of variation at the species level.
+The multi-group full simulation model has five additional random effects (predictor and intercept level random effect of species variance and their covariance, random intercept of group and random intercept of species-group interaction) compared to the single--group full model.
+Predictor and intercept level random effect of species are now applicable in the multi-group model setting because there are multiple (one in each group) observations per species, thus there exist some form of variation at the species level.
 Random intercept of species--group interactions describes whether the species within a group are more closely related on average than occurring by chance; which is equivalent to phylogenetic attraction as described in \cite{helmus2007separating}.
@@
 However, MCMC is often computationally expensive for GLMMs compared to platforms using deterministic optimization.
-\pkg{MCMCglmm} \citep{hadfield2010general} is the most commonly used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid} algorithm \citep{li2018fitting}.
+\pkg{MCMCglmm} \citep{hadfield2010general} is the most commonly used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC) algorithm \citep{duane1987hybrid}.
 
@@
 
-We used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \citep{li2017canfun}. 
+We used our method to reproduce the examples in [chapter 11] of \cite{garamszegi2014modern} using phylogenetic GLMMs based on `lme4` and `glmmTMB` (for more details see example in \cite{michael_li_2019_2639887}).
+We also used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \citep{li2017canfun}. 
 \pkg{lme4} and \pkg{pez} give identical results for fix and random effect estimates, our code is approximately 120 times faster than \pkg{pez}. 
@@
 In fact, our results showed that even simple single predictor--response multi-species model can incorporate multiple levels of complexity (e.g.
-including random slopes, intercept-slope correlations, and the phylogenetic analog) and fitting
+including random slopes, intercept-slope correlations, and the phylogenetic analog) and fitting simpler variations can lead to wrong unreliable estimates of the fixed effects, which is typically the main point of interest.
 
@@
 Models that cannot match the simulation (full) model are expected to perform poorly; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods available.
-Using models that include some form of phylogenetic variation and residual variation is necessary to detect evolutionary process.
+Using models that include some form of phylogenetic variation and residual variation is necessary to detect and account for correlations between species via evolutionary processes.
 Here are a few points researchers should think about when fitting phylogenetic regression models.
 
-\subsection{Process and observation variability and confounding}
+\subsection*{Process and observation variability and confounding}
 
@@
 Pagel's $\lambda$, Blomberg's $K$).
-If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation.
-However, one can summarize multiple measurements to a single measurement (for example, mean or median) per species to avoid confounding residual and tip variations.
+If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation, which is equilvent to Felsenstein's ``worst case".
+There are a few ways to deal with this case.
+For balanced data one can summarize multiple observations to a single measurement (for example, mean or median) per species to avoid confounding residual and tip variations.
+For unbalanced data, one can handle by weighting by number of observations per species, which is equivalent to assuming homoscedasticity using inverse-variance weights. 
+Alternatively, when within-species variance is actually of interest, accounting for within-species variance (i.e.
+adding species-level random effect in our example) can automatically handle multiple observations per species.
 
-\subsection{Random Slopes}
+\subsection*{Random effects}
 
-\bmb{I think you need a lot more here; what does a random slopes model mean?
+% \bmb{I think you need a lot more here; what does a random slopes model mean?
 You have one sentence that says ``people should be thinking about random slopes''.
 Then you talk about correlation (which is a much more subtle/less important issue, I think).}
-In classic GLMM (without phylogenetic relationships), random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+
+In classic GLMM, random effects are often used to handle group (or individual) level variations by allowing for group effects. 
+Random intercepts are often the ``go-to" method to handle these effects because it is controlling the group effects in the reponses level (i.e.
+the interest is the expected value of the response when all predictors are zero/unpresent) by allowing different intercepts.
+Alternatively, random effects can be in more complicated forms such as random slopes (i.e.
+allow fixed effects to vary for each group), and they are much harder to interpret and think about in practice.
+Random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
 Neglecting random slopes can lead to erroneous fixed effect estimates \citep{schielzeth2008conclusions} (parameters of interest) as shown in the simulations above. 
-When analyzing relationships among species between traits and environmental variables, it is entirely plausible that change in effect of predictor variables have on traits are changing in a phylogenetically correlated way.
-In fact, it is easier to interpret random slopes in the evolutionary process context (i.e, are size of different species phylogenetically correlated, or growth rate through time of different species phylogenetically correlated?)
-It may be easier to simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
+When analyzing relationships among species between traits, it is entirely plausible that change in effect of predictor variables have on traits are changing with respect to different grouping of species or even in phylogenetically correlated way (phylogenetic random slopes).
+Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random intercepts are more approriate than random slopes, or including both of them in the model.
+It may be easier to be conservative to include both of them but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
 Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
@@
 
-\subsection{Extension and alternatives}
+\subsection*{Extension and alternatives}
 
@@
 
-\section{Conclusion}
+\section*{Conclusion}
 
@@
 
-\section{Supplements -- translation}
+\section*{Supplements -- translation}
 

### 2019-05-31 phyloglmm_ms.tex
push

@@
 Researchers use these data to fit complex models describing species occurrence and traits.
-In particular, phylogenetic comparative methods (PCM) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or distributions.
+In particular, phylogenetic comparative methods (PCM) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or abundance distributions.
 % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
@@
 \ldots}
-Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
-Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs use phylogenetic relationships to estimate the amount of unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
+Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
+Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
-Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; and more recently becoming increasingly popular, they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
+Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; and more recently they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
 While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
@@
 % but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
-In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models through random effects. 
+In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models. 
 This method allow researchers to easily incorporate various complexities without sacrificing speed.
 
-\subsection*{Challenges in modeling phylogenetic process in multispecies data}
+\subsection*{Challenges in modeling phylogenetic process}
 
-Multispecies data sets are often very challenging to model using basic statistical frameworks due to the complexity and non-normality data. 
+Data sets in ecology and evolution are often challenging to model using basic statistical frameworks due to the non-normality of data. 
+% come back to complexity
 Most researchers typically use basic regression based methods such as generalized-linear models (GLMs) and suffocated ecologist would use generalized-linear mixed effect models \citep{bolker2009generalized} to control for different levels of variation.
-Most of the basic statistical techniques assume species are independent, which are false from a evolutionary perspective.
-The classic phylogenetic regression uses a statistical model that is equivalent to assuming that phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the trait that is the dependent variable evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
+Standard statistical techniques assume species respond independently, which is typically false in evolutionary settings.
+The classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the trait that is the dependent variable evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
 If the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
-More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for additional complexities in the evolutionary process.
-In particular, these methods can partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
-However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call tip variation as intercept-level variation) while the among species residual error is associated with among observations variation within each species.
+More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for different evolutionary processes.
+In particular, these methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
+However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call tip variation as intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
 %\bmb{This is a good point, well expressed.
@@
 Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such observations to partition variability more finely.
-Indeed, many existing methods cannot accommodate multiple observations per species, instead requiring users to collapse the data by taking species averages.
-Figure \ref{fig:variation} illustrates a visual categorization of various forms of variation for three species. 
+Indeed, many existing methods are restricted to single observations per species, requiring users to collapse multi-observation data by taking species averages.
+% Figure \ref{fig:variation} illustrates various forms of variation for three species. 
 %Despite having multiple observations per tip taxon (species), a typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
 % \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
-
-\begin{center}
-\begin{figure}[H]
-  \includegraphics[scale=0.8]{./phylo_diagram.png}
-  \caption{Visual categorization of various forms of phylogenetic variation of three species.
+% 
+% \begin{center}
+% \begin{figure}[H]
+%   \includegraphics[scale=0.8]{./phylo_diagram.png}
+%   \caption{Potential forms of phylogenetic variation of three species.
 Top left shows pure noise (i.e.
 assuming a star phylogeny); top right accounts for phylogeny, but without noise; bottom left accounts both phylogeny and noise at the tip, and finally bottom right shows multiple observations (with noise) at the tips.}
-\label{fig:variation}
-\end{figure}
-\end{center}
+% \label{fig:variation}
+% \end{figure}
+% \end{center}
 
 Classic PCMs usually examines the response (trait or distribution) to vary according to the phylogenetic relationship across species, but it is plausible the changes in effect (or levels) of the predictor variables may vary according to the phylogenetic relationship across species as well.
-Suppose we have exmained a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's example \citep{felsenstein1985phylogenies}) using mixed-effect model.
-It is plausible that one group is simply larger than the other (random intercepts) with similar ratio, or alternatively, one group having a larger ratio (random slopes). 
+Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's example \citep{felsenstein1985phylogenies}) using a mixed-effect model.
+It is plausible that taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes). 
 %\bmb{maybe we can focus on random slopes here?
 Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
-Several recent studies have incorporated phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
+Several recent studies have incorporated phylogenetic signals in different ways and looked at species response to phylogenetic signals with changes in environmental factors.
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach.
@@
 \textbf{Model} & \textbf{Method} & \textbf{Data} & \textbf{Platform} \\ \hline
-\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)\end{tabular}}        & Correlated residual                                                   & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular}                                   \\ \cline{2-4}
-                                                                                                  & \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
-\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single Observation\\ Balanced Design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4}
+\multirow{2}{*}{
+\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)
+\end{tabular}} & Correlated residual & Single observation & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4} 
+& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single observation & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
+\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez} \pkg{phyr}                                                                                            \\ \cline{3-4}
                                                                                                   &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}                                                                                  \\ \hline
-\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced Design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4}
+\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4}
                                                                                                   &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
@@
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
-Our new formulation can be incorporated in any framework that allows for random effect (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures by incorporating phylogenetic structures as part of the mean model.
+Our new formulation can be incorporated in any framework that allows for random effects (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model.
 %% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
-We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
+We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
 \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
@@
 \end{align}
-where $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector which is assumed to be multivariate normally distributed with mean $\bmu$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation matrix.
+where $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector which is assumed to be multivariate normally distributed with mean $\bmu$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation (PC) matrix.
 The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.
@@
 \begin{align}
-\bY & \sim \mathbf{f}(\bmu) + \besp \label{eq:glmm1} \\
+\bY & = \textrm{Distr}(g^{-1}(\bmu), \phi) \label{eq:glmm1} \\
 \bmu & = \bX \bbeta + \bZ \bb  \label{eq:glmm2} \\
 \bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
-\besp & \sim \textrm{N}(0,\sigma^{2}_{\epsilon}) \label{eq:glmm4}
 \end{align}
-where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G-side" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
-Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the generalized mixed model framework with Gaussian responses by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^{2}_{\epsilon} = 0$.
+where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the ``G-side" effect), representing the conditional modes is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\phi$ is a scaling parameter.
+% ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
+% Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the generalized mixed model framework with Gaussian responses by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^{2}_{\epsilon} = 0$.
 
-There are several advantages in modeling phylogenetic correlated traits in the mixed model framework.
+There are several ways to incorporate random variation in the mixed model framework.
 First, random intercepts can allow the response trait to vary across groups/levels within grouping predictors (e.g. patch, site, experiment, and etc). 
 Second, it can also allow the response (trait or distribution) to vary according to the phylogenetic relationships across species (i.e.
-similar species will have similar responses), and between species (tip variation).
+responses may vary between species, but species that are closely related may have similar responses), and between species (tip variation).
 The third type of variation that is often neglected or rarely considered are random slopes.
-Random slopes can allow fixed effects (size of the effect in continuous predictor variables) to vary for each species. 
-Analogous to modeling the response dependency via phylogenetic relationship among species, phylogenetic random slopes can allow fixed effects to account for the phylogenetic relations (i.e.
-similar species will have similar size of the effects on the response than distinct pairs).
+Random slopes can allow fixed effects (the relationship between predictors and responses) to vary for each group. 
+Analogous to modeling the response dependency via phylogenetic relationship among species, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e.
+similar species will have similiar predictor--response relationships).
 
@@
 
-An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
-In particular, suppose that the evolutionary process is a Brownian-motion, which means that evolution of a continuous trait evolve independently, following a standard Brownian-motion, along each branch of the phylogeny.
+Suppose that the evolution follows a Brownian-motion process, which means that continuous traits evolve independently, following a standard Brownian-motion, along each branch of the phylogeny.
 In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
@@
 %% FIXME: double check this
-The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree} and the $\ell_i$ are independent, homoscedastic Normal variates with zero mean and $\sigma^2$ variance (i.e.
-the variance for species 1 is $(\ell_1 \e_1 + \ell_2 \e_2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$).
+The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree} and the $\e_i$ are independent Normal variates with zero mean and $\sigma^2$ variance (i.e.
+the variance for species 1 is $\textrm{E}[(\ell_1 \e_1 + \ell_2 \e_2)^2] = (L_1 + L_2)\sigma^2$).
 The variance of Brownian-motion, is the evolutionary rate, where it is linear in evolution time.
@@
 
-The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and $\bell$ is a vector of square root branch lengths.
+The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and $\bell$ is a vector of square root branch lengths:
 
@@
 We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
-For simplicity, the response variable $\bY$ is conditional normally distributed (i.e. Gaussian distribution with an identity link function) , which is a linear mixed effect model. 
+For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e. Gaussian distribution with an identity link function) , corresponding to a linear mixed effect model. 
 For the first set of simulation, we simulate one observation per species.
@@
 The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance, phylogenetic random slope variance and covariance between phylogenetic random slope and intercept) and residual variance.
-The covariance between phylognetic random intercept and slope measures the correlation of phylogenetic dependency variability of the size in effect and outcome.
+The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic dependency variability of the size in effect and outcome (i.e.
+if the correlation is positive, this represents similar species have similiar relative intercept and slope).
 Predictor and intercept level random effect of species are not applicable in this simulation setting because there is no variation at the species level (i.e. single observation per species).
@@
 
-We extend the simulation model by adding multiple groups where each group has one observation per species. The multi-group full model is the following: 
+We extend the simulation model by adding multiple groups where each group has one observation per species. 
+The multi-group model is a generalization of multiple-site models used in the community ecology literature to model phylogenetic attraction (phylogenetically related species are more likely to occur in the same community).
+The multi-group full model is the following: 
 \begin{align}
@@
 (b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
-\sigma_{\mathrm{phy_{int},phy_{slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
+\Sigma_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
+\Sigma_{\mathrm{phy_{int},phy_{slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
 \end{bmatrix}
@@
 b_{\mathrm{group}} & \sim \textrm{MVN} ( 0 , \Sigma_{\mathrm{group}}) \\
-b_{\mathrm{sp:group}} & \sim \textrm{MVN} (0, \bf{I}_{\mathrm{group}} \otimes \sigma^2_{phy}) \\
+b_{\mathrm{sp:group}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \Sigma_{phy}) \\
 \besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
@@
 The multi-group full simulation model has five additional random effects (predictor and intercept level random effect of species variance and their covariance, random intercept of group and random intercept of species-group interaction) compared to the single--group full model.
-Predictor and intercept level random effect of species are now applicable in the multi-group model setting because there are multiple (one in each group) observations per species, thus there exist some form of variation at the species level.
-Random intercept of species--group interactions describes whether the species within a group are more closely related on average than occurring by chance; which is equivalent to phylogenetic attraction as described in \cite{helmus2007separating}.
+Predictor and intercept level random effect of species are now applicable in the multi-group model setting because there are multiple observations per species, thus there exist some form of variation at the species level.
+Random intercept of species--group interactions describes whether the species within a group are more similar on average than occurring by chance, equivalent to phylogenetic attraction as described in \cite{helmus2007separating}.
 
@@
 
-Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as part of a linear or random effects (Z matrix) in generalized linear model.
+Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as random effects in generalized linear mixed model. 
 We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}.
@@
 % \end{verbatim}
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it simply fits a linear model where the covariance structure between species is equivalent to assuming a Brownian-motion evolutionary process (or other evolutionary process) on the tree instead of treating the residual variation as independent and identically distributed.
-Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow observational residuals and as well as non-Gaussian response variables.
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes a Brownian-motion evolutionary process (or other evolutionary process) on the tree instead of treating the residual variation as independent and identically distributed.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow observational residuals and non-Gaussian response variables.
 Both packages can model other evolutionary processes and different correlation structures (i.e. Pagel's $\lambda$, Blomberg's $K$ etc.), but we restrict our PGLS fits to the simple BM correlation. 
@@
 One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle adding additional uncorrelated random slopes and interactions with intercept.
-Lastly, Bayesian phylogenetic GLMM using Markov chain Monte Carlo (MCMC) are available and can handle all of the cases described above. 
+Lastly, Bayesian phylogenetic GLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
 However, MCMC is often computationally expensive for GLMMs compared to platforms using deterministic optimization.
@@
 \resizebox{\textwidth}{!}{%
-\begin{tabular}{|c|c|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|c|c|c|}
 \hline
-\textbf{Package} & \pkg{nlme} & \pkg{phylolm} & \pkg{lme4}/\pkg{glmmTMB} & \pkg{pez} & \pkg{brms} & \pkg{MCMCglmm} \\ \hline
-Single Group & X & X & X &  & X & X \\ \hline
-\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-Intercept correlation\\ Residual\end{tabular} & X & \begin{tabular}[c]{@{}c@{}}X\\ \\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} \\ \hline
-Multi-group &  &  & X & X & X & X \\ \hline
-\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-intercept correlation\\ Phylo Species-group interaction\\ Species intercept\\ Species Slope \\ Species Slope-intercept correlation\\ Residual\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ \\ X\\ X\\ X\\ X\end{tabular} \\ \hline
+\textbf{Package} & \pkg{nlme} & \pkg{phylolm} & \pkg{lme4}/\pkg{glmmTMB} & \pkg{pez} & \pkg{phyr} & \pkg{brms} & \pkg{MCMCglmm} \\ \hline
+Single Group & X & X & X &  &  & X & X \\ \hline
+\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-Intercept correlation\\ Residual\end{tabular} & X & \begin{tabular}[c]{@{}c@{}}X\\ \\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} \\ \hline
+Multi-group &  &  & X & X & X & X & X \\ \hline
+\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-intercept correlation\\ Phylo Species-group interaction\\ Species intercept\\ Species Slope \\ Species Slope-intercept correlation\\ Residual\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ \\ X\\ X\\ X\\ X\end{tabular} \\ \hline
 \end{tabular}%
@@
 \subsection*{Simulation and evaluations}
-We simulated 100 phylogenetic trees for each sample size (n = 25,50,100 and additional n = 500 for multi-group model) and incorporated simulation parameters (illustrated in Figure \ref{ssplot} and Figure \ref{msplot}) to simulate responses from the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
-Each realization were fitted to all model variant using different platforms.
-Tables \ref{table:platform} shows the parameters that are estimable for each platform. 
+We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and additional n = 500 for multi-group model) and incorporated simulation parameters (illustrated in Figure \ref{ssplot} and Figure \ref{msplot}) to simulate responses from the simulation model.
+Each realization was fitted to all model variant using different platforms.
+Table \ref{table:platform} shows the parameters that are estimable for each platform. 
 We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
@@
 
-We used our method to reproduce the examples in [chapter 11] of \cite{garamszegi2014modern} using phylogenetic GLMMs based on `lme4` and `glmmTMB` (for more details see example in \cite{michael_li_2019_2639887}).
+We used our method to reproduce the examples in [chapter 11] of \cite{garamszegi2014modern} using phylogenetic GLMMs based on ``lme4'' and ``glmmTMB'' (for more details see example in \cite{michael_li_2019_2639887}).
 We also used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \citep{li2017canfun}. 
-\pkg{lme4} and \pkg{pez} give identical results for fix and random effect estimates, our code is approximately 120 times faster than \pkg{pez}. 
+\pkg{lme4} and \pkg{pez} give identical results for fixed and random effect estimates, our code is approximately 120 times faster than \pkg{pez}. 
 Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.
 
-\subsection*{Single Group models simulations}
+\subsection*{Single Group model simulations}
 
@@
 Estimates for fixed effect parameters ($\beta_0$ and $\beta_1$) approach nominal coverage as the number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not for other packages.
-\pkg{brms} have higher coverage than the nominal coverage at 95 \% because the prior distributions for the simulation parameters are cenetered at the true values.
+\pkg{brms} have higher coverage than the nominal coverage at 95 \% because the prior distributions for the simulation parameters are cenetered at the true values \citep{li2018fitting}.
 %% \mli{Is there a figure for this?....Ok, maybe We don't need it.}
@@
 Multi-group model fits on the other hand, are much more similar across platforms (only the more powerful subset of platforms can fit these models at all, and the fitting models are closer to the true simulation model).
-Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides and good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int,slope}}}$) for few number of species (i.e.
+Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int,slope}}}$) for small number of species (i.e.
 n = 25 and 50).
-The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, whereas \pkg{pez} underestimates many parameters.
+The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, whereas \pkg{pez} underestimates many parameters (Figure 6).
 
@@
 For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
-Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 (100) species model takes $\approx$ 9 (20) versus $\approx$ 200 (2000) seconds for \pkg{pez} and \pkg{phyr}.
-\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species for our study, where the computational speed did not scale linearly with sample size 25/50/100.
+Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 (100) species model takes $\approx$ 9 (20) versus $\approx$ 200 (2000) seconds for \pkg{pez} and \pkg{phyr} respectively.
+\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species simulation for our study, where the computational speed did not scale linearly with sample size 25/50/100.
 \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.

### 2019-06-03 phyloglmm_ms.tex
pushing

@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8,page=2]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=1]{./figure/sstime.png}
   \caption{Comparison of speed for all fitting packages variants: layout of the packages as in Figure \ref{ssplot}}
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8,page=3]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=1]{./figure/sscoverage.png}
   \caption{Comparison of coverage probablility for fixed effect parameters.
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8,page=5]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=1]{./figure/mstime.png}
   \caption{Comparison of multi-group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8,page=6]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=1]{./figure/mscoverage.png}
   \caption{Comparison of multi-group model coverage across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
@@
 For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
-Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 (100) species model takes $\approx$ 9 (20) versus $\approx$ 200 (2000) seconds for \pkg{pez} and \pkg{phyr} respectively.
+Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 species model takes $\approx$ 9 versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr} respectively.
 \pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species simulation for our study, where the computational speed did not scale linearly with sample size 25/50/100.
@@
 
-The simulations presented in this paper are an oversimplification of the real evolutionary processes in nature, but they are no simpler than what researchers typically use.
-In fact, our results showed that even simple single predictor--response multi-species model can incorporate multiple levels of complexity (e.g.
-including random slopes, intercept-slope correlations, and the phylogenetic analog) and fitting simpler variations can lead to wrong unreliable estimates of the fixed effects, which is typically the main point of interest.
+The simulations presented in this paper are an oversimplification of evolutionary processes in nature, but they are no simpler than what researchers typically use.
+In fact, our results showed that even simple single predictor--response multi-species model can incorporate multiple levels of complexity.  
+Fitting simpler variations can lead to unreliable estimates of the fixed effects, which are typically the quantities of interest.
 
 We have fitted models using different platforms that best match the simulation models. 
-Models that cannot match the simulation (full) model are expected to perform poorly; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods available.
+Models that cannot match the simulation (full) model perform poorly; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods available.
 Using models that include some form of phylogenetic variation and residual variation is necessary to detect and account for correlations between species via evolutionary processes.
 Here are a few points researchers should think about when fitting phylogenetic regression models.
+
+\subsection*{Incorporating different levels of variation}
+
+% \bmb{I think you need a lot more here; what does a random slopes model mean?
+You have one sentence that says ``people should be thinking about random slopes''.
+Then you talk about correlation (which is a much more subtle/less important issue, I think).}
+
+In classic GLMM, random effects are often use to handle group (or individual) level variation by allowing for group effects. 
+In the simplest experimental design with continuous reponse observations in different levels of a discrete grouping variable, where the response may vary according to different level of the grouping variable, fitting random intercepts are often the ``go-to'' method to handle these variations.
+The random intercept model controls the group effect in the response level by allowing different intercepts for each level of the grouping variable.
+However, random effects can be in more complicated forms as the experimental design increases (i.e observing more variables).  
+For example, imagine observing another continuous explanatory variable in the experimental design above, where the relationship of the response and the new explanatory variable may vary according to the grouping variable.
+In this case, fitting a random slopes model is most appropriate to handle this type of variation by allowing different slopes (relationship between continuous variables) for each level of the goruping variable.
+Random slopes models are much harder to think about in practice and require sufficient amount of data, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+Neglecting random slopes can lead to biased fixed effect estimates with poor coverage and type I errors \citep{schielzeth2008conclusions} (parameters of interest) as shown in the simulations above.
+
+CONT HERE
+
+When analyzing relationships among species between traits, it is entirely plausible that change in effect of predictor variables have on traits are changing with respect to different grouping of species or even in phylogenetically correlated way (phylogenetic random slopes).
+Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random intercepts are more approriate than random slopes, or including both of them in the model.
+It may be easier to be conservative to include both of them but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
+Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
+and Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}?
+% However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
 
@@
 adding species-level random effect in our example) can automatically handle multiple observations per species.
-
-\subsection*{Random effects}
-
-% \bmb{I think you need a lot more here; what does a random slopes model mean?
-You have one sentence that says ``people should be thinking about random slopes''.
-Then you talk about correlation (which is a much more subtle/less important issue, I think).}
-
-In classic GLMM, random effects are often used to handle group (or individual) level variations by allowing for group effects. 
-Random intercepts are often the ``go-to" method to handle these effects because it is controlling the group effects in the reponses level (i.e.
-the interest is the expected value of the response when all predictors are zero/unpresent) by allowing different intercepts.
-Alternatively, random effects can be in more complicated forms such as random slopes (i.e.
-allow fixed effects to vary for each group), and they are much harder to interpret and think about in practice.
-Random slopes models are not always appropriate, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
-Neglecting random slopes can lead to erroneous fixed effect estimates \citep{schielzeth2008conclusions} (parameters of interest) as shown in the simulations above. 
-When analyzing relationships among species between traits, it is entirely plausible that change in effect of predictor variables have on traits are changing with respect to different grouping of species or even in phylogenetically correlated way (phylogenetic random slopes).
-Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random intercepts are more approriate than random slopes, or including both of them in the model.
-It may be easier to be conservative to include both of them but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
-Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
-and Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}?
-% However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
 

### 2019-06-04 new_phylo_setup.R
Autocommit (phyloglmm)

@@ -207,7 +207,7 @@ mkBlist <- function (x, frloc, phylonm,phyloZ, drop.unused.levels = TRUE)
     # sm <- t(sm)
     }
     # bmat <- kronecker(kronecker(diag(lkr),phyloZ),diag(rkr))
-    bmat <- kronecker(diag(lkr),phyloZ)
+    bmat <- kronecker(phyloZ,diag(lkr))
  #   if(nrow(sm) == nspp*nsite){
   #    sm <- t(sm[ff,])
   #  }

### 2019-06-04 plot.R

@@ -156,22 +156,22 @@ gg_mscoverage <- (ggplot(data=ms_coverage
            , ymax=0.95 + 2*sqrt(0.95*0.05/100), alpha=0.2)
 )
 
-print(gg_mscoverage 	+ scale_color_manual(values=c("Gray","Purple","Dark Blue","Orange"))
+print(mscoverage<- gg_mscoverage 	+ scale_color_manual(values=c("Gray","Purple","Dark Blue","Orange"))
 )
 
-aa <- readRDS("datadir/lme4_ms_small_profile.RDS")
-aa_filter <- (aa
-  %>% filter(convergence==0)
-  %>% mutate(platform = "lme4_profile"
-      , size = factor(25)
-      , B0=between(0,B0_lower,B0_upper)
-      , B1 = between(0, B1_lower, B1_upper)
-             )
-  %>% group_by(platform,size)
-  %>% summarise(B0_coverage=mean(B0,na.rm=TRUE)
-                ,B1_coverage = mean(B1,na.rm=TRUE)
-                )
-  %>% gather(key=fixed_parameter, value=coverage, -c(platform,size))
-)
-
-profile_coverage <- (gg_mscoverage %+% rbind(ms_coverage,aa_filter))
+# aa <- readRDS("datadir/lme4_ms_small_profile.RDS")
+# aa_filter <- (aa
+#   %>% filter(convergence==0)
+#   %>% mutate(platform = "lme4_profile"
+#       , size = factor(25)
+#       , B0=between(0,B0_lower,B0_upper)
+#       , B1 = between(0, B1_lower, B1_upper)
+#              )
+#   %>% group_by(platform,size)
+#   %>% summarise(B0_coverage=mean(B0,na.rm=TRUE)
+#                 ,B1_coverage = mean(B1,na.rm=TRUE)
+#                 )
+#   %>% gather(key=fixed_parameter, value=coverage, -c(platform,size))
+# )
+# 
+# profile_coverage <- (gg_mscoverage %+% rbind(ms_coverage,aa_filter))

### 2019-06-04 phyloglmm_ms.tex

@@
 In particular, these methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
-However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call tip variation as intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
+However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation are now both part of the evolutionary process error (which we will call tip variation as intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
 %\bmb{This is a good point, well expressed.
@@
 Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
-Several recent studies have incorporated phylogenetic signals in different ways and looked at species response to phylogenetic signals with changes in environmental factors.
+Several recent studies have incorporated phylogenetic variations in different ways and looked at species response to phylogenetic variations with changes in environmental factors.
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach.
@@
 & \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single observation & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
-\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez} \pkg{phyr}                                                                                            \\ \cline{3-4}
+\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez}, \pkg{phyr}                                                                                            \\ \cline{3-4}
                                                                                                   &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}                                                                                  \\ \hline
@@
 We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
-\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
+\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species, as well as residual variation.
 As well as being flexible, we also expect that our approach will be fast, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
@@
 
-CONT HERE
-
 When analyzing relationships among species between traits, it is entirely plausible that change in effect of predictor variables have on traits are changing with respect to different grouping of species or even in phylogenetically correlated way (phylogenetic random slopes).
-Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random intercepts are more approriate than random slopes, or including both of them in the model.
-It may be easier to be conservative to include both of them but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
-Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
-and Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}?
-% However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
-
-\subsection*{Process and observation variability and confounding}
-
-% \bmb{explain: classical scenario for PIC is when we have species-level data (i.e., species averages or ``typical'' values).
-What do we do when we have multiple observations per species?
-Why can't we just average?
-(cf.
-\cite{murtaugh2007simplicity}) with a nested, balanced design, averaging by group and doing 1-way ANOVA gets the same answer for fixed effects as LMMs).
-\textbf{answers}: (1) unbalanced data (can handle by weighting by $n$, equivalent to assuming homoscedasticity and using inverse-variance weights); (2) non-Gaussian responses (averaging doesn't make sense) [can we use offsets to handle this case?
-don't know]; (3) maybe for equivalent of randomized block designs?
-(4) when within-species variance is actually of interest}
-Process and observation errors are highly confounded in phylogenetic regression if it is not handled correctly. 
-For single measurement per species model, any method that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e.
-Pagel's $\lambda$, Blomberg's $K$).
-If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation, which is equilvent to Felsenstein's ``worst case".
+Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random effects are more approriate than simple grouping random effects, or including all of them in the model.
+However, the model complexity depends on the experimental design and whether there are enough data to estimate these different levels of variation.
+Some of these variations can be highly confounded in phylogenetic regression if it is not handled correctly or if there are not enough data.
+For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ (where one parameter quantifying phylogenetic variation and one for the residual variation) will be sufficient to a first approximation.
+If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic variation by confounding tip variation with residual variation, which is equivalent to Felsenstein's ``worst case".
 There are a few ways to deal with this case.
@@
 
+It may be easier to be conservative to include both (phylogenetic random effects and grouping random effects) of them but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
+Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
+and Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}?
+% However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
+
 \subsection*{Extension and alternatives}
 
-Our analysis covers the classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models).
-Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-BM evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \cite{butler2004phylogenetic}), Bayesian approaches \cite{hadfield2010general}.
-More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis.
-This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as \pkg{MCMCglmm} \citep{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
-However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e.
+We have presented the range of classical phylogenetic comparative methods (i.e.
+phylogenetic least squares, linear and mixed models) to fit data that incorporates different levels of phylogenetic variation.
+Even within this scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-Brownian evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \citep{butler2004phylogenetic}), and Bayesian approaches \citep{hadfield2010general}.
+However, the simple approach we developed here offers a more efficient and mathematical equivalent way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models.
+This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using any platforms that supports independent latent variables such as \pkg{MCMCglmm} \citep{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
+However, in principle just like GLMM and most statistical models, users should be aware of the constraints of the amount of data they have relative to the complexity of the model they want to fit (i.e.
 What should we do when we don't have enough data?
-Should we use a complex model and overfit \citep{barr2013random}, the right balance of complexity and data \citep{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
-More importantly, this implementation in \pkg{lme4} allows users to fit phylogenetic mixed models to the fullest ( that other frequentist platforms lack of (i.e.
-large data, unbalanced species observations, complex random effects) and explore new ideas.
+Should we use a complex model and overfit \citep{barr2013random}, the right balance of complexity and data \citep{baayen2008mixed}, singular fits or possibility of Bayesian approaches \citep{hadfield2010mcmc}).
+More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.
 
@@
 We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical PCM to simulated data. 
-First, this new approach is magnitudes faster compared to existing phylogenetic mixed models without losing robustness/accuracy and can easily combined with any statistically mixed model framework. 
-Last and most importantly, it is more flexible in fitting large phylogenies, large volume of data, unbalanced data sets, and complex random effects such as slope correlations.
+Our approach is orders of magnitud faster than existing phylogenetic mixed models and it can easily combined with any statistical mixed model framework. 
+It is more flexible in fitting large phylogenies, large volumes of data, unbalanced data sets, and complex random effects such as slope correlations.
 

### 2019-06-04 phyloglmm_ms.tex
sending over to ben

@@
 \begin{figure}[H]
-  \includegraphics[scale=1]{./figure/sstime.png}
+  \includegraphics[scale=0.8]{./figure/sstime.pdf}
   \caption{Comparison of speed for all fitting packages variants: layout of the packages as in Figure \ref{ssplot}}
@@
 \begin{figure}[H]
-  \includegraphics[scale=1]{./figure/sscoverage.png}
+  \includegraphics[scale=0.8]{./figure/sscoverage.pdf}
   \caption{Comparison of coverage probablility for fixed effect parameters.
@@
 \begin{figure}[H]
-  \includegraphics[scale=1]{./figure/mstime.png}
+  \includegraphics[scale=0.8]{./figure/mstime.pdf}
   \caption{Comparison of multi-group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
@@
 \begin{figure}[H]
-  \includegraphics[scale=1]{./figure/mscoverage.png}
+  \includegraphics[scale=0.8]{./figure/mscoverage.pdf}
   \caption{Comparison of multi-group model coverage across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
@@
 n = 25 and 50).
-The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, whereas \pkg{pez} underestimates many parameters (Figure 6).
+The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, whereas \pkg{pez} and \pkg{phyr} underestimates many parameters (Figure \ref{msplot}).
 

### 2019-07-02 phyloglmm_ms.tex
sleeping now, con't in the am

@@
 \usepackage{blkarray}
-\usepackage{amsmath}
+\usepackage{bm}
+\usepackage{amsmath,empheq}
 \usepackage{alltt}
@@
 \ldots}
-Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
+Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signals (a measure of dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
 Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
-Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; and more recently they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
+Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
 While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
-In such cases, researchers must find ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
+In such cases, researchers typically search for ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
 %% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
@@
 In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models. 
-This method allow researchers to easily incorporate various complexities without sacrificing speed.
+This method allows researchers to easily incorporate various complexities without sacrificing speed.
 
@@
 
-Data sets in ecology and evolution are often challenging to model using basic statistical frameworks due to the non-normality of data. 
 % come back to complexity
-Most researchers typically use basic regression based methods such as generalized-linear models (GLMs) and suffocated ecologist would use generalized-linear mixed effect models \citep{bolker2009generalized} to control for different levels of variation.
-Standard statistical techniques assume species respond independently, which is typically false in evolutionary settings.
+Most researchers use regression based methods such as generalized-linear models (GLMs) (a regression approach that allows for the non-normality of the data) and sophisticated ecologist would use generalized-linear mixed effect models \citep{bolker2009generalized} to control for different levels of variation.
+However, standard statistical techniques assume species respond independently, which is typically false in evolutionary settings.
 The classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the trait that is the dependent variable evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
 If the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
-More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for different evolutionary processes.
-In particular, these methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
-However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation are now both part of the evolutionary process error (which we will call tip variation as intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
+More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, heritability $h^2$ \citep{housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for different evolutionary processes.
+In particular, these methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
+However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation are now both part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
 %\bmb{This is a good point, well expressed.
@@
 Indeed, many existing methods are restricted to single observations per species, requiring users to collapse multi-observation data by taking species averages.
+%% REMEMBER TO REVISIT IN DISCUSSION
+
 % Figure \ref{fig:variation} illustrates various forms of variation for three species. 
@@
 
-Classic PCMs usually examines the response (trait or distribution) to vary according to the phylogenetic relationship across species, but it is plausible the changes in effect (or levels) of the predictor variables may vary according to the phylogenetic relationship across species as well.
-Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's example \citep{felsenstein1985phylogenies}) using a mixed-effect model.
-It is plausible that taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes). 
+Classic PCMs usually examine the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects (or levels) of the predictor variables may evolve according to the phylogenetic relationship across species as well.
+Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example  using a mixed-effect model.
+Taxonomic groups with similar body sizes may vary in overall brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes). 
 %\bmb{maybe we can focus on random slopes here?
@@
 Several recent studies have incorporated phylogenetic variations in different ways and looked at species response to phylogenetic variations with changes in environmental factors.
-For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach.
-The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
-Table \ref{table:model} provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.
+For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach.
+The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
+Table \ref{table:model} summerises types of platforms, data constraints and provides model complexities for phylogenetic comparative analysis.
 
@@
 We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
-\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species, as well as residual variation.
-As well as being flexible, we also expect that our approach will be fast, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
+\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
+We expect that our approach will be fast ws well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
+%% TODO LAST LINE SEE PAGE 5
+
 %% FIXME: move to methods/simulations
@@
 
-\begin{align}
-\bmu & = \bX \bbeta  \label{eq:gls1} \\ 
-\bY & \sim \textrm{MVN}(\bmu,\sigma^{2} \bC), \label{eq:gls2}
-\end{align}
-where $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector which is assumed to be multivariate normally distributed with mean $\bmu$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation (PC) matrix.
+\begin{equation}
+\begin{aligned}
+\bY & = \bX \bbeta + \besp  \\
+\besp & \sim \textrm{MVN}(0,\sigma^{2} \bC), 
+\label{eq:gls}
+\end{aligned}
+\end{equation}
+where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times (m)$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation (PC) matrix.
 The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.
@@
 The typical GLMM has the form:
-\begin{align}
-\bY & = \textrm{Distr}(g^{-1}(\bmu), \phi) \label{eq:glmm1} \\
-\bmu & = \bX \bbeta + \bZ \bb  \label{eq:glmm2} \\
-\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
-\end{align}
-where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the ``G-side" effect), representing the conditional modes is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\phi$ is a scaling parameter.
+\begin{equation}
+\begin{aligned}
+\bY & = \bm{D}(g^{-1}(\bmu), \phi) \\
+\bmu & = \bX \bbeta + \bZ \bb  \\
+\bb & \sim \textrm{MVN}(0, \bSigma(\theta))  \\
+\label{eq:glmm}
+\end{aligned}
+\end{equation}
+where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the ``G-side" effect) representing the conditional modes, is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$ and $\phi$ is a scale parameter for the conditional distribution $\bm{D}$.
 % ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
@@
 There are several ways to incorporate random variation in the mixed model framework.
-First, random intercepts can allow the response trait to vary across groups/levels within grouping predictors (e.g. patch, site, experiment, and etc). 
-Second, it can also allow the response (trait or distribution) to vary according to the phylogenetic relationships across species (i.e.
-responses may vary between species, but species that are closely related may have similar responses), and between species (tip variation).
+First, random intercepts can allow the response trait to vary across groups within grouping variables (e.g. patches, sites, or experiments). 
+Second, random intercepts can also allow the response (trait or distribution) to vary between species (tip variation) or phylogenetically across species (i.e.
+species that are closely related tend to have similar responses).
 The third type of variation that is often neglected or rarely considered are random slopes.
 Random slopes can allow fixed effects (the relationship between predictors and responses) to vary for each group. 
-Analogous to modeling the response dependency via phylogenetic relationship among species, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e.
+Analogous to modeling correlations among responses via phylogenetic relationship among species, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e.
 similar species will have similiar predictor--response relationships).
@@
 
-Suppose that the evolution follows a Brownian-motion process, which means that continuous traits evolve independently, following a standard Brownian-motion, along each branch of the phylogeny.
+Suppose that the evolution follows a Brownian-motion process, i.e. continuous traits evolve independently, following a standard Brownian-motion, along each branch of the phylogeny.
 In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
-Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$, is equivalent to imposing a correlation $\bC$.
+Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$ is equivalent to imposing a correlation $\bC$.
 For example, for the phylogeny in figure \ref{fig:tree}, the corresponding $\bS$ takes the form:
@@
   \includegraphics[scale=0.8,page=1]{./figure/phylotree.png}
-  \caption{Three species phylogenetic tree.}
+  \caption{Three-species phylogenetic tree.}
 \label{fig:tree}
@@
 \begin{block}{c(cccc)}
-  t_1 & \ell_1 \e_1 & \ell_2 \e_2 & 0           & 0 \\
-  t_2 & \ell_1 \e_1 &  0          & \ell_3 \e_3 & 0 \\
-  t_3 & 0           &  0          & 0           & \ell_4 \e_4 \\
+  t_1 & \ell_1 & \ell_2 & 0           & 0 \\
+  t_2 & \ell_1 &  0          & \ell_3 & 0 \\
+  t_3 & 0           &  0          & 0           & \ell_4 \\
 \end{block}
 \end{blockarray}
+\hspace{0.5cm}
+\begin{blockarray}{(c)}
+  \e_1 \\
+  \e_2 \\
+  \e_3 \\
+  \e_4 \\
+\end{blockarray}
  \]
+% \begin{pmatrix}
+% \e_1 \\
+% \e_2 \\
+% \e_3 \\
+% \e_4
+% \end{pmatrix}
 
 %% FIXME: double check this
-The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree} and the $\e_i$ are independent Normal variates with zero mean and $\sigma^2$ variance (i.e.
+The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree}, and the $\e_i$ are independent Normal variates with zero mean and $\sigma^2$ evoulationary variance (i.e.
 the variance for species 1 is $\textrm{E}[(\ell_1 \e_1 + \ell_2 \e_2)^2] = (L_1 + L_2)\sigma^2$).
-The variance of Brownian-motion, is the evolutionary rate, where it is linear in evolution time.
+% The variance of Brownian-motion, is the evolutionary rate, where it is linear in evolution time.
 
@@
 
-The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and $\bell$ is a vector of square root branch lengths:
+The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and a vector $\bell$ of square roots of branch lengths:
 
@@
 
-The random-effect model matrix $\bZ$ can be decomposed into term-wise model matrices $\bZ_{i}$ as described in \citet{bates2015fitting}.
+In general, the random-effect model matrix $\bZ$ for a GLMM can be decomposed into term-wise model matrices $\bZ_{i}$ as described in \citet{bates2015fitting}.
 Analogous to the procedure described in \citet{bates2015fitting}, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is
@@
 \begin{equation}
-\bZ^{C}_{i} = (\bS^{T}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
+\bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top}, \label{eq:ZC}
 \end{equation}
-where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; and $\bX_{i}$ is the $n \times p_{i}$ raw random effects model matrix.
+where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions};, $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; and $\bX_{i}$ is the $n \times p_{i}$ raw random-effects model matrix.
 
@@
 \begin{align}
-\bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
+\bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top} =
 \left[
@@
 \end{bmatrix} 
-\right]^T
+\right]^{\top}
 \\
@@
 
-We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
+We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm}) with a single response variable $\bY$ and a single continuous predictor variable $\bX$ ($\bX$ is a $n \times 2$ matrix) for $n$ = 25, 50, and 100 species.
 For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e. Gaussian distribution with an identity link function) , corresponding to a linear mixed effect model. 
@@
 % For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
-The full simulation model is the following:
+Thus, the full simulation model is as follows:
 \begin{align}

### 2019-07-02 phyloglmm_ms.tex
con't in am
everything above results are done

@@
 Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework \citep{lynch1991methods}.
-The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorporate more than one kind of variability through random effects and for non-Gaussian responses.
+The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorporate more than one kind of variability through random effects and allows for non-Gaussian responses.
 The typical GLMM has the form:
@@
 
-\begin{align}
+\[
 \bS_{ind} = \begin{bmatrix}
@@
 \end{bmatrix} .
-\label{eq:S}
-\end{align}
+\]
 $\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
@@
 
-\begin{align}
+\[
 \bX = \begin{bmatrix}
@@
 \end{bmatrix} 
-\end{align}
+\]
 then the term-wise phylogenetic random effects model matrix is,
 
-\begin{align}
+\[
 \bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top} =
@@
 \end{bmatrix}.
-\end{align}
+\]
 
@@
 
-We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm}) with a single response variable $\bY$ and a single continuous predictor variable $\bX$ ($\bX$ is a $n \times 2$ matrix) for $n$ = 25, 50, and 100 species.
-For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e. Gaussian distribution with an identity link function) , corresponding to a linear mixed effect model. 
-For the first set of simulation, we simulate one observation per species.
+We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm}) with a single response variable $\bY$ and a single continuous predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
+For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e.
+$\bm{D}$ is a Gaussian distribution and $g$ is the identity link function) , corresponding to a linear mixed effect model.
+For the first set of simulations, we simulate one observation per species.
 % For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
 Thus, the full simulation model is as follows:
-\begin{align}
-\bY & = (\beta_0 + b_{\mathrm{phy_{int}}}) + (\beta_1 + b_{\mathrm{phy_{slope}}}) X + \besp  \label{eq:ss_sim}\\\
+\begin{equation}
+\begin{aligned}
+\bY & = (\beta_0 + b_{\mathrm{phy_{int}}}) + (\beta_1 + b_{\mathrm{phy_{slope}}}) \bm{t} + \besp \\\
 (b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
-\sigma_{\mathrm{phy_{int},phy_{slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
+\Sigma_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
+\Sigma_{\mathrm{phy_{int},phy_{slope}}} & \Sigma_{\mathrm{phy_{slope}}}
 \end{bmatrix} 
@@
 \besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2) .
-\end{align}
-The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance, phylogenetic random slope variance and covariance between phylogenetic random slope and intercept) and residual variance.
-The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic dependency variability of the size in effect and outcome (i.e.
-if the correlation is positive, this represents similar species have similiar relative intercept and slope).
-Predictor and intercept level random effect of species are not applicable in this simulation setting because there is no variation at the species level (i.e. single observation per species).
+\label{eq:single_glmm}
+\end{aligned}
+\end{equation}
+The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random slope and intercept $\Sigma_{\mathrm{phy_{int},phy_{slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).
+The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic dependency variability in regression effect ($b_{\mathrm{phy_{slope}}}$) and response ($b_{\mathrm{phy_{int}}}$); i.e.
+if a positive correlation indicates that similar species have similiar relative intercepts and slopes.
+Predictor-level and intercept-level random effects of species are not applicable in this simulation setting because there is no variation at the species level (i.e.
+there is only a single observation per species).
 
@@
 We extend the simulation model by adding multiple groups where each group has one observation per species. 
-The multi-group model is a generalization of multiple-site models used in the community ecology literature to model phylogenetic attraction (phylogenetically related species are more likely to occur in the same community).
-The multi-group full model is the following: 
-\begin{align}
-\bY & = (\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}) + (\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}) X + b_{\mathrm{sp:group}} + \besp \label{eq:ms_sim}\\
+The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction (measuing the tendency of phylogenetically related species to occur in the same community).
+The multi-group full model is as follows: 
+\begin{equation}
+\begin{aligned}
+\bY & = (\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}) + (\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}) \bm{t} + b_{\mathrm{sp:group}} + \besp \\
 (b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
 \Sigma_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
-\Sigma_{\mathrm{phy_{int},phy_{slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
+\Sigma_{\mathrm{phy_{int},phy_{slope}}} & \Sigma_{\mathrm{phy_{slope}}}
 \end{bmatrix}
@@
 \besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
-\end{align}
-where $\otimes$ is the Kronecker product.
+\label{eq:multiple_glmm}
+\end{aligned}
+\end{equation}
+where $\textrm{I}_{\textrm{group}}$ is a indicator matrix of groups; and $\otimes$ is the Kronecker product.
 
-The multi-group full simulation model has five additional random effects (predictor and intercept level random effect of species variance and their covariance, random intercept of group and random intercept of species-group interaction) compared to the single--group full model.
-Predictor and intercept level random effect of species are now applicable in the multi-group model setting because there are multiple observations per species, thus there exist some form of variation at the species level.
-Random intercept of species--group interactions describes whether the species within a group are more similar on average than occurring by chance, equivalent to phylogenetic attraction as described in \cite{helmus2007separating}.
+The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int},sp_{slope}}}$), random intercept of group ($b_{\mathrm{group}}$)and random intercept of species-group interaction ($b_{\mathrm{sp:group}}$)) compared to the single--group full model.
+Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species, thus we can quantify variation within-species distinct from residual variation.
+Random intercept of species--group interactions describes whether the species within a group are more similar on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
 
@@
 
-Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as random effects in generalized linear mixed model. 
+Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as random effects in a generalized linear mixed model.
 We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}.
@@
 % \end{verbatim}
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes a Brownian-motion evolutionary process (or other evolutionary process) on the tree instead of treating the residual variation as independent and identically distributed.
-Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow observational residuals and non-Gaussian response variables.
-Both packages can model other evolutionary processes and different correlation structures (i.e. Pagel's $\lambda$, Blomberg's $K$ etc.), but we restrict our PGLS fits to the simple BM correlation. 
-Neither PGLS nor PGLM can handle multiple observations within a species or random slopes.
-One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle adding additional uncorrelated random slopes and interactions with intercept.
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes a evolutionary process on the tree typically Brownian-motion, but other processes can be used instead of treating the residual variation as independent.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow for both phylogenetic and residual variation, and non-Gaussian response variables.
+Both \code{gls} and \code{phylolm} can model other evolutionary processes and different correlation structures (e.g.
+Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
+Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
+One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle additional random slopes ($\Sigma_{\mathrm{phy_{slope}}}$) and random intercept of species-group interactions ($b_{\mathrm{sp:group}}$) but cannot handle phylogenetic random slope-intercept covariance ($\Sigma_{\mathrm{phy_{int},phy_{slope}}}$).
 Lastly, Bayesian phylogenetic GLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
 However, MCMC is often computationally expensive for GLMMs compared to platforms using deterministic optimization.
-\pkg{MCMCglmm} \citep{hadfield2010general} is the most commonly used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC) algorithm \citep{duane1987hybrid}.
+\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC)\citep{duane1987hybrid}.
 
@@
 \subsection*{Simulation and evaluations}
-We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and additional n = 500 for multi-group model) and incorporated simulation parameters (illustrated in Figure \ref{ssplot} and Figure \ref{msplot}) to simulate responses from the simulation model.
-Each realization was fitted to all model variant using different platforms.
+We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to simulate responses from the simulation model (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
+Simulation parameters are illustrated in Figure \ref{ssplot} and Figure \ref{msplot}.
+Each realization was fitted to all model variants using a range of different platforms.
 Table \ref{table:platform} shows the parameters that are estimable for each platform. 
 We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
-For Bayesian GLMM, we combined two criteria to assess convergence: we required a value of the Gelman and Rubin statistic less than 1.1 and an effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0} \ \textrm{and} \ \beta_{1}$) for each replication.
+For Bayesian GLMM, to assess convergence we required a value of the Gelman-Rubin statistic less than the recommended threshold 1.1.
+We included another convergence criteria (due to the recent critique of the Gelman-Rubin recommended threshold \citep{vats2018revisiting}) of effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0} \ \textrm{and} \ \beta_{1}$) for each replication \citep{vehtari2019rank}.
 For each replication we sample two chains starting with 10000 iterations. 
 %% TODO MAKE SURE THIS IS CORRECT!
-We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms -- this gives us an idea of bias and variation (i.e.
+We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e.
 quality of the point estimate).
-Then we look at frequentist coverage, to assess quality of confidence intervals.
+We then examine at frequentist coverage, to assess the quality of the confidence intervals.
 Coverage refers to the frequency with which the computed confidence intervals include the true values of parameters.
-We used 90\% (Wald and profile) confidence intervals and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
+We used 95\% (Wald and profile) confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
 We also compare computational speed between different platforms to evaluate the efficiency of the platforms and methods.
 
+%% CONT HERE, taking a break
 \section*{Results}

### 2019-07-03 phyloglmm_ms.tex
need to fix up ref: (Rhat) citation
fixing symbols to match easy ggplot symbols

@@
 (b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\Sigma_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
-\Sigma_{\mathrm{phy_{int},phy_{slope}}} & \Sigma_{\mathrm{phy_{slope}}}
+\Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
+\Sigma_{\mathrm{phy_{int-slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
 \end{bmatrix} 
@@
 \end{equation}
-The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random slope and intercept $\Sigma_{\mathrm{phy_{int},phy_{slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).
+The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma^2_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma^2_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random slope and intercept $\Sigma_{\mathrm{phy_{int-slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).
 The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic dependency variability in regression effect ($b_{\mathrm{phy_{slope}}}$) and response ($b_{\mathrm{phy_{int}}}$); i.e.
@@
 (b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\Sigma_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
-\Sigma_{\mathrm{phy_{int},phy_{slope}}} & \Sigma_{\mathrm{phy_{slope}}}
+\Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
+\Sigma_{\mathrm{phy_{int-slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
 \end{bmatrix}
@@
 (b_{\mathrm{sp_{int}}}, b_{\mathrm{sp_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\sigma^2_{\mathrm{sp_{int}}} & \sigma_{\mathrm{sp_{int},sp_{slope}}} \\ 
-\sigma_{\mathrm{sp_{int},sp_{slope}}} & \sigma^2_{\mathrm{sp_{slope}}}
+\sigma^2_{\mathrm{sp_{int}}} & \sigma_{\mathrm{sp_{int-slope}}} \\ 
+\sigma_{\mathrm{sp_{int-slope}}} & \sigma^2_{\mathrm{sp_{slope}}}
 \end{bmatrix}
 \right) \\
-b_{\mathrm{group}} & \sim \textrm{MVN} ( 0 , \Sigma_{\mathrm{group}}) \\
-b_{\mathrm{sp:group}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \Sigma_{phy}) \\
+b_{\mathrm{group}} & \sim \textrm{MVN} ( 0 , \Sigma^2_{\mathrm{group}}) \\
+b_{\mathrm{sp:group}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \Sigma^2_{phy}) \\
 \besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
@@
 
-The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int},sp_{slope}}}$), random intercept of group ($b_{\mathrm{group}}$)and random intercept of species-group interaction ($b_{\mathrm{sp:group}}$)) compared to the single--group full model.
+The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($b_{\mathrm{group}}$)and random intercept of species-group interaction ($b_{\mathrm{sp:group}}$)) compared to the single--group full model.
 Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species, thus we can quantify variation within-species distinct from residual variation.
@@
 Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
-One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle additional random slopes ($\Sigma_{\mathrm{phy_{slope}}}$) and random intercept of species-group interactions ($b_{\mathrm{sp:group}}$) but cannot handle phylogenetic random slope-intercept covariance ($\Sigma_{\mathrm{phy_{int},phy_{slope}}}$).
+One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle additional random slopes ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercept of species-group interactions ($b_{\mathrm{sp:group}}$) but cannot handle phylogenetic random slope-intercept covariance ($\Sigma_{\mathrm{phy_{int-slope}}}$).
 Lastly, Bayesian phylogenetic GLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
@@
 
-We used our method to reproduce the examples in [chapter 11] of \cite{garamszegi2014modern} using phylogenetic GLMMs based on ``lme4'' and ``glmmTMB'' (for more details see example in \cite{michael_li_2019_2639887}).
+We used our method to reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on ``lme4'' and ``glmmTMB'' (for more details see example in supplement).
 We also used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \citep{li2017canfun}. 
-\pkg{lme4} and \pkg{pez} give identical results for fixed and random effect estimates, our code is approximately 120 times faster than \pkg{pez}. 
-Codes for fitting dune meadow analysis using \pkg{lme4} and \pkg{pez} are provided in the supplements.
+\pkg{lme4} and \pkg{pez} give identical results for fixed and random effect estimates, but our code runs approximately 120 times faster than \pkg{pez}. 
+Codes for analyzing the dune meadow data using \pkg{lme4} and \pkg{pez} are provided in the supplements.
 
@@
 Multi-group model fits on the other hand, are much more similar across platforms (only the more powerful subset of platforms can fit these models at all, and the fitting models are closer to the true simulation model).
-Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int,slope}}}$) for small number of species (i.e.
+Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int-slope}}}$) for small number of species (i.e.
 n = 25 and 50).

### 2019-07-03 phyloglmm_ms.tex
everything before discussion is done, need to fix multi-group figures. Discussion time

@@
 
-\[
+\begin{equation}
+\begin{aligned}
 \bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top} =
@@
 \end{bmatrix}.
-\]
+\end{aligned}
+\end{equation}
 
@@
 
-The full fitted model (which matches the simulation model that incorporates phylogenetic intercept, slope and correlation) provides estimates with low bias (the difference between the parameter estimates and true simulation parameters) and variance for all parameters.
+The full fitted model (which matches the simulation model that incorporates phylogenetic intercept, slope and correlation) provides estimates with low bias (average difference between the estimated parameters and the true simulation parameters) for all parameters.
 Estimates for fixed effect parameters ($\beta_0$ and $\beta_1$) approach nominal coverage as the number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not for other packages.
-\pkg{brms} have higher coverage than the nominal coverage at 95 \% because the prior distributions for the simulation parameters are cenetered at the true values \citep{li2018fitting}.
+\pkg{brms} has higher than nominal coverage (i.e., its confidence intervals are overly conservative) because the prior distributions for the simulation parameters are centered at the true values (see \citet{li2018fitting} for more details of biased coverage).
 %% \mli{Is there a figure for this?....Ok, maybe We don't need it.}
@@
 In general, models that are insufficiently flexible to match the true simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
-PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for phylogenetic intercept parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly conservative (90\% coverage of the 95\% CI with 100 species) and the fixed slope parameter ($\beta_1$) have poor coverage ($<$ 60\%).
+PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for the phylogenetic intercept standard deviation parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly conservative (90\% coverage of the 95\% CI with 100 species) and the fixed slope parameter ($\beta_1$) has poor coverage ($<$ 60\%).
 PGLS, with only one parameter available, confounds all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept and over-covering for $\beta_0$, and under-covering for $\beta_1$.
@@
 
-Multi-group model fits on the other hand, are much more similar across platforms (only the more powerful subset of platforms can fit these models at all, and the fitting models are closer to the true simulation model).
-Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model provides good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int-slope}}}$) for small number of species (i.e.
+In contrast, multi-group model fits are much more similar across platforms (only the more powerful platforms can fit these models at all, and the fitting models are closer to the true simulation model).
+Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model well and provide good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int-slope}}}$) for small numbers of species (i.e.
 n = 25 and 50).
-The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, whereas \pkg{pez} and \pkg{phyr} underestimates many parameters (Figure \ref{msplot}).
+The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, but underestimates the residual standard deviation (Figure \ref{msplot}).
 
 Although the parameter estimates are similar across platforms for the multi-group simulation fits, computational efficiency varies enormously across platforms and sample size.
-For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
-Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 species model takes $\approx$ 9 versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr} respectively.
-\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species simulation for our study, where the computational speed did not scale linearly with sample size 25/50/100.
+For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4} for the cases studied here.
+Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 species model is $\approx$ 9 versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr} respectively. 
+\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species simulation for our study, where the computational speed scaled faster than linearly with sample size.
 \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.

### 2019-07-03 phyloglmm_ms.tex
slight tweaks in the ms

@@
 In general, models that are insufficiently flexible to match the true simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
-PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for the phylogenetic intercept standard deviation parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly conservative (90\% coverage of the 95\% CI with 100 species) and the fixed slope parameter ($\beta_1$) has poor coverage ($<$ 60\%).
+PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for the phylogenetic intercept standard deviation parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly overconfident (the coverage $\approx$ 90\% with 100 species) and the fixed slope parameter ($\beta_1$) has poor coverage ($<$ 60\%).
 PGLS, with only one parameter available, confounds all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept and over-covering for $\beta_0$, and under-covering for $\beta_1$.

### 2019-07-04 phyloglmm_ms.tex
all new figures back in the ms, discussion timeg

@@
 \right) \\
-b_{\mathrm{group}} & \sim \textrm{MVN} ( 0 , \Sigma^2_{\mathrm{group}}) \\
+b_{\mathrm{group}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
 b_{\mathrm{sp:group}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \Sigma^2_{phy}) \\
@@
 
-The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($b_{\mathrm{group}}$)and random intercept of species-group interaction ($b_{\mathrm{sp:group}}$)) compared to the single--group full model.
+The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($b_{\mathrm{group}}$) and random intercept of species-group interaction ($b_{\mathrm{sp:group}}$)) compared to the single--group full model.
 Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species, thus we can quantify variation within-species distinct from residual variation.
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8,page=1]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=0.7,page=1]{./figure/ssplot.pdf}
   \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8]{./figure/sstime.pdf}
+  \includegraphics[scale=0.7]{./figure/sstime.pdf}
   \caption{Comparison of speed for all fitting packages variants: layout of the packages as in Figure \ref{ssplot}}
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8]{./figure/sscoverage.pdf}
+  \includegraphics[scale=0.7]{./figure/sscoverage.pdf}
   \caption{Comparison of coverage probablility for fixed effect parameters.
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8,page=4]{./git_push/plot.Rout.pdf}
+  \includegraphics[scale=0.7,page=1]{./figure/msplot.pdf}
   \caption{Comparison of multi-group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8]{./figure/mstime.pdf}
+  \includegraphics[scale=0.7]{./figure/mstime.pdf}
   \caption{Comparison of multi-group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}

### 2019-07-04 phyloglmm_ms.tex
temp pushing

@@
 We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to simulate responses from the simulation model (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
-Simulation parameters are illustrated in Figure \ref{ssplot} and Figure \ref{msplot}.
+All simulation parameters are illustrated with the results in Figure \ref{ssplot} and Figure \ref{msplot}.
+Correlation $\rho$ is used in place of covariances in the simulations.
 Each realization was fitted to all model variants using a range of different platforms.
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.7]{./figure/mstime.pdf}
+  \includegraphics[scale=0.8]{./figure/mstime.pdf}
   \caption{Comparison of multi-group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
@@
 
-The simulations presented in this paper are an oversimplification of evolutionary processes in nature, but they are no simpler than what researchers typically use.
+The simulations presented in this paper are an oversimplification of evolutionary processes in nature, but they are at least as complex as the models that researchers typically use.
 In fact, our results showed that even simple single predictor--response multi-species model can incorporate multiple levels of complexity.  
@@
 We have fitted models using different platforms that best match the simulation models. 
-Models that cannot match the simulation (full) model perform poorly; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods available.
+Models that cannot match the full simulation model perform poorly; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods.
 Using models that include some form of phylogenetic variation and residual variation is necessary to detect and account for correlations between species via evolutionary processes.
-Here are a few points researchers should think about when fitting phylogenetic regression models.
 
@@
 
-In classic GLMM, random effects are often use to handle group (or individual) level variation by allowing for group effects. 
-In the simplest experimental design with continuous reponse observations in different levels of a discrete grouping variable, where the response may vary according to different level of the grouping variable, fitting random intercepts are often the ``go-to'' method to handle these variations.
+In classic GLMMs, random effects are used to handle group (or individual) level variation by allowing for group effects. 
+In the simplest experimental design with continuous response observations in different levels of a discrete grouping variable, where the response may vary among levels of the grouping variable, fitting random intercepts are the ``go-to'' method to handle this variation.
 The random intercept model controls the group effect in the response level by allowing different intercepts for each level of the grouping variable.
-However, random effects can be in more complicated forms as the experimental design increases (i.e observing more variables).  
+However, random effects can take more complicated forms as the experimental design becomes more complex.  
 For example, imagine observing another continuous explanatory variable in the experimental design above, where the relationship of the response and the new explanatory variable may vary according to the grouping variable.
-In this case, fitting a random slopes model is most appropriate to handle this type of variation by allowing different slopes (relationship between continuous variables) for each level of the goruping variable.
-Random slopes models are much harder to think about in practice and require sufficient amount of data, but they are relevant over a wider range of scenarios than people are currently thinking about \cite{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
-Neglecting random slopes can lead to biased fixed effect estimates with poor coverage and type I errors \citep{schielzeth2008conclusions} (parameters of interest) as shown in the simulations above.
+A random slopes model, which allows different slopes (relationship between continuous variables) for each level of the grouping variable is most appropriate to handle this type of variation.
+Random slopes models require appropriate observation or experimental designs (multiple measurements of traits and responses within each evolutionary group), and often requre more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+Neglecting random slopes can lead to biased fixed effect estimates with poor coverage and type I errors \citep{schielzeth2008conclusions} as shown in the simulations above.
+
+%% CONT HERE
 

### 2019-07-08 phyloglmm_ms.tex
went through the whole thing once. Need to move a few pieces around and then sending it back to BB

@@
 
-Ever-increasing data collection capabilities (e.g.
-genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic data bases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
+Ever-increasing data collection capabilities (e.g., genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic databases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
 Researchers use these data to fit complex models describing species occurrence and traits.
@@
 \ldots}
-Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signals (a measure of dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
+Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signals (a measure of dependence among species responses due to their evolutionary relationships) in trait distributions or both.
 Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
@@
 % come back to complexity
-Most researchers use regression based methods such as generalized-linear models (GLMs) (a regression approach that allows for the non-normality of the data) and sophisticated ecologist would use generalized-linear mixed effect models \citep{bolker2009generalized} to control for different levels of variation.
+Most researchers use regression-based methods such as generalized linear models (GLMs) (a regression approach that allows for the non-normality of the data); sophisticated ecologist would use generalized-linear mixed effect models \citep{bolker2009generalized} to control for different levels of variation.
 However, standard statistical techniques assume species respond independently, which is typically false in evolutionary settings.
@@
 The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
-Table \ref{table:model} summerises types of platforms, data constraints and provides model complexities for phylogenetic comparative analysis.
+Table \ref{table:model} summarises types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.
 
@@
 \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
-We expect that our approach will be fast ws well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
+We expect that our approach will be fast as well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
 %% TODO LAST LINE SEE PAGE 5
@@
 Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework \citep{lynch1991methods}.
-The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorporate more than one kind of variability through random effects and allows for non-Gaussian responses.
+The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorporate more than one type of variability through random effects and allows for non-Gaussian responses.
 The typical GLMM has the form:
@@
 There are several ways to incorporate random variation in the mixed model framework.
-First, random intercepts can allow the response trait to vary across groups within grouping variables (e.g. patches, sites, or experiments). 
-Second, random intercepts can also allow the response (trait or distribution) to vary between species (tip variation) or phylogenetically across species (i.e.
-species that are closely related tend to have similar responses).
-The third type of variation that is often neglected or rarely considered are random slopes.
+First, random intercepts can allow the response trait to vary across groups within grouping variables (e.g., patches, sites, or experiments). 
+Second, random intercepts can also allow the response (trait or distribution) to vary between species (tip variation) or phylogenetically across species (i.e., species that are closely related tend to have similar responses).
+The third type of variation that is often neglected or rarely considered is random slopes.
 Random slopes can allow fixed effects (the relationship between predictors and responses) to vary for each group. 
-Analogous to modeling correlations among responses via phylogenetic relationship among species, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e.
-similar species will have similiar predictor--response relationships).
+Analogous to modeling correlations among responses via phylogenetic relationship among species, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).
 
@@
 
-Suppose that the evolution follows a Brownian-motion process, i.e. continuous traits evolve independently, following a standard Brownian-motion, along each branch of the phylogeny.
+Suppose that the evolution follows a Brownian-motion process, i.e., continuous traits evolve independently, following a standard Brownian-motion, along each branch of the phylogeny.
 In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
@@
 We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm}) with a single response variable $\bY$ and a single continuous predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
-For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e.
-$\bm{D}$ is a Gaussian distribution and $g$ is the identity link function) , corresponding to a linear mixed effect model.
+For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e., $\bm{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model.
 For the first set of simulations, we simulate one observation per species.
-% For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
+% For a broader range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
 Thus, the full simulation model is as follows:
@@
 if a positive correlation indicates that similar species have similiar relative intercepts and slopes.
-Predictor-level and intercept-level random effects of species are not applicable in this simulation setting because there is no variation at the species level (i.e.
-there is only a single observation per species).
+Predictor-level and intercept-level random effects of species are not applicable in this simulation setting because there is no variation at the species level (i.e., there is only a single observation per species).
 
@@
 We extend the simulation model by adding multiple groups where each group has one observation per species. 
-The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction (measuing the tendency of phylogenetically related species to occur in the same community).
+The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction (measuring the tendency of phylogenetically related species to occur in the same community).
 The multi-group full model is as follows: 
@@
 % \end{verbatim}
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes a evolutionary process on the tree typically Brownian-motion, but other processes can be used instead of treating the residual variation as independent.
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes an evolutionary process on the tree typically Brownian-motion, but other processes can be used instead of treating the residual variation as independent.
 Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow for both phylogenetic and residual variation, and non-Gaussian response variables.
-Both \code{gls} and \code{phylolm} can model other evolutionary processes and different correlation structures (e.g.
-Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
+Both \code{gls} and \code{phylolm} can model other evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
 Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
@@
 We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
-For Bayesian GLMM, to assess convergence we required a value of the Gelman-Rubin statistic less than the recommended threshold 1.1.
+For Bayesian GLMM, to assess convergence, we require values of the Gelman-Rubin statistic less than the recommended threshold 1.1.
 We included another convergence criteria (due to the recent critique of the Gelman-Rubin recommended threshold \citep{vats2018revisiting}) of effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0} \ \textrm{and} \ \beta_{1}$) for each replication \citep{vehtari2019rank}.
-For each replication we sample two chains starting with 10000 iterations. 
+For each replication, we sample two chains starting with 10000 iterations. 
 %% TODO MAKE SURE THIS IS CORRECT!
-We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e.
-quality of the point estimate).
+We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
 We then examine at frequentist coverage, to assess the quality of the confidence intervals.
@@
 
-The full fitted model (which matches the simulation model that incorporates phylogenetic intercept, slope and correlation) provides estimates with low bias (average difference between the estimated parameters and the true simulation parameters) for all parameters.
+The full fitted model (which matches the simulation model that incorporates phylogenetic intercept, slope, and correlation) provides estimates with low bias (average difference between the estimated parameters and the true simulation parameters) for all parameters.
 Estimates for fixed effect parameters ($\beta_0$ and $\beta_1$) approach nominal coverage as the number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not for other packages.
@@
 
-In contrast, multi-group model fits are much more similar across platforms (only the more powerful platforms can fit these models at all, and the fitting models are closer to the true simulation model).
+In contrast, the multi-group model fits are much more similar across platforms (only the more powerful platforms can fit these models at all, and the fitting models are closer to the true simulation model).
 Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model well and provide good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int-slope}}}$) for small numbers of species (i.e.
 n = 25 and 50).
-The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model, but underestimates the residual standard deviation (Figure \ref{msplot}).
+The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models do not appear to have a large effect on the estimates of the remaining parameters in the model but underestimates the residual standard deviation (Figure \ref{msplot}).
 
@@
 The simulations presented in this paper are an oversimplification of evolutionary processes in nature, but they are at least as complex as the models that researchers typically use.
-In fact, our results showed that even simple single predictor--response multi-species model can incorporate multiple levels of complexity.  
+Our results showed that even simple single predictor--response multi-species model could incorporate multiple levels of complexity.  
 Fitting simpler variations can lead to unreliable estimates of the fixed effects, which are typically the quantities of interest.
@@
 For example, imagine observing another continuous explanatory variable in the experimental design above, where the relationship of the response and the new explanatory variable may vary according to the grouping variable.
-A random slopes model, which allows different slopes (relationship between continuous variables) for each level of the grouping variable is most appropriate to handle this type of variation.
-Random slopes models require appropriate observation or experimental designs (multiple measurements of traits and responses within each evolutionary group), and often requre more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
-Neglecting random slopes can lead to biased fixed effect estimates with poor coverage and type I errors \citep{schielzeth2008conclusions} as shown in the simulations above.
+A random slopes model, which allows different slopes (the relationship between continuous variables) for each level of the grouping variable is most appropriate to handle this type of variation.
+Random slopes models require appropriate observation or experimental designs (multiple measurements of traits and responses within each evolutionary group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+Neglecting random slopes can lead to biased fixed effect estimates with inadequate coverage and type I errors \citep{schielzeth2008conclusions} as shown in the simulations above.
 
-%% CONT HERE
-
-When analyzing relationships among species between traits, it is entirely plausible that change in effect of predictor variables have on traits are changing with respect to different grouping of species or even in phylogenetically correlated way (phylogenetic random slopes).
-Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random effects are more approriate than simple grouping random effects, or including all of them in the model.
+% When analyzing relationships among species between traits, it is entirely plausible that change in the effect of predictor variables have on traits are changing with respect to different grouping of species or even in a phylogenetically correlated way (phylogenetic random slopes).
+Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random effects are more appropriate than simple grouping random effects, or including all of them in the model.
 However, the model complexity depends on the experimental design and whether there are enough data to estimate these different levels of variation.
 Some of these variations can be highly confounded in phylogenetic regression if it is not handled correctly or if there are not enough data.
-For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ (where one parameter quantifying phylogenetic variation and one for the residual variation) will be sufficient to a first approximation.
-If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic variation by confounding tip variation with residual variation, which is equivalent to Felsenstein's ``worst case".
-There are a few ways to deal with this case.
-For balanced data one can summarize multiple observations to a single measurement (for example, mean or median) per species to avoid confounding residual and tip variations.
-For unbalanced data, one can handle by weighting by number of observations per species, which is equivalent to assuming homoscedasticity using inverse-variance weights. 
-Alternatively, when within-species variance is actually of interest, accounting for within-species variance (i.e.
-adding species-level random effect in our example) can automatically handle multiple observations per species.
+For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ (where one parameter quantifies phylogenetic variation and quantifies for the residual variation) will be sufficient.
+However, if multiple observations are available per species, then using methods such as Pagel's $\lambda$ may confound tip variation with residual variation.
+In this case, multiple observations can summarized to a single measurement (for example, weighted-mean) per species to avoid confounding residual and tip variation.
+This is equivalent to assuming homoscedasticity using inverse-variance weights for unbalanced datasets. 
+Alternatively, when the within-species variance is actually of interest, accounting for within-species variation (i.e., adding species-level random effects in our example) can automatically handle multiple observations per species.
 
 It may be easier to be conservative to include both (phylogenetic random effects and grouping random effects) of them but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
-Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
-and Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}?
+
+Another simplifying alternative is to be conservative to include both phylogenetic random effects and grouping random effects but simplify the phylogenetic relationships and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
+However, this collapses phylogenetic structures to a standard random-slopes model. 
+Users should be aware of two essential questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
+Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}?
 % However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
@@
 We have presented the range of classical phylogenetic comparative methods (i.e.
-phylogenetic least squares, linear and mixed models) to fit data that incorporates different levels of phylogenetic variation.
-Even within this scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-Brownian evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \citep{butler2004phylogenetic}), and Bayesian approaches \citep{hadfield2010general}.
-However, the simple approach we developed here offers a more efficient and mathematical equivalent way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models.
-This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using any platforms that supports independent latent variables such as \pkg{MCMCglmm} \citep{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
-However, in principle just like GLMM and most statistical models, users should be aware of the constraints of the amount of data they have relative to the complexity of the model they want to fit (i.e.
+phylogenetic least squares, linear and mixed models) to fit models that incorporates different levels of phylogenetic variation.
+Even within this scope, there is additional room for exploration we neglected, such as phylogenetic multivariate response models; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck (OU) model which accounts for both selection and drift processes \citep{butler2004phylogenetic}); Bayesian approaches \citep{hadfield2010general}; and variable-rate model, where evolutionary parameters vary across the phylogeny.
+However, the simple approach we developed here offers an efficient way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models. 
+This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using any platforms that supports independent latent variables such as Stan.
+As always, users should be aware of the constraints of the amount of data they have relative to the complexity of the model they want to fit (i.e.
 What should we do when we don't have enough data?
@@
 
-\section*{Conclusion}
-
-We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical PCM to simulated data. 
-Our approach is orders of magnitud faster than existing phylogenetic mixed models and it can easily combined with any statistical mixed model framework. 
-It is more flexible in fitting large phylogenies, large volumes of data, unbalanced data sets, and complex random effects such as slope correlations.
+% \section*{Conclusion}
+% 
+% We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical PCM to simulated data. 
+% Our approach is orders of magnitud faster than existing phylogenetic mixed models and it can easily combined with any statistical mixed model framework. 
+% It is more flexible in fitting large phylogenies, large volumes of data, unbalanced data sets, and complex random effects such as slope correlations.
 

### 2019-07-08 phyloglmm_ms.tex
pushing to BB

@@
 
-%% CONT HERE, taking a break
 \section*{Results}
@@
 
-The simulations presented in this paper are an oversimplification of evolutionary processes in nature, but they are at least as complex as the models that researchers typically use.
-Our results showed that even simple single predictor--response multi-species model could incorporate multiple levels of complexity.  
-Fitting simpler variations can lead to unreliable estimates of the fixed effects, which are typically the quantities of interest.
-
 We have fitted models using different platforms that best match the simulation models. 
@@
 Using models that include some form of phylogenetic variation and residual variation is necessary to detect and account for correlations between species via evolutionary processes.
+
+The simulations presented in this paper are an oversimplification of evolutionary processes in nature, but they are at least as complex as the models that researchers typically use.
+Even a simple single predictor--response multi-species model like the one presented above could incorporate multiple levels of complexity.
+In theory, as long as there are a large amount of data and computational power to estimate model parameters (e.g., fixed and random effects), models that can incorporate a variety level of complexity always be better in capturing the system.
+However, it is rarely the case in real applications where ecologist and evolutionary biologist are constrained one way or another. 
+As always, users should be aware of the constraints of the amount of data they have relative to the complexity of the model they want to fit (i.e., What should we do when we don't have enough data?)
+and find the right balance of complexity and data \citep{baayen2008mixed}.
+Being overly conservative by fitting simple models can lead to unreliable estimates of the fixed effects, which are typically the quantities of interest.
+In contrast, being too ambitious can lead to overfitting where the data are unable to support. 
+However, random-effect models (random-slopes models in particular) are almost always theoretically justified, but a given set of data may be insufficient to fit the correct model. 
+In this case, overfitting the data with a maximal model may be better off than underfitting with a simpler model that have fewer data constraints \citep{barr2013random}.
+Another possibility is to use Bayesian approaches with appropriate prior assumptions to aid convergence \citep{hadfield2010mcmc}.
 
@@
 This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using any platforms that supports independent latent variables such as Stan.
-As always, users should be aware of the constraints of the amount of data they have relative to the complexity of the model they want to fit (i.e.
-What should we do when we don't have enough data?
-Should we use a complex model and overfit \citep{barr2013random}, the right balance of complexity and data \citep{baayen2008mixed}, singular fits or possibility of Bayesian approaches \citep{hadfield2010mcmc}).
 More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.

### 2019-07-08 phyloglmm_ms.tex
pushing again
puhsing yet again

@@
 \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
-We expect that our approach will be fast as well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
+% We expect that our approach will be fast as well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
 %% TODO LAST LINE SEE PAGE 5
@@
 However, the simple approach we developed here offers an efficient way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models. 
-This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using any platforms that supports independent latent variables such as Stan.
+This approach can in principle be combined flexibly with any platforms that supports independent latent variables such as Stan.
 More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.

### 2019-07-22 phyloglmm_ms.tex
minor change

@@
 \end{equation}
-where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions};, $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; and $\bX_{i}$ is the $n \times p_{i}$ raw random-effects model matrix.
+where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}; $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; and $\bX_{i}$ is the $n \times p_{i}$ raw random-effects model matrix.
 

### 2019-07-23 phyloglmm_ms.tex

@@
 \usepackage{alltt}
+%% hard to sort by year automatically ...
+%% https://tex.stackexchange.com/questions/377449/sorting-citations-and-bibliography-using-natbib-and-plainnat?rq=1
 \usepackage[sort]{natbib}
@@
 % come back to complexity
-Most researchers use regression-based methods such as generalized linear models (GLMs) (a regression approach that allows for the non-normality of the data); sophisticated ecologist would use generalized-linear mixed effect models \citep{bolker2009generalized} to control for different levels of variation.
+Most researchers use regression-based methods such as generalized linear models (GLMs) or generalized-linear mixed effect models \citep[GLMMs:][]{bolker2009generalized} to control for different levels of variation.
 However, standard statistical techniques assume species respond independently, which is typically false in evolutionary settings.
-The classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the trait that is the dependent variable evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
-If the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
-More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, heritability $h^2$ \citep{housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for different evolutionary processes.
+Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
+If the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
+More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for different evolutionary processes.
 In particular, these methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
-However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation are now both part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
+If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation are now both part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
 %\bmb{This is a good point, well expressed.
@@
 
-Classic PCMs usually examine the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects (or levels) of the predictor variables may evolve according to the phylogenetic relationship across species as well.
+Classic PCMs usually allow the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects (or levels) of the predictor variables may evolve according to the phylogenetic relationship across species as well.
 Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
@@
 Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
-Several recent studies have incorporated phylogenetic variations in different ways and looked at species response to phylogenetic variations with changes in environmental factors.
-For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach.
-The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
+Several recent studies have incorporated phylogenetic variation in different ways and looked at species response to phylogenetic variation with changes in environmental factors.
+For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach.
+The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2018brms}.
 Table \ref{table:model} summarises types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.
@@
 We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
-\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
+\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2018brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
 % We expect that our approach will be fast as well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
@@
 Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework \citep{lynch1991methods}.
-The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorporate more than one type of variability through random effects and allows for non-Gaussian responses.
+The generalized linear mixed effect model is an extension of the classic regression model, a hierarchical linear model that  allows for non-Gaussian responses and uses random effects to flexibly incorporate multiple types of variability.
 The typical GLMM has the form:
+\newcommand{\dist}{{\cal D}}
 \begin{equation}
 \begin{aligned}
-\bY & = \bm{D}(g^{-1}(\bmu), \phi) \\
+\bY & = \dist(g^{-1}(\bmu), \phi) \\
 \bmu & = \bX \bbeta + \bZ \bb  \\
@@
 \end{equation}
-where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the ``G-side" effect) representing the conditional modes, is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$ and $\phi$ is a scale parameter for the conditional distribution $\bm{D}$.
+where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional modes, is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
 % ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
@@
 There are several ways to incorporate random variation in the mixed model framework.
-First, random intercepts can allow the response trait to vary across groups within grouping variables (e.g., patches, sites, or experiments). 
-Second, random intercepts can also allow the response (trait or distribution) to vary between species (tip variation) or phylogenetically across species (i.e., species that are closely related tend to have similar responses).
-The third type of variation that is often neglected or rarely considered is random slopes.
-Random slopes can allow fixed effects (the relationship between predictors and responses) to vary for each group. 
-Analogous to modeling correlations among responses via phylogenetic relationship among species, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).
+First, random intercepts can allow the response trait to vary independently across groups other than species (e.g., patches, sites, or experiments). 
+Second, random intercepts can also allow the response (trait or distribution) to vary either independently among species (which we call ``tip variation'' since species represent the tips of the phylogenetic tree) or among species in a phylogenetically correlated way (i.e., species that are closely related tend to have similar responses).
+The third type of variation that is often neglected is random slopes.
+Random slopes allow fixed effects (the relationship between predictors and responses) to vary among groups.
+Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).
 
@@
 The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($b_{\mathrm{group}}$) and random intercept of species-group interaction ($b_{\mathrm{sp:group}}$)) compared to the single--group full model.
-Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species, thus we can quantify variation within-species distinct from residual variation.
-Random intercept of species--group interactions describes whether the species within a group are more similar on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
+Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species; thus we can quantify variation among species as distinct from residual variation.
+Variance in the intercept of species--group interactions ($b_{\mathrm{sp:group}}$) describes whether the species within a group are more similar on average than expected by chance \bmb{clarify?
+have more similar responses?}, equivalent to phylogenetic attraction \citep{helmus2007separating}.
 
@@
 We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}.
-We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, and \pkg{brms} \citep{burkner2016brms}.
+We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, and \pkg{brms} \citep{burkner2018brms}.
 % \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
@@
 For Bayesian GLMM, to assess convergence, we require values of the Gelman-Rubin statistic less than the recommended threshold 1.1.
-We included another convergence criteria (due to the recent critique of the Gelman-Rubin recommended threshold \citep{vats2018revisiting}) of effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0} \ \textrm{and} \ \beta_{1}$) for each replication \citep{vehtari2019rank}.
+Based on recent critique of typical Gelman-Rubin thresholds \citep{vats2018revisiting}), we included another convergence criterion of effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) for each replication \citep{vehtari2019rank}.
 For each replication, we sample two chains starting with 10000 iterations. 
@@
 We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
-We then examine at frequentist coverage, to assess the quality of the confidence intervals.
+We then compute frequentist coverage to assess the quality of the confidence intervals.
 Coverage refers to the frequency with which the computed confidence intervals include the true values of parameters.
 We used 95\% (Wald and profile) confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
-We also compare computational speed between different platforms to evaluate the efficiency of the platforms and methods.
+\bmb{clarify: did you use profile if available, otherwise Wald?
+You don't give any more detail in your coverage plots.
+[How different were Wald and profile for these results, or did you not compare them?]}
+We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.
 
@@
   \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
+Total simulations $N=100$ for each category.
 The horizontal line shows the true value of the parameters in the simulation model.
-Models capable of fitting all parameters (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) fit well for all parameters.}
+Models capable of fitting all parameters (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) fit well for all parameters.
+\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)}
+  }
 \label{ssplot}
@@
   \includegraphics[scale=0.7]{./figure/sstime.pdf}
-  \caption{Comparison of speed for all fitting packages variants: layout of the packages as in Figure \ref{ssplot}}
+  \caption{Comparison of speed for all fitting platforms: layout of the packages as in Figure \ref{ssplot} \bmb{violins?
+Consider tweaking y-axis labels (0.1, 10, 1000) (maybe tweak \code{scipen}, or specify tick pos/labels manually).
+Maybe reverse legend order \code{ guides(fill = guide\_legend(reverse = TRUE))}: \url{https://www.datanovia.com/en/blog/ggplot-legend-title-position-and-labels/} to match order in plot
+    }
+  }
 \label{ssplot_speed}
@@
 Models matching the simulation model (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) have coverage near the nominal value of 0.95.
-The black line shows the nominal coverage, and the grey ribbon the 95\% binomial confidence interval based on 100 simulated fits.}
+The black line shows the nominal coverage, and the grey ribbon the 95\% binomial confidence interval based on 100 simulated fits.
+\bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
+  }
 \label{ssplot_coverage}
@@
 Estimates for fixed effect parameters ($\beta_0$ and $\beta_1$) approach nominal coverage as the number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not for other packages.
-\pkg{brms} has higher than nominal coverage (i.e., its confidence intervals are overly conservative) because the prior distributions for the simulation parameters are centered at the true values (see \citet{li2018fitting} for more details of biased coverage).
-%% \mli{Is there a figure for this?....Ok, maybe We don't need it.}
+\pkg{brms} has higher than nominal coverage (i.e., its confidence intervals are overly conservative) because the prior distributions for the simulation parameters are centered at the true values (\citet{li2018fitting} discuss the interaction of informative priors and Bayesian calibration).
 
@@
   \includegraphics[scale=0.7,page=1]{./figure/msplot.pdf}
-  \caption{Comparison of multi-group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
+  \caption{Comparison of multi-group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.
+    \bmb{need to work harder to match up colours across plots - colours corresponding to specific platforms are different in different plots because different subsets are included (use \code{scale\_colour\_manual} appropriately)}
+  }
   \label{msplot}
@@
 In contrast, the multi-group model fits are much more similar across platforms (only the more powerful platforms can fit these models at all, and the fitting models are closer to the true simulation model).
-Similar to the single group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model well and provide good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int-slope}}}$) for small numbers of species (i.e.
+Similar to the single-group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model well and provide good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int-slope}}}$) for small numbers of species (i.e.
 n = 25 and 50).
@@
 Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 species model is $\approx$ 9 versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr} respectively. 
-\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for \pkg{pez} and \pkg{phyr} to fit 500-species simulation for our study, where the computational speed scaled faster than linearly with sample size.
+\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for us to fit 500-species models with \pkg{pez} and \pkg{phyr}, because computational speed scaled faster than linearly with sample size.
 \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
@@
 
-We have fitted models using different platforms that best match the simulation models. 
-Models that cannot match the full simulation model perform poorly; it is important to understand the limitations and performance of these simpler models that are restricted by the commonly used methods.
-Using models that include some form of phylogenetic variation and residual variation is necessary to detect and account for correlations between species via evolutionary processes.
+We have simulated relatively complex models containing phylogenetic variation in both intercepts and slopes, as well as within-species variation that is quantifiable because we allow multiple observations per species.
+These models are intrinsically more complex than some simple platforms for phylogenetic regression can handle, which may seem unfair; nevertheless, our models are certainly \emph{less} complex than evolutionary processes occurring in nature.
+Our results show that models that cannot match the full ``simulation world'' perform poorly even for the parameters they do measure; it is important to understand the limitations of these simpler, commonly used methods.
 
-The simulations presented in this paper are an oversimplification of evolutionary processes in nature, but they are at least as complex as the models that researchers typically use.
-Even a simple single predictor--response multi-species model like the one presented above could incorporate multiple levels of complexity.
-In theory, as long as there are a large amount of data and computational power to estimate model parameters (e.g., fixed and random effects), models that can incorporate a variety level of complexity always be better in capturing the system.
-However, it is rarely the case in real applications where ecologist and evolutionary biologist are constrained one way or another. 
-As always, users should be aware of the constraints of the amount of data they have relative to the complexity of the model they want to fit (i.e., What should we do when we don't have enough data?)
-and find the right balance of complexity and data \citep{baayen2008mixed}.
-Being overly conservative by fitting simple models can lead to unreliable estimates of the fixed effects, which are typically the quantities of interest.
-In contrast, being too ambitious can lead to overfitting where the data are unable to support. 
-However, random-effect models (random-slopes models in particular) are almost always theoretically justified, but a given set of data may be insufficient to fit the correct model. 
-In this case, overfitting the data with a maximal model may be better off than underfitting with a simpler model that have fewer data constraints \citep{barr2013random}.
-Another possibility is to use Bayesian approaches with appropriate prior assumptions to aid convergence \citep{hadfield2010mcmc}.
+Even our relatively simple models can incorporate many layers of complexity --- e.g.
+multiple spatial grouping variables as well as correlated phylogenetic variation in the effects of several different traits and environmental variables on a focal trait.
+In theory, as long as we have enough data and enough computational power, models that can incorporate more of the complexity will always describe a biological system better.
+However, real applications are almost always data-constrained.
+Deciding on a practically appropriate level of model complexity for a given
+problem and data set is an
+open and difficult general problem in statistical modeling, not just
+in phylogenetic studies.
+Should one use simple models that may be overly conservative or
+risk overfitting by using more ambitious models?
+How can one appropriately use the data themselves
+to choose model complexity \cite{roberts_cross-validation_2016}?
+What are the relative costs and benefits of using a step-down procedure
+starting from the most complex possible model \citep{barr2013random},
+choosing simpler models \emph{a priori} \citep{baayen2008mixed},
+or using Bayesian approaches with regularizing priors \citep{hadfield2010mcmc}?
 

### 2019-08-06 phyloglmm_ms.tex
fixed most of bmb's comments

@@
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+ \vspace{-8ex}
+  \date{}
 \begin{document}
@@
 \newcommand{\add}[1]{{\color{blue} ADD: #1}}
+
+\pagenumbering{gobble}
+
+% \renewcommand{\bmb}[1]{\relax}
+% \renewcommand{\mli}[1]{\relax}
 
@@
 Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species; thus we can quantify variation among species as distinct from residual variation.
-Variance in the intercept of species--group interactions ($b_{\mathrm{sp:group}}$) describes whether the species within a group are more similar on average than expected by chance \bmb{clarify?
-have more similar responses?}, equivalent to phylogenetic attraction \citep{helmus2007separating}.
+Variance in the intercept of species--group interactions ($b_{\mathrm{sp:group}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
 
@@
 Coverage refers to the frequency with which the computed confidence intervals include the true values of parameters.
-We used 95\% (Wald and profile) confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
-\bmb{clarify: did you use profile if available, otherwise Wald?
+We used 95\% Wald confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
+% \bmb{clarify: did you use profile if available, otherwise Wald?
 You don't give any more detail in your coverage plots.
@@
 Models capable of fitting all parameters (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) fit well for all parameters.
-\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)}
+%\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
 }
@@
   \includegraphics[scale=0.7]{./figure/sstime.pdf}
-  \caption{Comparison of speed for all fitting platforms: layout of the packages as in Figure \ref{ssplot} \bmb{violins?
-Consider tweaking y-axis labels (0.1, 10, 1000) (maybe tweak \code{scipen}, or specify tick pos/labels manually).
-Maybe reverse legend order \code{ guides(fill = guide\_legend(reverse = TRUE))}: \url{https://www.datanovia.com/en/blog/ggplot-legend-title-position-and-labels/} to match order in plot
-    }
-  }
+  \caption{Comparison of speed for all fitting platforms: layout of the packages as in Figure \ref{ssplot}}
 \label{ssplot_speed}
@@
 The black line shows the nominal coverage, and the grey ribbon the 95\% binomial confidence interval based on 100 simulated fits.
-\bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
+%  \bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
   }
@@
   \caption{Comparison of multi-group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.
-    \bmb{need to work harder to match up colours across plots - colours corresponding to specific platforms are different in different plots because different subsets are included (use \code{scale\_colour\_manual} appropriately)}
+%    \bmb{need to work harder to match up colours across plots - colours corresponding to specific platforms are different in different plots because different subsets are included (use \code{scale\_colour\_manual} appropriately)}
   }
@@
 
-\newpage
-
-\section*{Supplements -- translation}
-
-$1 \mid Sp_{phylo}$, $0 + X_{E} \mid Sp_{phylo}$ and $X_{E} \mid Sp_{phylo}$ means phylogenetically related have similar response,
-
-\begin{tabularx}{\textwidth}{|l|X|X|}
-\hline
-Formula & Statistics & Biology \\
-\hline
-$1 \mid Sp$ &
-random species intercept; variation within species in mean response across all factors &
-variation of how species respond \\
-\hline
-
-$0 + X_{E} \mid Sp$ &
-random slope of environment factor within species; variation in coefficient within species for the environmental factor &
-variation of how species respond to the same environmental factor \\
-\hline
-
-$1 + X_{E} \mid Sp$ &
-random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental factor with correlated mean response across all other factors &
-variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
-\hline
-
-$1 \mid Site:Sp $ &
-random variation in intercept among species within sites &
-variation of how species respond within sites \\
-\hline
-
-$1 | Sp_{Phylo} $ &
-variation among species in mean response across all factors demonstrate phylogenetic signal &
-phylogenetically related species respond similarly \\
-\hline
-
-$0 + X_{E} \mid Sp_{Phylo}$ &
-variation among species for environmental factors demonstrate phylogenetic signal &
-phylogenetically related species respond similar (share common response) to the same environmental factor \\
-\hline
-\end{tabularx}
+% \newpage
+% 
+% \section*{Supplements -- translation}
+% 
+% $1 \mid Sp_{phylo}$, $0 + X_{E} \mid Sp_{phylo}$ and $X_{E} \mid Sp_{phylo}$ means phylogenetically related have similar response,   
+% 
+% \begin{tabularx}{\textwidth}{|l|X|X|}
+% \hline
+% Formula & Statistics & Biology \\
+% \hline
+% $1 \mid Sp$ &
+% random species intercept; variation within species in mean response across all factors &
+% variation of how species respond \\
+% \hline
+% 
+% $0 + X_{E} \mid Sp$ &
+% random slope of environment factor within species; variation in coefficient within species for the environmental factor &
+% variation of how species respond to the same environmental factor \\
+% \hline
+% 
+% $1 + X_{E} \mid Sp$ &
+% random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental factor with correlated mean response across all other factors &
+% variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
+% \hline
+% 
+% $1 \mid Site:Sp $ &
+% random variation in intercept among species within sites &
+% variation of how species respond within sites \\
+% \hline
+% 
+% $1 | Sp_{Phylo} $ &
+% variation among species in mean response across all factors demonstrate phylogenetic signal &
+% phylogenetically related species respond similarly \\
+% \hline
+% 
+% $0 + X_{E} \mid Sp_{Phylo}$ &
+% variation among species for environmental factors demonstrate phylogenetic signal &
+% phylogenetically related species respond similar (share common response) to the same environmental factor \\
+% \hline
+% \end{tabularx}
+%             
+%             
+%             
+%             
+%             
+%
 

### 2019-08-06 plot.R

@@ -4,6 +4,7 @@ library(dplyr)
 library(ggplot2)
 library(tidyr)
 library(grid)
+library(scales)
 theme_set(theme_bw())
 zmargin <- theme(panel.spacing=grid::unit(0,"lines"))
 
@@ -62,7 +63,9 @@ gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform))
 	# + geom_violin(position=position_dodge(width=0.2),alpha=0.4)
 	+ geom_boxplot(outlier.colour = NULL, varwidth=TRUE)
 	# + scale_fill_brewer(palette = "Dark2")
-	+ scale_y_log10()
+	+ scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x)
+	    , labels = trans_format("log10", math_format(10^.x))
+	  )
 	+ scale_color_manual(values=c("Black","Red","Dark Blue","Dark Green","Orange"))
 	
 	+ ylab("Time (seconds)")
@@ -96,7 +99,8 @@ gg_sscoverage <- (ggplot(data=ss_coverage
 	+ scale_color_manual(values=c("Black","Red","Dark Blue","Dark Green","Orange"))
 	+ annotate("rect", xmin=0, xmax=4
 	           , ymin=0.95 - 2*sqrt(0.95*0.05/100)
-	           , ymax=0.95 + 2*sqrt(0.95*0.05/100), alpha=0.2)
+	           , ymax=0.95 + 2*sqrt(0.95*0.05/100), alpha=0.1
+	           , fill="blue")
   + theme(panel.spacing = unit(0,'lines'))
 	+ xlab("Number of Species")
 	+ ylab("Coverage")
@@ -180,7 +184,7 @@ gg_mscoverage <- (ggplot(data=ms_coverage
 # + ggtitle("Single Site coverage")
 + annotate("rect", xmin=0, xmax=5
            , ymin=0.95 - 2*sqrt(0.95*0.05/100)
-           , ymax=0.95 + 2*sqrt(0.95*0.05/100), alpha=0.2)
+           , ymax=0.95 + 2*sqrt(0.95*0.05/100), alpha=0.1, fill="blue")
 + theme(panel.spacing = unit(0,'lines'))
 )
 

### 2019-08-07 phyloglmm_ms.tex
formated for submission

@@
 \usepackage{pdfpages}
+\usepackage{lineno}
 %% maxwidth is the original width if it is less than linewidth
@@
 \makeatother
+
+\linenumbers
 
@@
 
-\newcommand{\fref}[1]{Figure~\ref{fig:#1}}
+\renewcommand{\figurename}{Fig.}
 
@@
 
-\pagenumbering{gobble}
+\newcommand{\pkg}[1]{{\tt #1}}
+\newcommand{\code}[1]{{\tt #1}}
+
+% \pagenumbering{gobble}
 
@@
 %\SweaveOpts{concordance=TRUE}
-\maketitle
+% \maketitle
+
+\begin{center}
+{\Large
+\textbf\newline{Reformulating phylogenetic mixed models to improve flexibility and speed} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+\vspace{2pt}
+Michael Li\textsuperscript{1*} and Benjamin M. Bolker\textsuperscript{1,2,3}
+\\
+\vspace{2pt}
+* Corresponding author: Michael Li; lim88@mcmaster.ca
+\end{center}
+\vspace{5pt}
+\textbf{1} Department of Biology, McMaster University, Hamilton, Ontario, Canada
+\\
+\textbf{2} Department of Mathematics and Statistics, McMaster University, Hamilton, Ontario, Canada
+\\
+\textbf{3} Institute for Infectious Diseases Research, McMaster University, Hamilton, Ontario, Canada
+\\
+\bigskip
+
+\section*{Abstract}
+
+\begin{enumerate}
+\item{Phylogenetic comparative methods (PCM) using phylogenetic (eco-evolutionary) regression is a powerful technique to explore relationships among species traits or abundance distributions.
+However, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.}
+\item{We propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible and computationally efficient.
+We have implemented this formulation in two R statistical packages (\pkg{lme4} and \pkg{glmmTMB}).}
+\item{Our reformulation of phylogenetic generalized linear mixed model is orders of magnitude faster than existing phylogenetic mixed models.}
+\item{This approach can, in principle, be combined flexibly with any platforms that support independent latent variables.
+More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.}
+\end{enumerate}
+
+\keywords{phylogenetic comparative methods, phylogenetic correlation, phyloglmm, species--branch matrix}
 
 \doublespacing
-
-\keywords{phyloglmm, species--branch matrix, phylogenetic correlation, phylogenetic comparative methods}
 
@@
 Table \ref{table:model} summarises types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.
-
-\newcommand{\pkg}[1]{{\tt #1}}
-\newcommand{\code}[1]{{\tt #1}}
 
@@
 
+\section*{Authors’ contributions}
+
+ML and BMB conceived the ideas and designed methodology; ML and BMB implimented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analysed the results; ML wrote the first draft of the manuscript.
+All authors contributed critically to the drafts and gave final approval for publication.
+
+\section*{Data Availability}
+
+All codes are available at DOI:10.5281/zenodo.2639887.
+
 % \section*{Conclusion}

### 2019-09-05 phyloglmm_ms.tex
pushing JD's comments

@@
 \bigskip
+
+\section*{JD's Comments}
+
+Sharpen opening framing somehow. Clarify the basic idea of
+taxonomic-based investigations. Intro could have two or three
+paragraphs, and could start by explaining either PCM or phylogenetic
+regression.
+
+Work on topic sentences. For example, the first paragraph
+in the Challenges section has absolutely the wrong topic sentence.
+Think of a paragraph as a unit of your story: you need more paragraphs
+in many places. This can be achieved with paragraph breaks and some
+sentence-shuffling.
+
+I frankly think you could make a lot of progress if you look at every
+section and ask: what are my paragraphs, and what should be their
+topic sentences?
+
+Clarify the hierarchical structure of the document. It's not
+completely clear whether or not challenges is nested under
+introduction or is the same level.
+
+In either case, some of the stuff in challenges should be postponed to
+methods. Everything pre methods should be aimed at getting across a
+high-level message of what this paper is adding.
+
+Challenges needs to be less detailed, and more clear about the
+fundamental relationship between methods. You say your method is not
+mathematically equivalent the other methods, which sort of implies
+that they're all mathematically equivalent to each other. But this
+last fact will come as a surprise or even a confusion to the reader.
+
+You should develop a clear point of view on the relationship between
+PCM and phylogenetic regression. Sometimes it seems like the former is
+broader than the ladder, and sometimes you seem to use them as
+synonyms. It kind of seems to me like you could  talk much less about
+PCM, I guess this is mostly a paper about phylogenetic regression. Or
+maybe it's a paper about "phylogenetic linear models", and we should
+just introduce this term?
+It kind of seems like methods needs an introductory paragraph before
+you jump into the details. I don't usually do that, but this is a
+pretty complicated method section.
+
+You need to say how the glmm approach relates conceptually to the
+simple phylogenetic regression above it.
+
+Be consistent about your transpose notation.
+
+The text around (3) is quite deep. It is not clear to me what you mean
+by term-wise. You have not explained what you mean by grouping
+factors.And the J Matrix, which apparently doesn't even need to be
+square, is just an identity Matrix. There's probably a way to present
+this that would lead to better intuition.
+
+Can you say a little more about the intuition or mechanism behind
+Khatri-Rao? KR seems to work on partitioned matrices, but I don't
+see clearly how these matrices are partitioned.
+
+The text around (5) should have less words and strive to be more
+explanatory. You can rely more on the equation.
+
+Be consistent with the exposition of simulation models. Groups are not
+well explained, and you suddenly talk under multi-group about one
+observation per species -- presumably that was true for the previous
+model?
+
+(6) is too clunky. Is there a way to simplify the notation for MVNs?
+Or just for the expandend covariance matrices, I guess.
+
+Platforms is a single huge ¶ where the topic s. is actually tangential
+(could literally be removed). See if you can get clear ideas across
+and rely more on the table.
+
+"We simulated 100 phylogenetic trees … to simulate responses …" Does
+anything after the first ellipsis carry information? What are the
+cites supporting?
+
+"Correlation $\rho$ is used in place of covariances" -- what does this mean?
+
+"Each realization was fitted to all model variants using a range of
+different platforms." Unnecessarily vague, tighten up here.
+
+Be careful with the word "identical". If pez is really identical, can
+you explain briefly why and maybe say "and therefore is not shown"? Is
+that why pez doesn't appear in F2?
+
+It should be more clear to the reader when you are using your method
+and when you are using older methods. Is F2 all implementations of the
+new method? Say so.
+
+F3. What is meant by layout of the packages?
+
+brms coverage is very high in F4. Are you using the right Bayesian method?
+
+In F2 maybe use dashed boxplots for the first two and explain why.
+Also, refer to the last three in the same order in the caption as the
+legend.
+
+I finally see from T2 that pez and phyr apparently "can't" do the
+single group. This should be briefly explained somewhere salient in
+text.
+
+F5 and 6 captions are switched. Also, they shouldn't both refer to F2.
+Not clear that the speed one should refer to anything, since nothing
+obvious seems to be explained about the speed (nor necessarily
+needed).
+
+F6. Why is pez fuzzed out? Does this relate to my identicality questions?
+
+F7. This "annotations" thing is actually getting annoying. Are the
+captions just unfinished?
+
+Yeah, I'm a stop now. You can take these comments for what they're
+worth, but I don't want to see this Chapter again until you've done
+all of the macro-editing of the Results section (figures in the right
+order, an attempt at captions, figure references correct).
 

### 2019-09-06 phyloglmm_ms.tex
fixed most of BMB's comments, commented out JD's comments for now

--- 2019-08-07 phyloglmm_ms.tex
@@
 \bigskip
+% 
+% \section*{JD's Comments}
+% 
+% Sharpen opening framing somehow. Clarify the basic idea of
+% taxonomic-based investigations. Intro could have two or three
+% paragraphs, and could start by explaining either PCM or phylogenetic
+% regression.
+% 
+% Work on topic sentences. For example, the first paragraph
+% in the Challenges section has absolutely the wrong topic sentence.
+% Think of a paragraph as a unit of your story: you need more paragraphs
+% in many places. This can be achieved with paragraph breaks and some
+% sentence-shuffling.
+% 
+% I frankly think you could make a lot of progress if you look at every
+% section and ask: what are my paragraphs, and what should be their
+% topic sentences?
+% 
+% Clarify the hierarchical structure of the document. It's not
+% completely clear whether or not challenges is nested under
+% introduction or is the same level.
+% 
+% In either case, some of the stuff in challenges should be postponed to
+% methods. Everything pre methods should be aimed at getting across a
+% high-level message of what this paper is adding.
+% 
+% Challenges needs to be less detailed, and more clear about the
+% fundamental relationship between methods. You say your method is not
+% mathematically equivalent the other methods, which sort of implies
+% that they're all mathematically equivalent to each other. But this
+% last fact will come as a surprise or even a confusion to the reader.
+% 
+% You should develop a clear point of view on the relationship between
+% PCM and phylogenetic regression. Sometimes it seems like the former is
+% broader than the ladder, and sometimes you seem to use them as
+% synonyms. It kind of seems to me like you could  talk much less about
+% PCM, I guess this is mostly a paper about phylogenetic regression. Or
+% maybe it's a paper about "phylogenetic linear models", and we should
+% just introduce this term?
+% It kind of seems like methods needs an introductory paragraph before
+% you jump into the details. I don't usually do that, but this is a
+% pretty complicated method section.
+% 
+% You need to say how the glmm approach relates conceptually to the
+% simple phylogenetic regression above it.
+% 
+% Be consistent about your transpose notation.
+% 
+% The text around (3) is quite deep. It is not clear to me what you mean
+% by term-wise. You have not explained what you mean by grouping
+% factors.And the J Matrix, which apparently doesn't even need to be
+% square, is just an identity Matrix. There's probably a way to present
+% this that would lead to better intuition.
+% 
+% 
+% Can you say a little more about the intuition or mechanism behind
+% Khatri-Rao? KR seems to work on partitioned matrices, but I don't
+% see clearly how these matrices are partitioned.
+% 
+% The text around (5) should have less words and strive to be more
+% explanatory. You can rely more on the equation.
+% 
+% Be consistent with the exposition of simulation models. Groups are not
+% well explained, and you suddenly talk under multi-group about one
+% observation per species -- presumably that was true for the previous
+% model?
+% 
+% (6) is too clunky. Is there a way to simplify the notation for MVNs?
+% Or just for the expandend covariance matrices, I guess.
+% 
+% Platforms is a single huge ¶ where the topic s. is actually tangential
+% (could literally be removed). See if you can get clear ideas across
+% and rely more on the table.
+% 
+% "We simulated 100 phylogenetic trees … to simulate responses …" Does
+% anything after the first ellipsis carry information? What are the
+% cites supporting?
+% 
+% "Correlation $\rho$ is used in place of covariances" -- what does this mean?
+% 
+% "Each realization was fitted to all model variants using a range of
+% different platforms." Unnecessarily vague, tighten up here.
+% 
+% Be careful with the word "identical". If pez is really identical, can
+% you explain briefly why and maybe say "and therefore is not shown"? Is
+% that why pez doesn't appear in F2?
+% 
+% It should be more clear to the reader when you are using your method
+% and when you are using older methods. Is F2 all implementations of the
+% new method? Say so.
+% 
+% F3. What is meant by layout of the packages?
+% 
+% brms coverage is very high in F4. Are you using the right Bayesian method?
+% 
+% In F2 maybe use dashed boxplots for the first two and explain why.
+% Also, refer to the last three in the same order in the caption as the
+% legend.
+% 
+% I finally see from T2 that pez and phyr apparently "can't" do the
+% single group. This should be briefly explained somewhere salient in
+% text.
+% 
+% F5 and 6 captions are switched. Also, they shouldn't both refer to F2.
+% Not clear that the speed one should refer to anything, since nothing
+% obvious seems to be explained about the speed (nor necessarily
+% needed).
+% 
+% F6. Why is pez fuzzed out? Does this relate to my identicality questions?
+% 
+% F7. This "annotations" thing is actually getting annoying. Are the
+% captions just unfinished?
+% 
+% Yeah, I'm a stop now. You can take these comments for what they're
+% worth, but I don't want to see this Chapter again until you've done
+% all of the macro-editing of the Results section (figures in the right
+% order, an attempt at captions, figure references correct).
 
@@
 \begin{enumerate}
-\item{Phylogenetic comparative methods (PCM) using phylogenetic (eco-evolutionary) regression is a powerful technique to explore relationships among species traits or abundance distributions.
+\item{Phylogenetic comparative methods (PCM) using phylogenetic regression are a powerful technique to explore relationships among related groups of species traits.
 However, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.}
-\item{We propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible and computationally efficient.
+\item{We propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but is more flexible in practice.
 We have implemented this formulation in two R statistical packages (\pkg{lme4} and \pkg{glmmTMB}).}
-\item{Our reformulation of phylogenetic generalized linear mixed model is orders of magnitude faster than existing phylogenetic mixed models.}
-\item{This approach can, in principle, be combined flexibly with any platforms that support independent latent variables.
-More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.}
+\item{Our reformulation of phylogenetic generalized linear mixed models is much more computationally efficient; it is orders of magnitude faster than existing methods for fitting phylogenetic mixed models.}
+\item{Our approach can, in principle, be implemented in any platforms.}
+\item{This implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to a broad range of previously difficult cases (e.g., large data, unbalanced species observations, complex random effects) and explore new ideas.}
 \end{enumerate}
@@
 Researchers use these data to fit complex models describing species occurrence and traits.
-In particular, phylogenetic comparative methods (PCM) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or abundance distributions.
+In particular, phylogenetic comparative methods (PCM) using phylogenetic regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or abundance distributions.
 % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
@@
 \ldots}
-Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signals (a measure of dependence among species responses due to their evolutionary relationships) in trait distributions or both.
+Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
 Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
-Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
+Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate an evolutionary framework in to applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
 While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
@@
 % but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
-In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models. 
-This method allows researchers to easily incorporate various complexities without sacrificing speed.
+In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships by extending exisiting software for fitting mixed effect models. 
+This method allows researchers to easily incorporate evolutionary and statistical complexities without sacrificing speed.
 
@@
 % come back to complexity
-Most researchers use regression-based methods such as generalized linear models (GLMs) or generalized-linear mixed effect models \citep[GLMMs:][]{bolker2009generalized} to control for different levels of variation.
-However, standard statistical techniques assume species respond independently, which is typically false in evolutionary settings.
+Researchers use regression-based methods such as generalized linear models (GLMs) or generalized linear mixed effect models \citep[GLMMs:][]{bolker2009generalized} to control for different levels of variation.
+However, standard statistical techniques do not allow for correlations among species responses due to their shared evolutionary history.
 Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
-If the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
-More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for different evolutionary processes.
-In particular, these methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
-If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation are now both part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
+If in addition the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
+More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than Brownian-motion.
+These methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
+If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation can both be considered part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
 %\bmb{This is a good point, well expressed.
@@
 %\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
-Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such observations to partition variability more finely.
-Indeed, many existing methods are restricted to single observations per species, requiring users to collapse multi-observation data by taking species averages.
+Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
+Indeed, many existing methods are restricted to single observations per species, requiring users to collapse multiple observations per species to species mean values.
 %% REMEMBER TO REVISIT IN DISCUSSION
@@
 Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
-Taxonomic groups with similar body sizes may vary in overall brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes). 
+The standard PCM allow for phylogenetic correlations in the intercept of the relationship between body and brain size. 
+However, species within taxonomic groups with similar body sizes may vary in overall brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes).
 %\bmb{maybe we can focus on random slopes here?
@@
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach.
-The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2018brms}.
+The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many biologist needing to fit random-slopes model have usually turned to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
 Table \ref{table:model} summarises types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.
@@
 \end{tabular}} & Correlated residual & Single observation & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4} 
-& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single observation & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
+& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} & Single observation & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
 \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez}, \pkg{phyr}                                                                                            \\ \cline{3-4}
@@
 In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
-Our new formulation can be incorporated in any framework that allows for random effects (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model.
+In particular, it easily allows our new formulation can be incorporated in any framework that allows for random effects (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model.
 %% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
@@
 \newcommand{\bmu}{{\boldsymbol \mu}}
-\newcommand{\bY}{{\mathbf Y}}
+\newcommand{\bY}{{\mathbf y}}  %% fixme later
 \newcommand{\bC}{{\mathbf C}}
@@
 
-We begin by describing the classic PCM in a linear regression setting.
-Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors $\bX$. 
+We begin by describing the classic phlyogenetic regression in a linear regression setting.
+Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors encoded in a model matrix $\bX$. 
 The standard phylogenetic regression can be formulated as
@@
 \end{equation}
-where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times (m)$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation (PC) matrix.
+where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation (PC) matrix.
 The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.
@@
 \subsection*{Phylogenetic generalized linear mixed model}
-Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework \citep{lynch1991methods}.
-The generalized linear mixed effect model is an extension of the classic regression model, a hierarchical linear model that  allows for non-Gaussian responses and uses random effects to flexibly incorporate multiple types of variability.
+Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework to define a wider range that includes the standard phylogenetic regression as a special case \citep{lynch1991methods}.
+The generalized linear mixed effect model allows for non-Gaussian responses and uses random effects to flexibly incorporate multiple types of variability.
 The typical GLMM has the form:
@@
 
-There are several ways to incorporate random variation in the mixed model framework.
+There are forms of random variation in the mixed model framework.
 First, random intercepts can allow the response trait to vary independently across groups other than species (e.g., patches, sites, or experiments). 
@@
 % \end{pmatrix}
-
 %% FIXME: double check this
-The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree}, and the $\e_i$ are independent Normal variates with zero mean and $\sigma^2$ evoulationary variance (i.e.
+The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree}, and the $\e_i$ are independent Normal variates with zero mean and $\sigma^2$ evolutionary variance (i.e.
 the variance for species 1 is $\textrm{E}[(\ell_1 \e_1 + \ell_2 \e_2)^2] = (L_1 + L_2)\sigma^2$).
@@
 
-We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm}) with a single response variable $\bY$ and a single continuous predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
+We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm}) with a single response variable $\bY$ and a single continuous normally distributed predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
 For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e., $\bm{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model.
@@
 The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic dependency variability in regression effect ($b_{\mathrm{phy_{slope}}}$) and response ($b_{\mathrm{phy_{int}}}$); i.e.
-if a positive correlation indicates that similar species have similiar relative intercepts and slopes.
-Predictor-level and intercept-level random effects of species are not applicable in this simulation setting because there is no variation at the species level (i.e., there is only a single observation per species).
+if a positive correlation indicates that similar species have similar relative intercepts and slopes.
+Predictor-level and intercept-level random effects of species are not applicable in this simulation setting because there is only a single observation per species, so within-species variation cannot be separated from tip variation.
 
@@
 The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($b_{\mathrm{group}}$) and random intercept of species-group interaction ($b_{\mathrm{sp:group}}$)) compared to the single--group full model.
-Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species; thus we can quantify variation among species as distinct from residual variation.
+Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species; thus we can quantify variation among species separately from residual variation.
 Variance in the intercept of species--group interactions ($b_{\mathrm{sp:group}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
@@
 % \end{verbatim}
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes an evolutionary process on the tree typically Brownian-motion, but other processes can be used instead of treating the residual variation as independent.
-Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow for both phylogenetic and residual variation, and non-Gaussian response variables.
-Both \code{gls} and \code{phylolm} can model other evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes an evolutionary process on the tree (typically Brownian-motion, but other processes can be used) instead of treating the residual error for each species as independent.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow for both phylogenetic and residual variation, as well as non-Gaussian response variables.
+Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
 Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
-One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle additional random slopes ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercept of species-group interactions ($b_{\mathrm{sp:group}}$) but cannot handle phylogenetic random slope-intercept covariance ($\Sigma_{\mathrm{phy_{int-slope}}}$).
+One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle additional random slopes ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercept of species-group interactions ($b_{\mathrm{sp:group}}$) but does not incorporate covariation between phylogenetic random slope-intercept ($\Sigma_{\mathrm{phy_{int-slope}}}$).
 Lastly, Bayesian phylogenetic GLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
-However, MCMC is often computationally expensive for GLMMs compared to platforms using deterministic optimization.
+However, MCMC is usually much more computationally expensive for GLMMs compared to platforms using deterministic optimization.
 \pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC)\citep{duane1987hybrid}.
@@
 We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
-For Bayesian GLMM, to assess convergence, we require values of the Gelman-Rubin statistic less than the recommended threshold 1.1.
-Based on recent critique of typical Gelman-Rubin thresholds \citep{vats2018revisiting}), we included another convergence criterion of effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) for each replication \citep{vehtari2019rank}.
+For Bayesian GLMM, we included only realizations where we require values of the Gelman-Rubin statistic less than the recommended threshold 1.1.
+Based on a recent critique of typical Gelman-Rubin thresholds \citep{vats2018revisiting}), we included another convergence criterion of effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) for each replication \citep{vehtari2019rank}.
 For each replication, we sample two chains starting with 10000 iterations. 
@@
 We then compute frequentist coverage to assess the quality of the confidence intervals.
-Coverage refers to the frequency with which the computed confidence intervals include the true values of parameters.
+Coverage refers to the proportion of simulations in which the computed confidence intervals include the true values of parameters.
 We used 95\% Wald confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
@@
 
-We used our method to reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on ``lme4'' and ``glmmTMB'' (for more details see example in supplement).
+We used our method to reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on \pkg{lme4} and \pkg{glmmTMB} (for more details see example in supplement).
 We also used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \citep{li2017canfun}. 
@@
 n = 25 and 50).
-The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models do not appear to have a large effect on the estimates of the remaining parameters in the model but underestimates the residual standard deviation (Figure \ref{msplot}).
+The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model but underestimates the residual standard deviation (Figure \ref{msplot}).
 
@@
 These models are intrinsically more complex than some simple platforms for phylogenetic regression can handle, which may seem unfair; nevertheless, our models are certainly \emph{less} complex than evolutionary processes occurring in nature.
-Our results show that models that cannot match the full ``simulation world'' perform poorly even for the parameters they do measure; it is important to understand the limitations of these simpler, commonly used methods.
+Our results show that models that cannot match the full ``simulation world'' perform poorly even for the parameters they do estimate; it is important to understand the limitations of these simpler, commonly used methods.
 
@@
 In theory, as long as we have enough data and enough computational power, models that can incorporate more of the complexity will always describe a biological system better.
-However, real applications are almost always data-constrained.
+However, real applications are always data-constrained.
 Deciding on a practically appropriate level of model complexity for a given
@@
 How can one appropriately use the data themselves
-to choose model complexity \cite{roberts_cross-validation_2016}?
+to choose model complexity \citep{roberts_cross-validation_2016}?
 What are the relative costs and benefits of using a step-down procedure
@@
 A random slopes model, which allows different slopes (the relationship between continuous variables) for each level of the grouping variable is most appropriate to handle this type of variation.
-Random slopes models require appropriate observation or experimental designs (multiple measurements of traits and responses within each evolutionary group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+Random slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each evolutionary group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
 Neglecting random slopes can lead to biased fixed effect estimates with inadequate coverage and type I errors \citep{schielzeth2008conclusions} as shown in the simulations above.

### 2019-09-09 phyloglmm_ms.tex
pushing ms

@@
 \makeatother
-
-\linenumbers
 
@@
 
-% \pagenumbering{gobble}
+\pagenumbering{gobble}
+% \linenumbers
 
@@
 \bigskip
-% 
-% \section*{JD's Comments}
-% 
-% Sharpen opening framing somehow. Clarify the basic idea of
-% taxonomic-based investigations. Intro could have two or three
-% paragraphs, and could start by explaining either PCM or phylogenetic
-% regression.
-% 
-% Work on topic sentences. For example, the first paragraph
-% in the Challenges section has absolutely the wrong topic sentence.
-% Think of a paragraph as a unit of your story: you need more paragraphs
-% in many places. This can be achieved with paragraph breaks and some
-% sentence-shuffling.
-% 
-% I frankly think you could make a lot of progress if you look at every
-% section and ask: what are my paragraphs, and what should be their
-% topic sentences?
-% 
-% Clarify the hierarchical structure of the document. It's not
-% completely clear whether or not challenges is nested under
-% introduction or is the same level.
-% 
-% In either case, some of the stuff in challenges should be postponed to
-% methods. Everything pre methods should be aimed at getting across a
-% high-level message of what this paper is adding.
-% 
-% Challenges needs to be less detailed, and more clear about the
-% fundamental relationship between methods. You say your method is not
-% mathematically equivalent the other methods, which sort of implies
-% that they're all mathematically equivalent to each other. But this
-% last fact will come as a surprise or even a confusion to the reader.
-% 
-% You should develop a clear point of view on the relationship between
-% PCM and phylogenetic regression. Sometimes it seems like the former is
-% broader than the ladder, and sometimes you seem to use them as
-% synonyms. It kind of seems to me like you could  talk much less about
-% PCM, I guess this is mostly a paper about phylogenetic regression. Or
-% maybe it's a paper about "phylogenetic linear models", and we should
-% just introduce this term?
-% It kind of seems like methods needs an introductory paragraph before
-% you jump into the details. I don't usually do that, but this is a
-% pretty complicated method section.
-% 
-% You need to say how the glmm approach relates conceptually to the
-% simple phylogenetic regression above it.
-% 
-% Be consistent about your transpose notation.
-% 
-% The text around (3) is quite deep. It is not clear to me what you mean
-% by term-wise. You have not explained what you mean by grouping
-% factors.And the J Matrix, which apparently doesn't even need to be
-% square, is just an identity Matrix. There's probably a way to present
-% this that would lead to better intuition.
-% 
-% 
-% Can you say a little more about the intuition or mechanism behind
-% Khatri-Rao? KR seems to work on partitioned matrices, but I don't
-% see clearly how these matrices are partitioned.
-% 
-% The text around (5) should have less words and strive to be more
-% explanatory. You can rely more on the equation.
-% 
-% Be consistent with the exposition of simulation models. Groups are not
-% well explained, and you suddenly talk under multi-group about one
-% observation per species -- presumably that was true for the previous
-% model?
-% 
-% (6) is too clunky. Is there a way to simplify the notation for MVNs?
-% Or just for the expandend covariance matrices, I guess.
-% 
-% Platforms is a single huge ¶ where the topic s. is actually tangential
-% (could literally be removed). See if you can get clear ideas across
-% and rely more on the table.
-% 
-% "We simulated 100 phylogenetic trees … to simulate responses …" Does
-% anything after the first ellipsis carry information? What are the
-% cites supporting?
-% 
-% "Correlation $\rho$ is used in place of covariances" -- what does this mean?
-% 
-% "Each realization was fitted to all model variants using a range of
-% different platforms." Unnecessarily vague, tighten up here.
-% 
-% Be careful with the word "identical". If pez is really identical, can
-% you explain briefly why and maybe say "and therefore is not shown"? Is
-% that why pez doesn't appear in F2?
-% 
-% It should be more clear to the reader when you are using your method
-% and when you are using older methods. Is F2 all implementations of the
-% new method? Say so.
-% 
-% F3. What is meant by layout of the packages?
-% 
-% brms coverage is very high in F4. Are you using the right Bayesian method?
-% 
-% In F2 maybe use dashed boxplots for the first two and explain why.
-% Also, refer to the last three in the same order in the caption as the
-% legend.
-% 
-% I finally see from T2 that pez and phyr apparently "can't" do the
-% single group. This should be briefly explained somewhere salient in
-% text.
-% 
-% F5 and 6 captions are switched. Also, they shouldn't both refer to F2.
-% Not clear that the speed one should refer to anything, since nothing
-% obvious seems to be explained about the speed (nor necessarily
-% needed).
-% 
-% F6. Why is pez fuzzed out? Does this relate to my identicality questions?
-% 
-% F7. This "annotations" thing is actually getting annoying. Are the
-% captions just unfinished?
-% 
-% Yeah, I'm a stop now. You can take these comments for what they're
-% worth, but I don't want to see this Chapter again until you've done
-% all of the macro-editing of the Results section (figures in the right
-% order, an attempt at captions, figure references correct).
 
@@
 \item{Our reformulation of phylogenetic generalized linear mixed models is much more computationally efficient; it is orders of magnitude faster than existing methods for fitting phylogenetic mixed models.}
-\item{Our approach can, in principle, be implemented in any platforms.}
+\item{Our approach can, in principle, be implemented on any PCM platform.}
 \item{This implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to a broad range of previously difficult cases (e.g., large data, unbalanced species observations, complex random effects) and explore new ideas.}
@@
 
+%%% PCM accounts for evolutionary relationships. More people are doing it.
+Phylogenetic comparative methods (PCM) is a powerful technique to explore relationships among related groups of species traits.
+Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
+
 Ever-increasing data collection capabilities (e.g., genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic databases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
 Researchers use these data to fit complex models describing species occurrence and traits.
-In particular, phylogenetic comparative methods (PCM) using phylogenetic regression \citep{hansen2012interpreting} is a powerful technique to explore relationships among species traits or abundance distributions.
+For example, ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate an evolutionary framework in to applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
 % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
@@
 \ldots}
-Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
-Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}.
+
+Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
-Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate an evolutionary framework in to applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
-While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
+While a wide range of tools is available for phylogenetic regression, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
 In such cases, researchers typically search for ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
@@
 
-% come back to complexity
-Researchers use regression-based methods such as generalized linear models (GLMs) or generalized linear mixed effect models \citep[GLMMs:][]{bolker2009generalized} to control for different levels of variation.
-However, standard statistical techniques do not allow for correlations among species responses due to their shared evolutionary history.
+% Researchers use regression-based methods such as generalized linear models (GLMs) or generalized linear mixed effect models \citep[GLMMs:][]{bolker2009generalized} to control for different levels of variation.
+Standard statistical regression techniques do not allow for correlations among species responses due to their shared evolutionary history.
 Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
@@
 
-Classic PCMs usually allow the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects (or levels) of the predictor variables may evolve according to the phylogenetic relationship across species as well.
+Classic phylogenetic regressions usually allow the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects (or levels) of the predictor variables may evolve according to the phylogenetic relationship across species as well.
 Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
-The standard PCM allow for phylogenetic correlations in the intercept of the relationship between body and brain size. 
-However, species within taxonomic groups with similar body sizes may vary in overall brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes).
+The standard phylognetic regression allow for phylogenetic correlations in the intercept of the relationship between body and brain size. 
+However, species within taxonomic groups with similar body sizes may vary in overall brain size, or taxonomic groups may vary in the relationship between brain and body size. 
 %\bmb{maybe we can focus on random slopes here?
@@
 
-In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
+In this paper, we will propose an alternative formulation of phylogenetic regression in a generalized linear mixed modeling framework that is mathematically equivalent to previous approaches, but more flexible.
 In particular, it easily allows our new formulation can be incorporated in any framework that allows for random effects (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model.
@@
 
+We generated test data using a simple framework that combines fixed effects, phylogenetic random intercept, slope, and their correlations from a phylogenetic tree, as well as residuals. 
+We then fit the test data from these simulations using our approach implemented using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB} that matches the assumptions of the simulation model, as well as various simplifications with other platforms.
+
 \subsection*{Phylogenetic regression}
@@
 \end{equation}
-where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation (PC) matrix.
+where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic covariance (PC) matrix.
 The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.
@@
 where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional modes, is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
+When $\dist$ is Gaussian, $g$ is the identity function, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
 % ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
@@
 First, random intercepts can allow the response trait to vary independently across groups other than species (e.g., patches, sites, or experiments). 
-Second, random intercepts can also allow the response (trait or distribution) to vary either independently among species (which we call ``tip variation'' since species represent the tips of the phylogenetic tree) or among species in a phylogenetically correlated way (i.e., species that are closely related tend to have similar responses).
+Second, random intercepts can also allow the response (trait or distribution) to vary either independently among species (since species represent the tips of the phylogenetic tree) or among species in a phylogenetically correlated way (i.e., species that are closely related tend to have similar responses).
 The third type of variation that is often neglected is random slopes.
@@
 
-The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and a vector $\bell$ of square roots of branch lengths:
+The $\bS$ matrix is the product of an $m \times b$ indicator matrix $\bS_{ind}$ of branch indices and a vector $\bell$ of square roots of branch lengths:
 
@@
 \end{equation}
-where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}; $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; and $\bX_{i}$ is the $n \times p_{i}$ raw random-effects model matrix.
+where $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; $\bX_{i}$ is the $n \times p_{i}$ raw random-effects model matrix; and $\ast$ is the Khatri-Rao product \citep{khatri1968solutions} partitioned at the observation level ($n$).
 
@@
 We extend the simulation model by adding multiple groups where each group has one observation per species. 
-The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction (measuring the tendency of phylogenetically related species to occur in the same community).
-The multi-group full model is as follows: 
+The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction \citep{helmus2007separating}. 
+The full multi-group model is as follows: 
 \begin{equation}
@@
 
-Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as random effects in a generalized linear mixed model.
-We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}.
 We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, and \pkg{brms} \citep{burkner2018brms}.
@@
 \subsection*{Simulation and evaluations}
-We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to simulate responses from the simulation model (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
+We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to simulate responses from the simulation model (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}) and each realization was fitted to all model variants.
 All simulation parameters are illustrated with the results in Figure \ref{ssplot} and Figure \ref{msplot}. 
 Correlation $\rho$ is used in place of covariances in the simulations.
-Each realization was fitted to all model variants using a range of different platforms.
 Table \ref{table:platform} shows the parameters that are estimable for each platform. 
@@
 The horizontal line shows the true value of the parameters in the simulation model.
-Models capable of fitting all parameters (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) fit well for all parameters.
+Models capable of fitting all parameters (\pkg{lme4},  \pkg{glmmTMB}, and \pkg{brms}) fit well for all parameters.
 %\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
@@
   \includegraphics[scale=0.7]{./figure/sstime.pdf}
-  \caption{Comparison of speed for all fitting platforms: layout of the packages as in Figure \ref{ssplot}}
+  \caption{Comparison of speed for all fitting platforms.}
 \label{ssplot_speed}
@@
   \includegraphics[scale=0.7,page=1]{./figure/msplot.pdf}
-  \caption{Comparison of multi-group model parameter estimates across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.
+  \caption{Comparison of multi-group model parameter estimates.
+The horizontal line shows the true value of the parameters in the simulation model.
+Models capable of fitting all parameters (lme4 and glmmTMB) fit well for all parameters.
+\pkg{pez} and \pkg{phyr} estimates for n = 100, and 500 are not available because the models did not converge within 30 minutes.
 %    \bmb{need to work harder to match up colours across plots - colours corresponding to specific platforms are different in different plots because different subsets are included (use \code{scale\_colour\_manual} appropriately)}
@@
   \includegraphics[scale=0.8]{./figure/mstime.pdf}
-  \caption{Comparison of multi-group model computational speed across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
+  \caption{Comparison of multi-group model computational speed.}
   \label{msplot_time}
@@
   \includegraphics[scale=0.8]{./figure/mscoverage.pdf}
-  \caption{Comparison of multi-group model coverage across different R packages in Table 2. Annotations as in Figure \ref{ssplot}.}
+  \caption{Comparison of multi-group model coverage.}
   \label{msplot_coverage}

### 2019-09-09 phyloglmm_ms.tex
Changes that need to be checked for accuracy

@@
 %%% PCM accounts for evolutionary relationships. More people are doing it.
-Phylogenetic comparative methods (PCM) is a powerful technique to explore relationships among related groups of species traits.
+Phylogenetic comparative methods (PCMs) are a powerful technique to explore relationships among related groups of species traits.
 Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
@@
 
-\subsection*{Challenges in modeling phylogenetic process}
+\subsection*{Challenges in modeling phylogenetic processes}
 
@@
 Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
-The standard phylognetic regression allow for phylogenetic correlations in the intercept of the relationship between body and brain size. 
+Standard phylogenetic regressions allow for phylogenetic correlations in the intercept of the relationship between body and brain size. 
 However, species within taxonomic groups with similar body sizes may vary in overall brain size, or taxonomic groups may vary in the relationship between brain and body size. 
@@
 In this paper, we will propose an alternative formulation of phylogenetic regression in a generalized linear mixed modeling framework that is mathematically equivalent to previous approaches, but more flexible.
-In particular, it easily allows our new formulation can be incorporated in any framework that allows for random effects (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model.
+In particular, our new formulation can be implemented in any framework that allows for random effects (such as random intercepts, slopes, and interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model.
 %% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
@@
 We generated test data using a simple framework that combines fixed effects, phylogenetic random intercept, slope, and their correlations from a phylogenetic tree, as well as residuals. 
-We then fit the test data from these simulations using our approach implemented using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB} that matches the assumptions of the simulation model, as well as various simplifications with other platforms.
+We then fit the test data from these simulations using our approach implemented using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}, using assumptions that match those of the simulation model.
+We also fit with other platforms, using standard simplifications when necessary for implementation.
 
@@
 
-We begin by describing the classic phlyogenetic regression in a linear regression setting.
+We begin by describing the classic phylogenetic regression in a linear regression setting.
 Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors encoded in a model matrix $\bX$. 

### 2019-09-09 phyloglmm_ms.tex
Fighting. Trying to simplify, but may not be sufficiently familiar with concepts.

@@
 where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional modes, is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
-When $\dist$ is Gaussian, $g$ is the identity function, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
+When $\dist$ is Gaussian and $g$ is the identity function, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
 % ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
@@
 
-In classic GLMMs, random effects are used to handle group (or individual) level variation by allowing for group effects. 
+In classic GLMMs, random effects are used to handle group (or individual) level variation. 
 In the simplest experimental design with continuous response observations in different levels of a discrete grouping variable, where the response may vary among levels of the grouping variable, fitting random intercepts are the ``go-to'' method to handle this variation.
@@
 % When analyzing relationships among species between traits, it is entirely plausible that change in the effect of predictor variables have on traits are changing with respect to different grouping of species or even in a phylogenetically correlated way (phylogenetic random slopes).
-Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random effects are more appropriate than simple grouping random effects, or including all of them in the model.
-However, the model complexity depends on the experimental design and whether there are enough data to estimate these different levels of variation.
-Some of these variations can be highly confounded in phylogenetic regression if it is not handled correctly or if there are not enough data.
-For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ (where one parameter quantifies phylogenetic variation and quantifies for the residual variation) will be sufficient.
-However, if multiple observations are available per species, then using methods such as Pagel's $\lambda$ may confound tip variation with residual variation.
+Nevertheless, it is hard to account for all forms of complexities and decide if when it is best to use phylogenetic random effects, simple grouping, or both. 
+Optimal model complexity depends on experimental design and whether the data provides enough signal to estimate these different levels of variation, which can be strongly confounded.
+For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ will be sufficient. 
+However, if multiple observations are available per species, then these simple methods may confound tip variation with residual variation.
 In this case, multiple observations can summarized to a single measurement (for example, weighted-mean) per species to avoid confounding residual and tip variation.

### 2019-09-10 phyloglmm_ms.tex

@@
 
-We have presented the range of classical phylogenetic comparative methods (i.e.
-phylogenetic least squares, linear and mixed models) to fit models that incorporates different levels of phylogenetic variation.
-Even within this scope, there is additional room for exploration we neglected, such as phylogenetic multivariate response models; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck (OU) model which accounts for both selection and drift processes \citep{butler2004phylogenetic}); Bayesian approaches \citep{hadfield2010general}; and variable-rate model, where evolutionary parameters vary across the phylogeny.
+We have presented a range of classical phylogenetic comparative methods (i.e.
+phylogenetic least squares, linear and mixed models) to fit models that incorporate different levels of phylogenetic variation.
+Even within this scope, there is additional room for exploration, such as phylogenetic multivariate response models; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck (OU) model which accounts for both selection and drift processes \citep{butler2004phylogenetic}); Bayesian approaches \citep{hadfield2010general}; and variable-rate model, where evolutionary parameters vary across the phylogeny.
 However, the simple approach we developed here offers an efficient way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models. 
@@
 
-ML and BMB conceived the ideas and designed methodology; ML and BMB implimented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analysed the results; ML wrote the first draft of the manuscript.
+ML and BMB conceived the ideas and designed methodology; ML and BMB implemented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analysed the results; ML wrote the first draft of the manuscript.
 All authors contributed critically to the drafts and gave final approval for publication.

### 2019-09-10 phyloglmm_ms.tex
Small stuff only.

@@
 
-Suppose that the evolution follows a Brownian-motion process, i.e., continuous traits evolve independently, following a standard Brownian-motion, along each branch of the phylogeny.
+Suppose that the evolution follows a Brownian-motion process, i.e., continuous traits evolve independently, following an unbiased, continuous-time random walk, along each branch of the phylogeny.
 In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
@@
 \subsection*{Simulation and evaluations}
-We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to simulate responses from the simulation model (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}) and each realization was fitted to all model variants.
-All simulation parameters are illustrated with the results in Figure \ref{ssplot} and Figure \ref{msplot}. 
+
+We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to and then modeld responses on each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
+Each realization was fitted using all model variants.
+All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
 Correlation $\rho$ is used in place of covariances in the simulations.
@@
 For Bayesian GLMM, we included only realizations where we require values of the Gelman-Rubin statistic less than the recommended threshold 1.1.
-Based on a recent critique of typical Gelman-Rubin thresholds \citep{vats2018revisiting}), we included another convergence criterion of effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) for each replication \citep{vehtari2019rank}.
+Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}), we included the additional convergence criterion of effective sample size (ESS) $>1000$ for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) for each replication \citep{vehtari2019rank}.
 For each replication, we sample two chains starting with 10000 iterations. 
@@
 %\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
+%\jd{Use dashed lines (or some other gg signifier) for the insufficient approaches}
 }
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8]{./figure/mscoverage.pdf}
+  \includegraphics[scale=0.7]{./figure/mscoverage.pdf}
   \caption{Comparison of multi-group model coverage.}

### 2019-09-10 phyloglmm_ms.tex
wrapping up for now
fixed a minor typo

@@
 
+\doublespacing
+
 \section*{Abstract}
@@
 \keywords{phylogenetic comparative methods, phylogenetic correlation, phyloglmm, species--branch matrix}
-
-\doublespacing
 
@@
 
-We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to and then modeld responses on each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
+We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to and then modelled responses on each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
 Each realization was fitted using all model variants.
@@
   \includegraphics[scale=0.7]{./figure/sstime.pdf}
-  \caption{Comparison of speed for all fitting platforms.}
+  \caption{Comparison of single-group model computational speed.}
 \label{ssplot_speed}
@@
   \caption{Comparison of coverage probablility for fixed effect parameters.
-Models matching the simulation model (\pkg{lme4}, \pkg{brms} and \pkg{glmmTMB}) have coverage near the nominal value of 0.95.
-The black line shows the nominal coverage, and the grey ribbon the 95\% binomial confidence interval based on 100 simulated fits.
+Models matching the simulation model (\pkg{lme4}, \pkg{glmmTMB} and pkg{brms}) have coverage near the nominal value of 0.95.
+The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits.
 %  \bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.7]{./figure/mscoverage.pdf}
+  \includegraphics[scale=0.5]{./figure/mscoverage.pdf}
   \caption{Comparison of multi-group model coverage.}
@@
 %   approximate power?}
-
-\newpage
 
@@
 %                                                               
+\singlespacing
 

### 2019-09-10 phyloglmm_ms.tex
spellcheck, abstract tweaks

@@
 We have implemented this formulation in two R statistical packages (\pkg{lme4} and \pkg{glmmTMB}).}
-\item{Our reformulation of phylogenetic generalized linear mixed models is much more computationally efficient; it is orders of magnitude faster than existing methods for fitting phylogenetic mixed models.}
-\item{Our approach can, in principle, be implemented on any PCM platform.}
-\item{This implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to a broad range of previously difficult cases (e.g., large data, unbalanced species observations, complex random effects) and explore new ideas.}
+\item{Our reformulation of phylogenetic generalized linear mixed models is computationally efficient, operating orders of magnitude faster than existing methods for fitting phylogenetic mixed models.}
+\item{Our approach can, in principle, be implemented in any platform for generalized mixed models.
+Our implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to a broad range of previously difficult cases (e.g., large data, unbalanced observational designs, complex random effects).}
 \end{enumerate}
@@
 % but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
-In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships by extending exisiting software for fitting mixed effect models. 
+In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships by extending existing software for fitting mixed effect models. 
 This method allows researchers to easily incorporate evolutionary and statistical complexities without sacrificing speed.
@@
 The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many biologist needing to fit random-slopes model have usually turned to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
-Table \ref{table:model} summarises types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.
+Table \ref{table:model} summarizes types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.
 
@@
   \includegraphics[scale=0.7]{./figure/sscoverage.pdf}
-  \caption{Comparison of coverage probablility for fixed effect parameters.
+  \caption{Comparison of coverage probability for fixed effect parameters.
 Models matching the simulation model (\pkg{lme4}, \pkg{glmmTMB} and pkg{brms}) have coverage near the nominal value of 0.95.
@@
 
-ML and BMB conceived the ideas and designed methodology; ML and BMB implemented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analysed the results; ML wrote the first draft of the manuscript.
+ML and BMB conceived the ideas and designed methodology; ML and BMB implemented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analyzed the results; ML wrote the first draft of the manuscript.
 All authors contributed critically to the drafts and gave final approval for publication.

### 2019-09-10 phyloglmm_ms.tex
more small fixes

@@
 Researchers use these data to fit complex models describing species occurrence and traits.
-For example, ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate an evolutionary framework in to applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
+For example, ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate evolutionary considerations in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
 % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
@@
 
-Classic phylogenetic regressions usually allow the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects (or levels) of the predictor variables may evolve according to the phylogenetic relationship across species as well.
+Classic phylogenetic regressions usually allow the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects of the predictor variables may evolve according to the phylogenetic relationship across species as well.
 Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
@@
 We begin by describing the classic phylogenetic regression in a linear regression setting.
-Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors encoded in a model matrix $\bX$. 
+Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors encoded in a model matrix $\bX$, where each species is measured exactly once. 
 The standard phylogenetic regression can be formulated as
@@
 where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional modes, is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
-When $\dist$ is Gaussian and $g$ is the identity function, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
+When $\dist$ is Gaussian, $g$ is the identity function, $\bZ$ is the identity matrix, and $\bSigma(\theta)=\sigma^2\bC$, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
 % ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.

### 2019-12-03 phyloglmm_ms.tex
lunching

@@
 
-Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
+Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental covariates) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
@@
 In such cases, researchers typically search for ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
-%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
-%% ord2010adaptation : used OU process with means (single obs per taxa),
-%%  i.e. ignoring within-taxon variation?
-%% fix me!?
-% which can lead to statistical issues and erroneous conclusions
-% \citep{felsenstein1985phylogenies, li2017statistical}.
-% \bmb{What precise ``issues'' do you have in mind?
-Don't think we need to state them here, but we should know.
-Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness?
-(2) not sure whether taking taxon-level means (Ord 2010) is actually a problem.
-Ignores some variation,
-% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
+
 In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships by extending existing software for fitting mixed effect models. 
@@
 
-% Researchers use regression-based methods such as generalized linear models (GLMs) or generalized linear mixed effect models \citep[GLMMs:][]{bolker2009generalized} to control for different levels of variation.
 Standard statistical regression techniques do not allow for correlations among species responses due to their shared evolutionary history.
 Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
-If in addition the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
+If in addition the residuals are normally distributed and observed without any additional error (e.g., from other covariates) or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
 More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than Brownian-motion.
@@
 % \end{center}
+
+%% CONT HERE
 

### 2019-12-04 phyloglmm_ms.tex
minor change and citations

@@
 
-%% CONT HERE
-
 Classic phylogenetic regressions usually allow the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects of the predictor variables may evolve according to the phylogenetic relationship across species as well.
-Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
-Standard phylogenetic regressions allow for phylogenetic correlations in the intercept of the relationship between body and brain size. 
-However, species within taxonomic groups with similar body sizes may vary in overall brain size, or taxonomic groups may vary in the relationship between brain and body size. 
-%\bmb{maybe we can focus on random slopes here?
-Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
-Several recent studies have incorporated phylogenetic variation in different ways and looked at species response to phylogenetic variation with changes in environmental factors.
+For example, suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
+The standard phylogenetic mixed-effect model allows for phylogenetic correlations in the intercept of the relationship between body and brain size. 
+However, species within taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes).
+Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors.
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach.
@@
 Table \ref{table:model} summarizes types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.
+
+%% CONT HERE
 
@@
 In this paper, we will propose an alternative formulation of phylogenetic regression in a generalized linear mixed modeling framework that is mathematically equivalent to previous approaches, but more flexible.
-In particular, our new formulation can be implemented in any framework that allows for random effects (such as random intercepts, slopes, and interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model.
+In particular, our new formulation can be implemented in any framework that allows for random effects (such as random intercepts, slopes, and interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model \citep{hefley2017basis}.
 %% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
 We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
-\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2018brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
+\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} \citep{phyr}, \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2018brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
 % We expect that our approach will be fast as well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 

### 2019-12-06 phyloglmm_ms.tex
going home, cont at Simulation and evaluations

@@
 
-\pagenumbering{gobble}
+% \pagenumbering{gobble}
 % \linenumbers
@@
 However, species within taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes).
+Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).
 Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors.
@@
 Table \ref{table:model} summarizes types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.
-
-%% CONT HERE
 
@@
 & \begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} & Single observation & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
-\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez}, \pkg{phyr}                                                                                            \\ \cline{3-4}
-                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}                                                                                  \\ \hline
+\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4}
+                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}, \pkg{phyr}                                                                                  \\ \hline
 \multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4}
@@
 In this paper, we will propose an alternative formulation of phylogenetic regression in a generalized linear mixed modeling framework that is mathematically equivalent to previous approaches, but more flexible.
-In particular, our new formulation can be implemented in any framework that allows for random effects (such as random intercepts, slopes, and interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model \citep{hefley2017basis}.
+In particular, it easily allows for random effects (such as random intercepts, random slopes, and random interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model \citep{hefley2017basis}.
 %% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
@@
 \end{equation}
-where $\bZ$ is an $n \times m$ model matrix for the $n$-dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional modes, is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
+where $\bZ$ is an $n \times q$ model matrix for the $q$-dimensional vector-valued random effects variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional mean (or mode) of the random effect, which is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
 When $\dist$ is Gaussian, $g$ is the identity function, $\bZ$ is the identity matrix, and $\bSigma(\theta)=\sigma^2\bC$, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
@@
 
-There are forms of random variation in the mixed model framework.
-First, random intercepts can allow the response trait to vary independently across groups other than species (e.g., patches, sites, or experiments). 
-Second, random intercepts can also allow the response (trait or distribution) to vary either independently among species (since species represent the tips of the phylogenetic tree) or among species in a phylogenetically correlated way (i.e., species that are closely related tend to have similar responses).
-The third type of variation that is often neglected is random slopes.
-Random slopes allow fixed effects (the relationship between predictors and responses) to vary among groups.
-Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes can allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).
+%% Check if the two s. above make sense.
+
+%%MLi: The P below does not feel like it belongs here. 
+% There are several forms of random variation in the mixed model framework.
+% First, random intercepts can allow the response trait to vary independently across groups other than species (e.g., patches, sites, or experiments). 
+% Second, random intercepts can also allow the response (trait or distribution) to vary either independently among species (since species represent the tips of the phylogenetic tree) or among species in a phylogenetically correlated way (i.e., species that are closely related tend to have similar responses).
+% The third type of variation are random slopes; they allow fixed effects (the relationship between predictors and responses) to vary among groups.
 
@@
 
+The duality of $\bSigma$ and $\bZ$ for equivalent model specifications is useful for efficient implementations of models that contains a functional covariance structure \citep{hefley2017basis}. 
+Similarly, we can construct phylogenetic correlations (PC) through $\bZ$.
 Suppose that the evolution follows a Brownian-motion process, i.e., continuous traits evolve independently, following an unbiased, continuous-time random walk, along each branch of the phylogeny.
@@
 We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm}) with a single response variable $\bY$ and a single continuous normally distributed predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
-For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e., $\bm{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model.
+For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e., $\cal{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model.
 For the first set of simulations, we simulate one observation per species.
@@
 \begin{aligned}
-\bY & = (\beta_0 + b_{\mathrm{phy_{int}}}) + (\beta_1 + b_{\mathrm{phy_{slope}}}) \bm{t} + \besp \\\
-(b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
+\bY & = \dist(g^{-1}(\bmu), \phi) \\
+\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}}}) \bm{t} + \besp \\\
+(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
 \Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
@@
 \begin{aligned}
-\bY & = (\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}) + (\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}) \bm{t} + b_{\mathrm{sp:group}} + \besp \\
-(b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
+\bY & = \dist(g^{-1}(\bmu), \phi) \\
+\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}}) \bm{t} + \bf{b_{\mathrm{sp:group}}} + \besp \\
+(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
 \Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
@@
 \right) \\
-(b_{\mathrm{sp_{int}}}, b_{\mathrm{sp_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
+(\bf{b_{\mathrm{sp_{int}}}, b_{\mathrm{sp_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
 \sigma^2_{\mathrm{sp_{int}}} & \sigma_{\mathrm{sp_{int-slope}}} \\ 
@@
 \right) \\
-b_{\mathrm{group}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
-b_{\mathrm{sp:group}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \Sigma^2_{phy}) \\
+\bf{b_{\mathrm{group}}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
+\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \Sigma^2_{phy}) \\
 \besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
@@
 
-The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($b_{\mathrm{group}}$) and random intercept of species-group interaction ($b_{\mathrm{sp:group}}$)) compared to the single--group full model.
-Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species; thus we can quantify variation among species separately from residual variation.
-Variance in the intercept of species--group interactions ($b_{\mathrm{sp:group}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
+The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$)) compared to the single--group full model.
+Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species; thus we can distinguish variation among species separately from residual variation.
+Variance in the intercept of species--group interactions ($\bf{b_{\mathrm{sp:group}}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
 
@@
 
-We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, and \pkg{brms} \citep{burkner2018brms}.
+We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
 % \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
@@
 Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes an evolutionary process on the tree (typically Brownian-motion, but other processes can be used) instead of treating the residual error for each species as independent.
-Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow for both phylogenetic and residual variation, as well as non-Gaussian response variables.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are a slightly more flexible variation of PGLS that can allow for both phylogenetic and residual variation, as well as non-Gaussian response variables.
 Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
 Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
-One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle additional random slopes ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercept of species-group interactions ($b_{\mathrm{sp:group}}$) but does not incorporate covariation between phylogenetic random slope-intercept ($\Sigma_{\mathrm{phy_{int-slope}}}$).
+One of the few packages that currently fit phylogenetic slopes to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle additional random slopes ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but does not incorporate covariation between phylogenetic random slope-intercept ($\Sigma_{\mathrm{phy_{int-slope}}}$).
 Lastly, Bayesian phylogenetic GLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
-However, MCMC is usually much more computationally expensive for GLMMs compared to platforms using deterministic optimization.
+However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
 \pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC)\citep{duane1987hybrid}.

### 2019-12-06 phyloglmm_ms.tex
On discussion again

@@
 % \pagenumbering{gobble}
-% \linenumbers
+\linenumbers
 
@@
 \end{center}
-
 \[
@@
 
-We simulated 100 phylogenetic trees for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to and then modelled responses on each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
+We simulated 100 random phylogenetic trees using R package \pkg{ape} \citep{ape} for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to generate and then modelled responses on each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
 Each realization was fitted using all model variants. 
 All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
-Correlation $\rho$ is used in place of covariances in the simulations.
+Correlation $\rho_{xy}$ is used in place of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations.
 Table \ref{table:platform} shows the parameters that are estimable for each platform. 
@@
 For each replication, we sample two chains starting with 10000 iterations. 
-%% TODO MAKE SURE THIS IS CORRECT!
 We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
@@
 We used our method to reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on \pkg{lme4} and \pkg{glmmTMB} (for more details see example in supplement).
-We also used our method and fitted the full model of the dune meadow data recently used with \pkg{pez} \citep{li2017canfun}. 
-\pkg{lme4} and \pkg{pez} give identical results for fixed and random effect estimates, but our code runs approximately 120 times faster than \pkg{pez}. 
-Codes for analyzing the dune meadow data using \pkg{lme4} and \pkg{pez} are provided in the supplements.
 
@@
 \end{center}
-
 \begin{center}
@@
 In general, models that are insufficiently flexible to match the true simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
-PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for the phylogenetic intercept standard deviation parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly overconfident (the coverage $\approx$ 90\% with 100 species) and the fixed slope parameter ($\beta_1$) has poor coverage ($<$ 60\%).
-PGLS, with only one parameter available, confounds all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept and over-covering for $\beta_0$, and under-covering for $\beta_1$.
+PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for the phylogenetic intercept standard deviation parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly overconfident (coverage $\approx$ 90\% with 100 species) and the fixed slope parameter ($\beta_1$) has poor coverage ($<$ 60\%).
+PGLS, with only one parameter available, confounds all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept variation, over-covering for $\beta_0$, and under-covering for $\beta_1$.
 

### 2019-12-07 phyloglmm_ms.tex
why can't I compile the ms

@@
 
-We have simulated relatively complex models containing phylogenetic variation in both intercepts and slopes, as well as within-species variation that is quantifiable because we allow multiple observations per species.
+We have simulated relatively complex models containing phylogenetic variation in both intercepts and slopes, as well as within-species variation that is quantifiable.
 These models are intrinsically more complex than some simple platforms for phylogenetic regression can handle, which may seem unfair; nevertheless, our models are certainly \emph{less} complex than evolutionary processes occurring in nature.
@@
 
-Even our relatively simple models can incorporate many layers of complexity --- e.g.
-multiple spatial grouping variables as well as correlated phylogenetic variation in the effects of several different traits and environmental variables on a focal trait.
+Even our relatively simple multi-group model described in \ref{eq:multiple_glmm} can incorporate many layers of complexity --- i.e., analogous to spatial grouping variables, as well as correlated phylogenetic variation in the effects of several different traits and environmental variables on a focal trait.
 In theory, as long as we have enough data and enough computational power, models that can incorporate more of the complexity will always describe a biological system better.
 However, real applications are always data-constrained.
-Deciding on a practically appropriate level of model complexity for a given
-problem and data set is an
-open and difficult general problem in statistical modeling, not just
-in phylogenetic studies.
+Deciding on a practically appropriate level of model complexity for a given problem and data set is an open and difficult general problem in statistical modeling, not just in phylogenetic studies.
 Should one use simple models that may be overly conservative or
@@
 
-\subsection*{Incorporating different levels of variation}
+\subsection*{Confounding levels of variation}
 
@@
 In classic GLMMs, random effects are used to handle group (or individual) level variation. 
-In the simplest experimental design with continuous response observations in different levels of a discrete grouping variable, where the response may vary among levels of the grouping variable, fitting random intercepts are the ``go-to'' method to handle this variation.
-The random intercept model controls the group effect in the response level by allowing different intercepts for each level of the grouping variable.
-However, random effects can take more complicated forms as the experimental design becomes more complex.  
-For example, imagine observing another continuous explanatory variable in the experimental design above, where the relationship of the response and the new explanatory variable may vary according to the grouping variable.
-A random slopes model, which allows different slopes (the relationship between continuous variables) for each level of the grouping variable is most appropriate to handle this type of variation.
-Random slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each evolutionary group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
-Neglecting random slopes can lead to biased fixed effect estimates with inadequate coverage and type I errors \citep{schielzeth2008conclusions} as shown in the simulations above.
+In the simplest experimental design with continuous response observations with variations between different levels of discrete grouping variable -- i.e., where the response may vary among levels of the grouping variable -- fitting random intercepts are the ``go-to'' method.
+% The random intercept model models the group effect in the response level by allowing different intercepts for each level of the grouping variable.
+However, random effects can take more complicated forms as the experimental design becomes more complex, e.g., random slopes models.
+Random slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+Neglecting random slopes can lead to biased fixed effect estimates with inadequate coverage and type I errors \citep{schielzeth2008conclusions}.
 
-% When analyzing relationships among species between traits, it is entirely plausible that change in the effect of predictor variables have on traits are changing with respect to different grouping of species or even in a phylogenetically correlated way (phylogenetic random slopes).
-Nevertheless, it is hard to account for all forms of complexities and decide if when it is best to use phylogenetic random effects, simple grouping, or both. 
+Nevertheless, for PGLMMs, it is hard to account for all forms of complexities and decide if when it is best to use phylogenetic random effects, simple grouping, or both. 
 Optimal model complexity depends on experimental design and whether the data provides enough signal to estimate these different levels of variation, which can be strongly confounded.
@@
 ML and BMB conceived the ideas and designed methodology; ML and BMB implemented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analyzed the results; ML wrote the first draft of the manuscript.
-All authors contributed critically to the drafts and gave final approval for publication.
+Both authors contributed critically to the drafts and gave final approval for publication.
 

### 2019-12-07 dune_pezlme4.R

@@ -31,11 +31,11 @@ dat <- (dat
 ## If you hack mkBlist in phylo_lmm and switch the order in the kronecker product step (bad!)
 lme4time_1 <- system.time(
   lme4fit_1 <- phylo_lmm(Y ~ 1 + log.sla + annual 
-+ (1|obs) 
-+ (1|sp)
+# + (1|obs) 
+# + (1|sp)
 + (1 | sp:site)
-+ (0 + log.sla | site)
-+ (1|site) 
+# + (0 + log.sla | site)
+# + (1|site) 
 , data=dat
 , phylonm = c("sp","sp:site")
 , phyloZ=phyZ
@@ -79,30 +79,30 @@ print(summary(lme4fit_1))
 print(peztime_1/lme4time_1)
 
 
-# peztime_1 <- system.time(
-#   pezfit_1 <-  communityPGLMM(formula = "Y ~ 1 + log.sla + annual"
-#      , data = dat
-#      , family = "gaussian"
-#      , sp = dat$sp
-#      , site = dat$site
-#      , random.effects = list(#re.sp.phy
-#     # , re.sp
-#     re.nested.phy
-#     # , re.sla
-#     # , re.site
-#      )
-#      , REML = F
-#      , verbose = F
-#      # , s2.init = c(1.5, rep(0.01, 4))
-#   )
-# )
-# 
+peztime_1 <- system.time(
+  pezfit_1 <-  communityPGLMM(formula = "Y ~ 1 + log.sla + annual"
+     , data = dat
+     , family = "gaussian"
+     , sp = dat$sp
+     , site = dat$site
+     , random.effects = list(#re.sp.phy
+    # , re.sp
+    re.nested.phy
+    # , re.sla
+    # , re.site
+     )
+     , REML = F
+     , verbose = F
+     # , s2.init = c(1.5, rep(0.01, 4))
+  )
+)
+
 chol_time <- system.time(cholphyt <- t(chol(phyZ %*% t(phyZ))))
 
 
 hackedmod <- glmmTMBhacked(Y ~ 1 + log.sla + annual
-  # + (1|obs)
-  + (1|sp)
+  + (1|obs:site)
+  # + (1|sp)
   # + (0 + log.sla | site)
   # + (1|site)
   # + (1 | sp:site)
@@ -115,3 +115,5 @@ hackedmod <- glmmTMBhacked(Y ~ 1 + log.sla + annual
   , REML = FALSE
   , lambhack=TRUE
 )
+
+hackedmod

### 2019-12-07 glmmTMBhacked.R

@@ -223,7 +223,8 @@ mkTMBStruchacked <- function (formula, ziformula, dispformula, combForm, mf, fr,
   parameters$b <- rep(0,ncol(data.tmb$Z))
   if(lambhack){
     # condList$Z <- t(condListhacked$reTrms$Zt %*% chol(phyloZ %*% t(phyloZ)))
-    condList$Z <- t(condListhacked$reTrms$Zt) %*% phyloZ
+    condList$Z <- t(condListhacked$reTrms$Zt) %*% kronecker(phyloZ, Diagonal(nsite))
+    condList$Z <- t(condListhacked$reTrms$Zt) %*% kronecker(Diagonal(nsite),phyloZ)
     data.tmb$Z <- condList$Z
     # parameters$b <- rep(0, length(data.tmb$Z != 0))
   }

### 2019-12-07 phylointeraction.R

@@ -1,0 +1,143 @@
+### simulate phylogenetic tree
+
+library(ape)
+library(MASS)
+library(sparseMVN)
+library(Matrix)
+library(dplyr)
+
+quadform <- function(sd_mat,cormat){
+  sd_mat %*% cormat %*% sd_mat
+}
+
+tree_seed <- 101
+set.seed(tree_seed)
+
+nspp <- 10
+nsite <- 5
+nrep <- 3
+
+phy <- rtree(n = nspp)
+
+Vphy <- vcv(phy)
+
+# Generate index data frame 
+# Species varying fastest and sites varying slowest
+# Species nested within site 
+
+interactions <- interaction(1:nspp,1:nsite)
+
+indexdat <- (data.frame(ints = levels(interactions))
+	%>% rowwise()
+	%>% mutate(ints = as.character(ints)
+		, site = unlist(strsplit(ints,"[.]"))[2]
+		, sp = unlist(strsplit(ints,"[.]"))[1]
+	)  
+)
+
+print(head(indexdat))
+
+# Species frame: 
+## Generate phylogenetic intercept and slope
+
+beta0 <- 0
+beta1 <- 0
+physd_B0 <- 10
+physd_B1 <- 8
+rho_phy <- 0.5
+
+phycormat <- matrix(c(1,rho_phy,rho_phy,1),2,2)
+physdmat <- diag(c(physd_B0,physd_B1))
+phycovmat <- quadform(physdmat,phycormat)
+
+phySigma <- kronecker(phycovmat,Vphy)  ## phylo blocks
+# phySigma <- kronecker(Vphy,phycovmat)
+
+b_phy <- MASS::mvrnorm(n=1
+	, mu=rep(c(beta0,beta1),each=nspp)
+	, Sigma=phySigma
+)
+
+
+# Generate random intercept and slope
+
+sd_B0 <- 6
+sd_B1 <- 3
+rho <- 0.7
+
+cormat <- matrix(c(1, rho, rho, 1), 2, 2)
+
+sdmat <- diag(c(sd_B0, sd_B1))
+covmat <- quadform(sdmat,cormat)
+
+bSigma <- kronecker(covmat,diag(nspp))
+
+b <- MASS::mvrnorm(n=nspp
+                   , mu=c(beta0, beta1)
+                   , Sigma = covmat
+                   , empirical = TRUE)
+
[…]

### 2019-12-07 plot.R

@@ -59,13 +59,16 @@ gg_ss <- (ggplot(data=ssdat, aes(x=size, y=sd, col=Platform))
 
 print(gg_ss)
 
+scaleFUN <- function(x) sprintf("%.2f", x)
+
 gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform))
 	# + geom_violin(position=position_dodge(width=0.2),alpha=0.4)
 	+ geom_boxplot(outlier.colour = NULL, varwidth=TRUE)
 	# + scale_fill_brewer(palette = "Dark2")
-	+ scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x)
-	    , labels = trans_format("log10", math_format(10^.x))
-	  )
+	# + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x)
+	#     , labels = trans_format("log10", math_format(10^.x))
+	#   )
+	+ scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),labels = scaleFUN)
 	+ scale_color_manual(values=c("Black","Red","Dark Blue","Dark Green","Orange"))
 	
 	+ ylab("Time (seconds)")

### 2020-01-14 phyloglmm_ms.tex
some tweaking

@@
 \item{Our reformulation of phylogenetic generalized linear mixed models is computationally efficient, operating orders of magnitude faster than existing methods for fitting phylogenetic mixed models.}
-\item{Our approach can, in principle, be implemented in any platform for generalized mixed models.
+\item{Our approach can be implemented in any platform for generalized mixed models.
 Our implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to a broad range of previously difficult cases (e.g., large data, unbalanced observational designs, complex random effects).}
@@
 %%% PCM accounts for evolutionary relationships. More people are doing it.
-Phylogenetic comparative methods (PCMs) are a powerful technique to explore relationships among related groups of species traits.
-Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
+Phylogenetic comparative methods (PCMs) are a powerful technique to account for evolutionary relatedness when analyzing species traits.
+Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
 
@@
 
-Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental covariates) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
+Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental covariates) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
 % While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
 While a wide range of tools is available for phylogenetic regression, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
-In such cases, researchers typically search for ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
+When faced with these constraints, researchers typically search for ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
 
@@
 
-Standard statistical regression techniques do not allow for correlations among species responses due to their shared evolutionary history.
+Standard statistical regression techniques neglect correlations among species responses due to their shared evolutionary history.
 Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
-If in addition the residuals are normally distributed and observed without any additional error (e.g., from other covariates) or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
-More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than Brownian-motion.
+If in addition the residuals are normally distributed and observed without any additional error (e.g., from other covariates \bmb{???}) or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
+\bmb{what's the second citation for?}
+More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than the simple Brownian motion model.
 These methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
@@
 Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
-Indeed, many existing methods are restricted to single observations per species, requiring users to collapse multiple observations per species to species mean values.
+Indeed, many existing methods are restricted to single observations per species, requiring users to collapse their observations to mean values for each species, which can be limiting if some covariates of interest vary within species.
 %% REMEMBER TO REVISIT IN DISCUSSION
@@
 
-Classic phylogenetic regressions usually allow the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects of the predictor variables may evolve according to the phylogenetic relationship across species as well.
+Classic phylogenetic regressions usually allow the uncertainty in the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects of the predictor variables may evolve according to the phylogenetic relationship across species as well.
 For example, suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
@@
 For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach.
-The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, many biologist needing to fit random-slopes model have usually turned to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
-Table \ref{table:model} summarizes types of platforms, data constraints, and provides model complexities for phylogenetic comparative analysis.
+The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, biologists needing to fit random-slopes model typically turn to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
+Table \ref{table:model} summarizes currently available methods for phylogenetic comparative analysis.
 
@@
 \end{tabular}
-\caption{List of phylogenetic generalized linear models and R packages.}
+\caption{GLM-based phylogenetic comparative methods and their implementations in R packages.
+\bmb{can we add ``per species'' to ``single observation'' without wrecking the table?}}
 \label{table:model}
@@
 
-In this paper, we will propose an alternative formulation of phylogenetic regression in a generalized linear mixed modeling framework that is mathematically equivalent to previous approaches, but more flexible.
-In particular, it easily allows for random effects (such as random intercepts, random slopes, and random interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model \citep{hefley2017basis}.
+In this paper, we propose an alternative formulation of phylogenetic regression in a generalized linear mixed modeling framework that is mathematically equivalent to previous approaches, but more flexible.
+In particular, it easily allows for complex phylogenetic effects (random intercepts, slopes, and interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model \citep{hefley2017basis}.
 %% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
@@
 
-We generated test data using a simple framework that combines fixed effects, phylogenetic random intercept, slope, and their correlations from a phylogenetic tree, as well as residuals. 
-We then fit the test data from these simulations using our approach implemented using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}, using assumptions that match those of the simulation model.
-We also fit with other platforms, using standard simplifications when necessary for implementation.
+We generated test data using a simple framework that combines fixed effects, phylogenetic random intercepts, slopes, and their correlations based on a simulated phylogenetic tree.
+We fit models to these test data using our approach built on the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}, using assumptions that match those of the simulation model.
+We also fit with other platforms (Table~\ref{table:model}), simplifying the statistical model when necessary.
 
@@
 
-We simulated 100 random phylogenetic trees using R package \pkg{ape} \citep{ape} for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) to generate and then modelled responses on each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
+Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) and then simulated the responses for each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
 Each realization was fitted using all model variants. 
 All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
-Correlation $\rho_{xy}$ is used in place of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations.
+Correlation $\rho_{xy}$ is used in place of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations \bmb{???}.
 Table \ref{table:platform} shows the parameters that are estimable for each platform. 
@@
 Models capable of fitting all parameters (\pkg{lme4},  \pkg{glmmTMB}, and \pkg{brms}) fit well for all parameters.
+    \bmb{add a light fill (possibly the same colour as the box edge with alpha=0.6?)? See https://serialmentor.com/dataviz/avoid-line-drawings.html}
 %\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
@@
 
-\subsection*{Extension and alternatives}
+\subsection*{Extensions and alternatives}
 
@@
 phylogenetic least squares, linear and mixed models) to fit models that incorporate different levels of phylogenetic variation.
-Even within this scope, there is additional room for exploration, such as phylogenetic multivariate response models; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck (OU) model which accounts for both selection and drift processes \citep{butler2004phylogenetic}); Bayesian approaches \citep{hadfield2010general}; and variable-rate model, where evolutionary parameters vary across the phylogeny.
-However, the simple approach we developed here offers an efficient way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models. 
-This approach can in principle be combined flexibly with any platforms that supports independent latent variables such as Stan.
+Even within this scope, there is additional room for exploration, such as phylogenetic multivariate response models; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck (OU) model which accounts for both selection and drift processes \citep{butler2004phylogenetic}); Bayesian approaches \citep{hadfield2010general}; and variable-rate models, where evolutionary rates vary across the phylogeny.
+The simple approach we developed here, while it cannot easily incorporate these complexities, does offer an efficient way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models.
+This approach can in principle be combined flexibly with any platform that supports independent latent variables.
 More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.

### 2020-01-14 phyloglmm_ms.tex
bmb tweaks/comments

@@
 
-The duality of $\bSigma$ and $\bZ$ for equivalent model specifications is useful for efficient implementations of models that contains a functional covariance structure \citep{hefley2017basis}. 
+The duality of $\bSigma$ and $\bZ$ for equivalent model specifications is useful for efficient implementations of models that contain a functional \bmb{???} covariance structure \citep{hefley2017basis}.
 Similarly, we can construct phylogenetic correlations (PC) through $\bZ$.
@@
 We have simulated relatively complex models containing phylogenetic variation in both intercepts and slopes, as well as within-species variation that is quantifiable.
-These models are intrinsically more complex than some simple platforms for phylogenetic regression can handle, which may seem unfair; nevertheless, our models are certainly \emph{less} complex than evolutionary processes occurring in nature.
+These models are intrinsically more complex than some simple platforms for phylogenetic regression can handle, which may seem unfair; nevertheless, our models are certainly less complex than evolutionary processes occurring in nature.
 Our results show that models that cannot match the full ``simulation world'' perform poorly even for the parameters they do estimate; it is important to understand the limitations of these simpler, commonly used methods.
@@
 Deciding on a practically appropriate level of model complexity for a given problem and data set is an open and difficult general problem in statistical modeling, not just in phylogenetic studies.
-Should one use simple models that may be overly conservative or
+For example, should one use simple models that may be overly conservative or
 risk overfitting by using more ambitious models?
-How can one appropriately use the data themselves
-to choose model complexity \citep{roberts_cross-validation_2016}?
 What are the relative costs and benefits of using a step-down procedure
@@
 or using Bayesian approaches with regularizing priors \citep{hadfield2010mcmc}?
+For predictive models, how can one appropriately evaluate out-of-sample accuracy for
+structured data \citep{roberts_cross-validation_2016}?
 
-\subsection*{Confounding levels of variation}
+\subsection*{Incorporating multiple levels of variation}
 
@@
 
-In classic GLMMs, random effects are used to handle group (or individual) level variation. 
-In the simplest experimental design with continuous response observations with variations between different levels of discrete grouping variable -- i.e., where the response may vary among levels of the grouping variable -- fitting random intercepts are the ``go-to'' method.
+In classic GLMMs, random effects are used to account for group-level variation. 
+In the simplest experimental designs where only the response and not the covariates vary
+within groups, random-intercept models (or even simpler analyses \cite{murtaugh2007simplicity})
+are appropriate.
 % The random intercept model models the group effect in the response level by allowing different intercepts for each level of the grouping variable.
-However, random effects can take more complicated forms as the experimental design becomes more complex, e.g., random slopes models.
-Random slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
-Neglecting random slopes can lead to biased fixed effect estimates with inadequate coverage and type I errors \citep{schielzeth2008conclusions}.
+However, random effects can take more complicated forms as the experimental design becomes more complex, e.g., random-slopes models.
+Random-slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+Neglecting random slopes can lead to biased fixed effect estimates with inadequate coverage and inflated type~I errors \citep{schielzeth2008conclusions}.
 
-Nevertheless, for PGLMMs, it is hard to account for all forms of complexities and decide if when it is best to use phylogenetic random effects, simple grouping, or both. 
+Nevertheless, for PGLMMs, it is hard to account for all forms of complexities and decide when it is best to use phylogenetic random effects, simple grouping, or both. 
 Optimal model complexity depends on experimental design and whether the data provides enough signal to estimate these different levels of variation, which can be strongly confounded.
@@
 
-It may be easier to be conservative to include both (phylogenetic random effects and grouping random effects) of them but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model.
+It may be easier to be conservative and include both phylogenetic random effects and grouping random effects
+\bmb{clarify? ``iid random effects''?}
+but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
+In this case, the PGLMM collapses to a standard random-slopes model where the group effects are considered to be independent and identically distributed.
 
-Another simplifying alternative is to be conservative to include both phylogenetic random effects and grouping random effects but simplify the phylogenetic relationships and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
+Another simplifying alternative is include both phylogenetic random effects and grouping random effects but simplify the phylogenetic relationships and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
 However, this collapses phylogenetic structures to a standard random-slopes model. 

### 2020-01-20 phyloglmm_ms.tex
tweaking text and plotting code

@@
 The horizontal line shows the true value of the parameters in the simulation model.
-Models capable of fitting all parameters (lme4 and glmmTMB) fit well for all parameters.
-\pkg{pez} and \pkg{phyr} estimates for n = 100, and 500 are not available because the models did not converge within 30 minutes.
-%    \bmb{need to work harder to match up colours across plots - colours corresponding to specific platforms are different in different plots because different subsets are included (use \code{scale\_colour\_manual} appropriately)}
+Models capable of fitting all parameters (\pkg{lme4} and \pkg{glmmTMB}) fit well for all parameters.
+\pkg{pez} and \pkg{phyr} estimates for n = 100 and 500 are not available because the models did not converge within 30 minutes.
   }
@@
 For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4} for the cases studied here.
-Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit 50 species model is $\approx$ 9 versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr} respectively. 
-\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical for us to fit 500-species models with \pkg{pez} and \pkg{phyr}, because computational speed scaled faster than linearly with sample size.
+Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit a 50-species model is $\approx$ 9 versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr} respectively. 
+\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical to fit 500-species models with \pkg{pez} and \pkg{phyr}, because computational speed scaled faster than linearly with sample size.
 \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.

### 2020-01-20 plot.R

@@ -41,16 +41,22 @@ ssdat <- (ssdat_raw
 	# %>% filter(platform != "brms")
 )
 
+## FIXME: could match these more programatically (use 'values' in
+## scale_colour_manual() ?)
+colvec_all <-  c(gls="Black",phylolm="Red",lme4="Dark Blue",
+                 glmmTMB="Dark Green",brms="Orange",pez="Gray",phyr="Purple")
+colvec  <- colvec_all[c("gls","phylolm","lme4","glmmTMB","brms")]
+colvec2 <- colvec_all[c("pez","phyr","lme4","glmmTMB")]
 
-gg_ss <- (ggplot(data=ssdat, aes(x=size, y=sd, col=Platform))
+gg_ss <- (ggplot(data=ssdat, aes(x=size, y=sd, col=Platform, fill=Platform))
   + facet_wrap(~sdtype, scale="free_y", labeller = label_parsed)
   # + geom_violin(position=position_dodge(width=0.5),alpha=0.4)
-  + geom_boxplot(outlier.colour = NULL, varwidth=TRUE)
-  + geom_hline(aes(yintercept = y_int))
-  # + scale_fill_manual(values=c("Dark Red","Dark Blue","Dark Green","Purple"))
+  + geom_boxplot(outlier.colour = NULL, varwidth=TRUE, alpha=0.2)
+  + geom_hline(aes(yintercept = y_int), lty=2)
+  + scale_fill_manual(values=colvec)
   # + scale_fill_brewer(palette = "Dark2")
   # + scale_color_brewer(palette = "Dark2")
-  + scale_color_manual(values=c("Black","Red","Dark Blue","Dark Green","Orange"))
+  + scale_color_manual(values=colvec)
   # + ggtitle("Single site")
   + theme_bw()
   + ylab("Parameter Estimates")
@@ -61,16 +67,17 @@ print(gg_ss)
 
 scaleFUN <- function(x) sprintf("%.2f", x)
 
-gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform))
+gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform,
+                                     fill=Platform))
 	# + geom_violin(position=position_dodge(width=0.2),alpha=0.4)
-	+ geom_boxplot(outlier.colour = NULL, varwidth=TRUE)
+	+ geom_boxplot(outlier.colour = NULL, varwidth=TRUE, alpha=0.2)
 	# + scale_fill_brewer(palette = "Dark2")
[…]

### 2020-01-20 phyloglmm_ms.tex
forgot to merge with BMB's recent changes, merging now

@@
 Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
-If in addition the residuals are normally distributed and observed without any additional error (e.g., from other covariates \bmb{???}) or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
-\bmb{what's the second citation for?}
+If in addition the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
 More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than the simple Brownian motion model.
@@
 \begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)
-\end{tabular}} & Correlated residual & Single observation & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4} 
-& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} & Single observation & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
-\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4}
+\end{tabular}} & Correlated residual & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4}
+& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
+\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation \\per species, \\ Balanced design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4}
                                                                                                   &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}, \pkg{phyr}                                                                                  \\ \hline

### 2020-02-07 phyloglmm_ms.tex
fixed all figures, just three more bmb comments left

@@
 \end{tabular}
-\caption{GLM-based phylogenetic comparative methods and their implementations in R packages.
-\bmb{can we add ``per species'' to ``single observation'' without wrecking the table?}}
+\caption{GLM-based phylogenetic comparative methods and their implementations in R packages.}
 \label{table:model}
@@
 All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
-Correlation $\rho_{xy}$ is used in place of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations \bmb{???}.
+Correlation $\rho_{xy}$ is used instead of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations \bmb{???}.
 Table \ref{table:platform} shows the parameters that are estimable for each platform. 
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.7,page=1]{./figure/ssplot.pdf}
+  \includegraphics[scale=0.7]{./figure/ssplot.pdf}
   \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
@@
 Models capable of fitting all parameters (\pkg{lme4},  \pkg{glmmTMB}, and \pkg{brms}) fit well for all parameters.
-    \bmb{add a light fill (possibly the same colour as the box edge with alpha=0.6?)? See https://serialmentor.com/dataviz/avoid-line-drawings.html}
 %\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.7,page=1]{./figure/msplot.pdf}
+  \includegraphics[scale=0.7]{./figure/msplot.pdf}
   \caption{Comparison of multi-group model parameter estimates.
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8]{./figure/mstime.pdf}
+  \includegraphics[scale=0.7]{./figure/mstime.pdf}
   \caption{Comparison of multi-group model computational speed.}
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.5]{./figure/mscoverage.pdf}
+  \includegraphics[scale=0.7]{./figure/mscoverage.pdf}
   \caption{Comparison of multi-group model coverage.}

### 2020-02-07 plot.R

@@ -64,6 +64,7 @@ gg_ss <- (ggplot(data=ssdat, aes(x=size, y=sd, col=Platform, fill=Platform))
 )
 
 print(gg_ss)
+ggsave(plot = gg_ss,filename = "figure/ssplot.pdf",width = 10, height=7)
 
 scaleFUN <- function(x) sprintf("%.2f", x)
 
@@ -84,6 +85,8 @@ gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform,
 )
 
 print(gg_sstime)
+ggsave(plot = gg_sstime,filename = "figure/sstime.pdf",width = 10, height=7)
+
 
 b_ci <- function(x,w) {
     x <- na.omit(x)
@@ -136,6 +139,7 @@ gg_sscoverage <- (ggplot(data=ss_coverage
 )
 
 print(gg_sscoverage)
+ggsave(plot = gg_sscoverage,filename = "figure/sscoverage.pdf",width = 10, height=5)
 
 mspar_df <- data.frame(
   sdtype = c("resid", "phylo_X", "phylo_int", "phylo_cor", "phylo_interaction"
@@ -175,10 +179,14 @@ msdat <- (msdat_raw
 gg_ms <- (gg_ss
 	%+% msdat
 	+ scale_color_manual(values=colvec2)
+	+ scale_fill_manual(values=colvec2)
 
 )
 
 print(gg_ms)
+ggsave(plot = gg_ms,filename = "figure/msplot.pdf",width = 10, height=7)
+
+
 
 gg_mstime <- (gg_sstime
 	%+% msdat
@@ -188,15 +196,18 @@ gg_mstime <- (gg_sstime
 )
 
 print(gg_mstime)
+ggsave(plot = gg_mstime,filename = "figure/mstime.pdf",width = 10, height=7)
+
 
 ## BMB: see revised code above
 ms_coverage <- (msdat
-                %>% filter(!(sd %in% c(-1,1)))
-                %>% group_by(Platform, size)
-                %>% summarise(B0_coverage = mean(B0, na.rm=TRUE)
-                              , B1_coverage = mean(B1, na.rm=TRUE)
-                )
-                %>% gather(key=fixed_parameter, value=coverage, -c(Platform, size))
+                # %>% filter(!(sd %in% c(-1,1)))
+                %>% as_tibble()
+                %>% tidyr::gather(key=fixed_parameter, value=cov, B0,B1)
[…]

### 2020-02-07 phyloglmm_ms.tex
pushing edits, I think we are done

@@
 
-The duality of $\bSigma$ and $\bZ$ for equivalent model specifications is useful for efficient implementations of models that contain a functional \bmb{???} covariance structure \citep{hefley2017basis}.
+The duality of $\bSigma$ and $\bZ$ for equivalent model specifications is useful for efficient implementations of models that contain a functional (e.g., autoregressive) covariance structure \citep{hefley2017basis}.
 Similarly, we can construct phylogenetic correlations (PC) through $\bZ$.
@@
 All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
-Correlation $\rho_{xy}$ is used instead of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations \bmb{???}.
+Correlation $\rho_{xy}$ is used instead of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations.
+%\bmb{???}
+%mli{Just to match what lme4/glmmtmb outputs.}
 Table \ref{table:platform} shows the parameters that are estimable for each platform. 
@@
 Alternatively, when the within-species variance is actually of interest, accounting for within-species variation (i.e., adding species-level random effects in our example) can automatically handle multiple observations per species.
-
-It may be easier to be conservative and include both phylogenetic random effects and grouping random effects
-\bmb{clarify? ``iid random effects''?}
-but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
+It may be easier to be conservative and include both phylogenetic random effects and iid grouping random effects by simplifying the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
 In this case, the PGLMM collapses to a standard random-slopes model where the group effects are considered to be independent and identically distributed.

### 2020-02-11 abstract.tex
first wordcutting attempt: down to 3601 words

--- 2020-02-07 phyloglmm_ms.tex
@@
-\documentclass[12pt]{article}
-\usepackage{graphicx}
-\usepackage{float}
-\usepackage{grffile}
-\usepackage{color}
-\usepackage{soul}
-\usepackage{tabularx}
-\usepackage{multirow}
-% use \xspace to allow for space after a macro as necessary
-\usepackage{xspace}
-\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
-\usepackage{pdfpages}
-\usepackage{lineno}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{commentcol}{rgb}{0.345, 0.345, 0.945}
-\newcommand{\comment}[1]{\textcolor{commentcol}{[[#1]]}}%
-
-% general-purpose mathrm macros
-\newcommand{\symsub}[2]{\ensuremath{#1_{\tiny \mathrm{#2}}}\xspace}
-\newcommand{\mrm}[1]{\ensuremath{\mathrm{#1}}
-\xspace}
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{randomcolor}{rgb}{.97, .27, .67}
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-%% hard to sort by year automatically ...
-%% https://tex.stackexchange.com/questions/377449/sorting-citations-and-bibliography-using-natbib-and-plainnat?rq=1
-\usepackage[sort]{natbib}
-\usepackage{blkarray}
-\usepackage{bm}
-\usepackage{amsmath,empheq}
-\usepackage{alltt}
-\usepackage{hyperref}
-\usepackage[utf8]{inputenc} % for accented characters
-%% stuff for editing
-%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
-%% to suppress notes & comments: \usepackage[final]{changes}
-\usepackage{setspace}
-\usepackage{changes}
-\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
-
-\bibliographystyle{chicago}
-\title{Reformulating phylogenetic mixed models to improve flexibility and speed}
-\author{Michael Li and Ben Bolker}
-\date{\today}
-
-\providecommand{\keywords}[1]{\textbf{\textit{Keywords:}} #1}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
- \vspace{-8ex}
-  \date{}
-\begin{document}
-\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}
-
-%% don't typeset BMB comments
-\newcommand{\bmbhide}[1]{}
-\newcommand{\bmb}[1]{{\color{blue} BB: #1}}
-
-\renewcommand{\figurename}{Fig.}
-
-\newcommand{\mli}[1]{{\color{red} ML: #1}}
-
-\newcommand{\add}[1]{{\color{blue} ADD: #1}}
-
-\newcommand{\pkg}[1]{{\tt #1}}
-\newcommand{\code}[1]{{\tt #1}}
-
-% \pagenumbering{gobble}
-\linenumbers
-
-% \renewcommand{\bmb}[1]{\relax}
-% \renewcommand{\mli}[1]{\relax}
-
-%\SweaveOpts{concordance=TRUE}
-%\SweaveOpts{concordance=TRUE}
-% \maketitle
-
-\begin{center}
-{\Large
-\textbf\newline{Reformulating phylogenetic mixed models to improve flexibility and speed} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-\vspace{2pt}
-Michael Li\textsuperscript{1*} and Benjamin M. Bolker\textsuperscript{1,2,3}
-\\
-\vspace{2pt}
-* Corresponding author: Michael Li; lim88@mcmaster.ca
-\end{center}
-\vspace{5pt}
-\textbf{1} Department of Biology, McMaster University, Hamilton, Ontario, Canada
-\\
-\textbf{2} Department of Mathematics and Statistics, McMaster University, Hamilton, Ontario, Canada
-\\
-\textbf{3} Institute for Infectious Diseases Research, McMaster University, Hamilton, Ontario, Canada
-\\
-\bigskip
-
-\doublespacing
-
 \section*{Abstract}
@@
 \keywords{phylogenetic comparative methods, phylogenetic correlation, phyloglmm, species--branch matrix}
-
-\section*{Introduction}
-
-%%% PCM accounts for evolutionary relationships. More people are doing it.
-Phylogenetic comparative methods (PCMs) are a powerful technique to account for evolutionary relatedness when analyzing species traits.
-Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
-
-Ever-increasing data collection capabilities (e.g., genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic databases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
-Researchers use these data to fit complex models describing species occurrence and traits.
-For example, ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate evolutionary considerations in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
-% \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
-how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature?
-\ldots}
-
-Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental covariates) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
-% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
-While a wide range of tools is available for phylogenetic regression, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
-When faced with these constraints, researchers typically search for ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
-
-In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships by extending existing software for fitting mixed effect models. 
-This method allows researchers to easily incorporate evolutionary and statistical complexities without sacrificing speed.
-
-\subsection*{Challenges in modeling phylogenetic processes}
-
-Standard statistical regression techniques neglect correlations among species responses due to their shared evolutionary history.
-Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
-If in addition the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
-More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than the simple Brownian motion model.
-These methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
-If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation can both be considered part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
-%\bmb{This is a good point, well expressed.
-Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel?
-Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
-%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
-Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
-Indeed, many existing methods are restricted to single observations per species, requiring users to collapse their observations to mean values for each species, which can be limiting if some covariates of interest vary within species.
-%% REMEMBER TO REVISIT IN DISCUSSION
-
-% Figure \ref{fig:variation} illustrates various forms of variation for three species. 
-%Despite having multiple observations per tip taxon (species), a typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
-% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
-% 
-% \begin{center}
-% \begin{figure}[H]
-%   \includegraphics[scale=0.8]{./phylo_diagram.png}
-%   \caption{Potential forms of phylogenetic variation of three species.
-Top left shows pure noise (i.e.
-assuming a star phylogeny); top right accounts for phylogeny, but without noise; bottom left accounts both phylogeny and noise at the tip, and finally bottom right shows multiple observations (with noise) at the tips.}
-% \label{fig:variation}
-% \end{figure}
-% \end{center}
-
-Classic phylogenetic regressions usually allow the uncertainty in the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects of the predictor variables may evolve according to the phylogenetic relationship across species as well.
-For example, suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
-The standard phylogenetic mixed-effect model allows for phylogenetic correlations in the intercept of the relationship between body and brain size. 
-However, species within taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes).
-Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).
-Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors.
-For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach.
-The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, biologists needing to fit random-slopes model typically turn to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
-Table \ref{table:model} summarizes currently available methods for phylogenetic comparative analysis.
-
-%% FIXME: chart looks ugly
-\begin{table}[]
-\begin{tabular}{|l|l|l|l|}
-\hline
-\textbf{Model} & \textbf{Method} & \textbf{Data} & \textbf{Platform} \\ \hline
-\multirow{2}{*}{
-\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)
-\end{tabular}} & Correlated residual & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4}
-& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
-\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation \\per species, \\ Balanced design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4}
-                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}, \pkg{phyr}                                                                                  \\ \hline
-\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4}
-                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
-\end{tabular}
-\caption{GLM-based phylogenetic comparative methods and their implementations in R packages.}
-\label{table:model}
-\end{table}
-
-In this paper, we propose an alternative formulation of phylogenetic regression in a generalized linear mixed modeling framework that is mathematically equivalent to previous approaches, but more flexible.
-In particular, it easily allows for complex phylogenetic effects (random intercepts, slopes, and interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model \citep{hefley2017basis}.
-%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
-We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
-\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} \citep{phyr}, \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2018brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
-% We expect that our approach will be fast as well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
-%% TODO LAST LINE SEE PAGE 5
-
-%% FIXME: move to methods/simulations
-% In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
-% We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.
-
-\section*{Materials and Methods}
-
-\newcommand{\bX}{{\mathbf X}}
-\newcommand{\bbeta}{{\boldsymbol \beta}}
-\newcommand{\bmu}{{\boldsymbol \mu}}
-\newcommand{\bY}{{\mathbf y}}  %% fixme later
-\newcommand{\bC}{{\mathbf C}}
-\newcommand{\bZ}{{\mathbf Z}}
-\newcommand{\bb}{{\mathbf b}}
-\newcommand{\besp}{{\boldsymbol \epsilon}}
-\newcommand{\bSigma}{{\boldsymbol \Sigma}}
-
-We generated test data using a simple framework that combines fixed effects, phylogenetic random intercepts, slopes, and their correlations based on a simulated phylogenetic tree.
-We fit models to these test data using our approach built on the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}, using assumptions that match those of the simulation model.
-We also fit with other platforms (Table~\ref{table:model}), simplifying the statistical model when necessary.
-
-\subsection*{Phylogenetic regression}
-
-We begin by describing the classic phylogenetic regression in a linear regression setting.
-Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors encoded in a model matrix $\bX$, where each species is measured exactly once. 
-The standard phylogenetic regression can be formulated as
-
-%\bmb{boldface matrices/vectors i.e.
-$\bX \bbeta$?
-or not worth the trouble?
-How about $\bmu=\bX \bbeta$; $\bY \sim \textrm{MVN}(\bmu,\sigma^2 \bC)$ for greater generality?
-(OK, I see that you move in this direction later.
-But I'm not sure it's worth the detour of using two different notations?
-even if you are going to extend to include fixed effects, $Z$, etc.
-later ...}:
-
-\begin{equation}
-\begin{aligned}
-\bY & = \bX \bbeta + \besp  \\
-\besp & \sim \textrm{MVN}(0,\sigma^{2} \bC), 
-\label{eq:gls}
-\end{aligned}
-\end{equation}
-where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic covariance (PC) matrix.
-The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.
-
-%\bmb{give a ref for this? maybe Paradis's ape book, or Garamszegi's book, or ?}
-%\mli{chapter 7 in garamszegi 2014}
-
-% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
-%\mli{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
-
-\subsection*{Phylogenetic generalized linear mixed model}
-Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework to define a wider range that includes the standard phylogenetic regression as a special case \citep{lynch1991methods}.
-The generalized linear mixed effect model allows for non-Gaussian responses and uses random effects to flexibly incorporate multiple types of variability.
-The typical GLMM has the form:
-\newcommand{\dist}{{\cal D}}
-\begin{equation}
-\begin{aligned}
-\bY & = \dist(g^{-1}(\bmu), \phi) \\
-\bmu & = \bX \bbeta + \bZ \bb  \\
-\bb & \sim \textrm{MVN}(0, \bSigma(\theta))  \\
-\label{eq:glmm}
-\end{aligned}
-\end{equation}
-where $\bZ$ is an $n \times q$ model matrix for the $q$-dimensional vector-valued random effects variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional mean (or mode) of the random effect, which is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
-When $\dist$ is Gaussian, $g$ is the identity function, $\bZ$ is the identity matrix, and $\bSigma(\theta)=\sigma^2\bC$, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
-% ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
-% Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the generalized mixed model framework with Gaussian responses by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^{2}_{\epsilon} = 0$.
-
-%% Check if the two s. above make sense.
-
-%%MLi: The P below does not feel like it belongs here. 
-% There are several forms of random variation in the mixed model framework.
-% First, random intercepts can allow the response trait to vary independently across groups other than species (e.g., patches, sites, or experiments). 
-% Second, random intercepts can also allow the response (trait or distribution) to vary either independently among species (since species represent the tips of the phylogenetic tree) or among species in a phylogenetically correlated way (i.e., species that are closely related tend to have similar responses).
-% The third type of variation are random slopes; they allow fixed effects (the relationship between predictors and responses) to vary among groups.
-
-% \bmb{worth including the \pkg{MCMCglmm}/\pkg{brms} inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
-% \mli{inverse VCV for BLUP? an alternative way to find the inverse of the vcv(phy)}
-% \mli{G vs R, PIC and simple variations (phylolm ...) uses R and no G}
-
-\subsection*{Reformulating the phylogenetic covariance matrix}
-% The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
-% Phylogenetic independent contrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
-% Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.
-
-\newcommand{\bS}{{\mathbf S}}
-\newcommand{\bJ}{{\mathbf J}}
-\newcommand{\bB}{{\mathbf B}}
-\newcommand{\bBadj}{{\mathbf B}_{\mbox{\tiny adj}}}
-\newcommand{\bomega}{{\boldsymbol \omega}}
-\newcommand{\bell}{{\boldsymbol \ell}}
-\newcommand{\e}{{ \epsilon}}
-
-The duality of $\bSigma$ and $\bZ$ for equivalent model specifications is useful for efficient implementations of models that contain a functional \bmb{???} covariance structure \citep{hefley2017basis}.
-Similarly, we can construct phylogenetic correlations (PC) through $\bZ$.
-Suppose that the evolution follows a Brownian-motion process, i.e., continuous traits evolve independently, following an unbiased, continuous-time random walk, along each branch of the phylogeny.
-In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
-Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$ is equivalent to imposing a correlation $\bC$.
-For example, for the phylogeny in figure \ref{fig:tree}, the corresponding $\bS$ takes the form:
-
-%% FIXME: need top/side labels
-\begin{center}
-\begin{figure}[H]
-  \includegraphics[scale=0.8,page=1]{./figure/phylotree.png}
-  \caption{Three-species phylogenetic tree.}
-\label{fig:tree}
-\end{figure}
-\end{center}
-\[
-  \begin{blockarray}{ccccc}
-  & L_1 & L_2 & L_3 & L_4  \\
-  \begin{block}{c(cccc)}
-  t_1 & \ell_1 & \ell_2 & 0           & 0 \\
-  t_2 & \ell_1 &  0          & \ell_3 & 0 \\
-  t_3 & 0           &  0          & 0           & \ell_4 \\
-  \end{block}
-  \end{blockarray}
-  \hspace{0.5cm}
-  \begin{blockarray}{(c)}
-  \e_1 \\
-  \e_2 \\
-  \e_3 \\
-  \e_4 \\
-  \end{blockarray}
-  \]
-% \begin{pmatrix}
-% \e_1 \\
-% \e_2 \\
-% \e_3 \\
-% \e_4
-% \end{pmatrix}
-%% FIXME: double check this
-The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree}, and the $\e_i$ are independent Normal variates with zero mean and $\sigma^2$ evolutionary variance (i.e.
-the variance for species 1 is $\textrm{E}[(\ell_1 \e_1 + \ell_2 \e_2)^2] = (L_1 + L_2)\sigma^2$).
-% The variance of Brownian-motion, is the evolutionary rate, where it is linear in evolution time.
-
-\subsubsection*{Constructing the species--branch random effects model matrix}
-
-The $\bS$ matrix is the product of an $m \times b$ indicator matrix $\bS_{ind}$ of branch indices and a vector $\bell$ of square roots of branch lengths:
-
-\[
-\bS_{ind} = \begin{bmatrix}
-1 & 1 & 0 & 0 \\ 
-1 & 0 & 1 & 0 \\ 
-0 & 0 & 0 & 1
-\end{bmatrix} , 
-\qquad
-\bell = \begin{bmatrix}
-\ell_1 \\
-\ell_2 \\
-\ell_3 \\
-\ell_4 
-\end{bmatrix} .
-\]
-$\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
-$\bS \bS^T$ gives the variance-covariance matrix of the phylogeny.
-
-In general, the random-effect model matrix $\bZ$ for a GLMM can be decomposed into term-wise model matrices $\bZ_{i}$ as described in \citet{bates2015fitting}.
-Analogous to the procedure described in \citet{bates2015fitting}, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is
-
-\begin{equation}
-\bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top}, \label{eq:ZC}
-\end{equation}
-where $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; $\bX_{i}$ is the $n \times p_{i}$ raw random-effects model matrix; and $\ast$ is the Khatri-Rao product \citep{khatri1968solutions} partitioned at the observation level ($n$).
-
-For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix corresponding to a random-slope model,
-
-\[
-\bX = \begin{bmatrix}
-1 & t_1  \\ 
-1 & t_2  \\ 
-1 & t_3 
-\end{bmatrix} 
-\]
-then the term-wise phylogenetic random effects model matrix is,
-
-\begin{equation}
-\begin{aligned}
-\bZ^{C}_{i} = (\bS^{\top}\bJ^{\top}_{i} \ast \bX^{\top}_{i})^{\top} =
-\left[
-\left(
-\begin{bmatrix}
-\ell_{1} & \ell_{1}  & 0 \\
-\ell_{2} &  0  & 0 \\
-0  &  \ell_{3} & 0 \\
-0 & 0 &  \ell_{4} 
-\end{bmatrix}
-\begin{bmatrix}
-1 & 0  & 0 \\
-0 & 1  & 0 \\
-0 & 0  & 1  
-\end{bmatrix}
-\right)
-\ast
-\begin{bmatrix}
-1   & 1   & 1  \\ 
-t_1 & t_2 & t_3
-\end{bmatrix} 
-\right]^{\top}
-\\
-= \begin{bmatrix}
-\ell_{1} & \ell_{1}t_1 & \ell_{2} & \ell_{2}t_1 & 0 & 0 & 0 & 0 \\
-\ell_{1} & \ell_{1}t_2 & 0 & 0 & \ell_{3} & \ell_{3}t_2 & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 0 & \ell_{4} & \ell_{4}t_3
-\end{bmatrix}.
-\end{aligned}
-\end{equation}
-
-% 
-% The scaled species--branch matrix $\bS_{adji}$ is scaled version of $\bS$ via the standardized generalized variance (SGV).
-% SGVs are used as an intrinsic/natural way to scale down the variance-covariance matrix measured in different scales (for example, phylogenetic distance, years and etc) compariable to other variation in the model \citep{sengupta1987tests}.
-% SGV is the $nth$ root of the generalized variance (GV), or the determinant of the variance-covariance matrix, where GV measures the overall size or volume of the variance-covariance matrix.  
-% The SGV is given by:
-% \begin{align}
-% \bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
-% \end{align}
-% where $n$ is the number of species in the phylogeny.
-% 
-% The analog in the $\bS$ is to scale the branch vector $\bell$ by square root of the SGV:
-% \begin{align}
-% \bell_{adj} & = \frac{\bell}{\sqrt{\bomega}}
-% \end{align} 
-% and the $\bS_{adj}$ is the matrix product of $\bS_{ind}$ and $\bell_{adj}$.
-
-\subsection*{Simulation}
-\subsubsection*{Single group model}
-
-We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm}) with a single response variable $\bY$ and a single continuous normally distributed predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
-For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e., $\cal{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model.
-For the first set of simulations, we simulate one observation per species.
-% For a broader range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
-Thus, the full simulation model is as follows:
-\begin{equation}
-\begin{aligned}
-\bY & = \dist(g^{-1}(\bmu), \phi) \\
-\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}}}) \bm{t} + \besp \\\
-(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
-\Sigma_{\mathrm{phy_{int-slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
-\end{bmatrix} 
-\right) \\ 
-\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2) .
-\label{eq:single_glmm}
-\end{aligned}
-\end{equation}
-The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma^2_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma^2_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random slope and intercept $\Sigma_{\mathrm{phy_{int-slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).
-The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic dependency variability in regression effect ($b_{\mathrm{phy_{slope}}}$) and response ($b_{\mathrm{phy_{int}}}$); i.e.
-if a positive correlation indicates that similar species have similar relative intercepts and slopes.
-Predictor-level and intercept-level random effects of species are not applicable in this simulation setting because there is only a single observation per species, so within-species variation cannot be separated from tip variation.
-
-\subsubsection*{Multi-group model}
-
-We extend the simulation model by adding multiple groups where each group has one observation per species. 
-The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction \citep{helmus2007separating}. 
-The full multi-group model is as follows: 
-\begin{equation}
-\begin{aligned}
-\bY & = \dist(g^{-1}(\bmu), \phi) \\
-\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}}) \bm{t} + \bf{b_{\mathrm{sp:group}}} + \besp \\
-(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
-\Sigma_{\mathrm{phy_{int-slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
-\end{bmatrix}
-\right) \\
-(\bf{b_{\mathrm{sp_{int}}}, b_{\mathrm{sp_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\sigma^2_{\mathrm{sp_{int}}} & \sigma_{\mathrm{sp_{int-slope}}} \\ 
-\sigma_{\mathrm{sp_{int-slope}}} & \sigma^2_{\mathrm{sp_{slope}}}
-\end{bmatrix}
-\right) \\
-\bf{b_{\mathrm{group}}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
-\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \Sigma^2_{phy}) \\
-\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
-\label{eq:multiple_glmm}
-\end{aligned}
-\end{equation}
-where $\textrm{I}_{\textrm{group}}$ is a indicator matrix of groups; and $\otimes$ is the Kronecker product.
-
-The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$)) compared to the single--group full model.
-Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species; thus we can distinguish variation among species separately from residual variation.
-Variance in the intercept of species--group interactions ($\bf{b_{\mathrm{sp:group}}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
-
-\subsection*{Platforms}
-
-We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
-% \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
-% \begin{verbatim}
-% grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
-% invert=TRUE,value=TRUE)
-% \end{verbatim}
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes an evolutionary process on the tree (typically Brownian-motion, but other processes can be used) instead of treating the residual error for each species as independent.
-Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are a slightly more flexible variation of PGLS that can allow for both phylogenetic and residual variation, as well as non-Gaussian response variables.
-Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
-Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
-One of the few packages that currently fit phylogenetic slopes to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle additional random slopes ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but does not incorporate covariation between phylogenetic random slope-intercept ($\Sigma_{\mathrm{phy_{int-slope}}}$).
-Lastly, Bayesian phylogenetic GLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
-However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
-\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC)\citep{duane1987hybrid}.
-
-% Table \mli{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
-\begin{table}[]
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{|c|c|c|c|c|c|c|c|}
-\hline
-\textbf{Package} & \pkg{nlme} & \pkg{phylolm} & \pkg{lme4}/\pkg{glmmTMB} & \pkg{pez} & \pkg{phyr} & \pkg{brms} & \pkg{MCMCglmm} \\ \hline
-Single Group & X & X & X &  &  & X & X \\ \hline
-\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-Intercept correlation\\ Residual\end{tabular} & X & \begin{tabular}[c]{@{}c@{}}X\\ \\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} \\ \hline
-Multi-group &  &  & X & X & X & X & X \\ \hline
-\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-intercept correlation\\ Phylo Species-group interaction\\ Species intercept\\ Species Slope \\ Species Slope-intercept correlation\\ Residual\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ \\ X\\ X\\ X\\ X\end{tabular} \\ \hline
-\end{tabular}%
-}
-\caption{List of estimable parameters for each R package.}
-\label{table:platform}
-\end{table}
-
-\subsection*{Simulation and evaluations}
-
-Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) and then simulated the responses for each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
-Each realization was fitted using all model variants. 
-All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
-Correlation $\rho_{xy}$ is used instead of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations \bmb{???}.
-Table \ref{table:platform} shows the parameters that are estimable for each platform. 
-We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
-For Bayesian GLMM, we included only realizations where we require values of the Gelman-Rubin statistic less than the recommended threshold 1.1.
-Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}), we included the additional convergence criterion of effective sample size (ESS) $>1000$ for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) for each replication \citep{vehtari2019rank}.
-For each replication, we sample two chains starting with 10000 iterations. 
-We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
-We then compute frequentist coverage to assess the quality of the confidence intervals.
-Coverage refers to the proportion of simulations in which the computed confidence intervals include the true values of parameters.
-We used 95\% Wald confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
-% \bmb{clarify: did you use profile if available, otherwise Wald?
-You don't give any more detail in your coverage plots.
-[How different were Wald and profile for these results, or did you not compare them?]}
-We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.
-
-\section*{Results}
-
-We used our method to reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on \pkg{lme4} and \pkg{glmmTMB} (for more details see example in supplement).
-
-\subsection*{Single Group model simulations}
-
-\begin{center}
-\begin{figure}[H]
-  \includegraphics[scale=0.7]{./figure/ssplot.pdf}
-  \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
-Total simulations $N=100$ for each category.
-The horizontal line shows the true value of the parameters in the simulation model.
-Models capable of fitting all parameters (\pkg{lme4},  \pkg{glmmTMB}, and \pkg{brms}) fit well for all parameters.
-%\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
-%\jd{Use dashed lines (or some other gg signifier) for the insufficient approaches}
-}
-\label{ssplot}
-\end{figure}
-\end{center}
-\begin{center}
-\begin{figure}[H]
-  \includegraphics[scale=0.7]{./figure/sstime.pdf}
-  \caption{Comparison of single-group model computational speed.}
-\label{ssplot_speed}
-\end{figure}
-\end{center}
-\begin{center}
-\begin{figure}[H]
-  \includegraphics[scale=0.7]{./figure/sscoverage.pdf}
-  \caption{Comparison of coverage probability for fixed effect parameters.
-Models matching the simulation model (\pkg{lme4}, \pkg{glmmTMB} and pkg{brms}) have coverage near the nominal value of 0.95.
-The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits.
-%  \bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
-  }
-\label{ssplot_coverage}
-\end{figure}
-\end{center}
-
-The full fitted model (which matches the simulation model that incorporates phylogenetic intercept, slope, and correlation) provides estimates with low bias (average difference between the estimated parameters and the true simulation parameters) for all parameters.
-Estimates for fixed effect parameters ($\beta_0$ and $\beta_1$) approach nominal coverage as the number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not for other packages.
-\pkg{brms} has higher than nominal coverage (i.e., its confidence intervals are overly conservative) because the prior distributions for the simulation parameters are centered at the true values (\citet{li2018fitting} discuss the interaction of informative priors and Bayesian calibration).
-
-In general, models that are insufficiently flexible to match the true simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
-PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for the phylogenetic intercept standard deviation parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly overconfident (coverage $\approx$ 90\% with 100 species) and the fixed slope parameter ($\beta_1$) has poor coverage ($<$ 60\%).
-PGLS, with only one parameter available, confounds all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept variation, over-covering for $\beta_0$, and under-covering for $\beta_1$.
-
-\subsection*{Multi-group model simulations}
-
-\begin{center}
-\begin{figure}[H]
-  \includegraphics[scale=0.7]{./figure/msplot.pdf}
-  \caption{Comparison of multi-group model parameter estimates.
-The horizontal line shows the true value of the parameters in the simulation model.
-Models capable of fitting all parameters (\pkg{lme4} and \pkg{glmmTMB}) fit well for all parameters.
-\pkg{pez} and \pkg{phyr} estimates for n = 100 and 500 are not available because the models did not converge within 30 minutes.
-  }
-  \label{msplot}
-\end{figure}
-\end{center}
-\begin{center}
-\begin{figure}[H]
-  \includegraphics[scale=0.7]{./figure/mstime.pdf}
-  \caption{Comparison of multi-group model computational speed.}
-  \label{msplot_time}
-\end{figure}
-\end{center}
-\begin{center}
-\begin{figure}[H]
-  \includegraphics[scale=0.7]{./figure/mscoverage.pdf}
-  \caption{Comparison of multi-group model coverage.}
-  \label{msplot_coverage}
-\end{figure}
-\end{center}
-
-In contrast, the multi-group model fits are much more similar across platforms (only the more powerful platforms can fit these models at all, and the fitting models are closer to the true simulation model).
-Similar to the single-group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model well and provide good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int-slope}}}$) for small numbers of species (i.e.
-n = 25 and 50).
-The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model but underestimates the residual standard deviation (Figure \ref{msplot}).
-
-Although the parameter estimates are similar across platforms for the multi-group simulation fits, computational efficiency varies enormously across platforms and sample size.
-For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4} for the cases studied here.
-Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit a 50-species model is $\approx$ 9 versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr} respectively. 
-\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical to fit 500-species models with \pkg{pez} and \pkg{phyr}, because computational speed scaled faster than linearly with sample size.
-\pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
-
-% \bmb{Can you put a lower bound on \pkg{pez}'s computation time? How long did you take to give up? How does it scale? can you either figure out from first principles
-%   or (easier) do a brute-force series of increasing sample size and
-%   compute a log-log regression of time vs sample size to find the
-%   approximate power?}
-
-\section*{Discussion}
-
-% \bmb{say something about the implications of this [e.g.
-good enough for getting an overall impression of phylog.
-whatever-it-is but not enough for detailed description/inference about the evolutionary process?] (or maybe that should be saved for the Discussion?)}
-
-We have simulated relatively complex models containing phylogenetic variation in both intercepts and slopes, as well as within-species variation that is quantifiable.
-These models are intrinsically more complex than some simple platforms for phylogenetic regression can handle, which may seem unfair; nevertheless, our models are certainly less complex than evolutionary processes occurring in nature.
-Our results show that models that cannot match the full ``simulation world'' perform poorly even for the parameters they do estimate; it is important to understand the limitations of these simpler, commonly used methods.
-
-Even our relatively simple multi-group model described in \ref{eq:multiple_glmm} can incorporate many layers of complexity --- i.e., analogous to spatial grouping variables, as well as correlated phylogenetic variation in the effects of several different traits and environmental variables on a focal trait.
-In theory, as long as we have enough data and enough computational power, models that can incorporate more of the complexity will always describe a biological system better.
-However, real applications are always data-constrained.
-Deciding on a practically appropriate level of model complexity for a given problem and data set is an open and difficult general problem in statistical modeling, not just in phylogenetic studies.
-For example, should one use simple models that may be overly conservative or
-risk overfitting by using more ambitious models?
-What are the relative costs and benefits of using a step-down procedure
-starting from the most complex possible model \citep{barr2013random},
-choosing simpler models \emph{a priori} \citep{baayen2008mixed},
-or using Bayesian approaches with regularizing priors \citep{hadfield2010mcmc}?
-For predictive models, how can one appropriately evaluate out-of-sample accuracy for
-structured data \citep{roberts_cross-validation_2016}?
-
-\subsection*{Incorporating multiple levels of variation}
-
-% \bmb{I think you need a lot more here; what does a random slopes model mean?
-You have one sentence that says ``people should be thinking about random slopes''.
-Then you talk about correlation (which is a much more subtle/less important issue, I think).}
-
-In classic GLMMs, random effects are used to account for group-level variation. 
-In the simplest experimental designs where only the response and not the covariates vary
-within groups, random-intercept models (or even simpler analyses \cite{murtaugh2007simplicity})
-are appropriate.
-% The random intercept model models the group effect in the response level by allowing different intercepts for each level of the grouping variable.
-However, random effects can take more complicated forms as the experimental design becomes more complex, e.g., random-slopes models.
-Random-slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
-Neglecting random slopes can lead to biased fixed effect estimates with inadequate coverage and inflated type~I errors \citep{schielzeth2008conclusions}.
-
-Nevertheless, for PGLMMs, it is hard to account for all forms of complexities and decide when it is best to use phylogenetic random effects, simple grouping, or both. 
-Optimal model complexity depends on experimental design and whether the data provides enough signal to estimate these different levels of variation, which can be strongly confounded.
-For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ will be sufficient. 
-However, if multiple observations are available per species, then these simple methods may confound tip variation with residual variation.
-In this case, multiple observations can summarized to a single measurement (for example, weighted-mean) per species to avoid confounding residual and tip variation.
-This is equivalent to assuming homoscedasticity using inverse-variance weights for unbalanced datasets. 
-Alternatively, when the within-species variance is actually of interest, accounting for within-species variation (i.e., adding species-level random effects in our example) can automatically handle multiple observations per species.
-
-It may be easier to be conservative and include both phylogenetic random effects and grouping random effects
-\bmb{clarify? ``iid random effects''?}
-but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
-In this case, the PGLMM collapses to a standard random-slopes model where the group effects are considered to be independent and identically distributed.
-
-Another simplifying alternative is include both phylogenetic random effects and grouping random effects but simplify the phylogenetic relationships and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
-However, this collapses phylogenetic structures to a standard random-slopes model. 
-Users should be aware of two essential questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
-Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}?
-% However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
-
-\subsection*{Extensions and alternatives}
-
-We have presented a range of classical phylogenetic comparative methods (i.e.
-phylogenetic least squares, linear and mixed models) to fit models that incorporate different levels of phylogenetic variation.
-Even within this scope, there is additional room for exploration, such as phylogenetic multivariate response models; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck (OU) model which accounts for both selection and drift processes \citep{butler2004phylogenetic}); Bayesian approaches \citep{hadfield2010general}; and variable-rate models, where evolutionary rates vary across the phylogeny.
-The simple approach we developed here, while it cannot easily incorporate these complexities, does offer an efficient way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models.
-This approach can in principle be combined flexibly with any platform that supports independent latent variables.
-More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.
-
-\section*{Authors’ contributions}
-
-ML and BMB conceived the ideas and designed methodology; ML and BMB implemented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analyzed the results; ML wrote the first draft of the manuscript.
-Both authors contributed critically to the drafts and gave final approval for publication.
-
-\section*{Data Availability}
-
-All codes are available at DOI:10.5281/zenodo.2639887.
-
-% \section*{Conclusion}
-% 
-% We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical PCM to simulated data. 
-% Our approach is orders of magnitud faster than existing phylogenetic mixed models and it can easily combined with any statistical mixed model framework. 
-% It is more flexible in fitting large phylogenies, large volumes of data, unbalanced data sets, and complex random effects such as slope correlations.
-
-% \newpage
-% 
-% \section*{Supplements -- translation}
-% 
-% $1 \mid Sp_{phylo}$, $0 + X_{E} \mid Sp_{phylo}$ and $X_{E} \mid Sp_{phylo}$ means phylogenetically related have similar response,   
-% 
-% \begin{tabularx}{\textwidth}{|l|X|X|}
-% \hline
-% Formula & Statistics & Biology \\
-% \hline
-% $1 \mid Sp$ &
-% random species intercept; variation within species in mean response across all factors &
-% variation of how species respond \\
-% \hline
-% 
-% $0 + X_{E} \mid Sp$ &
-% random slope of environment factor within species; variation in coefficient within species for the environmental factor &
-% variation of how species respond to the same environmental factor \\
-% \hline
-% 
-% $1 + X_{E} \mid Sp$ &
-% random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental factor with correlated mean response across all other factors &
-% variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
-% \hline
-% 
-% $1 \mid Site:Sp $ &
-% random variation in intercept among species within sites &
-% variation of how species respond within sites \\
-% \hline
-% 
-% $1 | Sp_{Phylo} $ &
-% variation among species in mean response across all factors demonstrate phylogenetic signal &
-% phylogenetically related species respond similarly \\
-% \hline
-% 
-% $0 + X_{E} \mid Sp_{Phylo}$ &
-% variation among species for environmental factors demonstrate phylogenetic signal &
-% phylogenetically related species respond similar (share common response) to the same environmental factor \\
-% \hline
-% \end{tabularx}
-%             
-%             
-%             
-%             
-%             
-%                                                               
-\singlespacing
-
-\bibliography{phyloglmm}
-
-\end{document}

### 2020-02-11 phyloglmm_ms.tex
need to read over from top to bottom
last round of fixes before sending it in
TAGGING: going to reduce text for MEE
first wordcutting attempt: down to 3601 words

@@
 
-\section*{Abstract}
-
-\begin{enumerate}
-\item{Phylogenetic comparative methods (PCM) using phylogenetic regression are a powerful technique to explore relationships among related groups of species traits.
-However, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.}
-\item{We propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but is more flexible in practice.
-We have implemented this formulation in two R statistical packages (\pkg{lme4} and \pkg{glmmTMB}).}
-\item{Our reformulation of phylogenetic generalized linear mixed models is computationally efficient, operating orders of magnitude faster than existing methods for fitting phylogenetic mixed models.}
-\item{Our approach can be implemented in any platform for generalized mixed models.
-Our implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to a broad range of previously difficult cases (e.g., large data, unbalanced observational designs, complex random effects).}
-\end{enumerate}
-
-\keywords{phylogenetic comparative methods, phylogenetic correlation, phyloglmm, species--branch matrix}
+%\include{abstract}
 
@@
 
-%%% PCM accounts for evolutionary relationships. More people are doing it.
 Phylogenetic comparative methods (PCMs) are a powerful technique to account for evolutionary relatedness when analyzing species traits.
@@
 
-Ever-increasing data collection capabilities (e.g., genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic databases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
-Researchers use these data to fit complex models describing species occurrence and traits.
-For example, ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently they have begun to integrate evolutionary considerations in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
-% \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e.
-how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature?
-\ldots}
+Ever-increasing data collection capabilities, in combination with large-scale synthetic databases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
+Researchers use these data to fit complex models describing species occurrence and traits \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic, winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
 
 Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental covariates) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
-% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}.
 While a wide range of tools is available for phylogenetic regression, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
-When faced with these constraints, researchers typically search for ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
-
-In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships by extending existing software for fitting mixed effect models. 
-This method allows researchers to easily incorporate evolutionary and statistical complexities without sacrificing speed.
+When faced with these constraints, researchers typically search for ways to simplify their analyses.
 
@@
 If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation can both be considered part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
-%\bmb{This is a good point, well expressed.
-Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel?
-Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
-%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
 Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
 Indeed, many existing methods are restricted to single observations per species, requiring users to collapse their observations to mean values for each species, which can be limiting if some covariates of interest vary within species.
-%% REMEMBER TO REVISIT IN DISCUSSION
-
-% Figure \ref{fig:variation} illustrates various forms of variation for three species. 
-%Despite having multiple observations per tip taxon (species), a typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
-% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}
-% 
-% \begin{center}
-% \begin{figure}[H]
-%   \includegraphics[scale=0.8]{./phylo_diagram.png}
-%   \caption{Potential forms of phylogenetic variation of three species.
-Top left shows pure noise (i.e.
-assuming a star phylogeny); top right accounts for phylogeny, but without noise; bottom left accounts both phylogeny and noise at the tip, and finally bottom right shows multiple observations (with noise) at the tips.}
-% \label{fig:variation}
-% \end{figure}
-% \end{center}
 
@@
 Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).
-Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors.
-For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites when modeling plant abundance via a phylogenetic GLMM approach.
+Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors \cite{nowakowski2018phylogenetic, li2017canfun}.
 The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, biologists needing to fit random-slopes model typically turn to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
@@
 
-%% FIXME: chart looks ugly
 \begin{table}[]
@@
 In particular, it easily allows for complex phylogenetic effects (random intercepts, slopes, and interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model \citep{hefley2017basis}.
-%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
 We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
@@
 \newcommand{\bSigma}{{\boldsymbol \Sigma}}
-
-We generated test data using a simple framework that combines fixed effects, phylogenetic random intercepts, slopes, and their correlations based on a simulated phylogenetic tree.
-We fit models to these test data using our approach built on the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}, using assumptions that match those of the simulation model.
-We also fit with other platforms (Table~\ref{table:model}), simplifying the statistical model when necessary.
 
@@
 \subsection*{Reformulating the phylogenetic covariance matrix}
-% The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
-% Phylogenetic independent contrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
-% Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.
 
@@
 [How different were Wald and profile for these results, or did you not compare them?]}
+% \mli{Profile and wald are approx the same for the ~20+?? that I tested. Profile takes way too long.}
 We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.
@@
 
-% \bmb{I think you need a lot more here; what does a random slopes model mean?
-You have one sentence that says ``people should be thinking about random slopes''.
-Then you talk about correlation (which is a much more subtle/less important issue, I think).}
-
 In classic GLMMs, random effects are used to account for group-level variation. 
@@
 are appropriate.
-% The random intercept model models the group effect in the response level by allowing different intercepts for each level of the grouping variable.
 However, random effects can take more complicated forms as the experimental design becomes more complex, e.g., random-slopes models.
@@
 
-% \section*{Conclusion}
-% 
-% We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical PCM to simulated data. 
-% Our approach is orders of magnitud faster than existing phylogenetic mixed models and it can easily combined with any statistical mixed model framework. 
-% It is more flexible in fitting large phylogenies, large volumes of data, unbalanced data sets, and complex random effects such as slope correlations.
-
-% \newpage
-% 
-% \section*{Supplements -- translation}
-% 
-% $1 \mid Sp_{phylo}$, $0 + X_{E} \mid Sp_{phylo}$ and $X_{E} \mid Sp_{phylo}$ means phylogenetically related have similar response,   
-% 
-% \begin{tabularx}{\textwidth}{|l|X|X|}
-% \hline
-% Formula & Statistics & Biology \\
-% \hline
-% $1 \mid Sp$ &
-% random species intercept; variation within species in mean response across all factors &
-% variation of how species respond \\
-% \hline
-% 
-% $0 + X_{E} \mid Sp$ &
-% random slope of environment factor within species; variation in coefficient within species for the environmental factor &
-% variation of how species respond to the same environmental factor \\
-% \hline
-% 
-% $1 + X_{E} \mid Sp$ &
-% random slope of environmental factor within species with correlated intercept; variation in coefficient within species for the environmental factor with correlated mean response across all other factors &
-% variation of how species respond in the same environmental factors and the correlation of the variation of how they respond in general \\
-% \hline
-% 
-% $1 \mid Site:Sp $ &
-% random variation in intercept among species within sites &
-% variation of how species respond within sites \\
-% \hline
-% 
-% $1 | Sp_{Phylo} $ &
-% variation among species in mean response across all factors demonstrate phylogenetic signal &
-% phylogenetically related species respond similarly \\
-% \hline
-% 
-% $0 + X_{E} \mid Sp_{Phylo}$ &
-% variation among species for environmental factors demonstrate phylogenetic signal &
-% phylogenetically related species respond similar (share common response) to the same environmental factor \\
-% \hline
-% \end{tabularx}
-%             
-%             
-%             
-%             
-%             
-%                                                               
-\singlespacing
-
 \bibliography{phyloglmm}

### 2020-02-11 phyloglmm_ms.tex
cutted down to 3342, it is getting harder to cut down

@@
 
-Standard statistical regression techniques neglect correlations among species responses due to their shared evolutionary history.
 Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
@@
 Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
-Indeed, many existing methods are restricted to single observations per species, requiring users to collapse their observations to mean values for each species, which can be limiting if some covariates of interest vary within species.
 
@@
 However, species within taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes).
-Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way (i.e., similar species will have similar predictor--response relationships).
+Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way.
 Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors \cite{nowakowski2018phylogenetic, li2017canfun}.
@@
 \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} \citep{phyr}, \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2018brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
-% We expect that our approach will be fast as well as flexible, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
-%% TODO LAST LINE SEE PAGE 5
-
-%% FIXME: move to methods/simulations
-% In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
-% We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.
 
@@
 
-We begin by describing the classic phylogenetic regression in a linear regression setting.
 Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors encoded in a model matrix $\bX$, where each species is measured exactly once. 
 The standard phylogenetic regression can be formulated as
-
-%\bmb{boldface matrices/vectors i.e.
-$\bX \bbeta$?
-or not worth the trouble?
-How about $\bmu=\bX \bbeta$; $\bY \sim \textrm{MVN}(\bmu,\sigma^2 \bC)$ for greater generality?
-(OK, I see that you move in this direction later.
-But I'm not sure it's worth the detour of using two different notations?
-even if you are going to extend to include fixed effects, $Z$, etc.
-later ...}:
 
@@
 \end{equation}
-where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic covariance (PC) matrix.
+where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic covariance (PC) matrix.
 The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.
-
-%\bmb{give a ref for this? maybe Paradis's ape book, or Garamszegi's book, or ?}
-%\mli{chapter 7 in garamszegi 2014}
-
-% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
-%\mli{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
 
@@
 \end{equation}
-where $\bZ$ is an $n \times q$ model matrix for the $q$-dimensional vector-valued random effects variables; $\bb$ (sometimes referred to as the ``G-side" effect) representing the conditional mean (or mode) of the random effect, which is assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
+where $\bZ$ is an $n \times q$ model matrix for the $q$-dimensional vector-valued random effects variables; $\bb$ representing the conditional mean (or mode) of the random effect, which is assumed to be multivariate normally distributed with a covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
 When $\dist$ is Gaussian, $g$ is the identity function, $\bZ$ is the identity matrix, and $\bSigma(\theta)=\sigma^2\bC$, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
-% ; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
-% Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the generalized mixed model framework with Gaussian responses by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^{2}_{\epsilon} = 0$.
-
-%% Check if the two s. above make sense.
-
-%%MLi: The P below does not feel like it belongs here. 
-% There are several forms of random variation in the mixed model framework.
-% First, random intercepts can allow the response trait to vary independently across groups other than species (e.g., patches, sites, or experiments). 
-% Second, random intercepts can also allow the response (trait or distribution) to vary either independently among species (since species represent the tips of the phylogenetic tree) or among species in a phylogenetically correlated way (i.e., species that are closely related tend to have similar responses).
-% The third type of variation are random slopes; they allow fixed effects (the relationship between predictors and responses) to vary among groups.
-
-% \bmb{worth including the \pkg{MCMCglmm}/\pkg{brms} inverse-VCV formulation here? And/or contacting Eric Pedersen to find out about his Markov Random Field phylog. stuff in GAM?}
-% \mli{inverse VCV for BLUP? an alternative way to find the inverse of the vcv(phy)}
-% \mli{G vs R, PIC and simple variations (phylolm ...) uses R and no G}
 
@@
 $\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
-$\bS \bS^T$ gives the variance-covariance matrix of the phylogeny.
+$\bS \bS^T$ gives the covariance matrix of the phylogeny.
 
@@
 \end{equation}
-
-% 
-% The scaled species--branch matrix $\bS_{adji}$ is scaled version of $\bS$ via the standardized generalized variance (SGV).
-% SGVs are used as an intrinsic/natural way to scale down the variance-covariance matrix measured in different scales (for example, phylogenetic distance, years and etc) compariable to other variation in the model \citep{sengupta1987tests}.
-% SGV is the $nth$ root of the generalized variance (GV), or the determinant of the variance-covariance matrix, where GV measures the overall size or volume of the variance-covariance matrix.  
-% The SGV is given by:
-% \begin{align}
-% \bomega & = \textrm{det}(\bSigma_{phy})^{1/n},
-% \end{align}
-% where $n$ is the number of species in the phylogeny.
-% 
-% The analog in the $\bS$ is to scale the branch vector $\bell$ by square root of the SGV:
-% \begin{align}
-% \bell_{adj} & = \frac{\bell}{\sqrt{\bomega}}
-% \end{align} 
-% and the $\bS_{adj}$ is the matrix product of $\bS_{ind}$ and $\bell_{adj}$.
 
@@
 We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
-% \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
-% \begin{verbatim}
-% grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
-% invert=TRUE,value=TRUE)
-% \end{verbatim}
 Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes an evolutionary process on the tree (typically Brownian-motion, but other processes can be used) instead of treating the residual error for each species as independent.
@@
 We used 95\% Wald confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
-% \bmb{clarify: did you use profile if available, otherwise Wald?
-You don't give any more detail in your coverage plots.
-[How different were Wald and profile for these results, or did you not compare them?]}
-% \mli{Profile and wald are approx the same for the ~20+?? that I tested. Profile takes way too long.}
 We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.
@@
 
-% \bmb{say something about the implications of this [e.g.
-good enough for getting an overall impression of phylog.
-whatever-it-is but not enough for detailed description/inference about the evolutionary process?] (or maybe that should be saved for the Discussion?)}
-
 We have simulated relatively complex models containing phylogenetic variation in both intercepts and slopes, as well as within-species variation that is quantifiable.
@@
 
-Even our relatively simple multi-group model described in \ref{eq:multiple_glmm} can incorporate many layers of complexity --- i.e., analogous to spatial grouping variables, as well as correlated phylogenetic variation in the effects of several different traits and environmental variables on a focal trait.
+Even our relatively simple multi-group model described in (\ref{eq:multiple_glmm}) can incorporate many layers of complexity.
 In theory, as long as we have enough data and enough computational power, models that can incorporate more of the complexity will always describe a biological system better.
@@
 Deciding on a practically appropriate level of model complexity for a given problem and data set is an open and difficult general problem in statistical modeling, not just in phylogenetic studies.
-For example, should one use simple models that may be overly conservative or
+%For example, should one use simple models that may be overly conservative or
 risk overfitting by using more ambitious models?
-What are the relative costs and benefits of using a step-down procedure
-starting from the most complex possible model \citep{barr2013random},
-choosing simpler models \emph{a priori} \citep{baayen2008mixed},
-or using Bayesian approaches with regularizing priors \citep{hadfield2010mcmc}?
-For predictive models, how can one appropriately evaluate out-of-sample accuracy for
-structured data \citep{roberts_cross-validation_2016}?
+%What are the relative costs and benefits of using a step-down procedure
+%starting from the most complex possible model \citep{barr2013random},
+%choosing simpler models \emph{a priori} \citep{baayen2008mixed},
+%or using Bayesian approaches with regularizing priors \citep{hadfield2010mcmc}?
+%For predictive models, how can one appropriately evaluate out-of-sample accuracy for
+%structured data \citep{roberts_cross-validation_2016}?
 
@@
 
-In classic GLMMs, random effects are used to account for group-level variation. 
-In the simplest experimental designs where only the response and not the covariates vary
-within groups, random-intercept models (or even simpler analyses \cite{murtaugh2007simplicity})
-are appropriate.
-However, random effects can take more complicated forms as the experimental design becomes more complex, e.g., random-slopes models.
+In classic GLMMs, random effects are used to account for group-level variation and they take more complicated forms as the experimental design becomes more complex, e.g., random-slopes models.
 Random-slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
@@
 It may be easier to be conservative and include both phylogenetic random effects and iid grouping random effects by simplifying the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
-In this case, the PGLMM collapses to a standard random-slopes model where the group effects are considered to be independent and identically distributed.
-
-Another simplifying alternative is include both phylogenetic random effects and grouping random effects but simplify the phylogenetic relationships and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
-However, this collapses phylogenetic structures to a standard random-slopes model. 
 Users should be aware of two essential questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?

### 2020-02-11 phyloglmm_ms.tex
cutted down to 3300 words

@@
 
-Ever-increasing data collection capabilities, in combination with large-scale synthetic databases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
-Researchers use these data to fit complex models describing species occurrence and traits \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic, winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
-
 Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental covariates) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
@@
 Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way.
-Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors \cite{nowakowski2018phylogenetic, li2017canfun}.
+Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors \citep{nowakowski2018phylogenetic, li2017canfun}.
 The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, biologists needing to fit random-slopes model typically turn to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
@@
 
-We used our method to reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on \pkg{lme4} and \pkg{glmmTMB} (for more details see example in supplement).
+We used our method to reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on \pkg{lme4} and \pkg{glmmTMB} (for more details see Data Availability section).
 

### 2020-02-17 phyloglmm_ms.tex
started working on trimming

@@
 Phylogenetic comparative methods (PCMs) are a powerful technique to account for evolutionary relatedness when analyzing species traits.
-Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.
+Given a known phylogeny, PCMs model the relationships among species traits while accounting for relatedness; they can be used to control statistically for phylogeny, to quantify phylogenetic signal in traits (dependence among traits due to evolutionary relationships), or both.
 
-Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental covariates) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
+Unlike standard statistical models, where all of the predictor variables of interest (such as species traits and environmental covariates) are directly observable, PCMs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
 While a wide range of tools is available for phylogenetic regression, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
-When faced with these constraints, researchers typically search for ways to simplify their analyses.
+When faced with these constraints, researchers often simplify their analyses, at the risk of neglecting important processes.
 
@@
 
-Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
-If in addition the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
-More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than the simple Brownian motion model.
-These methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
-If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation can both be considered part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
+In classic phylogenetic regression, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
+If the residuals are normally distributed and observed without additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
+More recent approaches --- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than Brownian motion.
+These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
+If each species' traits are measured multiple times, possibly under different conditions, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process while the observation error can be independently identified from within-species variation.
 Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
 
-Classic phylogenetic regressions usually allow the uncertainty in the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects of the predictor variables may evolve according to the phylogenetic relationship across species as well.
-For example, suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model.
+Classic phylogenetic regressions allow the residuals to evolve along the phylogeny, but the effects of the predictor variables may evolve along the phylogeny as well; extending PCMs to allow for multiple sources of variation lead to the
+For example, suppose we wish to fit a PCM to predict species' brain size (Y) from their body size (X) \cite{felsenstein1985phylogenies} using a mixed-effect model.
+
 The standard phylogenetic mixed-effect model allows for phylogenetic correlations in the intercept of the relationship between body and brain size. 

### 2020-02-17 phyloglmm_ms.tex
trimming (3055 words)

@@
 \usepackage{multirow}
+\usepackage{amssymb}
+\newcommand{\y}{\checkmark}
 % use \xspace to allow for space after a macro as necessary
@@
 
-Phylogenetic comparative methods (PCMs) are a powerful technique to account for evolutionary relatedness when analyzing species traits.
-Given a known phylogeny, PCMs model the relationships among species traits while accounting for relatedness; they can be used to control statistically for phylogeny, to quantify phylogenetic signal in traits (dependence among traits due to evolutionary relationships), or both.
+Phylogenetic comparative methods (PCMs) account for evolutionary relatedness when analyzing species traits.
+Given a known phylogeny, PCMs model the relationships among species traits while incorporating relatedness; they can be used to control statistically for phylogeny, to quantify phylogenetic signal in traits, or both.
 
-Unlike standard statistical models, where all of the predictor variables of interest (such as species traits and environmental covariates) are directly observable, PCMs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
-While a wide range of tools is available for phylogenetic regression, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
+In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PCMs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
+While a wide range of tools is available for PCM, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets.
 When faced with these constraints, researchers often simplify their analyses, at the risk of neglecting important processes.
@@
 
-In classic phylogenetic regression, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}.
+In classic PCMs, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion (BM) evolutionary model \citep{felsenstein1985phylogenies}.
 If the residuals are normally distributed and observed without additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
-More recent approaches --- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than Brownian motion.
+More recent approaches --- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than BM.
 These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
-If each species' traits are measured multiple times, possibly under different conditions, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process while the observation error can be independently identified from within-species variation.
+If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation.
 Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
 
-Classic phylogenetic regressions allow the residuals to evolve along the phylogeny, but the effects of the predictor variables may evolve along the phylogeny as well; extending PCMs to allow for multiple sources of variation lead to the
-For example, suppose we wish to fit a PCM to predict species' brain size (Y) from their body size (X) \cite{felsenstein1985phylogenies} using a mixed-effect model.
+Classic PCMs allow the residuals to evolve along the phylogeny, but the effects of the predictor variables may evolve along the phylogeny as well; extending PCMs to allow for multiple sources of variation leads to the  \emph{phylogenetic mixed model} \citep{housworth2004phylogenetic}.
+For example, suppose we wish to fit a PCM to predict species' brain size from their body size \cite{felsenstein1985phylogenies} using a mixed-effect model.
+The standard PCM allows for phylogenetic correlations in the residuals of the relationship between body and brain size; the phylogenetic mixed model can allow incorporate both
+variation in brain size among taxa with similar body sizes (random intercepts) and
+variation in the \emph{relationship} between predictors and responses among taxa (random slopes).
 
-The standard phylogenetic mixed-effect model allows for phylogenetic correlations in the intercept of the relationship between body and brain size. 
-However, species within taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes).
-Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way.
-Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors \citep{nowakowski2018phylogenetic, li2017canfun}.
-The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, biologists needing to fit random-slopes model typically turn to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
-Table \ref{table:model} summarizes currently available methods for phylogenetic comparative analysis.
+Several recent studies have used phylogenetic mixed models in
+community ecology \citep{nowakowski2018phylogenetic, li2017canfun}.
+However, the phylogenetic mixed modeling tools that allow extensions such as random slopes
+and separation of tip and observation error in a frequentist framework are inflexible; thus, biologists needing to fit random-slopes models typically turn to Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms} (Table \ref{table:model}).
 
@@
 
-In this paper, we propose an alternative formulation of phylogenetic regression in a generalized linear mixed modeling framework that is mathematically equivalent to previous approaches, but more flexible.
-In particular, it easily allows for complex phylogenetic effects (random intercepts, slopes, and interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model \citep{hefley2017basis}.
-We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e.
-\pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} \citep{phyr}, \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2018brms}), fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species as well as residual variation.
+We propose an alternative, more flexible formulation of the phylogenetic mixed model that is mathematically equivalent to previous approaches.
+In particular, it allows for complex phylogenetic effects (random intercepts, slopes, and interactions), without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model \citep{hefley2017basis}.
+We compare our technique (built on the R packages \pkg{lme4} and \pkg{glmmTMB}) with existing R packages, fitting models to simulated data that incorporates random slopes, random intercepts (tip variation), and residual variation.
 
@@
 
-Consider a simple linear regression model of observable trait $\bY$ as a function of some predictors encoded in a model matrix $\bX$, where each species is measured exactly once. 
-The standard phylogenetic regression can be formulated as
-
+Suppose a species trait $\bY$ is a linear function of some predictors encoded in a model matrix $\bX$, where each species is measured exactly once. 
+The standard phylogenetic regression can be expressed as
 \begin{equation}
@@
 \end{equation}
-where $\bY$ is an $n \times 1$ response vector; $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\besp$ is an $n \times 1$ vector which is assumed to be multivariate normally distributed with mean $0$ and covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic covariance (PC) matrix.
-The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.
+where $\bY$ is an length-$n$ response vector; $\bX$ is an $n \times m$ matrix, describing $n$ observations of $m$ predictor variables; $\bbeta$ is an $m$-vector of coefficients; $\besp$ is a multivariate normally distributed $n$-vector with mean $0$ and covariance matrix $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation (PC) matrix that quantifies the proportion of shared evolution between any pair of taxa in the phylogeny \citep{garamszegi2014modern}.
 
 \subsection*{Phylogenetic generalized linear mixed model}
-Alternatively, one can use the generalized linear mixed effects modeling (GLMM) framework to define a wider range that includes the standard phylogenetic regression as a special case \citep{lynch1991methods}.
-The generalized linear mixed effect model allows for non-Gaussian responses and uses random effects to flexibly incorporate multiple types of variability.
-The typical GLMM has the form:
+
+The phylogenetic generalized linear mixed model (PGLMM) framework defines a wider range of models that includes the standard phylogenetic regression as a special case \citep{lynch1991methods}.
+The PGLMM allows for non-Gaussian responses and uses incorporates multiple components of variability.
+The PGLMM has the form:
 \newcommand{\dist}{{\cal D}}
@@
 \end{equation}
-where $\bZ$ is an $n \times q$ model matrix for the $q$-dimensional vector-valued random effects variables; $\bb$ representing the conditional mean (or mode) of the random effect, which is assumed to be multivariate normally distributed with a covariance matrix given by $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
+where in addition to the terms from (\ref{eq:gls}) $\bZ$ is an $n \times q$ model matrix for the $q$-dimensional vector-valued random effects; $\bb$ represents the conditional mean (or mode) of the random effect, which is multivariate normally distributed with covariance matrix $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
 When $\dist$ is Gaussian, $g$ is the identity function, $\bZ$ is the identity matrix, and $\bSigma(\theta)=\sigma^2\bC$, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
@@
 
-The duality of $\bSigma$ and $\bZ$ for equivalent model specifications is useful for efficient implementations of models that contain a functional (e.g., autoregressive) covariance structure \citep{hefley2017basis}.
-Similarly, we can construct phylogenetic correlations (PC) through $\bZ$.
-Suppose that the evolution follows a Brownian-motion process, i.e., continuous traits evolve independently, following an unbiased, continuous-time random walk, along each branch of the phylogeny.
-In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
+In general, correlations within statistical models can be integrated either in the 
+covariance matrix $\bSigma$ or in the structure of the model matrix $\bZ$ \citep{hefley2017basis};
+thus, we can use $\bZ$ to incorporate phylogenetic correlations.
+Suppose evolution follows a BM process, i.e., continuous traits evolve independently, following an unbiased random walk along each branch of the phylogeny.
+Then the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
 Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$ is equivalent to imposing a correlation $\bC$.
@@
 
-%% FIXME: need top/side labels
 \begin{center}
@@
   \hspace{0.5cm}
-  \begin{blockarray}{(c)}
+  \begin{blockarray}{c}
+    \vphantom{L_1} \\
+    \begin{block}{(c)}
       \e_1 \\
@@
       \e_4 \\
+    \end{block}
   \end{blockarray}
@@
 %% FIXME: double check this
-The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree}, and the $\e_i$ are independent Normal variates with zero mean and $\sigma^2$ evolutionary variance (i.e.
-the variance for species 1 is $\textrm{E}[(\ell_1 \e_1 + \ell_2 \e_2)^2] = (L_1 + L_2)\sigma^2$).
+For example, the phylogenetic effect for species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in Figure \ref{fig:tree}, and the $\e_i$ are independent Normal deviates with zero mean and variance $\sigma^2$ (i.e.
+the phylogenetic variance for species 1 is $\textrm{E}[(\ell_1 \e_1 + \ell_2 \e_2)^2] = (L_1 + L_2)\sigma^2$).
 % The variance of Brownian-motion, is the evolutionary rate, where it is linear in evolution time.
@@
 
-In general, the random-effect model matrix $\bZ$ for a GLMM can be decomposed into term-wise model matrices $\bZ_{i}$ as described in \citet{bates2015fitting}.
-Analogous to the procedure described in \citet{bates2015fitting}, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is
+In general, the random-effect model matrix $\bZ$ for a mixed model can be decomposed into term-wise model matrices $\bZ_{i}$ as described in \citet{bates2015fitting}.
+Analogous to this procedure, the phylogenetic random-effect matrix $\bZ^{C}_{i}$ is
 
@@
 
-For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix corresponding to a random-slope model,
+For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a model matrix corresponding to intercept and slope terms,
 
@@
 
-We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm}) with a single response variable $\bY$ and a single continuous normally distributed predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
-For simplicity, the response variable $\bY$ is conditionally normally distributed (i.e., $\cal{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model.
+We generated test data based on (\ref{eq:glmm}) with a single response variable $\bY$ and a single normally distributed predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
+The response variable $\bY$ is conditionally normally distributed (i.e., $\cal{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model. 
 For the first set of simulations, we simulate one observation per species.
@@
 \right) \\ 
-\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2) .
+\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2) \quad.
 \label{eq:single_glmm}
@@
 \end{equation}
-The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma^2_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma^2_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random slope and intercept $\Sigma_{\mathrm{phy_{int-slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).
-The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic dependency variability in regression effect ($b_{\mathrm{phy_{slope}}}$) and response ($b_{\mathrm{phy_{int}}}$); i.e.
-if a positive correlation indicates that similar species have similar relative intercepts and slopes.
+The model has two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma^2_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma^2_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random slope and intercept $\Sigma_{\mathrm{phy_{int-slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).
+The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic effects on the slope ($b_{\mathrm{phy_{slope}}}$) and intercept ($b_{\mathrm{phy_{int}}}$) for each branch of the phylogeny; i.e.
+a positive correlation indicates that species with similar intercepts also have similar slopes.
 Predictor-level and intercept-level random effects of species are not applicable in this simulation setting because there is only a single observation per species, so within-species variation cannot be separated from tip variation.
@@
 \end{equation}
-where $\textrm{I}_{\textrm{group}}$ is a indicator matrix of groups; and $\otimes$ is the Kronecker product.
+where $\textrm{I}_{\textrm{group}}$ is a indicator matrix for groups and $\otimes$ is the Kronecker product.
 
-The multi-group full simulation model has five additional random effects (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$)) compared to the single--group full model.
-Predictor-level and intercept-level random effects of species are applicable in the multi-group model setting because there are multiple observations per species; thus we can distinguish variation among species separately from residual variation.
+Compared to the single--group model, the multi-group simulation model has five additional random effect parameters (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$))
+Because each species has multiple observations we can distinguish variation among species from residual variation, and thus we can include predictor-level and intercept-level random effects of species.
 Variance in the intercept of species--group interactions ($\bf{b_{\mathrm{sp:group}}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
@@
 
-We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it fits a linear model where the covariance structure between species assumes an evolutionary process on the tree (typically Brownian-motion, but other processes can be used) instead of treating the residual error for each species as independent.
-Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are a slightly more flexible variation of PGLS that can allow for both phylogenetic and residual variation, as well as non-Gaussian response variables.
+We compare our approach with five other R packages that can fit PCMs: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PCM models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
 Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
 Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
-One of the few packages that currently fit phylogenetic slopes to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle additional random slopes ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but does not incorporate covariation between phylogenetic random slope-intercept ($\Sigma_{\mathrm{phy_{int-slope}}}$).
-Lastly, Bayesian phylogenetic GLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
+Among the few packages that currently fit phylogenetic slopes to predictor level variation are \pkg{pez} and \pkg{phyr}, which can handle random-slope models ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but do not incorporate covariation between phylogenetic random slopes and intercepts ($\Sigma_{\mathrm{phy_{int-slope}}}$).
+Lastly, Bayesian PGLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
 However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
-\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC)\citep{duane1987hybrid}.
+\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique.
 
@@
 \hline
-\textbf{Package} & \pkg{nlme} & \pkg{phylolm} & \pkg{lme4}/\pkg{glmmTMB} & \pkg{pez} & \pkg{phyr} & \pkg{brms} & \pkg{MCMCglmm} \\ \hline
-Single Group & X & X & X &  &  & X & X \\ \hline
-\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-Intercept correlation\\ Residual\end{tabular} & X & \begin{tabular}[c]{@{}c@{}}X\\ \\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} \\ \hline
-Multi-group &  &  & X & X & X & X & X \\ \hline
-\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-intercept correlation\\ Phylo Species-group interaction\\ Species intercept\\ Species Slope \\ Species Slope-intercept correlation\\ Residual\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ \\ X\\ X\\ X\\ X\end{tabular} \\ \hline
+ & \pkg{nlme} & \pkg{phylolm} & \pkg{lme4}/\pkg{glmmTMB} & \pkg{pez} & \pkg{phyr} & \pkg{brms} & \pkg{MCMCglmm} \\ \hline
+Single Group & \y & \y & \y &  &  & \y & \y \\ \hline
+\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-Intercept correlation\\ Residual\end{tabular} & \y & \begin{tabular}[c]{@{}c@{}}\y\\ \\ \\ \y\end{tabular} & \begin{tabular}[c]{@{}c@{}}\y\\ \y\\ \y\\ \y\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}\y\\ \y\\ \y\\ \y\end{tabular} & \begin{tabular}[c]{@{}c@{}}\y\\ \y\\ \y\\ \y\end{tabular} \\ \hline
+Multi-group &  &  & \y & \y & \y & \y & \y \\ \hline
+\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-intercept correlation\\ Phylo Species-group interaction\\ Species intercept\\ Species Slope \\ Species Slope-intercept correlation\\ Residual\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}\y\\ \y\\ \y\\ \y\\ \y\\ \y\\ \y\\ \y\end{tabular} & \begin{tabular}[c]{@{}c@{}}\y\\ \y\\ \\ \y\\ \y\\ \y\\ \\ \y\end{tabular} & \begin{tabular}[c]{@{}c@{}}\y\\ \y\\ \\ \y\\ \y\\ \y\\ \\ \y\end{tabular} & \begin{tabular}[c]{@{}c@{}}\y\\ \y\\ \y\\ \y\\ \y\\ \y\\ \y\\ \y\end{tabular} & \begin{tabular}[c]{@{}c@{}}\y\\ \y\\ \y\\ \\ \y\\ \y\\ \y\\ \y\end{tabular} \\ \hline
 \end{tabular}%
@@
 
-Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) and then simulated the responses for each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
+Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) and then simulated the responses for each tree (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
 Each realization was fitted using all model variants. 
 All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
-Correlation $\rho_{xy}$ is used instead of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations.
+We parameterize the models with correlation $\rho_{xy}$ rather than covariance $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$).
 %\bmb{???}
@@
 We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
-For Bayesian GLMM, we included only realizations where we require values of the Gelman-Rubin statistic less than the recommended threshold 1.1.
-Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}), we included the additional convergence criterion of effective sample size (ESS) $>1000$ for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) for each replication \citep{vehtari2019rank}.
-For each replication, we sample two chains starting with 10000 iterations. 
+For Bayesian fits, we evaluate realizations with Gelman-Rubin statistic $<1.1$.
+Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}), we additionally restricted results to fits with  effective sample size $>1000$ for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) \citep{vehtari2019rank}.
+For each replicate, we sample two chains starting with 10000 iterations. 
 We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
-We then compute frequentist coverage to assess the quality of the confidence intervals.
-Coverage refers to the proportion of simulations in which the computed confidence intervals include the true values of parameters.
-We used 95\% Wald confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
+We computed 95\% Wald confidence intervals for frequentist methods and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
+\bmb{no profiles CIs?
+is that problematic?}
 We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.
 
+In supplementary materials, we reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on \pkg{lme4} and \pkg{glmmTMB}.
+
 \section*{Results}
-
-We used our method to reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on \pkg{lme4} and \pkg{glmmTMB} (for more details see Data Availability section).
 
@@
 Estimates for fixed effect parameters ($\beta_0$ and $\beta_1$) approach nominal coverage as the number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not for other packages.
-\pkg{brms} has higher than nominal coverage (i.e., its confidence intervals are overly conservative) because the prior distributions for the simulation parameters are centered at the true values (\citet{li2018fitting} discuss the interaction of informative priors and Bayesian calibration).
+\pkg{brms} has higher than nominal coverage (i.e., its confidence intervals are too conservative) because the prior distributions for the simulation parameters are centered at the true values (\citet{li2018fitting}).
 
-In general, models that are insufficiently flexible to match the true simulation model (PGLM and PGLS) will try to fit the data with the parameters available. 
-PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for the phylogenetic intercept standard deviation parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly overconfident (coverage $\approx$ 90\% with 100 species) and the fixed slope parameter ($\beta_1$) has poor coverage ($<$ 60\%).
-PGLS, with only one parameter available, confounds all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept variation, over-covering for $\beta_0$, and under-covering for $\beta_1$.
+In general, models that are insufficiently flexible to match the data (PGLM and PGLS) will lead to bias in some parameters.
+PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for the phylogenetic intercept standard deviation parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly overconfident (coverage $\approx$ 90\% with 100 species) and the fixed slope parameter ($\beta_1$) has extremely poor coverage ($<$ 60\%).
+PGLS, which uses only one parameter, combines all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept variation, over-covering for $\beta_0$, and under-covering for $\beta_1$.
 
@@
 Models capable of fitting all parameters (\pkg{lme4} and \pkg{glmmTMB}) fit well for all parameters.
-\pkg{pez} and \pkg{phyr} estimates for n = 100 and 500 are not available because the models did not converge within 30 minutes.
+\pkg{pez} and \pkg{phyr} estimates for n = 100 and 500 are missing because the models did not converge within 30 minutes.
+\bmb{add English strip labels as in Fig 2?}
   }
@@
 
-In contrast, the multi-group model fits are much more similar across platforms (only the more powerful platforms can fit these models at all, and the fitting models are closer to the true simulation model).
-Similar to the single-group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model well and provide good estimates for all parameters except the correlation ($\sigma_{\mathrm{sp_{int-slope}}}$) for small numbers of species (i.e.
+The multi-group model estimates are much more similar across platforms (only a subset of platforms can fit these models at all, and the fitting models are closer to the true model).
+As with the single-group fits, \pkg{lme4} and \pkg{glmmTMB} match the simulation model well and provide good estimates for all parameters, except the correlation ($\sigma_{\mathrm{sp_{int-slope}}}$) for small numbers of species (i.e.
 n = 25 and 50).
-The lack of correlations in \pkg{pez} and \pkg{phyr}'s statistical models does not appear to have a large effect on the estimates of the remaining parameters in the model but underestimates the residual standard deviation (Figure \ref{msplot}).
+The absence of correlations in \pkg{pez} and \pkg{phyr}'s statistical models has little effect on the other parameter estimates but leads to underestimates of the residual standard deviation (Figure \ref{msplot}).
 
-Although the parameter estimates are similar across platforms for the multi-group simulation fits, computational efficiency varies enormously across platforms and sample size.
-For example, the new formulation is implemented in both \pkg{lme4} and \pkg{glmmTMB}, but \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4} for the cases studied here.
-Comparing \pkg{glmmTMB} to \pkg{pez} and \pkg{phyr}, the median time for \pkg{glmmTMB} to fit a 50-species model is $\approx$ 9 versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr} respectively. 
-\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was not practical to fit 500-species models with \pkg{pez} and \pkg{phyr}, because computational speed scaled faster than linearly with sample size.
-\pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
+Although the parameter estimates are similar across platforms, computational efficiency varies enormously across platforms and sample size.
+Comparing our two implementations, \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
+\pkg{glmmTMB} is also faster than \pkg{pez} and \pkg{phyr}: the median time for \pkg{glmmTMB} to fit a 50-species model is $\approx$ 9, versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr}.
+\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was impractical to fit 500-species models with \pkg{pez} and \pkg{phyr}.
 
@@
 
-We have simulated relatively complex models containing phylogenetic variation in both intercepts and slopes, as well as within-species variation that is quantifiable.
-These models are intrinsically more complex than some simple platforms for phylogenetic regression can handle, which may seem unfair; nevertheless, our models are certainly less complex than evolutionary processes occurring in nature.
-Our results show that models that cannot match the full ``simulation world'' perform poorly even for the parameters they do estimate; it is important to understand the limitations of these simpler, commonly used methods.
+We have simulated relatively complex models containing phylogenetic variation in species intercepts and slopes, as well as within-species variation.
+Comparing our fits with simple platforms for phylogenetic regression that cannot handle these complexities may seem unfair; nevertheless, our models are certainly less complex than evolutionary processes occurring in nature.
+Models that cannot match the full simulations perform poorly even for the parameters they do estimate.
 
-Even our relatively simple multi-group model described in (\ref{eq:multiple_glmm}) can incorporate many layers of complexity.
+Even the relatively simple multi-group model described in (\ref{eq:multiple_glmm}) can incorporate many layers of complexity.
 In theory, as long as we have enough data and enough computational power, models that can incorporate more of the complexity will always describe a biological system better.
 However, real applications are always data-constrained.
-Deciding on a practically appropriate level of model complexity for a given problem and data set is an open and difficult general problem in statistical modeling, not just in phylogenetic studies.
+Establishing the practical level of model complexity for a given problem and data set is an open and difficult general problem throughout statistical modeling, not just in phylogenetic studies.
 %For example, should one use simple models that may be overly conservative or
-risk overfitting by using more ambitious models?
+% risk overfitting by using more ambitious models?
 %What are the relative costs and benefits of using a step-down procedure
@@
 
-In classic GLMMs, random effects are used to account for group-level variation and they take more complicated forms as the experimental design becomes more complex, e.g., random-slopes models.
-Random-slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each group) and often require more data overall for reliable estimates, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
+Random-slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each group) and generally require more data overall, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
 Neglecting random slopes can lead to biased fixed effect estimates with inadequate coverage and inflated type~I errors \citep{schielzeth2008conclusions}.
 
-Nevertheless, for PGLMMs, it is hard to account for all forms of complexities and decide when it is best to use phylogenetic random effects, simple grouping, or both. 
-Optimal model complexity depends on experimental design and whether the data provides enough signal to estimate these different levels of variation, which can be strongly confounded.
-For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ will be sufficient. 
-However, if multiple observations are available per species, then these simple methods may confound tip variation with residual variation.
-In this case, multiple observations can summarized to a single measurement (for example, weighted-mean) per species to avoid confounding residual and tip variation.
-This is equivalent to assuming homoscedasticity using inverse-variance weights for unbalanced datasets. 
-Alternatively, when the within-species variance is actually of interest, accounting for within-species variation (i.e., adding species-level random effects in our example) can automatically handle multiple observations per species.
-It may be easier to be conservative and include both phylogenetic random effects and iid grouping random effects by simplifying the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}).
-Users should be aware of two essential questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes?
-Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}?
+Nevertheless, it is impossible to account for all possible complexities.
+The best model --- whether using phylogenetic random effects, simple grouping, or both --- depends on the experimental design and whether there are enough data to separate different levels of variation, which can be strongly confounded.
+If multiple observations are available per species, then simple methods like Pagel's $\lambda$ will confound tip variation with residual variation. \bmb{cite boettiger blog??}
+Multiple observations can be collapsed to a single value (such as the mean) per species, with analyses weighting each species by its number of observations.
+Alternatively, if the within-species variance is of interest, a phylogenetic mixed model can separate tip variation from within-species variation and measurement error.
+More generally, phylogenetic mixed models can be simplified to ordinary mixed models, at the cost of taxonomic detail, by simplifying the phylogenetic tree to a strictly hierarchical set of nested higher-level taxa \citep{bunnefeld2012island}).
+Users should be aware of two essential questions when fitting random-slope models: do they have enough information to practically estimate the random slopes, and what are the potential costs of ignoring them \citep{schielzeth2008conclusions}?
 % However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
@@
 
-We have presented a range of classical phylogenetic comparative methods (i.e.
-phylogenetic least squares, linear and mixed models) to fit models that incorporate different levels of phylogenetic variation.
-Even within this scope, there is additional room for exploration, such as phylogenetic multivariate response models; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck (OU) model which accounts for both selection and drift processes \citep{butler2004phylogenetic}); Bayesian approaches \citep{hadfield2010general}; and variable-rate models, where evolutionary rates vary across the phylogeny.
-The simple approach we developed here, while it cannot easily incorporate these complexities, does offer an efficient way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models.
-This approach can in principle be combined flexibly with any platform that supports independent latent variables.
-More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.
+The range of PCMs presented here can incorporate many levels and types of phylogenetic variation.
+Of course, there are further biological complexities we have neglected, such as multivariate responses; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck model \citep{butler2004phylogenetic}); and variable-rate models, which allow evolutionary rates to vary across the phylogeny.
+While it cannot easily incorporate these complexities, the approach here does offer an efficient way to analyze a wide range of evolutionary scenarios.
+The general principle of encoding phylogenetic structure in the random effects model matrix can be implemented with any platform that supports continuous latent variables.
+Our implementation allows users to explore new ideas by fitting phylogenetic mixed models with complex random effects to large data sets.
 
@@
 
-ML and BMB conceived the ideas and designed methodology; ML and BMB implemented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analyzed the results; ML wrote the first draft of the manuscript.
+ML and BMB conceived the ideas and designed methodology; ML and BMB implemented the code in \pkg{lme4} and \pkg{glmmTMB}; ML ran all simulations; ML and BMB analyzed the results; ML wrote the first draft.
 Both authors contributed critically to the drafts and gave final approval for publication.

### 2020-02-19 phyloglmm_ms.tex
fixed all BMB comments, reply to profile CI

@@
 \end{equation}
-The model has two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma^2_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma^2_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random slope and intercept $\Sigma_{\mathrm{phy_{int-slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).
+The model has two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma^2_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma^2_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random intercept and slope $\Sigma_{\mathrm{phy_{int-slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).
 The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic effects on the slope ($b_{\mathrm{phy_{slope}}}$) and intercept ($b_{\mathrm{phy_{int}}}$) for each branch of the phylogeny; i.e.
@@
 \bmb{no profiles CIs?
-is that problematic?}
+is that problematic?} \mli{Profile is EXTREMELY slow, but I did double check with an handful of examples ~50 simulations and compared them to wald}
 We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.
@@
 \pkg{pez} and \pkg{phyr} estimates for n = 100 and 500 are missing because the models did not converge within 30 minutes.
-\bmb{add English strip labels as in Fig 2?}
   }
@@
 The best model --- whether using phylogenetic random effects, simple grouping, or both --- depends on the experimental design and whether there are enough data to separate different levels of variation, which can be strongly confounded.
-If multiple observations are available per species, then simple methods like Pagel's $\lambda$ will confound tip variation with residual variation. \bmb{cite boettiger blog??}
+If multiple observations are available per species, then simple methods like Pagel's $\lambda$ will confound tip variation with residual variation \citep{boettiger2013is}.
 Multiple observations can be collapsed to a single value (such as the mean) per species, with analyses weighting each species by its number of observations.

### 2020-02-19 plot.R

@@ -123,7 +123,7 @@ gg_sscoverage <- (ggplot(data=ss_coverage
 	)
 	+ facet_wrap(~Platform, nrow = 1)
         + geom_point(size=4, alpha=0.5)
-        + geom_linerange(aes(ymin=lwr,ymax=upr))
+        # + geom_linerange(aes(ymin=lwr,ymax=upr))
 	+ scale_shape_discrete("Parameters",labels=c(expression(beta[0])
 	                              , expression(beta[1])))
 	+ geom_hline(aes(yintercept=0.95))
@@ -163,23 +163,25 @@ msdat <- (msdat_raw
 			, sdtype = factor(sdtype, levels=c("phylo_int","phylo_cor","phylo_X"
 			                                   , "species_int", "species_cor", "species_X"
 			                                   , "phylo_interaction", "site_int", "resid")
-			                  , labels=c(expression(paste(Sigma[phy[int]]))
-			                             , expression(paste(rho[phy[int-slope]]))
-			                             , expression(paste(Sigma[phy[slope]]))
-			                             , expression(paste(sigma[sp[int]]))
-			                             , expression(paste(rho[sp[int-slope]]))
-			                             , expression(paste(sigma[sp[slope]]))
-			                             , expression(paste(Sigma[phy[sp:site]]))
-			                             , expression(paste(sigma[site]))
-			                             , expression(sigma[epsilon]))
+			                  , labels=c(expression(paste("Phylogenetic random intercept ",Sigma[phy[int]]))
+			                             , expression(paste("Phylo random intercept-slope correlation ",rho[phy[int-slope]]))
+			                             , expression(paste("Phylogenetic random slope ",Sigma[phy[slope]]))
+			                             , expression(paste("Species random intercept ",sigma[sp[int]]))
+			                             , expression(paste("Species random intercept-slope correlation ",rho[sp[int-slope]]))
+			                             , expression(paste("Species random slope ",sigma[sp[slope]]))
+			                             , expression(paste("Phylo random species-group interaction ",Sigma[phy[sp:site]]))
+			                             , expression(paste("random group ",sigma[site]))
+			                             , expression(paste("Residual ",sigma[epsilon])))
 			)
 		)
 )
+
 
 gg_ms <- (gg_ss
 	%+% msdat
[…]

### 2020-02-19 phyloglmm_ms.tex
deleting profile comment

@@
 We computed 95\% Wald confidence intervals for frequentist methods and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
-\bmb{no profiles CIs?
-is that problematic?} \mli{Profile is EXTREMELY slow, but I did double check with an handful of examples ~50 simulations and compared them to wald}
 We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.

### 2020-02-20 plot.R
caption and x-label tweaks

@@ -79,7 +79,8 @@ gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform,
 	+ scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),labels = scaleFUN)
 	+ scale_color_manual(values=colvec)
 	+ scale_fill_manual(values=colvec)
-	+ ylab("Time (seconds)")
+    + ylab("Time (seconds)")
+    + xlab("Number of Species")
 	# + ggtitle("Single site time")
   + theme_bw()
 )
@@ -229,7 +230,8 @@ gg_mscoverage <- (ggplot(data=ms_coverage
 + annotate("rect", xmin=0, xmax=5
            , ymin=0.95 - 2*sqrt(0.95*0.05/100)
            , ymax=0.95 + 2*sqrt(0.95*0.05/100), alpha=0.1, fill="blue")
-+ theme(panel.spacing = unit(0,'lines'))
+    + theme(panel.spacing = unit(0,'lines'))
+    + xlab("Number of Species")
 )
 
 print(mscoverage<- gg_mscoverage 	+ scale_color_manual(values=colvec2))

### 2020-02-20 phyloglmm_ms.tex
caption and x-label tweaks
submitted, need to chat about merging with lme4/glmmTMB

@@
 
-%\include{abstract}
+\include{abstract}
 
@@
   \includegraphics[scale=0.7]{./figure/mscoverage.pdf}
-  \caption{Comparison of multi-group model coverage.}
+  \caption{Comparison of multi-group model coverage. The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits.}
   \label{msplot_coverage}
@@
 
+\section*{Achnowledgements}
+
+We would like to thank Jonathan Dushoff for thoughtful comments.
+This study was funded by NSERC Discovery Grant.
+
 \section*{Authors’ contributions}

### 2020-10-05 response.md
downloading response

Dear Authors,

Thank you for submitting your research to MEE and apologies for the long time getting back to you, which partly a consequence of the ongoing lock-down work-from-home situation. Your manuscript has been reviewed by two highly qualified reviewers who found your work interesting while raising several points about the presentation and extent of the simulations, the scalability of the method and suggesting the inclusion of empirical examples. Based on the reviewers’ judgment and my own take on the paper I cannot recommend the acceptance of this paper at this point, but would encourage you to resubmit a thoroughly re-worked version.

As currently presented I found this paper a little bit in between an application note and a research paper but lacking some key components from both, but turning it into a research paper would require, in my opinion, the addition of an empirical example (see also comments from reviewer 1).

The benchmarking of the new method is interesting but is mixing too many confounding factors linked with the different implementations of the various packages (see also comment from reviewer 2). What is the performance of the new model all else being equal? How much of the performance results is actually due to differences the optimization among the different R packages?

In the Introduction, a broader perspective about PCMs should be given before jumping into the specifics of the models tested here. Phylogenetic comparative methods include lots of other things, see e.g. Luke Harmon’s book: https://lukejharmon.github.io/pcm/chapters/. In general, a better reference to previous work is required and some claims of the paper should be either demonstrated or removed. For example, stating that “existing [PCM] procedures are either insufficiently flexible or too computationally demanding to analyze large data sets” is simply not true. There are many research papers applying PCMs to large datasets (and as a side note: how large should a dataset be to be considered large?).

A sentence like “Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely” needs references supporting it. For example, Kostikova et al. 2016 (DOI:10.1093/sysbio/syw010) present a model that seems relevant here. Further, claiming that ‘brms’ is more efficient than ‘MCMCglmm’ needs to be backed up by either a reference or a test (see also comments from the reviewers).

In the methods it should be better clarified what is the state of the art and what are the novel parts you are presenting in this study.

In lines 74-75 “phylogenetic mixed models in community ecology” are mentioned but it is unclear what the link is to the previous paragraph and the to the following sentences.

Why are the parameter estimates of the two implementations of the new model different in Fig. 5? And, in the same figure, why were runs of alternative implementations killed after 30 min?

I am not sure about the point raised in the Discussion: “Establishing the practical level of model complexity for a given problem and data set is an open and difficult general problem throughout statistical modeling, not just in phylogenetic studies.” Model complexity is typically addressed by model testing, which is not mentioned here.

In the discussion about within-species variance and measurement errors (l. 272-274) you could make several references to previous work, including several papers published in MEE.

I hope you will find these and the reviewers’ comments useful.
Best wishes,

Daniele Silvestro

Reviewer(s)' Comments to Author:
Reviewer: 1

Comments to the Corresponding Author
The authors propose a reformulation of the phylogenetic mixed model in order to make that class of PCM faster and more flexible. Given, their results, their goal has been reached. I found the paper well written and informative and their method of interest for a large audience. I don’t have any remarks about the scientific aspect nor about the benchmarking of that application which are thorough and well detailed.

However, I think the article could fail to reach the extent of its audience by missing several key elements:

- Because that work isn’t a theoretical work but instead the description of an applied statistics method, adding a few examples, such as the one provided at l68-73, would help the reader understanding the value of those models. I think examples should particularly be aimed at the “Simulation” § l132-155, where they could shed light on the subtle differences between the parameters
- In the same spirit of applied statistics, a brief description (or table) of which functions to use in the libraries lme4 and glmmTMB to make these models work would make it easier to start. There should also be mention of the tutorial available in the repository
- It would also be beneficial to add an application of the method with biological data directly in the article. The more your article is associated with biological ideas, the more users you will reach in my opinion.
- Naming that class of models (even though it is still pglmm) could help the reader follow the results section and identify the method as well.

These remarks might sound cosmetic but I think they might help to reach a broader audience and help all kind of users getting started with those models.

I recommend the authors to think on those ideas to improve the visibility of their method and the journal to ask them to do so in order to increase the impact of the article.

In a purely scientific perspective, I have no objections to publish that article.

Minor comments:

L26. In the Abstract, a claim has been made about the benefit of the method against unbalanced observational designs, but we lack insight on that point throughout the article. I think that sort of claim should be tested and I expect the Bayesian methods to perform better with that kind of problems.

L10. Small grammatical error: “can allow to incorporate”

Reviewer: 2

Comments to the Corresponding Author
This paper proposes an alternative formulation for the phylogenetic generalized linear mixed model which can utilize existing R packages for linear mixed models (lme4 and glmmTMB). The authors illustrate the advantages of their proposed method through simulations. The paper is well written and easy to follow. My main concern is that the proposed method is not scalable. That is, the computational efficiency of the proposed method only relies on the highly optimized implementation of lme4 and glmmTMB and its algorithm is actually "slower" than existing algorithms in phylogenetics. The followings are my specific comments:

1. What is the computational complexity of the proposed method? Does the alternative formulation help avoid matrix inversion such that the complexity is linear with respect to the number of species? Note that this is the case for many phylogenetic packages including phylolm and MCMCglmm. I am aware that it is hard or even impossible to achieve linear-complexity for all models. However, could linear-complexity be achieved by the proposed methods for some simple models such as the standard phylogenetic linear regression (equation 1)?

2. In the paper, the authors replace the package MCMCglmm by the package brms. Although the implementation of MCMCglmm is not as efficient as brms, the complexity of MCMCglmm is linear with respect to the number of species while brms should be cubic (since it doesn't have a special method to avoid matrix inversion). It may be a good idea to add a simulation with a series of trees from dozens to thousands of species for showing the scalability of these methods (including MCMCglmm).

3. The models in simulations are not sufficiently described. What are the formulae for the phylogenetic random intercept variance, phylogenetic random slope variance, and covariance between phylogenetic random intercept and slope, and so on? Do they depend on the phylogenetic tree? The authors provide a clear explanation for how their proposed method can be applied to the standard phylogenetic regression (equation 1) but do not include it in the simulation section. I suggest the authors add this model to the simulation part for illustrating the applicability of their proposed method for this simple case.

+++++
RESUBMISSION INSTRUCTIONS
Please note that resubmitting your manuscript does not guarantee eventual acceptance, and that your resubmission may be subject to re-review before a decision is rendered. Please also ensure that your altered manuscript still conforms to our word limit of 3000 for applications.

Once you have made the suggested changes, go to https://mc.manuscriptcentral.com/mee-besjournals and login to your Author Centre. Click on "Manuscripts with Decisions," and then click on "Create a Resubmission" located next to the manuscript number. Then, follow the steps for resubmitting your manuscript.

Because we are trying to facilitate timely publication of manuscripts submitted to Methods in Ecology and Evolution, your new manuscript should be uploaded within 12 weeks. The deadline for your resubmission is 05-Aug-2020. If it is not possible for you to submit your manuscript by that date, please get in touch with the editorial office, otherwise we will consider your paper as a completely new submission.

### 2020-10-12 response.md
pushing first pass, need to start working on it

@@
 
+MLi: Thank you for letting us resubmit.
+
 As currently presented I found this paper a little bit in between an application note and a research paper but lacking some key components from both, but turning it into a research paper would require, in my opinion, the addition of an empirical example (see also comments from reviewer 1).
+
+MLi: This paper is intended as an application/software paper.
 
@@
 How much of the performance results is actually due to differences the optimization among the different R packages?
+
+MLi: The model is the mathematically equivalent, thus, the difference is the speed/efficiency.
+MLi: Comparing pez and phyr uses lme4 correlation structure. This is the most direct comparision.
 
@@
 
+Todo: Revisit and rewrite.
+
 A sentence like “Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely” needs references supporting it.
@@
 
+Todo: add reference!
+
 In the methods it should be better clarified what is the state of the art and what are the novel parts you are presenting in this study.
+
+Todo:... MCMCglmm.
 
@@
 
+Todo: revisit
+
 Why are the parameter estimates of the two implementations of the new model different in Fig. 5? And, in the same figure, why were runs of alternative implementations killed after 30 min?
+
+Todo: Help? lme4 vs glmmTMB differences?
 
@@
 
+Todo: What do they mean by model testing here? This feels like snooping and cheating if we test and select models.
+
 In the discussion about within-species variance and measurement errors (l. 272-274) you could make several references to previous work, including several papers published in MEE.
+
+Todo: find references.
 
@@
 I don’t have any remarks about the scientific aspect nor about the benchmarking of that application which are thorough and well detailed.
+
+MLi: Thanks.
 
@@
 
+Todo: Thanks, these are wonderful and helpful suggestion. We have included a tutorial in the released version on zenodo/github. We have now included the link in the paper....
+
 Minor comments:
@@
 I think that sort of claim should be tested and I expect the Bayesian methods to perform better with that kind of problems.
+
+Todo: Table 1 shows all the possbilities (double check) with existing package.
 
@@
 
+Todo: Why is it slower? Do we need to profile the pieces?
+
 1.
@@
 
+Todo: Exactly, we are avoiding the matrix inversion step.
+
 2.
@@
 It may be a good idea to add a simulation with a series of trees from dozens to thousands of species for showing the scalability of these methods (including MCMCglmm).
+
+Todo: Will look into it.
 
@@
 I suggest the authors add this model to the simulation part for illustrating the applicability of their proposed method for this simple case.
+
+Todo: what? go back to the paper and see if we say this clearly, this should be trival and clear.
 

### 2020-10-26 response.md
minor comments

@@
 MLi: This paper is intended as an application/software paper. 
+TODO: didn't we already do this? double check. We have included the example provided in Chapter 11 Garamszegi and Li et al. Dune mellow example.
 
@@
 Todo: Revisit and rewrite.
+It is true many research papers used PCMs to large datasets, but it depends on the complexity of the model.
+I.e.
+If we simply use GLM methods in Table 1, it is relative fast even with large datasets.
+If we have more complicated models, then it will be very computationally demanding.
 
@@
 
-Todo:... MCMCglmm.
+Todo:... MCMCglmm?
 

### 2020-12-07 response.md
starting reviews

--- 2020-10-05 response.md
@@
 
+**MLi: Thank you for letting us resubmit.**
+
 As currently presented I found this paper a little bit in between an application note and a research paper but lacking some key components from both, but turning it into a research paper would require, in my opinion, the addition of an empirical example (see also comments from reviewer 1).
+
+**MLi: This paper is intended as an application/software paper. 
+TODO: didn't we already do this? double check. We have included the example provided in Chapter 11 Garamszegi and Li et al. Dune mellow example.**
 
@@
 How much of the performance results is actually due to differences the optimization among the different R packages?
+
+**MLi: The model is the mathematically equivalent, thus, the difference is the speed/efficiency.**
+
+**MLi: Comparing pez and phyr uses lme4 correlation structure. This is the most direct comparision. **
 
@@
 
+**Todo: Revisit and rewrite.
+It is true many research papers used PCMs to large datasets, but it depends on the complexity of the model.
+I.e.
+If we simply use GLM methods in Table 1, it is relative fast even with large datasets.
+If we have more complicated models, then it will be very computationally demanding.**
+
 A sentence like “Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely” needs references supporting it.
@@
 
+**Todo: add reference!**
+
 In the methods it should be better clarified what is the state of the art and what are the novel parts you are presenting in this study.
+
+**Todo:... MCMCglmm is probably the state of the art**
 
@@
 
+**Todo: revisit**
+
 Why are the parameter estimates of the two implementations of the new model different in Fig. 5? And, in the same figure, why were runs of alternative implementations killed after 30 min?
+
+**Todo: Help? lme4 vs glmmTMB differences?**
 
@@
 
+**Todo: What do they mean by model testing here? This feels like snooping and cheating if we test and select models.**
+
 In the discussion about within-species variance and measurement errors (l. 272-274) you could make several references to previous work, including several papers published in MEE.
+
+**Todo: find references.**
 
@@
 I don’t have any remarks about the scientific aspect nor about the benchmarking of that application which are thorough and well detailed.
+
+**MLi: Thanks.**
 
@@
 
+**Todo: Thanks, these are wonderful and helpful suggestion. We have included a tutorial in the released version on zenodo/github. We have now included the link in the paper.... **
+
 Minor comments:
@@
 I think that sort of claim should be tested and I expect the Bayesian methods to perform better with that kind of problems.
+
+**Todo: Table 1 shows all the possbilities (double check) with existing package.**
 
@@
 
+**Todo: Why is it slower? Do we need to profile the pieces?**
+
 1.
@@
 
+**Todo: Exactly, we are avoiding the matrix inversion step.**
+
 2.
@@
 It may be a good idea to add a simulation with a series of trees from dozens to thousands of species for showing the scalability of these methods (including MCMCglmm).
+
+**Todo: Will look into it.**
 
@@
 I suggest the authors add this model to the simulation part for illustrating the applicability of their proposed method for this simple case.
+
+**Todo: what? go back to the paper and see if we say this clearly, this should be trival and clear. **
 

### 2020-12-07 phylotreelike.rmd
starting reviews

@@
              , LR3 = null-null2
-             , LRT = ll > null # abs(LR) > c/2
+             , LRT = ll > null # abs(LR) > qchisq(0.95,df=i-2)
              , LRT1 = ll > wll
@@
              , LRT3 = null > null2 # abs(LR3) > (qchisq(0.95,df=(i-1))/2)
-             , LRT4 = wll > null
-             , LRT5 = wll > null2
+             , LRT4 = wll > null2
              )
@@
 ddprop$star_test[i-2] <- mean(dd2$LRT3)
-ddprop$wrongtree_cor[i-2] <- mean(dd2$LRT4)
-ddprop$wrongtree_test[i-2] <- mean(dd2$LRT5)
+ddprop$wrongtree_test[i-2] <- mean(dd2$LRT4)
 }

### 2020-12-07 hacked_code/1-data.R

@@ -41,7 +41,7 @@ tree_dune_phylocom = ladderize(tree_dune_phylocom)
 tree_dune_phylocom$tip.label = unname(sapply(tree_dune_phylocom$tip.label, simpleCap))
 
 # read traits
-dune.traits = read.table(file = input_files[[1]], header = T)
+dune.traits = read.table(file = "./data_clean/dune_traits_Z.txt", header = T)
 names(dune.traits)[1] = "abr3"
 # update species names
 dune.traits$abr3[dune.traits$abr3 == "Pot_pal"] = "Com_pal" # Potentilla palustris -- Compalu pal...

### 2020-12-07 plot.R

@@ -170,8 +170,8 @@ msdat <- (msdat_raw
 			                             , expression(paste("Species random intercept ",sigma[sp[int]]))
 			                             , expression(paste("Species random intercept-slope correlation ",rho[sp[int-slope]]))
 			                             , expression(paste("Species random slope ",sigma[sp[slope]]))
-			                             , expression(paste("Phylo random species-group interaction ",Sigma[phy[sp:site]]))
-			                             , expression(paste("random group ",sigma[site]))
+			                             , expression(paste("Phylo random species-group interaction ",Sigma[phy[sp:group]]))
+			                             , expression(paste("Random group ",sigma[group]))
 			                             , expression(paste("Residual ",sigma[epsilon])))
 			)
 		)

### 2020-12-29 response.md
pushing to BMB and work in parallel now

@@
 
-**MLi: This paper is intended as an application/software paper. 
-TODO: didn't we already do this? double check. We have included the example provided in Chapter 11 Garamszegi and Li et al. Dune mellow example.**
+**MLi: This paper is intended as an application/software paper.** 
+**TODO: didn't we already do this? We have included the example provided in Chapter 11 Garamszegi. We can add Li et al. Dune mellow example if we want.**
 
@@
 
-**MLi: Comparing pez and phyr uses lme4 correlation structure. This is the most direct comparision. **
+**MLi: Comparing pez and phyr uses lme4/glmer correlation structure. This is the most direct comparision of the proposed method (i.e. correlation vs random effect matrix structure. **
 
@@
 
-**Todo: add reference!**
+** MLi: This line is now removed. We have confirmed that the suggested example is relevant to our framework in partitioning different levels of variation. **
 
@@
 
-**Todo:... MCMCglmm is probably the state of the art**
+**MLi: Bayesian methods are the state of the art for more complex. I've now reworded the last parameter in the intro before methods.**
 
@@
 
-**Todo: revisit**
+**MLi: This refers to the recent developed frequentist tools for phyloglmm and used in community ecology applications. We have now reword and clarify this sentence. **
 
@@
 
-**Todo: Help? lme4 vs glmmTMB differences?**
+**To BMB: Help? lme4 vs glmmTMB differences? Maybe convergence settings? Should we bother if they are close enough?**
+**MLi: The parameters that are in both single and multiple site models remain the same.
+We have now added one more alternative implementation run for the 100 species simulation with the slow platforms.
+The goal of this paper is to show the inflexiably of existing tools and the efficiency of the new implemented approach.
+**
 
@@
 
-**Todo: What do they mean by model testing here? This feels like snooping and cheating if we test and select models.**
+**To BMB: What do they mean by model testing here? This feels like snooping and cheating if we test and select models.**
 
@@
 
-**Todo: find references.**
+**MLi: added one already, going to find another one.**
 
@@
 I think examples should particularly be aimed at the “Simulation” § l132-155, where they could shed light on the subtle differences between the parameters
+
+**Todo**
+
 - In the same spirit of applied statistics, a brief description (or table) of which functions to use in the libraries lme4 and glmmTMB to make these models work would make it easier to start.
 There should also be mention of the tutorial available in the repository
+
+**MLi: we have now added a tutorial**
+
 - It would also be beneficial to add an application of the method with biological data directly in the article.
 The more your article is associated with biological ideas, the more users you will reach in my opinion.
+
+**Going to be a little annoying looking for an example with replicates.**
+
 - Naming that class of models (even though it is still pglmm) could help the reader follow the results section and identify the method as well.
+
+**TO BMB: What do they mean by class?**
 
@@
 **Todo: Table 1 shows all the possbilities (double check) with existing package.** 
+**Todo: Add it to the tutorial?**
 
 L10. Small grammatical error: “can allow to incorporate”
+
+**MLi: Fixed.**
 
@@
 
-**Todo: Why is it slower? Do we need to profile the pieces?**
+**To BMB: Why is it slower? Do we need to profile the pieces?**
 
@@
 
-**Todo: Exactly, we are avoiding the matrix inversion step.**
+**MLi: Exactly, we are avoiding the matrix inversion step.**
 
@@
 
-**Todo: Will look into it.**
+**MLi: Thanks.
+We have now added MCMCglmm to our simulation.
+For our simple simulation test, MCMCglmm is turned out to be faster based on your convergence critrion which heavily depend on the number of iteration.
+However, the goal is to look at the granlarity of frequentist approaches.
+Even for simple linear models, Bayesian approaches are much slowly due to MCMC compared to least squares.**
 
@@
 
-**Todo: what? go back to the paper and see if we say this clearly, this should be trival and clear. **
+**Todo: what?
+go back to the paper and see if we say this clearly, this should be trival and clear.
+The formulas are clear!
+We are using the variance of the phylogenetic tree when simulating the b's.**
 

### 2020-12-29 phyloglmm_ms_resubmit.tex
pushing to BMB and work in parallel now

--- 2020-02-20 phyloglmm_ms.tex
@@
 These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
-If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation.
-Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.
+If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging}.
 
@@
 For example, suppose we wish to fit a PCM to predict species' brain size from their body size \cite{felsenstein1985phylogenies} using a mixed-effect model.
-The standard PCM allows for phylogenetic correlations in the residuals of the relationship between body and brain size; the phylogenetic mixed model can allow incorporate both
+The standard PCM allows for phylogenetic correlations in the residuals of the relationship between body and brain size; the phylogenetic mixed model can allow to incorporate both
 variation in brain size among taxa with similar body sizes (random intercepts) and
@@
 
-Several recent studies have used phylogenetic mixed models in
-community ecology \citep{nowakowski2018phylogenetic, li2017canfun}.
+Several recent studies developed new tools to fit phylogenetic mixed models in
+community ecology applications \citep{nowakowski2018phylogenetic, li2017canfun}.
 However, the phylogenetic mixed modeling tools that allow extensions such as random slopes
-and separation of tip and observation error in a frequentist framework are inflexible; thus, biologists needing to fit random-slopes models typically turn to Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms} (Table \ref{table:model}).
+and separation of tip and observation error in a frequentist framework are inflexible; thus, biologists needing to fit more complex models typically turn to Bayesian approaches which is the state of the art, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms, kostikova2016bridging} (Table \ref{table:model}).
 
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.8,page=1]{./figure/phylotree.png}
+  \includegraphics[scale=0.8]{./figure/phylotree.png}
   \caption{Three-species phylogenetic tree.}
@@
 
+\section*{Results}
+
 In supplementary materials, we reproduce the examples in chapter 11 of \cite{garamszegi2014modern} using phylogenetic GLMMs based on \pkg{lme4} and \pkg{glmmTMB}.
-
-\section*{Results}
 
@@
 Multiple observations can be collapsed to a single value (such as the mean) per species, with analyses weighting each species by its number of observations.
-Alternatively, if the within-species variance is of interest, a phylogenetic mixed model can separate tip variation from within-species variation and measurement error.
+Alternatively, if the within-species variance is of interest, a phylogenetic mixed model can separate tip variation from within-species variation and measurement error \citep{kostikova2016bridging}.
 More generally, phylogenetic mixed models can be simplified to ordinary mixed models, at the cost of taxonomic detail, by simplifying the phylogenetic tree to a strictly hierarchical set of nested higher-level taxa \citep{bunnefeld2012island}).

### 2020-12-29 phyloglmm_ms_resubmit.tex
minor change, deleting a line

@@
 However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
-\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique.
+\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, and recently \pkg{brms}, which uses a more computationally efficient MCMC technique.
 
@@
 All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
-We parameterize the models with correlation $\rho_{xy}$ rather than covariance $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$).
-%\bmb{???}
-%mli{Just to match what lme4/glmmtmb outputs.}
 Table \ref{table:platform} shows the parameters that are estimable for each platform. 

### 2020-12-29 phyloglmm_ms_resubmit.tex
adding more citations

@@
 These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
-If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging}.
+If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging, de2012bayesian}.
 

### 2020-12-30 example.rmd
example running ok again

@@
 datR <- (datR 
-  %>% mutate(sp = species
-        , animals = phylo
+  %>% mutate(sp = factor(species)
+        , animals = factor(phylo)
         )
@@
 
-dat <- dat %>% mutate(sp=phylo) 
+dat <- dat %>% mutate(sp=factor(phylo)) 
 phylo_glmm_fit <- phylo_glmm(phen_pois~cofactor+(1|sp)+(1|obs)

### 2020-12-30 glmmTMB_phylo_setup.R

@@ -2,7 +2,7 @@ glmmTMBphylo <- function (formula, data = NULL, family = gaussian(), ziformula =
                            dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL, 
                            phylonm = NULL,
                            na.action = na.fail, se = TRUE, verbose = FALSE, doFit = TRUE, 
-                           control = glmmTMBControl(), REML = FALSE) 
+                           control = glmmTMBControl(), REML = FALSE, map=NULL, sparseX = NULL) 
 {
   call <- mf <- mc <- match.call()
   if (is.character(family)) {
@@ -90,7 +90,7 @@ glmmTMBphylo <- function (formula, data = NULL, family = gaussian(), ziformula =
   TMBStruc <- mkTMBStrucphylo(formula, ziformula, dispformula, combForm, 
                                mf, fr, yobs = y, respCol, weights, contrasts = contrasts, 
                                family = family, se = se, call = call, verbose = verbose, 
-                               REML = REML, phylonm=phylonm, phyloZ=phyloZ)
+                               REML = REML, phylonm=phylonm, phyloZ=phyloZ, map=map, sparseX=sparseX)
   TMBStruc$control <- lapply(control, eval, envir = TMBStruc)
   if (!doFit) 
     return(TMBStruc)
@@ -102,7 +102,7 @@ mkTMBStrucphylo <- function (formula, ziformula, dispformula, combForm, mf, fr,
                               yobs, respCol, weights, contrasts=contrasts, size = NULL, family, se = NULL, phyloZ = phyloZ, 
                               phylonm=phylonm,
                               call = NULL, verbose = NULL, ziPredictCode = "corrected", 
-                              doPredict = 0, whichPredict = integer(0), REML = FALSE) 
+                              doPredict = 0, whichPredict = integer(0), REML = FALSE, map =NULL, sparseX=NULL) 
 {
   if (!is(family, "family")) {
     if (is.list(family)) {
@@ -226,7 +226,7 @@ mkTMBStrucphylo <- function (formula, ziformula, dispformula, combForm, mf, fr,
                           grpVar, condList, ziList, dispList, condReStruc, ziReStruc, 
                           family, contrasts, respCol, allForm = lme4:::namedList(combForm, 
                                                                                  formula, ziformula, dispformula), fr, se, call, verbose, 
-                          REML))
[…]

### 2021-01-29 response.md
adding more citations
more minor things in the response

@@
 **MLi: This paper is intended as an application/software paper.** 
-**TODO: didn't we already do this? We have included the example provided in Chapter 11 Garamszegi. We can add Li et al. Dune mellow example if we want.**
+**TODO: didn't we already do this?
+We have included the example provided in Chapter 11 Garamszegi (DONE) in our release branch.
+We can add Li et al.
+Dune mellow example if we want, however, one of the reason we went through the trouble is because we found some bugs in their example with phylo-interactions.**
 
@@
 
-**MLi: added one already, going to find another one.**
+**Done. We added more citation on this topic.**
 
@@
 **Going to be a little annoying looking for an example with replicates.**
+**Skip?**
 

### 2021-02-01 vignette.Rmd
updating example

--- 2018-05-02 vignette.Rmd
@@
 # Basic parameters
-nspp <- 3
+nspp <- 10
 
@@
 
-fitmod <- phylo_lmm2(Y ~ X + (1 |sp) + (1 + X|sp)
+fitmod <- phylo_lmm(Y ~ X + (1 + X|sp)
 			, data=dat 
 			, phylonm = "sp" # wishy-washy
+			, phylo = phy
 			, phyloZ = phyZ # Maybe we just want the phy object and compute this internally
-			, nsp = nspp # make default check
+			# , nsp = nspp # make default check
 			, control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")  ## make default

### 2021-02-11 response.md
way too much figure tweaking

@@
 
-**MLi: Thank you for letting us resubmit.**
+**We appreciate the opportunity to resubmit a revised version.**
 
@@
 
-**MLi: This paper is intended as an application/software paper.** 
-**TODO: didn't we already do this?
-We have included the example provided in Chapter 11 Garamszegi (DONE) in our release branch.
-We can add Li et al.
-Dune mellow example if we want, however, one of the reason we went through the trouble is because we found some bugs in their example with phylo-interactions.**
+**We intended the paper as an application/software paper, not as a research paper: there is some novelty here, but as our primary innovation is to recast the model in a mathematically equivalent but more convenient/efficient form, we would prefer to treat this as an application.**
+
+**BMB: I think if we have a non-problematic example we should include it**
 
@@
 **MLi: The model is the mathematically equivalent, thus, the difference is the speed/efficiency.**
-
 **MLi: Comparing pez and phyr uses lme4/glmer correlation structure. This is the most direct comparision of the proposed method (i.e. correlation vs random effect matrix structure. **

### 2021-02-11 phyloglmm_ms_resubmit.tex
way too much figure tweaking

--- 2020-12-29 phyloglmm_ms_resubmit.tex
@@
 \makeatother
-
-\definecolor{commentcol}{rgb}{0.345, 0.345, 0.945}
-\newcommand{\comment}[1]{\textcolor{commentcol}{[[#1]]}}%
 
@@
 \usepackage{changes}
-\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
+% \usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
 
@@
 
-Phylogenetic comparative methods (PCMs) account for evolutionary relatedness when analyzing species traits.
-Given a known phylogeny, PCMs model the relationships among species traits while incorporating relatedness; they can be used to control statistically for phylogeny, to quantify phylogenetic signal in traits, or both.
+\bmb{change PCM to PR (phylogenetic regression) throughout to narrow the scope? I made this change in the first para, would need to replace throughout if we think it's a good idea}
+Phylogenetic regressions (PR) account for evolutionary relatedness when analyzing relationships among characteristics of species.
+Given a known phylogeny, PRs model the relationships among species traits while incorporating relatedness; they can be used to control statistically for phylogeny, to quantify phylogenetic signal in traits, or both.
 
@@
 These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
-If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging}.
+If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging, devillemereuil2012bayesian}.
 
@@
   \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
+\bmb{we need a better colour for MCMCglmm.  colorspace package? }
     Total simulations $N=100$ for each category.

### 2021-02-11 plot.R

@@ -1,5 +1,4 @@
 ## plots for the ms
-
 library(dplyr)
 library(ggplot2)
 library(tidyr)
@@ -18,6 +17,7 @@ sub_pkgs2 <- setdiff(all_pkgs, c("gls","phylolm","brms","MCMCglmm"))
 colvec_all <- setNames(colorspace::qualitative_hcl(n=length(all_pkgs)),
                        all_pkgs)
 colvec  <- colvec_all[sub_pkgs]
+colvec2  <- colvec_all[sub_pkgs2]
 ##    c(gls="Black",phylolm="Red",lme4="Dark Blue",
 ##                 glmmTMB="Dark Green",brms="Orange",pez="Gray",phyr="Purple",MCMCglmm="Yellow")
 
@@ -92,7 +92,7 @@ sspos <- ssdat %>% group_by(Platform,size) %>%
                              Platform=="glmmTMB" ~ 1.5,
                              TRUE ~ 1))
 
-gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform,
+gg_sstime0 <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform,
                                      fill=Platform))
 	# + geom_violin(position=position_dodge(width=0.2),alpha=0.4)
 	+ geom_boxplot(outlier.colour = NULL, varwidth=TRUE, alpha=0.2)
@@ -111,15 +111,21 @@ gg_sstime <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform,
     + stat_summary(fun=median,geom="line",aes(group=Platform),
                    position=position_dodge(width=0.75),
                    size=3,alpha=0.1)
-    + geom_label(data=sspos,
+)
+
+gg_sstime <- (gg_sstime0
+          + geom_label(data=sspos,
                  aes(x=nsize,y=time,label=Platform,colour=Platform),
                  ## direction="y",
                  hjust="left",
                  nudge_x=0.2,
                  fill=NA)
-    + theme(legend.position="none")
+          + theme(legend.position="none")
 )
-print(gg_sstime)
+
+print(
+    gg_sstime
+)
 ## TODO: spend more time piddling with horizontal placement
 ## to match position_dodge effects?
 
[…]

### 2021-04-09 garamszegi_example.rmd
not sure what changed, it works perfectly fine on a fresh computer

--- 2020-12-30 example.rmd
@@
 ---
-title: "Example"
+title: "Garamszegi example"
 author: "Michael Li and Ben Bolker"
@@
 
-Similarily, fitting using glmmTMB:
+Similarly, fitting using glmmTMB:
 
@@
 
+print(packageVersion("glmmTMB"))
+print(packageVersion("lme4"))
+
 ```

### 2021-06-04 pkg/R/data.R
first pass at package

@@ -1,0 +1,6 @@
+#' data
+#'
+#' @name garamszegi_phy
+NULL
+
+

### 2021-06-04 pkg/R/glmmTMB_phylo_setup.R

@@ -1,0 +1,280 @@
+#' hacked version of glmmTMB
+#' @export
+glmmTMBphylo <- function (formula, data = NULL, family = gaussian(), ziformula = ~0,
+                           dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL,
+                           phylonm = NULL,
+                           na.action = na.fail, se = TRUE, verbose = FALSE, doFit = TRUE,
+                           control = glmmTMBControl(), REML = FALSE, map=NULL, sparseX = NULL)
+{
+  call <- mf <- mc <- match.call()
+  if (is.character(family)) {
+    if (family == "beta") {
+      family <- "beta_family"
+      warning("please use ", sQuote("beta_family()"), " rather than ",
+              sQuote("\"beta\""), " to specify a Beta-distributed response")
+    }
+    family <- get(family, mode = "function", envir = parent.frame())
+  }
+  if (is.function(family)) {
+    family <- family()
+  }
+  if (is.null(family$family)) {
+    print(family)
+    stop("'family' not recognized")
+  }
+  fnames <- names(family)
+  if (!all(c("family", "link") %in% fnames))
+    stop("'family' must contain at least 'family' and 'link' components")
+  if (length(miss_comp <- setdiff(c("linkfun", "variance"),
+                                  fnames)) > 0) {
+    warning("some components missing from ", sQuote("family"),
+            ": downstream methods may fail")
+  }
+  if (grepl("^quasi", family$family))
+    stop("\"quasi\" families cannot be used in glmmTMB")
+  link <- family$link
+  environment(formula) <- parent.frame()
+  call$formula <- mc$formula <- formula
+  if (!is.null(eval(substitute(offset), data, enclos = environment(formula)))) {
+    formula <- addForm0(formula, makeOp(substitute(offset),
+                                        op = quote(offset)))
+  }
+  environment(ziformula) <- environment(formula)
+  call$ziformula <- ziformula
+  environment(dispformula) <- environment(formula)
+  call$dispformula <- dispformula
+  m <- match(c("data", "subset", "weights", "na.action", "offset"),
+             names(mf), 0L)
+  mf <- mf[c(1L, m)]
+  mf$drop.unused.levels <- TRUE
[…]

### 2021-06-04 pkg/R/lme4_phylo_setup.R

@@ -1,9 +1,12 @@
-## new phyloglmm_setup
-if (getRversion() < "4.0.0") stop("need R version >= 4.0.x (for deparse1)")
-
-phylo.to.Z <- function(r,stand=FALSE){
+##' construct a Z matrix from a phylogeny
+##' @name phylo_machinery
+##' @param r a \code{phylo} object
+##' @param stand (?)
+##' @export
+##' @importFrom Matrix t diag
+phylo.to.Z <- function(r, stand=FALSE){
   ntip <- length(r$tip.label)
-  Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
+  Zid <- Matrix::Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
   nodes <- (ntip+1):max(r$edge)
   root <- nodes[!(nodes %in% r$edge[,2])]
   for (i in 1:ntip){
@@ -17,24 +20,38 @@ phylo.to.Z <- function(r,stand=FALSE){
   V <- vcv(r)
   # V <- V/max(V)
   sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
-  # sig <- det(V)^(1/ntip)
-  Z <- t(sqrt(r$edge.length) * t(Zid))
-  if(stand){Z <- t(sqrt(r$edge.length/sig) * t(Zid))}
+  ## sig <- det(V)^(1/ntip)
+  tZid <- Matrix::t(Zid)
+  Z <- Matrix::t(sqrt(r$edge.length) * tZid)
+  if (stand) { Z <- Matrix::t(sqrt(r$edge.length/sig) * tZid) }
   rownames(Z) <- r$tip.label
   colnames(Z) <- 1:length(r$edge.length)
   return(Z)
 }
 
-phylo_lmm <- function(formula,data,phylo,phylonm=NULL,phyloZ=NULL,control,REML,doFit,...){
-  lmod <- lFormula(formula=formula,data = data,control=control, REML=REML,phylonm=phylonm, phyloZ=phyloZ,...)
+##' phylogenetic linear mixed model
+phylo_lmm <- function(formula, data, phylo, phylonm=NULL, phyloZ=NULL, control, REML, doFit) {
+  lmod <- lFormula(formula=formula,data = data,control=control, REML=REML,phylonm=phylonm, phyloZ=phyloZ)
   devfun <- do.call(mkLmerDevfun, lmod)
[…]

### 2021-06-05 pkg/R/glmmTMB_phylo_setup.R
more pkg cleanup

@@ -1,4 +1,6 @@
 #' hacked version of glmmTMB
+#' @importFrom stats make.link na.fail update as.formula terms model.weights gaussian model.matrix
+#' @importFrom methods is new
 #' @export
 glmmTMBphylo <- function (formula, data = NULL, family = gaussian(), ziformula = ~0,
                            dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL,

### 2021-06-05 pkg/R/lme4_phylo_setup.R

@@ -30,6 +30,10 @@ phylo.to.Z <- function(r, stand=FALSE){
 }
 
 ##' phylogenetic linear mixed model
+##' @export
+##' @inheritParams lme4_utils
+##' @importFrom lme4 optimizeLmer optimizeGlmer mkMerMod mkLmerDevfun mkGlmerDevfun
+##' @importFrom lme4 findbars nobars lmerControl glmerControl updateGlmerDevfun
 phylo_lmm <- function(formula, data, phylo, phylonm=NULL, phyloZ=NULL, control, REML, doFit) {
   lmod <- lFormula(formula=formula,data = data,control=control, REML=REML,phylonm=phylonm, phyloZ=phyloZ)
   devfun <- do.call(mkLmerDevfun, lmod)
@@ -38,6 +42,8 @@ phylo_lmm <- function(formula, data, phylo, phylonm=NULL, phyloZ=NULL, control, 
 }
 
 ##' hacked version of lme4::lFormula
+##' @name lme4_utils
+##' @rdname phylo_lmm
 ##' @param formula mixed-model formula
 ##' @param data data frame
 ##' @param REML use restricted max likelihood?
@@ -122,6 +128,8 @@ lFormula <- function (formula, data = NULL, REML = TRUE, subset, weights,
 
 
 ##' hacked version of lme4::mkReTrms
+##' @rdname lme4_utils
+##' @importFrom Matrix KhatriRao fac2sparse sparseMatrix
 ##' @export
 mkReTrms <- function(bars, fr, phylonm,phyloZ,drop.unused.levels = TRUE){
   if (!length(bars))
@@ -188,6 +196,8 @@ mkReTrms <- function(bars, fr, phylonm,phyloZ,drop.unused.levels = TRUE){
 }
 
 ##' hacked version of mkBlist
+##' @rdname lme4_utils
+##' @inheritParams lme4_utils
 mkBlist <- function (x, frloc, phylonm,phyloZ, drop.unused.levels = TRUE)
 {
   frloc <- factorize(x, frloc)
@@ -243,6 +253,9 @@ mkBlist <- function (x, frloc, phylonm,phyloZ, drop.unused.levels = TRUE)
 }
 
 #' phylogenetic GLMM
+#' @rdname phylo_lmm
+#' @inheritParams lme4_utils
+#' @export
 phylo_glmm <- function(formula,data,phylo,phylonm=NULL,phyloZ=NULL,control,family){
[…]

### 2021-06-05 pkg/R/data.R
futzing with package

@@ -1,6 +1,19 @@
 #' data
 #'
-#' @name garamszegi_phy
-NULL
+#' #' Simple data from Garamszegi chapter 11
+"garamszegi_phy"
 
+#' @rdname garamszegi_phy
+"garamszegi_pois_missing"
 
+#' @rdname garamszegi_phy
+"garamszegi_repeat"
+
+#' @rdname garamszegi_phy
+"garamszegi_simple"
+
+#' @rdname garamszegi_phy
+"garamszegi_pois"
+
+#' @rdname garamszegi_phy
+"garamszegi_effect"

### 2021-06-05 pkg/R/glmmTMB_phylo_setup.R

@@ -2,18 +2,19 @@
 #' @importFrom stats make.link na.fail update as.formula terms model.weights gaussian model.matrix
 #' @importFrom methods is new
 #' @export
-glmmTMBphylo <- function (formula, data = NULL, family = gaussian(), ziformula = ~0,
-                           dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL,
-                           phylonm = NULL,
-                           na.action = na.fail, se = TRUE, verbose = FALSE, doFit = TRUE,
-                           control = glmmTMBControl(), REML = FALSE, map=NULL, sparseX = NULL)
-{
+glmmTMBphylo <- function(formula, data = NULL, family = gaussian(), ziformula = ~0,
+                         dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL,
+                         phylonm = NULL,
+                         na.action = na.fail, se = TRUE, verbose = FALSE, doFit = TRUE,
+                         control = glmmTMBControl(), REML = FALSE, map = NULL, sparseX = NULL) {
   call <- mf <- mc <- match.call()
   if (is.character(family)) {
     if (family == "beta") {
       family <- "beta_family"
-      warning("please use ", sQuote("beta_family()"), " rather than ",
-              sQuote("\"beta\""), " to specify a Beta-distributed response")
+      warning(
+        "please use ", sQuote("beta_family()"), " rather than ",
+        sQuote("\"beta\""), " to specify a Beta-distributed response"
+      )
     }
     family <- get(family, mode = "function", envir = parent.frame())
   }
@@ -25,34 +26,45 @@ glmmTMBphylo <- function (formula, data = NULL, family = gaussian(), ziformula =
     stop("'family' not recognized")
   }
   fnames <- names(family)
-  if (!all(c("family", "link") %in% fnames))
+  if (!all(c("family", "link") %in% fnames)) {
     stop("'family' must contain at least 'family' and 'link' components")
-  if (length(miss_comp <- setdiff(c("linkfun", "variance"),
-                                  fnames)) > 0) {
-    warning("some components missing from ", sQuote("family"),
-            ": downstream methods may fail")
   }
-  if (grepl("^quasi", family$family))
[…]

### 2021-06-05 pkg/R/lme4_phylo_setup.R

@@ -1,29 +1,35 @@
 ##' construct a Z matrix from a phylogeny
 ##' @name phylo_machinery
 ##' @param r a \code{phylo} object
-##' @param stand (?)
+##' @param stand standardize edge lengths by determinant of phylogenetic covariance matrix
 ##' @export
 ##' @importFrom Matrix t diag
-phylo.to.Z <- function(r, stand=FALSE){
+##' @importFrom ape vcv
+phylo.to.Z <- function(r, stand = FALSE) {
   ntip <- length(r$tip.label)
-  Zid <- Matrix::Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
-  nodes <- (ntip+1):max(r$edge)
-  root <- nodes[!(nodes %in% r$edge[,2])]
-  for (i in 1:ntip){
-    cn <- i  ## current node
-    while (cn != root){
-      ce <- which(r$edge[,2]==cn)   ## find current edge
-      Zid[i,ce] <- 1   ## set Zid to 1
-      cn <- r$edge[ce,1]            ## find previous node
+  Zid <- Matrix::Matrix(0.0, ncol = length(r$edge.length), nrow = ntip)
+  nodes <- (ntip + 1):max(r$edge)
+  root <- nodes[!(nodes %in% r$edge[, 2])]
+  for (i in 1:ntip) {
+    cn <- i ## current node
+    while (cn != root) {
+      ce <- which(r$edge[, 2] == cn) ## find current edge
+      Zid[i, ce] <- 1 ## set Zid to 1
+      cn <- r$edge[ce, 1] ## find previous node
     }
   }
-  V <- vcv(r)
-  # V <- V/max(V)
-  sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
-  ## sig <- det(V)^(1/ntip)
-  tZid <- Matrix::t(Zid)
-  Z <- Matrix::t(sqrt(r$edge.length) * tZid)
-  if (stand) { Z <- Matrix::t(sqrt(r$edge.length/sig) * tZid) }
+  tZid <- t(Zid)
+  browser()
+  Z <- t(sqrt(r$edge.length) * tZid)
+  if (stand) {
+    V <- ape::vcv(r)
+    ## V <- V/max(V)
+    sig <- exp(as.numeric(determinant(V)["modulus"]) / ntip)
+    ## sig <- det(V)^(1/ntip)
[…]

### 2021-06-06 pkg/R/glmmTMB_phylo_setup.R
working on package: basic tests, hack_function etc.

@@ -1,8 +1,29 @@
 #' hacked version of glmmTMB
-#' @importFrom stats make.link na.fail update as.formula terms model.weights gaussian model.matrix
-#' @importFrom methods is new
+## @params copied from glmmTMB as well ...
+##' @inheritParams phylo_lmm
+##' @param formula combined fixed and random effects formula, following lme4 syntax.
+##' @param data optional data frame containing model variables.
+##' @param family a family function, a character string naming a family function, or the result of a call to a family function (variance/link function) information. See \code{\link{family}} for a generic discussion of families or \code{\link{family_glmmTMB}} for details of \code{glmmTMB}-specific families.
+##' @param ziformula a \emph{one-sided} (i.e., no response variable) formula for zero-inflation combining fixed and random effects: the default \code{~0} specifies no zero-inflation. Specifying \code{~.} sets the zero-inflation formula identical to the right-hand side of \code{formula} (i.e., the conditional effects formula); terms can also be added or subtracted. \strong{When using \code{~.} as the zero-inflation formula in models where the conditional effects formula contains an offset term, the offset term will automatically be dropped}. The zero-inflation model uses a logit link.
+##' @param dispformula a \emph{one-sided} formula for dispersion containing only fixed effects: the default \code{~1} specifies the standard dispersion given any family. The argument is ignored for families that do not have a dispersion parameter. For an explanation of the dispersion parameter for each family, see \code{\link{sigma}}. The dispersion model uses a log link. In Gaussian mixed models, \code{dispformula=~0} fixes the residual variance to be 0 (actually a small non-zero value), forcing variance into the random effects. The precise value can be controlled via \code{control=glmmTMBControl(zero_dispval=...)}; the default value is \code{sqrt(.Machine$double.eps)}.
+##' @param weights weights, as in \code{glm}. Not automatically scaled to have sum 1.
+##' @param offset offset for conditional model (only).
+##' @param contrasts an optional list, e.g., \code{list(fac1="contr.sum")}. See the \code{contrasts.arg} of \code{\link{model.matrix.default}}.
+##' @param na.action how to handle missing values, see \code{\link{na.action}} and \code{\link{model.frame}}. From \code{\link{lm}}: \dQuote{The default is set by the \code{\link{na.action}} setting of \code{\link{options}}, and is \code{\link{na.fail}} if that is unset. The \sQuote{factory-fresh} default is \code{\link{na.omit}}.}
[…]

### 2021-06-06 pkg/R/lme4_phylo_setup.R

@@ -19,7 +19,6 @@ phylo.to.Z <- function(r, stand = FALSE) {
     }
   }
   tZid <- t(Zid)
-  browser()
   Z <- t(sqrt(r$edge.length) * tZid)
   if (stand) {
     V <- ape::vcv(r)
@@ -36,32 +35,45 @@ phylo.to.Z <- function(r, stand = FALSE) {
 }
 
 ##' phylogenetic linear mixed model
-##' @export
-##' @inheritParams lme4_utils
 ##' @importFrom lme4 optimizeLmer optimizeGlmer mkMerMod mkLmerDevfun mkGlmerDevfun
-##' @importFrom lme4 findbars nobars lmerControl glmerControl updateGlmerDevfun
-phylo_lmm <- function(formula, data, phylo, phylonm = NULL, phyloZ = NULL, control, REML, doFit) {
+##' @importFrom lme4 findbars nobars lmerControl glmerControl updateGlmerDevfun subbars
+##' @param formula mixed-model formula
+##' @param data data frame
+##' @param phylo phylogenetic tree in \code{\link{phylo}} format
+##' @param phylonm name of phylogenetic grouping variable
+##' @param phyloZ phylogenetic Z-matrix (see \code{\link{phylo.to.Z}})
+##' @param control control
+##' @param REML use restricted max likelihood?
+## unimplemented ... pass via ... ?
+## @param subset subset expression
+## @param weights weights
+## @param na.action na.action
+## @param offset offset
+## @param contrasts contrasts
+##' @export
+phylo_lmm <- function(formula, data, phylo, phylonm = NULL, phyloZ = NULL, control, REML) {
   lmod <- lFormula(formula = formula, data = data, control = control, REML = REML, phylonm = phylonm, phyloZ = phyloZ)
   devfun <- do.call(mkLmerDevfun, lmod)
   opt <- optimizeLmer(devfun, control = control$optCtrl)
   mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
 }
 
-##' hacked version of lme4::lFormula
-##' @name lme4_utils
-##' @rdname phylo_lmm
-##' @param formula mixed-model formula
-##' @param data data frame
-##' @param REML use restricted max likelihood?
-##' @param subset subset expression
-##' @param weights weights
-##' @param na.action na.action
-##' @param offset offset
-##' @param contrasts contrasts
-##' @param control control
[…]

### 2021-06-06 pkg/tests/testthat/test-basic.R

@@ -1,0 +1,26 @@
+library("ape")
+library("testthat")
+library("phyloglmm")
+
+phyloZ <- phylo.to.Z(garamszegi_phy)
+datG <- transform(garamszegi_simple,
+                  obs = factor(seq(nrow(garamszegi_simple))),
+                  sp = phylo)
+
+test_that("basic phylo.to.Z", {
+  expect_s4_class(phyloZ, "Matrix")
+  expect_equal(nrow(phyloZ), ape::Ntip(garamszegi_phy))
+  expect_equal(ncol(phyloZ), ape::Ntip(garamszegi_phy) + ape::Nnode(garamszegi_phy) - 1)
+})
+
+test_that("basic lme4 LMM", {
+  phylo_lmm_fit <- phylo_lmm(phen~cofactor+(1|sp)
+                           , data=datG
+                           , phylonm = "sp"
+                           , phylo = phylo
+                           , phyloZ = phyloZ
+                           , REML = TRUE
+                           , control = lme4::lmerControl(check.nobs.vs.nlev = "ignore", check.nobs.vs.nRE = "ignore")
+                             )
+  expect_equal(c(lme4::VarCorr(phylo_lmm_fit)$sp), 207.02517)
+})

### 2021-06-07 pkg/R/glmmTMB_phylo_setup.R
TODO; test names in phylo vs data

@@ -20,15 +20,19 @@
 ##' @param sparseX a named logical vector containing (possibly) elements named "cond", "zi", "disp" to indicate whether fixed-effect model matrices for particular model components should be generated as sparse matrices, e.g. \code{c(cond=TRUE)}. Default is all \code{FALSE}
 ##' @importFrom stats make.link na.fail update as.formula terms model.weights gaussian model.matrix
 ##' @importFrom methods is new
+##' @importFrom glmmTMB inForm extractForm dropHead noSpecials addForm splitForm getReStruc glmmTMBControl getGrpVar fitTMB
 #' @export
-phylo_glmmTMB <-
-
-  function(formula, data = NULL, family = gaussian(), ziformula = ~0,
+phylo_glmmTMB <-  function(formula, data = NULL, family = gaussian(), ziformula = ~0,
                          dispformula = ~1, weights = NULL, offset = NULL, contrasts = NULL, phyloZ = NULL,
                          phylonm = NULL,
                          na.action = na.fail, se = TRUE, verbose = FALSE, doFit = TRUE,
                          control = glmmTMBControl(), REML = FALSE, map = NULL, sparseX = NULL) {
+
   call <- mf <- mc <- match.call()
+
+  ## hack_function
+  check_phylo_names(phyloZ, data[[phylonm]])
+
   if (is.character(family)) {
     if (family == "beta") {
       family <- "beta_family"
@@ -66,7 +70,7 @@ phylo_glmmTMB <-
   environment(formula) <- parent.frame()
   call$formula <- mc$formula <- formula
   if (!is.null(eval(substitute(offset), data, enclos = environment(formula)))) {
-    formula <- addForm0(formula, makeOp(substitute(offset),
+    formula <- glmmTMB:::addForm0(formula, glmmTMB:::makeOp(substitute(offset),
       op = quote(offset)
     ))
   }
@@ -83,7 +87,7 @@ phylo_glmmTMB <-
   mf[[1]] <- as.name("model.frame")
   if (inForm(ziformula, quote(.))) {
     ziformula <- update(
-      glmmTMB:::RHSForm(drop.special2(formula), as.form = TRUE),
+      glmmTMB:::RHSForm(glmmTMB:::drop.special2(formula), as.form = TRUE),
[…]

### 2021-06-07 pkg/R/lme4_phylo_setup.R

@@ -1,39 +1,3 @@
-##' construct a Z matrix from a phylogeny
-##' @name phylo_machinery
-##' @param r a \code{phylo} object
-##' @param stand standardize edge lengths by determinant of phylogenetic covariance matrix
-##' @export
-##' @importFrom Matrix t diag
-##' @importFrom ape vcv
-phylo.to.Z <- function(r, stand = FALSE) {
-  ntip <- length(r$tip.label)
-  Zid <- Matrix::Matrix(0.0, ncol = length(r$edge.length), nrow = ntip)
-  nodes <- (ntip + 1):max(r$edge)
-  root <- nodes[!(nodes %in% r$edge[, 2])]
-  for (i in 1:ntip) {
-    cn <- i ## current node
-    while (cn != root) {
-      ce <- which(r$edge[, 2] == cn) ## find current edge
-      Zid[i, ce] <- 1 ## set Zid to 1
-      cn <- r$edge[ce, 1] ## find previous node
-    }
-  }
-  tZid <- t(Zid)
-  Z <- t(sqrt(r$edge.length) * tZid)
-  if (stand) {
-    V <- ape::vcv(r)
-    ## V <- V/max(V)
-    sig <- exp(as.numeric(determinant(V)["modulus"]) / ntip)
-    ## sig <- det(V)^(1/ntip)
-    ## all.equal(Z/sqrt(sig),
-    ##  t(sqrt(r$edge.length / sig) * tZid), tolerance=2e-16)
-    Z <- Z/sqrt(sig)
-  }
-  rownames(Z) <- r$tip.label
-  colnames(Z) <- 1:length(r$edge.length)
-  return(Z)
-}
-
 ##' phylogenetic linear mixed model
 ##' @importFrom lme4 optimizeLmer optimizeGlmer mkMerMod mkLmerDevfun mkGlmerDevfun
 ##' @importFrom lme4 findbars nobars lmerControl glmerControl updateGlmerDevfun subbars
@@ -52,6 +16,7 @@ phylo.to.Z <- function(r, stand = FALSE) {
 ## @param contrasts contrasts
 ##' @export
 phylo_lmm <- function(formula, data, phylo, phylonm = NULL, phyloZ = NULL, control, REML) {
+  check_phylo_names(phyloZ, data[[phylonm]])
   lmod <- lFormula(formula = formula, data = data, control = control, REML = REML, phylonm = phylonm, phyloZ = phyloZ)
[…]

### 2021-06-07 pkg/R/utils.R

@@ -1,0 +1,58 @@
+##' construct a Z matrix from a phylogeny
+##' @name phylo_machinery
+##' @param r a \code{phylo} object
+##' @param stand standardize edge lengths by determinant of phylogenetic covariance matrix
+##' @export
+##' @importFrom Matrix t diag
+##' @importFrom ape vcv
+phylo.to.Z <- function(r, stand = FALSE) {
+  ntip <- length(r$tip.label)
+  Zid <- Matrix::Matrix(0.0, ncol = length(r$edge.length), nrow = ntip)
+  nodes <- (ntip + 1):max(r$edge)
+  root <- nodes[!(nodes %in% r$edge[, 2])]
+  for (i in 1:ntip) {
+    cn <- i ## current node
+    while (cn != root) {
+      ce <- which(r$edge[, 2] == cn) ## find current edge
+      Zid[i, ce] <- 1 ## set Zid to 1
+      cn <- r$edge[ce, 1] ## find previous node
+    }
+  }
+  tZid <- t(Zid)
+  Z <- t(sqrt(r$edge.length) * tZid)
+  if (stand) {
+    V <- ape::vcv(r)
+    ## V <- V/max(V)
+    sig <- exp(as.numeric(determinant(V)["modulus"]) / ntip)
+    ## sig <- det(V)^(1/ntip)
+    ## all.equal(Z/sqrt(sig),
+    ##  t(sqrt(r$edge.length / sig) * tZid), tolerance=2e-16)
+    Z <- Z/sqrt(sig)
+  }
+  rownames(Z) <- r$tip.label
+  colnames(Z) <- 1:length(r$edge.length)
+  return(Z)
+}
+
+check_phylo_names <- function(phyloZ, data_sp) {
+  in_phy_nin_data <- setdiff(rownames(phyloZ), unique(data_sp))
+  in_data_nin_phy <- setdiff(unique(data_sp), rownames(phyloZ))
+  if (length(in_phy_nin_data) > 0 ||
+      length(in_data_nin_phy) > 0) {
+    msg <- "mismatch between tip names in phylogeny/Z-matrix and names in data:"
+    if (length(in_phy_nin_data) > 0) {
+      msg <- c(msg,
+               "\nin phyloZ but not data: ",
+               paste(in_phy_nin_data, collapse = ", "))
+    }
+    if (length(in_data_nin_phy) > 0) {
+      msg <- c(msg,
+               "\nin data but not phyloZ: ",
[…]

### 2021-06-07 pkg/tests/testthat/test-basic.R

@@ -3,9 +3,22 @@ library("testthat")
 library("phyloglmm")
 
 phyloZ <- phylo.to.Z(garamszegi_phy)
-datG <- transform(garamszegi_simple,
-                  obs = factor(seq(nrow(garamszegi_simple))),
-                  sp = phylo)
+
+augment_data <- function(x) {
+  transform(x,
+            obs = factor(seq(nrow(x))),
+            sp = factor(phylo)
+            )
+}
+
+simp_VC <- function(x) {
+  v <- lme4::VarCorr(x)
+  if (inherits(x, "glmmTMB")) v <- v$cond
+  unlist(lapply(v, c))
+}
+
+datG <- augment_data(garamszegi_simple)
+datP <- augment_data(garamszegi_pois)
 
 test_that("basic phylo.to.Z", {
   expect_s4_class(phyloZ, "Matrix")
@@ -23,4 +36,45 @@ test_that("basic lme4 LMM", {
                            , control = lme4::lmerControl(check.nobs.vs.nlev = "ignore", check.nobs.vs.nRE = "ignore")
                              )
   expect_equal(c(lme4::VarCorr(phylo_lmm_fit)$sp), 207.02517)
+})
+
+
+test_that("phylo_glmm", {
+  phylo_glmm_fit <- phylo_glmm(phen_pois~cofactor+(1|sp)+(1|obs)
+                             , data=datP
+                             , phylonm = "sp"
+                             , family = poisson
+                             , phylo = phylo
+                             , phyloZ=phyloZ
+                             , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore")
+                               )
+  phylo_glmmTMB_fit <- phylo_glmmTMB(phen_pois~cofactor+(1|sp)+(1|obs)
+                             , data=datP
+                             , phylonm = "sp"
+                             , family = poisson
+                             , phyloZ=phyloZ
+                               )
+## Warning message:
+## In condReStruc[i] <- `*vtmp*` :
+##   number of items to replace is not a multiple of replacement length
+
+  expect_equal(simp_VC(phylo_glmm_fit), simp_VC(phylo_glmmTMB_fit),
+               tolerance = 1e-3)
+})
+
+sf <- function(x) system.file("test_data", x, package="phyloglmm", mustWork=TRUE)
+
+dd <- read.table(sf("GH6_data.txt"), header=TRUE)
[…]

### 2021-06-24 response.md
minor changes in the revision and response

@@
 **BMB: I think if we have a non-problematic example we should include it**
+**MLI: I think we cleaned up garamszegi's example for the package already. DOUBLE CHECK**
 
@@
 
-**MLi: Bayesian methods are the state of the art for more complex. I've now reworded the last parameter in the intro before methods.**
+**MLi: Bayesian methods are the state of the art for more complex. I've now reworded the last paragraph in the intro before methods.**
 
@@
 **Skip?**
+**garamszegi's example is real**
 

### 2021-06-24 phyloglmm_ms_resubmit.tex
minor changes in the revision and response

@@
 
-\bmb{change PCM to PR (phylogenetic regression) throughout to narrow the scope? I made this change in the first para, would need to replace throughout if we think it's a good idea}
 Phylogenetic regressions (PR) account for evolutionary relatedness when analyzing relationships among characteristics of species.
@@
 
-In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PCMs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
-While a wide range of tools is available for PCM, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets.
+In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PRs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
+While a wide range of tools is available for PR, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets.
 When faced with these constraints, researchers often simplify their analyses, at the risk of neglecting important processes.
@@
 
-In classic PCMs, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion (BM) evolutionary model \citep{felsenstein1985phylogenies}.
+In classic PRs, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion (BM) evolutionary model \citep{felsenstein1985phylogenies}.
 If the residuals are normally distributed and observed without additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
@@
 
-Classic PCMs allow the residuals to evolve along the phylogeny, but the effects of the predictor variables may evolve along the phylogeny as well; extending PCMs to allow for multiple sources of variation leads to the  \emph{phylogenetic mixed model} \citep{housworth2004phylogenetic}.
-For example, suppose we wish to fit a PCM to predict species' brain size from their body size \cite{felsenstein1985phylogenies} using a mixed-effect model.
-The standard PCM allows for phylogenetic correlations in the residuals of the relationship between body and brain size; the phylogenetic mixed model can allow to incorporate both
+Classic PRs allow the residuals to evolve along the phylogeny, but the effects of the predictor variables may evolve along the phylogeny as well; extending PRs to allow for multiple sources of variation leads to the  \emph{phylogenetic mixed model} \citep{housworth2004phylogenetic}.
+For example, suppose we wish to fit a PR to predict species' brain size from their body size \cite{felsenstein1985phylogenies} using a mixed-effect model.
+The standard PR allows for phylogenetic correlations in the residuals of the relationship between body and brain size; the phylogenetic mixed model can allow to incorporate both
 variation in brain size among taxa with similar body sizes (random intercepts) and
@@
 
-We compare our approach with five other R packages that can fit PCMs: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
-Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PCM models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
+We compare our approach with five other R packages that can fit PRs: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, \pkg{phyr} \citep{phyr} and \pkg{brms} \citep{burkner2018brms}.
+Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PR models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
 Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
@@
 
-The range of PCMs presented here can incorporate many levels and types of phylogenetic variation.
+The range of PRs presented here can incorporate many levels and types of phylogenetic variation.
 Of course, there are further biological complexities we have neglected, such as multivariate responses; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck model \citep{butler2004phylogenetic}); and variable-rate models, which allow evolutionary rates to vary across the phylogeny.

### 2021-07-15 response.md
pushing new changes in the response and added glmmTMB citation

@@
 
-**We intended the paper as an application/software paper, not as a research paper: there is some novelty here, but as our primary innovation is to recast the model in a mathematically equivalent but more convenient/efficient form, we would prefer to treat this as an application.**
-
-**BMB: I think if we have a non-problematic example we should include it**
-**MLI: I think we cleaned up garamszegi's example for the package already. DOUBLE CHECK**
+**We intended the paper as an application/software paper, not as a research paper: there is some novelty here, but as our primary innovation is to recast the model in a mathematically equivalent but more convenient/efficient form, we would prefer to treat this as an application.
+We have provided an example in the vignette.**
 
@@
 
-** MLi: This line is now removed. We have confirmed that the suggested example is relevant to our framework in partitioning different levels of variation. **
+** This line is now removed. We have confirmed that the suggested example is relevant to our framework in partitioning different levels of variation and cited appropriately. **
 
@@
 
-**MLi: Bayesian methods are the state of the art for more complex. I've now reworded the last paragraph in the intro before methods.**
+**Bayesian methods are the state of the art for more complex. I've now reworded the last paragraph in the intro before methods.**
 
@@
 
-**MLi: This refers to the recent developed frequentist tools for phyloglmm and used in community ecology applications. We have now reword and clarify this sentence. **
+**This refers to the recent developed frequentist tools for phyloglmm and used in community ecology applications. We have now reword and clarify this sentence. **
 
@@
 
-**To BMB: Help? lme4 vs glmmTMB differences? Maybe convergence settings? Should we bother if they are close enough?**
 **MLi: The parameters that are in both single and multiple site models remain the same.
@@
 The goal of this paper is to show the inflexiably of existing tools and the efficiency of the new implemented approach.
-**
+The implementation of glmmTMB designed to maximize speed and flexibility.
+All else equal, the computational performance of glmmTMB are expected to outperform lme4.
+We added a text in the results that cites this expectiation.**
 
@@
 
-**Todo**
+**MLi: We have added an example in the vignette for this package**
 
@@
 
-**MLi: we have now added a tutorial**
+**MLi: We have added an example in the vignette for this package**
 
@@
 
-**Going to be a little annoying looking for an example with replicates.**
-**Skip?**
-**garamszegi's example is real**
+**MLi: Thanks for the suggestion.
+This paper is written to be an application paper rather than a research paper.
+We have provided a real example from Garamszegi's ``Modern phylogenetic comparative methods and their application in evolutionary biology: concepts and practice'' in the repository.**
 
@@
 
-**TO BMB: What do they mean by class?**
+**Thanks. We have now renamed everything under phylogenetic regression**
 
@@
 
-**Todo: Thanks, these are wonderful and helpful suggestion. We have included a tutorial in the released version on zenodo/github. We have now included the link in the paper.... **
+**Thanks, these are wonderful and helpful suggestion. We have included a tutorial example in the released version on zenodo/github. We have now included the link in the paper.... **
 

### 2021-07-15 phyloglmm_ms_resubmit.tex
pushing new changes in the response and added glmmTMB citation

@@
 Models capable of fitting all parameters (\pkg{lme4} and \pkg{glmmTMB}) fit well for all parameters.
-\pkg{pez} and \pkg{phyr} estimates for n = 100 and 500 are missing because the models did not converge within 30 minutes.
+\pkg{pez} and \pkg{phyr} estimates for n = 500 are missing because the models did not converge within 30 minutes.
   }
@@
 Although the parameter estimates are similar across platforms, computational efficiency varies enormously across platforms and sample size.
-Comparing our two implementations, \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
+Comparing our two implementations, \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}; \pkg{glmmTMB} is expected to out perform \pkg{lme4} as it is designed to maximize flexibility and speed \citep{brooks2017glmmTMB}.
 \pkg{glmmTMB} is also faster than \pkg{pez} and \pkg{phyr}: the median time for \pkg{glmmTMB} to fit a 50-species model is $\approx$ 9, versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr}.

### 2021-09-22 pkg/R/data.R
major package cleanup

@@ -1,0 +1,19 @@
+#' data
+#'
+#' #' Simple data from Garamszegi chapter 11
+"garamszegi_phy"
+
+#' @rdname garamszegi_phy
+"garamszegi_pois_missing"
+
+#' @rdname garamszegi_phy
+"garamszegi_repeat"
+
+#' @rdname garamszegi_phy
+"garamszegi_simple"
+
+#' @rdname garamszegi_phy
+"garamszegi_pois"
+
+#' @rdname garamszegi_phy
+"garamszegi_effect"

### 2021-09-22 pkg/R/glmmTMB_phylo_setup.R

@@ -1,0 +1,356 @@
+#' hacked version of glmmTMB
+## @params copied from glmmTMB as well ...
+##' @inheritParams phylo_lmm
+##' @param formula combined fixed and random effects formula, following lme4 syntax.
+##' @param data optional data frame containing model variables.
+##' @param family a family function, a character string naming a family function, or the result of a call to a family function (variance/link function) information. See \code{\link{family}} for a generic discussion of families or \code{\link{family_glmmTMB}} for details of \code{glmmTMB}-specific families.
+##' @param ziformula a \emph{one-sided} (i.e., no response variable) formula for zero-inflation combining fixed and random effects: the default \code{~0} specifies no zero-inflation. Specifying \code{~.} sets the zero-inflation formula identical to the right-hand side of \code{formula} (i.e., the conditional effects formula); terms can also be added or subtracted. \strong{When using \code{~.} as the zero-inflation formula in models where the conditional effects formula contains an offset term, the offset term will automatically be dropped}. The zero-inflation model uses a logit link.
+##' @param dispformula a \emph{one-sided} formula for dispersion containing only fixed effects: the default \code{~1} specifies the standard dispersion given any family. The argument is ignored for families that do not have a dispersion parameter. For an explanation of the dispersion parameter for each family, see \code{\link{sigma}}. The dispersion model uses a log link. In Gaussian mixed models, \code{dispformula=~0} fixes the residual variance to be 0 (actually a small non-zero value), forcing variance into the random effects. The precise value can be controlled via \code{control=glmmTMBControl(zero_dispval=...)}; the default value is \code{sqrt(.Machine$double.eps)}.
+##' @param weights weights, as in \code{glm}. Not automatically scaled to have sum 1.
+##' @param offset offset for conditional model (only).
+##' @param contrasts an optional list, e.g., \code{list(fac1="contr.sum")}. See the \code{contrasts.arg} of \code{\link{model.matrix.default}}.
[…]

### 2021-09-22 pkg/R/lme4_phylo_setup.R

@@ -1,0 +1,447 @@
+##' phylogenetic linear mixed model
+##' @importFrom lme4 optimizeLmer optimizeGlmer mkMerMod mkLmerDevfun mkGlmerDevfun
+##' @importFrom lme4 findbars nobars lmerControl glmerControl updateGlmerDevfun subbars
+##' @param formula mixed-model formula
+##' @param data data frame
+##' @param phylo phylogenetic tree in \code{\link{phylo}} format
+##' @param phylonm name of phylogenetic grouping variable
+##' @param phyloZ phylogenetic Z-matrix (see \code{\link{phylo.to.Z}}): optional, will be computed
+##' internally from \code{phylo} if not specified here. (For large phylogenies that are going to
+##' be used in multiple models it may make sense to compute the Z matrix first and pass it to
+##' the modeling functions to avoid recomputing it each time.)
+##' to compute \code{phyloZ} fi
+##' @param control control
+##' @param REML use restricted max likelihood?
+## unimplemented ... pass via ... ?
+## @param subset subset expression
+## @param weights weights
+## @param na.action na.action
+## @param offset offset
+## @param contrasts contrasts
+##' @export
+phylo_lmm <- function(formula, data, phylo = NULL, phylonm = NULL, phyloZ = NULL, control, REML = FALSE) {
+  phyloZ <- get_phyloZ(phylo, phyloZ, data[[phylonm]])
+  lmod <- lFormula(formula = formula, data = data, control = control, REML = REML, phylonm = phylonm, phyloZ = phyloZ)
+  devfun <- do.call(mkLmerDevfun, lmod)
+  opt <- optimizeLmer(devfun, control = control$optCtrl)
+  mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
+}
+
+## DON'T document externally
+## hacked version of lme4::lFormula
+## @name lme4_utils
+## @rdname phylo_lmm
+## @param formula mixed-model formula
+## @param data data frame
+## @param REML use restricted max likelihood?
+## @param subset subset expression
+## @param weights weights
+## @param na.action na.action
[…]

### 2021-09-22 pkg/R/utils.R

@@ -1,0 +1,84 @@
+##' construct a Z matrix from a phylogeny
+##' @name phylo_machinery
+##' @param r a \code{phylo} object
+##' @param stand standardize edge lengths by determinant of phylogenetic covariance matrix
+##' @export
+##' @importFrom Matrix t diag
+##' @importFrom ape vcv
+## FIXME:: rename to phylo_to_Z ?
+phylo.to.Z <- function(r, stand = FALSE) {
+  ntip <- length(r$tip.label)
+  Zid <- Matrix::Matrix(0.0, ncol = length(r$edge.length), nrow = ntip)
+  nodes <- (ntip + 1):max(r$edge)
+  root <- nodes[!(nodes %in% r$edge[, 2])]
+  for (i in 1:ntip) {
+    cn <- i ## current node
+    while (cn != root) {
+      ce <- which(r$edge[, 2] == cn) ## find current edge
+      Zid[i, ce] <- 1 ## set Zid to 1
+      cn <- r$edge[ce, 1] ## find previous node
+    }
+  }
+  tZid <- t(Zid)
+  Z <- t(sqrt(r$edge.length) * tZid)
+  if (stand) {
+    V <- ape::vcv(r)
+    ## V <- V/max(V)
+    sig <- exp(as.numeric(determinant(V)["modulus"]) / ntip)
+    ## sig <- det(V)^(1/ntip)
+    ## all.equal(Z/sqrt(sig),
+    ##  t(sqrt(r$edge.length / sig) * tZid), tolerance=2e-16)
+    Z <- Z/sqrt(sig)
+  }
+  rownames(Z) <- r$tip.label
+  colnames(Z) <- 1:length(r$edge.length)
+  return(Z)
+}
+
+check_phylo_names <- function(phyloZ, data_sp) {
+  in_phy_nin_data <- setdiff(rownames(phyloZ), unique(data_sp))
+  in_data_nin_phy <- setdiff(unique(data_sp), rownames(phyloZ))
+  if (length(in_phy_nin_data) > 0 ||
+      length(in_data_nin_phy) > 0) {
+    msg <- "mismatch between tip names in phylogeny/Z-matrix and names in data:"
+    if (length(in_phy_nin_data) > 0) {
+      msg <- c(msg,
+               "\nin phyloZ but not data: ",
+               paste(in_phy_nin_data, collapse = ", "))
+    }
+    if (length(in_data_nin_phy) > 0) {
+      msg <- c(msg,
+               "\nin data but not phyloZ: ",
[…]

### 2021-09-22 pkg/misc/get_garamszegi.R

@@ -1,0 +1,16 @@
+tt <- tempdir()
+download.file("http://mpcm-evolution.com/OPM/Chapter11_OPM/data.zip",
+              dest=file.path(tt, "data.zip"))
+old.dir <- setwd(tt)
+untar("data.zip")
+garamszegi_phy <- ape::read.nexus("phylo.nex")
+ff <- list.files(pattern="data_.*\\.txt")
+for (f in ff) {
+  x <- read.table(f, header=TRUE)
+  n <- gsub("\\.txt$","",
+            gsub("data_","garamszegi_",f))
+  assign(n, x)
+}
+setwd(old.dir)
+save(list = ls(pattern="garamszegi"), file="garamszegi.rda")
+unlink(tt)

### 2021-09-22 pkg/tests/testthat.R

@@ -1,0 +1,4 @@
+library(testthat)
+library(phyloglmm)
+
+test_check("phyloglmm")

### 2021-09-22 pkg/tests/testthat/test-basic.R

@@ -1,0 +1,118 @@
+library("ape")
+library("testthat")
+library("phyloglmm")
+
+cc <- lme4::lmerControl(check.nobs.vs.nlev = "ignore", check.nobs.vs.nRE = "ignore")
+phyloZ0 <- phylo.to.Z(garamszegi_phy)
+
+augment_data <- function(x) {
+  transform(x,
+            obs = factor(seq(nrow(x))),
+            sp = factor(phylo)
+            )
+}
+
+simp_VC <- function(x) {
+  v <- lme4::VarCorr(x)
+  if (inherits(x, "glmmTMB")) v <- v$cond
+  unlist(lapply(v, c))
+}
+
+datG <- augment_data(garamszegi_simple)
+datP <- augment_data(garamszegi_pois)
+
+phyloZ <- phyloZ0[levels(datG$sp), ]
+
+test_that("basic phylo.to.Z", {
+  expect_s4_class(phyloZ, "Matrix")
+  expect_equal(nrow(phyloZ), ape::Ntip(garamszegi_phy))
+  expect_equal(ncol(phyloZ), ape::Ntip(garamszegi_phy) + ape::Nnode(garamszegi_phy) - 1)
+})
+
+phylo_lmm_fit <- phylo_lmm(phen~cofactor+(1|sp)
+                         , data=datG
+                         , phylonm = "sp"
+                         , phylo = garamszegi_phy
+                         , REML = TRUE
+                         , control = cc
+                           )
+
+test_that("basic lme4 LMM", {
+  expect_equal(c(lme4::VarCorr(phylo_lmm_fit)$sp), 207.02517)
+})
+
+test_that("fit with external phyloZ", {
+  phylo_lmm_fitZ <- phylo_lmm(phen~cofactor+(1|sp)
+                           , data=datG
+                           , phylonm = "sp"
+                           , phyloZ = phyloZ
+                           , REML = TRUE
+                           , control = cc
+                             )
+  ## too tedious to check identity with all these environments & calls
+  ## floating around ...
+  expect_identical(simp_VC(phylo_lmm_fit), simp_VC(phylo_lmm_fitZ))
+})
+
+test_that("detect phyloZ/sp mismatch", {
+  expect_error(phylo_lmm(phen~cofactor+(1|sp)
+          , data=datG
+          , phylonm = "sp"
+          , phyloZ = phyloZ0
+          , REML = TRUE
+          , control = cc
+            ),
+            "row names of phyloZ must match")
+})
+
+test_that("require sp to be a factor", {
[…]

### 2021-09-22 response.md
working on response

@@
 
-**To BMB: What do they mean by model testing here? This feels like snooping and cheating if we test and select models.**
+**Todo: (for response and possibly add something to discussion) Problem 1: effects of model selection on coverage etc.
+(i.e.
+snooping), maybe not a problem if the aspects of the model being selected are not focal (e.g.
+Matuschek et al.); problem 2: understanding what levels of complexity are computationally practical**
 
@@
 This paper is written to be an application paper rather than a research paper.
-We have provided a real example from Garamszegi's ``Modern phylogenetic comparative methods and their application in evolutionary biology: concepts and practice'' in the repository.**
+We have provided a real example from Garamszegi's ``Modern phylogenetic comparative methods and their application in evolutionary biology: concepts and practice'' in the repository (we could add this if the editor thought it was appropriate, but we're up against the length limits for an application paper and would prefer not to open the can of worms of making this into a research paper ...**
 
@@
 
-**Todo: Table 1 shows all the possbilities (double check) with existing package.** 
-**Todo: Add it to the tutorial?**
+**Table 1 identifies several broadly used packages that are restricted to balanced designs.
+(We are not discussing the performance (e.g.
+power or efficiency) of different methods applied to unbalanced designs, simply the fact that some software packages for PCM cannot handle unbalanced designs.)**
 
@@
 
-**To BMB: Why is it slower? Do we need to profile the pieces?**
+**Todo: Why is it slower? Do we need to profile the pieces? Waffle!**
 
@@
 **Todo: what?
-go back to the paper and see if we say this clearly, this should be trival and clear.
+go back to the paper and see if we say this clearly, this should be trivial and clear.
 The formulas are clear!

### 2021-09-24 response.md
working on response

@@
 
-**MLi: The model is the mathematically equivalent, thus, the difference is the speed/efficiency.**
-**MLi: Comparing pez and phyr uses lme4/glmer correlation structure. This is the most direct comparision of the proposed method (i.e. correlation vs random effect matrix structure. **
+**MLi: The model is the mathematically equivalent, thus, the difference is the speed/efficiency.
+The most direct comparison is between `phyloglmm` and the `phyr` or `pez` packages: all three methods are built on the mixed-models machinery of `lme4`, but `phyloglmm` modifies the Z matrix while `phyr` and `pez` modify the correlation structure of the random effects**.
+(TODO/MLi: is my restatement correct?)
 
@@
 There are many research papers applying PCMs to large datasets (and as a side note: how large should a dataset be to be considered large?).
+
+**We have switched our focal set of models from "phylogenetic comparative methods" (PCM) to the narrower scope of "phylogenetic regression" (PR).
+The reviewer is correct that PCM is too broad a scope for describing our goals here.
+We demonstrate in our simulations that the fastest other methods in our set of comparisons take >30 minutes for a
 
@@
 
-**Bayesian methods are the state of the art for more complex. I've now reworded the last paragraph in the intro before methods.**
+**Bayesian methods are the state of the art for more complex models. I've now reworded the last paragraph in the intro before methods.**
 

### 2021-09-24 phyloglmm_ms_resubmit.tex
working on response

@@
 In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PRs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
-While a wide range of tools is available for PR, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets.
+While a wide range of tools is available for PR, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets (e.g.
+random-slopes phylogenetic regressions with hundreds of species).
 When faced with these constraints, researchers often simplify their analyses, at the risk of neglecting important processes.
@@
 (\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
-\Sigma_{\mathrm{phy_{int-slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
+\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int-slope}}} \\ 
+\sigma_{\mathrm{phy_{int-slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
 \end{bmatrix}
@@
 \bf{b_{\mathrm{group}}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
-\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \Sigma^2_{phy}) \\
+\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \sigma^2_{phy}) \\
 \besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
@@
 Neither PGLS nor PGLM can handle random slopes or multiple observations within a species.
-Among the few packages that currently fit phylogenetic slopes to predictor level variation are \pkg{pez} and \pkg{phyr}, which can handle random-slope models ($\Sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but do not incorporate covariation between phylogenetic random slopes and intercepts ($\Sigma_{\mathrm{phy_{int-slope}}}$).
+Among the few packages that currently fit phylogenetic slopes to predictor level variation are \pkg{pez} and \pkg{phyr}, which can handle random-slope models ($\sigma^2_{\mathrm{phy_{slope}}}$) and random intercepts of species-group interactions ($\bf{b_{\mathrm{sp:group}}}$) but do not incorporate covariation between phylogenetic random slopes and intercepts ($\sigma_{\mathrm{phy_{int-slope}}}$).
 Lastly, Bayesian PGLMMs using Markov chain Monte Carlo (MCMC) can handle all of the cases described above. 
@@
   \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
-\bmb{we need a better colour for MCMCglmm.  colorspace package? }
     Total simulations $N=100$ for each category.

### 2021-10-01 response.md
lots of work on figs

@@
 
-**We appreciate the opportunity to resubmit a revised version.**
+**We appreciate the opportunity to resubmit a revised version. We recognize that due to the extremely long delay, this version is likely to be treated as a new submission.**
 
@@
 
-**We intended the paper as an application/software paper, not as a research paper: there is some novelty here, but as our primary innovation is to recast the model in a mathematically equivalent but more convenient/efficient form, we would prefer to treat this as an application.
-We have provided an example in the vignette.**
+**We intended the paper as an application/software paper, not as a research paper: there is some novelty here, but as our primary innovation is to recast existing models in a mathematically equivalent but more convenient/efficient form, we would prefer to treat this as an application.
+We have provided an example application to empirical data in a package vignette.**
 
@@
 
-**MLi: The model is the mathematically equivalent, thus, the difference is the speed/efficiency.
+**The model is the mathematically equivalent; thus, the difference is the speed/efficiency.
 The most direct comparison is between `phyloglmm` and the `phyr` or `pez` packages: all three methods are built on the mixed-models machinery of `lme4`, but `phyloglmm` modifies the Z matrix while `phyr` and `pez` modify the correlation structure of the random effects**.
@@
 **We have switched our focal set of models from "phylogenetic comparative methods" (PCM) to the narrower scope of "phylogenetic regression" (PR).
-The reviewer is correct that PCM is too broad a scope for describing our goals here.
-We demonstrate in our simulations that the fastest other methods in our set of comparisons take >30 minutes for a
-
-**Todo: Revisit and rewrite.
-It is true many research papers used PCMs to large datasets, but it depends on the complexity of the model.
-I.e.
-If we simply use GLM methods in Table 1, it is relative fast even with large datasets.
-If we have more complicated models, then it will be very computationally demanding.**
+The reviewer is correct that PCM is too broad a scope this paper.
+We believe the results in our paper back up the other claims in the introduction: at present, researchers have a choice between slow, flexible methods (mostly Bayesian) and fast, inflexible (mostly frequentist) methods.**
 
@@
 
-** This line is now removed. We have confirmed that the suggested example is relevant to our framework in partitioning different levels of variation and cited appropriately. **
+**We have removed this statement. We agree that the suggested example is relevant to our discussion and have cited it appropriately.**
 
@@
 
-**This refers to the recent developed frequentist tools for phyloglmm and used in community ecology applications. We have now reword and clarify this sentence. **
+**This refers to the recent developed frequentist tools for phyloglmm and used in community ecology applications. We have now reword and clarify this sentence.**
 
@@
 
-**MLi: The parameters that are in both single and multiple site models remain the same.
-We have now added one more alternative implementation run for the 100 species simulation with the slow platforms.
-The goal of this paper is to show the inflexiably of existing tools and the efficiency of the new implemented approach.
-The implementation of glmmTMB designed to maximize speed and flexibility.
-All else equal, the computational performance of glmmTMB are expected to outperform lme4.
-We added a text in the results that cites this expectiation.**
+**In complex models, different implementations of the same model/algorithm will often give slightly different results.
+Indeed, the estimates from pez and phyr (which implement the same statistical model) also differ from each other.
+Since lme4 and glmmTMB use very different underlying estimation algorithms, it is not surprising that the phyloglmm implementations based on these two platforms converge to slightly different parameter estimates.**
 
@@
 
-**Todo: (for response and possibly add something to discussion) Problem 1: effects of model selection on coverage etc.
-(i.e.
-snooping), maybe not a problem if the aspects of the model being selected are not focal (e.g.
-Matuschek et al.); problem 2: understanding what levels of complexity are computationally practical**
+**We have expanded on this discussion to clarify our point.**
 
@@
 
-**Done. We added more citation on this topic.**
+**Done. We have added more citations on this topic.**
 
@@
 
-**Todo: Why is it slower? Do we need to profile the pieces? Waffle!**
-
 1.
@@
 
-**MLi: Thanks.
-We have now added MCMCglmm to our simulation.
-For our simple simulation test, MCMCglmm is turned out to be faster based on your convergence critrion which heavily depend on the number of iteration.
-However, the goal is to look at the granlarity of frequentist approaches.
+**We have now added MCMCglmm to our simulation.
+For our simple simulation test, MCMCglmm turned out to be faster based on your convergence criterion which heavily depend on the number of iteration.
+However, the goal is to look at the granularity of frequentist approaches.
 Even for simple linear models, Bayesian approaches are much slowly due to MCMC compared to least squares.**

### 2021-10-01 phyloglmm_ms_resubmit.tex
lots of work on figs

@@
 
+\newcommand{\plmefour}{phyloglmm/lme4}
+\newcommand{\pgTMB}{phyloglmm/glmmTMB}
+
 % \pagenumbering{gobble}
@@
 \\
-\textbf{3} Institute for Infectious Diseases Research, McMaster University, Hamilton, Ontario, Canada
+\textbf{3} Institute for Infectious Disease Research, McMaster University, Hamilton, Ontario, Canada
 \\
@@
 
-Phylogenetic regressions (PR) account for evolutionary relatedness when analyzing relationships among characteristics of species.
+Phylogenetic regressions (PR), a subset of phylogenetic comparative methods, account for evolutionary relatedness when analyzing relationships among morphological, physiological, or ecological characteristics of species.
 Given a known phylogeny, PRs model the relationships among species traits while incorporating relatedness; they can be used to control statistically for phylogeny, to quantify phylogenetic signal in traits, or both.
 
-In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PRs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
-While a wide range of tools is available for PR, existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets (e.g.
+In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PRs incorporate phylogenetic relationships to account for the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
+While a wide range of tools is available for PR, most existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets (e.g.
 random-slopes phylogenetic regressions with hundreds of species).
@@
 \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation \\per species, \\ Balanced design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4}
-                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}, \pkg{phyr}                                                                                  \\ \hline
+                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{\plmefour}, \pkg{\pgTMB}, \pkg{phyr}                                                                                  \\ \hline
 \multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4}
@@
 \hline
- & \pkg{nlme} & \pkg{phylolm} & \pkg{lme4}/\pkg{glmmTMB} & \pkg{pez} & \pkg{phyr} & \pkg{brms} & \pkg{MCMCglmm} \\ \hline
+& \pkg{nlme} & \pkg{phylolm} & \pkg{phyloglmm} & \pkg{pez} & \pkg{phyr} & \pkg{brms} & \pkg{MCMCglmm} \\
+&            &               & (this paper)    &           &            &            &                \\  \hline
 Single Group & \y & \y & \y &  &  & \y & \y \\ \hline
@@
 }
-\caption{List of estimable parameters for each R package.}
+\caption{List of estimable models for each R package.}
 \label{table:platform}
@@
 The horizontal line shows the true value of the parameters in the simulation model.
-Models capable of fitting all parameters (\pkg{lme4},  \pkg{glmmTMB}, and \pkg{brms}) fit well for all parameters.
+Models capable of fitting all parameters (\pkg{\plmefour},  \pkg{\pgTMB}, and \pkg{brms}) fit well for all parameters.
 %\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
@@
   \caption{Comparison of coverage probability for fixed effect parameters.
-Models matching the simulation model (\pkg{lme4}, \pkg{glmmTMB} and pkg{brms}) have coverage near the nominal value of 0.95.
-The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits.
+Models matching the simulation model (\pkg{\plmefour}, \pkg{\pgTMB} and \pkg{brms}) have coverage near the nominal value of 0.95.
+The black line shows the nominal coverage; the blue ribbon shows the 95\% binomial confidence interval around the nominal coverage based on fits to 100 simulated data sets.
 %  \bmb{really picky, but maybe change grid lines or ribbon colour slightly so they're not confusable (maybe use blue ribbon with alpha=0.1?)}
@@
 The horizontal line shows the true value of the parameters in the simulation model.
-Models capable of fitting all parameters (\pkg{lme4} and \pkg{glmmTMB}) fit well for all parameters.
+Models capable of fitting all parameters (\pkg{\plmefour} and \pkg{\pgTMB}) fit well for all parameters.
 \pkg{pez} and \pkg{phyr} estimates for n = 500 are missing because the models did not converge within 30 minutes.
@@
 However, real applications are always data-constrained.
-Establishing the practical level of model complexity for a given problem and data set is an open and difficult general problem throughout statistical modeling, not just in phylogenetic studies.
+Establishing the practical level of model complexity for a given problem and data set is an open and difficult general problem throughout statistical modeling,
+not just in phylogenetic studies.
+While information-theoretic or stepwise selection methods are often used to decide the complexity of models in ecology and evolution \cite{matuschek2017balancing,darriba2020modeltest}, sufficiently complex models may simply be too complex to fit without some form of regularization
+\cite{uriarte2009preaching}, and data-driven model selection may affect inference in unexpected and unwanted ways \cite{hurvich1990impact,morin2020is}.
+
 %For example, should one use simple models that may be overly conservative or

### 2021-10-01 plot.R

@@ -8,6 +8,8 @@ library(colorspace)
 library(directlabels)
 library(ggrepel)
 
+plot_width <- 7
+
 theme_set(theme_bw())
 zmargin <- theme(panel.spacing=grid::unit(0,"lines"))
 
@@ -33,24 +35,34 @@ sspar_df <- data.frame(
   , y_int = c(sd.resid^2, physd.B0^2, physd.B1^2, phyrho.B01*physd.B0*physd.B1)
 )
 
+trans_platform <- function(platform) {
+  ifelse(platform %in% c("lme4","glmmTMB"),
+         sprintf("phyloglmm/\n%s", platform),
+         platform)
+}
+
+
 ssdat <- (ssdat_raw
-	%>% separate(model,c("platform", "sites", "size", "seed","saveformat"), "[.]")
-	%>% dplyr:::select(time, platform, size, seed, resid, phylo_int, phylo_X, phylo_cor, B0, B1)
-	%>% gather(key = sdtype, value = sd, -c(platform, size, seed, time, B0, B1))
-	%>% left_join(.,sspar_df, by="sdtype")
-	%>% mutate(size = factor(size
-	      , levels=c("small","med","large","xlarge"), labels=c("25","50","100","500")
-			)
-			, Platform = factor(platform, levels=sub_pkgs)
-			, sdtype = factor(sdtype, levels=c("phylo_int","phylo_cor","phylo_X","resid")
-			                  , labels=c(expression(paste("Phylogenetic random intercept ", Sigma[phy[int]]))
-			                             , expression(paste("Phylogenetic random intercept-slope correlation ", rho[phy[int-slope]]))
-			                             , expression(paste("Phylogenetic random slope ", Sigma[phy[slope]]))
-			                             , expression(paste("Residual ", sigma[epsilon])))
-		)
-	)
+  %>% as_tibble()
+  %>% separate(model,c("platform", "sites", "size", "seed","saveformat"), "[.]")
+  %>% dplyr:::select(time, platform, size, seed, resid, phylo_int, phylo_X, phylo_cor, B0, B1)
+  %>% gather(key = sdtype, value = sd, -c(platform, size, seed, time, B0, B1))
+  %>% left_join(.,sspar_df, by="sdtype")
+  %>% mutate(size = factor(size
+                         , levels=c("small","med","large","xlarge"), labels=c("25","50","100","500")
+                           )
+           , Platform = factor(trans_platform(platform), levels=trans_platform(sub_pkgs))
[…]

### 2022-08-02 plot.R
fixing up plots (not quite done)

@@ -100,10 +100,11 @@ get_pos <- function(dat,
                     bwid = 0.02,
                     discrete = TRUE) {
   sspos <- dat %>%
-    group_by(Platform,size) %>%
-    summarise(time=median(time), .groups="drop_last") %>%
-    filter(time==max(time)) %>%
-    ungroup() %>%
+      group_by(Platform,size) %>%
+      summarise(time=median(time), .groups="drop_last") %>%
+      filter(time==max(time)) %>%
+      slice(n()) %>%  ## take only last row, in case there are ties on max(time)
+      ungroup() %>%
     mutate(nplatform=drop(scale(as.numeric(Platform), scale=FALSE)))
   if (discrete) {
     ## x-position calculation: 'size' is going to be the size
@@ -139,14 +140,16 @@ gg_sstime0 <- (ggplot(data=ssdat, aes(x=size, y=time, col=Platform,
                  size=3, alpha=0.1)
 )
 
+scaleposx <- c(80, 150)
+scaleposy <- 0.5*c(scaleposx[1]/200, scaleposx[2]^2/100*0.85)
 boxwid <- 0.125
 gg_cstime0 <- (ggplot(data=ssdat, aes(x=as_num(size),
                                       group=interaction(size,Platform),
                                       y=time, col=Platform,
                                       fill=Platform))
   + geom_function(fun=function(x) x/200, linetype=2, col="darkgray")
-  + annotate(x=c(100, 200),
-             y= 0.5*c(100/200, 200^2/100*0.85),
+  + annotate(x=scaleposx,
+             y= scaleposy,
              colour = "black",
              label=c('"time" %prop% S', '"time" %prop% S^2'),
              parse=TRUE,
@@ -165,18 +168,18 @@ gg_cstime0 <- (ggplot(data=ssdat, aes(x=as_num(size),
                  position=position_dodge(width = boxwid),
                  size=3, alpha=0.1)
 )
-csspos <- get_pos(ssdat, discrete=FALSE, y_tweak=c(glmmTMB=2.5, lme4=0.6), bwid=0.03)
-
+csspos <- get_pos(ssdat, discrete=FALSE, y_tweak=c(glmmTMB=2.5, lme4=0.6, gls = 2), bwid=0.03)
 
 gg_csstime <- (gg_cstime0
   + theme(legend.position="none")
-  + expand_limits(x=150)
-  + geom_text(data=csspos,
-              size = 3,
[…]

### 2022-08-03 response.md
tweak responses

@@
 
-**The model is the mathematically equivalent; thus, the difference is the speed/efficiency.
+**For models that are sufficiently flexible to fit a model structure of given complexity (e.g.
+brms/pez), all of the models are mathematically equivalent; the difference is in the precise formulation and implementation, leading to the observed significant differences in speed.
 The most direct comparison is between `phyloglmm` and the `phyr` or `pez` packages: all three methods are built on the mixed-models machinery of `lme4`, but `phyloglmm` modifies the Z matrix while `phyr` and `pez` modify the correlation structure of the random effects**.
-(TODO/MLi: is my restatement correct?)
 
@@
 
-**Done. We have added more citations on this topic.**
-
-I hope you will find these and the reviewers’ comments useful.
-Best wishes,
-
-Daniele Silvestro
+**We have added more citations on this topic.**
 
@@
 
-**MLi: Thanks.**
+**Thank you.**
 
@@
 
-**MLi: We have added an example in the vignette for this package**
+**We have added an example in the vignette for this package.**
 
@@
 
-**MLi: Thanks for the suggestion.
+**Thanks you for the suggestion.
 This paper is written to be an application paper rather than a research paper.
@@
 
-**Thanks. We have now renamed everything under phylogenetic regression**
+**We have now renamed everything under phylogenetic regression**
 
@@
 
-**Thanks, these are wonderful and helpful suggestion. We have included a tutorial example in the released version on zenodo/github. We have now included the link in the paper.... **
+**These are very useful suggestions. We have included a tutorial example in the released version on zenodo/github; the link is included n the paper.**
 
@@
 
-**MLi: Fixed.**
+**Fixed.**
 
@@
 
-**MLi: Exactly, we are avoiding the matrix inversion step.**
+**Exactly, we are avoiding the matrix inversion step.**
 
@@
 
-**We have now added MCMCglmm to our simulation.
-For our simple simulation test, MCMCglmm turned out to be faster based on your convergence criterion which heavily depend on the number of iteration.
-However, the goal is to look at the granularity of frequentist approaches.
-Even for simple linear models, Bayesian approaches are much slowly due to MCMC compared to least squares.**
+**We have now added MCMCglmm to our set of simulation.
+For our simple simulation test, MCMCglmm was generally faster than brms, although both Bayesian approaches (MCMCglmm/brms) were still significantly slower than frequentist methods, and scaled slightly worse.**
 
@@
 
-**Todo: what?
-go back to the paper and see if we say this clearly, this should be trivial and clear.
-The formulas are clear!
-We are using the variance of the phylogenetic tree when simulating the b's.**
-
-+++++
-RESUBMISSION INSTRUCTIONS
-Please note that resubmitting your manuscript does not guarantee eventual acceptance, and that your resubmission may be subject to re-review before a decision is rendered.
-Please also ensure that your altered manuscript still conforms to our word limit of 3000 for applications.
-
-Once you have made the suggested changes, go to https://mc.manuscriptcentral.com/mee-besjournals and login to your Author Centre.
-Click on "Manuscripts with Decisions," and then click on "Create a Resubmission" located next to the manuscript number.
-Then, follow the steps for resubmitting your manuscript.
-
-Because we are trying to facilitate timely publication of manuscripts submitted to Methods in Ecology and Evolution, your new manuscript should be uploaded within 12 weeks.
-The deadline for your resubmission is 05-Aug-2020.
-If it is not possible for you to submit your manuscript by that date, please get in touch with the editorial office, otherwise we will consider your paper as a completely new submission.
+**FIXME: make sure we state that the simulations follow the estimation model exactly, so that all the formulas for simulation are the same as those for estimation (and maybe emphasize that the Z/Sigma setup generalizes standardized phylogenetic regression to all of the more complex cases).**

### 2022-08-08 phyloglmm_ms_resubmit.tex
fixing up plots (not quite done)
pushing minor edits in ms and updated todo

@@
 
-Compared to the single--group model, the multi-group simulation model has five additional random effect parameters (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$))
+Compared to the single--group model, the multi-group simulation model has five additional random effect parameters (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$)).
 Because each species has multiple observations we can distinguish variation among species from residual variation, and thus we can include predictor-level and intercept-level random effects of species.
@@
 \begin{figure}[H]
-  \includegraphics[scale=0.7]{./figure/sstime.pdf}
+  \includegraphics[scale=0.7]{./figure/csstime.pdf}
   \caption{Comparison of single-group model computational speed.}
@@
 \end{center}
+
 \begin{center}

### 2022-08-09 abstract.tex
way too much figure tweaking
small ms edits

@@
 \begin{enumerate}
-\item{Phylogenetic comparative methods (PCM) using phylogenetic regression are a powerful technique to explore relationships among related groups of species traits.
+\item{Phylogenetic regression is a powerful technique for exploring relationships among characteristics of related species.
 However, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.}
@@
 We have implemented this formulation in two R statistical packages (\pkg{lme4} and \pkg{glmmTMB}).}
-\item{Our reformulation of phylogenetic generalized linear mixed models is computationally efficient, operating orders of magnitude faster than existing methods for fitting phylogenetic mixed models.}
+\item{Our reformulation of phylogenetic generalized linear mixed models is computationally efficient, operating orders of magnitude faster than existing comparably flexible methods.}
 \item{Our approach can be implemented in any platform for generalized mixed models.

### 2022-08-09 phyloglmm_ms_resubmit.tex
small ms edits

@@
 \vspace{2pt}
-* Corresponding author: Michael Li; lim88@mcmaster.ca
+* Corresponding author: Michael Li; lim88@mcmaster.ca. Current address: Public Health Risk Science Division, Public Health Agency of Canada 
 \end{center}
@@
 
-In contrast to standard statistical models, where all of the predictor variables (such as species traits and environmental covariates) are directly observable, PRs incorporate phylogenetic relationships to account for the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
-While a wide range of tools is available for PR, most existing procedures are either insufficiently flexible or too computationally demanding to analyze large data sets (e.g.
+In contrast to standard statistical models, where all observations are assumed to be independent, PRs incorporate phylogenetic relationships to account for correlations between observations driven by the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}.
+While a wide range of tools is available for PR, most existing procedures are either inflexible or too computationally demanding to analyze large data sets (e.g.
 random-slopes phylogenetic regressions with hundreds of species).
@@
 If the residuals are normally distributed and observed without additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
-More recent approaches --- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than BM.
+More recent approaches --- including phylogenetic generalized linear mixed models  \citep[PGLMM:][]{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than BM.
 These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.

### 2022-08-09 phyloglmm_ms_resubmit.tex
cleanup, tweaks

@@
 \newcommand{\symsub}[2]{\ensuremath{#1_{\tiny \mathrm{#2}}}\xspace}
-\newcommand{\mrm}[1]{\ensuremath{\mathrm{#1}}
-\xspace}
+\newcommand{\mrm}[1]{\ensuremath{\mathrm{#1}}\xspace}
 
@@
 \newcommand{\pgTMB}{phyloglmm/glmmTMB}
-
+\newcommand{\phyvar}{\sigma} %% phylogenetic var: allow sigma/Sigma switch
+\newcommand{\phyv}[1]{\ensuremath{\phyvar^2_{\mathrm{phy}_\mathrm{#1}}}}
+\newcommand{\phys}[1]{\ensuremath{\phyvar_{\mathrm{phy}_\mathrm{#1}}}}
+\newcommand{\bphys}[1]{\ensuremath{{\bf b}_{\mathrm{phy}_{\mathrm{#1}}}}}
+\newcommand{\bsps}[1]{\ensuremath{{\bf b}_{\mathrm{sp}_{\mathrm{#1}}}}}
 % \pagenumbering{gobble}
@@
 More recent approaches --- including phylogenetic generalized linear mixed models  \citep[PGLMM:][]{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than BM.
-These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
-If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process error while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging, devillemereuil2012bayesian}.
+These methods partition residual variation into two components: (1) independent residual variation (tip variation, which may be confounded with observation error) and (2) phylogenetic signal  \citep[evolutionary process error:][]{hansen2012interpreting, housworth2004phylogenetic}.
+If each species' traits are measured multiple times, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation both contribute to the evolutionary variation while the observation error can be independently identified from within-species variation \citep{kostikova2016bridging, devillemereuil2012bayesian}.
 
@@
 
-Several recent studies developed new tools to fit phylogenetic mixed models in
+Several recent studies have developed new tools to fit phylogenetic mixed models in
 community ecology applications \citep{nowakowski2018phylogenetic, li2017canfun}.
 However, the phylogenetic mixed modeling tools that allow extensions such as random slopes
-and separation of tip and observation error in a frequentist framework are inflexible; thus, biologists needing to fit more complex models typically turn to Bayesian approaches which is the state of the art, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms, kostikova2016bridging} (Table \ref{table:model}).
+and separation of tip and observation error in a frequentist framework may not allow other extensions like non-normal response distributions or additional random effects; thus, biologists needing to fit more complex models typically turn to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms, kostikova2016bridging} (Table \ref{table:model}).
 
-\begin{table}[]
-\begin{tabular}{|l|l|l|l|}
-\hline
-\textbf{Model} & \textbf{Method} & \textbf{Data} & \textbf{Platform} \\ \hline
-\multirow{2}{*}{
-\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)
-\end{tabular}} & Correlated residual & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4}
-& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} & \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
-\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation \\per species, \\ Balanced design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4}
-                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{\plmefour}, \pkg{\pgTMB}, \pkg{phyr}                                                                                  \\ \hline
-\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4}
-                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
-\end{tabular}
-\caption{GLM-based phylogenetic comparative methods and their implementations in R packages.}
-\label{table:model}
-\end{table}
+\input table1.tex
 
@@
 The phylogenetic generalized linear mixed model (PGLMM) framework defines a wider range of models that includes the standard phylogenetic regression as a special case \citep{lynch1991methods}.
-The PGLMM allows for non-Gaussian responses and uses incorporates multiple components of variability.
+The PGLMM allows for non-Gaussian responses and incorporates multiple components of variability.
 The PGLMM has the form:
@@
 \begin{aligned}
-\bY & = \dist(g^{-1}(\bmu), \phi) \\
-\bmu & = \bX \bbeta + \bZ \bb  \\
+\bY & \sim \dist(\bmu, \phi) \\
+\bmu & = g^{-1}(\bX \bbeta + \bZ \bb)  \\
 \bb & \sim \textrm{MVN}(0, \bSigma(\theta))  \\
@@
 where in addition to the terms from (\ref{eq:gls}) $\bZ$ is an $n \times q$ model matrix for the $q$-dimensional vector-valued random effects; $\bb$ represents the conditional mean (or mode) of the random effect, which is multivariate normally distributed with covariance matrix $\bSigma(\theta)$; and $\phi$ is a scale parameter for the conditional distribution $\dist$.
-When $\dist$ is Gaussian, $g$ is the identity function, $\bZ$ is the identity matrix, and $\bSigma(\theta)=\sigma^2\bC$, (\ref{eq:glmm}) reduces to (\ref{eq:gls}).
+The PGLMM reduces to the simple PR model (\ref{eq:gls}) when
+$\dist$ is Gaussian, $g$ is the identity function, $\bZ$ is the identity matrix,
+$\bSigma(\theta)=\sigma^2\bC$, and $\dist$ is Gaussian.
+In addition, we need $\phi=\sigma_r^2 = 0$ so that the residual variance disappears and the only variance comes from the phylogenetic covariance matrix; otherwise, the additional residual variance term corresponds to one implementation of Pagel's $\lambda$.
 
@@
 thus, we can use $\bZ$ to incorporate phylogenetic correlations.
-Suppose evolution follows a BM process, i.e., continuous traits evolve independently, following an unbiased random walk along each branch of the phylogeny.
+Suppose evolution follows a BM process, i.e., continuous traits evolve independently at a constant rate, following an unbiased random walk along each branch of the phylogeny.
 Then the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
@@
 \end{center}
-\[
-  \begin{blockarray}{ccccc}
-  & L_1 & L_2 & L_3 & L_4  \\
-  \begin{block}{c(cccc)}
-  t_1 & \ell_1 & \ell_2 & 0           & 0 \\
-  t_2 & \ell_1 &  0          & \ell_3 & 0 \\
-  t_3 & 0           &  0          & 0           & \ell_4 \\
-  \end{block}
-  \end{blockarray}
-  \hspace{0.5cm}
-  \begin{blockarray}{c}
-    \vphantom{L_1} \\
-    \begin{block}{(c)}
-      \e_1 \\
-      \e_2 \\
-      \e_3 \\
-      \e_4 \\
-    \end{block}
-  \end{blockarray}
-  \]
+
+\input matmult.tex
+
 % \begin{pmatrix}
@@
 \]
-$\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
+$\bS_{ind}$ is a binary matrix that describes whether a particular branch occurs in the history of a focal species. 
 $\bS \bS^T$ gives the covariance matrix of the phylogeny.
@@
 \end{aligned}
+\label{eq:zmatrix}
 \end{equation}
@@
 
-We generated test data based on (\ref{eq:glmm}) with a single response variable $\bY$ and a single normally distributed predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
+Using the formulation described in (\ref{eq:glmm}--\ref{eq:zmatrix}), we generated test data with a single response variable $\bY$ and a single normally distributed predictor variable $\bm{t}$ for $n$ = 25, 50, and 100 species.
 The response variable $\bY$ is conditionally normally distributed (i.e., $\cal{D}$ is a Gaussian distribution, and $g$ is the identity link function), corresponding to a linear mixed effect model. 
@@
 \bY & = \dist(g^{-1}(\bmu), \phi) \\
-\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}}}) \bm{t} + \besp \\\
-(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\Sigma^2_{\mathrm{phy_{int}}} & \Sigma_{\mathrm{phy_{int-slope}}} \\ 
-\Sigma_{\mathrm{phy_{int-slope}}} & \Sigma^2_{\mathrm{phy_{slope}}}
+\bmu & = (\beta_0 + \bphys{int}) + (\beta_1 + \bphys{slope}) \bm{t} + \besp \\\
+(\bphys{int}, \bphys{slope}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
+\phyv{int} & \phys{int-slope} \\ 
+\phys{int-slope} & \phyv{slope}
 \end{bmatrix} 
@@
 \end{equation}
-The model has two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance $\Sigma^2_{\mathrm{phy}_{int}}$, phylogenetic random slope variance $\Sigma^2_{\mathrm{phy}_{slope}}$ and covariance between phylogenetic random intercept and slope $\Sigma_{\mathrm{phy_{int-slope}}}$) and residual variance ($\sigma_{\epsilon}^2$).
+The model has two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance (\phyv{int}), phylogenetic random slope variance (\phyv{slope}) and covariance between phylogenetic random intercept and slope (\phys{int-slope}) and residual variance ($\sigma_{\epsilon}^2$).
 The covariance between phylogenetic random intercept and slope measures the correlation of phylogenetic effects on the slope ($b_{\mathrm{phy_{slope}}}$) and intercept ($b_{\mathrm{phy_{int}}}$) for each branch of the phylogeny; i.e.
@@
 We extend the simulation model by adding multiple groups where each group has one observation per species. 
-The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction \citep{helmus2007separating}. 
+The multi-group model is a generalization of multiple-site models used in community ecology to model phylogenetic attraction and repulsion \citep{helmus2007separating}. 
 The full multi-group model is as follows: 
@@
 \bY & = \dist(g^{-1}(\bmu), \phi) \\
-\bmu & = (\bf{\beta_0 + b_{\mathrm{phy_{int}}} + b_{\mathrm{sp_{int}}} + b_{\mathrm{group}}}) + (\bf{\beta_1 + b_{\mathrm{phy_{slope}}} + b_{\mathrm{sp_{slope}}}}) \bm{t} + \bf{b_{\mathrm{sp:group}}} + \besp \\
-(\bf{b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
-\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int-slope}}} \\ 
-\sigma_{\mathrm{phy_{int-slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
+\bmu & = (\beta_0 + \bphys{int} + \bsps{int} + {\bf b}_{\mathrm{group}}}) + (\beta_1 + \bphys{slope} + \bsps{slope}) \bm{t} + {\bf b}_{\mathrm{sp:group}}} + \besp \\
+(\bphys{int}, \bphys{slope}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
+\phyv{int} & \phys{int-slope} \\ 
+\phys{int-slope} & \phyv{slope}
 \end{bmatrix}
@@
 \bf{b_{\mathrm{group}}} & \sim \textrm{MVN} ( 0 , \sigma^2_{\mathrm{group}}) \\
-\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \sigma^2_{phy}) \\
+\bf{b_{\mathrm{sp:group}}} & \sim \textrm{MVN} (0, \mathrm{I}_{\mathrm{group}} \otimes \sigma^2_{\mathrm{phy}}) \\
 \besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2),
@@
 
-Compared to the single--group model, the multi-group simulation model has five additional random effect parameters (predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) random effect of species variance and their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$), random intercept of group ($\bf{b_{\mathrm{group}}}$) and random intercept of species-group interaction ($\bf{b_{\mathrm{sp:group}}}$)).
+\bmb{ML: please check notation; in particular, is $\sigma^2_{\mathrm{phy}}$ correct??}
+Compared to the single--group model, the multi-group simulation model has five additional random effect parameters: predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) among-species variances;  their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$);  among-group variation in the intercept ($\sigma^2_{\mathrm{group}}$); and variation among species-group combinations in the intercept ($\sigma^2_{\mathrm{phy}}$).
 Because each species has multiple observations we can distinguish variation among species from residual variation, and thus we can include predictor-level and intercept-level random effects of species.
-Variance in the intercept of species--group interactions ($\bf{b_{\mathrm{sp:group}}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
+Variance in the intercept of species--group interactions ($\sigma^2_{\mathrm{phy}}$) describes whether the species within a group have more similar responses on average than expected by chance, equivalent to phylogenetic attraction \citep{helmus2007separating}.
 
@@
 Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used PR models; it fits a linear model with a covariance structure that assumes an evolutionary process on the tree (typically BM, but other processes can be used) instead of treating the residual error for each species as independent.
-Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) are extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
+Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in the \pkg{phylolm} package) extend PGLS by allowing for both phylogenetic and residual variation, as well as non-Gaussian response variables.
 Both \code{gls} and \code{phylolm} can model non-Brownian evolutionary processes and different correlation structures (e.g., Pagel's $\lambda$ or Blomberg's $K$), but we restrict our PGLS fits to the simple BM correlation.
@@
 However, MCMC is usually much more computationally expensive for GLMMs than platforms using deterministic optimization.
-\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM, and recently \pkg{brms}, which uses a more computationally efficient MCMC technique.
+\pkg{MCMCglmm} \citep{hadfield2010general} is the most widely used Bayesian phylogenetic GLMM;  the more recent\pkg{brms} package is extremely flexible and uses Hamilton Monte Carlo, which is often more computationally efficient (although below we find that \pkg{MCMCglmm} is faster in this application).
 
@@
 
-Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) and then simulated the responses for each tree (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
+Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size ($n$ = 25, 50, 100 and an additional $n$ = 500 for the multi-group model) and then simulated the responses for each tree (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}).
 Each realization was fitted using all model variants. 
@@
 For Bayesian fits, we evaluate realizations with Gelman-Rubin statistic $<1.1$.
-Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}), we additionally restricted results to fits with  effective sample size $>1000$ for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) \citep{vehtari2019rank}.
+Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}, we additionally restricted results to fits with  effective sample size $>1000$ for the fixed effect parameters  \citep[$\beta_{0}$ and $\beta_{1}$: ][]{vehtari2019rank}.
 For each replicate, we sample two chains starting with 10000 iterations. 
 We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
-We computed 95\% Wald confidence intervals for frequentist methods and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
-We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.
+We computed 95\% Wald confidence intervals for frequentist methods \bmb{check: I would think lme4 would only give profile CIs for RE parameters?} and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
+We also compare computational speed between different platforms.
 
@@
   \caption{Comparison of single group model parameter estimates across different R packages in Table \ref{table:platform}.
-    Total simulations $N=100$ for each category.
+    Total simulations $n=100$ for each category.
 The horizontal line shows the true value of the parameters in the simulation model.
 Models capable of fitting all parameters (\pkg{\plmefour},  \pkg{\pgTMB}, and \pkg{brms}) fit well for all parameters.
+    \bmb{are we setting a boundary at 200? why are some methods pegging out here?}
 %\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)

### 2022-08-09 table1.tex
cleanup, tweaks

\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model} & \textbf{Method} & \textbf{Data} & \textbf{Platform} \\
\hline
\multirow{2}{*}{
  \begin{tabular}[c]{@{}l@{}}
    Generalized Linear \\ Model (GLM)
\end{tabular}} &
Correlated residual &
\begin{tabular}[c]{@{}l@{}} Single observation \\ per species \end{tabular} &
\begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic \end{tabular} \\
\cline{2-4} &
\begin{tabular}[c]{@{}l@{}}Residual \\ + phylogenetic intercept\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Single observation \\ per species \end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $K$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\
  \hline 
  \multirow{2}{*}{
    \begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} &

  \multirow{2}{*}{Random effect}
  & \begin{tabular}[c]{@{}l@{}}Single observation \\per species, \\ Balanced design\end{tabular} &
      \pkg{pez}                                                     \\ \cline{3-4} 
      &  & Unrestricted   &
      \begin{tabular}[c]{@{}l@{}}  \pkg{\plmefour}, \\ \pkg{\pgTMB}, \\ \pkg{phyr} \end{tabular}
      \\
      \hline
         \multirow{2}{*}{Bayesian GLMM}
         & \multirow{2}{*}{Random effect}
         & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4} 
                                                                                                   &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
\end{tabular}
\end{table}

### 2022-08-09 phyloglmm_ms_resubmit.tex
almost-final tweaks

@@
 For Bayesian fits, we evaluate realizations with Gelman-Rubin statistic $<1.1$.
-Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}, we additionally restricted results to fits with  effective sample size $>1000$ for the fixed effect parameters  \citep[$\beta_{0}$ and $\beta_{1}$: ][]{vehtari2019rank}.
+Based on recent concerns about Gelman-Rubin thresholds \citep{vats2021revisiting}, we additionally restricted results to fits with  effective sample size $>1000$ for the fixed effect parameters  \citep[$\beta_{0}$ and $\beta_{1}$: ][]{vehtari2021rank}.
 For each replicate, we sample two chains starting with 10000 iterations. 
@@
   \includegraphics[scale=0.7]{./figure/csstime.pdf}
-  \caption{Comparison of single-group model computational speed.}
+  \caption{Single-group model computational speed.
+Dashed lines indicate scaling relationships: lines parallel to the lower line would indicate linear scaling of computational time with number of species,
+  while those parallel to the upper line would indicate quadratic scaling.}
 \label{ssplot_speed}
@@
   \includegraphics[scale=0.7]{./figure/sscoverage.pdf}
-  \caption{Comparison of coverage probability for fixed effect parameters.
+  \caption{Coverage probability for fixed effect parameters.
 Models matching the simulation model (\pkg{\plmefour}, \pkg{\pgTMB} and \pkg{brms}) have coverage near the nominal value of 0.95.
@@
 Estimates for fixed effect parameters ($\beta_0$ and $\beta_1$) approach nominal coverage as the number of species increases for \pkg{lme4} and \pkg{glmmTMB} but not for other packages.
-\pkg{brms} has higher than nominal coverage (i.e., its confidence intervals are too conservative) because the prior distributions for the simulation parameters are centered at the true values (\citet{li2018fitting}).
+\pkg{brms} has higher than nominal coverage for the slope parameter $\beta_1$ (i.e., its confidence intervals are too conservative) because the prior distributions for the simulation parameters are centered at the true values \citep{li2018fitting}.
 
@@
 PGLM (which lacks the phylogenetic slope parameter) provides reasonably good estimates for the phylogenetic intercept standard deviation parameter ($\sigma_{\mathrm{phy_{int}}}$) but overestimates the residual standard deviation; the estimates for the intercept ($\beta_0$) are slightly overconfident (coverage $\approx$ 90\% with 100 species) and the fixed slope parameter ($\beta_1$) has extremely poor coverage ($<$ 60\%).
-PGLS, which uses only one parameter, combines all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter, resulting in overestimating the phylogenetic intercept variation, over-covering for $\beta_0$, and under-covering for $\beta_1$.
+PGLS, which uses only one parameter, combines all variation (phylogenetic intercept, slope and residual variation) into the phylogenetic intercept parameter; as a result, it overestimates the phylogenetic intercept variation, over-covers for $\beta_0$, and under-covers for $\beta_1$.
 
@@
   \includegraphics[scale=0.7]{./figure/msplot.pdf}
-  \caption{Comparison of multi-group model parameter estimates.
+  \caption{Multi-group model parameter estimates.
 The horizontal line shows the true value of the parameters in the simulation model.
 Models capable of fitting all parameters (\pkg{\plmefour} and \pkg{\pgTMB}) fit well for all parameters.
-\pkg{pez} and \pkg{phyr} estimates for n = 500 are missing because the models did not converge within 30 minutes.
+\pkg{pez} and \pkg{phyr} estimates for $n$ = 500 are missing because the models did not converge within 30 minutes.
   }
@@
   \includegraphics[scale=0.7]{./figure/mstime.pdf}
-  \caption{Comparison of multi-group model computational speed.}
+  \caption{Multi-group model computational speed.}
   \label{msplot_time}
@@
 
-We have simulated relatively complex models containing phylogenetic variation in species intercepts and slopes, as well as within-species variation.
+We have simulated complex models containing phylogenetic variation in species intercepts and slopes, as well as within-species variation.
 Comparing our fits with simple platforms for phylogenetic regression that cannot handle these complexities may seem unfair; nevertheless, our models are certainly less complex than evolutionary processes occurring in nature.
@@
 not just in phylogenetic studies.
-While information-theoretic or stepwise selection methods are often used to decide the complexity of models in ecology and evolution \cite{matuschek2017balancing,darriba2020modeltest}, sufficiently complex models may simply be too complex to fit without some form of regularization
-\cite{uriarte2009preaching}, and data-driven model selection may affect inference in unexpected and unwanted ways \cite{hurvich1990impact,morin2020is}.
+While information-theoretic or stepwise selection methods are often used to decide the complexity of models in ecology and evolution \citep{matuschek2017balancing,darriba2020modeltest}, sufficiently complex models may be impossible to fit without some form of regularization
+\citep{uriarte2009preaching}, and data-driven model selection may affect inference in unexpected and unwanted ways \citep{hurvich1990impact,morin2020is}.
 
@@
 Random-slopes models require appropriate observational or experimental designs (i.e., multiple measurements of traits and responses within each group) and generally require more data overall, but they are relevant over a wide range of scenarios \citep{schielzeth2008conclusions, cleasby2015quantifying,ord2010adaptation}.
-Neglecting random slopes can lead to biased fixed effect estimates with inadequate coverage and inflated type~I errors \citep{schielzeth2008conclusions}.
+Neglecting random slopes can lead to biased fixed-effect estimates with inadequate coverage and inflated type~I errors \citep{schielzeth2008conclusions}.
 
 Nevertheless, it is impossible to account for all possible complexities.
-The best model --- whether using phylogenetic random effects, simple grouping, or both --- depends on the experimental design and whether there are enough data to separate different levels of variation, which can be strongly confounded.
+The best model --- whether using phylogenetic random effects, simple grouping, or both --- depends on the experimental design and whether there is enough data to separate different levels of variation, which can be strongly confounded.
 If multiple observations are available per species, then simple methods like Pagel's $\lambda$ will confound tip variation with residual variation \citep{boettiger2013is}.
-Multiple observations can be collapsed to a single value (such as the mean) per species, with analyses weighting each species by its number of observations.
+Multiple observations can be collapsed to a single value (such as the mean) per species, with analyses weighting each species by its number of observations \citep{murtaugh2007simplicity}.
 Alternatively, if the within-species variance is of interest, a phylogenetic mixed model can separate tip variation from within-species variation and measurement error \citep{kostikova2016bridging}.
 More generally, phylogenetic mixed models can be simplified to ordinary mixed models, at the cost of taxonomic detail, by simplifying the phylogenetic tree to a strictly hierarchical set of nested higher-level taxa \citep{bunnefeld2012island}).
-Users should be aware of two essential questions when fitting random-slope models: do they have enough information to practically estimate the random slopes, and what are the potential costs of ignoring them \citep{schielzeth2008conclusions}?
+Users should be aware of two essential questions when fitting random-slope models: do they have enough information to reliably estimate the random slopes, and what are the potential costs of ignoring them \citep{schielzeth2008conclusions}?
 % However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.
@@
 The range of PRs presented here can incorporate many levels and types of phylogenetic variation.
-Of course, there are further biological complexities we have neglected, such as multivariate responses; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck model \citep{butler2004phylogenetic}); and variable-rate models, which allow evolutionary rates to vary across the phylogeny.
+Of course, we have neglected further biological complexities, such as multivariate responses; non-Brownian evolutionary processes such as the Ornstein-Uhlenbeck model \citep{butler2004phylogenetic}); and variable-rate models, which allow evolutionary rates to vary across the phylogeny.
 While it cannot easily incorporate these complexities, the approach here does offer an efficient way to analyze a wide range of evolutionary scenarios.
@@
 We would like to thank Jonathan Dushoff for thoughtful comments.
-This study was funded by NSERC Discovery Grant.
+This study was funded by an NSERC Discovery Grant.
 

### 2022-08-09 plot.R

@@ -53,9 +53,9 @@ ssdat <- (ssdat_raw
                            )
            , Platform = factor(trans_platform(platform), levels=trans_platform(sub_pkgs))
            , sdtype = factor(sdtype, levels=c("phylo_int","phylo_cor","phylo_X","resid")
-                           , labels=c(expression(paste("Phylogenetic random intercept ", Sigma[phy[int]]))
+                           , labels=c(expression(paste("Phylogenetic random intercept ", sigma[phy[int]]))
                                     , expression(paste("Phylogenetic random intercept-slope correlation ", rho[phy[int-slope]]))
-                                    , expression(paste("Phylogenetic random slope ", Sigma[phy[slope]]))
+                                    , expression(paste("Phylogenetic random slope ", sigma[phy[slope]]))
                                     , expression(paste("Residual ", sigma[epsilon])))
                              )
              )
@@ -271,13 +271,13 @@ lev_vec <- c("phylo_int","phylo_cor","phylo_X"
            , "species_int", "species_cor", "species_X"
            , "phylo_interaction", "site_int", "resid")
 
-lab_vec <- c(expression(atop("Phylogenetic random intercept ",Sigma[phy[int]]))
+lab_vec <- c(expression(atop("Phylogenetic random intercept ",sigma[phy[int]]))
            , expression(atop("Phylo random intercept-slope corr",rho[phy[int-slope]]))
-           , expression(atop("Phylogenetic random slope ",Sigma[phy[slope]]))
+           , expression(atop("Phylogenetic random slope ",sigma[phy[slope]]))
            , expression(atop("Species random intercept ",sigma[sp[int]]))
            , expression(atop("Species random intercept-slope corr",rho[sp[int-slope]]))
            , expression(atop("Species random slope ",sigma[sp[slope]]))
-           , expression(atop("Phylo random species-group interact",Sigma[phy[sp:group]]))
+           , expression(atop("Phylo random species-group interact",sigma[phy[sp:group]]))
            , expression(atop("Random group ",sigma[group]))
            , expression(atop("Residual ",sigma[epsilon])))
 

### 2022-08-09 response.md
more TODO etc

@@
 
-**FIXME: make sure we state that the simulations follow the estimation model exactly, so that all the formulas for simulation are the same as those for estimation (and maybe emphasize that the Z/Sigma setup generalizes standardized phylogenetic regression to all of the more complex cases).**
+**We have to admit that we are confused by this comment.
+We tried to clearly indicate that the simulations follow precisely the same structures as we are using for estimation, which is described in detail in the first section ...**

### 2022-08-09 phyloglmm_ms_resubmit.tex
more TODO etc

@@
 
-\bmb{ML: please check notation; in particular, is $\sigma^2_{\mathrm{phy}}$ correct??}
+\bmb{ML: please check notation; in particular, is $\sigma^2_{\mathrm{phy}}$ correct for the sp-group variance??}
 Compared to the single--group model, the multi-group simulation model has five additional random effect parameters: predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) among-species variances;  their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$);  among-group variation in the intercept ($\sigma^2_{\mathrm{group}}$); and variation among species-group combinations in the intercept ($\sigma^2_{\mathrm{phy}}$).

### 2022-08-09 plot.R
misc TODO/fix multi-time plot

@@ -43,16 +43,16 @@ trans_platform <- function(platform) {
 
 
 ssdat <- (ssdat_raw
-  %>% as_tibble()
-  %>% separate(model,c("platform", "sites", "size", "seed","saveformat"), "[.]")
-  %>% dplyr:::select(time, platform, size, seed, resid, phylo_int, phylo_X, phylo_cor, B0, B1)
-  %>% gather(key = sdtype, value = sd, -c(platform, size, seed, time, B0, B1))
-  %>% left_join(.,sspar_df, by="sdtype")
-  %>% mutate(size = factor(size
-                         , levels=c("small","med","large","xlarge"), labels=c("25","50","100","500")
-                           )
-           , Platform = factor(trans_platform(platform), levels=trans_platform(sub_pkgs))
-           , sdtype = factor(sdtype, levels=c("phylo_int","phylo_cor","phylo_X","resid")
+    %>% as_tibble()
+    %>% separate(model,c("platform", "sites", "size", "seed","saveformat"), "[.]")
+    %>% dplyr:::select(time, platform, size, seed, resid, phylo_int, phylo_X, phylo_cor, B0, B1)
+    %>% gather(key = sdtype, value = sd, -c(platform, size, seed, time, B0, B1))
+    %>% left_join(.,sspar_df, by="sdtype")
+    %>% mutate(size = factor(size
+                           , levels=c("small","med","large","xlarge"), labels=c("25","50","100","500")
+                             )
+             , Platform = factor(trans_platform(platform), levels=trans_platform(sub_pkgs))
+             , sdtype = factor(sdtype, levels=c("phylo_int","phylo_cor","phylo_X","resid")
                            , labels=c(expression(paste("Phylogenetic random intercept ", sigma[phy[int]]))
                                     , expression(paste("Phylogenetic random intercept-slope correlation ", rho[phy[int-slope]]))
                                     , expression(paste("Phylogenetic random slope ", sigma[phy[slope]]))
@@ -64,9 +64,11 @@ ssdat <- (ssdat_raw
 
 colvec_trans <- setNames(colvec, trans_platform(names(colvec)))
 
+ssdat |> group_by(Platform) |> summarise(bad = mean(sd > 200, na.rm = TRUE))
+mean(ssdat$sd > 200, na.rm = TRUE)
[…]

### 2022-08-11 phyloglmm_ms_resubmit.tex
misc TODO/fix multi-time plot
MLi tweaks and fixing some technical problems

--- 2022-08-09 phyloglmm_ms_resubmit.tex
@@
 
-\bmb{ML: please check notation; in particular, is $\sigma^2_{\mathrm{phy}}$ correct??}
 Compared to the single--group model, the multi-group simulation model has five additional random effect parameters: predictor-level ($\sigma^2_{\mathrm{sp_{slope}}}$) and intercept-level ($\sigma^2_{\mathrm{sp_{int}}}$) among-species variances;  their covariance ($\sigma_{\mathrm{sp_{int-slope}}}$);  among-group variation in the intercept ($\sigma^2_{\mathrm{group}}$); and variation among species-group combinations in the intercept ($\sigma^2_{\mathrm{phy}}$).
@@
 We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
-We computed 95\% Wald confidence intervals for frequentist methods \bmb{check: I would think lme4 would only give profile CIs for RE parameters?} and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
+We computed 95\% Wald confidence intervals for frequentist methods and quantile-based intervals for Bayesian methods, then computed coverage --- the proportion of simulations in which the computed confidence intervals include the true values of parameters --- to assess the quality of the confidence intervals.
 We also compare computational speed between different platforms.
@@
 Models capable of fitting all parameters (\pkg{\plmefour},  \pkg{\pgTMB}, and \pkg{brms}) fit well for all parameters.
-    \bmb{are we setting a boundary at 200? why are some methods pegging out here?}
+Values above 200 (very common for \pkg{gls} and \pkg{phylolm}, less than 1\% of results otherwise) are censored to 200.
     %\bmb{consider tweaking aspect ratio (slightly wider?) Consider including n=25, n=50, n=100 in x-axis tick labels (I don't have all the output data so can't rebuild figs myself ...)
@@
   \includegraphics[scale=0.7]{./figure/mstime.pdf}
-  \caption{Multi-group model computational speed.}
-  \label{msplot_time}
+  \caption{Multi-group model computational speed. Linear and quadratic scaling lines as in Figure~\ref{ssplot_speed}.}
+  \label{msplot_speed}
 \end{figure}

### 2022-08-11 collect.R

@@ -392,7 +392,7 @@ glmmTMBms_results <- function(tt){
     print(i)
     glmmTMB_obj <- readRDS(paste(glmmTMB_path,tt[i],sep=""))
     covobj <- VarCorr(glmmTMB_obj[[1]])[["cond"]]
-    glmmTMBms_df[i,"resid"] <- attr(covobj,"sc")
+    glmmTMBms_df[i,"resid"] <- attr(covobj,"sc")^2
     glmmTMBms_df[i, "phylo_X"] <- covobj[["sp"]]["X","X"]
     glmmTMBms_df[i, "phylo_int"] <- covobj[["sp"]]["(Intercept)","(Intercept)"]
     glmmTMBms_df[i, "phylo_cor"] <- covobj[["sp"]]["(Intercept)","X"]

### 2022-08-11 plot.R

@@ -53,10 +53,10 @@ ssdat <- (ssdat_raw
                              )
              , Platform = factor(trans_platform(platform), levels=trans_platform(sub_pkgs))
              , sdtype = factor(sdtype, levels=c("phylo_int","phylo_cor","phylo_X","resid")
-                           , labels=c(expression(paste("Phylogenetic random intercept ", sigma[phy[int]]))
-                                    , expression(paste("Phylogenetic random intercept-slope correlation ", rho[phy[int-slope]]))
-                                    , expression(paste("Phylogenetic random slope ", sigma[phy[slope]]))
-                                    , expression(paste("Residual ", sigma[epsilon])))
+                           , labels=c(expression(paste("Phylogenetic random intercept ", sigma[phy[int]]^2))
+                                    , expression(paste("Phylogenetic random intercept-slope correlation ", sigma[phy[int-slope]]))
+                                    , expression(paste("Phylogenetic random slope ", sigma[phy[slope]]^2))
+                                    , expression(paste("Residual ", sigma[epsilon]^2)))
                              )
              )
 	# %>% filter(platform != "brms")
@@ -274,15 +274,15 @@ lev_vec <- c("phylo_int","phylo_cor","phylo_X"
            , "species_int", "species_cor", "species_X"
            , "phylo_interaction", "site_int", "resid")
 
-lab_vec <- c(expression(atop("Phylogenetic random intercept ",sigma[phy[int]]))
-           , expression(atop("Phylo random intercept-slope corr",rho[phy[int-slope]]))
-           , expression(atop("Phylogenetic random slope ",sigma[phy[slope]]))
-           , expression(atop("Species random intercept ",sigma[sp[int]]))
-           , expression(atop("Species random intercept-slope corr",rho[sp[int-slope]]))
-           , expression(atop("Species random slope ",sigma[sp[slope]]))
-           , expression(atop("Phylo random species-group interact",sigma[phy[sp:group]]))
-           , expression(atop("Random group ",sigma[group]))
-           , expression(atop("Residual ",sigma[epsilon])))
+lab_vec <- c(expression(atop("Phylogenetic random intercept ",sigma[phy[int]]^2))
+           , expression(atop("Phylo random intercept-slope cov",sigma[phy[int-slope]]))
+           , expression(atop("Phylogenetic random slope ",sigma[phy[slope]]^2))
[…]

### 2023-10-24 phyloglmm_ms_resubmit.tex
latex tweaks

@@
 \newcommand{\phys}[1]{\ensuremath{\phyvar_{\mathrm{phy}_\mathrm{#1}}}}
-\newcommand{\bphys}[1]{\ensuremath{{\bf b}_{\mathrm{phy}_{\mathrm{#1}}}}}
-\newcommand{\bsps}[1]{\ensuremath{{\bf b}_{\mathrm{sp}_{\mathrm{#1}}}}}
+\newcommand{\bphys}[1]{\ensuremath{{\mathbf b}_{\mathrm{phy}_{\mathrm{#1}}}}}
+\newcommand{\bsps}[1]{\ensuremath{{\mathbf b}_{\mathrm{sp}_{\mathrm{#1}}}}}
+\newcommand{\bgrp}[1]{{{\mathbf b}_{\mathrm{#1}}}}
 % \pagenumbering{gobble}
@@
 The full multi-group model is as follows:
+
 \begin{equation}
@@
 \bY & = \dist(g^{-1}(\bmu), \phi) \\
-\bmu & = (\beta_0 + \bphys{int} + \bsps{int} + {\bf b}_{\mathrm{group}}}) + (\beta_1 + \bphys{slope} + \bsps{slope}) \bm{t} + {\bf b}_{\mathrm{sp:group}}} + \besp \\
+\bmu & = (\beta_0 + \bphys{int} + \bsps{int} + \bgrp{group}) + (\beta_1 + \bphys{slope} + \bsps{slope}) \bm{t} + \bgrp{sp:group} + \besp \\
 (\bphys{int}, \bphys{slope}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
